### 2020-04-21 dev/fake_data.Rmd
WIP fake data simulation

---
title: "Tests and development of fake data simulations"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
breathing_data <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>% 
  select(-Remdesivir_Last_Day) %>%
  pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
  filter(!is.na(breathing)) %>%
  transmute(
    patient = Patient,
    breathing = if_else(breathing %in% c("LoO2", "HiO2"), "Oxygen", breathing),
    day = as.integer(day)) %>%
  group_by(patient) %>% 
  mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "Oxygen", "NIPPV", "MV", "ECMO")])], breathing)) %>%
  ungroup() %>%                           
  mutate(    
    breathing = factor(breathing, levels = c("Discharge", "AA", "Oxygen", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
  )

if(any(is.na(breathing_data$breathing)) || any(is.na(breathing_data$day))){
  stop("Error")
}

breathing_summary <- breathing_data %>% group_by(patient) %>% 
  summarise(worst_condition = max(breathing), 
            first_day_icu = as.integer(min(c(1e4, day[breathing >= "MV" & breathing < "Death"]))),
            last_day_icu = as.integer(max(c(-1, day[breathing >= "MV" & breathing < "Death"])))
            ) %>%
  mutate(first_day_icu = if_else(first_day_icu > 1e3, NA_integer_, first_day_icu),
         last_day_icu = if_else(last_day_icu < 0, NA_integer_, last_day_icu))

patient_data <- data_grein %>% group_by(patient) %>%
  top_n(1, day) %>%
  inner_join(breathing_summary, by = c("patient" = "patient")) %>%
  group_by(patient) %>%
  mutate(outcome = case_when(breathing == "Discharge" ~ "Dischared",
                             breathing == "Death" ~ "Death",
                             runif(n()) < 0.01 ~ "Transferred",
                             TRUE ~ "Hospitalized"),
         site = sample(c("A","B","C"), size = n(), replace = TRUE, prob = c(0.7,0.2,0.1)),
         sex = if_else(runif(n()) < 0.5, "M", "F" ),
         days_from_symptom_onset = rnbinom(n(), mu = 4, size = 1),
         myocardial_infarction = rpois(n(), 0.05 + (site != "A") * 0.08),
         hypertension_drugs = rnbinom(n(), mu = 0.1 + (site == "B") * 0.6 , size = 0.3),
         smoking = runif(n()) < 0.4 + (site == "C") * 0.15,
         bmi = rnorm(n(), 28, 4) + case_when(site == "A" ~ 0, site == "B" ~ 2.7, site == "C" ~ -0.5) ,
         took_hcq = runif(n()) < 0.6,
         took_az = took_hcq & runif(n()) < 0.8,
         took_kaletra = !took_hcq & site == "B" & runif(n()) < 0.2,
         first_day_antivirals = if_else(took_az | took_kaletra, as.integer(rdunif(n(), a = 0, b = 3)), NA_integer_),
         took_tocilizumab = worst_condition >= "MV" & site == "A" & runif(n()) < 0.1,
         first_day_tocilizumab = first_day_icu + as.integer(rdunif(n(), a = 0, b = 1))
  ) %>% 
  ungroup() %>%
  select(-breathing, )

breathing_data <- breathing_data %>% filter(!(breathing %in% c("Discharge", "Death"))) %>%
  mutate(breathing = droplevels(breathing))

marker_data_list <- list()
next_id <- 1
for(p in patient_data$patient) {
  breathing_p <- breathing_data %>% filter(patient == p)
  patient_p <- patient_data %>% filter(patient == p)

  #Fit a spline to breathing status
  sp <- smooth.spline(breathing_p$day, as.integer(breathing_p$breathing), nknots = 6)

  oxygen_days <- breathing_p %>% filter(breathing == "Oxygen") %>% pull(day)
  if(length(oxygen_days) > 0) {
    oxygen_flow <- round( (runif(1, 0, 3) + runif(length(oxygen_days)) + predict(sp, oxygen_days)$y - 3) * 2 ) / 2 + 2
    marker_data_list[[next_id]] <- data.frame(patient = p, day = oxygen_days, marker = "oxygen_flow", value = oxygen_flow)
    next_id <- next_id + 1
  }

  if(patient_p$took_hcq) {
    hcq_days <- patient_p$first_day_antivirals:min(patient_p$first_day_antivirals + 5, max(breathing_p$day))
    if(patient_p$site == "B") {
      hcq_dose <- rep(400, length(hcq_days))
      hcq_dose[1] <- 800
    } else {
      hcq_dose <- rep(200, length(hcq_days))
      hcq_dose[1] <- 400
    }
    marker_data_list[[next_id]] <- data.frame(patient = p, day = hcq_days, marker = "hcq", value = hcq_dose,  censored = "none")
    next_id <- next_id + 1

    if(patient_p$took_az) {
      az_dose <- rep(250, length(hcq_days))
      az_dose[1] <- 500                     
      marker_data_list[[next_id]] <- data.frame(patient = p, day = hcq_days, marker = "az", value = az_dose,  censored = "none")
      next_id <- next_id + 1
    }
  }
  if(patient_p$took_kaletra) {
    kaletra_max_days <- rdunif(1, a = 4, b = 8)
    kaletra_days <- patient_p$first_day_antivirals:min(patient_p$first_day_antivirals + kaletra_max_days, max(breathing_p$day))
    if(patient_p$took_kaletra) {
      marker_data_list[[next_id]] <- data.frame(patient = p, day = kaletra_days, marker = "kaletra", value = 400,  censored = "none")
      next_id <- next_id + 1
    }
  }
  if(patient_p$took_tocilizumab) {
      marker_data_list[[next_id]] <- data.frame(patient = p, day = patient_p$first_day_tocilizumab, marker = "tocilizumab", value = if_else(site == "C", 6, 7),  censored = "none")
      next_id <- next_id + 1
  }

  pcr_days <- c(rbinom(1, size = 2, prob = 0.3), which(runif(max(breathing_p$day) - 5) < 0.1) + 2) 
  pcr_positive <- runif(pcr_days) < 0.93 
  if(site == "A") {
    pcr_value <- 35 - (((predict(sp, pcr_days[pcr_positive] - 2)$y / 5) * 21) + 12 + rnorm(sum(pcr_positive), sd = 9))
    pcr_value[pcr_value < 8] <- runif(length(pcr_value < 8), 8, 10)
    pcr_value[pcr_value > 35] <- runif(length(pcr_value > 35), 33, 35)

    marker_data_list[[next_id]] <- data.frame(patient = p, day = pcr_days, marker = "pcr_value", value = pcr_value,  censored = "none")
      next_id <- next_id + 1

  }

  if(patient_p$outcome == "Discharged") {
    first_negative <- rdunif(1, max(pcr_days) + 1, max(breathing_p$day) - 2)
    pcr_days <- c(pcr_days, first_negative, first_negative + 1 + (runif(1) < 0.4))
    pcr_positive <- c(pcr_positive, FALSE, FALSE)
  }

  marker_data_list[[next_id]] <- data.frame(patient = p, day = pcr_positive, marker = "pcr_positive", value = pcr_positive,  censored = "none")
      next_id <- next_id + 1

}

marker_data <- do.call(rbind, marker_data_list)
marker_data_wide

```

### 2020-04-21 R/fake_data.R

@@ -1,0 +1,1 @@
+feka_data

### 2020-04-22 dev/fake_data.Rmd
First working version of fake data generation and data documentation

@@
 library(tidyverse)
+library(cowplot)
+theme_set(theme_cowplot())
 ```
 
-Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.
-
-When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
-
-The preview shows you a rendered HTML copy of the contents of the editor.
-Consequently, unlike *Knit*, *Preview* does not run any R code chunks.
-Instead, the output of the chunk when it was last run in the editor is displayed.
+```{r}
+fake_data <- fake_data_grein()
+fake_data$breathing_data %>% ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5)
+```
 
 ```{r}
-breathing_data <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>% 
-  select(-Remdesivir_Last_Day) %>%
-  pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
-  filter(!is.na(breathing)) %>%
-  transmute(
-    patient = Patient,
-    breathing = if_else(breathing %in% c("LoO2", "HiO2"), "Oxygen", breathing),
-    day = as.integer(day)) %>%
-  group_by(patient) %>% 
-  mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "Oxygen", "NIPPV", "MV", "ECMO")])], breathing)) %>%
-  ungroup() %>%                           
-  mutate(    
-    breathing = factor(breathing, levels = c("Discharge", "AA", "Oxygen", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
-  )
-
-if(any(is.na(breathing_data$breathing)) || any(is.na(breathing_data$day))){
-  stop("Error")
-}
-
-breathing_summary <- breathing_data %>% group_by(patient) %>% 
-  summarise(worst_condition = max(breathing), 
-            first_day_icu = as.integer(min(c(1e4, day[breathing >= "MV" & breathing < "Death"]))),
-            last_day_icu = as.integer(max(c(-1, day[breathing >= "MV" & breathing < "Death"])))
-            ) %>%
-  mutate(first_day_icu = if_else(first_day_icu > 1e3, NA_integer_, first_day_icu),
-         last_day_icu = if_else(last_day_icu < 0, NA_integer_, last_day_icu))
-
-patient_data <- data_grein %>% group_by(patient) %>%
-  top_n(1, day) %>%
-  inner_join(breathing_summary, by = c("patient" = "patient")) %>%
-  group_by(patient) %>%
-  mutate(outcome = case_when(breathing == "Discharge" ~ "Dischared",
-                             breathing == "Death" ~ "Death",
-                             runif(n()) < 0.01 ~ "Transferred",
-                             TRUE ~ "Hospitalized"),
-         site = sample(c("A","B","C"), size = n(), replace = TRUE, prob = c(0.7,0.2,0.1)),
-         sex = if_else(runif(n()) < 0.5, "M", "F" ),
-         days_from_symptom_onset = rnbinom(n(), mu = 4, size = 1),
-         myocardial_infarction = rpois(n(), 0.05 + (site != "A") * 0.08),
-         hypertension_drugs = rnbinom(n(), mu = 0.1 + (site == "B") * 0.6 , size = 0.3),
-         smoking = runif(n()) < 0.4 + (site == "C") * 0.15,
-         bmi = rnorm(n(), 28, 4) + case_when(site == "A" ~ 0, site == "B" ~ 2.7, site == "C" ~ -0.5) ,
-         took_hcq = runif(n()) < 0.6,
-         took_az = took_hcq & runif(n()) < 0.8,
-         took_kaletra = !took_hcq & site == "B" & runif(n()) < 0.2,
-         first_day_antivirals = if_else(took_az | took_kaletra, as.integer(rdunif(n(), a = 0, b = 3)), NA_integer_),
-         took_tocilizumab = worst_condition >= "MV" & site == "A" & runif(n()) < 0.1,
-         first_day_tocilizumab = first_day_icu + as.integer(rdunif(n(), a = 0, b = 1))
-  ) %>% 
-  ungroup() %>%
-  select(-breathing, )
-
-breathing_data <- breathing_data %>% filter(!(breathing %in% c("Discharge", "Death"))) %>%
-  mutate(breathing = droplevels(breathing))
-
-marker_data_list <- list()
-next_id <- 1
-for(p in patient_data$patient) {
-  breathing_p <- breathing_data %>% filter(patient == p)
-  patient_p <- patient_data %>% filter(patient == p)
-
-  #Fit a spline to breathing status
-  sp <- smooth.spline(breathing_p$day, as.integer(breathing_p$breathing), nknots = 6)
-
-  oxygen_days <- breathing_p %>% filter(breathing == "Oxygen") %>% pull(day)
-  if(length(oxygen_days) > 0) {
-    oxygen_flow <- round( (runif(1, 0, 3) + runif(length(oxygen_days)) + predict(sp, oxygen_days)$y - 3) * 2 ) / 2 + 2
-    marker_data_list[[next_id]] <- data.frame(patient = p, day = oxygen_days, marker = "oxygen_flow", value = oxygen_flow)
-    next_id <- next_id + 1
-  }
-
-  if(patient_p$took_hcq) {
-    hcq_days <- patient_p$first_day_antivirals:min(patient_p$first_day_antivirals + 5, max(breathing_p$day))
-    if(patient_p$site == "B") {
-      hcq_dose <- rep(400, length(hcq_days))
-      hcq_dose[1] <- 800
-    } else {
-      hcq_dose <- rep(200, length(hcq_days))
-      hcq_dose[1] <- 400
-    }
-    marker_data_list[[next_id]] <- data.frame(patient = p, day = hcq_days, marker = "hcq", value = hcq_dose,  censored = "none")
-    next_id <- next_id + 1
-
-    if(patient_p$took_az) {
-      az_dose <- rep(250, length(hcq_days))
-      az_dose[1] <- 500                     
-      marker_data_list[[next_id]] <- data.frame(patient = p, day = hcq_days, marker = "az", value = az_dose,  censored = "none")
-      next_id <- next_id + 1
-    }
-  }
-  if(patient_p$took_kaletra) {
-    kaletra_max_days <- rdunif(1, a = 4, b = 8)
-    kaletra_days <- patient_p$first_day_antivirals:min(patient_p$first_day_antivirals + kaletra_max_days, max(breathing_p$day))
-    if(patient_p$took_kaletra) {
-      marker_data_list[[next_id]] <- data.frame(patient = p, day = kaletra_days, marker = "kaletra", value = 400,  censored = "none")
-      next_id <- next_id + 1
-    }
-  }
-  if(patient_p$took_tocilizumab) {
-      marker_data_list[[next_id]] <- data.frame(patient = p, day = patient_p$first_day_tocilizumab, marker = "tocilizumab", value = if_else(site == "C", 6, 7),  censored = "none")
-      next_id <- next_id + 1
-  }
-
-  pcr_days <- c(rbinom(1, size = 2, prob = 0.3), which(runif(max(breathing_p$day) - 5) < 0.1) + 2) 
-  pcr_positive <- runif(pcr_days) < 0.93 
-  if(site == "A") {
-    pcr_value <- 35 - (((predict(sp, pcr_days[pcr_positive] - 2)$y / 5) * 21) + 12 + rnorm(sum(pcr_positive), sd = 9))
-    pcr_value[pcr_value < 8] <- runif(length(pcr_value < 8), 8, 10)
-    pcr_value[pcr_value > 35] <- runif(length(pcr_value > 35), 33, 35)
-
-    marker_data_list[[next_id]] <- data.frame(patient = p, day = pcr_days, marker = "pcr_value", value = pcr_value,  censored = "none")
-      next_id <- next_id + 1
-
-  }
-
-  if(patient_p$outcome == "Discharged") {
-    first_negative <- rdunif(1, max(pcr_days) + 1, max(breathing_p$day) - 2)
-    pcr_days <- c(pcr_days, first_negative, first_negative + 1 + (runif(1) < 0.4))
-    pcr_positive <- c(pcr_positive, FALSE, FALSE)
-  }
-
-  marker_data_list[[next_id]] <- data.frame(patient = p, day = pcr_positive, marker = "pcr_positive", value = pcr_positive,  censored = "none")
-      next_id <- next_id + 1
-
-}
-
-marker_data <- do.call(rbind, marker_data_list)
-marker_data_wide
-
+fake_data$marker_data %>% filter(!(marker %in% c("hcq","az", "tocilizumab", "kaletra"))) %>% ggplot(aes(x = day, y = value, group = patient)) + geom_line() + facet_wrap(~marker, scales = "free")
 ```

### 2020-04-22 R/fake_data.R

@@ -1,1 +1,219 @@
-feka_data
+#' Generates fake dataset starting from breathing support histories taken from
+#' Gren et al.: https://www.nejm.org/doi/full/10.1056/NEJMoa2007016
+#' @export
+#' @import dplyr
+#' @import tidyr
+#' @importFrom magrittr "%>%"
+fake_data_grein <- function() {
+  breathing_data_orig <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
+    select(-Remdesivir_Last_Day) %>%
+    pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
+    filter(!is.na(breathing)) %>%
+    transmute(
+      patient = Patient,
+      breathing = if_else(breathing %in% c("LoO2", "HiO2"), "Oxygen", breathing),
+      day = as.integer(day)) %>%
+    group_by(patient) %>%
+    mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "Oxygen", "NIPPV", "MV", "ECMO")])], breathing)) %>%
+    ungroup() %>%
+    mutate(
+      breathing = factor(breathing, levels = c("Discharge", "AA", "Oxygen", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
+    )
+
+  #Generate "Lead-in" data
+  breathing_data_list <- list()
+  for(p in unique(breathing_data_orig$patient)) {
+    breathing_p <- breathing_data_orig %>% filter(patient == p)
+    start_value <- breathing_p %>% filter(day == 0) %>% pull(breathing)
+    if(start_value >= "NIPPV") {
+      n_days_before <- 1 + rnbinom(1, mu = (as.integer(start_value) - 2) * 2.5, size = 1)
+      n_steps <- as.integer(start_value) - 3
+      step_times <- sort(purrr::rdunif(n_steps, a = 1, b = n_days_before))
+      breathing_new <- character(n_days_before)
+      start <- 1
+      for(s in 1:n_steps) {
+        if(start <= step_times[s]) {
+          breathing_new[start:(step_times[s])] <- levels(breathing_data_orig$breathing)[s + 1]
+          start <- step_times[s] + 1
+        }
+      }
[…]

### 2020-04-29 dev/hmm_model_devel.Rmd
Simulator and WIP rewrite of the rate-based HMM model

---
title: "HMM Model development"
output: html_notebook
---

```{r setup}
library(rstan)
library(tidyverse)
devtools::load_all()
```

```{r}
model <- stan_model(here::here("hmm.stan"))
```

```{r}
data <- hmm_simulator(10, 20, 2)

improving_data <- data$true$true_improving %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "improving") %>% filter(improving)

data$true$true_base_states %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "breathing") %>% mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
  ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + geom_tile(data = improving_data, width = 1, height = 0.1, fill = "gray") + ggtitle("True - gray lines for improving")

data$observed$observations %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "breathing") %>% mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
  ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + ggtitle("Observed")
```

### 2020-04-29 R/hmm_simulator.R

@@ -1,0 +1,153 @@
+hmm_simulator <- function(N_patients, N_time, N_mid_states) {
+
+  # States and transitions
+
+  N_states_hidden <- 2 + N_mid_states * 2
+  N_states_observed <- 2 + N_mid_states
+
+  s_discharged_hid <- 1
+  s_improving_hid <- 2:(N_mid_states + 1)
+  s_worsening_hid <- (N_mid_states + 2):(N_mid_states * 2 + 1)
+  s_death_hid <- N_states_hidden
+
+  s_discharged_obs <- 1
+  s_mid_obs <- 2:(N_mid_states + 1)
+  s_death_obs <- N_states_observed
+
+  N_rates <- N_mid_states + #worsening to death
+    2 * N_mid_states + #worsening <-> improving
+    2 * (N_mid_states - 1) + #worsen/improve by one
+    1 #discharge
+
+  rates_from <- array(0, N_rates)
+  rates_to <- array(0, N_rates)
+  rates_prior_mean <- numeric(N_rates)
+
+  next_rate <- 1
+  rates_death <- next_rate:(next_rate + N_mid_states - 1)
+  rates_from[rates_death] <- s_worsening_hid
+  rates_to[rates_death] <- s_death_hid
+  rates_prior_mean[rates_death] <- 0.01
+  next_rate <- next_rate + length(rates_death)
+
+  rates_to_improving <- next_rate:(next_rate + N_mid_states - 1)
+  rates_from[rates_to_improving] <- s_worsening_hid
+  rates_to[rates_to_improving] <- s_improving_hid
+  rates_prior_mean[rates_to_improving] <- 0.1
+  next_rate <- next_rate + length(rates_to_improving)
+
+  rates_to_worsening <- next_rate:(next_rate + N_mid_states - 1)
+  rates_from[rates_to_worsening] <- s_improving_hid
+  rates_to[rates_to_worsening] <- s_worsening_hid
+  rates_prior_mean[rates_to_worsening] <- 0.1
+  next_rate <- next_rate + length(rates_to_worsening)
+
+  rates_worsen_one <- next_rate:(next_rate + N_mid_states - 2)
+  rates_from[rates_worsen_one] <- s_worsening_hid[1:(length(s_worsening_hid) - 1)]
+  rates_to[rates_worsen_one] <- s_worsening_hid[2:length(s_worsening_hid)]
+  rates_prior_mean[rates_worsen_one] <- 0.2
+  next_rate <- next_rate + length(rates_worsen_one)
+
[…]

### 2020-04-30 dev/hmm_model_devel.Rmd
First rate HMM version passing SBC

@@
 library(tidyverse)
+library(cowplot)
+theme_set(theme_cowplot())
+
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+
 devtools::load_all()
-```
-
-```{r}
-model <- stan_model(here::here("hmm.stan"))
 ```
@@
 ```
+
+```{r}
+model <- stan_model(here::here("hmm.stan"))
+```
+
+```{r}
+adapt_delta <- 0.8
+
+data <- hmm_simulator(10, 20, 2)
+
+if(inherits(model, "stanmodel")) {
+  fit <- sampling(model, data = data$observed, control = list(adapt_delta = adapt_delta))
+} else {
+  cmdstan_fit <- model$sample(data = data$observed, adapt_delta = adapt_delta)
+  fit <- rstan::read_stan_csv(cmdstan_fit$output_files())
+}
+evaluation_summary(fit, data$true)
+
+```
+
+```{r}
+sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
+
+saveRDS(sbc_res, "sbc.rds")
+
+sbc_res$params %>% filter(grepl("rates", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("sensitivity", param_name)) %>% plot_sbc_params()
+#sbc_res$params %>% filter(grepl("initial", param_name)) %>% plot_sbc_params()
+
+#sbc_res$params %>% plot_sbc_params()
+summarise_sbc_diagnostics(sbc_res)
+sbc_res$diagnostics
+
+```

### 2020-05-07 dev/hmm_model_devel.Rmd
A slightly better observational model

@@
 ```{r}
-data <- hmm_simulator(10, 20, 2)
+data <- hmm_simulator(10, 20, 2, use_noisy_states = FALSE)
 
@@
 
-data <- hmm_simulator(10, 20, 2)
+data <- hmm_simulator(10, 20, 2, use_noisy_states = TRUE)
 
 if(inherits(model, "stanmodel")) {
-  fit <- sampling(model, data = data$observed, control = list(adapt_delta = adapt_delta))
+  fit <- sampling(model, data = data$observed, control = list(adapt_delta = adapt_delta), chains = 1)
 } else {
@@
 ```{r}
-sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
+sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2, use_noisy_states = TRUE) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
 
@@
 ```
+
+```{r}
+library(brms)
+test_data <- data.frame(a = c(1,2,3), b = c(0.5,1.0,1.5), x1 = c("A"), x2 = c(0,1,2))
+f <- cbind(a,b) ~ 0 + (0 + x2 | a | x1)
+make_stancode(f, data = test_data)
+make_standata(f, data = test_data)
+```

### 2020-05-07 R/hmm_simulator.R

@@ -1,7 +1,9 @@
-hmm_simulator <- function(N_patients, N_time, N_mid_states) {
+hmm_simulator <- function(N_patients, N_time, N_mid_states, use_noisy_states = FALSE) {
 
   # States and transitions
-
+  if(N_mid_states < 1) {
+    stop("Need at least one mid state")
+  }
   N_states_hidden <- 2 + N_mid_states * 2
   N_states_observed <- 2 + N_mid_states
 
@@ -75,17 +77,41 @@ hmm_simulator <- function(N_patients, N_time, N_mid_states) {
 
 
   # Observation model
-  sensitivity_low_bound <- 0.9
+  sensitivity_low_bound <- 0.8
   sensitivity <- runif(N_mid_states, min = sensitivity_low_bound, max = 1)
   corresponding_observation <- array(NA_integer_, N_states_hidden)
   corresponding_observation[s_discharged_hid] <- s_discharged_obs
   corresponding_observation[s_worsening_hid] <- s_mid_obs
   corresponding_observation[s_improving_hid] <- s_mid_obs
   corresponding_observation[s_death_hid] <-  s_death_obs
-  N_noisy_states <- N_mid_states
-  noisy_states <- s_mid_obs
-  noisy_state_id <- array(NA_integer_, N_states_observed)
-  noisy_state_id[noisy_states] <- 1:N_noisy_states
+  if(use_noisy_states) {
+    N_noisy_states <- N_mid_states
+    noisy_states <- s_mid_obs
+    noisy_state_id <- array(NA_integer_, N_states_observed)
+    noisy_state_id[noisy_states] <- 1:N_noisy_states
+
+    N_other_observations <- 2
+    noisy_states_other_obs <- array(NA_integer_, c(N_noisy_states, N_other_observations))
+    noisy_states_other_obs[1, ] <- c(noisy_states[1] + 1,noisy_states[1] + 2)
+    noisy_states_other_obs[N_noisy_states, ] <- c(noisy_states[N_noisy_states] - 2,noisy_states[N_noisy_states] - 1)
+    for(ns in 2:(N_noisy_states - 1)) {
+      noisy_states_other_obs[ns, ] <- c(noisy_states[ns] - 1, noisy_states[ns] + 1)
+    }
+
+    other_observations_probs <- array(NA_real_, c(N_noisy_states, N_other_observations))
+    for(ns in 1:N_noisy_states) {
[…]

### 2020-05-07 dev/hmm_model_devel.Rmd
Preparing scaffold for brms predictors

--- 2020-04-30 dev/hmm_model_devel.Rmd
@@
 ```{r}
-data <- hmm_simulator(10, 20, 2)
+data <- hmm_simulator(10, 20, 4, use_noisy_states = TRUE)
 
-improving_data <- data$true$true_improving %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "improving") %>% filter(improving)
+improving_data <- data.frame(day = data$observed$times, patient = data$observed$patients, improving = data$true$true_improving) %>% filter(improving)
 
-data$true$true_base_states %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "breathing") %>% mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
+data.frame(day = data$observed$times, patient = data$observed$patients, breathing = data$true$true_base_states) %>%
+  mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
   ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + geom_tile(data = improving_data, width = 1, height = 0.1, fill = "gray") + ggtitle("True - gray lines for improving")
 
-data$observed$observations %>% as.data.frame() %>% mutate(day = 1:n()) %>% pivot_longer(starts_with("V"), names_to = "patient", values_to = "breathing") %>% mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
+data.frame(day = data$observed$times, patient = data$observed$patients, breathing = data$observed$obs_states) %>%
+  mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
   ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + ggtitle("Observed")
@@
 
-data <- hmm_simulator(10, 20, 2)
+data <- hmm_simulator(10, 20, 2, use_noisy_states = TRUE)
 
@@
 ```{r}
-sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
+sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2, use_noisy_states = FALSE) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
 
@@
 sbc_res$params %>% filter(grepl("sensitivity", param_name)) %>% plot_sbc_params()
-#sbc_res$params %>% filter(grepl("initial", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("other_observations_probs", param_name)) %>% plot_sbc_params()
 

### 2020-05-07 R/hmm_simulator.R

@@ -62,18 +62,29 @@ hmm_simulator <- function(N_patients, N_time, N_mid_states, use_noisy_states = F
   rates_prior_mean[rate_discharge] <- 0.1
 
 
-  rates_prior_alpha <- rep(3, N_rates)
-  rates_prior_beta <- rates_prior_alpha / rates_prior_mean
+  # Predictors
+  N_predictor_sets <- 2
+  N <- N_predictor_sets * N_rates
+  rate_predictors <- array(1:(N_rates * N_predictor_sets), c(N_predictor_sets,N_rates))
 
-  rates <- rgamma(N_rates, rates_prior_alpha, rates_prior_beta)
-
-  rate_matrix <- matrix(0, nrow = N_states_hidden, ncol = N_states_hidden)
-  for(r in 1:N_rates) {
-    rate_matrix[rates_from[r], rates_to[r]] <- rates[r]
-    rate_matrix[rates_from[r], rates_from[r]] <- rate_matrix[rates_from[r], rates_from[r]] - rates[r]
+  intercept_prior_logmean <- array(NA_real_, N)
+  for(ps in 1:N_predictor_sets) {
+    intercept_prior_logmean[rate_predictors[ps,]] <- log(rates_prior_mean)
   }
+  intercept_prior_sd <- array(1, N);
+  mu <- rnorm(N, intercept_prior_logmean, intercept_prior_sd)
 
-  transition_matrix <- expm::expm(rate_matrix)
+  transition_matrices <- array(NA_real_, c(N_predictor_sets,N_states_hidden, ncol = N_states_hidden))
+
+  for(ps in 1:N_predictor_sets) {
+    rate_matrix <- matrix(0, nrow = N_states_hidden, ncol = N_states_hidden)
+    for(r in 1:N_rates) {
+      rate_value <- exp( mu[ rate_predictors[ps, r] ] )
+      rate_matrix[rates_from[r], rates_to[r]] <- rate_value
+      rate_matrix[rates_from[r], rates_from[r]] <- rate_matrix[rates_from[r], rates_from[r]] - rate_value
+    }
+    transition_matrices[ps, ,] <- expm::expm(rate_matrix)
+  }
 
 
   # Observation model
@@ -90,12 +101,21 @@ hmm_simulator <- function(N_patients, N_time, N_mid_states, use_noisy_states = F
     noisy_state_id <- array(NA_integer_, N_states_observed)
[…]

### 2020-05-07 dev/hmm_model_devel.Rmd
WIP using brms to have predictors for HMM

@@
 
-improving_data <- data.frame(day = data$observed$times, patient = data$observed$patients, improving = data$true$true_improving) %>% filter(improving)
+improving_data <- data.frame(day = data$observed$times, patient = data$observed$series, improving = data$true$true_improving) %>% filter(improving)
 
-data.frame(day = data$observed$times, patient = data$observed$patients, breathing = data$true$true_base_states) %>%
+data.frame(day = data$observed$times, patient = data$observed$series, breathing = data$true$true_base_states) %>%
   mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
@@
 
-data.frame(day = data$observed$times, patient = data$observed$patients, breathing = data$observed$obs_states) %>%
+data.frame(day = data$observed$times, patient = data$observed$series, breathing = data$observed$obs_states) %>%
   mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
@@
 
-sbc_res$params %>% filter(grepl("rates", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("mu", param_name)) %>% plot_sbc_params()
 sbc_res$params %>% filter(grepl("sensitivity", param_name)) %>% plot_sbc_params()
@@
 library(brms)
+```
+
+```{r}
+data <- hmm_simulator(10, 20, 2, use_noisy_states = TRUE)
+cc <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
+
+model2 <- stan_model(model_code = cc)
+
+prep_data <- make_data_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
+
+all_data <- c(make_standata( make_brms_formula_hmm(data$observed_structured$formula), data = prep_data$brmsdata),
+              prep_data$standata)
+
+sampling(model2, data = all_data)
+
+```
+
+```{r}
+
 test_data <- data.frame(a = c(1,2,3), b = c(0.5,1.0,1.5), x1 = c("A"), x2 = c(0,1,2))
-f <- cbind(a,b) ~ 0 + (0 + x2 | a | x1)
-make_stancode(f, data = test_data)
-make_standata(f, data = test_data)
+# f <- cbind(a,b) ~ 0 + (0 + x2 | a | x1)
+# make_stancode(f, data = test_data)
+# make_standata(f, data = test_data)
+
+stan_llh.rate_hmm <- function( ... ) {
+  ""
+}
+
+.family_rate_hmm <- function() {
+
+}
+
+ff <-
+
+make_stancode(a ~ 0 + x2, family = ff, data = test_data)
+
 ```

### 2020-05-07 R/make_stancode_hmm.R

@@ -1,0 +1,247 @@
+make_brms_formula_hmm <- function(formula) {
+  if(!is.null(brms:::lhs(formula))) {
+    stop("Formula needs to be one-sided")
+  }
+  update.formula(formula, .dummy ~ .)
+}
+
+make_stancode_hmm <- function(formula, serie_data, rate_data, state_data, prior = NULL) {
+  formula <- make_brms_formula_hmm(formula)
+  data <- make_data_hmm(formula, serie_data, rate_data, state_data)
+  brms::make_stancode(formula, family = rate_hmm_family,
+                      data = data$brmsdata,
+                      stanvars = rate_hmm_stanvars(data$standata))
+}
+
+rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars = "mu",
+                           lb = NA, ub = NA, type = "real", vars = NULL, specials = NULL,
+                           log_lik = NULL, posterior_predict = NULL, posterior_epred = NULL),
+                      class = c("rate_hmm", "brmsfamily"))
+
+.family_rate_hmm <- function() {
+
+}
+
+rate_hmm_stanvars <- function(standata) {
+
+  brms::stanvar(scode = rate_hmm_functions_code, block = "functions") +
+  brms::stanvar(x = standata, scode = rate_hmm_data_code, block = "data") +
+  brms::stanvar(scode = rate_hmm_tdata_code, block = "tdata") +
+  brms::stanvar(scode = rate_hmm_parameters_code, block = "parameters")
+}
+
+make_data_hmm <- function(formula, serie_data, rate_data, state_data) {
+
+  serie_data <- serie_data %>% mutate(.predictor_set = 1:n())
+  rate_data <- rate_data %>% mutate(.rate_id = 1:n())
+  brmsdata <- crossing(serie_data, rate_data) %>% mutate(.dummy = 0)
+
+
+
+  N_states_hidden <- length(unique(c(rate_data$.from, rate_data$.to)))
+
+  N_rates = nrow(rate_data)
+
+  state_data <- state_data %>% arrange(.id)
+  if(!identical(state_data$.id, 1:nrow(state_data))) {
+    stop("State ids need to be consecutive and not duplicated")
+  }
+  #TODO allow for noisy states
+  N_states_observed <- length(unique(state_data$.corresponding_obs))
+
+  N_series <- max(serie_data$.serie)
[…]

### 2020-05-14 dev/hmm_model_devel.Rmd
First working version of using brms predictors

@@
 ```{r}
-data <- hmm_simulator(10, 20, 2, use_noisy_states = TRUE)
-cc <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
+data <- hmm_simulator(5, 10, 2, use_noisy_states = FALSE)
+cc <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, prior= data$observed_structured$prior)
 
@@
 
-all_data <- c(make_standata( make_brms_formula_hmm(data$observed_structured$formula), data = prep_data$brmsdata),
-              prep_data$standata)
+all_data <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
 
-sampling(model2, data = all_data)
-
+fit_raw <- sampling(model2, data = all_data)
+#TODO - compactification is necessary!
 ```
@@
 test_data <- data.frame(a = c(1,2,3), b = c(0.5,1.0,1.5), x1 = c("A"), x2 = c(0,1,2))
-# f <- cbind(a,b) ~ 0 + (0 + x2 | a | x1)
+f <- a ~ 0 + (0 + x2 | a | x1)
 # make_stancode(f, data = test_data)
-# make_standata(f, data = test_data)
+#brms::make_standata(f, data = test_data)
+xx <- brms::brmsformula(f)
+
+xx$resp
 

### 2020-05-14 R/hmm_simulator.R

@@ -237,6 +237,7 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
         .from = rates_from,
         .to = rates_to
       ),
+      prior = brms::set_prior("normal(0,1)") + brms::set_prior("normal(0,1)", "Intercept"),
       state_data = data.frame(.id = 1:N_states_hidden,
                               .corresponding_obs = corresponding_observation,
                               .initial_prob = initial_states_prob)

### 2020-05-14 dev/hmm_model_devel.Rmd
Including noisy states

@@
 ```{r}
-model <- stan_model(here::here("hmm.stan"))
-```
+adapt_delta <- 0.9
 
-```{r}
-adapt_delta <- 0.8
+data <- hmm_simulator(30, 20, 2, use_noisy_states = TRUE)
 
-data <- hmm_simulator(10, 20, 2, use_noisy_states = TRUE)
+single_code <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, prior= data$observed_structured$prior)
 
-if(inherits(model, "stanmodel")) {
-  fit <- sampling(model, data = data$observed, control = list(adapt_delta = adapt_delta))
+model_single <- stan_model(model_code = single_code)
+
+prepdata <- make_data_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
+
+standata <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
+
+if(inherits(model_single, "stanmodel")) {
+  fit <- sampling(model_single, data = standata, control = list(adapt_delta = adapt_delta))
 } else {
-  cmdstan_fit <- model$sample(data = data$observed, adapt_delta = adapt_delta)
+  cmdstan_fit <- model_single$sample(data = standata, adapt_delta = adapt_delta)
   fit <- rstan::read_stan_csv(cmdstan_fit$output_files())
@@
 ```{r}
-sbc_res <- sbc(model, generator = function() {hmm_simulator(10, 20, 2, use_noisy_states = FALSE) }, N_steps = 100, control = list(adapt_delta = adapt_delta))
+noisy_states_in_sbc <- FALSE
+sbc_generator <- function() {
+  data <- hmm_simulator(50, 20, 3, use_noisy_states = noisy_states_in_sbc)
+  standata <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
+
+  list(observed = standata,
+       true = data$true)
+}
+
+sbc_res <- sbc(model_single, generator = sbc_generator, N_steps = 200, control = list(adapt_delta = adapt_delta))
 
@@
 
-sbc_res$params %>% filter(grepl("mu", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("b|sd", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("r_1", param_name)) %>% plot_sbc_params()
+sbc_res$params %>% filter(grepl("r_2", param_name)) %>% plot_sbc_params()
+if(noisy_states_in_sbc) {
   sbc_res$params %>% filter(grepl("sensitivity", param_name)) %>% plot_sbc_params()
   sbc_res$params %>% filter(grepl("other_observations_probs", param_name)) %>% plot_sbc_params()
+}
 
@@
 ```{r}
-data <- hmm_simulator(5, 10, 2, use_noisy_states = FALSE)
-cc <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, prior= data$observed_structured$prior)
-
-model2 <- stan_model(model_code = cc)
-
-prep_data <- make_data_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
-
-all_data <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data)
-
-fit_raw <- sampling(model2, data = all_data)
-#TODO - compactification is necessary!
 ```

### 2020-05-14 R/hmm_simulator.R

@@ -23,43 +23,63 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
 
   rates_from <- array(0, N_rates)
   rates_to <- array(0, N_rates)
-  rates_prior_mean <- numeric(N_rates)
+  rates_group <- character(N_rates)
 
   next_rate <- 1
   rates_death <- next_rate:(next_rate + N_mid_states - 1)
   rates_from[rates_death] <- s_worsening_hid
   rates_to[rates_death] <- s_death_hid
-  rates_prior_mean[rates_death] <- 0.01
+  rates_group[rates_death] <- "death"
   next_rate <- next_rate + length(rates_death)
 
   rates_to_improving <- next_rate:(next_rate + N_mid_states - 1)
   rates_from[rates_to_improving] <- s_worsening_hid
   rates_to[rates_to_improving] <- s_improving_hid
-  rates_prior_mean[rates_to_improving] <- 0.1
+  rates_group[rates_to_improving] <- "to_improving"
   next_rate <- next_rate + length(rates_to_improving)
 
   rates_to_worsening <- next_rate:(next_rate + N_mid_states - 1)
   rates_from[rates_to_worsening] <- s_improving_hid
   rates_to[rates_to_worsening] <- s_worsening_hid
-  rates_prior_mean[rates_to_worsening] <- 0.1
+  rates_group[rates_to_worsening] <- "to_worsening"
   next_rate <- next_rate + length(rates_to_worsening)
 
   rates_worsen_one <- next_rate:(next_rate + N_mid_states - 2)
   rates_from[rates_worsen_one] <- s_worsening_hid[1:(length(s_worsening_hid) - 1)]
   rates_to[rates_worsen_one] <- s_worsening_hid[2:length(s_worsening_hid)]
-  rates_prior_mean[rates_worsen_one] <- 0.2
+  rates_group[rates_worsen_one] <- "worsen_one"
   next_rate <- next_rate + length(rates_worsen_one)
 
   rates_improve_one <- next_rate:(next_rate + N_mid_states - 2)
   rates_from[rates_improve_one] <- s_improving_hid[2:length(s_improving_hid)]
   rates_to[rates_improve_one] <- s_improving_hid[1:(length(s_improving_hid) - 1)]
-  rates_prior_mean[rates_improve_one] <- 0.2
+  rates_group[rates_improve_one] <- "improve_one"
   next_rate <- next_rate + length(rates_improve_one)
 
   rate_discharge <- next_rate
[…]

### 2020-05-14 R/make_stancode_hmm.R

@@ -30,7 +30,7 @@ rate_hmm_stanvars <- function(standata) {
   brms::stanvar(scode = rate_hmm_parameters_code, block = "parameters")
 }
 
-make_data_hmm <- function(formula_processed, serie_data, rate_data, state_data) {
+make_data_hmm <- function(formula_processed, serie_data, rate_data, state_data, sensitivity_low_bound = 0.5) {
 
   rate_data <- rate_data %>% mutate(.rate_id = factor(1:n()))
 
@@ -40,12 +40,37 @@ make_data_hmm <- function(formula_processed, serie_data, rate_data, state_data) 
 
   N_rates = nrow(rate_data)
 
-  state_data <- state_data %>% arrange(.id)
-  if(!identical(state_data$.id, 1:nrow(state_data))) {
+  state_data <- state_data %>% arrange(id)
+  if(!identical(state_data$id, 1:nrow(state_data))) {
     stop("State ids need to be consecutive and not duplicated")
   }
-  #TODO allow for noisy states
-  N_states_observed <- length(unique(state_data$.corresponding_obs))
+  possibly_observed_states <- state_data %>% select(corresponding_obs, starts_with("other_obs_")) %>%
+    as.matrix() %>% as.integer() %>% unique() %>% sort()
+  possibly_observed_states <- possibly_observed_states[!is.na(possibly_observed_states)]
+  if(!identical(possibly_observed_states, 1:length(possibly_observed_states))) {
+    stop("Possibly observed states have gaps")
+  }
+  N_states_observed <- length(possibly_observed_states)
+
+  N_noisy_states <- sum(state_data$is_noisy)
+  if(N_noisy_states > 0) {
+    noisy_states <- state_data %>% filter(is_noisy) %>% pull(id)
+
+    N_other_observations <- state_data %>% select(starts_with("other_obs_")) %>% length()
+    if(N_other_observations < 1) {
+      stop("Noisy states require other_obs_x columns")
+    }
+
+    noisy_states_other_obs <-  state_data %>%
+      filter(is_noisy) %>%
+      select(starts_with("other_obs_")) %>%
+      as.matrix()
+  } else {
+    noisy_states = numeric(0)
+    N_other_observations = 1
+    noisy_states_other_obs = array(0, c(0, 1))
+  }
+
 
   N_series <- max(serie_data$.serie)
[…]

### 2020-05-15 dev/hmm_model_devel.Rmd
Moving noisiness back to observed states

@@
 
-single_code <- make_stancode_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, prior= data$observed_structured$prior)
+single_code <- make_stancode_hmm(data$observed_structured$formula, 
+                                 serie_data = data$observed_structured$serie_data, 
+                                 rate_data = data$observed_structured$rate_data, 
+                                 hidden_state_data = data$observed_structured$hidden_state_data, 
+                                 observed_state_data = data$observed_structured$observed_state_data,
+                                 prior= data$observed_structured$prior)
 
@@
 
-prepdata <- make_data_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
-
-standata <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
+standata <- make_standata_hmm(data$observed_structured$formula, 
+                              serie_data = data$observed_structured$serie_data, 
+                              rate_data = data$observed_structured$rate_data, 
+                              hidden_state_data = data$observed_structured$hidden_state_data, 
+                              observed_state_data = data$observed_structured$observed_state_data,
+                              sensitivity_low_bound = data$observed$sensitivity_low_bound)
 

### 2020-05-15 R/hmm_simulator.R

@@ -117,29 +117,25 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
   corresponding_observation[s_improving_hid] <- s_mid_obs
   corresponding_observation[s_death_hid] <-  s_death_obs
   if(use_noisy_states) {
-    #TODO switch to using hidden states as base
-    N_noisy_states <- 2 * N_mid_states
-    noisy_states <- c(s_worsening_hid, s_improving_hid)
-    noisy_state_id <- array(NA_integer_, N_states_hidden)
+    N_noisy_states <- N_mid_states
+    noisy_states <- s_mid_obs
+    noisy_state_id <- array(NA_integer_, N_states_observed)
     noisy_state_id[noisy_states] <- 1:N_noisy_states
 
-    if(N_mid_states == 1) {
+    if(N_noisy_states == 1) {
       stop("Need at least 2 mid states for noisy states")
-    } else if(N_mid_states == 2) {
+    } else if(N_noisy_states == 2) {
       N_other_observations <- 1
       noisy_states_other_obs <- array(NA_integer_, c(N_noisy_states, N_other_observations))
-      noisy_states_other_obs[c(1,3), ] <- s_mid_obs[2]
-      noisy_states_other_obs[c(2,4), ] <- s_mid_obs[1]
+      noisy_states_other_obs[1, ] <- noisy_states[2]
+      noisy_states_other_obs[2, ] <- noisy_states[1]
     } else {
       N_other_observations <- 2
       noisy_states_other_obs <- array(NA_integer_, c(N_noisy_states, N_other_observations))
-      noisy_states_other_obs[noisy_state_id[s_improving_hid[1]], ] <- s_mid_obs[c(2,3)]
-      noisy_states_other_obs[noisy_state_id[s_worsening_hid[1]], ] <- s_mid_obs[c(2,3)]
-      noisy_states_other_obs[noisy_state_id[s_improving_hid[N_mid_states]], ] <- s_mid_obs[c(N_mid_states - 1,N_mid_states - 2)]
-      noisy_states_other_obs[noisy_state_id[s_worsening_hid[N_mid_states]], ] <- s_mid_obs[c(N_mid_states - 1,N_mid_states - 2)]
-      for(ns in 2:(N_mid_states - 1)) {
[…]

### 2020-05-15 R/make_stancode_hmm.R

@@ -5,9 +5,9 @@ make_brms_formula_hmm <- function(formula) {
   update.formula(formula, .dummy ~ .)
 }
 
-make_stancode_hmm <- function(formula, serie_data, rate_data, state_data, prior = NULL) {
+make_stancode_hmm <- function(formula, serie_data, rate_data, hidden_state_data, observed_state_data = NULL, prior = NULL) {
   formula <- make_brms_formula_hmm(formula)
-  data <- make_data_hmm(formula, serie_data, rate_data, state_data)
+  data <- make_data_hmm(formula, serie_data, rate_data, hidden_state_data, observed_state_data)
   brms::make_stancode(formula, family = rate_hmm_family,
                       data = data$brmsdata,
                       stanvars = rate_hmm_stanvars(data$standata), prior = prior)
@@ -30,7 +30,7 @@ rate_hmm_stanvars <- function(standata) {
   brms::stanvar(scode = rate_hmm_parameters_code, block = "parameters")
 }
 
-make_data_hmm <- function(formula_processed, serie_data, rate_data, state_data, sensitivity_low_bound = 0.5) {
+make_data_hmm <- function(formula_processed, serie_data, rate_data, hidden_state_data, observed_state_data = NULL, sensitivity_low_bound = 0.5) {
 
   rate_data <- rate_data %>% mutate(.rate_id = factor(1:n()))
 
@@ -40,28 +40,46 @@ make_data_hmm <- function(formula_processed, serie_data, rate_data, state_data, 
 
   N_rates = nrow(rate_data)
 
-  state_data <- state_data %>% arrange(id)
-  if(!identical(state_data$id, 1:nrow(state_data))) {
-    stop("State ids need to be consecutive and not duplicated")
+  hidden_state_data <- hidden_state_data %>% arrange(id)
+  if(!identical(hidden_state_data$id, 1:N_states_hidden)) {
+    stop("Hiden state ids need to be consecutive and not duplicated and span the same range as rate_data")
   }
-  possibly_observed_states <- state_data %>% select(corresponding_obs, starts_with("other_obs_")) %>%
-    as.matrix() %>% as.integer() %>% unique() %>% sort()
-  possibly_observed_states <- possibly_observed_states[!is.na(possibly_observed_states)]
-  if(!identical(possibly_observed_states, 1:length(possibly_observed_states))) {
-    stop("Possibly observed states have gaps")
-  }
[…]

### 2020-05-15 dev/hmm_model_devel.Rmd
Known initial states

@@
 
-data <- hmm_simulator(30, 20, 2, use_noisy_states = TRUE)
+data <- hmm_simulator(5, 10, 2, use_noisy_states = TRUE)
 
-single_code <- make_stancode_hmm(data$observed_structured$formula, 
-                                 serie_data = data$observed_structured$serie_data, 
-                                 rate_data = data$observed_structured$rate_data, 
-                                 hidden_state_data = data$observed_structured$hidden_state_data, 
-                                 observed_state_data = data$observed_structured$observed_state_data,
-                                 prior= data$observed_structured$prior)
+single_code <- make_stancode_hmm(data$observed$formula, 
+                                 serie_data = data$observed$serie_data, 
+                                 rate_data = data$observed$rate_data, 
+                                 hidden_state_data = data$observed$hidden_state_data, 
+                                 initial_states = data$observed$initial_states,
+                                 observed_state_data = data$observed$observed_state_data,
+                                 prior= data$observed$prior)
 
@@
 
-standata <- make_standata_hmm(data$observed_structured$formula, 
-                              serie_data = data$observed_structured$serie_data, 
-                              rate_data = data$observed_structured$rate_data, 
-                              hidden_state_data = data$observed_structured$hidden_state_data, 
-                              observed_state_data = data$observed_structured$observed_state_data,
+standata <- make_standata_hmm(data$observed$formula, 
+                              serie_data = data$observed$serie_data, 
+                              rate_data = data$observed$rate_data, 
+                              hidden_state_data = data$observed$hidden_state_data, 
+                              initial_states = data$observed$initial_states,
+                              observed_state_data = data$observed$observed_state_data,
                               sensitivity_low_bound = data$observed$sensitivity_low_bound)
@@
   data <- hmm_simulator(50, 20, 3, use_noisy_states = noisy_states_in_sbc)
-  standata <- make_standata_hmm(data$observed_structured$formula, serie_data = data$observed_structured$serie_data, rate_data = data$observed_structured$rate_data, state_data = data$observed_structured$state_data, sensitivity_low_bound = data$observed$sensitivity_low_bound)
+  standata <- make_standata_hmm(data$observed$formula, 
+                                serie_data = data$observed$serie_data, 
+                                rate_data = data$observed$rate_data, 
+                                hidden_state_data = data$observed$hidden_state_data, 
+                                initial_states = data$observed$initial_states,
+                                observed_state_data = data$observed$observed_state_data,
+                                sensitivity_low_bound = data$observed$sensitivity_low_bound)
 

### 2020-05-15 R/hmm_simulator.R

@@ -157,14 +157,11 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
   sensitivity <- runif(N_noisy_states, min = sensitivity_low_bound, max = 1)
 
 
-  initial_states_prob <- numeric(N_states_hidden)
-  initial_states_prob[s_worsening_hid] <- 1 / N_mid_states
+  hidden_state_data <- data.frame(id = 1:N_states_hidden,
+                          corresponding_obs = corresponding_observation)
 
-  hidden_state_data = data.frame(id = 1:N_states_hidden,
-                          corresponding_obs = corresponding_observation,
-                          initial_prob = initial_states_prob)
 
-  observed_state_data = data.frame(id = 1:N_states_observed,
+  observed_state_data <- data.frame(id = 1:N_states_observed,
                           is_noisy = (1:N_states_observed) %in% noisy_states)
   if(N_noisy_states > 0) {
     for(other_obs in 1:N_other_observations) {
@@ -181,6 +178,8 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
   predictor_sets <- array(NA_integer_, N_observations)
   treated <- logical(N_observations)
 
+  initial_states <- sample(s_worsening_hid, N_series, replace = TRUE)
+
 
   true_base_states <- array(NA_integer_, N_observations)
   true_improving <- array(FALSE, N_observations)
@@ -191,7 +190,7 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
     } else {
       max_time <- N_time - 1
     }
-    state <- sample(1:N_states_hidden, 1, prob = initial_states_prob)
+    state <- initial_states[p]
     for(t in 1:max_time) {
       series[next_observation] <- p
       times[next_observation] <- t
@@ -231,38 +230,6 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
 
   list(
     observed = loo::nlist(
-      N,
-      N_states_hidden,
-      N_states_observed,
-
-      N_rates,
-      rates_from,
-      rates_to,
-
-      corresponding_observation,
-
-      N_noisy_states,
-      noisy_states,
-      N_other_observations,
-      noisy_states_other_obs,
-
-      sensitivity_low_bound,
-
-      initial_states_prob,
-
-      N_observations,
-      N_series,
-      N_time,
[…]

### 2020-05-15 R/make_stancode_hmm.R

@@ -5,9 +5,13 @@ make_brms_formula_hmm <- function(formula) {
   update.formula(formula, .dummy ~ .)
 }
 
-make_stancode_hmm <- function(formula, serie_data, rate_data, hidden_state_data, observed_state_data = NULL, prior = NULL) {
+make_stancode_hmm <- function(formula,
+                              serie_data, rate_data,
+                              hidden_state_data, initial_states,
+                              observed_state_data = NULL, prior = NULL) {
   formula <- make_brms_formula_hmm(formula)
-  data <- make_data_hmm(formula, serie_data, rate_data, hidden_state_data, observed_state_data)
+  data <- make_data_hmm(formula, serie_data, rate_data,
+                        hidden_state_data, initial_states, observed_state_data)
   brms::make_stancode(formula, family = rate_hmm_family,
                       data = data$brmsdata,
                       stanvars = rate_hmm_stanvars(data$standata), prior = prior)
@@ -30,7 +34,10 @@ rate_hmm_stanvars <- function(standata) {
   brms::stanvar(scode = rate_hmm_parameters_code, block = "parameters")
 }
 
-make_data_hmm <- function(formula_processed, serie_data, rate_data, hidden_state_data, observed_state_data = NULL, sensitivity_low_bound = 0.5) {
+make_data_hmm <- function(formula_processed,
+                          serie_data, rate_data,
+                          hidden_state_data, initial_states,
+                          observed_state_data = NULL, sensitivity_low_bound = 0.5) {
 
   rate_data <- rate_data %>% mutate(.rate_id = factor(1:n()))
 
@@ -148,13 +155,13 @@ make_data_hmm <- function(formula_processed, serie_data, rate_data, hidden_state
     noisy_states_other_obs,
 
     sensitivity_low_bound,
-    initial_states_prob = hidden_state_data$initial_prob,
 
     N_observations = nrow(serie_data),
     N_series,
     N_time,
     N_predictor_sets,
 
+    initial_states,
     series = serie_data$.serie,
     times = serie_data$.time,
     obs_states,
@@ -166,10 +173,14 @@ make_data_hmm <- function(formula_processed, serie_data, rate_data, hidden_state
 }
 
 make_standata_hmm <- function(formula, serie_data, rate_data,
-                              hidden_state_data, observed_state_data = NULL,
+                              hidden_state_data, initial_states,
[…]

### 2020-05-15 dev/hmm_model_devel.Rmd
Preparing gen. quants for posterior predictive stuff

@@
 ```{r}
-noisy_states_in_sbc <- FALSE
+noisy_states_in_sbc <- TRUE
 sbc_generator <- function() {
-  data <- hmm_simulator(50, 20, 3, use_noisy_states = noisy_states_in_sbc)
+  data <- hmm_simulator(20, 15, 2, use_noisy_states = noisy_states_in_sbc)
   standata <- make_standata_hmm(data$observed$formula, 
@@
 
-sbc_res <- sbc(model_single, generator = sbc_generator, N_steps = 200, control = list(adapt_delta = adapt_delta))
+sbc_res <- sbc(model_single, generator = sbc_generator, N_steps = 50, control = list(adapt_delta = adapt_delta))
 
@@
 
-```{r}
-library(brms)
-```
-
-```{r}
-```
+## Apply to Grein et al. dataset
 
@@
 
-test_data <- data.frame(a = c(1,2,3), b = c(0.5,1.0,1.5), x1 = c("A"), x2 = c(0,1,2))
-f <- a ~ 0 + (0 + x2 | a | x1)
-# make_stancode(f, data = test_data)
-#brms::make_standata(f, data = test_data)
-xx <- brms::brmsformula(f)
-
-xx$resp
-
-stan_llh.rate_hmm <- function( ... ) {
-  ""
-}
-
-.family_rate_hmm <- function() {
-
-}
-
-ff <-
-
-make_stancode(a ~ 0 + x2, family = ff, data = test_data)
+  breathing_data_orig <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
+    select(-Remdesivir_Last_Day) %>%
+    pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
+    filter(!is.na(breathing)) %>%
+    transmute(
+      patient = Patient,
+      breathing = if_else(breathing %in% c("LoO2", "HiO2"), "Oxygen", breathing),
+      day = as.integer(day)) %>%
+    group_by(patient) %>%
+    mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "Oxygen", "NIPPV", "MV", "ECMO")])], breathing)) %>%
+    ungroup() %>%
+    mutate(
+      breathing = factor(breathing, levels = c("Discharge", "AA", "Oxygen", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
+    )
 

### 2020-05-15 dev/hmm_model_devel.Rmd
Making the HMM family more compatible with brms

@@
 ```{r}
+
+prepdata <- make_data_hmm(data$observed$formula, 
+                              serie_data = data$observed$serie_data, 
+                              rate_data = data$observed$rate_data, 
+                              hidden_state_data = data$observed$hidden_state_data, 
+                              initial_states = data$observed$initial_states,
+                              observed_state_data = data$observed$observed_state_data,
+                              sensitivity_low_bound = data$observed$sensitivity_low_bound)
+
+bfit <- brm(formula = make_brms_formula_hmm(data$observed$formula), family = rate_hmm_family, data = prepdata$brmsdata, prior = data$observed$prior, stanvars = rate_hmm_stanvars(prepdata$standata))
+
+bfit <- brms:::brmsfit(formula = make_brms_formula_hmm(data$observed$formula), data = prepdata$brmsdata, prior = data$observed$prior, stanvars = rate_hmm_stanvars(prepdata$standata), fit = fit)
+
+bfit
+
+posterior_epred(bfit)
+
+brms::make_standata(formula = make_brms_formula_hmm(data$observed$formula), family = rate_hmm_family, data = prepdata$brmsdata, prior = data$observed$prior, stanvars = rate_hmm_stanvars(prepdata$standata))
+
+```
+
+```{r}
 noisy_states_in_sbc <- TRUE

### 2020-05-15 R/make_stancode_hmm.R

@@ -2,7 +2,8 @@ make_brms_formula_hmm <- function(formula) {
   if(!is.null(brms:::lhs(formula))) {
     stop("Formula needs to be one-sided")
   }
-  update.formula(formula, .dummy ~ .)
+
+  brms::brmsformula(update.formula(formula, .dummy ~ .), family = rate_hmm_family)
 }
 
 make_stancode_hmm <- function(formula,
@@ -19,6 +20,7 @@ make_stancode_hmm <- function(formula,
 
 rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars = "mu",
                            lb = NA, ub = NA, type = "real", vars = NULL, specials = NULL,
+                           ybounds = c(-Inf, Inf),
                            log_lik = NULL, posterior_predict = NULL, posterior_epred = NULL),
                       class = c("rate_hmm", "brmsfamily"))
 
@@ -29,7 +31,7 @@ rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars 
 rate_hmm_stanvars <- function(standata) {
 
   brms::stanvar(scode = rate_hmm_functions_code, block = "functions") +
-  brms::stanvar(x = standata, scode = rate_hmm_data_code, block = "data") +
+  rate_hmm_stanvars_data(standata) +
   brms::stanvar(scode = rate_hmm_tdata_code, block = "tdata") +
   brms::stanvar(scode = rate_hmm_parameters_code, block = "parameters") +
   brms::stanvar(scode = rate_hmm_genquant_code, block = "genquant")
@@ -80,41 +82,38 @@ rate_hmm_functions_code <- "
   }
 "
 
-rate_hmm_data_code <- '
-  // HMM data
-  int<lower=1> N_states_hidden;
-  int<lower=1> N_states_observed;
+rate_hmm_stanvars_data  <- function(standata) {
+  brms::stanvar(x = standata$N_states_hidden, name = "N_states_hidden", scode = "  // HMM data\n  int<lower=1> N_states_hidden;", block = "data") +
+  brms::stanvar(x = standata$N_states_observed, name = "N_states_observed", scode = "  int<lower=1> N_states_observed;", block = "data") +
 
-  int<lower=1> N_rates;
-  int<lower=1, upper=N_states_hidden> rates_from[N_rates];
-  int<lower=1, upper=N_states_hidden> rates_to[N_rates];
+  brms::stanvar(x = standata$N_rates, name = "N_rates", scode = "  int<lower=1> N_rates;", block = "data") +
[…]

### 2020-05-18 dev/hmm_model_devel.Rmd
Towards better model wrapping, posterior predictive checks

--- 2020-05-15 dev/hmm_model_devel.Rmd
@@
 
-improving_data <- data.frame(day = data$observed$times, patient = data$observed$series, improving = data$true$true_improving) %>% filter(improving)
+improving_data <- data.frame(day = data$observed$serie_data$.time, patient = data$observed$serie_data$.serie, improving = data$true$true_improving) %>% filter(improving)
 
-data.frame(day = data$observed$times, patient = data$observed$series, breathing = data$true$true_base_states) %>%
-  mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
+data.frame(day = data$observed$serie_data$.time, patient = data$observed$serie_data$.serie, breathing = data$true$true_base_states) %>%
+  mutate(breathing = factor(breathing, 1:nrow(data$observed$observed_state_data), ordered = TRUE)) %>%
   ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + geom_tile(data = improving_data, width = 1, height = 0.1, fill = "gray") + ggtitle("True - gray lines for improving")
 
-data.frame(day = data$observed$times, patient = data$observed$series, breathing = data$observed$obs_states) %>%
-  mutate(breathing = factor(breathing, 1:data$observed$N_states_observed, ordered = TRUE)) %>%
+data.frame(day = data$observed$serie_data$.time, patient = data$observed$serie_data$.serie, breathing = data$observed$serie_data$.observed) %>%
+  mutate(breathing = factor(breathing, 1:nrow(data$observed$observed_state_data), ordered = TRUE)) %>%
   ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile(width = 1, height = 0.5) + ggtitle("Observed")
@@
 
-single_code <- make_stancode_hmm(data$observed$formula, 
-                                 serie_data = data$observed$serie_data, 
-                                 rate_data = data$observed$rate_data, 
-                                 hidden_state_data = data$observed$hidden_state_data, 
-                                 initial_states = data$observed$initial_states,
-                                 observed_state_data = data$observed$observed_state_data,
-                                 prior= data$observed$prior)
+single_code <- make_stancode_hmm(data$observed)
 
@@
 
-standata <- make_standata_hmm(data$observed$formula, 
-                              serie_data = data$observed$serie_data, 
-                              rate_data = data$observed$rate_data, 
-                              hidden_state_data = data$observed$hidden_state_data, 
-                              initial_states = data$observed$initial_states,
-                              observed_state_data = data$observed$observed_state_data,
-                              sensitivity_low_bound = data$observed$sensitivity_low_bound)
+standata <- make_standata_hmm(data$observed)
 
@@
 ```{r}
-noisy_states_in_sbc <- FALSE
+data <- hmm_simulator(5, 10, 2, use_noisy_states = TRUE)
+
+bfit <- brmhmm(data$observed)
+
+summary(bfit)
+
+xx <- brms::posterior_epred(bfit)
+dim(xx)
+```
+
+```{r}
+noisy_states_in_sbc <- TRUE
 sbc_generator <- function() {
-  data <- hmm_simulator(50, 20, 3, use_noisy_states = noisy_states_in_sbc)
-  standata <- make_standata_hmm(data$observed$formula, 
-                                serie_data = data$observed$serie_data, 
-                                rate_data = data$observed$rate_data, 
-                                hidden_state_data = data$observed$hidden_state_data, 
-                                initial_states = data$observed$initial_states,
-                                observed_state_data = data$observed$observed_state_data,
-                                sensitivity_low_bound = data$observed$sensitivity_low_bound)
+  data <- hmm_simulator(40, 25, 2, use_noisy_states = noisy_states_in_sbc)
+  standata <- make_standata_hmm(data$observed)
 
   list(observed = standata,
-       true = data$true)
+       true = data$true,
+       observed_raw = data$observed)
 }
 
-sbc_res <- sbc(model_single, generator = sbc_generator, N_steps = 200, control = list(adapt_delta = adapt_delta))
+model_sbc <-  stan_model(model_code = make_stancode_hmm(sbc_generator()$observed_raw))
+
+sbc_res <- sbc(model_sbc, generator = sbc_generator, N_steps = 50, control = list(adapt_delta = adapt_delta))
 
@@
 
-```{r}
-library(brms)
-```
+## Apply to Grein et al. dataset
 
 ```{r}
+data_grein <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
+  select(-Remdesivir_Last_Day) %>%
+  pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
+  filter(!is.na(breathing)) %>%
+  transmute(
+    patient = Patient,
+    breathing = breathing,
+    day = as.integer(day)) %>%
+  group_by(patient) %>%
+  mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
+  ungroup() %>%
+  mutate(
+    breathing = factor(breathing, levels = c("Discharge", "AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
+  )
+
 ```
@@
 
-test_data <- data.frame(a = c(1,2,3), b = c(0.5,1.0,1.5), x1 = c("A"), x2 = c(0,1,2))
-f <- a ~ 0 + (0 + x2 | a | x1)
-# make_stancode(f, data = test_data)
-#brms::make_standata(f, data = test_data)
-xx <- brms::brmsformula(f)
-
-xx$resp
-
-stan_llh.rate_hmm <- function( ... ) {
-  ""
-}
-
-.family_rate_hmm <- function() {
-
-}
-
-ff <-
-
-make_stancode(a ~ 0 + x2, family = ff, data = test_data)
-
 ```

### 2020-05-18 R/brmshmmdata.R

@@ -1,0 +1,65 @@
+brmshmmdata <- function(formula,
+                        serie_data, rate_data,
+                        hidden_state_data, initial_states,
+                        observed_state_data = NULL, prior = NULL,
+                        sensitivity_low_bound = 0.5) {
+
+  validate_brmshmmdata(
+    structure(loo::nlist(
+      formula = make_brms_formula_hmm(formula),
+      serie_data,
+      rate_data,
+      hidden_state_data,
+      initial_states,
+      observed_state_data,
+      prior,
+      sensitivity_low_bound),
+      class = "brmshmmdata"
+      ))
+}
+
+validate_brmshmmdata <- function(d) {
+  if(!inherits(d, "brmshmmdata")) {
+    stop("Must be of class brmshmmdata")
+  }
+
+  if(is.null(d$rate_data$.rate_id)) {
+    d$rate_data <- d$rate_data %>% mutate(.rate_id = factor(1:n()))
+  } else {
+    if(!is.factor(d$rate_data$.rate_id)) {
+      stop("rate_data$.rate_id must be a factor")
+    }
+  }
+
+  #TODO: test for duplicate rates
+  #TODO: test for states with no rates
+
+  d$hidden_state_data <- d$hidden_state_data %>% arrange(id)
+  if(!identical(d$hidden_state_data$id, 1:nrow(d$hidden_state_data))) {
+    stop("Hiden state ids need to be consecutive and not duplicated and span the same range as rate_data")
+  }
+
+  all_corresponding_states <- d$hidden_state_data %>% pull(corresponding_obs) %>%
+    unique() %>% sort()
+
+
+  if(is.null(d$observed_state_data)) {
+
+    if(!identical(all_corresponding_states, 1:length(all_corresponding_states))) {
+      stop("If observed_state_data is NULL, corresponding_obs cannot have gaps")
+    }
+
+    d$observed_state_data <- data.frame(id = all_corresponding_states, is_noisy = FALSE)
+  }
+
+  d$observed_state_data <- d$observed_state_data %>% arrange(id)
+  if(!identical(d$observed_state_data$id, 1:nrow(d$observed_state_data))) {
+    stop("Observed state ids need to be consecutive.")
+  }
+
+  if(!all(all_corresponding_states %in% d$observed_state_data$id)) {
+    stop("Some corresponding observations are not in observed_state_data")
+  }
+
+  d
+}

### 2020-05-18 R/brmshmmfit.R

@@ -1,0 +1,90 @@
+brmhmm <- function(brmshmmdata) {
+  d <- validate_brmshmmdata(brmshmmdata)
+
+  prepdata <- make_data_hmm(d)
+
+  bfit <- brms::brm(
+    formula = d$formula,
+    family = rate_hmm_family,
+    data = prepdata$brmsdata,
+    prior = d$prior,
+    stanvars = rate_hmm_stanvars(prepdata$standata)
+    )
+
+  structure(list(
+    brmsfit = bfit,
+    data = brmshmmdata
+  ), class = "brmshmmfit")
+}
+
+
+validate_brmshmmfit <- function(fit) {
+  if(!inherits(fit, "brmshmmfit")) {
+    stop("Not of class brmshmmfit")
+  }
+  fit
+}
+
+posterior_epred.brmshmmfit <- function(fit, nsamples = NULL) {
+  validate_brmshmmfit(fit)
+  brms:::contains_samples(fit$brmsfit)
+
+  pred_rawdata <- fit$data
+
+  data_hmm <- make_data_hmm(pred_rawdata)
+
+  epred_mu <- brms::posterior_epred(fit$brmsfit, newdata = data_hmm$brmsdata, nsamples = nsamples)
+  transition_matrices <- compute_all_transition_matrices(data_hmm, epred_mu)
+
+  if(is.null(nsamples)) {
+    nsamples <- dim(epred_mu)[1]
+  }
+
+  N_states_hidden <- data_hmm$standata$N_states_hidden
+
+  result_rect <- array(NA_integer_, c(max(pred_rawdata$serie_data$.time), max(pred_rawdata$serie_data$.serie), nsamples))
+  for(s in unique(pred_rawdata$serie_data$.serie)) {
+
+    state <- pred_rawdata$initial_states[s]
+    result_rect[1, s, ] <- state
+
+    max_time <- fit$data$serie_data %>% filter(.serie == s) %>% pull(.time) %>% max()
+    for(i in 1:nsamples) {
+      for(t in 2:max_time) {
+        ps <- data_hmm$standata$predictor_sets_rect[s, t]
+        state <- sample(1:N_states_hidden, size = 1, prob = transition_matrices[state, , ps, i])
+        result_rect[t, s, i] <- state
+      }
+    }
+  }
+
+  result <- array(NA_integer_, c(nsamples, nrow(pred_rawdata$serie_data)))
+
+  for(o in 1:nrow(pred_rawdata$serie_data)) {
[…]

### 2020-05-18 R/hmm_simulator.R

@@ -229,34 +229,34 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
   }
 
   list(
-    observed = loo::nlist(
-      formula = ~ 0 + Intercept + rate_death + rate_to_worsening + rate_improve_one + rate_worsen_one + (1 || .rate_id) + (0 + treated || group),
-      prior =
-        brms::set_prior("normal(-2.3,1)", "b", coef = "Intercept") +
-        brms::set_prior("normal(-2.3,1)", "b", coef = "rate_death") +
-        brms::set_prior("normal(0,1)", "b", coef = "rate_to_worsening") +
-        brms::set_prior("normal(0.7,1)", "b", coef = "rate_improve_one") +
-        brms::set_prior("normal(0.7,1)", "b", coef = "rate_worsen_one") +
-        brms::set_prior("normal(0, 0.5)", "sd", group = ".rate_id") +
-        brms::set_prior("normal(0, 1)", "sd", group = "group"),
-      serie_data = data.frame(
-        .serie = series,
-        .time = times,
-        .observed = obs_states,
-        treated = as.numeric(treated)),
-      rate_data = data.frame(
-        .from = rates_from,
-        .to = rates_to,
-        group = rates_group,
-        rate_death = as.numeric(rates_group == "death"),
-        rate_to_worsening = as.numeric(rates_group == "to_worsening"),
-        rate_improve_one = as.numeric(rates_group == "improve_one"),
-        rate_worsen_one = as.numeric(rates_group == "worsen_one")
-      ),
-      hidden_state_data,
-      initial_states,
-      sensitivity_low_bound,
-      observed_state_data
+    observed = brmshmmdata(
+        formula = ~ 0 + Intercept + rate_death + rate_to_worsening + rate_improve_one + rate_worsen_one + (1 || .rate_id) + (0 + treated || group),
+        prior =
+          brms::set_prior("normal(-2.3,1)", "b", coef = "Intercept") +
+          brms::set_prior("normal(-2.3,1)", "b", coef = "rate_death") +
+          brms::set_prior("normal(0,1)", "b", coef = "rate_to_worsening") +
+          brms::set_prior("normal(0.7,1)", "b", coef = "rate_improve_one") +
+          brms::set_prior("normal(0.7,1)", "b", coef = "rate_worsen_one") +
+          brms::set_prior("normal(0, 0.5)", "sd", group = ".rate_id") +
[…]

### 2020-05-18 R/make_stancode_hmm.R

@@ -6,16 +6,13 @@ make_brms_formula_hmm <- function(formula) {
   brms::brmsformula(update.formula(formula, .dummy ~ .), family = rate_hmm_family)
 }
 
-make_stancode_hmm <- function(formula,
-                              serie_data, rate_data,
-                              hidden_state_data, initial_states,
-                              observed_state_data = NULL, prior = NULL) {
-  formula <- make_brms_formula_hmm(formula)
-  data <- make_data_hmm(formula, serie_data, rate_data,
-                        hidden_state_data, initial_states, observed_state_data)
-  brms::make_stancode(formula, family = rate_hmm_family,
+make_stancode_hmm <- function(brmshmmdata) {
+  d <- validate_brmshmmdata(brmshmmdata)
+
+  data <- make_data_hmm(d)
+  brms::make_stancode(d$formula, family = rate_hmm_family,
                       data = data$brmsdata,
-                      stanvars = rate_hmm_stanvars(data$standata), prior = prior)
+                      stanvars = rate_hmm_stanvars(data$standata), prior = d$prior)
 }
 
 rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars = "mu",
@@ -24,9 +21,18 @@ rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars 
                            log_lik = NULL, posterior_predict = NULL, posterior_epred = NULL),
                       class = c("rate_hmm", "brmsfamily"))
 
+summary.rate_hmm <- function(f, link = FALSE) {
+  "Rate-based HMM"
+}
+
 .family_rate_hmm <- function() {
 
 }
+
+posterior_epred_rate_hmm <- function(prep) {
+  brms:::posterior_epred_gaussian(prep)
+}
+
 
 rate_hmm_stanvars <- function(standata) {
 
@@ -100,18 +106,16 @@ rate_hmm_stanvars_data  <- function(standata) {
   brms::stanvar(x = standata$sensitivity_low_bound, name = "sensitivity_low_bound", scode = "  real<lower=0, upper=1> sensitivity_low_bound;", block = "data") +
 
 
-  brms::stanvar(x = standata$N_observations, name = "N_observations", scode = "  // Observations\n  int<lower=1> N_observations;", block = "data") +
[…]

### 2020-05-18 R/make_standata_hmm.R

@@ -1,44 +1,18 @@
-make_data_hmm <- function(formula_processed,
-                          serie_data, rate_data,
-                          hidden_state_data, initial_states,
-                          observed_state_data = NULL, sensitivity_low_bound = 0.5) {
+make_data_hmm <- function(brmshmmdata) {
+  d <- validate_brmshmmdata(brmshmmdata)
 
-  rate_data <- rate_data %>% mutate(.rate_id = factor(1:n()))
-
-
+  formula_processed <- d$formula
+  serie_data <- d$serie_data
+  rate_data <- d$rate_data
+  hidden_state_data <- d$hidden_state_data
+  observed_state_data <- d$observed_state_data
 
   N_states_hidden <- length(unique(c(rate_data$.from, rate_data$.to)))
 
   N_rates = nrow(rate_data)
 
-  hidden_state_data <- hidden_state_data %>% arrange(id)
-  if(!identical(hidden_state_data$id, 1:N_states_hidden)) {
-    stop("Hiden state ids need to be consecutive and not duplicated and span the same range as rate_data")
-  }
-
-  all_corresponding_states <- hidden_state_data %>% pull(corresponding_obs) %>%
-    unique() %>% sort()
-  if(is.null(observed_state_data)) {
-
-    if(!identical(all_corresponding_states, 1:length(all_corresponding_states))) {
-      stop("If observed_state_data is NULL, corresponding_obs cannot have gaps")
-    }
-
-    observed_state_data <- data.frame(id = all_corresponding_states, is_noisy = FALSE)
-  }
 
   N_states_observed <- nrow(observed_state_data)
-  observed_state_data <- observed_state_data %>% arrange(id)
-  if(!identical(observed_state_data$id, 1:N_states_observed)) {
-    stop("Observed state ids need to be consecutive.")
-  }
-
-  if(!all(all_corresponding_states %in% observed_state_data$id)) {
-    stop("Some corresponding observations are not in observed_state_data")
-  }
-
-
-
 
 
   N_noisy_states <- sum(observed_state_data$is_noisy)
@@ -69,7 +43,7 @@ make_data_hmm <- function(formula_processed,
   N_time <- max(serie_data$.time)
 
   #TODO be smart about this, using allvars (probably pass something from make_standata)
-  all_vars_needed <- brms:::all_vars(formula_processed)
+  all_vars_needed <- all.vars(as.formula(formula_processed))
 
[…]

### 2020-05-19 R/predict_hmm.R
Roughly working posterior predictve

@@ -1,0 +1,108 @@
+posterior_epred.brmshmmfit <- function(fit, nsamples = NULL, to_corresponding_states = FALSE) {
+  validate_brmshmmfit(fit)
+  brms:::contains_samples(fit$brmsfit)
+
+  pred_rawdata <- fit$data
+
+  data_hmm <- make_data_hmm(pred_rawdata)
+
+  epred_mu <- brms::posterior_epred(fit$brmsfit, newdata = data_hmm$brmsdata, nsamples = nsamples)
+  transition_matrices <- compute_all_transition_matrices(data_hmm, epred_mu)
+
+  if(is.null(nsamples)) {
+    nsamples <- dim(epred_mu)[1]
+  }
+
+  N_states_hidden <- data_hmm$standata$N_states_hidden
+
+  result_rect <- array(NA_integer_, c(max(pred_rawdata$serie_data$.time), max(pred_rawdata$serie_data$.serie), nsamples))
+  for(s in unique(pred_rawdata$serie_data$.serie)) {
+
+    state <- pred_rawdata$initial_states[s]
+    result_rect[1, s, ] <- state
+
+    max_time <- fit$data$serie_data %>% filter(.serie == s) %>% pull(.time) %>% max()
+    for(i in 1:nsamples) {
+      for(t in 2:max_time) {
+        ps <- data_hmm$standata$predictor_sets_rect[s, t]
+        state <- sample(1:N_states_hidden, size = 1, prob = transition_matrices[state, , ps, i])
+        result_rect[t, s, i] <- state
+      }
+    }
+  }
+
+  result <- array(NA_integer_, c(nsamples, nrow(pred_rawdata$serie_data)))
+
+  for(o in 1:nrow(pred_rawdata$serie_data)) {
+    result[, o] <- result_rect[pred_rawdata$serie_data$.time[o], pred_rawdata$serie_data$.serie[o], ]
+  }
+
+  if(to_corresponding_states) {
+    data_hmm$standata$corresponding_observation[result]
+  } else {
+    result
+  }
+}
+
+posterior_predict.brmshmmfit <- function(fit, nsamples = NULL) {
+  validate_brmshmmfit(fit)
+  epred <- posterior_epred.brmshmmfit(fit, nsamples = nsamples, to_corresponding_states = FALSE)
+
+  if(is.null(nsamples)) {
+    nsamples = dim(epred)[1]
+  }
+
+  observation_probs_samples <- rstan::extract(fit$brmsfit$fit, "observation_probs")$observation_probs
+
[…]

### 2020-05-20 dev/hmm_model_devel.Rmd
Roughly working posterior predictve
Fixing posterior predictive, plots from posterior

@@
 ```{r}
-data <- hmm_simulator(5, 10, 2, use_noisy_states = TRUE)
+data <- hmm_simulator(20, 15, 2, use_noisy_states = FALSE, N_treatments = 0)
 
@@
 
-xx <- brms::posterior_epred(bfit)
-dim(xx)
+bfit$brmsfit$fit
+```
+
+```{r}
+#xx <- brms::posterior_epred(bfit) %>% hidden_to_corresponding_observed(bfit$data_processed, .)
+xx <- brms::posterior_predict(bfit)
+
+xy <- prediction_to_wide_format(bfit$data, xx)
+
+step_size <- 6
+for(step in 1:ceiling(bfit$data_processed$standata$N_series / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(xy, series_id, bfit$data) %>% print()
+
+}
 ```

### 2020-05-20 R/brmshmmfit.R

@@ -13,7 +13,8 @@ brmhmm <- function(brmshmmdata) {
 
   structure(list(
     brmsfit = bfit,
-    data = brmshmmdata
+    data = brmshmmdata,
+    data_processed = prepdata
   ), class = "brmshmmfit")
 }
 

### 2020-05-20 R/make_standata_hmm.R

@@ -42,21 +42,25 @@ make_data_hmm <- function(brmshmmdata) {
 
   N_time <- max(serie_data$.time)
 
-  #TODO be smart about this, using allvars (probably pass something from make_standata)
   all_vars_needed <- all.vars(as.formula(formula_processed))
 
   serie_data_vars <- intersect(all_vars_needed, names(serie_data))
+  if(length(serie_data_vars) > 0) {
+    serie_data_distinct <- serie_data %>% select(one_of(serie_data_vars)) %>%
+      dplyr::distinct() %>%
+      mutate(.predictor_set = 1:n())
 
-  serie_data_distinct <- serie_data %>% select(one_of(serie_data_vars)) %>%
-    dplyr::distinct() %>%
-    mutate(.predictor_set = 1:n())
+    N_predictor_sets <- nrow(serie_data_distinct)
 
-  N_predictor_sets <- nrow(serie_data_distinct)
-
-  serie_data_raw <- serie_data
-  serie_data <- serie_data %>% dplyr::left_join(serie_data_distinct, by = serie_data_vars)
-  if(any(is.na(serie_data$.predictor_set)) || nrow(serie_data_raw) != nrow(serie_data)) {
-    stop("Failed join")
+    serie_data_raw <- serie_data
+    serie_data <- serie_data %>% dplyr::left_join(serie_data_distinct, by = serie_data_vars)
+    if(any(is.na(serie_data$.predictor_set)) || nrow(serie_data_raw) != nrow(serie_data)) {
+      stop("Failed join")
+    }
+  } else {
+    N_predictor_sets <- 1
+    serie_data <- serie_data %>% mutate(.predictor_set = 1)
+    serie_data_distinct <- data.frame(.predictor_set = 1)
   }
 
   brmsdata_all <- crossing(serie_data_distinct, rate_data)

### 2020-05-20 R/posterior_plots_hmm.R

@@ -1,0 +1,28 @@
+model_color <- "#3993dd"
+data_color <- "#91171f"
+decoration_color <- "#ca5310"
+
+posterior_state_plot <- function(predictions_wide, series_id = 1:max(predictions_wide$.serie), observed_data = NULL) {
+  predictions_grouped <- predictions_wide %>%
+    filter(.serie %in% series_id) %>%
+    group_by_at(setdiff(names(predictions_wide), ".sample")) %>%
+    summarise(n_samples = n()) %>%
+    group_by_at(setdiff(names(predictions_wide), c(".predicted", ".sample"))) %>%
+    mutate(state_prob = n_samples / sum(n_samples)) %>%
+    ungroup()
+
+  if(!is.null(observed_data)) {
+    observed_data <- validate_brmshmmdata(observed_data)
+    observed_data <- observed_data$serie_data %>% filter(.serie %in% series_id)
+    observed_geom <- geom_line(data = observed_data, aes(x = .time, y = .observed), color = data_color, inherit.aes = FALSE, size = 2)
+  } else {
+    observed_geom <- NULL
+  }
+
+  print(predictions_grouped %>%
+          ggplot(aes(x = .time, y = .predicted, fill = state_prob)) + geom_raster() +
+          observed_geom +
+          scale_fill_gradient(low = "white", high = model_color, limits = c(0,1)) +
+          facet_wrap(~.serie)
+  )
+}

### 2020-05-20 dev/hmm_model_devel.Rmd
More encapsulation, fitting to Grein et al. dataset

@@
 ```{r}
-data <- hmm_simulator(20, 15, 2, use_noisy_states = FALSE, N_treatments = 0)
+data <- hmm_simulator(20, 15, 2, use_noisy_states = TRUE, N_treatments = 0)
 
@@
 ```{r}
+
+disease_levels <- c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")
+breathing_levels <- c("Discharge", disease_levels, "Death")
+
 data_grein <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
@@
   group_by(patient) %>%
-  mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
+  mutate(breathing_imputed = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
   ungroup() %>%
   mutate(
-    breathing = factor(breathing, levels = c("Discharge", "AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
+    breathing = factor(breathing, levels = breathing_levels, ordered = TRUE),
+    breathing_imputed = factor(breathing, levels = breathing_levels, ordered = TRUE)
   )
@@
 
+series_data <- data_grein %>% transmute(.serie = as.integer(patient), .time = day + 1, .observed = breathing)
+
+hidden_state_data <- data.frame(id = factor(breathing_levels, levels = breathing_levels)) %>% mutate( corresponding_obs = id)
+
+rate_data <- rbind(
+  data.frame(.from = disease_levels, .to = "Death", rate_group = "death"),
+  data.frame(.from = "AA", .to = "Discharge", rate_group = "improve_one"),
+  data.frame(.from = disease_levels[2 : length(disease_levels)], .to = disease_levels[1 : (length(disease_levels) - 1)], rate_group = "improve_one"),
+  data.frame(.from = disease_levels[1 : (length(disease_levels) - 1)], .to = disease_levels[2 : length(disease_levels)], rate_group = "worsen_one")
+) %>%
+  mutate(.from = factor(.from, levels = breathing_levels),
+         .to = factor(.to, levels = breathing_levels),
+         )
+
+initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+
+prior = brms::set_prior("normal(-2, 5)", "Intercept") +
+  brms::set_prior("normal(0, 3)", "b") +
+  brms::set_prior("normal(0, 1)", "sd", group = ".rate_id")
+
+grein_hmm <- brmshmmdata(~ rate_group + (1|.rate_id), series_data, rate_data, hidden_state_data, initial_states, prior = prior)
+
+fit_grein <- brmhmm(grein_hmm)
+
 ```
+
+```{r}
+# fit_grein$data <- grein_hmm
+# fit_grein$data_processed <- make_data_hmm(grein_hmm)
+
+#posterior_grein <- brms::posterior_epred(fit_grein)
+posterior_grein <- brms::posterior_predict(fit_grein)
+
+posterior_grein_wide <- prediction_to_wide_format(fit_grein$data, posterior_grein)
+
+posterior_grein_wide <- posterior_grein_wide %>% mutate(.predicted = factor(.predicted, levels = 1:length(levels(fit_grein$data$serie_data$.observed)), ordered = TRUE, labels = levels(fit_grein$data$serie_data$.observed)))
+
+step_size <- 6
+for(step in 1:ceiling(fit_grein$data_processed$standata$N_series / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(posterior_grein_wide, series_id, fit_grein$data) %>% print()
+
+}
+
+#TODO ppchecks: last state, pairwise state transitions, step up/down
+
+```
+
+```{r}
+d1 <- data.frame(x = 1, y = factor(1:3, levels = 1:3, labels = c("First","After first","After after first"), ordered = TRUE), f = c(0.1,0.2,0.3))
+
+d2 <- data.frame(x = c(1,2), y2 = factor(c(2,3), levels = 1:3, labels = c("First","After first","After after first"), ordered = TRUE), f = 0.5)
+
+ggplot(d1, aes(x = x, y = y, fill = f)) + geom_raster() + 
+  geom_line(data = d2, inherit.aes = FALSE, aes(x = x, y = y2, group = 1))
+
+```

### 2020-05-20 R/brmshmmdata.R

@@ -18,6 +18,45 @@ brmshmmdata <- function(formula,
       ))
 }
 
+validate_id <- function(id, name_for_message,
+                        force_unique = FALSE, force_no_gaps = FALSE, force_no_na = TRUE,
+                        reference_levels = NULL, reference_levels_for_message = NULL) {
+  if(force_no_na && any(is.na(id))) {
+    stop(paste0(name_for_message, " contains NA values"))
+  }
+  if(!is.factor(id)) {
+    if(is.double(id)) {
+      if(!identical(round(id), id)) {
+        stop(paste0(name_for_message, " includes non-integer numbers"))
+      }
+      id <- as.integer(id)
+    }
+    if(is.integer(id) || is.character(id)) {
+      if(is.null(reference_levels)) {
+        id <- factor(id, levels = sort(unique(id)))
+      } else {
+        na_orig <- is.na(id)
+        id <- factor(id, levels = reference_levels)
+        if(any(is.na(id) & !na_orig)) {
+          stop(paste0("Some components of ", name_for_message, " do not fit into reference levels (", reference_levels_for_message, " )."))
+        }
+      }
+    } else {
+      stop(paste0(name_for_message, " must be a factor, integer or character"))
+    }
+  } else if(!is.null(reference_levels) && !all(levels(id) == reference_levels)) {
+    stop(paste0("Levels of ", name_for_message, " are different than reference levels (", reference_levels_for_message, " )."))
+  }
+
+  if(force_unique && !identical(sort(unique(as.integer(id))), 1:length(id))) {
+    stop(paste0(name_for_message, " has to be unique and without gaps"))
+  } else if(force_no_gaps && !identical(sort(unique(as.integer(id))), 1:max(as.integer(id)))) {
+    stop(paste0(name_for_message, " has to be without gaps"))
+  }
+
+  id
+}
+
 validate_brmshmmdata <- function(d) {
   if(!inherits(d, "brmshmmdata")) {
     stop("Must be of class brmshmmdata")
@@ -26,40 +65,63 @@ validate_brmshmmdata <- function(d) {
   if(is.null(d$rate_data$.rate_id)) {
     d$rate_data <- d$rate_data %>% mutate(.rate_id = factor(1:n()))
   } else {
-    if(!is.factor(d$rate_data$.rate_id)) {
-      stop("rate_data$.rate_id must be a factor")
-    }
[…]

### 2020-05-20 R/hmm_simulator.R

@@ -252,13 +252,13 @@ hmm_simulator <- function(N_series, N_time, N_mid_states, use_noisy_states = FAL
         formula = model_formula,
         prior = prior,
         serie_data = data.frame(
-          .serie = series,
-          .time = times,
-          .observed = obs_states,
+          .serie = as.integer(series),
+          .time = as.integer(times),
+          .observed = as.integer(obs_states),
           treated = as.numeric(treated)),
         rate_data = data.frame(
-          .from = rates_from,
-          .to = rates_to,
+          .from = as.integer(rates_from),
+          .to = as.integer(rates_to),
           group = rates_group,
           rate_death = as.numeric(rates_group == "death"),
           rate_to_worsening = as.numeric(rates_group == "to_worsening"),

### 2020-05-20 R/make_stancode_hmm.R

@@ -175,7 +175,12 @@ stan_llh.rate_hmm <- function( ... ) {
     for(t in 2:serie_max_time[s]) {
       real col_norm;
       int ps = predictor_sets_rect[s, t - 1];
-      alpha[, t] = observation_probs[, obs_states_rect[s, t]] .* (transition_matrices_t[ps] * alpha[, t - 1]);
+      vector[N_states_hidden] transition_probs = (transition_matrices_t[ps] * alpha[, t - 1]);
+      if(obs_states_rect[s,t] != 0) {
+        alpha[, t] = observation_probs[, obs_states_rect[s, t]] .* transition_probs;
+      } else {
+        alpha[, t] = transition_probs;
+      }
       col_norm = max(alpha[, t]);
       alpha[, t] /= col_norm;
       alpha_log_norms[t] = log(col_norm) + alpha_log_norms[t - 1];

### 2020-05-20 R/make_standata_hmm.R

@@ -21,12 +21,13 @@ make_data_hmm <- function(brmshmmdata) {
 
     N_other_observations <- observed_state_data %>% select(starts_with("other_obs_")) %>% length()
     if(N_other_observations < 1) {
-      stop("Noisy states require other_obs_x columns")
+      stop("Noisy states require other_obs_XX columns")
     }
 
     noisy_states_other_obs <- observed_state_data %>%
       filter(is_noisy) %>%
       select(starts_with("other_obs_")) %>%
+      mutate_all(as.integer) %>%
       as.matrix()
   } else {
     noisy_states = numeric(0)
@@ -35,10 +36,7 @@ make_data_hmm <- function(brmshmmdata) {
   }
 
 
-  N_series <- max(serie_data$.serie)
-  if(!identical(sort(unique(serie_data$.serie)), 1:N_series)) {
-    stop("Patient IDs need to be consecutive and not duplicated")
-  }
+  N_series <- max(as.integer(serie_data$.serie))
 
   N_time <- max(serie_data$.time)
 
@@ -73,9 +71,9 @@ make_data_hmm <- function(brmshmmdata) {
   # brmsdata <- brmsdata_all %>% left_join(
   brmsdata <- brmsdata_all %>%
     arrange(.predictor_set, .rate_id) %>%
-    mutate(.predictor_id = 1:n(), .dummy = rnorm(n()))
+    mutate(.predictor_id = 1:n(), .dummy = seq(0, 1, length.out = n()))
 
-  obs_states <- serie_data$.observed
+  obs_states <- as.integer(serie_data$.observed)
   obs_states[is.na(serie_data$.observed)] <- 0
 
   rate_predictors <- array(NA_integer_, c(N_predictor_sets, N_rates))
@@ -88,12 +86,12 @@ make_data_hmm <- function(brmshmmdata) {
   serie_max_time <- array(0, N_series)
 
   for(o in 1:nrow(serie_data)) {
-    s <- serie_data$.serie[o]
+    s <- as.integer(serie_data$.serie[o])
     t <- serie_data$.time[o]
     if(!is.na(serie_data$.observed[o])) {
-      obs_states_rect[s, t] = serie_data$.observed[o];
-      serie_max_time[s] = max(serie_max_time[s], t);
+      obs_states_rect[s, t] = as.integer(serie_data$.observed[o]);
     }
+    serie_max_time[s] = max(serie_max_time[s], t);
     predictor_sets_rect[s, t] = serie_data$.predictor_set[o];
   }
 
[…]

### 2020-05-20 R/posterior_plots_hmm.R

@@ -4,25 +4,26 @@ decoration_color <- "#ca5310"
 
 posterior_state_plot <- function(predictions_wide, series_id = 1:max(predictions_wide$.serie), observed_data = NULL) {
   predictions_grouped <- predictions_wide %>%
+    select(-`.observed`) %>%
     filter(.serie %in% series_id) %>%
-    group_by_at(setdiff(names(predictions_wide), ".sample")) %>%
+    group_by_at(setdiff(names(predictions_wide), c(".sample", ".observed"))) %>%
     summarise(n_samples = n()) %>%
-    group_by_at(setdiff(names(predictions_wide), c(".predicted", ".sample"))) %>%
+    group_by_at(setdiff(names(predictions_wide), c(".observed", ".predicted", ".sample"))) %>%
     mutate(state_prob = n_samples / sum(n_samples)) %>%
     ungroup()
 
   if(!is.null(observed_data)) {
     observed_data <- validate_brmshmmdata(observed_data)
-    observed_data <- observed_data$serie_data %>% filter(.serie %in% series_id)
-    observed_geom <- geom_line(data = observed_data, aes(x = .time, y = .observed), color = data_color, inherit.aes = FALSE, size = 2)
+    observed_data_plot <- observed_data$serie_data %>% filter(.serie %in% series_id, !is.na(.observed))
+    observed_geom <- geom_line(data = observed_data_plot, aes(x = .time, y = .observed, group = .serie), color = data_color, inherit.aes = FALSE, size = 2)
   } else {
     observed_geom <- NULL
   }
 
-  print(predictions_grouped %>%
-          ggplot(aes(x = .time, y = .predicted, fill = state_prob)) + geom_raster() +
+  predictions_grouped %>%
+          ggplot(aes(x = .time, y = .predicted, fill = state_prob)) +
+          geom_raster() +
           observed_geom +
           scale_fill_gradient(low = "white", high = model_color, limits = c(0,1)) +
           facet_wrap(~.serie)
-  )
 }

### 2020-05-20 R/predict_hmm.R

@@ -15,18 +15,25 @@ posterior_epred.brmshmmfit <- function(fit, nsamples = NULL) {
 
   N_states_hidden <- data_hmm$standata$N_states_hidden
 
-  result_rect <- array(NA_integer_, c(max(pred_rawdata$serie_data$.time), max(pred_rawdata$serie_data$.serie), nsamples))
-  for(s in unique(pred_rawdata$serie_data$.serie)) {
+  result_rect <- array(NA_integer_, c(max(pred_rawdata$serie_data$.time), max(as.integer(pred_rawdata$serie_data$.serie)), nsamples))
+  for(s in unique(as.integer(pred_rawdata$serie_data$.serie))) {
 
-    result_rect[1, s, ] <- pred_rawdata$initial_states[s]
+    result_rect[1, s, ] <- as.integer(pred_rawdata$initial_states[s])
 
-    max_time <- fit$data$serie_data %>% filter(.serie == s) %>% pull(.time) %>% max()
+    max_time <- pred_rawdata$serie_data %>% filter(as.integer(.serie) == s) %>% pull(.time) %>% max()
+
     for(i in 1:nsamples) {
       state <- pred_rawdata$initial_states[s]
       for(t in 2:max_time) {
         ps <- data_hmm$standata$predictor_sets_rect[s, t]
+        if(ps == 0) {
+          stop(paste0("Element with no predictor at serie ",s,", time ", t))
+        }
         #cat(state, ":", ps, ",", i, ":", transition_matrices[state, , ps, i], "\n")
-        state <- sample(1:N_states_hidden, size = 1, prob = transition_matrices[state, , ps, i])
+        state <- sample.int(N_states_hidden, size = 1, prob = transition_matrices[state, , ps, i])
+        if(is.na(state)) {
+          stop("NA state")
+        }
         result_rect[t, s, i] <- state
       }
     }
@@ -35,9 +42,16 @@ posterior_epred.brmshmmfit <- function(fit, nsamples = NULL) {
   result <- array(NA_integer_, c(nsamples, nrow(pred_rawdata$serie_data)))
 
   for(o in 1:nrow(pred_rawdata$serie_data)) {
-    result[, o] <- result_rect[pred_rawdata$serie_data$.time[o], pred_rawdata$serie_data$.serie[o], ]
+    result_for_row <- result_rect[pred_rawdata$serie_data$.time[o], as.integer(pred_rawdata$serie_data$.serie[o]), ]
+    if(any(is.na(result_for_row))) {
[…]

### 2020-05-21 dev/hmm_grein.Rmd
Fitting multiple models to Grein et al. dataset

---
title: "Testing with Grein et al. data"
output: html_notebook
---

```{r setup}
library(rstan)
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

devtools::load_all()

cache_dir <- here::here("local_temp_data")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}
```

## Load the data

```{r}

disease_levels <- c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")
breathing_levels <- c("Discharge", disease_levels, "Death")

data_grein <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
  select(-Remdesivir_Last_Day) %>%
  pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
  filter(!is.na(breathing)) %>%
  transmute(
    patient = Patient,
    breathing = breathing,
    day = as.integer(day)) %>%
  group_by(patient) %>%
  mutate(breathing_imputed = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
  ungroup() %>%
  mutate(
    breathing = factor(breathing, levels = breathing_levels, ordered = TRUE),
    breathing_imputed = factor(breathing, levels = breathing_levels, ordered = TRUE)
  )

series_data <- data_grein %>% transmute(.serie = as.integer(patient), .time = day + 1, .observed = breathing)

```

## Only increase/decrease

```{r}
m1_hidden_state_data <- tibble(id = factor(breathing_levels, levels = breathing_levels)) %>% mutate( corresponding_obs = id)

m1_observed_state_data <- tibble(id = factor(breathing_levels, levels = breathing_levels), is_noisy = FALSE)

recode_group <- function(rate_data, group_name) {
  is_column <- paste0("is_", group_name) 
  id_column <- paste0(group_name, "_id")

  rate_data %>% mutate(
    !!is_column := as.double(rate_group == group_name),
    !!id_column := factor(
      if_else(rate_group == group_name, as.integer(.rate_id) - min(as.integer(.rate_id[rate_group == group_name])), as.integer(0)), 
      levels = 0:(sum(rate_group == group_name) - 1), 
      ordered = TRUE)
  )
}

m1_rate_data <- rbind(
  tibble(.from = disease_levels, .to = "Death", rate_group = "death"),
  tibble(.from = "AA", .to = "Discharge", rate_group = "improve_one"),
  tibble(.from = disease_levels[2 : length(disease_levels)], .to = disease_levels[1 : (length(disease_levels) - 1)], rate_group = "improve_one"),
  tibble(.from = disease_levels[1 : (length(disease_levels) - 1)], .to = disease_levels[2 : length(disease_levels)], rate_group = "worsen_one")
) %>%
  mutate(.rate_id = factor(1:n()),
         .from = factor(.from, levels = breathing_levels),
         .to = factor(.to, levels = breathing_levels)
  ) %>%
  recode_group("death") %>%
  recode_group("improve_one") %>%
  recode_group("worsen_one")

m1_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)

m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
  brms::set_prior("normal(0, 5)", "b") +
  brms::set_prior("normal(0, 2)", "sd")

m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
    (0 + is_improve_one | improve_one_id)
m1 <- brmshmmdata(m1_formula, series_data, 
                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
                  observed_state_data = m1_observed_state_data)

#make_stancode_hmm(m1)
#xx <- brms::make_standata(m1$formula, data = make_data_hmm(m1)$brmsdata)
m1_fit <- brmhmm(m1, cache_file = paste0(cache_dir,"/grein_m1.rds"))

summary(m1_fit$brmsfit)
```

```{r}
m1_epred_rect <- posterior_epred_rect(m1_fit)
m1_predicted_rect <- posterior_epred_to_predicted(m1_fit, m1_epred_rect)
m1_predicted <- posterior_rect_to_long(m1_fit, m1_predicted_rect)
```

```{r}
m1_predicted_df <- posterior_long_to_df(m1_fit$data, m1_predicted)

step_size <- 6
for(step in 1:ceiling(m1_fit$data_processed$standata$N_series / step_size)) {
  series_id <- ((step - 1) * step_size + 1) : (step* step_size)

  posterior_state_plot(m1_predicted_df, series_id, m1_fit$data) %>% print()

}
```

```{r}
pp_check_last_state(m1_fit, m1_predicted_rect)
pp_check_transitions_direction(m1_fit, m1_predicted_rect)
```

```{r}
pp_check_transitions(m1_fit, m1_predicted_rect, states_from = 2:4)
pp_check_transitions(m1_fit, m1_predicted_rect, states_from = 5:7)

```

## Increase / decrease + noise

```{r}
m2_hidden_state_data <- m1_hidden_state_data

m2_observed_state_data <- m1_observed_state_data

m2_rate_data <- rbind(
  tibble(.from = paste0(disease_levels, "_worsening"), .to = "Death", rate_group = "death"),
  tibble(.from = "AA_improving", .to = "Discharge", rate_group = "improve_one"),
  tibble(.from = paste0(disease_levels[2 : length(disease_levels)], "_improving"), 
             .to = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_improving"), 
             rate_group = "improve_one"),
  tibble(.from = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_worsening"), 
             .to = paste0(disease_levels[2 : length(disease_levels)], "_worsening"), 
             rate_group = "worsen_one"),
  tibble(.from = paste0(disease_levels,"_worsening"), 
             .to = paste0(disease_levels, "_improving"), 
             rate_group = "to_improving"),
  tibble(.from = paste0(disease_levels,"_improving"), 
             .to = paste0(disease_levels, "_worsening"), 
             rate_group = "to_worsening")
  ) %>%
  mutate(.rate_id = factor(1:n())) %>%
  recode_group("death") %>%
  recode_group("improve_one") %>%
  recode_group("worsen_one") %>%
  recode_group("to_improving") %>%
  recode_group("to_worsening")

m2_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")

m2_prior = m1_prior

m2_formula <- update.formula(m1_formula,
  ~ . + is_to_improving + is_to_worsening + (0 + is_to_improving | to_improving_id) + 
    (0 + is_to_worsening | to_worsening_id))

m2 <- brmshmmdata(m2_formula, series_data, 
                  m2_rate_data, m2_hidden_state_data, m2_initial_states, prior = m2_prior,
                  observed_state_data = m1_observed_state_data)

m2_fit <- brmhmm(m2, cache_file = paste0(cache_dir,"/grein_m2.rds"), control = list(adapt_delta = 0.9))

summary(m2_fit$brmsfit)
```

```{r}
m2_epred_rect <- posterior_epred_rect(m2_fit)
m2_predicted_rect <- posterior_epred_to_predicted(m2_fit, m2_epred_rect)
m2_predicted <- posterior_rect_to_long(m2_fit, m2_predicted_rect)
```

```{r}
m2_predicted_df <- posterior_long_to_df(m2_fit$data, m2_predicted)

step_size <- 6
for(step in 1:ceiling(m2_fit$data_processed$standata$N_series / step_size)) {
  series_id <- ((step - 1) * step_size + 1) : (step* step_size)

  posterior_state_plot(m2_predicted_df, series_id, m2_fit$data) %>% print()

}
```

```{r}
pp_check_last_state(m2_fit, m2_predicted_rect)
pp_check_transitions_direction(m2_fit, m2_predicted_rect)
pp_check_transitions_direction(m2_fit, m2_predicted_rect, states_from = 2:7)
```

```{r}
pp_check_transitions(m2_fit, m2_predicted_rect, states_from = 2:4)
pp_check_transitions(m2_fit, m2_predicted_rect, states_from = 5:7)

```

## Improving/worsening + noise

```{r}
m3_hidden_state_data <- m1_hidden_state_data

m3_observed_state_data <- tibble(id =  breathing_levels, 
         is_noisy = c(FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE),
         other_obs_1 = c(NA, disease_levels[2:6], disease_levels[1], NA), 
         other_obs_2 = c(NA, disease_levels[3:6], disease_levels[1:2], NA),
         other_obs_3 = c(NA, disease_levels[4:6], disease_levels[1:3], NA),
         other_obs_4 = c(NA, disease_levels[5:6], disease_levels[1:4], NA),
         other_obs_5 = c(NA, disease_levels[6], disease_levels[1:5], NA)

  ) %>% mutate(id = factor(id, levels = breathing_levels))

m3_rate_data <- m1_rate_data

m3_initial_states <- m1_initial_states

m3_prior = m1_prior

m3_formula <- m1_formula

m3 <- brmshmmdata(m3_formula, series_data, 
                  m3_rate_data, m3_hidden_state_data, m3_initial_states, prior = m3_prior,
                  observed_state_data = m3_observed_state_data)

m3_fit <- brmhmm(m3, cache_file = paste0(cache_dir,"/grein_m3.rds"), control = list(adapt_delta = 0.9))

summary(m3_fit)
summary(fit$brmsfit$fit, pars =  c("sensitivity", "other_observations_probs"), probs = c(0.025,0.975))$summary
```

```{r}
m3_epred_rect <- posterior_epred_rect(m3_fit)
m3_predicted_rect <- posterior_epred_to_predicted(m3_fit, m3_epred_rect)
m3_predicted <- posterior_rect_to_long(m3_fit, m3_predicted_rect)
```

```{r}
m3_predicted_df <- posterior_long_to_df(m3_fit$data, m3_predicted)

step_size <- 6
for(step in 1:ceiling(m3_fit$data_processed$standata$N_series / step_size)) {
  series_id <- ((step - 1) * step_size + 1) : (step* step_size)

  posterior_state_plot(m3_predicted_df, series_id, m3_fit$data) %>% print()

}
```

```{r}
pp_check_last_state(m3_fit, m3_predicted_rect)
pp_check_transitions_direction(m3_fit, m3_predicted_rect)
pp_check_transitions_direction(m3_fit, m3_predicted_rect, states_from = 2:7)
```

```{r}
pp_check_transitions(m3_fit, m3_predicted_rect, states_from = 2:4)
pp_check_transitions(m3_fit, m3_predicted_rect, states_from = 5:7)

```

### 2020-05-21 dev/hmm_model_devel.Rmd
Fitting multiple models to Grein et al. dataset

--- 2020-05-20 dev/hmm_model_devel.Rmd
@@
 ```
-
-## Apply to Grein et al. dataset
-
-```{r}
-data_grein <- readxl::read_excel(here::here("public_data", "Grein_et_al.xlsx"), na = c("")) %>%
-  select(-Remdesivir_Last_Day) %>%
-  pivot_longer(starts_with("Day_"), names_prefix = "Day_", names_to = "day", values_to = "breathing") %>%
-  filter(!is.na(breathing)) %>%
-  transmute(
-    patient = Patient,
-    breathing = breathing,
-    day = as.integer(day)) %>%
-  group_by(patient) %>%
-  mutate(breathing = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
-  ungroup() %>%
-  mutate(
-    breathing = factor(breathing, levels = c("Discharge", "AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO", "Death"), ordered = TRUE)
-  )
-
-```
-
-```{r}
-
-```

### 2020-05-21 R/brms_tools.R

@@ -1,0 +1,28 @@
+brm_with_cache <- function(cache_file = NULL, ...) {
+  fit <- NULL
+  stancode <- brms::make_stancode(...)
+  standata <- brms::make_standata(...)
+  if(!is.null(cache_file) && file.exists(cache_file)) {
+    cache_contents <- readRDS(cache_file)
+    if(!is.list(cache_contents) || is.null(cache_contents$stancode) ||
+       is.null(cache_contents$standata) || is.null(cache_contents$fit)) {
+      message("Invalid cache content")
+    } else {
+      if(identical(stancode, cache_contents$stancode) &&
+         identical(standata, cache_contents$standata)) {
+        fit <- cache_contents$fit
+      } else {
+        message("Cache file out of date, refitting")
+      }
+    }
+  }
+
+  if(is.null(fit)) {
+    fit <- brms::brm(...)
+    if(!is.null(cache_file) && brms:::contains_samples(fit)) {
+      saveRDS(loo::nlist(stancode, standata, fit), file = cache_file)
+    }
+  }
+
+  fit
+}

### 2020-05-21 R/brmshmmdata.R

@@ -85,13 +85,12 @@ validate_brmshmmdata <- function(d) {
 
 
   if(is.null(d$observed_state_data)) {
+
     d$hidden_state_data$corresponding_obs <-
-      validate_id(d$hidden_state_data$corresponding_obs, "corresponding_obs", force_no_gaps = TRUE)
+      validate_id(d$hidden_state_data$corresponding_obs, "corresponding_obs (without observed data)", force_no_gaps = TRUE)
 
-    all_corresponding_states <- d$hidden_state_data %>% pull(corresponding_obs) %>%
-      unique() %>% sort()
 
-    d$observed_state_data <- data.frame(id = all_corresponding_states, is_noisy = FALSE)
+    d$observed_state_data <- data.frame(id = levels(d$hidden_state_data$corresponding_obs), is_noisy = FALSE)
   } else {
     d$hidden_state_data$corresponding_obs <-
       validate_id(d$hidden_state_data$corresponding_obs, "corresponding_obs", force_no_gaps = TRUE,

### 2020-05-21 R/brmshmmfit.R

@@ -1,9 +1,10 @@
-brmhmm <- function(brmshmmdata, ...) {
+brmhmm <- function(brmshmmdata, cache_file = NULL, ...) {
   d <- validate_brmshmmdata(brmshmmdata)
 
   prepdata <- make_data_hmm(d)
 
-  bfit <- brms::brm(
+  bfit <- brm_with_cache(
+    cache_file = cache_file,
     formula = d$formula,
     family = rate_hmm_family,
     data = prepdata$brmsdata,
@@ -27,4 +28,17 @@ validate_brmshmmfit <- function(fit) {
   fit
 }
 
+summary.brmshmmfit <- function(fit) {
+  validate_brmshmmfit(fit)
+  structure(list(
+    brmssummary = summary(fit$brmsfit)#,
+    #hmmsummary = summary(fit$brmsfit$fit, pars =  c("sensitivity", "other_observations_probs"))
+    ),
+    class = "summary.brmshmmfit"
+  )
+}
 
+print.summary.brmshmmfit <- function(s) {
+  print(s$brmssummary)
+  #print(s$hmmsummary$summary)
+}

### 2020-05-21 R/evaluation_tools.R

@@ -1,8 +1,17 @@
 missing_arg <- function() quote(expr=)
 
+n_missing_args <- function(n) {
+  res <- list()
+  for(i in 1:n) {
+    res[[i]] <- missing_arg()
+  }
+  res
+
+}
+
 order_within_samples <- function(x, samples, n_samples = 99) {
   thinned_samples <- sample(samples, n_samples)
-  # For discrete parameters, I need to uniformly samples across the positions that are equal 
+  # For discrete parameters, I need to uniformly samples across the positions that are equal
   sum(thinned_samples < x) + rdunif(1, a = 0, b = sum(thinned_samples == x)) + 1
 }
 

### 2020-05-21 R/posterior_plots_hmm.R

@@ -2,14 +2,35 @@ model_color <- "#3993dd"
 data_color <- "#91171f"
 decoration_color <- "#ca5310"
 
-posterior_state_plot <- function(predictions_wide, series_id = 1:max(predictions_wide$.serie), observed_data = NULL) {
-  predictions_grouped <- predictions_wide %>%
+posterior_state_plot <- function(predictions_df, series_id = 1:max(predictions_df$.serie),
+                                 observed_data = NULL,
+                                 predicted_labels = NULL) {
+
+  if(is.factor(predictions_df$.predicted) && !is.null(predicted_labels)) {
+    warning("predicted_labels ignored as .predicted is already a factor")
+  }
+
+  if(!is.factor(predictions_df$.predicted)) {
+    if(!is.null(observed_data) && is.null(predicted_labels)) {
+      predicted_labels <- levels(observed_data$serie_data$.observed)
+    }
+
+    if(!is.null(predicted_labels)) {
+      predicted_levels <- 1:length(predicted_labels)
+    } else {
+      predicted_levels <- 1:max(predictions_df$.predicted)
+      predicted_labels <- as.character(predicted_levels)
+    }
+  }
+
+  predictions_grouped <- predictions_df %>%
     select(-`.observed`) %>%
     filter(.serie %in% series_id) %>%
-    group_by_at(setdiff(names(predictions_wide), c(".sample", ".observed"))) %>%
+    group_by_at(setdiff(names(predictions_df), c(".sample", ".observed"))) %>%
     summarise(n_samples = n()) %>%
-    group_by_at(setdiff(names(predictions_wide), c(".observed", ".predicted", ".sample"))) %>%
-    mutate(state_prob = n_samples / sum(n_samples)) %>%
+    group_by_at(setdiff(names(predictions_df), c(".observed", ".predicted", ".sample"))) %>%
+    mutate(state_prob = n_samples / sum(n_samples),
+           .predicted = factor(.predicted, levels = predicted_levels, labels = predicted_labels)) %>%
     ungroup()
 
   if(!is.null(observed_data)) {
@@ -26,4 +47,110 @@ posterior_state_plot <- function(predictions_wide, series_id = 1:max(predictions
           observed_geom +
           scale_fill_gradient(low = "white", high = model_color, limits = c(0,1)) +
[…]

### 2020-05-29 dev/hmm_grein.Rmd
Testing m4 with stable states

@@
 ```{r}
-m2_hidden_state_data <- m1_hidden_state_data
+m2_hidden_state_data <- rbind(
+  tibble(base = disease_levels) %>% mutate(corresponding_obs = base) %>% 
+    crossing(tibble(state_group = c("_improving","_worsening"))) %>% 
+    mutate(id = factor(paste0(base, state_group))) %>%
+    select(-base, -state_group),
+  tibble(id = c("Death","Discharge")) %>% mutate(corresponding_obs = id)
+)
 
@@
 ```
+
+## Improving/worsening/stable
+
+```{r}
+m4_hidden_state_data <- rbind(
+  tibble(base = disease_levels) %>% mutate(corresponding_obs = base) %>% 
+    crossing(tibble(state_group = c("_improving", "_stable", "_worsening"))) %>% 
+    mutate(id = factor(paste0(base, state_group))) %>%
+    select(-base, -state_group),
+  tibble(id = c("Death","Discharge")) %>% mutate(corresponding_obs = id)
+)
+
+m4_observed_state_data <- m1_observed_state_data
+
+m4_rate_data <- rbind(
+  tibble(.from = paste0(disease_levels, "_worsening"), .to = "Death", rate_group = "death"),
+  tibble(.from = "AA_improving", .to = "Discharge", rate_group = "improve_one"),
+  tibble(.from = paste0(disease_levels[2 : length(disease_levels)], "_improving"), 
+             .to = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_improving"), 
+             rate_group = "improve_one"),
+  tibble(.from = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_worsening"), 
+             .to = paste0(disease_levels[2 : length(disease_levels)], "_worsening"), 
+             rate_group = "worsen_one"),
+  tibble(.from = paste0(disease_levels,"_stable"), 
+             .to = paste0(disease_levels, "_improving"), 
+             rate_group = "to_improving"),
+  tibble(.from = paste0(disease_levels,"_stable"), 
+             .to = paste0(disease_levels, "_worsening"), 
+             rate_group = "to_worsening"),
+  tibble(.from = paste0(disease_levels,"_worsening"), 
+             .to = paste0(disease_levels, "_stable"), 
+             rate_group = "worsening_to_stable"),
+  tibble(.from = paste0(disease_levels,"_improving"), 
+             .to = paste0(disease_levels, "_stable"), 
+             rate_group = "improving_to_stable")
+  ) %>%
+  mutate(.rate_id = factor(1:n())) %>%
+  recode_group("death") %>%
+  recode_group("improve_one") %>%
+  recode_group("worsen_one") %>%
+  recode_group("to_improving") %>%
+  recode_group("to_worsening") %>%
+  recode_group("improving_to_stable") %>%
+  recode_group("worsening_to_stable")
+
+m4_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")
+
+m4_prior = m1_prior
+
+m4_formula <- update.formula(m2_formula,
+  ~ . + is_improving_to_stable + is_worsening_to_stable + (0 + is_improving_to_stable | improving_to_stable_id) + 
+    (0 + is_worsening_to_stable | worsening_to_stable_id))
+
+m4 <- brmshmmdata(m4_formula, series_data, 
+                  m4_rate_data, m4_hidden_state_data, m4_initial_states, prior = m4_prior,
+                  observed_state_data = m1_observed_state_data)
+
+m4_fit <- brmhmm(m4, cache_file = paste0(cache_dir,"/grein_m4.rds"), control = list(adapt_delta = 0.9))
+
+summary(m4_fit$brmsfit)
+```
+
+```{r}
+m4_epred_rect <- posterior_epred_rect(m4_fit)
+m4_predicted_rect <- posterior_epred_to_predicted(m4_fit, m4_epred_rect)
+m4_predicted <- posterior_rect_to_long(m4_fit, m4_predicted_rect)
+```
+
+```{r}
+m4_predicted_df <- posterior_long_to_df(m4_fit$data, m4_predicted)
+
+step_size <- 6
+for(step in 1:ceiling(m4_fit$data_processed$standata$N_series / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m4_predicted_df, series_id, m4_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m4_fit, m4_predicted_rect)
+pp_check_transitions_direction(m4_fit, m4_predicted_rect)
+pp_check_transitions_direction(m4_fit, m4_predicted_rect, states_from = 2:7)
+```
+
+```{r}
+pp_check_transitions(m4_fit, m4_predicted_rect, states_from = 2:4)
+pp_check_transitions(m4_fit, m4_predicted_rect, states_from = 5:7)
+
+```

### 2020-06-27 dev/hmm_grein.Rmd
Tweaks to m4

@@
              .to = paste0(disease_levels, "_stable"), 
-             rate_group = "improving_to_stable")
+             rate_group = "improving_to_stable"),
+  tibble(.from = paste0(disease_levels,"_improving"), 
+             .to = paste0(disease_levels, "_worsening"), 
+             rate_group = "improving_to_worsening"),
+  tibble(.from = paste0(disease_levels,"_worsening"), 
+             .to = paste0(disease_levels, "_improving"), 
+             rate_group = "worsening_to_improving")
   ) %>%
@@
   recode_group("improving_to_stable") %>%
-  recode_group("worsening_to_stable")
+  recode_group("worsening_to_stable") %>%
+  recode_group("improving_to_worsening") %>%
+  recode_group("worsening_to_improving")
 
@@
 m4_formula <- update.formula(m2_formula,
-  ~ . + is_improving_to_stable + is_worsening_to_stable + (0 + is_improving_to_stable | improving_to_stable_id) + 
-    (0 + is_worsening_to_stable | worsening_to_stable_id))
+  ~ . + is_improving_to_stable + is_worsening_to_stable + is_improving_to_worsening + is_worsening_to_improving +
+    (0 + is_improving_to_stable | improving_to_stable_id) + 
+    (0 + is_worsening_to_stable | worsening_to_stable_id) +
+    (0 + is_improving_to_worsening | worsening_to_improving_id) + 
+    (0 + is_worsening_to_improving | improving_to_worsening_id) 
+  )
 
@@
 
-m4_fit <- brmhmm(m4, cache_file = paste0(cache_dir,"/grein_m4.rds"), control = list(adapt_delta = 0.9))
+m4_fit <- brmhmm(m4, cache_file = paste0(cache_dir,"/grein_m4.rds"), control = list(adapt_delta = 0.9), cores = 1)
 
 summary(m4_fit$brmsfit)
+```
+
+```{r}
+# m4_hidden_state_data <- rbind(
+#   tibble(base = disease_levels) %>% mutate(corresponding_obs = base) %>% 
+#     crossing(tibble(state_group = c("_improving", "_stable", "_worsening"))) %>% 
+#     mutate(id = factor(paste0(base, state_group))) %>%
+#     select(-base, -state_group),
+#   tibble(id = c("Death","Discharge")) %>% mutate(corresponding_obs = id)
+# )
+# 
+# m4_observed_state_data <- m1_observed_state_data
+# 
+# m4_rate_data <- rbind(
+#   tibble(.from = paste0(disease_levels, "_worsening"), .to = "Death", rate_group = "death"),
+#   tibble(.from = "AA_improving", .to = "Discharge", rate_group = "improve_one"),
+#   tibble(.from = paste0(disease_levels[2 : length(disease_levels)], "_improving"), 
+#              .to = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_improving"), 
+#              rate_group = "improve_one"),
+#   tibble(.from = paste0(disease_levels[1 : (length(disease_levels) - 1)],"_worsening"), 
+#              .to = paste0(disease_levels[2 : length(disease_levels)], "_worsening"), 
+#              rate_group = "worsen_one"),
+#   tibble(.from = paste0(disease_levels,"_stable"), 
+#              .to = paste0(disease_levels, "_improving"), 
+#              rate_group = "to_improving"),
+#   tibble(.from = paste0(disease_levels,"_stable"), 
+#              .to = paste0(disease_levels, "_worsening"), 
+#              rate_group = "to_worsening"),
+#   tibble(.from = paste0(disease_levels,"_worsening"), 
+#              .to = paste0(disease_levels, "_stable"), 
+#              rate_group = "worsening_to_stable"),
+#   tibble(.from = paste0(disease_levels,"_improving"), 
+#              .to = paste0(disease_levels, "_stable"), 
+#              rate_group = "improving_to_stable"),
+#   tibble(.from = paste0(disease_levels,"_improving"), 
+#              .to = paste0(disease_levels, "_worsening"), 
+#              rate_group = "improving_to_worsening"),
+#   tibble(.from = paste0(disease_levels,"_worsening"), 
+#              .to = paste0(disease_levels, "_improving"), 
+#              rate_group = "worsening_to_improving")
+#   ) %>%
+#   mutate(.rate_id = factor(1:n())) %>%
+#   recode_group("death") %>%
+#   recode_group("improve_one") %>%
+#   recode_group("worsen_one") %>%
+#   recode_group("to_improving") %>%
+#   recode_group("to_worsening") %>%
+#   recode_group("improving_to_stable") %>%
+#   recode_group("worsening_to_stable")
+#  
+# 
+# m4_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")
+# 
+# m4_prior = m1_prior
+# 
+# m4_formula <- update.formula(m2_formula,
+#   ~ . + is_improving_to_stable + is_worsening_to_stable + (0 + is_improving_to_stable | improving_to_stable_id) + 
+#     (0 + is_worsening_to_stable | worsening_to_stable_id))
+#   
+# m4 <- brmshmmdata(m4_formula, series_data, 
+#                   m4_rate_data, m4_hidden_state_data, m4_initial_states, prior = m4_prior,
+#                   observed_state_data = m1_observed_state_data)
+# 
+# m4_fit <- brmhmm(m4, cache_file = paste0(cache_dir,"/grein_m4.rds"), control = list(adapt_delta = 0.9))
+# 
+# summary(m4_fit$brmsfit)
 ```

### 2020-07-09 dev/data_preprocessing.Rmd
Work on preprocessing

---
title: "Developing data load and processing"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
```

```{r}
raw_data_dir <- "H:/raw/"
if(!dir.exists(raw_data_dir)) {
  stop("Raw data dir not found")
}

raw_data_map <- readxl::read_excel(paste0(raw_data_dir,"mapping.xlsx"), sheet = "HospitalMapping") %>% filter(!is.na(File))
```

```{r}
file_relative <- raw_data_map$File[2]
hospital_id <- raw_data_map$HospitalID[2]

file <- paste0(raw_data_dir, file_relative)

patient_col_types <- c("text","numeric","text","date","date", #ID - Date of admission
                                   rep("text", 12), #Transferred from - Smoking
                                   rep("numeric", 5), # BMI - Albumin
                                   "text", "text", "date", "text", "text") #Discontinued medication - Transferred to

patient_data_raw <- readxl::read_excel(file, sheet = "Patients", range = "A2:AA100", na = "NA", col_types = patient_col_types)

logical_column <- function(x) {
  if(any( !is.na(x) && !(tolower(x) %in% c("yes","no"))) ) {
    stop("Invalid logical column contents")
  }
  tolower(x) == "yes"
}

outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")

patient_data_typed <- patient_data_raw %>% 
  filter(!is.na(Age), !is.na(Sex)) %>%
  transmute(
    hospital_id = hospital_id, 
    age = Age, 
    sex = factor(Sex, levels = c("M","F")), 
    symptom_onset = `Date of symptom onset`, 
    admission = `Date of admission`, 
    transferred_from = `Transferred from`, 
    admitted_for_covid = logical_column(`Admitted for Covid`), 
    ischemic_heart_disease = logical_column(`Ischemic Heart Disease`),
    has_hypertension_drugs = tolower(`Hypertension drugs`) != "no",
    n_hypertension_drugs = if_else(has_hypertension_drugs,
                                   as.integer(if_else(has_hypertension_drugs, `Hypertension drugs`, NA_character_)), #Inner if_else to avoid warning
                                    0L),
    heart_failure = logical_column(`Heart failure`),
    COPD = logical_column(COPD),
    asthma = logical_column(`Asthma Bronchiale`),
    other_lung_disease = logical_column(`Other lung disease`),
    diabetes = logical_column(Diabetes),
    renal_disease = logical_column(`Renal Disease`),
    liver_disease = logical_column(`Liver Disease`),
    smoking = logical_column(Smoking),
    BMI = BMI,
    NYHA = NYHA,
    creatinin = `Creatinin [μmol/L]`,
    pt_inr = `PT INR (Quick)`,
    albumin = `Albumin [g/L]`,
    discontinued_medication = tolower(`Discontinued medication`) != "no",
    discontinued_medication_reason = if_else(discontinued_medication, `Discontinued medication`, NA_character_),
    best_supportive_care_from = if_else(tolower(`Best supportive care from:`) == "no", lubridate::as_date(NA), 
                                        janitor::excel_numeric_to_date(as.numeric(
                                          if_else(tolower(`Best supportive care from:`) == "no", NA_character_, `Best supportive care from:`) # Inner if_else to avoid warning
                                          ))),
    last_record = `Date of last record`,
    outcome = factor(tolower(`Outcome (Discharged/Hospitalized/Transferred/Death)`), levels = tolower(outcome_labels), labels = outcome_labels),
    transferred_to = `Transferred to`
  )
```

### 2020-07-09 dev/hmm_grein.Rmd
Work on preprocessing

--- 2020-05-29 dev/hmm_grein.Rmd
@@
     day = as.integer(day)) %>%
-  group_by(patient) %>%
-  mutate(breathing_imputed = if_else(breathing == "ND", breathing[which.max(day[breathing %in% c("AA", "LoO2", "HiO2", "NIPPV", "MV", "ECMO")])], breathing)) %>%
-  ungroup() %>%
   mutate(
     breathing = factor(breathing, levels = breathing_levels, ordered = TRUE),
-    breathing_imputed = factor(breathing, levels = breathing_levels, ordered = TRUE)
   )
@@
 
+data_grein %>% ggplot(aes(x = day, y = patient, fill = breathing)) + geom_tile()
+
+data_grein %>% group_by(breathing) %>% summarise(pocet = n())
+mean(data_grein$breathing == "LoO2", na.rm = TRUE)
 ```
@@
 ```
+
+## No NIPPV, improving/worsening/stable
+
+```{r}
+
+disease_levels_m5 <- c("AA", "O2", "MV", "ECMO")
+breathing_levels_m5 <- c("Discharge", disease_levels_m5, "Death")
+
+m5_observed_state_data <- tibble(id = factor(breathing_levels_m5, levels = breathing_levels_m5), is_noisy = FALSE)
+
+m5_hidden_state_data <- m4_hidden_state_data %>% filter(corresponding_obs != "NIPPV") %>% droplevels()
+
+m5_hidden_state_data <- rbind(
+  tibble(base = disease_levels_m5) %>% mutate(corresponding_obs = base) %>% 
+    crossing(tibble(state_group = c("_improving", "_stable", "_worsening"))) %>% 
+    mutate(id = factor(paste0(base, state_group))) %>%
+    select(-base, -state_group),
+  tibble(id = c("Death","Discharge")) %>% mutate(corresponding_obs = id)
+)
+
+m5_rate_data <- rbind(
+  tibble(.from = paste0(disease_levels_m5, "_worsening"), .to = "Death", rate_group = "death"),
+  tibble(.from = "AA_improving", .to = "Discharge", rate_group = "improve_one"),
+  tibble(.from = paste0(disease_levels_m5[2 : length(disease_levels_m5)], "_improving"), 
+             .to = paste0(disease_levels_m5[1 : (length(disease_levels_m5) - 1)],"_improving"), 
+             rate_group = "improve_one"),
+  tibble(.from = paste0(disease_levels_m5[1 : (length(disease_levels_m5) - 1)],"_worsening"), 
+             .to = paste0(disease_levels_m5[2 : length(disease_levels_m5)], "_worsening"), 
+             rate_group = "worsen_one"),
+  tibble(.from = paste0(disease_levels_m5,"_stable"), 
+             .to = paste0(disease_levels_m5, "_improving"), 
+             rate_group = "to_improving"),
+  tibble(.from = paste0(disease_levels_m5,"_stable"), 
+             .to = paste0(disease_levels_m5, "_worsening"), 
+             rate_group = "to_worsening"),
+  tibble(.from = paste0(disease_levels_m5,"_worsening"), 
+             .to = paste0(disease_levels_m5, "_stable"), 
+             rate_group = "worsening_to_stable"),
+  tibble(.from = paste0(disease_levels_m5,"_improving"), 
+             .to = paste0(disease_levels_m5, "_stable"), 
+             rate_group = "improving_to_stable")
+  ) %>%
+  mutate(.rate_id = factor(1:n())) %>%
+  recode_group("death") %>%
+  recode_group("improve_one") %>%
+  recode_group("worsen_one") %>%
+  recode_group("to_improving") %>%
+  recode_group("to_worsening") %>%
+  recode_group("improving_to_stable") %>%
+  recode_group("worsening_to_stable")
+
+series_data_m5 <- series_data %>% mutate(.observed = factor(case_when(
+  .observed == "NIPPV" ~ "MV", 
+  .observed %in% c("LoO2","HiO2") ~ "O2",
+  TRUE ~ as.character(.observed)), 
+  levels = breathing_levels_m5, ordered = TRUE))
+
+m5_initial_states <- series_data_m5 %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")
+
+m5_prior <- brms::set_prior("normal(-2, 5)", "Intercept") +
+  brms::set_prior("normal(0, 5)", "b")
+
+m5_formula <- ~ factor(.rate_id) #m4_formula
+
+m5 <- brmshmmdata(m5_formula, series_data_m5, 
+                  m5_rate_data, m5_hidden_state_data, m5_initial_states, prior = m5_prior,
+                  observed_state_data = m5_observed_state_data)
+
+m5_fit <- brmhmm(m5, cache_file = paste0(cache_dir,"/grein_m5.rds"), control = list(adapt_delta = 0.9))
+
+summary(m5_fit$brmsfit)
+```
+
+```{r}
+m5_epred_rect <- posterior_epred_rect(m5_fit)
+m5_predicted_rect <- posterior_epred_to_predicted(m5_fit, m5_epred_rect)
+m5_predicted <- posterior_rect_to_long(m5_fit, m5_predicted_rect)
+```
+
+```{r}
+m5_predicted_df <- posterior_long_to_df(m5_fit$data, m5_predicted)
+
+step_size <- 6
+for(step in 1:ceiling(m5_fit$data_processed$standata$N_series / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m5_predicted_df, series_id, m5_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m5_fit, m5_predicted_rect)
+pp_check_transitions_direction(m5_fit, m5_predicted_rect)
+pp_check_transitions_direction(m5_fit, m5_predicted_rect, states_from = 2:6)
+```
+
+```{r}
+pp_check_transitions(m5_fit, m5_predicted_rect, states_from = 2:4)
+pp_check_transitions(m5_fit, m5_predicted_rect, states_from = 5:6)
+
+```

### 2020-07-27 dev/data_preprocessing.Rmd
Work on preprocessing

@@
 ```{r}
-file_relative <- raw_data_map$File[2]
-hospital_id <- raw_data_map$HospitalID[2]
+i <- 1
+file_relative <- raw_data_map$File[1]
+hospital_id <- raw_data_map$HospitalID[1]
 
@@
 
+patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
+
 logical_column <- function(x) {
@@
 
-patient_data_typed <- patient_data_raw %>% 
-  filter(!is.na(Age), !is.na(Sex)) %>%
+patient_data_typed <- patient_data_raw_usable %>% 
   transmute(
     hospital_id = hospital_id, 
+    patient_id = `Patient ID`,
     age = Age, 
     sex = factor(Sex, levels = c("M","F")), 
-    symptom_onset = `Date of symptom onset`, 
-    admission = `Date of admission`, 
+    symptom_onset = lubridate::as_date(`Date of symptom onset`), 
+    admission = lubridate::as_date(`Date of admission`), 
     transferred_from = `Transferred from`, 
@@
                                           ))),
-    last_record = `Date of last record`,
+    last_record = lubridate::as_date(`Date of last record`),
     outcome = factor(tolower(`Outcome (Discharged/Hospitalized/Transferred/Death)`), levels = tolower(outcome_labels), labels = outcome_labels),
@@
   )
+
+invalid_outcome <- is.na(patient_data_raw_usable$`Outcome (Discharged/Hospitalized/Transferred/Death)`) != is.na(patient_data_typed$outcome)
+if(any(invalid_outcome)) {
+  print(patient_data_raw_usable %>% filter(invalid_outcome))
+  stop("Invalid outcome")
+}
 ```
+
+```{r}
+progression_data_raw <- readxl::read_excel(file, sheet = "Disease Progression", range = "A1:CU2539", na = c("NA",""), col_types = c(c("text","text", "skip", "text", "text"), rep("text", 94)), col_names = TRUE)
+
+progression_data <- progression_data_raw %>% 
+  rename(patient_id = `Patient ID`, indicator = Indicator, first_day = First_Day, unit = Unit) %>%
+  filter(!is.na(indicator))
+
+# Remove the "Don't edit examples" comment
+progression_data$patient_id[2:9] <- NA_character_
+
+# Spread Patient ID and Start date
+last_patient_id <- progression_data$patient_id[1]
+last_first_day <- progression_data$first_day[1]
+for(i in 2:nrow(progression_data)) {
+  if(!is.na(progression_data$patient_id[i])) {
+    last_patient_id <- progression_data$patient_id[i]
+    last_first_day <- progression_data$first_day[i]
+  } else {
+    progression_data$patient_id[i] <- last_patient_id
+    progression_data$first_day[i] <- last_first_day
+  }
+}
+
+progression_data <- progression_data %>% filter(patient_id %in% patient_data_typed$patient_id)
+
+progression_data <- progression_data %>% 
+  pivot_longer(starts_with("Day_"), names_to = "day", values_to = "value", names_prefix = "Day_") %>%
+  mutate(day = as.integer(day))
+
+progression_data <- progression_data %>% 
+  filter(!is.na(indicator))
+
+# Check dates
+dates_data <- progression_data %>% 
+  filter(indicator == "Date")
+test_dates <- dates_data %>%
+  mutate(value = as.integer(value), first_day = as.integer(first_day)) %>%
+  filter(value != first_day + day - 1 )
+
+if(nrow(test_dates) > 0) {
+  stop("Dates check failed")
+}
+
+breathing_data <- progression_data %>% filter(indicator == "Breathing") %>%
+  mutate(breathing = factor(tolower(value), levels = tolower(c("AA","Oxygen", "NIPPV","MV","ECMO")),labels = c("AA","Oxygen", "NIPPV","MV","ECMO"), ordered = TRUE))
+
+breathing_unrecognized <- breathing_data %>% filter(is.na(breathing) != is.na(value))
+if(nrow(breathing_unrecognized) > 0) {
+  stop("Breathing NAs check failed")
+}
+
+breathing_data <- breathing_data %>% select(-value, -indicator, -unit)
+
+correct_first_day_admission <- function(patient_data, progression_data) {
+  nrow_before <- nrow(progression_data)
+  progression_data <- progression_data %>% 
+    inner_join(patient_data %>% select(patient_id, admission), by = "patient_id") %>%
+    mutate(first_day = janitor::excel_numeric_to_date(as.numeric(first_day)),
+           correction = as.integer(first_day - admission),
+           day = day + correction - 1)
+
+  if(nrow(progression_data) != nrow_before) {
+    stop("Invalid join")
+  }
+
+  progression_data %>%
+    select(-admission, -first_day, -correction)
+}
+
+breathing_data <- correct_first_day_admission(patient_data_typed, breathing_data) %>% filter(!is.na(breathing))
+
+adverse_event_data <- progression_data %>% 
+  filter(indicator == "Adverse events")
+
+dummy_markers <- c("Drugs", "Other marker (optional)", "Other Covid drug", "Yet another drug")
+
+marker_data <- progression_data %>% filter(!(indicator %in% c("Date", "Breathing", "Adverse events", dummy_markers))) %>% rename(marker = indicator)
+marker_data <- correct_first_day_admission(patient_data_typed, marker_data)
+
+# Correct Excel error
+marker_data <- marker_data %>%
+  mutate(marker = case_when(
+    grepl("^Supp\\. O[0-9]*$", marker) ~ "Supp. O2",
+    grepl("^SpO[0-9]*$", marker) ~ "SpO2",
+    grepl("^IL-[0-9]*$", marker) ~ "IL-6",
+    TRUE ~ marker))
+
+if(any(is.na(marker_data$marker))) {
+  stop("NA in markers")
+}
+
+marker_map <- c("PCR" = "pcr_value", 
+                "Supp. O2" = "oxygen_flow",
+                "SpO2" = "SpO2",
+                "SpO2 native" = "SpO2_native",
+                "Horowitz index" = "Horowitz_index",
+                "PEEP" = "PEEP",
+                "Ferritin" = "ferritin",
+                "D-dimer" = "d_dimer",
+                "CRP" = "CRP",
+                "IL-6" = "IL_6",
+                "Lymphocyte count" = "lymphocyte_count",
+                "Hydroxychloroquine" = "hcq",
+                "Azithromycin" = "az",
+                "Kaletra" = "kaletra",
+                "Tocilizumab" = "tocilizumab",
+                "Amoxicillin/Clavulanate" = "amoxiclav",
+                "Piperacilin/tazobactam" = "piperacilin_tazobactam",
+                "Clarithromycin" = "clarithromycin",
+                "ofloxacin" = "ofloxacin",
+                "ceftriaxone" = "ceftriaxone",
+                "cefuroxim" = "cefuroxim",
+                "ampicillin/sulbactam" = "ampicillin_sulbactam"
+                )
+#Check NAs (TODO)
+unrecgonized_markers <- marker_data %>% filter(!(tolower(marker) %in% tolower(names(marker_map))))
+
+if(nrow(unrecgonized_markers) > 0) {
+  print(unrecgonized_markers)
+  stop("Unrecognized markers")
+}
+
+names(marker_map) <- tolower(names(marker_map))
+
+marker_data <- marker_data %>% mutate(marker = marker_map[tolower(marker)])
+
+#TODO - handle PCR specifically
+
+```

### 2020-08-16 dev/data_preprocessing.Rmd
More preprocessing

@@
 ```{r}
-i <- 1
-file_relative <- raw_data_map$File[1]
-hospital_id <- raw_data_map$HospitalID[1]
+i <- 2
+file_relative <- raw_data_map$File[i]
+hospital_id <- raw_data_map$HospitalID[i]
 
@@
 outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")
+
+# Special handling of badly filled outcomes
+if(hospital_id == "QKuFp") {
+  patient_data_raw_usable <- patient_data_raw_usable %>%
+    mutate(`Outcome (Discharged/Hospitalized/Transferred/Death)` =
+             if_else(grepl("^439[0-9]{2}$", `Outcome (Discharged/Hospitalized/Transferred/Death)`), 
+                     "Hospitalized",
+                     `Outcome (Discharged/Hospitalized/Transferred/Death)`))
+}
 
@@
                 "Lymphocyte count" = "lymphocyte_count",
+                "Leukocyte count" = "leukocyte_count",
+                "Leukocytes" = "leukocyte_count",
+                "Leukocyte" = "leukocyte_count",
                 "Hydroxychloroquine" = "hcq",
@@
                 "cefuroxim" = "cefuroxim",
-                "ampicillin/sulbactam" = "ampicillin_sulbactam"
+                "cefotaxim" = "cefotaxim",
+                "cefepim" = "cefepime",
+                "cefepime" = "cefepime",
+                "Vancomycin" = "vancomycin",
+                "ampicillin/sulbactam" = "ampicillin_sulbactam",
+                "zinkorot" = "zinc",
+                "sulfamethoxazol" = "sulfamethoxazol",
+                "trimethoprim" = "trimethoprim",
+                "Meropenem" = "meropenem",
+                "Fluconazol" = "fluconazol",
+                "Covalescent plasma" = "convalescent_plasma",
+                "Anidulafungin" = "anidulafungin"
                 )
@@
 if(nrow(unrecgonized_markers) > 0) {
+  print(unique(unrecgonized_markers$marker))
   print(unrecgonized_markers)
@@
 
-#TODO - handle PCR specifically
+# Special handling for PCR
+pcr_data <- marker_data %>% filter(marker == "pcr_value")
+
+pcr_values_positive <- c("pos","neg", "neg and pos", "susp.", "susp", "poz")
+pcr_value_negative <- "neg"
+
+wrong_pcr <- pcr_data %>% filter(!is.na(value),!(tolower(value) %in% pcr_values_positive), !tolower(value) %in% pcr_value_negative, !grepl("^[0-9]+$",value))
+if(nrow(wrong_pcr) > 0) {
+  print(wrong_pcr)
+  stop("Wrong PCR")
+}
+
+pcr_data_long <- pcr_data %>% mutate(pcr_positive = as.numeric(tolower(value) %in% pcr_values_positive),
+                    pcr_value = as.numeric(if_else(grepl("^[0-9]+$",value), value, NA_character_))) %>%
+  select(-value, -marker) %>%
+  pivot_longer(c("pcr_value","pcr_positive"), names_to = "marker", values_to = "value") %>%
+  mutate(censored = "none")
+
+# Handle NT for SpO2_native
+marker_data <- marker_data %>% 
+  mutate(value = if_else(marker == "SpO2_native" & value == "NT", "< 92", value) )
+
+# Some special cases
+# TODO ask at hospitel
+if(hospital_id == "QKuFp") {
+  marker_data <- marker_data %>% mutate(value = if_else(marker == "SpO2" & value == "NEG", NA_character_, value))
+}
+
+# Handle censoring on everything except PCR
+marker_data <- marker_data %>% filter(marker != "pcr_value") %>%
+  mutate(censored = case_when(startsWith(value, "<") ~ "right",
+                              startsWith(value, ">") ~ "left",
+                              TRUE ~ "none"),
+         value_parsed = as.numeric(gsub("^[<>]? *", "", value)))
+
+bad_parse_value <- marker_data %>% filter(is.na(value_parsed) & !is.na(value))
+if(nrow(bad_parse_value) > 0) {
+  print(bad_parse_value)
+  stop("Bad parse value")
+}
+
+# Join with PCR data
+marker_data <- marker_data %>%
+  mutate(value = value_parsed) %>%
+  select(-value_parsed) %>%
+  rbind(pcr_data_long)
+
+# Some special handling
+if(hospital_id == "QKuFp") {
+  marker_data <- marker_data %>% 
+    mutate(unit = if_else(marker == "leukocyte_count" & "unit" == "unit", "10^9/l", unit))
+}
 
 ```
+
+```{r}
+# TODO check units (and convert if necessary)
+marker_data %>% group_by(marker)
+
+```

### 2020-08-17 dev/data_preprocessing.Rmd
Translation for preprocessing

@@
 
+Notes to resolve:
+- Missing "Covid medication before hospitalization" in "Old" version.
+
 ```{r}
-i <- 2
+i <- 5
 file_relative <- raw_data_map$File[i]
 hospital_id <- raw_data_map$HospitalID[i]
+lang <- raw_data_map$Lang[i]
+file_version <- raw_data_map$Version[i]
 
@@
 
+if(file_version == "From Sheet") {
+  file_version <- readxl::read_excel(file, sheet = version_sheet(lang), range = "C8:C8", col_types = "text", col_names = "version")$version 
+}
+
+if(!grepl("^(Old|1\\.[2-4]\\.[0-9])$", file_version)) {
+  print(file_version)
+  stop("Invalid version")
+}
+
+if(file_version == "Old") {
+  n_middle_text_columns <- 12
+  last_column <- "Z"
+} else {
+  n_middle_text_columns <- 13
+  last_column <- "AA"
+}
+
 patient_col_types <- c("text","numeric","text","date","date", #ID - Date of admission
-                                   rep("text", 12), #Transferred from - Smoking
+                                   rep("text", n_middle_text_columns), #Transferred from - Smoking
                                    rep("numeric", 5), # BMI - Albumin
-                                   "text", "text", "date", "text", "text") #Discontinued medication - Transferred to
+                                   "text", "text", "date", "text") #Discontinued medication - Outcome
 
-patient_data_raw <- readxl::read_excel(file, sheet = "Patients", range = "A2:AA100", na = "NA", col_types = patient_col_types)
+patients_range <- paste0("A2:",last_column, "100")
+
+patient_data_raw <- readxl::read_excel(file, sheet = patient_sheet(lang), range = patients_range, na = "NA", col_types = patient_col_types) %>% 
+  translate_patient_columns(lang)
+
+patient_data_raw <- patient_data_raw %>%
+  mutate(`Outcome (Discharged/Hospitalized/Transferred/Death)` = translate_outcomes(`Outcome (Discharged/Hospitalized/Transferred/Death)`, lang))
+
+if(file_version == "Old") {
+  patient_data_raw <- patient_data_raw %>% mutate(`Covid Medications before admission` = NA_character_)
+}
+```
+
+```{r}
 
@@
 logical_column <- function(x) {
-  if(any( !is.na(x) && !(tolower(x) %in% c("yes","no"))) ) {
+  if(any( !is.na(x) && !(tolower(x) %in% c(yes_variants, no_variants))) ) {
     stop("Invalid logical column contents")
   }
-  tolower(x) == "yes"
+  tolower(x) %in% yes_variants
 }
@@
     ischemic_heart_disease = logical_column(`Ischemic Heart Disease`),
-    has_hypertension_drugs = tolower(`Hypertension drugs`) != "no",
+    has_hypertension_drugs = !(tolower(`Hypertension drugs`) %in% no_variants),
     n_hypertension_drugs = if_else(has_hypertension_drugs,
@@
     albumin = `Albumin [g/L]`,
-    discontinued_medication = tolower(`Discontinued medication`) != "no",
+    discontinued_medication = !(tolower(`Discontinued medication`) %in% no_variants),
     discontinued_medication_reason = if_else(discontinued_medication, `Discontinued medication`, NA_character_),
-    best_supportive_care_from = if_else(tolower(`Best supportive care from:`) == "no", lubridate::as_date(NA), 
+    best_supportive_care_from = if_else(tolower(`Best supportive care from:`) %in% no_variants, lubridate::as_date(NA), 
                                         janitor::excel_numeric_to_date(as.numeric(
-                                          if_else(tolower(`Best supportive care from:`) == "no", NA_character_, `Best supportive care from:`) # Inner if_else to avoid warning
+                                          if_else(tolower(`Best supportive care from:`) %in% no_variants, NA_character_, `Best supportive care from:`) # Inner if_else to avoid warning
                                           ))),
     last_record = lubridate::as_date(`Date of last record`),
-    outcome = factor(tolower(`Outcome (Discharged/Hospitalized/Transferred/Death)`), levels = tolower(outcome_labels), labels = outcome_labels),
-    transferred_to = `Transferred to`
+    outcome = factor(tolower(`Outcome (Discharged/Hospitalized/Transferred/Death)`), levels = tolower(outcome_labels), labels = outcome_labels)
   )
@@
 ```{r}
-progression_data_raw <- readxl::read_excel(file, sheet = "Disease Progression", range = "A1:CU2539", na = c("NA",""), col_types = c(c("text","text", "skip", "text", "text"), rep("text", 94)), col_names = TRUE)
+progression_data_raw <- readxl::read_excel(file, sheet = progression_sheet(lang), range = "A1:CU2539", na = c("NA",""), col_types = c(c("text","text", "skip", "text", "text"), rep("text", 94)), col_names = TRUE)
 
 progression_data <- progression_data_raw %>% 
+  translate_progression_columns(lang) %>%
   rename(patient_id = `Patient ID`, indicator = Indicator, first_day = First_Day, unit = Unit) %>%
-  filter(!is.na(indicator))
+  filter(!is.na(indicator)) %>%
+  mutate(indicator = translate_markers(indicator, lang))
 
@@
 breathing_data <- progression_data %>% filter(indicator == "Breathing") %>%
-  mutate(breathing = factor(tolower(value), levels = tolower(c("AA","Oxygen", "NIPPV","MV","ECMO")),labels = c("AA","Oxygen", "NIPPV","MV","ECMO"), ordered = TRUE))
+  mutate(
+    # Special cases
+    value = case_when(value == "Oxygen/NIPPV" ~ "NIPPV",
+                      TRUE ~ value),
+    # Translate to factor
+    breathing = factor(tolower(value), levels = tolower(c("AA","Oxygen", "NIPPV","MV","ECMO")),labels = c("AA","Oxygen", "NIPPV","MV","ECMO"), ordered = TRUE))
 
@@
 if(nrow(breathing_unrecognized) > 0) {
+  print(unique(breathing_unrecognized$value))
   stop("Breathing NAs check failed")
@@
                 "IL-6" = "IL_6",
+                "PCT" = "procalcitonin",
                 "Lymphocyte count" = "lymphocyte_count",
@@
                 "Covalescent plasma" = "convalescent_plasma",
-                "Anidulafungin" = "anidulafungin"
+                "Anidulafungin" = "anidulafungin",
+                "Isoprinosine" = "isoprinosine"
                 )
@@
 marker_data <- marker_data %>% 
-  mutate(value = if_else(marker == "SpO2_native" & value == "NT", "< 92", value) )
+  mutate(value = if_else(marker == "SpO2_native" & tolower(value) == "nt", "< 92", value) )
 
 # Some special cases
-# TODO ask at hospitel
+# TODO ask at hospital
 if(hospital_id == "QKuFp") {
   marker_data <- marker_data %>% mutate(value = if_else(marker == "SpO2" & value == "NEG", NA_character_, value))
+} else if(hospital_id == "YqNbe") {
+  marker_data <- marker_data %>% mutate(value = if_else(marker %in% c("hcq","az") & value == "ano", ">100", value))
 }

### 2020-08-17 R/translation.R

@@ -1,0 +1,77 @@
+translations_file <-  here::here("public_data", "Translations.xlsx")
+sheet_name_translations <- readxl::read_excel(translations_file, "SheetNames")
+patient_columns_translations <- readxl::read_excel(translations_file, "PatientColumns")
+outcome_translations <- readxl::read_excel(translations_file, "Outcomes")
+progression_columns_translations <- readxl::read_excel(translations_file, "ProgressionColumns")
+marker_translations <- readxl::read_excel(translations_file, "Markers")
+
+
+yes_variants <- c("yes","y", "ano", "a")
+no_variants <- c("no","n", "ne")
+
+
+version_sheet <- function(lang) {
+  if_else(lang == "En", "I need help",
+          sheet_name_translations %>% filter(Sheet == "version") %>% pull(Cz)
+  )
+}
+
+patients_sheet <- function(lang) {
+  if_else(lang == "En", "Patients", "Pacienti")
+}
+
+progression_sheet <- function(lang) {
+  if_else(lang == "En", "Disease Progression",
+          sheet_name_translations %>% filter(Sheet == "progression") %>% pull(Cz)
+  )
+}
+
+translate_patient_columns <- function(patient_data_raw, lang) {
+  if(lang == "En") {
+    patient_data_raw
+  } else {
+    translation <- patient_columns_translations$Column
+    names(translation) <- patient_columns_translations$Cz
+    names(patient_data_raw) <- translation[names(patient_data_raw)]
+
+    patient_data_raw
+  }
+}
+
+translate_outcomes <- function(outcomes, lang) {
+  if(lang == "En") {
+    outcomes
+  } else {
+    translation <- outcome_translations$Outcome
+    names(translation) <- outcome_translations$Cz
+
+    translation[outcomes]
+  }
+
+}
+
+translate_progression_columns <- function(progression_data_raw, lang) {
+  if(lang == "En") {
+    progression_data_raw
+  } else {
+    translation <- progression_columns_translations$Column
+    names(translation) <- progression_columns_translations$Cz
+    names_to_change <- names(progression_data_raw) %in% names(translation)
+    names(progression_data_raw)[names_to_change] <- translation[names(progression_data_raw)[names_to_change]]
+
+    progression_data_raw
+  }
+}
+
+translate_markers <- function(markers, lang) {
+  if(lang == "En") {
+    markers
+  } else {
+    translation <- marker_translations$Marker
[…]

### 2020-08-18 dev/data_preprocessing.Rmd
Preprocessing refactor

--- 2020-07-27 dev/data_preprocessing.Rmd
@@
 
+Notes to resolve:
+- Missing "Covid medication before hospitalization" in "Old" version.
+
 ```{r}
-i <- 1
-file_relative <- raw_data_map$File[1]
-hospital_id <- raw_data_map$HospitalID[1]
+for(i in 1:nrow(raw_data_map)) {
+  file_relative <- raw_data_map$File[i]
+  hospital_id <- raw_data_map$HospitalID[i]
+  lang <- raw_data_map$Lang[i]
+  file_version <- raw_data_map$Version[i]
 
@@
 
-patient_col_types <- c("text","numeric","text","date","date", #ID - Date of admission
-                                   rep("text", 12), #Transferred from - Smoking
-                                   rep("numeric", 5), # BMI - Albumin
-                                   "text", "text", "date", "text", "text") #Discontinued medication - Transferred to
+  cat("Processing ", i, "-", hospital_id, "\n")
 
-patient_data_raw <- readxl::read_excel(file, sheet = "Patients", range = "A2:AA100", na = "NA", col_types = patient_col_types)
-
-patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
-
-logical_column <- function(x) {
-  if(any( !is.na(x) && !(tolower(x) %in% c("yes","no"))) ) {
-    stop("Invalid logical column contents")
-  }
-  tolower(x) == "yes"
+  read_raw_data(file, hospital_id, lang, file_version)
 }
 
-outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")
-
-patient_data_typed <- patient_data_raw_usable %>% 
-  transmute(
-    hospital_id = hospital_id, 
-    patient_id = `Patient ID`,
-    age = Age, 
-    sex = factor(Sex, levels = c("M","F")), 
-    symptom_onset = lubridate::as_date(`Date of symptom onset`), 
-    admission = lubridate::as_date(`Date of admission`), 
-    transferred_from = `Transferred from`, 
-    admitted_for_covid = logical_column(`Admitted for Covid`), 
-    ischemic_heart_disease = logical_column(`Ischemic Heart Disease`),
-    has_hypertension_drugs = tolower(`Hypertension drugs`) != "no",
-    n_hypertension_drugs = if_else(has_hypertension_drugs,
-                                   as.integer(if_else(has_hypertension_drugs, `Hypertension drugs`, NA_character_)), #Inner if_else to avoid warning
-                                    0L),
-    heart_failure = logical_column(`Heart failure`),
-    COPD = logical_column(COPD),
-    asthma = logical_column(`Asthma Bronchiale`),
-    other_lung_disease = logical_column(`Other lung disease`),
-    diabetes = logical_column(Diabetes),
-    renal_disease = logical_column(`Renal Disease`),
-    liver_disease = logical_column(`Liver Disease`),
-    smoking = logical_column(Smoking),
-    BMI = BMI,
-    NYHA = NYHA,
-    creatinin = `Creatinin [μmol/L]`,
-    pt_inr = `PT INR (Quick)`,
-    albumin = `Albumin [g/L]`,
-    discontinued_medication = tolower(`Discontinued medication`) != "no",
-    discontinued_medication_reason = if_else(discontinued_medication, `Discontinued medication`, NA_character_),
-    best_supportive_care_from = if_else(tolower(`Best supportive care from:`) == "no", lubridate::as_date(NA), 
-                                        janitor::excel_numeric_to_date(as.numeric(
-                                          if_else(tolower(`Best supportive care from:`) == "no", NA_character_, `Best supportive care from:`) # Inner if_else to avoid warning
-                                          ))),
-    last_record = lubridate::as_date(`Date of last record`),
-    outcome = factor(tolower(`Outcome (Discharged/Hospitalized/Transferred/Death)`), levels = tolower(outcome_labels), labels = outcome_labels),
-    transferred_to = `Transferred to`
-  )
-
-invalid_outcome <- is.na(patient_data_raw_usable$`Outcome (Discharged/Hospitalized/Transferred/Death)`) != is.na(patient_data_typed$outcome)
-if(any(invalid_outcome)) {
-  print(patient_data_raw_usable %>% filter(invalid_outcome))
-  stop("Invalid outcome")
-}
 ```
@@
 ```{r}
-progression_data_raw <- readxl::read_excel(file, sheet = "Disease Progression", range = "A1:CU2539", na = c("NA",""), col_types = c(c("text","text", "skip", "text", "text"), rep("text", 94)), col_names = TRUE)
-
-progression_data <- progression_data_raw %>% 
-  rename(patient_id = `Patient ID`, indicator = Indicator, first_day = First_Day, unit = Unit) %>%
-  filter(!is.na(indicator))
-
-# Remove the "Don't edit examples" comment
-progression_data$patient_id[2:9] <- NA_character_
-
-# Spread Patient ID and Start date
-last_patient_id <- progression_data$patient_id[1]
-last_first_day <- progression_data$first_day[1]
-for(i in 2:nrow(progression_data)) {
-  if(!is.na(progression_data$patient_id[i])) {
-    last_patient_id <- progression_data$patient_id[i]
-    last_first_day <- progression_data$first_day[i]
-  } else {
-    progression_data$patient_id[i] <- last_patient_id
-    progression_data$first_day[i] <- last_first_day
-  }
-}
-
-progression_data <- progression_data %>% filter(patient_id %in% patient_data_typed$patient_id)
-
-progression_data <- progression_data %>% 
-  pivot_longer(starts_with("Day_"), names_to = "day", values_to = "value", names_prefix = "Day_") %>%
-  mutate(day = as.integer(day))
-
-progression_data <- progression_data %>% 
-  filter(!is.na(indicator))
-
-# Check dates
-dates_data <- progression_data %>% 
-  filter(indicator == "Date")
-test_dates <- dates_data %>%
-  mutate(value = as.integer(value), first_day = as.integer(first_day)) %>%
-  filter(value != first_day + day - 1 )
-
-if(nrow(test_dates) > 0) {
-  stop("Dates check failed")
-}
-
-breathing_data <- progression_data %>% filter(indicator == "Breathing") %>%
-  mutate(breathing = factor(tolower(value), levels = tolower(c("AA","Oxygen", "NIPPV","MV","ECMO")),labels = c("AA","Oxygen", "NIPPV","MV","ECMO"), ordered = TRUE))
-
-breathing_unrecognized <- breathing_data %>% filter(is.na(breathing) != is.na(value))
-if(nrow(breathing_unrecognized) > 0) {
-  stop("Breathing NAs check failed")
-}
-
-breathing_data <- breathing_data %>% select(-value, -indicator, -unit)
-
-correct_first_day_admission <- function(patient_data, progression_data) {
-  nrow_before <- nrow(progression_data)
-  progression_data <- progression_data %>% 
-    inner_join(patient_data %>% select(patient_id, admission), by = "patient_id") %>%
-    mutate(first_day = janitor::excel_numeric_to_date(as.numeric(first_day)),
-           correction = as.integer(first_day - admission),
-           day = day + correction - 1)
-
-  if(nrow(progression_data) != nrow_before) {
-    stop("Invalid join")
-  }
-
-  progression_data %>%
-    select(-admission, -first_day, -correction)
-}
-
-breathing_data <- correct_first_day_admission(patient_data_typed, breathing_data) %>% filter(!is.na(breathing))
-
-adverse_event_data <- progression_data %>% 
-  filter(indicator == "Adverse events")
-
-dummy_markers <- c("Drugs", "Other marker (optional)", "Other Covid drug", "Yet another drug")
-
-marker_data <- progression_data %>% filter(!(indicator %in% c("Date", "Breathing", "Adverse events", dummy_markers))) %>% rename(marker = indicator)
-marker_data <- correct_first_day_admission(patient_data_typed, marker_data)
-
-# Correct Excel error
-marker_data <- marker_data %>%
-  mutate(marker = case_when(
-    grepl("^Supp\\. O[0-9]*$", marker) ~ "Supp. O2",
-    grepl("^SpO[0-9]*$", marker) ~ "SpO2",
-    grepl("^IL-[0-9]*$", marker) ~ "IL-6",
-    TRUE ~ marker))
-
-if(any(is.na(marker_data$marker))) {
-  stop("NA in markers")
-}
-
-marker_map <- c("PCR" = "pcr_value", 
-                "Supp. O2" = "oxygen_flow",
-                "SpO2" = "SpO2",
-                "SpO2 native" = "SpO2_native",
-                "Horowitz index" = "Horowitz_index",
-                "PEEP" = "PEEP",
-                "Ferritin" = "ferritin",
-                "D-dimer" = "d_dimer",
-                "CRP" = "CRP",
-                "IL-6" = "IL_6",
-                "Lymphocyte count" = "lymphocyte_count",
-                "Hydroxychloroquine" = "hcq",
-                "Azithromycin" = "az",
-                "Kaletra" = "kaletra",
-                "Tocilizumab" = "tocilizumab",
-                "Amoxicillin/Clavulanate" = "amoxiclav",
-                "Piperacilin/tazobactam" = "piperacilin_tazobactam",
-                "Clarithromycin" = "clarithromycin",
-                "ofloxacin" = "ofloxacin",
-                "ceftriaxone" = "ceftriaxone",
-                "cefuroxim" = "cefuroxim",
-                "ampicillin/sulbactam" = "ampicillin_sulbactam"
-                )
-#Check NAs (TODO)
-unrecgonized_markers <- marker_data %>% filter(!(tolower(marker) %in% tolower(names(marker_map))))
-
-if(nrow(unrecgonized_markers) > 0) {
-  print(unrecgonized_markers)
-  stop("Unrecognized markers")
-}
-
-names(marker_map) <- tolower(names(marker_map))
-
-marker_data <- marker_data %>% mutate(marker = marker_map[tolower(marker)])
-
-#TODO - handle PCR specifically
 
 ```
+
+```{r}
+
+```
+
+```{r}
+# TODO check units (and convert if necessary)
+marker_data %>% group_by(marker)
+
+```

### 2020-08-18 R/data_mapping.R

@@ -1,0 +1,66 @@
+dummy_markers <- c("Drugs", "Other marker (optional)", "Other Covid drug", "Yet another drug")
+
+
+marker_map <- c(
+      ## Markers
+        "PCR" = "pcr_value",
+        "Supp. O2" = "oxygen_flow",
+        "SpO2" = "SpO2",
+        "SpO2 native" = "SpO2_native",
+        "Horowitz index" = "Horowitz_index",
+        "PEEP" = "PEEP",
+        "Ferritin" = "ferritin",
+        "D-dimer" = "d_dimer",
+        "CRP" = "CRP",
+        "IL-6" = "IL_6",
+        "PCT" = "procalcitonin",
+        "Lymphocyte count" = "lymphocyte_count",
+        "Leukocyte count" = "leukocyte_count",
+        "Leukocytes" = "leukocyte_count",
+        "Leukocyte" = "leukocyte_count",
+        "urea" = "urea",
+      ## Covid-specific drugs
+        "Hydroxychloroquine" = "hcq",
+        "Kaletra" = "kaletra",
+        "Azithromycin" = "az",
+        "Tocilizumab" = "tocilizumab",
+        "Covalescent plasma" = "convalescent_plasma",
+      ## Antibiotics
+        "Amoxicillin/Clavulanate" = "amoxiclav",
+        "Augmentin" = "amoxiclav",
+        "Piperacilin/tazobactam" = "piperacilin_tazobactam",
+        "Clarithromycin" = "clarithromycin",
+        "ofloxacin" = "ofloxacin",
+        "ceftriaxone" = "ceftriaxone",
+        "cefuroxim" = "cefuroxim",
+        "cefotaxim" = "cefotaxim",
+        "cefepim" = "cefepime",
+        "cefepime" = "cefepime",
+        "Vancomycin" = "vancomycin",
+        "ampicillin/sulbactam" = "ampicillin_sulbactam",
+        "trimethoprim" = "trimethoprim",
+        "Ciprofloxacin" = "ciprofloxacin",
+        "Meropenem" = "meropenem",
+        "Metronidazol" = "Metronidazol",
+      ## Other drugs
+        "Fluconazol" = "fluconazol",
+        "zinkorot" = "zinc",
+        "Anidulafungin" = "anidulafungin",
+        "sulfamethoxazol" = "sulfamethoxazol",
[…]

### 2020-08-18 R/data_preprocessing.R
More preprocessing

@@ -1,13 +1,16 @@
 #Throughout we use global variables from data_mapping.R
 
-read_raw_data <- function(file, hospital_id, lang, file_version) {
+read_raw_data <- function(file, hospital_id, lang, file_version, remove_examples = TRUE) {
   file_version <- update_and_check_version(file, lang, file_version)
-  patient_data <- read_patient_data(file, hospital_id, lang, file_version)
+  patient_data <- read_patient_data(file, hospital_id, lang, file_version, remove_examples)
+
   progression_data <- read_progression_data(file, hospital_id, lang, file_version, patient_data)
 
+  # TODO code adverse events
   loo::nlist(patient_data,
              breathing_data = progression_data$breathing_data,
-             marker_data = progression_data$marker_data)
+             marker_data = progression_data$marker_data,
+             adverse_events_data = progression_data$adverse_events_data)
 }
 
 
@@ -24,10 +27,15 @@ update_and_check_version <- function(file, lang, file_version) {
   file_version
 }
 
-read_patient_data <- function(file, hospital_id, lang, file_version) {
+read_patient_data <- function(file, hospital_id, lang, file_version, remove_examples) {
   patient_data_raw <- read_patient_data_raw(file, hospital_id, lang, file_version)
 
   patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
+
+  if(remove_examples) {
+    patient_data_raw_usable <- patient_data_raw_usable %>%
+      filter(!startsWith(`Patient ID`, "Example"), !grepl("^P.*klad", `Patient ID`))
+  }
 
 
   logical_column <- function(x) {
@@ -135,7 +143,8 @@ read_progression_data <- function(file, hospital_id, lang, file_version, patient
     translate_progression_columns(lang) %>%
     rename(patient_id = `Patient ID`, indicator = Indicator, first_day = First_Day, unit = Unit) %>%
     filter(!is.na(indicator)) %>%
-    mutate(indicator = translate_markers(indicator, lang))
+    mutate(indicator = translate_markers(indicator, lang),
+           hospital_id = hospital_id)
 
   # Remove the "Don't edit examples" comment
   progression_data$patient_id[2:9] <- NA_character_
@@ -209,7 +218,7 @@ read_progression_data <- function(file, hospital_id, lang, file_version, patient
 
[…]

### 2020-08-27 dev/data_preprocessing.Rmd
More preprocessing
Merging patients/wards

@@
 
-raw_data_map <- readxl::read_excel(paste0(raw_data_dir,"mapping.xlsx"), sheet = "HospitalMapping") %>% filter(!is.na(File))
+mapping_file <- paste0(raw_data_dir,"mapping.xlsx")
+
+raw_data_map <- readxl::read_excel(mapping_file, sheet = "HospitalMapping") %>% filter(!is.na(File))
 ```
@@
 ```{r}
+hospital_data <- list()
 for(i in 1:nrow(raw_data_map)) {
@@
 
-  read_raw_data(file, hospital_id, lang, file_version)
+  hospital_data[[i]] <- read_raw_data(file, hospital_id, lang, file_version)
 }
@@
 ```{r}
+complete_data <- join_hospital_data(hospital_data)
+# TODO transfers
+# TODO check:last day vs. breathing data
+# MAx. day vs. read columns
+# TODO check units (and convert if necessary)
+differing_units <- complete_data$marker_data %>% group_by(marker) %>% filter(length(unique(unit)) > 1) %>% 
+  group_by(marker, unit) %>% summarise(count = n(), .groups = "drop")
+
+if(nrow(differing_units) > 0) {
+  print(differing_units)
+  print(complete_data$marker_data %>% group_by(marker) %>% filter(length(unique(unit)) > 1, !is.na(value)) %>% 
+  ggplot(aes(x = value)) + geom_histogram() + facet_wrap(marker ~ unit, scales = "free")
+  )
+
+  stop("Differing units")
+}
 
@@
 
+## Checking unit conversions
+
+Check values for all markers with converted units
+
 ```{r}
+for(conversion in unit_conversions) {
+  for(marker_to_check in conversion$markers) {
+    cat(" ==== ", marker_to_check, "=====\n")
+    complete_data$marker_data %>% filter(marker == marker_to_check, !is.na(value)) %>% 
+      pull(value) %>% unique() %>% print()
+    print(
+      complete_data$marker_data %>% filter(marker == marker_to_check, !is.na(value)) %>% 
+        ggplot(aes(x = value)) + geom_histogram() + facet_wrap(marker ~ unit, scales = "free")
+    )
+  }
+}
+```
+
+```{r}
+complete_data$marker_data %>% filter(marker == "CRP", unit == "mg/l") %>% pull(hospital_id) %>%   unique()
 
@@
 
+## Merging patients/wards
+
 ```{r}
-# TODO check units (and convert if necessary)
-marker_data %>% group_by(marker)
+patients_to_merge <- readxl::read_excel(mapping_file, sheet = "PatientsToMerge")
+hospitals_to_merge <- readxl::read_excel(mapping_file, sheet = "HospitalsToMerge")
+```
 
+Check for candidates to merge
+
+```{r}
+i <- 2; j <- 3
+for(i in 1:(length(hospital_data) - 1)) {
+  for(j in (i + 1):length(hospital_data)) {
+    #cat(i,",", j,"\n")
+
+    candidate_overlap <-  check_patient_overlap(hospital_data[[i]], hospital_data[[j]], 
+                                                patients_to_merge, hospitals_to_merge) %>% 
+      select(hospital_id1, patient_id1, hospital_id2, patient_id2, 
+             symptom_onset1, symptom_onset2, outcome1, outcome2, 
+             BMI1, BMI2, last_record1, last_record2,
+             transferred_from1, transferred_to2, transferred_from2, transferred_to1) 
+    if(nrow(candidate_overlap) > 0) {
+      # Workaround, issue https://github.com/rstudio/rstudio/issues/4439
+      for(n in names(candidate_overlap)) {
+        names(candidate_overlap[[n]]) <- NULL
+      }
+      print(candidate_overlap)
+    }
+  }
+}
+#  mutate(d = as.numeric(symptom_onset2 - symptom_onset1))
 ```
+
+```{r}
+complete_data <- merge_patients(complete_data, patients_to_merge)
+
+complete_data$patient_data %>% filter(hospital_id == "Oriik", patient_id == 3)
+```
+
+```{r}
+complete_data <- merge_hospitals(complete_data, hospitals_to_merge)
+
+complete_data$patient_data %>% filter(hospital_id == "Oriik")
+```

### 2020-08-27 R/data_mapping.R

@@ -19,6 +19,7 @@ marker_map <- c(
         "Leukocytes" = "leukocyte_count",
         "Leukocyte" = "leukocyte_count",
         "urea" = "urea",
+        "IgG" = "IgG",
       ## Covid-specific drugs
         "Hydroxychloroquine" = "hcq",
         "Kaletra" = "kaletra",
@@ -26,14 +27,18 @@ marker_map <- c(
         "Tocilizumab" = "tocilizumab",
         "Covalescent plasma" = "convalescent_plasma",
       ## Antibiotics
+        "Amoxicillin/Clavulanic" = "amoxiclav",
         "Amoxicillin/Clavulanate" = "amoxiclav",
         "Augmentin" = "amoxiclav",
         "Piperacilin/tazobactam" = "piperacilin_tazobactam",
+        "Piperacillin/Tazobactam" = "piperacilin_tazobactam",
         "Clarithromycin" = "clarithromycin",
         "ofloxacin" = "ofloxacin",
         "ceftriaxone" = "ceftriaxone",
-        "cefuroxim" = "cefuroxim",
-        "cefotaxim" = "cefotaxim",
+        "cefuroxime" = "cefuroxime",
+        "cefuroxim" = "cefuroxime",
+        "cefotaxim" = "cefotaxime",
+        "cefotaxime" = "cefotaxime",
         "cefepim" = "cefepime",
         "cefepime" = "cefepime",
         "Vancomycin" = "vancomycin",
@@ -56,11 +61,32 @@ pcr_values_positive <- c("pos","neg", "neg and pos", "susp.", "susp", "poz")
 pcr_value_negative <- "neg"
 
 unit_overwrites <- list(
-  QKuFp = list(
-    list(marker = "leukocyte_count", old_unit = "unit", new_unit = "10^9/l")
-    ),
-  YqNbe = list(
-    list(marker = "d_dimer", old_unit = "ng/ml DDU", new_unit = "μg/l"),
+  all = list(
     list(marker = "CRP", old_unit = "ng/l", new_unit = "mg/l")
-   )
+  ),
+  QKuFp = list(
+    list(marker = "leukocyte_count", old_unit = "unit", new_unit = "10^9/l"),
+    list(marker = "vancomycin", old_unit = "", new_unit = "g/day"),
+    #TODO check the IgG with hospital
[…]

### 2020-08-27 R/data_preprocessing.R

@@ -66,6 +66,7 @@ read_patient_data <- function(file, hospital_id, lang, file_version, remove_exam
       symptom_onset = lubridate::as_date(`Date of symptom onset`),
       admission = lubridate::as_date(`Date of admission`),
       transferred_from = `Transferred from`,
+      transferred_to = `Transferred to`,
       admitted_for_covid = logical_column(`Admitted for Covid`),
       ischemic_heart_disease = logical_column(`Ischemic Heart Disease`),
       has_hypertension_drugs = !(tolower(`Hypertension drugs`) %in% no_variants),
@@ -102,22 +103,29 @@ read_patient_data <- function(file, hospital_id, lang, file_version, remove_exam
     stop("Invalid outcome")
   }
 
+  should_not_be_NA <- c("age", "sex")
+  for(no_NAs in should_not_be_NA) {
+    if(any(is.na(patient_data_typed[[no_NAs]]))) {
+      stop(paste0("NAs in ", no_NAs))
+    }
+  }
+
   patient_data_typed
 }
 
 read_patient_data_raw <- function(file, hospital_id, lang, file_version) {
   if(file_version == "Old") {
     n_middle_text_columns <- 12
-    last_column <- "Z"
+    last_column <- "AA"
   } else {
     n_middle_text_columns <- 13
-    last_column <- "AA"
+    last_column <- "AB"
   }
 
   patient_col_types <- c("text","numeric","text","date","date", #ID - Date of admission
                          rep("text", n_middle_text_columns), #Transferred from - Smoking
                          rep("numeric", 5), # BMI - Albumin
-                         "text", "text", "date", "text") #Discontinued medication - Outcome
+                         "text", "text", "date", "text", "text") #Discontinued medication - Transferred to
 
   patients_range <- paste0("A2:",last_column, "100")
 
@@ -137,17 +145,29 @@ read_patient_data_raw <- function(file, hospital_id, lang, file_version) {
 
 
 read_progression_data <- function(file, hospital_id, lang, file_version, patient_data_typed) {
-  progression_data_raw <- readxl::read_excel(file, sheet = progression_sheet(lang), range = "A1:CU2539", na = c("NA",""), col_types = c(c("text","text", "skip", "text", "text"), rep("text", 94)), col_names = TRUE)
+  max_day <- 172
[…]

### 2020-08-27 dev/data_preprocessing.Rmd
Anonymization for analysis

@@
 ```
+
+TODO: code adverse events, code discontinued_medication_reason
 
@@
 complete_data <- join_hospital_data(hospital_data)
-# TODO transfers
 # TODO check:last day vs. breathing data
 # MAx. day vs. read columns
-# TODO check units (and convert if necessary)
 differing_units <- complete_data$marker_data %>% group_by(marker) %>% filter(length(unique(unit)) > 1) %>% 
@@
 ```
+
+## Final processing steps
+
+```{r}
+no_data_pts <- 
+  complete_data$breathing_data %>% group_by(hospital_id, patient_id) %>%
+  summarise(count = n(), .groups = "drop") %>%
+  filter(count < 1)
+
+if(nrow(no_data_pts) > 0) {
+  print(no_data_pts)
+  stop("Some patients have no data")
+}
+```
+
+```{r}
+anonymized_data <- anonymize_for_analysis(complete_data)
+```
+
+```{r}
+for(n in names(anonymized_data) ) {
+  write_csv(anonymized_data[[n]], path = here::here("private_data", paste0(n,".csv")))
+}
+```

### 2020-08-27 R/translation.R

@@ -43,9 +43,17 @@ translate_outcomes <- function(outcomes, lang) {
     outcomes
   } else {
     translation <- outcome_translations$Outcome
-    names(translation) <- outcome_translations$Cz
+    names(translation) <- tolower(outcome_translations$Cz)
 
-    translation[outcomes]
+    res <- translation[tolower(as.character(outcomes))]
+
+    badly_translated <- is.na(res) != is.na(outcomes)
+    if(any(badly_translated)) {
+      print(outcomes[badly_translated])
+      stop("Badly translated outcome")
+    }
+
+    res
   }
 
 }

### 2020-09-01 manuscript/descriptive.Rmd
More preprocessing, descriptive statistics

---
title: "Descriptive stats and visuals"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
```

```{r}
nrow(data$patient_data)
data$breathing_data %>% group_by(patient_id) %>% summarise(n_days_data = n(), .groups = "drop") %>%
  ggplot(aes(x = n_days_data)) + geom_histogram(binwidth = 5)
```

```{r}
disease_levels <- c("Discharged", breathing_levels, "Death")
outcome_data <- data$patient_data %>% 
  filter(outcome %in% c("Discharged", "Death")) %>%
  transmute(patient_id = patient_id, 
            hospital_id = hospital_id,
            day = last_record,
            breathing = as.character(outcome))

data_for_plot <- data$breathing_data %>% 
  mutate(breathing = as.character(breathing)) %>%
  rbind(outcome_data) %>%
  mutate(breathing = factor(breathing, levels = disease_levels, ordered = TRUE))

data_for_plot %>% ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
  geom_tile() + 
  scale_fill_viridis_d(direction = -1) +
  theme(axis.text.y = element_blank())
```
```{r}

```

### 2020-09-01 R/data_preprocessing.R

@@ -1,500 +1,52 @@
-#Throughout we use global variables from data_mapping.R
-
-read_raw_data <- function(file, hospital_id, lang, file_version, remove_examples = TRUE) {
-  file_version <- update_and_check_version(file, lang, file_version)
-  patient_data <- read_patient_data(file, hospital_id, lang, file_version, remove_examples)
-
-  progression_data <- read_progression_data(file, hospital_id, lang, file_version, patient_data)
-
-  # TODO code adverse events
-  loo::nlist(patient_data,
-             breathing_data = progression_data$breathing_data,
-             marker_data = progression_data$marker_data,
-             adverse_events_data = progression_data$adverse_events_data)
-}
-
-
-update_and_check_version <- function(file, lang, file_version) {
-  if(file_version == "From Sheet") {
-    file_version <- readxl::read_excel(file, sheet = version_sheet(lang), range = "C8:C8", col_types = "text", col_names = "version")$version
-  }
-
-  if(!grepl("^(Old|1\\.[2-4]\\.[0-9])$", file_version)) {
-    print(file_version)
-    stop("Invalid version")
-  }
-
-  file_version
-}
-
-read_patient_data <- function(file, hospital_id, lang, file_version, remove_examples) {
-  patient_data_raw <- read_patient_data_raw(file, hospital_id, lang, file_version)
-
-  patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
-
-  if(remove_examples) {
-    patient_data_raw_usable <- patient_data_raw_usable %>%
-      filter(!startsWith(`Patient ID`, "Example"), !grepl("^P.*klad", `Patient ID`))
-  }
-
-
-  logical_column <- function(x) {
-    if(any( !is.na(x) && !(tolower(x) %in% c(yes_variants, no_variants))) ) {
-      stop("Invalid logical column contents")
-    }
-    tolower(x) %in% yes_variants
-  }
-
-  outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")
-
-  # Special handling of badly filled outcomes
-  if(hospital_id == "QKuFp") {
-    patient_data_raw_usable <- patient_data_raw_usable %>%
-      mutate(`Outcome (Discharged/Hospitalized/Transferred/Death)` =
-               if_else(grepl("^439[0-9]{2}$", `Outcome (Discharged/Hospitalized/Transferred/Death)`),
-                       "Hospitalized",
[…]

### 2020-09-01 R/raw_data_preprocessing.R

@@ -1,0 +1,504 @@
+#Throughout we use global variables from data_mapping.R
+
+read_raw_data <- function(file, hospital_id, lang, file_version, remove_examples = TRUE) {
+  file_version <- update_and_check_version(file, lang, file_version)
+  patient_data <- read_patient_data(file, hospital_id, lang, file_version, remove_examples)
+
+  progression_data <- read_progression_data(file, hospital_id, lang, file_version, patient_data)
+
+  # TODO code adverse events
+  loo::nlist(patient_data,
+             breathing_data = progression_data$breathing_data,
+             marker_data = progression_data$marker_data,
+             adverse_events_data = progression_data$adverse_events_data)
+}
+
+
+update_and_check_version <- function(file, lang, file_version) {
+  if(file_version == "From Sheet") {
+    file_version <- readxl::read_excel(file, sheet = version_sheet(lang), range = "C8:C8", col_types = "text", col_names = "version")$version
+  }
+  if(file_version == "From Sheet C7") {
+    file_version <- readxl::read_excel(file, sheet = version_sheet(lang), range = "C7:C7", col_types = "text", col_names = "version")$version
+  }
+
+  if(!grepl("^(Old|1\\.[2-4]\\.[0-9])$", file_version)) {
+    print(file_version)
+    stop("Invalid version")
+  }
+
+  file_version
+}
+
+read_patient_data <- function(file, hospital_id, lang, file_version, remove_examples) {
+  patient_data_raw <- read_patient_data_raw(file, hospital_id, lang, file_version)
+
+  patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
+
+  if(remove_examples) {
+    patient_data_raw_usable <- patient_data_raw_usable %>%
+      filter(!startsWith(`Patient ID`, "Example"), !grepl("^P.*klad", `Patient ID`))
+  }
+
+
+  logical_column <- function(x) {
+    if(any( !is.na(x) && !(tolower(x) %in% c(yes_variants, no_variants))) ) {
+      stop("Invalid logical column contents")
+    }
+    tolower(x) %in% yes_variants
+  }
+
+  outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")
+
+  # Special handling of badly filled outcomes
+  if(hospital_id == "QKuFp") {
+    patient_data_raw_usable <- patient_data_raw_usable %>%
[…]

### 2020-09-08 dev/hmm_grein.Rmd
First results on real data

@@
 
-#make_stancode_hmm(m1)
+make_stancode_hmm(m1)
 #xx <- brms::make_standata(m1$formula, data = make_data_hmm(m1)$brmsdata)

### 2020-09-08 dev/raw_data_preprocessing.Rmd
More preprocessing, descriptive statistics
First results on real data

--- 2020-08-27 dev/data_preprocessing.Rmd
@@
 TODO: code adverse events, code discontinued_medication_reason
+TODO: check last day, add outcome to breathing data, possibly
 
@@
 
-Check values for all markers with converted units
+Check values for all markers with converted units (Note the 0 for Amoxiclav has been checked and is in the original data as well)
 
@@
 }
+```
+```{r}
+complete_data$marker_data %>% filter(marker == "amoxiclav", value == 0)
 ```
@@
 ```{r}
-i <- 2; j <- 3
 for(i in 1:(length(hospital_data) - 1)) {

### 2020-09-08 manuscript/competing_risks.Rmd
First results on real data

---
title: "Survival - frequentist"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
library(survival)

theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- prepare_marker_data_wide(data)

hypothesis_res_list <- list()

cache_dir <- here::here("local_temp_data")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

# Competing risks death vs. discharged

```{r}
data_surv <- wide %>%  
  filter(day > 0) %>%
  mutate(daym1 = day - 1,
    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>% factor(ordered = FALSE) %>% fct_relevel("Hospitalized", "Discharged", "Death") %>% fct_explicit_na("Hospitalized")
  ) %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))

f_competing_hcq <- Surv(daym1, day,status) ~ age + sex + took_hcq

#survcheck(f_competing_hcq, data = data_surv, id = patient_id)
fit_competing_hcq <- coxph(f_competing_hcq, data = data_surv, id = patient_id)
fit_competing_hcq

hypothesis_res_list[["competing_hcq"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex")
  )

hypothesis_res_list[["competing_hcq"]]
```

```{r}
f_competing_hcq_az <- update.formula(f_competing_hcq, . ~ . + took_az)
fit_competing_hcq_az <- coxph(f_competing_hcq_az, data = data_surv, id = patient_id)

fit_competing_hcq_az

hypothesis_res_list[["competing_hcq_az"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az")
  )
hypothesis_res_list[["competing_hcq_az"]]
```

```{r}
f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + hospital_id)
fit_competing_hcq_az_hospital <- coxph(f_competing_hcq_az_hospital, data = data_surv, id = patient_id)
fit_competing_hcq_az_hospital

hypothesis_res_list[["competing_hcq_az_hospital"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_hospital,
                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az")
  )
hypothesis_res_list[["competing_hcq_az_hospital"]]

```
```{r}
hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
```

# PCR positivity predicts

# Lymphocyte count predicts?

### 2020-09-08 manuscript/descriptive.Rmd
First results on real data

@@
 ```{r}
-disease_levels <- c("Discharged", breathing_levels, "Death")
-outcome_data <- data$patient_data %>% 
-  filter(outcome %in% c("Discharged", "Death")) %>%
-  transmute(patient_id = patient_id, 
-            hospital_id = hospital_id,
-            day = last_record,
-            breathing = as.character(outcome))
+# disease_levels <- c("Discharged", breathing_levels, "Death")
+# outcome_data <- data$patient_data %>% 
+#   filter(outcome %in% c("Discharged", "Death")) %>%
+#   transmute(patient_id = patient_id, 
+#             hospital_id = hospital_id,
+#             day = last_record,
+#             breathing = as.character(outcome))
+#                                                 
+# data_for_plot <- data$breathing_data %>% 
+#   mutate(breathing = as.character(breathing)) %>%
+#   rbind(outcome_data) %>%
+#   mutate(breathing = factor(breathing, levels = disease_levels, ordered = TRUE))
+# 
+# if(any(is.na(data_for_plot$day)) || any(is.na(data_for_plot$breathing))) {
+#   stop("Problem")
+# }
 
-data_for_plot <- data$breathing_data %>% 
-  mutate(breathing = as.character(breathing)) %>%
-  rbind(outcome_data) %>%
-  mutate(breathing = factor(breathing, levels = disease_levels, ordered = TRUE))
-
-data_for_plot %>% ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
+data$breathing_data %>% ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
   geom_tile() + 
   scale_fill_viridis_d(direction = -1) +
-  theme(axis.text.y = element_blank())
+  theme(axis.text.y = element_blank()) +
+  facet_wrap(~hospital_id, scales = "free")
+
 ```
 ```{r}
+data$breathing_data %>% group_by(hospital_id, patient_id) %>%
+  summarise(first = min(day, na.rm = TRUE)) %>%
+  filter(first > 0)
+```
+
+```{r}
+wide <- prepare_marker_data_wide(data)
+wide  %>% ggplot(aes(x = day, y = patient_id, fill = took_convalescent_plasma)) + 
+  geom_tile() + 
+  scale_fill_viridis_d(direction = -1) +
+  theme(axis.text.y = element_blank()) +
+  facet_wrap(~hospital_id, scales = "free")
+
+wide  %>% ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
+  geom_tile() + 
+  scale_fill_viridis_d(direction = -1) +
+  theme(axis.text.y = element_blank()) +
+  facet_wrap(~hospital_id, scales = "free")
 
 ```
+
+```{r}
+outcome_data %>% filter(day > 30)
+data$patient_data %>% filter(patient_id == "kOw52xii4l")
+data_for_plot %>% filter(is.na(day))
+```
+```{r}
+data$patient_data %>% filter(is.na(last_record))
+```
+
+```{r}
+data$patient_data %>% group_by(hospital_id) %>% summarise(count = n())
+```
+Where do we have measurements for the individual markers (tiles show where we have breathing data, study sites are coded via fill color, crosses show available data).
+```{r}
+all_markers <- data$marker_data %>% group_by(marker) %>% summarise(n_obs = n(), .groups = "drop") %>%
+  arrange(desc(n_obs)) %>% filter(n_obs > 10) %>% pull(marker)
+
+breathing_data_ordered <- data$breathing_data %>%  mutate(patient_id = factor(patient_id) %>% fct_reorder(hospital_id, .fun = unique))
+
+n_markers <- length(all_markers)
+step_size <- 4
+for(i in 1:ceiling(n_markers / step_size)) {
+  markers_to_show <- all_markers[((i - 1) * step_size + 1):min(n_markers, i * step_size)]
+  print(
+    data$marker_data %>%
+      filter(marker %in% markers_to_show)  %>%
+      ggplot(aes(x = day, y = patient_id)) +
+        geom_tile(data = breathing_data_ordered, aes(fill = hospital_id)) +
+        geom_point(shape = "cross") +
+        theme(axis.text.y = element_blank()) +
+        scale_fill_brewer(type = "qual", palette = 2, guide = FALSE) +
+        facet_wrap(~marker, nrow = 1, scales = "free_x")
+
+  )
+}
+```

### 2020-09-08 manuscript/hmm.Rmd
First results on real data

---
title: "HMM Model"
output: html_document
---

```{r}
devtools::load_all()
library(tidyverse)
options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)

theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- prepare_marker_data_wide(data)

cache_dir <- here::here("local_temp_data")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

#Temporary hack
registerS3method("stan_llh", class = "rate_hmm", method = stan_llh.rate_hmm, envir = asNamespace("brms"))
```

```{r}
series_data <- wide %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
  rename(.time = day, .serie = patient_id, .observed = breathing) %>%
  mutate(.time = .time +1) %>% 
  filter(.time >= 1) %>%
  group_by(.serie) %>%
  filter(sum(!is.na(.observed)) > 1) %>%
  ungroup()
```

```{r}
data$patient_data %>% filter(!is.na(best_supportive_care_from)) %>% select(outcome)
```

```{r}
m1_hidden_state_data <- tibble(id = factor(disease_levels, levels = disease_levels)) %>% mutate( corresponding_obs = id)

m1_observed_state_data <- tibble(id = factor(disease_levels, levels = disease_levels), is_noisy = FALSE)

recode_group <- function(rate_data, group_name) {
  is_column <- paste0("is_", group_name) 
  id_column <- paste0(group_name, "_id")

  rate_data %>% mutate(
    !!is_column := as.double(rate_group == group_name),
    !!id_column := factor(
      if_else(rate_group == group_name, as.integer(.rate_id) - min(as.integer(.rate_id[rate_group == group_name])), as.integer(0)), 
      levels = 0:(sum(rate_group == group_name) - 1), 
      ordered = TRUE)
  )
}

m1_rate_data <- rbind(
  tibble(.from = breathing_levels, .to = "Death", rate_group = "death"),
  tibble(.from = "AA", .to = "Discharged", rate_group = "improve_one"),
  tibble(.from = breathing_levels[2 : length(breathing_levels)], .to = breathing_levels[1 : (length(breathing_levels) - 1)], rate_group = "improve_one"),
  tibble(.from = breathing_levels[1 : (length(breathing_levels) - 1)], .to = breathing_levels[2 : length(breathing_levels)], rate_group = "worsen_one")
) %>%
  mutate(.rate_id = factor(1:n()),
         .from = factor(.from, levels = disease_levels),
         .to = factor(.to, levels = disease_levels)
  ) %>%
  recode_group("death") %>%
  recode_group("improve_one") %>%
  recode_group("worsen_one")

m1_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)

m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
  brms::set_prior("normal(0, 5)", "b") +
  brms::set_prior("normal(0, 2)", "sd")

m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group
m1 <- brmshmmdata(m1_formula, series_data, 
                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
                  observed_state_data = m1_observed_state_data)

#make_stancode_hmm(m1)
#xx <- brms::make_standata(m1$formula, data = make_data_hmm(m1)$brmsdata)
m1_fit <- brmhmm(m1, cache_file = paste0(cache_dir,"/hcq_m1.rds"))

summary(m1_fit$brmsfit)

```
```{r}
m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
m2 <- brmshmmdata(m2_formula, series_data, 
                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
                  observed_state_data = m1_observed_state_data)

m2_fit <- brmhmm(m2, cache_file = paste0(cache_dir,"/hcq_m2.rds"))

summary(m2_fit$brmsfit)
```

```{r}
m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
m3 <- brmshmmdata(m3_formula, series_data, 
                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
                  observed_state_data = m1_observed_state_data)

m3_fit <- brmhmm(m3, cache_file = paste0(cache_dir,"/hcq_m3.rds"))

summary(m3_fit$brmsfit)
```
```{r}
m3_epred_rect <- posterior_epred_rect(m3_fit)
m3_predicted_rect <- posterior_epred_to_predicted(m3_fit, m3_epred_rect)
m3_predicted <- posterior_rect_to_long(m3_fit, m3_predicted_rect)
```

```{r}
m3_predicted_df <- posterior_long_to_df(m3_fit$data, m3_predicted)

all_ids <- m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
  distinct() %>% arrange(hospital_id) %>% pull(.serie)
step_size <- 6
for(step in 1:ceiling(length(all_ids) / step_size)) {
  series_id <- ((step - 1) * step_size + 1) : (step* step_size)

  posterior_state_plot(m3_predicted_df, all_ids[series_id], m3_fit$data) %>% print()

}
```

```{r}
pp_check_last_state(m3_fit, m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
pp_check_transitions_direction(m3_fit, m3_predicted_rect)
pp_check_transitions_direction(m3_fit, m3_predicted_rect, states_from = 2:6)
```

### 2020-09-08 R/brmshmmfit.R

@@ -5,8 +5,7 @@ brmhmm <- function(brmshmmdata, cache_file = NULL, ...) {
 
   bfit <- brm_with_cache(
     cache_file = cache_file,
-    formula = d$formula,
-    family = rate_hmm_family,
+    formula = brms::brmsformula(d$formula, family = rate_hmm_family),
     data = prepdata$brmsdata,
     prior = d$prior,
     stanvars = rate_hmm_stanvars(prepdata$standata),

### 2020-09-08 R/data_mapping.R

@@ -93,6 +93,7 @@ unit_map <- c(
   )
 
 breathing_levels <- c("AA","Oxygen", "NIPPV","MV","ECMO")
+disease_levels <- c("Discharged", breathing_levels, "Death")
 
 unit_conversions <- list(
   list(markers = c("cefotaxime","cefuroxime", "ampicillin_sulbactam", "amoxiclav"),

### 2020-09-08 R/data_preprocessing.R

@@ -27,14 +27,15 @@ read_data_for_analysis <- function() {
                              best_supportive_care_from = col_integer(),
                              last_record = col_integer(),
                              outcome = col_factor()
-                           ))
+                           )) %>%
+    mutate(age_norm = (age - 65) / 20)
 
   breathing_data <- read_csv(here::here("private_data", "breathing_data.csv"),
                            col_types = cols(
                             patient_id = col_character(),
                             hospital_id = col_character(),
                             day = col_integer(),
-                            breathing = col_factor(levels = breathing_levels, ordered = TRUE)
+                            breathing = col_factor(levels = disease_levels, ordered = TRUE)
                           ))
 
   marker_data <- read_csv(here::here("private_data", "marker_data.csv"),
@@ -49,4 +50,98 @@ read_data_for_analysis <- function() {
                             ))
 
   loo::nlist(patient_data, breathing_data, marker_data)
+}
+
+prepare_marker_data_wide <- function(data) {
+  patient_max_days <- data$breathing_data %>%
+    group_by(hospital_id, patient_id) %>%
+    summarise(max_day = max(day), .groups = "drop")
+
+  complete_max_day <- max(patient_max_days$max_day)
+
+  # Add _censored columns only for markers where censoring was observed
+  wider_spec_censored <-
+    tibble(
+      marker = data$marker_data %>% filter(censored != "none") %>% pull(marker) %>% unique()
+    ) %>%
+    mutate(.name = paste0(marker, "_censored"), .value = "censored")
+
+  wider_spec <-
+    build_wider_spec(data$marker_data, names_from = "marker", values_from = "value") %>%
+    rbind(wider_spec_censored)
+  marker_wide <- data$marker_data %>%
+    select(-unit) %>% pivot_wider_spec(wider_spec, id_cols = one_of("patient_id", "hospital_id", "day"), values_fn = function(x) { if(length(x) > 1) { "ERROR" } else { x } })
+
+  which((marker_wide %>% select(oxygen_flow:az_censored) %>% as.matrix()) > 1, arr.ind = TRUE)
+
+  all_values <- marker_wide %>% select(oxygen_flow:az_censored) %>% unlist()
+  if(any(grepl("ERROR", all_values))) {
[…]

### 2020-09-08 R/hypotheses.R

@@ -1,0 +1,23 @@
+hypotheses <-
+  list(hcq_reduces_death = list(caption = "HCQ reduces risk of death"),
+       hcq_increases_discharged = list(caption = "HCQ reduces time in hospital")
+       ) %>% purrr::imap(
+         function(def, name) {
+           def$name <- name
+           def
+         })
+
+
+frequentist_hypothesis_res_from_coxph <- function(
+  hypothesis, coxph_fit, coefficient_name, transition, adjusted) {
+
+  summ <- summary(coxph_fit)
+  coefficient_id <- summ$cmap[coefficient_name, transition]
+  data.frame(hypothesis = hypothesis$name,
+             model = "coxph",
+             adjusted = "age, sex",
+             p_value = summ$coefficients[coefficient_id, "Pr(>|z|)"],
+             ci_low = summ$conf.int[coefficient_id, "lower .95"],
+             ci_high = summ$conf.int[coefficient_id, "upper .95"]
+  )
+}

### 2020-09-08 R/make_stancode_hmm.R

@@ -15,11 +15,15 @@ make_stancode_hmm <- function(brmshmmdata) {
                       stanvars = rate_hmm_stanvars(data$standata), prior = d$prior)
 }
 
-rate_hmm_family <- structure(list(family = "rate_hmm", link = "identity", dpars = "mu",
+rate_hmm_family <- function() {
+  structure(list(family = "rate_hmm",
+                                  link = "identity", dpars = "mu",
                            lb = NA, ub = NA, type = "real", vars = NULL, specials = NULL,
                            ybounds = c(-Inf, Inf),
                            log_lik = NULL, posterior_predict = NULL, posterior_epred = NULL),
+                      #class = c("rate_hmm", "brmsfamily", "family"))
                       class = c("rate_hmm", "brmsfamily"))
+}
 
 summary.rate_hmm <- function(f, link = FALSE) {
   "Rate-based HMM"
@@ -28,6 +32,10 @@ summary.rate_hmm <- function(f, link = FALSE) {
 .family_rate_hmm <- function() {
 
 }
+
+# family_info.rate_hmm <- function(x, y, ...) {
+#   NULL
+# }
 
 posterior_epred_rate_hmm <- function(prep) {
   brms:::posterior_epred_gaussian(prep)
@@ -43,7 +51,6 @@ rate_hmm_stanvars <- function(standata) {
   brms::stanvar(scode = rate_hmm_genquant_code, block = "genquant")
 
 }
-
 
 rate_hmm_functions_code <- "
   // Compute a single transition matrix
@@ -138,11 +145,15 @@ rate_hmm_tdata_code <- '
   }
 '
 
+##' @export rate_hmm_parameters_code
 rate_hmm_parameters_code <- "
   vector<lower=sensitivity_low_bound, upper=1>[N_noisy_states] sensitivity;
   simplex[N_other_observations] other_observations_probs[N_noisy_states];
 "
 
+##' @export stan_llh.rate_hmm
+##' @method stan_llh rate_hmm
+##' @importFrom brms stan_llh
 stan_llh.rate_hmm <- function( ... ) {
 "
   matrix[N_states_hidden, N_states_observed] observation_probs = compute_observation_matrix(
@@ -190,6 +201,7 @@ stan_llh.rate_hmm <- function( ... ) {
 
 "
 }
+
 
 rate_hmm_genquant_code <- '
     matrix[N_states_hidden, N_states_observed] observation_probs = compute_observation_matrix(

### 2020-09-11 manuscript/competing_risks.Rmd
Additional models

@@
 data <- read_data_for_analysis()
-wide <- prepare_marker_data_wide(data)
+wide <- data$marker_data_wide
 
@@
   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_hospital,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
+                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, hospital"),
   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_hospital,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az")
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital")
   )
@@
 
+```{r}
+f_competing_hcq_az_comorbidities_sum <- update.formula(f_competing_hcq_az, . ~ . + comorbidities_sum)
+fit_competing_hcq_az_comorbidities_sum <- coxph(f_competing_hcq_az_comorbidities_sum, data = data_surv, id = patient_id)
+fit_competing_hcq_az_comorbidities_sum
+
+hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]] <-
+  rbind(
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_comorbidities_sum,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_comorbidities_sum,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)")
+  )
+hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
+```
+
+```{r}
+hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
+write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+```
+
 # PCR positivity predicts

### 2020-09-11 manuscript/hmm.Rmd
Additional models

@@
 ```{r}
+m1_epred_rect <- posterior_epred_rect(m1_fit)
+m1_predicted_rect <- posterior_epred_to_predicted(m1_fit, m1_epred_rect)
+m1_predicted <- posterior_rect_to_long(m1_fit, m1_predicted_rect)
+```
+
+```{r}
+m1_predicted_df <- posterior_long_to_df(m1_fit$data, m1_predicted)
+
+all_ids <- m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m1_predicted_df, all_ids[series_id], m1_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m1_fit, m1_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(m1_fit, m1_predicted_rect)
+pp_check_transitions_direction(m1_fit, m1_predicted_rect, states_from = 2:6)
+```
+
+```{r}
 m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
@@
 summary(m2_fit$brmsfit)
+```
+
+```{r}
+m2_epred_rect <- posterior_epred_rect(m2_fit)
+m2_predicted_rect <- posterior_epred_to_predicted(m2_fit, m2_epred_rect)
+m2_predicted <- posterior_rect_to_long(m2_fit, m2_predicted_rect)
+```
+
+```{r}
+m2_predicted_df <- posterior_long_to_df(m2_fit$data, m2_predicted)
+
+all_ids <- m2_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m2_predicted_df, all_ids[series_id], m2_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m2_fit, m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(m2_fit, m2_predicted_rect)
+pp_check_transitions_direction(m2_fit, m2_predicted_rect, states_from = 2:6)
 ```
@@
 ```
+
+```{r}
+m4_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
+m4 <- brmshmmdata(m4_formula, series_data, 
+                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
+                  observed_state_data = m1_observed_state_data)
+
+m4_fit <- brmhmm(m4, cache_file = paste0(cache_dir,"/hcq_m4.rds"))
+
+summary(m4_fit$brmsfit)
+```
+
+```{r}
+m4_epred_rect <- posterior_epred_rect(m4_fit)
+m4_predicted_rect <- posterior_epred_to_predicted(m4_fit, m4_epred_rect)
+m4_predicted <- posterior_rect_to_long(m4_fit, m4_predicted_rect)
+```
+
+```{r}
+m4_predicted_df <- posterior_long_to_df(m4_fit$data, m4_predicted)
+
+all_ids <- m4_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m4_predicted_df, all_ids[series_id], m4_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m4_fit, m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(m4_fit, m4_predicted_rect)
+pp_check_transitions_direction(m4_fit, m4_predicted_rect, states_from = 2:6)
+```
+
+## M5 -
+
+```{r}
+m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + mo(comorbidities_sum) * rate_group
+m5 <- brmshmmdata(m5_formula, series_data, 
+                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
+                  observed_state_data = m1_observed_state_data)
+
+m5_fit <- brmhmm(m5, cache_file = paste0(cache_dir,"/hcq_m5.rds"))
+
+summary(m5_fit$brmsfit)
+```
+
+```{r}
+m5_epred_rect <- posterior_epred_rect(m5_fit)
+m5_predicted_rect <- posterior_epred_to_predicted(m5_fit, m5_epred_rect)
+m5_predicted <- posterior_rect_to_long(m5_fit, m5_predicted_rect)
+```
+
+```{r}
+m5_predicted_df <- posterior_long_to_df(m5_fit$data, m5_predicted)
+
+all_ids <- m5_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(m5_predicted_df, all_ids[series_id], m5_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(m5_fit, m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(m5_fit, m5_predicted_rect)
+pp_check_transitions_direction(m5_fit, m5_predicted_rect, states_from = 2:6)
+```

### 2020-09-11 manuscript/jm.Rmd
Additional models

---
title: "Joint modelling"
output: html_notebook
---

```{r, setup}
devtools::load_all()
library(tidyverse)
library(survival)
library(rstanarm)

theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide

cache_dir <- here::here("local_temp_data")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

```{r}

data_surv <- wide %>%  
  filter(day > 0) %>%
  mutate(daym1 = day - 1,
    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>% factor(ordered = FALSE) %>% fct_relevel("Hospitalized", "Discharged", "Death") %>% fct_explicit_na("Hospitalized")
  ) %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))

tt <- stan_jm(formulaLong = CRP ~ day + (1 | patient_id),
              dataLong = wide,
              formulaEvent = Surv(daym1, day,status) ~ age + sex + took_hcq,
              dataEvent = data_surv,
              id_var = "patient_id",
              time_var = "day")
```

### 2020-09-11 R/data_preprocessing.R

@@ -49,7 +49,50 @@ read_data_for_analysis <- function() {
                               censored = col_character()
                             ))
 
-  loo::nlist(patient_data, breathing_data, marker_data)
+  data <- loo::nlist(patient_data, breathing_data, marker_data)
+  data$marker_data_wide <- prepare_marker_data_wide(data)
+
+  data <- compute_derived_quantities_patients(data)
+
+  data
+}
+
+#' @importFrom tidy replace_na
+compute_derived_quantities_patients <- function(data) {
+  data$patient_data <- data$patient_data %>%
+    mutate(
+      heart_problems = NYHA > 1,
+      obesity = BMI > 30,
+      comorbidities_sum =
+        replace_na(ischemic_heart_disease, FALSE) +
+        replace_na(has_hypertension_drugs, FALSE) +
+        replace_na(heart_failure, FALSE) +
+        replace_na(COPD, FALSE) +
+        replace_na(asthma, FALSE) +
+        replace_na(other_lung_disease, FALSE) +
+        replace_na(diabetes, FALSE) +
+        replace_na(renal_disease, FALSE) +
+        replace_na(liver_disease, FALSE) +
+        replace_na(smoking, FALSE) +
+        replace_na(heart_problems, FALSE) +
+        replace_na(obesity, FALSE),
+      comorbidities_sum_na = 2 * (
+        replace_na(ischemic_heart_disease, 0.5) +
+        replace_na(has_hypertension_drugs, 0.5) +
+        replace_na(heart_failure, 0.5) +
+        replace_na(COPD, 0.5) +
+        replace_na(asthma, 0.5) +
+        replace_na(other_lung_disease, 0.5) +
+        replace_na(diabetes, 0.5) +
+        replace_na(renal_disease, 0.5) +
+        replace_na(liver_disease, 0.5) +
+        replace_na(smoking, 0.5) +
+        replace_na(heart_problems, 0.5) +
+        replace_na(obesity, 0.5))
+    )
+
+
+  data
 }
 
 prepare_marker_data_wide <- function(data) {

### 2020-09-12 manuscript/manuscript.Rmd
Started writing manuscript

---
title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: A dataset and exploratory analysis" 
output: pdf_document
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom."
bibliography: Covid19.json
---

```
{r setup}
devtools::load_all()
library(tidyverse)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
```

# Introduction

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths. Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR; TODO]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; http://zotero.org/users/5567156/items/5VAKEW58]. While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@https://www.acpjournals.org/doi/10.7326/M20-4207] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; http://zotero.org/users/5567156/items/GMR9FX9Z]. Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; http://zotero.org/users/5567156/items/BHGT88US] For additional summary see (live review)

In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e. g., @http://zotero.org/users/5567156/items/WZPNBRYN; http://zotero.org/users/5567156/items/4HR7MUJ6; http://zotero.org/users/5567156/items/NXC9CDI9]

High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/ID7F3XIH].
Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed TODO.

An ongoing challenge in evalutating Covid-19 treatments is that the analysis and interpretation of the data is often inappropriate or misleading, most notably interpreting lack of evidence due to small sample size as evidence of no effect [see e.g., @http://zotero.org/users/5567156/items/5PZEEPJV; @http://zotero.org/users/5567156/items/AZDZHLJV].

# Methods

A convenience sample of patients... The study was approved by the Ethical committees of TODO.

Sex differences are to be expected for both sociological and biological [@http://zotero.org/users/5567156/items/5KQZ3XEE] reasons.

Important to consider appropriate estimands [@http://zotero.org/users/5567156/items/MH797GNC].

We performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V]

# Results

```
{r}
n_patients <- nrow(data$patient_data)
n_hospitals <- length(unique(data$patient_data$hospital_id))
```

# References

### 2020-09-12 manuscript/manuscript.Rmd
More work on manuscript

@@
 
-A convenience sample of patients... The study was approved by the Ethical committees of TODO.
+## Data Collection
+A convenience sample of patients...
+Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.
+Full protocol for data collection is attached in the supplementary material.
+The study was approved by the Ethical committees of TODO.
 
-Sex differences are to be expected for both sociological and biological [@http://zotero.org/users/5567156/items/5KQZ3XEE] reasons.
+## Statistical analysis
 
-Important to consider appropriate estimands [@http://zotero.org/users/5567156/items/MH797GNC].
+Since many details of analysis may influence the conclusions made, we performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V] and report results for all the hypothesis tested across multiple different models using both frequentist and Bayesian paradigms.
+For each model class we worked with several possible sets of adjustments and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC].
+All analysis wes performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualisations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
+Complete code for all analyses can be found at TODO.
 
-We performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V]
+First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
+
+`brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
+The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths.
+
+`stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5]
+
+```{r}
+#Sex differences are to be expected for both sociological and biological [@http://zotero.org/users/5567156/items/5KQZ3XEE] reasons.
+```
 

### 2020-09-12 dev/raw_data_preprocessing.Rmd
Data revision, renamed vars in hmm.Rmd

@@
 }
+write_csv(data.frame(version = date()), path = here::here("data_revision.csv"))
 ```

### 2020-09-12 manuscript/hmm.Rmd
Data revision, renamed vars in hmm.Rmd

--- 2020-09-08 manuscript/hmm.Rmd
@@
 ```{r}
-m1_hidden_state_data <- tibble(id = factor(disease_levels, levels = disease_levels)) %>% mutate( corresponding_obs = id)
+hcq_m1_hidden_state_data <- tibble(id = factor(disease_levels, levels = disease_levels)) %>% mutate( corresponding_obs = id)
 
-m1_observed_state_data <- tibble(id = factor(disease_levels, levels = disease_levels), is_noisy = FALSE)
+hcq_m1_observed_state_data <- tibble(id = factor(disease_levels, levels = disease_levels), is_noisy = FALSE)
 
@@
 
-m1_rate_data <- rbind(
+hcq_m1_rate_data <- rbind(
   tibble(.from = breathing_levels, .to = "Death", rate_group = "death"),
@@
 
-m1_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+hcq_m1_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
+hcq_m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
   brms::set_prior("normal(0, 5)", "b") +
@@
 
-m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+hcq_m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group
-m1 <- brmshmmdata(m1_formula, series_data, 
-                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
-                  observed_state_data = m1_observed_state_data)
+hcq_m1 <- brmshmmdata(hcq_m1_formula, series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
 
-#make_stancode_hmm(m1)
-#xx <- brms::make_standata(m1$formula, data = make_data_hmm(m1)$brmsdata)
-m1_fit <- brmhmm(m1, cache_file = paste0(cache_dir,"/hcq_m1.rds"))
+#make_stancode_hmm(hcq_m1)
+#xx <- brms::make_standata(hcq_m1$formula, data = make_data_hmm(hcq_m1)$brmsdata)
+hcq_m1_fit <- brmhmm(hcq_m1, cache_file = paste0(cache_dir,"/hcq_m1.rds"))
 
-summary(m1_fit$brmsfit)
+summary(hcq_m1_fit$brmsfit)
 
@@
 ```{r}
-m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
-m2 <- brmshmmdata(m2_formula, series_data, 
-                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
-                  observed_state_data = m1_observed_state_data)
-
-m2_fit <- brmhmm(m2, cache_file = paste0(cache_dir,"/hcq_m2.rds"))
-
-summary(m2_fit$brmsfit)
+hcq_m1_epred_rect <- posterior_epred_rect(hcq_m1_fit)
+hcq_m1_predicted_rect <- posterior_epred_to_predicted(hcq_m1_fit, hcq_m1_epred_rect)
+hcq_m1_predicted <- posterior_rect_to_long(hcq_m1_fit, hcq_m1_predicted_rect)
 ```
@@
 ```{r}
-m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-m3 <- brmshmmdata(m3_formula, series_data, 
-                  m1_rate_data, m1_hidden_state_data, m1_initial_states, prior = m1_prior,
-                  observed_state_data = m1_observed_state_data)
+hcq_m1_predicted_df <- posterior_long_to_df(hcq_m1_fit$data, hcq_m1_predicted)
 
-m3_fit <- brmhmm(m3, cache_file = paste0(cache_dir,"/hcq_m3.rds"))
-
-summary(m3_fit$brmsfit)
-```
-```{r}
-m3_epred_rect <- posterior_epred_rect(m3_fit)
-m3_predicted_rect <- posterior_epred_to_predicted(m3_fit, m3_epred_rect)
-m3_predicted <- posterior_rect_to_long(m3_fit, m3_predicted_rect)
-```
-
-```{r}
-m3_predicted_df <- posterior_long_to_df(m3_fit$data, m3_predicted)
-
-all_ids <- m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+all_ids <- hcq_m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
   distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(m3_predicted_df, all_ids[series_id], m3_fit$data) %>% print()
+  posterior_state_plot(hcq_m1_predicted_df, all_ids[series_id], hcq_m1_fit$data) %>% print()
 
@@
 ```{r}
-pp_check_last_state(m3_fit, m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(m3_fit, m3_predicted_rect)
-pp_check_transitions_direction(m3_fit, m3_predicted_rect, states_from = 2:6)
+pp_check_last_state(hcq_m1_fit, hcq_m1_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect)
+pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect, states_from = 2:6)
 ```
+
+```{r}
+hcq_m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
+hcq_m2 <- brmshmmdata(hcq_m2_formula, series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m2_fit <- brmhmm(hcq_m2, cache_file = paste0(cache_dir,"/hcq_m2.rds"))
+
+summary(hcq_m2_fit$brmsfit)
+```
+
+```{r}
+hcq_m2_epred_rect <- posterior_epred_rect(hcq_m2_fit)
+hcq_m2_predicted_rect <- posterior_epred_to_predicted(hcq_m2_fit, hcq_m2_epred_rect)
+hcq_m2_predicted <- posterior_rect_to_long(hcq_m2_fit, hcq_m2_predicted_rect)
+```
+
+```{r}
+hcq_m2_predicted_df <- posterior_long_to_df(hcq_m2_fit$data, hcq_m2_predicted)
+
+all_ids <- hcq_m2_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m2_predicted_df, all_ids[series_id], hcq_m2_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m2_fit, hcq_m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect)
+pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect, states_from = 2:6)
+```
+
+```{r}
+hcq_m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
+hcq_m3 <- brmshmmdata(hcq_m3_formula, series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m3_fit <- brmhmm(hcq_m3, cache_file = paste0(cache_dir,"/hcq_m3.rds"))
+
+summary(hcq_m3_fit$brmsfit)
+```
+```{r}
+hcq_m3_epred_rect <- posterior_epred_rect(hcq_m3_fit)
+hcq_m3_predicted_rect <- posterior_epred_to_predicted(hcq_m3_fit, hcq_m3_epred_rect)
+hcq_m3_predicted <- posterior_rect_to_long(hcq_m3_fit, hcq_m3_predicted_rect)
+```
+
+```{r}
+hcq_m3_predicted_df <- posterior_long_to_df(hcq_m3_fit$data, hcq_m3_predicted)
+
+all_ids <- hcq_m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m3_predicted_df, all_ids[series_id], hcq_m3_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m3_fit, hcq_m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect)
+pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect, states_from = 2:6)
+```
+
+```{r}
+hcq_m4_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
+hcq_m4 <- brmshmmdata(hcq_m4_formula, series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m4_fit <- brmhmm(hcq_m4, cache_file = paste0(cache_dir,"/hcq_m4.rds"))
+
+summary(hcq_m4_fit$brmsfit)
+```
+
+```{r}
+hcq_m4_epred_rect <- posterior_epred_rect(hcq_m4_fit)
+hcq_m4_predicted_rect <- posterior_epred_to_predicted(hcq_m4_fit, hcq_m4_epred_rect)
+hcq_m4_predicted <- posterior_rect_to_long(hcq_m4_fit, hcq_m4_predicted_rect)
+```
+
+```{r}
+hcq_m4_predicted_df <- posterior_long_to_df(hcq_m4_fit$data, hcq_m4_predicted)
+
+all_ids <- hcq_m4_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m4_predicted_df, all_ids[series_id], hcq_m4_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m4_fit, hcq_m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect)
+pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect, states_from = 2:6)
+```
+
+## hcq_m5 -
+
+```{r}
+hcq_m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + mo(comorbidities_sum) * rate_group
+hcq_m5 <- brmshmmdata(hcq_m5_formula, series_data, ☻
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m5_fit <- brmhmm(hcq_m5, cache_file = paste0(cache_dir,"/hcq_m5.rds"))
+
+summary(hcq_m5_fit$brmsfit)
+```
+
+```{r}
+hcq_m5_epred_rect <- posterior_epred_rect(hcq_m5_fit)
+hcq_m5_predicted_rect <- posterior_epred_to_predicted(hcq_m5_fit, hcq_m5_epred_rect)
+hcq_m5_predicted <- posterior_rect_to_long(hcq_m5_fit, hcq_m5_predicted_rect)
+```
+
+```{r}
+hcq_m5_predicted_df <- posterior_long_to_df(hcq_m5_fit$data, hcq_m5_predicted)
+
+all_ids <- hcq_m5_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m5_predicted_df, all_ids[series_id], hcq_m5_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m5_fit, hcq_m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect)
+pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect, states_from = 2:6)
+```

### 2020-09-12 dev/raw_data_preprocessing.Rmd
First working stan_jm models

--- 2020-09-08 dev/raw_data_preprocessing.Rmd
@@
 }
+write_lines(digest::digest(anonymized_data), path = here::here("private_data", "data_revision.txt"))
 ```

### 2020-09-12 manuscript/jm.Rmd
First working stan_jm models

@@
 
-```{r, setup}
+```{r setup}
 devtools::load_all()
@@
 wide <- data$marker_data_wide
+
+options(mc.cores = parallel::detectCores())
 
@@
   mutate(daym1 = day - 1,
-    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>% factor(ordered = FALSE) %>% fct_relevel("Hospitalized", "Discharged", "Death") %>% fct_explicit_na("Hospitalized")
+    event_death = !is.na(breathing) & breathing == "Death",
+    event_discharged = !is.na(breathing) & breathing == "Discharged",
   ) %>%
-  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))
+  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))  %>%
+  arrange("patient_id", "day")
 
-tt <- stan_jm(formulaLong = CRP ~ day + (1 | patient_id),
-              dataLong = wide,
-              formulaEvent = Surv(daym1, day,status) ~ age + sex + took_hcq,
-              dataEvent = data_surv,
+#TODO: impute the competing risk as censored long in the future
+```
+
+```{r}
+data_d_dimer <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(d_dimer))) %>%
+  ungroup() %>%
+    select(event_death, event_discharged, 
+         day, daym1, patient_id, hospital_id, 
+         age, sex, took_hcq,
+         d_dimer) %>%
+  mutate(log_d_dimer = log(d_dimer + 1) - 4)
+
+data_d_dimer %>% filter(d_dimer == 0)
+# data_d_dimer %>% group_by(sex, ever_hcq) %>% summarise(pts = length(unique(patient_id)), n_dimer = sum(!is.na(d_dimer)))
+
+fit_d_dimer_age_sex_hcq <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
+              dataLong = data_d_dimer %>% filter(!is.na(d_dimer)),
+              formulaEvent = Surv(daym1, day, event_death) ~ age + sex + took_hcq,
+              dataEvent = data_d_dimer,
               id_var = "patient_id",
-              time_var = "day")
+              time_var = "day",
+              cache_file = paste0(cache_dir, "/jm_d_dimer_age_sex_hcq.rds"))
+
+fit_d_dimer_age_sex_hcq
+
+#length(unique(data_d_dimer$patient_id))
+to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
+  filter(n_d_dimer > 1) %>% pull(patient_id)
+#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])
+
 ```
+```{r}
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[1:11])
+plot(p1, plot_observed = TRUE)                     
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[12:14])
+plot(p1, plot_observed = TRUE)                     
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[15:18])
+plot(p1, plot_observed = TRUE)
+
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[19:20])
+plot(p1, plot_observed = TRUE)                     
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[21:22])
+plot(p1, plot_observed = TRUE)                     
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[22:26])
+plot(p1, plot_observed = TRUE)
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[27:29])
+plot(p1, plot_observed = TRUE)                     
+p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[30:32])
+plot(p1, plot_observed = TRUE)                     
+```
+
+```{r}
+data_IL_6 <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(IL_6))) %>%
+  ungroup() %>% 
+  select(event_death, event_discharged, 
+         day, daym1, patient_id, hospital_id, 
+         age, sex, took_hcq,
+         IL_6) %>%
+  mutate(log_IL_6 = log(IL_6))
+
+fit_IL_6_age_sex_hcq <- stan_jm_with_cache(formulaLong = log_IL_6 ~ day + (1 + day | patient_id),
+              dataLong = data_IL_6 %>% filter(!is.na(IL_6)),
+              formulaEvent = Surv(daym1, day, event_death) ~ age + sex + took_hcq,
+              dataEvent = data_IL_6,
+              id_var = "patient_id",
+              time_var = "day",
+              cache_file = paste0(cache_dir, "/jm_IL_6_age_sex_hcq.rds"))
+
+fit_IL_6_age_sex_hcq
+
+```
+```{r}
+for(id in unique(data_IL_6$patient_id)) {
+  p1 <- posterior_traj(fit_IL_6_age_sex_hcq, ids = id)
+  print(plot(p1, plot_observed = TRUE)                     )
+}
+
+```

### 2020-09-12 manuscript/manuscript.Rmd
First working stan_jm models

@@
 
-```
-{r setup}
+```{r setup}
 devtools::load_all()
@@
 data <- read_data_for_analysis()
+wide <- prepare_marker_data_wide(data)
 ```
@@
 
-```
-{r}
+```{r}
 n_patients <- nrow(data$patient_data)
@@
 
+```{r}
+my_paste <- function(x) {
+  paste0(as.integer(x), collapse = "")
+}
+wide  %>% 
+  arrange(hospital_id, patient_id, day) %>%
+  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing,  .fun = my_paste, .desc = FALSE)) %>%
+  ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
+  geom_tile() + 
+  scale_fill_viridis_d(direction = -1) +
+  theme(axis.text.y = element_blank()) +
+  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank())
+
+```
+
 # References

### 2020-09-12 R/data_preprocessing.R

@@ -59,6 +59,10 @@ read_data_for_analysis <- function() {
 
 #' @importFrom tidy replace_na
 compute_derived_quantities_patients <- function(data) {
+  derived_from_wide <- data$marker_data_wide %>%
+    group_by(patient_id) %>%
+    summarise(ever_hcq = any(!is.na(took_hcq) & took_hcq))
+
   data$patient_data <- data$patient_data %>%
     mutate(
       heart_problems = NYHA > 1,
@@ -90,6 +94,9 @@ compute_derived_quantities_patients <- function(data) {
         replace_na(heart_problems, 0.5) +
         replace_na(obesity, 0.5))
     )
+
+  data$patient_data <- data$patient_data %>%
+    left_join(derived_from_wide, by = "patient_id")
 
 
   data
@@ -179,12 +186,33 @@ impute_marker_constant <- function(markers_wide, column_name, initial_value = NA
     if(markers_wide$patient_id[i] != last_patient_id) {
       last_value = initial_value
     }
-    if(is.na(markers_wide[[ccolumn_name]][i])) {
-      markers_wide[[ccolumn_name]][i] <- last_value
+    if(is.na(markers_wide[[column_name]][i])) {
+      markers_wide[[column_name]][i] <- last_value
     } else {
-      last_value <- markers_wide[[ccolumn_name]][i]
+      last_value <- markers_wide[[column_name]][i]
     }
   }
+
+  markers_wide
+}
+
+compute_marker_peak <- function(markers_wide, column_name, new_column_name, initial_value = NA) {
+  markers_wide <- markers_wide %>% arrange(patient_id, day)
+
+  last_patient_id <- ""
+  current_max <- NA
+  peaks <- numeric(nrow(markers_wide))
+  for(i in 1:nrow(markers_wide)) {
+    if(markers_wide$patient_id[i] != last_patient_id) {
+      current_max = initial_value
+    }
+    if(!is.na(markers_wide[[column_name]][i])) {
+      current_max = max(current_max, markers_wide[[ccolumn_name]][i])
+    }
+    peaks[i] <- current_max
+  }
+
+  markers_wide[[new_column_name]] <- peaks
 
   markers_wide
 }

### 2020-09-12 R/rstanarm_tools.R

@@ -1,0 +1,34 @@
+stan_jm_with_cache <- function(formulaLong, dataLong, formulaEvent, dataEvent, cache_file = NULL, ...) {
+  fit <- NULL
+  if(!is.null(cache_file) && file.exists(cache_file)) {
+    cache_contents <- readRDS(cache_file)
+    if(!is.list(cache_contents) || is.null(cache_contents$formulaLong) ||
+       is.null(cache_contents$dataLong) ||
+       is.null(cache_contents$formulaEvent) ||
+       is.null(cache_contents$dataEvent) ||
+       is.null(cache_contents$fit)) {
+      message("Invalid cache content")
+    } else {
+      if(identical(formulaLong, cache_contents$formulaLong) &&
+         identical(dataLong, cache_contents$dataLong) &&
+         identical(formulaEvent, cache_contents$formulaEvent) &&
+         identical(dataEvent, cache_contents$dataEvent)
+         ) {
+        fit <- cache_contents$fit
+      } else {
+        message("Cache file out of date, refitting")
+      }
+    }
+  }
+
+  if(is.null(fit)) {
+    fit <- rstanarm::stan_jm(formulaLong = formulaLong, dataLong = dataLong,
+                              formulaEvent = formulaEvent, dataEvent = dataEvent,
+                              ...)
+    if(!is.null(cache_file)) {
+      saveRDS(loo::nlist(formulaLong, dataLong, formulaEvent, dataEvent, fit), file = cache_file)
+    }
+  }
+
+  fit
+}

### 2020-09-12 manuscript/hmm.Rmd
More HMM models, work on manuscript

@@
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
-  rename(.time = day, .serie = patient_id, .observed = breathing) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
   mutate(.time = .time +1) %>% 
@@
   ungroup()
+
 ```
@@
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + mo(comorbidities_sum) * rate_group
-hcq_m5 <- brmshmmdata(hcq_m5_formula, series_data, ☻
+hcq_m5 <- brmshmmdata(hcq_m5_formula, series_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
 ```
+
+## hcq_m6 - D-dimer
+
+```{r}
+hcq_m6_series_data <- series_data %>% 
+  group_by(.serie) %>%
+  filter(any(!is.na(d_dimer))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "d_dimer", new_column_name = "peak_d_dimer", initial_value = 0) %>%
+  mutate(log_peak_d_dimer = log(peak_d_dimer + 1) - 4)
+
+hcq_m6_initial_states <- hcq_m6_series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+
+hcq_m6_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_d_dimer * rate_group
+
+hcq_m6 <- brmshmmdata(hcq_m6_formula, hcq_m6_series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m6_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m6_fit <- brmhmm(hcq_m6, cache_file = paste0(cache_dir,"/hcq_m6.rds"))
+
+summary(hcq_m6_fit$brmsfit)
+```
+
+```{r}
+hcq_m6_epred_rect <- posterior_epred_rect(hcq_m6_fit)
+hcq_m6_predicted_rect <- posterior_epred_to_predicted(hcq_m6_fit, hcq_m6_epred_rect)
+hcq_m6_predicted <- posterior_rect_to_long(hcq_m6_fit, hcq_m6_predicted_rect)
+```
+
+```{r}
+hcq_m6_predicted_df <- posterior_long_to_df(hcq_m6_fit$data, hcq_m6_predicted)
+
+all_ids <- hcq_m6_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m6_predicted_df, all_ids[series_id], hcq_m6_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m6_fit, hcq_m6_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect)
+pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect, states_from = 2:6)
+```
+
+## hcq_m7 - IL-6
+
+```{r}
+hcq_m7_series_data <- series_data %>% 
+  group_by(.serie) %>%
+  filter(any(!is.na(IL_6))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "IL_6", new_column_name = "peak_IL_6", initial_value = 1) %>%
+  mutate(log_peak_IL_6 = log(peak_IL_6))
+
+hcq_m7_initial_states <- hcq_m7_series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+
+hcq_m7_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_IL_6 * rate_group
+
+hcq_m7 <- brmshmmdata(hcq_m7_formula, hcq_m7_series_data, 
+                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m7_initial_states, prior = hcq_m1_prior,
+                  observed_state_data = hcq_m1_observed_state_data)
+
+hcq_m7_fit <- brmhmm(hcq_m7, cache_file = paste0(cache_dir,"/hcq_m7.rds"))
+
+summary(hcq_m7_fit$brmsfit)
+```
+
+```{r}
+hcq_m7_epred_rect <- posterior_epred_rect(hcq_m7_fit)
+hcq_m7_predicted_rect <- posterior_epred_to_predicted(hcq_m7_fit, hcq_m7_epred_rect)
+hcq_m7_predicted <- posterior_rect_to_long(hcq_m7_fit, hcq_m7_predicted_rect)
+```
+
+```{r}
+hcq_m7_predicted_df <- posterior_long_to_df(hcq_m7_fit$data, hcq_m7_predicted)
+
+all_ids <- hcq_m7_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  distinct() %>% arrange(hospital_id) %>% pull(.serie)
+step_size <- 6
+for(step in 1:ceiling(length(all_ids) / step_size)) {
+  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+  posterior_state_plot(hcq_m7_predicted_df, all_ids[series_id], hcq_m7_fit$data) %>% print()
+
+}
+```
+
+```{r}
+pp_check_last_state(hcq_m7_fit, hcq_m7_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect)
+pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect, states_from = 2:6)
+```

### 2020-09-12 manuscript/manuscript.Rmd
More HMM models, work on manuscript

@@
 
-```{r setup}
+```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+knitr::opts_chunk$set(echo = FALSE)
+
 devtools::load_all()
@@
 ```{r}
+
+data_table_1 <- rbind(
+  data$patient_data %>% mutate(group = "All"),
+  data$patient_data %>% mutate(group = if_else(ever_hcq, "HCQ", "No HCQ")),
+  data$patient_data %>% filter(any_d_dimer) %>% mutate(group = "Measured D-dimer"),
+  data$patient_data %>% filter(any_IL_6) %>% mutate(group = "Measured IL-6")
+  )
+
+sum_percent <- function(x) {
+  paste0(sum(x, na.rm = TRUE), " (", scales::percent(mean(x, na.rm = TRUE), accuracy = 1), ")")
+}
+
+mean_iqr <- function(x) {
+  paste0(round(mean(x, na.rm = TRUE)), " (", round(quantile(x, 0.25, na.rm = TRUE)), " - ", round(quantile(x, 0.75, na.rm = TRUE)), ")")  
+}
+
+data_table_1 %>% group_by(group) %>% 
+  summarise(
+    N = n(),
+    `Distinct sites` = length(unique(hospital_id)),
+    Female = sum_percent(sex == "F"),
+    `Age (mean, IQR)` = mean_iqr(age),
+    `Admitted for Covid` = sum_percent(admitted_for_covid),
+    `Took Azithromycin` = sum_percent(ever_az),
+    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
+    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
+    `Heart Failure` = sum_percent(heart_failure),
+    COPD = sum_percent(COPD),
+    Asthma = sum_percent(asthma),
+    `Other lung disease` = sum_percent(other_lung_disease),
+    Diabetes = sum_percent(diabetes),
+    `Renal Disease` = sum_percent(renal_disease),
+    `Liver Disease` = sum_percent(liver_disease),
+    `Smoking` = sum_percent(smoking),
+    `BMI (mean, IQR)` = mean_iqr(BMI),
+    .groups = "drop"
+  ) %>%
+  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Measured D-dimer","Measured IL-6"), ordered = TRUE)) %>%
+  arrange(group) %>%
+  as.data.frame() %>%
+  column_to_rownames("group") %>%
+  t() %>% 
+  knitr::kable()
+```
+
+```{r}
 n_patients <- nrow(data$patient_data)

### 2020-09-13 manuscript/manuscript.Rmd
Bit of writing

@@
 
-`brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
+Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
 The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths.
+This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e.
+that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
 

### 2020-09-13 manuscript/hmm.Rmd
Work towards counterfactual HMM predictions

@@
 ```{r}
-series_data <- wide %>%
+serie_data <- wide %>%
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
@@
 
-hcq_m1_initial_states <- series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+hcq_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
@@
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group
-hcq_m1 <- brmshmmdata(hcq_m1_formula, series_data, 
+hcq_m1 <- brmshmmdata(hcq_m1_formula, serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
 ```
+```{r}
+
+counterfactual_max_days <- 90
+counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
+  inner_join(data$patient_data, by = c("patient_id")) %>% 
+  left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+  mutate(.time = .time +1) %>% 
+  filter(.time >= 1) %>%
+  group_by(.serie) %>%
+  mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
+  ungroup()
+
+counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
+counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
+
+hcq_m1_counterfactual_took_hcq <- hcq_m1
+hcq_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
+
+hcq_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_took_hcq)
+
+hcq_m1_counterfactual_no_hcq <- hcq_m1
+hcq_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
+
+hcq_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_no_hcq)
+
+```
+
 ```{r}
@@
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
-hcq_m2 <- brmshmmdata(hcq_m2_formula, series_data, 
+hcq_m2 <- brmshmmdata(hcq_m2_formula, serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m3 <- brmshmmdata(hcq_m3_formula, series_data, 
+hcq_m3 <- brmshmmdata(hcq_m3_formula, serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m4 <- brmshmmdata(hcq_m4_formula, series_data, 
+hcq_m4 <- brmshmmdata(hcq_m4_formula, serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
 hcq_m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + mo(comorbidities_sum) * rate_group
-hcq_m5 <- brmshmmdata(hcq_m5_formula, series_data, 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + comorbidities_sum * rate_group #mo(comorbidities_sum) * rate_group
+hcq_m5 <- brmshmmdata(hcq_m5_formula, serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
@@
 ```{r}
-hcq_m6_series_data <- series_data %>% 
+hcq_m6_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m6_initial_states <- hcq_m6_series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+hcq_m6_initial_states <- hcq_m6_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
@@
 
-hcq_m6 <- brmshmmdata(hcq_m6_formula, hcq_m6_series_data, 
+hcq_m6 <- brmshmmdata(hcq_m6_formula, hcq_m6_serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m6_initial_states, prior = hcq_m1_prior,
@@
 ```{r}
-hcq_m7_series_data <- series_data %>% 
+hcq_m7_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m7_initial_states <- hcq_m7_series_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+hcq_m7_initial_states <- hcq_m7_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
@@
 
-hcq_m7 <- brmshmmdata(hcq_m7_formula, hcq_m7_series_data, 
+hcq_m7 <- brmshmmdata(hcq_m7_formula, hcq_m7_serie_data, 
                   hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m7_initial_states, prior = hcq_m1_prior,

### 2020-09-13 R/brms_tools.R

@@ -8,8 +8,9 @@ brm_with_cache <- function(cache_file = NULL, ...) {
        is.null(cache_contents$standata) || is.null(cache_contents$fit)) {
       message("Invalid cache content")
     } else {
-      if(identical(stancode, cache_contents$stancode) &&
+      if(#identical(stancode, cache_contents$stancode) &&
          identical(standata, cache_contents$standata)) {
+        warning("Currently not checking stancode, TODO")
         fit <- cache_contents$fit
       } else {
         message("Cache file out of date, refitting")
@@ -26,3 +27,5 @@ brm_with_cache <- function(cache_file = NULL, ...) {
 
   fit
 }
+
+#normalize_stancode <- function

### 2020-09-13 R/brmshmmdata.R

@@ -121,6 +121,10 @@ validate_brmshmmdata <- function(d) {
   d$serie_data$.serie <- validate_id(d$serie_data$.serie, "serie_data$.serie",
                                         force_no_gaps = TRUE)
 
+  #TODO make the binding between initial states and series explicit
+  if(length(d$initial_state) != length(unique(d$serie_data$.serie))) {
+    stop("Incorrect number of initial states")
+  }
 
   d
 }

### 2020-09-13 R/posterior_plots_hmm.R

@@ -2,7 +2,7 @@ model_color <- "#3993dd"
 data_color <- "#91171f"
 decoration_color <- "#ca5310"
 
-posterior_state_plot <- function(predictions_df, series_id = 1:max(predictions_df$.serie),
+posterior_state_plot <- function(predictions_df, series_id = unique(predictions_df$.serie),
                                  observed_data = NULL,
                                  predicted_labels = NULL) {
 

### 2020-09-13 R/predict_hmm.R

@@ -1,9 +1,13 @@
-posterior_epred_rect <- function(fit, nsamples = NULL) {
+posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL) {
   validate_brmshmmfit(fit)
 
   brms:::contains_samples(fit$brmsfit)
 
-  pred_rawdata <- fit$data
+  if(is.null(newdata)) {
+    pred_rawdata <- fit$data
+  } else {
+    pred_rawdata <- newdata
+  }
 
   data_hmm <- make_data_hmm(pred_rawdata)
 
@@ -43,12 +47,16 @@ posterior_epred_rect <- function(fit, nsamples = NULL) {
   result_rect
 }
 
-posterior_rect_to_long <- function(fit, posterior_rect) {
+posterior_rect_to_long <- function(fit, posterior_rect, newdata = NULL) {
   validate_brmshmmfit(fit)
   brms:::contains_samples(fit$brmsfit)
 
   nsamples <- dim(posterior_rect)[3]
-  pred_rawdata <- fit$data
+  if(is.null(newdata)) {
+    pred_rawdata <- fit$data
+  } else {
+    pred_rawdata <- newdata
+  }
 
   result <- array(NA_integer_, c(nsamples, nrow(pred_rawdata$serie_data)))
 

### 2020-09-16 manuscript/competing_risks.Rmd
Started writing manuscript
First usable set of hypothesis results

@@
 
+# Peak IL-6, peak D-dimer
+
 ```{r}
-hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+data_surv_IL_6 <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(IL_6))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "IL_6", new_column_name = "peak_IL_6", initial_value = 1) %>%
+  mutate(log_peak_IL_6 = log(peak_IL_6))
+
+data_surv_IL_6 %>% group_by(patient_id) %>% summarise(m_b = max(breathing)) %>% group_by(m_b) %>% summarise(n())
+
+f_competing_IL_6 <- Surv(daym1, day,status) ~ log_peak_IL_6
+
+fit_competing_IL_6 <- coxph(f_competing_IL_6, data = data_surv_IL_6, id = patient_id)
+fit_competing_IL_6
+
+# hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]] <-
+#   rbind(
+#   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_comorbidities_sum,
+#                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
+#   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_comorbidities_sum,
+#                                         coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)")
+#   )
+# hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
+```
+```{r}
+data_surv_d_dimer <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(d_dimer))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "d_dimer", new_column_name = "peak_d_dimer", initial_value = 0) %>%
+  mutate(log_peak_d_dimer = log(peak_d_dimer + 1) - 4)
+
+data_surv_d_dimer %>% group_by(patient_id) %>% summarise(m_b = max(breathing)) %>% group_by(m_b) %>% summarise(n())
+
+f_competing_d_dimer <- Surv(daym1, day,status) ~ log_peak_d_dimer
+
+fit_competing_d_dimer <- coxph(f_competing_d_dimer, data = data_surv_d_dimer, id = patient_id)
+fit_competing_d_dimer
+
+# hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]] <-
+#   rbind(
+#   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_comorbidities_sum,
+#                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
+#   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_comorbidities_sum,
+#                                         coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)")
+#   )
+# hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
 ```
@@
 # Lymphocyte count predicts?
+
+```{r}
+hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
+write_csv(hypothesis_res_all, path = here::here("manuscript", "competing_risks_res.csv"))
+```

### 2020-09-16 manuscript/hmm.Rmd
First usable set of hypothesis results

@@
 }
+
+do_pp_checks <- FALSE
+hmm_hypothesis_res_list <- list()
 
@@
 
-counterfactual_max_days <- 90
-counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
-  inner_join(data$patient_data, by = c("patient_id")) %>% 
-  left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
-  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
-  mutate(.time = .time +1) %>% 
-  filter(.time >= 1) %>%
-  group_by(.serie) %>%
-  mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
-  ungroup()
-
-counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
-counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
-
-hcq_m1_counterfactual_took_hcq <- hcq_m1
-hcq_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
-
-hcq_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_took_hcq)
-
-hcq_m1_counterfactual_no_hcq <- hcq_m1
-hcq_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
-
-hcq_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_no_hcq)
+# counterfactual_max_days <- 90
+# counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
+#   inner_join(data$patient_data, by = c("patient_id")) %>% 
+#   left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
+#   mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+#   mutate(.time = .time +1) %>% 
+#   filter(.time >= 1) %>%
+#   group_by(.serie) %>%
+#   mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
+#   ungroup()
+# 
+# 
+# counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
+# counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
+# 
+# 
+# 
+# hcq_m1_counterfactual_took_hcq <- hcq_m1
+# hcq_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
+# 
+# hcq_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_took_hcq)
+# 
+# hcq_m1_counterfactual_no_hcq <- hcq_m1
+# hcq_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
+# 
+# hcq_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_no_hcq)
 
@@
 ```{r}
+draws <- tidybayes::tidy_draws(hcq_m1_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m1_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "supportive"
+  )
+
+hmm_hypothesis_res_list[["hcq_m1_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "supportive"
+  )
+```
+
+```{r}
+if(do_pp_checks) {
   hcq_m1_epred_rect <- posterior_epred_rect(hcq_m1_fit)
@@
   hcq_m1_predicted <- posterior_rect_to_long(hcq_m1_fit, hcq_m1_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m1_predicted_df <- posterior_long_to_df(hcq_m1_fit$data, hcq_m1_predicted)
@@
   }
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m1_fit, hcq_m1_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect, states_from = 2:6)
+}
 ```
@@
 summary(hcq_m2_fit$brmsfit)
+```
+
+```{r}
+draws <- tidybayes::tidy_draws(hcq_m2_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m2_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive"
+  )
+
+hmm_hypothesis_res_list[["hcq_m2_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive"
+  )
 ```
@@
 ```
+
+```{r}
+draws <- tidybayes::tidy_draws(hcq_m3_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m3_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital"
+  )
+
+hmm_hypothesis_res_list[["hcq_m3_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital"
+  )
+```
+
 ```{r}
@@
 summary(hcq_m4_fit$brmsfit)
+```
+
+```{r}
+draws <- tidybayes::tidy_draws(hcq_m4_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m4_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital, az"
+  )
+
+hmm_hypothesis_res_list[["hcq_m4_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, az"
+  )
+
+hmm_hypothesis_res_list[["hcq_m4_death_az"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_azTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital, hcq"
+  )
+
+hmm_hypothesis_res_list[["hcq_m4_discharged_az"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_azTRUE + draws$`b_rate_groupimprove_one:took_azTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, hcq"
+  )
 ```
@@
 summary(hcq_m5_fit$brmsfit)
+```
+
+```{r}
+draws <- tidybayes::tidy_draws(hcq_m5_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m5_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+  )
+
+hmm_hypothesis_res_list[["hcq_m5_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+  )
 ```
@@
 ```{r}
+draws <- tidybayes::tidy_draws(hcq_m6_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m6_d_dimer_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_d_dimer,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$d_dimer_increases_death,
+    adjusted = "age, sex, hospital"
+  )
+
+```
+
+```{r}
 hcq_m6_epred_rect <- posterior_epred_rect(hcq_m6_fit)
@@
 ```{r}
+draws <- tidybayes::tidy_draws(hcq_m7_fit$brmsfit)
+hmm_hypothesis_res_list[["hcq_m7_IL_6_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_IL_6,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$IL_6_increases_death,
+    adjusted = "age, sex, hospital"
+  )
+
+```
+
+```{r}
 hcq_m7_epred_rect <- posterior_epred_rect(hcq_m7_fit)
@@
 ```
+
+# Environment
+
+```{r}
+hmm_hypothesis_res_all <- do.call(rbind, hmm_hypothesis_res_list)
+write_csv(hmm_hypothesis_res_all, path = here::here("manuscript", "hmm_res.csv"))
+```
+
+```{r}
+sessionInfo()
+```

### 2020-09-16 manuscript/jm.Rmd
First usable set of hypothesis results

@@
 library(rstanarm)
+library(tidybayes)
 
@@
 options(mc.cores = parallel::detectCores())
+
+jm_hypothesis_res_list <- list()
 
@@
 
-data_d_dimer %>% filter(d_dimer == 0)
+#data_d_dimer %>% filter(d_dimer == 0)
 # data_d_dimer %>% group_by(sex, ever_hcq) %>% summarise(pts = length(unique(patient_id)), n_dimer = sum(!is.na(d_dimer)))
+fit_d_dimer_age_sex <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
+              dataLong = data_d_dimer %>% filter(!is.na(d_dimer)),
+              formulaEvent = Surv(daym1, day, event_death) ~ age + sex,
+              dataEvent = data_d_dimer,
+              id_var = "patient_id",
+              time_var = "day",
+              cache_file = paste0(cache_dir, "/jm_d_dimer_age_sex.rds"))
 
+fit_d_dimer_age_sex
+
+#length(unique(data_d_dimer$patient_id))
+# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
+#   filter(n_d_dimer > 1) %>% pull(patient_id)
+#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])
+
+```
+
+```{r}
+jm_hypothesis_res_list[["d_dimer_age_sex"]] <-
+  bayesian_hypothesis_res_from_jm(
+    hypotheses$d_dimer_increases_death,
+    fit_d_dimer_age_sex,
+    coefficient_name = "Assoc|Long1|etavalue",
+    adjusted  = "age, sex")
+
+jm_hypothesis_res_list[["d_dimer_age_sex"]]
+```
+
+```{r}
 fit_d_dimer_age_sex_hcq <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
@@
 #length(unique(data_d_dimer$patient_id))
-to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
-  filter(n_d_dimer > 1) %>% pull(patient_id)
+# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
+#   filter(n_d_dimer > 1) %>% pull(patient_id)
 #data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])
@@
 ```
+
+```{r}
+jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]] <-
+  bayesian_hypothesis_res_from_jm(
+    hypotheses$d_dimer_increases_death,
+    fit_d_dimer_age_sex_hcq,
+    coefficient_name = "Assoc|Long1|etavalue",
+    adjusted  = "age, sex, hcq")
+
+jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]]
+```
+
 ```{r}
@@
 
+fit_IL_6_age_sex <- stan_jm_with_cache(formulaLong = log_IL_6 ~ day + (1 + day | patient_id),
+              dataLong = data_IL_6 %>% filter(!is.na(IL_6)),
+              formulaEvent = Surv(daym1, day, event_death) ~ age + sex,
+              dataEvent = data_IL_6,
+              id_var = "patient_id",
+              time_var = "day",
+              cache_file = paste0(cache_dir, "/jm_IL_6_age_sex.rds"))
+
+fit_IL_6_age_sex
+
+```
+```{r}
+jm_hypothesis_res_list[["IL_6_age_sex"]] <-
+  bayesian_hypothesis_res_from_jm(
+    hypotheses$IL_6_increases_death,
+    fit_IL_6_age_sex,
+    coefficient_name = "Assoc|Long1|etavalue",
+    adjusted  = "age, sex")
+
+jm_hypothesis_res_list[["IL_6_age_sex"]]
+
+```
+
+```{r}
 fit_IL_6_age_sex_hcq <- stan_jm_with_cache(formulaLong = log_IL_6 ~ day + (1 + day | patient_id),
@@
 fit_IL_6_age_sex_hcq
+```
+
+```{r}
+jm_hypothesis_res_list[["IL_6_age_sex_hcq"]] <-
+  bayesian_hypothesis_res_from_jm(
+    hypotheses$IL_6_increases_death,
+    fit_IL_6_age_sex_hcq,
+    coefficient_name = "Assoc|Long1|etavalue",
+    adjusted  = "age, sex, hcq")
+
+jm_hypothesis_res_list[["IL_6_age_sex_hcq"]]
 
 ```
+
 ```{r}
@@
 ```
+
+```{r}
+jm_hypothesis_res_all <- do.call(rbind, jm_hypothesis_res_list)
+write_csv(jm_hypothesis_res_all, path = here::here("manuscript", "jm_res.csv"))
+```

### 2020-09-16 manuscript/manuscript.Rmd
Visualising multiverse

@@
 ---
-title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: A dataset and exploratory analysis" 
-output: pdf_document
+title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
+output: 
+  #rticles::arxiv_article
+  pdf_document
+  # bookdown::pdf_document2: 
+  #   toc: false
+  ##bookdown::pdf_book: 
+  #   base_format: rticles::arxiv_article
 abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2.
@@
 Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR; TODO].
-Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; http://zotero.org/users/5567156/items/5VAKEW58].
-While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@https://www.acpjournals.org/doi/10.7326/M20-4207] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; http://zotero.org/users/5567156/items/GMR9FX9Z].
+Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
+While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z].
 Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
 Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
-No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; http://zotero.org/users/5567156/items/BHGT88US] For additional summary see (live review)
+No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US] For additional summary see [@http://zotero.org/users/5567156/items/YP62P6XJ].
 
 In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e.
-g., @http://zotero.org/users/5567156/items/WZPNBRYN; http://zotero.org/users/5567156/items/4HR7MUJ6; http://zotero.org/users/5567156/items/NXC9CDI9]
+g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9]
 
@@
 
-```{r}
+```{r summarytable, echo=FALSE}
 
@@
     N = n(),
-    `Distinct sites` = length(unique(hospital_id)),
-    Female = sum_percent(sex == "F"),
+   # `Distinct sites` = length(unique(hospital_id)),
+    Male = sum_percent(sex == "M"),
     `Age (mean, IQR)` = mean_iqr(age),
@@
     `BMI (mean, IQR)` = mean_iqr(BMI),
+    Deceased = sum_percent(outcome == "Death"),
+    Discharged = sum_percent(outcome == "Discharged"),
     .groups = "drop"
@@
   t() %>% 
-  knitr::kable()
+  knitr::kable(caption = "Patient characterstics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
 ```
+
+See Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.
 
@@
 
-```{r}
+In Figure \@ref(fig:progression) we show the disease progression for all patients.
+
+```{r progression, fig.cap="Disease progression for all patients included in the study"}
 my_paste <- function(x) {
@@
 wide  %>% 
+  filter(!is.na(breathing)) %>%
   arrange(hospital_id, patient_id, day) %>%
@@
 
+```{r multiverse_results}
+bayesian_cols <- cols(
+  hypothesis = col_character(),
+  model = col_character(),
+  adjusted = col_character(),
+  estimand = col_character(),
+  widest_CI_excl_reference = col_double(),
+  point_estimate = col_double(),
+  ci_low = col_double(),
+  ci_high = col_double()
+)
+res_bayesian <- 
+  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
+        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
+        )
+
+frequentist_cols <- cols(
+  hypothesis = col_character(),
+  model = col_character(),
+  adjusted = col_character(),
+  p_value = col_double(),
+  ci_low = col_double(),
+  ci_high = col_double()
+)
+
+res_frequentist <- 
+  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols))
+```
+
+```{r, fig.height=6, fig.width=}
+rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
+      res_frequentist %>% select(-p_value)) %>%
+  mutate(
+    model_adjusted = paste0(model, " - ", adjusted),
+    label = factor(model_adjusted, levels = model_adjusted, labels = str_wrap(model_adjusted, 30)),
+    hypothesis = factor(hypothesis, levels = names(hypotheses), labels = purrr::map_chr(hypotheses,.f = ~ .x$caption))) %>%
+  ggplot(aes(x = point_estimate, xmin = ci_low, xmax = ci_high, y = label)) +
+  geom_vline(xintercept = 0, color = "blue") +
+  geom_point() +
+  geom_errorbarh() +
+  facet_wrap(~hypothesis, ncol = 1, scales = "free") 
+  #theme(axis.text.y = element_text(size = 6))
+```
+
 # References

### 2020-09-16 R/hypotheses.R

@@ -24,16 +24,16 @@ frequentist_hypothesis_res_from_coxph <- function(
              estimand = "log(HR)",
              p_value = summ$coefficients[coefficient_id, "Pr(>|z|)"],
              point_estimate = summ$coefficients[coefficient_id, "coef"],
-             ci_low = summ$conf.int[coefficient_id, "lower .95"],
-             ci_high = summ$conf.int[coefficient_id, "upper .95"]
+             ci_low = log(summ$conf.int[coefficient_id, "lower .95"]),
+             ci_high = log(summ$conf.int[coefficient_id, "upper .95"])
   )
 }
 
 
 bayesian_hypothesis_res_from_jm <- function(
   hypothesis, jm_fit, coefficient_name, adjusted) {
-  bayesian_hypothesis_res_from_fit(
-    draws = tidybayes::tidy_draws(fit)[[coefficient_name]],
+  bayesian_hypothesis_res_from_draws(
+    draws = tidybayes::tidy_draws(jm_fit)[[coefficient_name]],
     model = "jm",
     estimand = "log(HR)",
     hypothesis = hypothesis,

### 2020-09-17 manuscript/competing_risks.Rmd
Workable draft

--- 2020-09-11 manuscript/competing_risks.Rmd
@@
   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az")
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az,
+                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq")
   )
 hypothesis_res_list[["competing_hcq_az"]]
+```
+
+```{r}
+f_competing_hcq_az_admitted <- update.formula(f_competing_hcq_az, . ~ . + admitted_for_covid)
+fit_competing_hcq_az_admitted <- coxph(f_competing_hcq_az_admitted, data = data_surv, id = patient_id)
+
+fit_competing_hcq_az_admitted
+
+hypothesis_res_list[["competing_hcq_az_admitted"]] <-
+  rbind(
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, admitted, az"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, admitted, az"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, admitted, hcq"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, admitted, hcq")
+  )
+hypothesis_res_list[["competing_hcq_az_admitted"]]
 ```
@@
   frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_hospital,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital")
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az_hospital,
+                                        coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq, hospital"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az_hospital,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq, hospital")
   )
@@
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+write_csv(hypothesis_res_all, path = here::here("manuscript", "competing_risks_res.csv"))
 ```
@@
 
+# Peak IL-6, peak D-dimer
+
 # PCR positivity predicts

### 2020-09-17 manuscript/hmm.Rmd
Workable draft

@@
 ```{r}
+if(do_pp_checks) {
   hcq_m2_epred_rect <- posterior_epred_rect(hcq_m2_fit)
@@
   hcq_m2_predicted <- posterior_rect_to_long(hcq_m2_fit, hcq_m2_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m2_predicted_df <- posterior_long_to_df(hcq_m2_fit$data, hcq_m2_predicted)
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m2_fit, hcq_m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect, states_from = 2:6)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m3_epred_rect <- posterior_epred_rect(hcq_m3_fit)
@@
   hcq_m3_predicted <- posterior_rect_to_long(hcq_m3_fit, hcq_m3_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m3_predicted_df <- posterior_long_to_df(hcq_m3_fit$data, hcq_m3_predicted)
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m3_fit, hcq_m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect, states_from = 2:6)
+}
 ```
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
+    hypothesis = hypotheses$az_reduces_death,
     adjusted = "age, sex, supportive, hospital, hcq"
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
+    hypothesis = hypotheses$az_increases_discharged,
     adjusted = "age, sex, supportive, hospital, hcq"
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m4_epred_rect <- posterior_epred_rect(hcq_m4_fit)
@@
   hcq_m4_predicted <- posterior_rect_to_long(hcq_m4_fit, hcq_m4_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m4_predicted_df <- posterior_long_to_df(hcq_m4_fit$data, hcq_m4_predicted)
@@
   }
+}  
 ```
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m4_fit, hcq_m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect, states_from = 2:6)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m5_epred_rect <- posterior_epred_rect(hcq_m5_fit)
@@
   hcq_m5_predicted <- posterior_rect_to_long(hcq_m5_fit, hcq_m5_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m5_predicted_df <- posterior_long_to_df(hcq_m5_fit$data, hcq_m5_predicted)
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m5_fit, hcq_m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect, states_from = 2:6)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m6_epred_rect <- posterior_epred_rect(hcq_m6_fit)
@@
   hcq_m6_predicted <- posterior_rect_to_long(hcq_m6_fit, hcq_m6_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m6_predicted_df <- posterior_long_to_df(hcq_m6_fit$data, hcq_m6_predicted)
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m6_fit, hcq_m6_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect, states_from = 2:6)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m7_epred_rect <- posterior_epred_rect(hcq_m7_fit)
@@
   hcq_m7_predicted <- posterior_rect_to_long(hcq_m7_fit, hcq_m7_predicted_rect)
+}
 ```
@@
 ```{r}
+if(do_pp_checks) {
   hcq_m7_predicted_df <- posterior_long_to_df(hcq_m7_fit$data, hcq_m7_predicted)
@@
 ```{r}
+if(do_pp_checks) {
   pp_check_last_state(hcq_m7_fit, hcq_m7_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
@@
   pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect, states_from = 2:6)
+}
 ```

### 2020-09-17 manuscript/manuscript.Rmd
Workable draft

@@
   #rticles::arxiv_article
-  pdf_document
-  # bookdown::pdf_document2: 
-  #   toc: false
-  ##bookdown::pdf_book: 
+  #pdf_document
+  bookdown::pdf_document2: 
+     toc: false
+  # bookdown::pdf_book: 
   #    base_format: rticles::arxiv_article
-abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2.
+abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
@@
 library(tidyverse)
+library(patchwork)
 theme_set(cowplot::theme_cowplot())
@@
 First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
+We primarily use a model with competing risks for death and discharge from hospital (see Figure \@ref(fig:modelstates)a).
 
@@
 that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
+We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening.
+The states and available transitions are shown in Figure \@ref(fig:modelstates)b.
 
-`stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5]
+```{r modelstates, fig.cap="States used in the competing risk model (a) and in the hidden Markov model (b).
+AA = Ambient air, Oxygen = Nasal oxygen, NIPPV = Non-invasive positive-pressure ventilation, ECMO = Extra-corporeal membrane oxygenation.
+In both models the 'Death' and 'Discharged' states are terminal.
+In the hidden Markov model, the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
+knitr::include_graphics(here::here("manuscript","static_figs","model_states.pdf"))
+```
+
+When markers are used as predictors in both the hidden Markov models and the competing risks model, we impute the missing values using either the maximum of the marker so far or the last-known value of the marker.
+
+Additionally we used joint models for longitudinal and time-to-event data from `stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5].
+Those models include a longitudinal component to model the evolution over time of marker for each patient and a time-to-event component which assumes proportional-hazards for an event.
+The estimated value of the marker is then used as one of the predictors in the time-to-event component.
+
+```{r}
+#Since `stan_jm` does not support competing risks, we imputed the second event...
+```
+
+Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g. the effect of taking a drug is only included for the days after the drug was taken.
 
@@
 # Results
+
+```{r}
+n_patients <- nrow(data$patient_data)
+n_hospitals <- length(unique(data$patient_data$hospital_id))
+```
+
+In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.
 
@@
   t() %>% 
-  knitr::kable(caption = "Patient characterstics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
-```
-
-See Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.
-
-```{r}
-n_patients <- nrow(data$patient_data)
-n_hospitals <- length(unique(data$patient_data$hospital_id))
+  knitr::kable(caption = "Patient characteristics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
 ```
@@
 
-```{r progression, fig.cap="Disease progression for all patients included in the study"}
+```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
+AA = Ambient air, Oxygen = Nasal oxygen, NIPPV = Non-invasive positive-pressure ventilation, ECMO = Extra-corporeal membrane oxygenation."}
 my_paste <- function(x) {
@@
   ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
-  geom_tile() + 
-  scale_fill_viridis_d(direction = -1) +
+  geom_tile(height = 0.7) + 
+  scale_fill_viridis_d("", direction = -1) +
+  scale_x_continuous("Days since hospitalization") +
   theme(axis.text.y = element_blank()) +
@@
 
-```{r multiverse_results}
+```{r}
 bayesian_cols <- cols(
@@
   rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols))
+
+all_hypo <- rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
+      res_frequentist %>% select(-p_value)) %>%
+    inner_join(hypotheses_df, by = c("hypothesis" = "name"))
+
 ```
 
-```{r, fig.height=6, fig.width=}
-rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
-      res_frequentist %>% select(-p_value)) %>%
-  mutate(
+```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between treatments and main outcomes.
+Each row represents a model - HMM = Bayesian hidden-markov model, coxph = Frequentist competing risks model.
+Additionally the factors the model adjusted for are listed - supportive = best supportive care initiated, comorbidities_sum = total number of comorbidities, admitted = admitted for Covid-19, az = took Azithromycin, hcq = took Hydroxychloroquine.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval."}
+plot_multiverse <- function(hypo_res) {
+  hypo_res %>% mutate(
+    id = rev(1:n()),
     model_adjusted = paste0(model, " - ", adjusted),
     label = factor(model_adjusted, levels = model_adjusted, labels = str_wrap(model_adjusted, 30)),
-    hypothesis = factor(hypothesis, levels = names(hypotheses), labels = purrr::map_chr(hypotheses,.f = ~ .x$caption))) %>%
-  ggplot(aes(x = point_estimate, xmin = ci_low, xmax = ci_high, y = label)) +
+    model_adjusted_fct = fct_reorder(factor(model_adjusted),  id)) %>%
+  ggplot(aes(x = point_estimate, xmin = ci_low, xmax = ci_high, y = model_adjusted_fct)) +
   geom_vline(xintercept = 0, color = "blue") +
@@
   geom_errorbarh() + 
-  facet_wrap(~hypothesis, ncol = 1, scales = "free") 
+  facet_wrap(~caption, ncol = 1, scales = "free_y") +
+  scale_x_continuous("Model coefficient estimate") +
+  scale_y_discrete("")
   #theme(axis.text.y = element_text(size = 6))
+}
+
+multiverse_hcq <- plot_multiverse(all_hypo %>% filter(group == "hcq"))
+multiverse_az <- plot_multiverse(all_hypo %>% filter(group == "az"))
+
+(multiverse_hcq / multiverse_az) + plot_layout(heights = c(2.2,1))
+
 ```
 
+```{r multiverseresultsmarkers, fig.height=2.5, fig.width=6, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
+Each row represents a model - HMM = Bayesian hidden-markov model, jm = Joint longitudinal and time-to-event model.
+Additionally the factors the model adjusted for are listed - hcq = took Hydroxychloroquine.
+We show posterior mean and 95% credible interval."}
+plot_multiverse(all_hypo %>% filter(group == "markers"))
+```
+
+The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure \@ref(fig:multiverseresultsdrugs).
+In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
+Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged).
+After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome.
+In all cases, the results seem to not be strongly affected by model choice.
+
+We additionally investigated the association of several markers with disease progression.
+Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice.
+All of the coefficient estimates can be seen in Figure \@ref(fig:multiverseresultsmarkers)
+
+# Conclusions
+
+Will be written once we have the full dataset.
+
+Where our resutls contradict randomized trials, the most likely explanation is systematic bias in our dataset.
+
 # References

### 2020-09-17 manuscript/manuscript.Rmd
ACP score - basic treatment

@@
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
-To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom."
+To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
+Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
 bibliography: Covid19.json
+date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
 ---
@@
 
+Additionally, many methods for predicting disease severity of Covid-19 were published, but the methods are at high risk of bias and lack external validation [@http://zotero.org/users/5567156/items/TS57HJ3T].
+
 # Methods
@@
 First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
-We primarily use a model with competing risks for death and discharge from hospital (see Figure \@ref(fig:modelstates)a).
+We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).
 
@@
 We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening.
-The states and available transitions are shown in Figure \@ref(fig:modelstates)b.
+The states and available transitions are shown in Figure&nbsp;\@ref(fig:modelstates)b.
 
@@
 ```
+
+## Evaluating prognostic models
+
+We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e.
+where we have gathered all the input features).
+This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5].
 
@@
 
-In Figure \@ref(fig:progression) we show the disease progression for all patients.
+In Figure&nbsp;\@ref(fig:progression) we show the disease progression for all patients.
 
@@
 
+## Association between treatment, markers and outcomes
+
 ```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between treatments and main outcomes.
@@
 
-The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure \@ref(fig:multiverseresultsdrugs).
+The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
 In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
@@
 Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice.
-All of the coefficient estimates can be seen in Figure \@ref(fig:multiverseresultsmarkers)
+All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)
+
+## Evaluating prediction models
+
+The ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] is a simple score that categorizes the patients into 3 categories based on age and C-reactive protein (CRP): grade 1 (age $<$ 60 years and CRP $<$ 34 mg/L), grade 2 (age $\geq$ 60
+years and CRP $<$ 34 mg/L or age $<$ 60 years and CRP $\geq$ 34 mg/L), grade 3 (age $\geq$ 60 years and CRP $\geq$ 34 mg/L) and was evaluated primarily on 12-day mortality.
+Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in rough agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).
+
+```{r acpresults, echo=FALSE}
+acp_data <- wide %>%  
+  group_by(patient_id) %>%
+  filter(any(!is.na(CRP))) %>%
+  summarise(
+    first_CRP_day = min(day[!is.na(CRP)]),
+    first_CRP = CRP[day == first_CRP_day],
+    worst_breathing_12 = max(breathing[day >= first_CRP_day & day <= first_CRP_day + 12]),
+    .groups = "drop") %>% 
+  inner_join(data$patient_data, by = c("patient_id")) %>%
+  filter(outcome %in% c("Death","Discharged") | last_record - first_CRP_day >= 12) %>%
+  mutate(    
+    alive_12 = last_record > first_CRP_day + 12 | outcome == "Discharged",
+    ACP_grade = case_when(
+       age < 60 & first_CRP < 34 ~ 1,
+       age >= 60 & first_CRP >= 34 ~ 3,
+       TRUE ~ 2
+    ) 
+  )
+
+acp_data %>%
+  group_by(ACP_grade) %>%
+  summarise(`patients` = n(), `12-day mortality` = scales::percent(mean(!alive_12), accuracy = 0.1), .groups = "drop") %>%
+  arrange(ACP_grade) %>%
+  mutate(`Li et al. 95% CI` = c("0%", "0 - 11.3%", "19.8 - 44.3%")) %>%
+  rename(`ACP grade` = ACP_grade) %>%
+  knitr::kable(booktabs = TRUE, caption = "Comparing the 12-day mortality based on ACP score (where available) with the 95\\% confidence intervals (CI) for 12-day mortality reported by Lu et al.")
+
+# acp_data %>%
+#   group_by(ACP_grade, worst_breathing_12) %>%
+#   summarise(mean(alive_12))
+```
+
+(TODO: the other prediction models)
 

### 2020-09-17 R/data_preprocessing.R

@@ -65,7 +65,9 @@ compute_derived_quantities_patients <- function(data) {
               ever_az = any(took_az),
               ever_convalescent_plasma = any(took_convalescent_plasma),
               any_d_dimer = any(!is.na(d_dimer)),
-              any_IL_6 = any(!is.na(IL_6))
+              any_IL_6 = any(!is.na(IL_6)),
+              worst_breathing = max(breathing),
+              .groups = "drop"
     )
 
   data$patient_data <- data$patient_data %>%

### 2020-09-20 manuscript/risk_scores.Rmd
Notes on risk scores

---
title: "Validating risk scores"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE)

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- prepare_marker_data_wide(data)
```

# ACP

# Bello-Chavolla

- Requires pneumonia diagnosis (which we probably can't infer)... Maybe from using antibiotics?
- Early onset diabetes = diabetes & 40 years

- Age ≥ 65 years  3 
Diabetes  1
Diabetes*Age < 40 years 5
Age < 40 years -6
Obesity 1
Pneumonia 7
CKD 3
COPD 1
Immunosuppression 1

# Chen and Liu

𝑅𝑖𝑠𝑘𝑆𝑐𝑜𝑟𝑒𝑠=−4.211+0.013×𝐶𝑅𝑃+0.059×𝐴𝑔𝑒+0.112×𝐷𝐷−1.984×𝐿𝑌𝑀

Results for the validation dataset:
AUROC = 0.881
Cutoff (0.5)
Sensitivity = 0.839
Specificity = 0.794

```{r}

wide$ly
```

# Shi et al.

1 point for Age >= 50, Male, Hypertension

Not sure what "severe" means

### 2020-09-21 manuscript/jm.Rmd
Added model_check and data_version to results CSV files

@@
     coefficient_name = "Assoc|Long1|etavalue",
-    adjusted  = "age, sex")
+    adjusted  = "age, sex",
+    model_check = "Suspicious")
 
@@
     coefficient_name = "Assoc|Long1|etavalue",
-    adjusted  = "age, sex, hcq")
+    adjusted  = "age, sex, hcq",
+    model_check = "Problematic")
 
@@
     coefficient_name = "Assoc|Long1|etavalue",
-    adjusted  = "age, sex, hcq")
+    adjusted  = "age, sex, hcq",
+    model_check = "Suspicious")
 

### 2020-09-21 R/data_preprocessing.R

@@ -227,3 +227,7 @@ compute_marker_peak <- function(markers_wide, column_name, new_column_name, init
 
   markers_wide
 }
+
+get_data_version <- function() {
+  readLines(here::here("private_data", "data_revision.txt"))
+}

### 2020-09-21 R/hypotheses.R

@@ -16,7 +16,7 @@ hypotheses_df <- purrr::map_dfr(hypotheses, as.data.frame) %>%
   mutate(name = factor(name, levels = name))
 
 frequentist_hypothesis_res_from_coxph <- function(
-  hypothesis, coxph_fit, coefficient_name, transition, adjusted) {
+  hypothesis, coxph_fit, coefficient_name, transition, adjusted, model_check = "OK") {
 
   summ <- summary(coxph_fit)
   coefficient_id <- summ$cmap[coefficient_name, transition]
@@ -27,24 +27,27 @@ frequentist_hypothesis_res_from_coxph <- function(
              p_value = summ$coefficients[coefficient_id, "Pr(>|z|)"],
              point_estimate = summ$coefficients[coefficient_id, "coef"],
              ci_low = log(summ$conf.int[coefficient_id, "lower .95"]),
-             ci_high = log(summ$conf.int[coefficient_id, "upper .95"])
+             ci_high = log(summ$conf.int[coefficient_id, "upper .95"]),
+             model_check = model_check,
+             data_version = get_data_version()
   )
 }
 
 
 bayesian_hypothesis_res_from_jm <- function(
-  hypothesis, jm_fit, coefficient_name, adjusted) {
+  hypothesis, jm_fit, coefficient_name, adjusted, model_check = "OK") {
   bayesian_hypothesis_res_from_draws(
     draws = tidybayes::tidy_draws(jm_fit)[[coefficient_name]],
     model = "jm",
     estimand = "log(HR)",
     hypothesis = hypothesis,
-    adjusted = adjusted)
+    adjusted = adjusted,
+    model_check = model_check)
 }
 
 bayesian_hypothesis_res_from_draws <- function(
   draws, model, estimand,
-  hypothesis, adjusted) {
+  hypothesis, adjusted, model_check = "OK") {
 
   reference_point <- 0
   reference_location <- ecdf(draws)(reference_point)
@@ -56,6 +59,8 @@ bayesian_hypothesis_res_from_draws <- function(
              point_estimate = mean(draws),
              ci_low = quantile(draws, 0.025),
              ci_high = quantile(draws, 0.975),
+             model_check = model_check,
+             data_version = get_data_version(),
              row.names = NULL
   )
 }

### 2020-09-21 manuscript/descriptive.Rmd
More work on manuscript
First working stan_jm models
Tweaks to derived data, updated data_description.md

@@
 ```{r}
+data$patient_data %>% filter(hospital_id == "QKuFp") %>% select(first_day_invasive, last_day_invasive)
+```
+
+```{r}
 data$breathing_data %>% group_by(hospital_id, patient_id) %>%
@@
 
-```{r}
-outcome_data %>% filter(day > 30)
-data$patient_data %>% filter(patient_id == "kOw52xii4l")
-data_for_plot %>% filter(is.na(day))
-```
 ```{r}

### 2020-09-21 R/data_mapping.R

@@ -48,7 +48,7 @@ marker_map <- c(
         "ampicillin/sulbactam" = "ampicillin_sulbactam",
         "Ciprofloxacin" = "ciprofloxacin",
         "Meropenem" = "meropenem",
-        "Metronidazol" = "Metronidazol",
+        "Metronidazol" = "metronidazol",
       ## Other drugs
         "Fluconazol" = "fluconazol",
         "zinkorot" = "zinc",
@@ -57,6 +57,27 @@ marker_map <- c(
         "Tamiflu" = "tamiflu",
         "Lexaurin" = "lexaurin"
 )
+
+all_antibiotics <- c(
+  "az",
+  "amoxiclav",
+  "piperacilin_tazobactam",
+  "clarithromycin",
+  "sulfamethoxazol",
+  "trimethoprim",
+  "ofloxacin",
+  "ceftriaxone",
+  "cefuroxime",
+  "cefotaxime",
+  "cefepime",
+  "vancomycin",
+  "ampicillin_sulbactam",
+  "ciprofloxacin",
+  "meropenem",
+  "metronidazol"
+)
+
+all_macrolides <- c("az", "clarithromycin")
 
 pcr_values_positive <- c("pos","neg", "neg and pos", "susp.", "susp", "poz")
 pcr_value_negative <- "neg"

### 2020-09-21 R/data_preprocessing.R

@@ -28,7 +28,10 @@ read_data_for_analysis <- function() {
                              last_record = col_integer(),
                              outcome = col_factor()
                            )) %>%
-    mutate(age_norm = (age - 65) / 20)
+    mutate(age_norm = (age - 65) / 20,
+           high_creatinin = if_else(sex == "M", creatinin > 115, creatinin > 97),
+           high_pt_inr = pt_inr > 1.2,
+           low_albumin = albumin < 36)
 
   breathing_data <- read_csv(here::here("private_data", "breathing_data.csv"),
                            col_types = cols(
@@ -49,10 +52,32 @@ read_data_for_analysis <- function() {
                               censored = col_character()
                             ))
 
+  multiple_units <- data$marker_data %>%
+    group_by(marker) %>%
+    summarise(n_units = length(unique(unit))) %>%
+    filter(n_units > 1)
+  if(nrow(multiple_units) > 1) {
+    print(multiple_units)
+    stop("Multiple units for some markers")
+  }
+
+
   data <- loo::nlist(patient_data, breathing_data, marker_data)
   data$marker_data_wide <- prepare_marker_data_wide(data)
 
   data <- compute_derived_quantities_patients(data)
+
+  # TODO Reconstruct long from wide (because wide has more derived quantities and covers all days)
+  # Would require keeping units around in marker_data_wide
+
+
+  # Remove unsed xxx_censored fields from wide
+  censored_columns <- names(data$marker_data_wide)[grepl("_censored$", names(data$marker_data_wide))]
+  for(censored_col in censored_columns) {
+    if(all(is.na(data$marker_data_wide[[censored_col]]) | data$marker_data_wide[[censored_col]] == "none")) {
+      data$marker_data_wide[[censored_col]] <- NULL
+    }
+  }
 
   data
 }
@@ -64,9 +89,13 @@ compute_derived_quantities_patients <- function(data) {
     summarise(ever_hcq = any(took_hcq),
               ever_az = any(took_az),
               ever_convalescent_plasma = any(took_convalescent_plasma),
+              ever_antibiotics = any(took_antibiotics),
               any_d_dimer = any(!is.na(d_dimer)),
               any_IL_6 = any(!is.na(IL_6)),
               worst_breathing = max(breathing),
[…]

### 2020-09-22 manuscript/manuscript.Rmd
Cite the new prophylaxis preprint

@@
 ---
-title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
+title: "Detailed disease progression of 97 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
 output: 
@@
 data <- read_data_for_analysis()
-wide <- prepare_marker_data_wide(data)
+wide <- data$marker_data_wide
 ```
@@
 Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
-While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI] and as post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z].
+While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
 Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].

### 2020-10-02 manuscript/risk_scores.Rmd
Further work on risk scores

@@
 title: "Validating risk scores"
-output: html_notebook
+output: 
+  bookdown::pdf_document2: 
+     toc: false
+bibliography: Covid19.json
+date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
 ---
@@
 data <- read_data_for_analysis()
-wide <- prepare_marker_data_wide(data)
+```
+
+```{r}
+baseline <- data$patient_data %>%
+  mutate(alive_12_admission = case_when(
+    last_record >= 12 ~ TRUE,
+    outcome == "Discharged" ~ TRUE,
+    outcome == "Death" ~ FALSE,
+    TRUE ~ NA),
+    alive_30_admission = case_when(
+    last_record >= 30 ~ TRUE,
+    outcome == "Discharged" ~ TRUE,
+    outcome == "Death" ~ FALSE,
+    TRUE ~ NA)
+  )
+
 ```
@@
 
-# Bello-Chavolla
+The ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] is a simple score that categorizes the patients into 3 categories based on age and C-reactive protein (CRP): grade 1 (age $<$ 60 years and CRP $<$ 34 mg/L), grade 2 (age $\geq$ 60
+years and CRP $<$ 34 mg/L or age $<$ 60 years and CRP $\geq$ 34 mg/L), grade 3 (age $\geq$ 60 years and CRP $\geq$ 34 mg/L) and was evaluated primarily on 12-day mortality.
+Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).
 
-- Requires pneumonia diagnosis (which we probably can't infer)... Maybe from using antibiotics?
-- Early onset diabetes = diabetes & 40 years
+```{r acpresults, echo=FALSE}
+acp_data <- data$marker_data_wide %>%  
+  group_by(patient_id) %>%
+  filter(any(!is.na(CRP))) %>%
+  summarise(
+    first_CRP_day = min(day[!is.na(CRP)]),
+    first_CRP = CRP[day == first_CRP_day],
+    worst_breathing_12 = max(breathing[day >= first_CRP_day & day <= first_CRP_day + 12]),
+    .groups = "drop") %>% 
+  inner_join(data$patient_data, by = c("patient_id")) %>%
+  filter(outcome %in% c("Death","Discharged") | last_record - first_CRP_day >= 12) %>%
+  mutate(    
+    alive_12 = case_when(last_record >= first_CRP_day + 12 | outcome == "Discharged" ~ TRUE,
+                         outcome == "Death" ~ FALSE,
+                         TRUE ~ NA),
+    alive_30 = case_when(last_record >= first_CRP_day + 30 | outcome == "Discharged" ~ TRUE,
+                         outcome == "Death" ~ FALSE,
+                         TRUE ~ NA),
+    ACP_grade = case_when(
+       age < 60 & first_CRP < 34 ~ 1,
+       age >= 60 & first_CRP >= 34 ~ 3,
+       TRUE ~ 2
+    ) 
+  )
 
-- Age ≥ 65 years  3 
-Diabetes  1
-Diabetes*Age < 40 years 5
-Age < 40 years -6
-Obesity 1
-Pneumonia 7
-CKD 3
-COPD 1
-Immunosuppression 1
+acp_data %>%
+  filter(!is.na(alive_12)) %>%
+  group_by(ACP_grade) %>%
+  summarise(`patients` = n(), `12-day mortality` = sum_percent(!alive_12), .groups = "drop") %>%
+  arrange(ACP_grade) %>%
+  mutate(`Li et al. 95% CI` = c("0%", "0 - 11.3%", "19.8 - 44.3%")) %>%
+  rename(`ACP grade` = ACP_grade) %>%
+  knitr::kable(booktabs = TRUE, caption = "Comparing the 12-day mortality based on ACP score (where available) with the 95\\% confidence intervals (CI) for 12-day mortality reported by Lu et al.")
+
+# acp_data %>%
+#   group_by(ACP_grade, worst_breathing_12) %>%
+#   summarise(mean(alive_12))
+```
+```{r}
+li_et_al_data <- data.frame(
+  ACP_grade = c(rep(1, 194), rep(2, 145), rep(3, 95)),
+  alive_12 = c(rep(TRUE, 194), #Grade 1, 0% mortality
+               rep(FALSE, 8), rep(TRUE, 145-8), #Gred 2, 5.6% mortality
+               rep(FALSE, 32), rep(TRUE, 95 - 32) #Grade 3, 33.2% mortality
+  ))
+
+acp_auc_orig <- my_auc(alive_12 ~ ACP_grade, li_et_al_data)
+acp_auc_orig$summary
+```
+
+```{r}
+outcome_caption = c("alive_12" = "12-day mortality", "alive_30" = "30-day mortality")
+score_caption = c("ACP_grade" = "Li et al. (ACP)")
+
+my_auc <- function(f, d) {
+  roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
+  ci = pROC::ci.auc(roc)
+
+  outcome = as.character(formula.tools::lhs(f))
+  score = as.character(formula.tools::rhs(f))
+
+  list(roc = roc,
+       summary = data.frame(
+         auc = ci[2],
+         auc_low = ci[1],
+         auc_high = ci[3],
+         outcome = outcome,
+         score = score
+       )
+  )
+}
+
+plot_my_auc <- function(my_auc) {
+  caption = paste0(score_caption[my_auc$summary$score], ", ", outcome_caption[my_auc$summary$outcome])
+  plot(my_auc$roc, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE, main = caption)
+}
+
+plot_all_auc <- function(my_auc_list) {
+  my_auc_list %>% walk(plot_my_auc)
+}
+```
+
+```{r}
+acp_auc <- list(
+  alive_12 = my_auc(alive_12 ~ ACP_grade, acp_data),
+  alive_30 = my_auc(alive_30 ~ ACP_grade, acp_data)
+)
+plot_all_auc(acp_auc)
+
+```
 
@@
 
-𝑅𝑖𝑠𝑘𝑆𝑐𝑜𝑟𝑒𝑠=−4.211+0.013×𝐶𝑅𝑃+0.059×𝐴𝑔𝑒+0.112×𝐷𝐷−1.984×𝐿𝑌𝑀
+$$
+Score = −4.211+0.013 \times CRP+0.059 \times Age +0.112 \times DD −1.984 \times LYM
+$$
+
+The paper does not have units, but the web app
+
+> patients  who  were  with  >10%  missing  values,  stayed  in  the  hospital  <  7  days,  afflicted  by  a  severe  disease  before  admission  (e.g.,  cancer,  aplastic  anemia,  and  uremia),  were  unconscious  at  admission, and were directly admitted to the intensive care unit (ICU) were excluded.
+
+> Non-normally   distributed   continuous   variables   were   transformed   using   a   Box-Cox   transformation.
+
+Slightly unclear how they computed the outcome.
+
+CRP: mg/l
+D-dimer: ug / mL FEU (we use ng/ml DDU, FEU value is double the DDU)
+Lymphocytes: 10^9 / L
+
+```{r}
+data$marker_data %>% filter(marker %in% c("CRP", "d_dimer", "lymphocyte_count")) %>%
+  select(marker, unit) %>% distinct()
+```
 
@@
 
+Precise coefficient values and the CI range determined by reverse engineering the webapp at https://phenomics.fudan.edu.cn/risk_scores/
+
 ```{r}
+chen_liu_data <- data$marker_data_wide %>%  
+  mutate(has_all_values = !is.na(CRP) & !is.na(d_dimer) & !is.na(lymphocyte_count),
+         d_dimer_converted = d_dimer * 2 / 1000) %>%
+  group_by(patient_id) %>%
+  filter(any(has_all_values)) %>%
+  summarise(
+    first_day = min(day[has_all_values]),
+    first_CRP = CRP[day == first_day],
+    first_d_dimer_converted = d_dimer_converted[day == first_day],
+    first_lymphocyte_count = lymphocyte_count[day == first_day],
+    breathing_first = breathing[day == first_day],
+    .groups = "drop") %>% 
+  inner_join(data$patient_data, by = c("patient_id")) %>%
+  mutate(    
+    inclusion_criterium = breathing_first < "NIPPV" & (outcome %in% c("Discharged", "Death") | last_record >= first_day + 7),
+    score =  -4.21167 + 0.058699 * age + 0.013391 * first_CRP + 0.112055 * first_d_dimer_converted - 1.983733 * first_lymphocyte_count,
+    risk = 1 / (1 + exp(-score)),
+    risk_lower = 1 / (1 + exp(-score + 1.96*0.1721)),
+    risk_upper = 1 / (1 + exp(-score - 1.96*0.1721)),
+    alive_12 = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
+                         outcome == "Death" ~ FALSE,
+                         TRUE ~ NA),
+    alive_30 = case_when(last_record >= first_day + 30 | outcome == "Discharged" ~ TRUE,
+                         outcome == "Death" ~ FALSE,
+                         TRUE ~ NA),
+    alive_all = case_when(outcome == "Discharged" ~ TRUE,
+                         outcome == "Death" ~ FALSE,
+                         TRUE ~ NA),
+    alive_imputed = outcome != "Death"
 
-wide$ly
+  ) %>% select(age, starts_with("first"), score, starts_with("risk"), breathing_first, starts_with("alive"), inclusion_criterium)
+```
+
+```{r}
+chen_liu_data %>% ggplot(aes(x = score, y = alive_30)) + geom_point()
+```
+
+```{r}
+roc_chen_liu_12 <- pROC::roc(alive_12 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_chen_liu_30 <- pROC::roc(alive_30 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_chen_liu_all <- pROC::roc(alive_all ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+
+roc_chen_liu_12_incl <- pROC::roc(alive_12 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_chen_liu_30_incl <- pROC::roc(alive_30 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_chen_liu_all_incl <- pROC::roc(alive_all ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_chen_liu_imputed_incl <- pROC::roc(alive_imputed ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+
 ```
@@
 Not sure what "severe" means
+
+# Caramelo
+
+???
+
+# Bello-Chavolla
+
+The index given in [@http://zotero.org/users/5567156/items/INZVABQI] is built from following elements:
+
+```{r}
+bello_chavolla_table <- rbind(
+  data.frame(factor = "Age >= 65 years", score = 3),
+  data.frame(factor = "Diabetes", score = 1),
+  data.frame(factor = "Diabetes*Age < 40 years", score = 5),
+  data.frame(factor = "Age < 40 years", score = -6),
+  data.frame(factor = "Obesity", score = 1),
+  data.frame(factor = "Pneumonia", score = 7),
+  data.frame(factor = "Chronic Kidney Disease", score = 3),
+  data.frame(factor = "COPD", score = 1),
+  data.frame(factor = "Immunosuppression", score = 1)
+)
+
+bello_chavolla_table
+
+```
+
+Of those, our data does not contain Immunosuppresion and Pneumonia.
+We can however use taking antibiotics within 1 day of hospitalization as an imperfect proxy for pneumonia.
+We will ignore immunosuppresion in our test.
+Both of those choices could reduce the performance of the score.
+
+The authors do not define obesity, so we will use BMI > 30 as threshold for obesity.
+Our data contains an indicator for "Renal disease" which might be classified slightly differently than "Chronic Kidney Disease".
+
+The authors report C-statistic = 0.830 for mortality on validation dataset - it is unclear how they treat censoring, but they mention 30-day mortality in other part of the paper, so we will focus on 30-day mortality.
+
+```{r}
+atb_first_day <- data$marker_data_wide %>% select(patient_id, day, took_antibiotics) %>% 
+  group_by(patient_id) %>%
+  summarise(atb0 = any(took_antibiotics[day <= 0]), 
+            atb1 = any(took_antibiotics[day <= 1]), .groups = "drop")
+
+bello_chavolla_data <- baseline %>%
+  inner_join(atb_first_day, by = "patient_id") %>%
+  mutate(bc_base =  3 *  (age >= 65) +
+           diabetes +
+           5 * (diabetes & age < 40) +
+           (-6) * (age < 40) +
+           obesity +
+           3 * renal_disease +
+           COPD,
+         bc0 = bc_base + 7 * atb0,
+         bc1 = bc_base + 7 * atb1
+  )
+
+```
+
+```{r}
+bello_chavolla_data %>% ggplot(aes(x = bc0, y = as.integer(alive_30_admission))) +
+  geom_smooth() + geom_jitter(width = 0.3, height = 0.3) 
+bello_chavolla_data %>% ggplot(aes(x = bc_base, y = as.integer(alive_30_admission))) +
+  geom_smooth() + geom_jitter(width = 0.3, height = 0.3)
+
+```
+
+```{r}
+roc_bc_base_30 <- pROC::roc(alive_30_admission ~ bc_base, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_bc_base_12 <- pROC::roc(alive_12_admission ~ bc_base, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_bc0_30 <- pROC::roc(alive_30_admission ~ bc0, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_bc0_12 <- pROC::roc(alive_12_admission ~ bc0, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_bc1_30 <- pROC::roc(alive_30_admission ~ bc1, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+roc_bc1_12 <- pROC::roc(alive_12_admission ~ bc1, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+
+```

### 2020-10-02 R/data_mapping.R

@@ -107,6 +107,7 @@ unit_overwrites <- list(
     list(marker = "IgG", old_unit = "", new_unit = "TODO")
   ),
   YqNbe = list(
+    #TODO: check that d_dimer is DDU, not FEU
     #list(marker = "d_dimer", old_unit = "ng/ml DDU", new_unit = "\U03BCg/l"),
     list(marker = "d_dimer", old_unit = "ug/l", new_unit = "ng/ml DDU"),
     list(marker = "procalcitonin", old_unit = "unit", new_unit = "\U03BCg/l")

### 2020-10-02 R/manuscript_tools.R

@@ -1,0 +1,7 @@
+sum_percent <- function(x) {
+  paste0(sum(x, na.rm = TRUE), " (", scales::percent(mean(x, na.rm = TRUE), accuracy = 1), ")")
+}
+
+mean_iqr <- function(x) {
+  paste0(round(mean(x, na.rm = TRUE)), " (", round(quantile(x, 0.25, na.rm = TRUE)), " - ", round(quantile(x, 0.75, na.rm = TRUE)), ")")
+}

### 2020-10-04 manuscript/analysis_brms.Rmd
initial attempts to analyse effects of HCQ and AZ

---
title: "Analysis of Retrospective COVID-19 data"
author: "Paul Bürkner"
output:
  html_document:
    theme: default
encoding: UTF-8
editor_options: 
  chunk_output_type: console
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

---

# {.tabset}

## Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 10)
```

```{r general_options, cache=FALSE}
devtools::load_all()
library(tidyverse)
library(tidybayes)
library(brms)
options(width = 160)
options(brms.backend = "cmdstanr")
options(mc.cores = 4)
theme_set(bayesplot::theme_default())

cache_dir <- here::here("local_temp_data")
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}
```

```{r functions}
scale2 <- function(x, na.rm = TRUE) {
  (x - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
}

center <- function(x, na.rm = TRUE) {
  x - mean(x, na.rm = na.rm)
}

SW <- suppressWarnings

sum_coding <- function(x, lvls = levels(x)) {
  # codes the first category with -1
  nlvls <- length(lvls)
  stopifnot(nlvls > 1)
  cont <- diag(nlvls)[, -nlvls, drop = FALSE]
  cont[nlvls, ] <- -1
  cont <- cont[c(nlvls, 1:(nlvls - 1)), , drop = FALSE]
  colnames(cont) <- lvls[-1]
  x <- factor(x, levels = lvls)
  contrasts(x) <- cont
  x
}

dummy_coding <- function(x, lvls = levels(x)) {
  x <- factor(x, levels = lvls)
  contrasts(x) <- contr.treatment(levels(x))
  x
}
```

## Data preparation

```{r}
data <- read_data_for_analysis()
```

```{r}
data$patient_data <- data$patient_data %>%
  mutate(
    survival = outcome != "Death",
    discharged = outcome == "Discharged"
  )
```

```{r}
data$surv_data <- data$marker_data_wide %>%  
  filter(day > 0) %>%
  mutate(
    daym1 = day - 1,
    status = fct_collapse(breathing, Hospitalized = breathing_levels) %>%
      factor(ordered = FALSE) %>% 
      fct_relevel("Hospitalized", "Discharged", "Death") %>%
      fct_explicit_na("Hospitalized"),
    censored = case_when(
      status == "Death" ~ "none",
      status %in% c("Hospitalized", "Discharged") ~ "right"
    ),
    censored_discharged = case_when(
      status == "Discharged" ~ "none",
      # approximation since patients who died cannot be 
      # discharged in the future
      status %in% c("Hospitalized", "Death") ~ "right"
    )
  ) %>%
  inner_join(data$patient_data, by = c("hospital_id", "patient_id"))

data$surv_data_last_day <- data$surv_data %>%
  group_by(patient_id) %>%
  filter(day == max(day)) %>%
  ungroup()
```

```{r}
hypothesis_res_list <- list()
```

### Notes:

- first and last day invasive: Martin will check
- days_from_symptom_onset. Days from symptom onset until hospitalization. 
Uncertain marker but contains some information. 
Required for anti viral (< 7 days) hypothesis.
- first_day_*: time from day 0 (arrived at hospital) to getting antivirals
- some sites only support patients with mild disease and some only with
strong disease symptoms

## Analysis: Survival

### Basic binary model

```{r, results="hide"}
form_survival_base <- bf(
  survival ~ sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az +
    (1 | hospital_id),
  family = bernoulli("logit")
)

prior_survival_base <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_survival_base <- brm(
  formula = form_survival_base,
  data = data$patient_data,
  prior = prior_survival_base,
  file = "models/fit_survival_base"
)
fit_survival_base <- add_criterion(fit_survival_base, "loo")
```

```{r}
summary(fit_survival_base)
# ce <- conditional_effects(fit_survival_base)
# plot(ce, ask = FALSE)
```

```{r}
draws <- tidy_draws(fit_survival_base)
hypothesis_res_list[["competing_hcq_az_survival_base"]] <- rbind(
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_hcqTRUE,
    model = "brms_binary",
    estimand = "logit(survival)",
    hypothesis = hypotheses$hcq_reduces_death,
    adjusted = "az, sex, age_norm, smoking, BMI"
  ),
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_azTRUE,
    model = "brms_binary",
    estimand = "logit(survival)",
    hypothesis = hypotheses$az_reduces_death,
    adjusted = "hcq, sex, age_norm, smoking, BMI"
  )
)
```

### TODOs

- model varying slopes?
- model interactions?
- use horseshoe prior
- use projpred
- add ever_keltara as soon as there is a lest one patient who took it.

## Analysis: Survival time

```{r, results="hide"}
form_survival_time <- bf(
  day | cens(censored) ~ sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az + 
    (1 | hospital_id),
  family = brmsfamily("cox", "log")
)

prior_survival_time <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_survival_time <- brm(
  formula = form_survival_time,
  data = data$surv_data_last_day,
  prior = prior_survival_time,
  inits = 0,
  file = "models/fit_survival_time"
)
fit_survival_time <- add_criterion(fit_survival_time, "loo")
```

```{r}
summary(fit_survival_time)
```

```{r}
draws <- tidy_draws(fit_survival_time)
hypothesis_res_list[["competing_hcq_az_survival_time"]] <- rbind(
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_hcqTRUE,
    model = "brms_cox",
    estimand = "log(HR)",
    hypothesis = hypotheses$hcq_reduces_death,
    adjusted = "az, sex, age_norm, smoking, BMI"
  ),
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_azTRUE,
    model = "brms_cox",
    estimand = "log(HR)",
    hypothesis = hypotheses$az_reduces_death,
    adjusted = "hcq, sex, age_norm, smoking, BMI"
  )
)
```

### TODOs:

- How does that relate to days of hospitalization? (Answer: could be analysed;
could be an interesting end point when people are discharged)

## Analysis: Discharged

### Basic binary model

```{r, results="hide"}
form_discharged_base <- bf(
  discharged ~ sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az +
    (1 | hospital_id),
  family = bernoulli("logit")
)

prior_discharged_base <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_discharged_base <- brm(
  formula = form_discharged_base,
  data = data$patient_data,
  prior = prior_discharged_base,
  file = "models/fit_discharged_base"
)
fit_discharged_base <- add_criterion(fit_discharged_base, "loo")
```

```{r}
summary(fit_discharged_base)
# ce <- conditional_effects(fit_discharged_base)
# plot(ce, ask = FALSE)
```

```{r}
draws <- tidy_draws(fit_discharged_base)
hypothesis_res_list[["competing_hcq_az_discharged_base"]] <- rbind(
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_hcqTRUE,
    model = "brms_binary",
    estimand = "logit(discharged)",
    hypothesis = hypotheses$hcq_increases_discharged,
    adjusted = "az, sex, age_norm, smoking, BMI"
  ),
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_azTRUE,
    model = "brms_binary",
    estimand = "logit(discharged)",
    hypothesis = hypotheses$az_increases_discharged,
    adjusted = "hcq, sex, age_norm, smoking, BMI"
  )
)
```

### TODOs

- model varying slopes?
- model interactions?
- use horseshoe prior
- use projpred
- add ever_keltara as soon as there is a lest one patient who took it.

## Analysis: discharged time

```{r, results="hide"}
form_discharged_time <- bf(
  day | cens(censored_discharged) ~ sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az + 
    (1 | hospital_id),
  family = brmsfamily("cox", "log")
)

prior_discharged_time <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_discharged_time <- brm(
  formula = form_discharged_time,
  data = data$surv_data_last_day,
  prior = prior_discharged_time,
  inits = 0,
  control = list(adapt_delta = 0.95),
  file = "models/fit_discharged_time"
)
fit_discharged_time <- add_criterion(fit_discharged_time, "loo")
```

```{r}
summary(fit_discharged_time)
```

```{r}
draws <- tidy_draws(fit_discharged_time)
hypothesis_res_list[["competing_hcq_az_discharged_time"]] <- rbind(
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_hcqTRUE,
    model = "brms_cox",
    estimand = "log(HR)",
    hypothesis = hypotheses$hcq_increases_discharged,
    adjusted = "az, sex, age_norm, smoking, BMI"
  ),
  bayesian_hypothesis_res_from_draws(
    draws = draws$b_ever_azTRUE,
    model = "brms_cox",
    estimand = "log(HR)",
    hypothesis = hypotheses$az_increases_discharged,
    adjusted = "hcq, sex, age_norm, smoking, BMI"
  )
)
```

## Analysis: Worst Breathing {.tabset}

### Basic ordinal model

```{r, results="hide"}
form_worst_breathing_base <- bf(
  worst_breathing ~ sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az +
    (1 | hospital_id),
  family = sratio("logit")
)

prior_worst_breathing_base <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_worst_breathing_base <- brm(
  formula = form_worst_breathing_base,
  data = data$patient_data,
  prior = prior_worst_breathing_base,
  file = "models/fit_worst_breathing_base"
)
fit_worst_breathing_base <- add_criterion(fit_worst_breathing_base, "loo")
```

```{r}
summary(fit_worst_breathing_base)
# ce <- conditional_effects(fit_worst_breathing_base, categorical = TRUE)
# plot(ce, ask = FALSE)
```

### Category-specific ordinal model

```{r, results="hide"}
form_worst_breathing_cs <- bf(
  worst_breathing ~ cs(sex + age_norm + 
    smoking + BMI + ever_hcq + ever_az) +
    (1 | hospital_id),
  family = sratio("logit")
)

prior_worst_breathing_cs <- 
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 5), class = "sd")

fit_worst_breathing_cs <- brm(
  formula = form_worst_breathing_cs,
  data = data$patient_data,
  prior = prior_worst_breathing_cs,
  inits = 0,
  file = "models/fit_worst_breathing_cs"
)
fit_worst_breathing_cs <- add_criterion(fit_worst_breathing_cs, "loo")
```

```{r}
summary(fit_worst_breathing_cs)
# ce <- conditional_effects(fit_worst_breathing_cs, categorical = TRUE)
# plot(ce, ask = FALSE)
```

### Model Comparison

```{r}
loo(fit_worst_breathing_base, fit_worst_breathing_cs)
```

### TODOs

- How to handle missing values in `worst_breathing`?
- decide which predictors to include 
- model interactions?
- model varying slopes?
- use horseshoe or similar sparsity priors?

```{r}
hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
write_csv(hypothesis_res_all, path = paste0(cache_dir, "/analysis_brms.csv"))
```

### 2020-10-04 manuscript/analysis_brms.Rmd
add 3-outcome categorical analysis

@@
     survival = outcome != "Death",
-    discharged = outcome == "Discharged"
+    discharged = outcome == "Discharged",
+    outcome3 = fct_collapse(outcome, Discharged = c("Discharged", "Transferred")) %>%
+      fct_relevel("Hospitalized", "Discharged", "Death")
   )
@@
 draws <- tidy_draws(fit_survival_base)
-hypothesis_res_list[["competing_hcq_az_survival_base"]] <- rbind(
+hypothesis_res_list[["hcq_az_survival_base"]] <- rbind(
   bayesian_hypothesis_res_from_draws(
@@
 draws <- tidy_draws(fit_survival_time)
-hypothesis_res_list[["competing_hcq_az_survival_time"]] <- rbind(
+hypothesis_res_list[["hcq_az_survival_time"]] <- rbind(
   bayesian_hypothesis_res_from_draws(
@@
 draws <- tidy_draws(fit_discharged_base)
-hypothesis_res_list[["competing_hcq_az_discharged_base"]] <- rbind(
+hypothesis_res_list[["hcq_az_discharged_base"]] <- rbind(
   bayesian_hypothesis_res_from_draws(
@@
 draws <- tidy_draws(fit_discharged_time)
-hypothesis_res_list[["competing_hcq_az_discharged_time"]] <- rbind(
+hypothesis_res_list[["hcq_az_discharged_time"]] <- rbind(
   bayesian_hypothesis_res_from_draws(
@@
     estimand = "log(HR)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+## Analysis: Hospitalized vs. Discharged vs. Death
+
+### Basic Categorical model
+
+```{r, results="hide"}
+form_outcome_base <- bf(
+  outcome3 ~ sex + age_norm + 
+    smoking + BMI + ever_hcq + ever_az +
+    (1 | hospital_id),
+  family = categorical("logit")
+)
+
+prior_outcome_base <- 
+  prior(normal(0, 5), class = "b", dpar = "muDeath") +
+  prior(normal(0, 5), class = "sd", dpar = "muDeath") +
+  prior(normal(0, 5), class = "b", dpar = "muDischarged") +
+  prior(normal(0, 5), class = "sd", dpar = "muDischarged")
+
+fit_outcome_base <- brm(
+  formula = form_outcome_base,
+  data = data$patient_data,
+  prior = prior_outcome_base,
+  file = "models/fit_outcome_base"
+)
+fit_outcome_base <- add_criterion(fit_outcome_base, "loo")
+```
+
+```{r}
+summary(fit_outcome_base)
+# ce <- conditional_effects(fit_outcome_base)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_outcome_base)
+hypothesis_res_list[["hcq_az_outcome_base"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_hcqTRUE,
+    model = "brms_categorical",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_azTRUE,
+    model = "brms_categorical",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_hcqTRUE,
+    model = "brms_categorical",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_azTRUE,
+    model = "brms_categorical",
+    estimand = "logit(discharged)",
     hypothesis = hypotheses$az_increases_discharged,

### 2020-10-09 manuscript/analysis_brms.Rmd
run more sensitivity analysis

@@
 ```{r}
+# all patients for which we have data for at least 7 days
+data$surv_data_7 <- data$surv_data_last_day %>%
+  filter(!(day < 7 & outcome3 == "Hospitalized"))
+
+# all patients for which we have data for at least 7 days
+data$surv_data_28 <- data$surv_data_last_day %>%
+  filter(!(day < 28 & outcome3 == "Hospitalized"))
+```
+
+```{r}
 hypothesis_res_list <- list()
@@
     model = "brms_binary",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+### Basic binary model (7 days)
+
+```{r, results="hide"}
+fit_survival_base_7 <- brm(
+  formula = form_survival_base,
+  data = data$surv_data_7,
+  prior = prior_survival_base,
+  file = "models/fit_survival_base_7"
+)
+fit_survival_base_7 <- add_criterion(fit_survival_base_7, "loo")
+```
+
+```{r}
+summary(fit_survival_base_7)
+# ce <- conditional_effects(fit_survival_base_7)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_survival_base_7)
+hypothesis_res_list[["hcq_az_survival_base_7"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+### Basic binary model (28 days)
+
+```{r, results="hide"}
+fit_survival_base_28 <- brm(
+  formula = form_survival_base,
+  data = data$surv_data_28,
+  prior = prior_survival_base,
+  file = "models/fit_survival_base_28"
+)
+fit_survival_base_28 <- add_criterion(fit_survival_base_28, "loo")
+```
+
+```{r}
+summary(fit_survival_base_28)
+# ce <- conditional_effects(fit_survival_base_28)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_survival_base_28)
+hypothesis_res_list[["hcq_az_survival_base_28"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_binary_28",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_binary_28",
     estimand = "logit(survival)",
@@
     model = "brms_binary",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+### Basic binary model (7 Days)
+
+```{r, results="hide"}
+fit_discharged_base_7 <- brm(
+  formula = form_discharged_base,
+  data = data$surv_data_7,
+  prior = prior_discharged_base,
+  file = "models/fit_discharged_base_7"
+)
+fit_discharged_base_7 <- add_criterion(fit_discharged_base_7, "loo")
+```
+
+```{r}
+summary(fit_discharged_base_7)
+# ce <- conditional_effects(fit_discharged_base_7)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_discharged_base_7)
+hypothesis_res_list[["hcq_az_discharged_base_7"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+### Basic binary model (28 Days)
+
+```{r, results="hide"}
+fit_discharged_base_28 <- brm(
+  formula = form_discharged_base,
+  data = data$surv_data_28,
+  prior = prior_discharged_base,
+  file = "models/fit_discharged_base_28"
+)
+fit_discharged_base_28 <- add_criterion(fit_discharged_base_28, "loo")
+```
+
+```{r}
+summary(fit_discharged_base_28)
+# ce <- conditional_effects(fit_discharged_base_28)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_discharged_base_28)
+hypothesis_res_list[["hcq_az_discharged_base_28"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_binary_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_binary_28",
     estimand = "logit(discharged)",
@@
 
+### Basic Categorical model (7 Days)
+
+```{r, results="hide"}
+fit_outcome_base_7 <- brm(
+  formula = form_outcome_base,
+  data = data$surv_data_7,
+  prior = prior_outcome_base,
+  file = "models/fit_outcome_base_7"
+)
+fit_outcome_base_7 <- add_criterion(fit_outcome_base_7, "loo")
+```
+
+```{r}
+summary(fit_outcome_base_7)
+# ce <- conditional_effects(fit_outcome_base_7)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_outcome_base_7)
+hypothesis_res_list[["hcq_az_outcome_base_7"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_hcqTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_azTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_hcqTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_azTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
+### Basic Categorical model (28 Days)
+
+```{r, results="hide"}
+fit_outcome_base_28 <- brm(
+  formula = form_outcome_base,
+  data = data$surv_data_28,
+  prior = prior_outcome_base,
+  file = "models/fit_outcome_base_28"
+)
+fit_outcome_bas_28 <- add_criterion(fit_outcome_base_28, "loo")
+```
+
+```{r}
+summary(fit_outcome_base_28)
+# ce <- conditional_effects(fit_outcome_base_28)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_outcome_base_28)
+hypothesis_res_list[["hcq_az_outcome_base_28"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_hcqTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_azTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_hcqTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "az, sex, age_norm, smoking, BMI"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_azTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "hcq, sex, age_norm, smoking, BMI"
+  )
+)
+```
+
 ## Analysis: Worst Breathing {.tabset}

### 2020-10-09 manuscript/risk_scores.Rmd
Further work on risk scores

@@
 ```{r}
-baseline <- data$patient_data %>%
-  mutate(alive_12_admission = case_when(
+at_admission <- data$patient_data %>%
+  mutate(alive_12 = case_when(
     last_record >= 12 ~ TRUE,
@@
     TRUE ~ NA),
-    alive_30_admission = case_when(
+    alive_30 = case_when(
     last_record >= 30 ~ TRUE,
@@
 ```
+
 ```{r}
@@
 ```{r}
-outcome_caption = c("alive_12" = "12-day mortality", "alive_30" = "30-day mortality")
-score_caption = c("ACP_grade" = "Li et al. (ACP)")
-
-my_auc <- function(f, d) {
-  roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
-  ci = pROC::ci.auc(roc)
-
-  outcome = as.character(formula.tools::lhs(f))
-  score = as.character(formula.tools::rhs(f))
-
-  list(roc = roc,
-       summary = data.frame(
-         auc = ci[2],
-         auc_low = ci[1],
-         auc_high = ci[3],
-         outcome = outcome,
-         score = score
-       )
-  )
-}
-
-plot_my_auc <- function(my_auc) {
-  caption = paste0(score_caption[my_auc$summary$score], ", ", outcome_caption[my_auc$summary$outcome])
-  plot(my_auc$roc, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE, main = caption)
-}
-
-plot_all_auc <- function(my_auc_list) {
-  my_auc_list %>% walk(plot_my_auc)
-}
-```
-
-```{r}
 acp_auc <- list(
@@
 
+```
+
+```{r}
+compare_auc_to_orig(list(acp_auc_orig), list(acp_auc))
+#acp_auc
 ```
@@
     inclusion_criterium = breathing_first < "NIPPV" & (outcome %in% c("Discharged", "Death") | last_record >= first_day + 7),
-    score =  -4.21167 + 0.058699 * age + 0.013391 * first_CRP + 0.112055 * first_d_dimer_converted - 1.983733 * first_lymphocyte_count,
-    risk = 1 / (1 + exp(-score)),
-    risk_lower = 1 / (1 + exp(-score + 1.96*0.1721)),
-    risk_upper = 1 / (1 + exp(-score - 1.96*0.1721)),
+    chen_liu = -4.21167 + 0.058699 * age + 0.013391 * first_CRP + 0.112055 * first_d_dimer_converted - 1.983733 * first_lymphocyte_count,
+    risk = 1 / (1 + exp(-chen_liu)),
+    risk_lower = 1 / (1 + exp(-chen_liu + 1.96*0.1721)),
+    risk_upper = 1 / (1 + exp(-chen_liu - 1.96*0.1721)),
     alive_12 = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
@@
 
-  ) %>% select(age, starts_with("first"), score, starts_with("risk"), breathing_first, starts_with("alive"), inclusion_criterium)
+  ) %>% select(age, starts_with("first"), chen_liu, starts_with("risk"), breathing_first, starts_with("alive"), inclusion_criterium)
 ```
@@
 ```{r}
-chen_liu_data %>% ggplot(aes(x = score, y = alive_30)) + geom_point()
+chen_liu_data %>% ggplot(aes(x = chen_liu, y = alive_30)) + geom_point()
 ```
@@
 ```{r}
-roc_chen_liu_12 <- pROC::roc(alive_12 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_chen_liu_30 <- pROC::roc(alive_30 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_chen_liu_all <- pROC::roc(alive_all ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+chen_liu_data_inclusion <- chen_liu_data %>% filter(inclusion_criterium)
+chen_liu_auc <- list(
+  my_auc(alive_12 ~ chen_liu, chen_liu_data),
+  my_auc(alive_30 ~ chen_liu, chen_liu_data),
+  my_auc(alive_all ~ chen_liu, chen_liu_data),
+  my_auc(alive_imputed ~ chen_liu, chen_liu_data),
+  my_auc(alive_12 ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc(alive_30 ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc(alive_imputed ~ chen_liu, chen_liu_data_inclusion, "incl")
+)
 
-roc_chen_liu_12_incl <- pROC::roc(alive_12 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_chen_liu_30_incl <- pROC::roc(alive_30 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_chen_liu_all_incl <- pROC::roc(alive_all ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_chen_liu_imputed_incl <- pROC::roc(alive_imputed ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+plot_all_auc(chen_liu_auc)
+
+```
+
+```{r}
+chen_liu_auc_orig <- my_auc_from_estimate(0.881, "alive", "chen_liu", "incl")
+compare_auc_to_orig(list(chen_liu_auc_orig), list(chen_liu_auc))
 
@@
 
-Not sure what "severe" means
+Not sure what "severe" means, will test both "not ventilated", "not needing Oxygen"
+
+```{r}
+shi_et_al_data <- data$marker_data_wide %>%  
+  group_by(patient_id) %>%
+  summarise(
+    breathing_admission = max(breathing[!is.na(breathing) & day < 2]),
+    not_ventilated_admission = breathing_admission < "NIPPV",
+    not_oxygen_admission = breathing_admission < "Oxygen",
+    .groups = "drop") %>% 
+  inner_join(at_admission, by = c("patient_id")) %>%
+  mutate(shi = (age >= 50) + (sex == "M") + has_hypertension_drugs,
+         not_ventilated_all = case_when(
+           outcome == "Death" ~ FALSE,
+           outcome != "Discharged" ~ NA,
+           worst_breathing < "NIPPV" ~ TRUE
+         ),
+         not_oxygen_all = case_when(
+           outcome == "Death" ~ FALSE,
+           outcome != "Discharged" ~ NA,
+           worst_breathing < "Oxygen" ~ TRUE
+         )
+         )
+
+shi_et_al_auc <- list(
+  my_auc(alive_12 ~ shi, shi_et_al_data),
+  my_auc(alive_30 ~ shi, shi_et_al_data),
+  my_auc(not_ventilated_all ~ shi, shi_et_al_data),
+  my_auc(not_oxygen_all ~ shi, shi_et_al_data),
+  my_auc(not_ventilated_admission ~ shi, shi_et_al_data),
+  my_auc(not_oxygen_admission ~ shi, shi_et_al_data)
+)
+
+plot_all_auc(shi_et_al_auc)
+```
+
+From Fig1 B,C:
+```{r}
+shi_et_al_reported <- data.frame(
+  score = c(0,1,2,3), shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429)
+)
+
+shi_et_al_compare <- shi_et_al_data %>%
+  group_by(shi) %>%
+  summarise(
+    across(
+      one_of(c("not_ventilated_admission", "not_ventilated_all", "not_oxygen_admission", "not_oxygen_all")), 
+      ~ mean(.x, na.rm = TRUE), .names = "our_{col}"), 
+    n_our_admission = sum(!is.na(not_ventilated_admission)), 
+    n_our_all = sum(!is.na(not_ventilated_all)),
+    .groups = "drop") %>%
+  inner_join(shi_et_al_reported, by = c("shi" = "score"))
+
+shi_et_al_compare %>%
+  transmute(score = shi, `Shi et al.
+(admission)` = shi_severe_admission, `Ventilated (admission)` = 1 - our_not_ventilated_admission, `Oxygen (admission)` = 1 - our_not_oxygen_admission, N = n_our_admission)
+
+shi_et_al_compare %>%
+  transmute(score = shi, `Shi et al.
+(hospitalization)` = shi_severe_all, `Ventilated (hospitalization)` = 1 - our_not_ventilated_all, `Oxygen (hospitalization)` = 1 - our_not_oxygen_all, N = n_our_all)
+
+```
 
@@
 
-bello_chavolla_data <- baseline %>%
+bello_chavolla_data <- at_admission %>%
   inner_join(atb_first_day, by = "patient_id") %>%

### 2020-10-09 R/auc_tools.R

@@ -1,0 +1,104 @@
+outcome_caption = c("alive_12" = "12-day mortality", "alive_30" = "30-day mortality",
+                    "alive_all" = "Non-censored mortality",
+                    "alive_imputed" = "Alive at data collection",
+                    "not_ventilated_all" = "Not ventilated (uncensored only)",
+                    "not_oxygen_all" = "Not on oxygen (uncensored only)",
+                    "not_ventilated_admission" = "Not ventilated on admission",
+                    "not_oxygen_admission" = "Not on oxygen on admission")
+score_caption = c("ACP_grade" = "Li et al. (ACP)",
+                  "chen_liu" = "Chen & Liu",
+                  "chen_liu_incl" = "Chen & Liu",
+                  "shi" = "Shi et al.",
+                  "caramelo" = "Caramelo et al.",
+                  "bello_chavolla", "Bello-Chavolla et al.")
+
+outcome_note_caption = c("incl" = " (inclusion crit.)")
+
+my_auc <- function(f, d, outcome_note = "") {
+  roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
+  ci = pROC::ci.auc(roc)
+
+  outcome = as.character(formula.tools::lhs(f))
+  score = as.character(formula.tools::rhs(f))
+
+
+  list(roc = roc,
+       summary = tibble::tibble(
+         auc = ci[2],
+         auc_low = ci[1],
+         auc_high = ci[3],
+         outcome = outcome,
+         score = score,
+         outcome_note = outcome_note
+       )
+  )
+}
+
+plot_my_auc <- function(auc_obj) {
+  if(!(auc_obj$summary$score %in% names(score_caption))) {
+    stop("Score does not have caption")
+  }
+  if(!(auc_obj$summary$outcome %in% names(outcome_caption))) {
+    stop("Outcome does not have caption")
+  }
+
+  if(auc_obj$summary$outcome_note != "") {
+    if(!(auc_obj$summary$outcome_note %in% names(outcome_note_caption))) {
+      stop("Outcome note does not have caption")
+    }
+    outcome_note_caption_value <- outcome_note_caption[auc_obj$summary$outcome_note]
+  } else {
+    outcome_note_caption_value <- ""
+  }
+
+
+  caption = paste0(score_caption[auc_obj$summary$score], ", ", outcome_caption[auc_obj$summary$outcome],
[…]

### 2020-10-10 manuscript/risk_scores.Rmd
Original risk score for Shi et al, tweaks to plots

@@
 ```{r}
-li_et_al_data <- data.frame(
-  ACP_grade = c(rep(1, 194), rep(2, 145), rep(3, 95)),
-  alive_12 = c(rep(TRUE, 194), #Grade 1, 0% mortality
-               rep(FALSE, 8), rep(TRUE, 145-8), #Gred 2, 5.6% mortality
-               rep(FALSE, 32), rep(TRUE, 95 - 32) #Grade 3, 33.2% mortality
-  ))
+auc_data_from_counts <- function(score_values, positive_counts, negative_counts) {
+  res <- list()
+  for(i in 1:length(score_values)) {
+    res[[i]] <- rbind(
+      data.frame(score = rep(score_values[i], positive_counts[i]), outcome = rep(TRUE, positive_counts[i])),
+      data.frame(score = rep(score_values[i], negative_counts[i]), outcome = rep(FALSE, negative_counts[i]))
+    )
+  }
+  do.call(rbind, res)
+}
+
+#Grade 1, 0% mortality,Grade 2, 5.6% mortality, Grade 3, 33.2% mortality
+li_et_al_data <-  auc_data_from_counts(c(1,2,3), 
+                                            positive_counts = c(194, 145-8, 95-32), 
+                                            negative_counts = c(0, 8, 32)) %>%
+                         transmute(alive_12 = outcome, ACP_grade = score)
 
@@
 acp_auc_orig$summary
+
 ```
@@
 
-plot_all_auc(chen_liu_auc)
+#plot_all_auc(chen_liu_auc)
 
@@
   my_auc(not_oxygen_all ~ shi, shi_et_al_data),
-  my_auc(not_ventilated_admission ~ shi, shi_et_al_data),
-  my_auc(not_oxygen_admission ~ shi, shi_et_al_data)
+  my_auc(not_ventilated_admission ~ shi, shi_et_al_data, subgroup = " (at admission)"),
+  my_auc(not_oxygen_admission ~ shi, shi_et_al_data, subgroup = " (at admission)")
 )
 
-plot_all_auc(shi_et_al_auc)
+#plot_all_auc(shi_et_al_auc)
 ```
@@
 shi_et_al_reported <- data.frame(
-  score = c(0,1,2,3), shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429)
+  score = c(0,1,2,3), shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429),
+  mild_cases_admission = c(118, 199, 98, 21), all_cases_admission = c(118, 211, 121, 35),
+  mild_cases_hospitalization = c(11, 25, 11, 4), all_cases_hospitalization = c(12, 29, 18, 7)
 )
+
+# Check that the proportions match
+# shi_et_al_reported %>% mutate(sev2 = (all_cases_admission - mild_cases_admission)/all_cases_admission) %>%
+#   select(score, shi_severe_admission,sev2)
+# 
+# shi_et_al_reported %>% mutate(sev2 = (all_cases_hospitalization - mild_cases_hospitalization)/all_cases_hospitalization) %>%
+#   select(score, shi_severe_all,sev2)
+shie_et_al_auc_data_admission <- auc_data_from_counts(score_values = shi_et_al_reported$score,
+                                                      positive_counts = shi_et_al_reported$mild_cases_admission,
+                                                      negative_counts = shi_et_al_reported$all_cases_admission - shi_et_al_reported$mild_cases_admission) %>% 
+  transmute(shi = score, severe_admission = outcome)
+
+shi_et_al_auc_reported_admission <- my_auc(severe_admission ~ shi, shie_et_al_auc_data_admission, subgroup = " (at admission)")
+#shi_et_al_auc_reported_admission$summary
+
+shie_et_al_auc_data_all <- auc_data_from_counts(score_values = shi_et_al_reported$score,
+                                                      positive_counts = shi_et_al_reported$mild_cases_hospitalization,
+                                                      negative_counts = shi_et_al_reported$all_cases_hospitalization - shi_et_al_reported$mild_cases_hospitalization) %>% 
+  transmute(shi = score, severe_hospitalized = outcome)
+
+shi_et_al_auc_reported_all <- my_auc(severe_hospitalized ~ shi, shie_et_al_auc_data_all)
+
+compare_auc_to_orig(list(shi_et_al_auc_reported_admission, shi_et_al_auc_reported_all), list(shi_et_al_auc), ncol = 1)
+```
+
+```{r}
 

### 2020-10-10 R/auc_tools.R

@@ -14,7 +14,7 @@ score_caption = c("ACP_grade" = "Li et al. (ACP)",
 
 outcome_note_caption = c("incl" = " (inclusion crit.)")
 
-my_auc <- function(f, d, outcome_note = "") {
+my_auc <- function(f, d, outcome_note = "", subgroup = "") {
   roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
   ci = pROC::ci.auc(roc)
 
@@ -29,7 +29,8 @@ my_auc <- function(f, d, outcome_note = "") {
          auc_high = ci[3],
          outcome = outcome,
          score = score,
-         outcome_note = outcome_note
+         outcome_note = outcome_note,
+         subgroup = subgroup
        )
   )
 }
@@ -61,7 +62,7 @@ plot_all_auc <- function(my_auc_list) {
   my_auc_list %>% walk(plot_my_auc)
 }
 
-my_auc_from_estimate <- function(auc_estimate, outcome, score, outcome_note) {
+my_auc_from_estimate <- function(auc_estimate, outcome, score, outcome_note, subgroup = "") {
   list(roc = NULL,
        summary = tibble::tibble(
          auc = auc_estimate,
@@ -69,32 +70,37 @@ my_auc_from_estimate <- function(auc_estimate, outcome, score, outcome_note) {
          auc_high = NA_real_,
          outcome = outcome,
          score = score,
-         outcome_note = outcome_note
+         outcome_note = outcome_note,
+         subgroup = subgroup
        )
   )
 
 }
 
-compare_auc_to_orig <- function(orig_auc_list, our_auc_list, show_outcome = TRUE) {
+compare_auc_to_orig <- function(orig_auc_list, our_auc_list, show_outcome = TRUE, ...) {
   get_summaries <- function(x) {
     map_df(x, ~ .x$summary)
   }
+
+  source_factor <- function(x) {
+    factor(x, levels = c("our","orig"), labels = c("Present study", "Original study"))
+  }
   all_auc <- our_auc_list %>% map_df(get_summaries) %>%
-    mutate(source = "our", score = score_caption[score])
-  all_orig <- orig_auc_list %>% get_summaries() %>% mutate(source = "orig", score = score_caption[score])
+    mutate(source = source_factor("our"), score = paste(score_caption[score], subgroup))
[…]

### 2020-10-11 manuscript/risk_scores.Rmd
Risk scores bascially complete (some polish remaining)

@@
 
+In all of the following we focus on the C-statistic, also often called area under curve (AUC). We compute AUC using the  `pROC` package, which also reports bootstrapped confidence intervals.
+
 # ACP
@@
 years and CRP $<$ 34 mg/L or age $<$ 60 years and CRP $\geq$ 34 mg/L), grade 3 (age $\geq$ 60 years and CRP $\geq$ 34 mg/L) and was evaluated primarily on 12-day mortality.
-Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).
+Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in rough agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).
 
@@
 
+The original authors do not report AUC (C-statistic), but they report the overall counts of alive and death patients, letting us compute it. The counts are:
+
 ```{r}
@@
 
-acp_auc_orig <- my_auc(alive_12 ~ ACP_grade, li_et_al_data)
-acp_auc_orig$summary
+li_et_al_data
+
+```
+Giving us this AUC:
+
+```{r}
+acp_auc_orig <- my_auc_summary(alive_12 ~ ACP_grade, li_et_al_data)
+acp_auc_orig %>% transmute(`C-static` = auc, `95% CI` = paste0("(",auc_low, ", ",auc_high, ")"))
+
+```
+
+So we can now compare AUC computed on our data for both 12-day mortality (as in the original study) and 30-day mortality.
+
+```{r, fig.cap="The AUC and it's 95% credible intervals for the ACP score.
+The thick blue vertical line marks AUC of 0.5, i.e.
+the point where reverting the score would yield better predictions.
+The thin gray dashed line marks the AUC from the original study for easier comparison."}
+acp_auc <- rbind(
+  my_auc_summary(alive_12 ~ ACP_grade, acp_data),
+  my_auc_summary(alive_30 ~ ACP_grade, acp_data)
+)
+compare_auc_to_orig(acp_auc_orig, acp_auc)
 
@@
 ```{r}
-acp_auc <- list(
-  alive_12 = my_auc(alive_12 ~ ACP_grade, acp_data),
-  alive_30 = my_auc(alive_30 ~ ACP_grade, acp_data)
-)
-plot_all_auc(acp_auc)
-
-```
-
-```{r}
-compare_auc_to_orig(list(acp_auc_orig), list(acp_auc))
 #acp_auc
@@
 chen_liu_data_inclusion <- chen_liu_data %>% filter(inclusion_criterium)
-chen_liu_auc <- list(
-  my_auc(alive_12 ~ chen_liu, chen_liu_data),
-  my_auc(alive_30 ~ chen_liu, chen_liu_data),
-  my_auc(alive_all ~ chen_liu, chen_liu_data),
-  my_auc(alive_imputed ~ chen_liu, chen_liu_data),
-  my_auc(alive_12 ~ chen_liu, chen_liu_data_inclusion, "incl"),
-  my_auc(alive_30 ~ chen_liu, chen_liu_data_inclusion, "incl"),
-  my_auc(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl"),
-  my_auc(alive_imputed ~ chen_liu, chen_liu_data_inclusion, "incl")
+chen_liu_auc <- rbind(
+  my_auc_summary(alive_12 ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_30 ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_all ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_imputed ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_12 ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc_summary(alive_30 ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc_summary(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc_summary(alive_imputed ~ chen_liu, chen_liu_data_inclusion, "incl")
 )
@@
 ```{r}
-chen_liu_auc_orig <- my_auc_from_estimate(0.881, "alive", "chen_liu", "incl")
-compare_auc_to_orig(list(chen_liu_auc_orig), list(chen_liu_auc))
+chen_liu_auc_orig <- my_auc_summary_from_estimate(0.881, "alive", "chen_liu", "incl")
+compare_auc_to_orig(chen_liu_auc_orig, chen_liu_auc)
 
@@
 
-shi_et_al_auc <- list(
-  my_auc(alive_12 ~ shi, shi_et_al_data),
-  my_auc(alive_30 ~ shi, shi_et_al_data),
-  my_auc(not_ventilated_all ~ shi, shi_et_al_data),
-  my_auc(not_oxygen_all ~ shi, shi_et_al_data),
-  my_auc(not_ventilated_admission ~ shi, shi_et_al_data, subgroup = " (at admission)"),
-  my_auc(not_oxygen_admission ~ shi, shi_et_al_data, subgroup = " (at admission)")
+shi_et_al_auc <- rbind(
+  my_auc_summary(alive_12 ~ shi, shi_et_al_data),
+  my_auc_summary(alive_30 ~ shi, shi_et_al_data),
+  my_auc_summary(not_ventilated_all ~ shi, shi_et_al_data),
+  my_auc_summary(not_oxygen_all ~ shi, shi_et_al_data),
+  my_auc_summary(not_ventilated_admission ~ shi, shi_et_al_data, subgroup = " (at admission)"),
+  my_auc_summary(not_oxygen_admission ~ shi, shi_et_al_data, subgroup = " (at admission)")
 )
@@
 
-shi_et_al_auc_reported_admission <- my_auc(severe_admission ~ shi, shie_et_al_auc_data_admission, subgroup = " (at admission)")
+shi_et_al_auc_reported_admission <- my_auc_summary(severe_admission ~ shi, shie_et_al_auc_data_admission, subgroup = " (at admission)")
 #shi_et_al_auc_reported_admission$summary
@@
 
-shi_et_al_auc_reported_all <- my_auc(severe_hospitalized ~ shi, shie_et_al_auc_data_all)
+shi_et_al_auc_reported_all <- my_auc_summary(severe_hospitalized ~ shi, shie_et_al_auc_data_all)
 
-compare_auc_to_orig(list(shi_et_al_auc_reported_admission, shi_et_al_auc_reported_all), list(shi_et_al_auc), ncol = 1)
+compare_auc_to_orig(rbind(shi_et_al_auc_reported_admission, shi_et_al_auc_reported_all), shi_et_al_auc, ncol = 1)
 ```
@@
 
-???
+Uses simulations to get patient-level characterstics from aggregate data.
+
+```{r}
+caramelo_data <- at_admission %>%
+  mutate(
+    caramelo_base = 0 + 
+           case_when(age < 20 ~ 0,
+                     age < 30 ~ 0.2017,
+                     age < 40 ~ 0.3271,
+                     age < 50 ~ 5.6030,
+                     age < 60 ~ 6.7626,
+                     age < 70 ~ 18.8161,
+                     age < 80 ~ 43.7291,
+                     TRUE ~ 86.8680) +
+           if_else(sex == "M", 1.8518, 0) +
+           if_else(has_hypertension_drugs, 7.4191, 0) +
+           if_else(diabetes, 9.0329, 0),
+    caramelo1 = caramelo_base +            
+           if_else(heart_failure | ischemic_heart_disease | NYHA > 1, 12.8328, 0) +
+           if_else(COPD | asthma, 7.7925, 0),
+    caramelo2 = caramelo_base +            
+           if_else(heart_failure | ischemic_heart_disease, 12.8328, 0) +
+           if_else(COPD, 7.7925, 0)
+         )
+
+caramelo_auc <- rbind(
+  my_auc_summary(alive_12 ~ caramelo1, caramelo_data),
+  my_auc_summary(alive_30 ~ caramelo1, caramelo_data),
+  my_auc_summary(alive_12 ~ caramelo2, caramelo_data, outcome_note = "v2"),
+  my_auc_summary(alive_30 ~ caramelo2, caramelo_data, outcome_note = "v2"),
+  my_auc_summary(alive_12 ~ caramelo_base, caramelo_data, outcome_note = "base"),
+  my_auc_summary(alive_30 ~ caramelo_base, caramelo_data, outcome_note = "base")
+)
+
+compare_auc_to_orig(tibble(), caramelo_auc)
+```
 
@@
 ```{r}
-bello_chavolla_data %>% ggplot(aes(x = bc0, y = as.integer(alive_30_admission))) +
+bello_chavolla_data %>% ggplot(aes(x = bc0, y = as.integer(alive_30))) +
   geom_smooth() + geom_jitter(width = 0.3, height = 0.3) 
-bello_chavolla_data %>% ggplot(aes(x = bc_base, y = as.integer(alive_30_admission))) +
+bello_chavolla_data %>% ggplot(aes(x = bc_base, y = as.integer(alive_30))) +
   geom_smooth() + geom_jitter(width = 0.3, height = 0.3)
@@
 ```{r}
-roc_bc_base_30 <- pROC::roc(alive_30_admission ~ bc_base, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_bc_base_12 <- pROC::roc(alive_12_admission ~ bc_base, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_bc0_30 <- pROC::roc(alive_30_admission ~ bc0, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_bc0_12 <- pROC::roc(alive_12_admission ~ bc0, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_bc1_30 <- pROC::roc(alive_30_admission ~ bc1, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
-roc_bc1_12 <- pROC::roc(alive_12_admission ~ bc1, bello_chavolla_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
+bc_auc <- rbind(
+  my_auc_summary(alive_30 ~ bc_base, bello_chavolla_data, outcome_note = "base"),
+  my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, outcome_note = "base"),
+  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data),
+  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data),
+  my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, outcome_note = "day1"),
+  my_auc_summary(alive_12 ~ bc1, bello_chavolla_data, outcome_note = "day1")
+)
+
+bc_auc_orig <- my_auc_summary_from_estimate(0.830, "alive_30", "bc0")
+
+compare_auc_to_orig(bc_auc_orig, bc_auc)
+```
+# Age only
+
+As a simple baseline we use age in years as the sole predictor.
+
+```{r}
+age_auc <-rbind(
+  my_auc_summary(alive_30 ~ age, at_admission),
+  my_auc_summary(alive_12 ~ age, at_admission)
+)
+
+plot_my_auc(alive_30 ~ age, at_admission)
+plot_my_auc(alive_12 ~ age, at_admission)
 
 ```
+
+# All together now
+
+```{r}
+compare_auc_to_orig(rbind(acp_auc_orig, chen_liu_auc_orig, shi_et_al_auc_reported_admission,
+                         shi_et_al_auc_reported_all, bc_auc_orig),
+                    rbind(acp_auc, chen_liu_auc, shi_et_al_auc, caramelo_auc, bc_auc), show_outcome = FALSE)
+
+```

### 2020-10-11 R/auc_tools.R

@@ -7,14 +7,20 @@ outcome_caption = c("alive_12" = "12-day mortality", "alive_30" = "30-day mortal
                     "not_oxygen_admission" = "Not on oxygen on admission")
 score_caption = c("ACP_grade" = "Li et al. (ACP)",
                   "chen_liu" = "Chen & Liu",
-                  "chen_liu_incl" = "Chen & Liu",
                   "shi" = "Shi et al.",
                   "caramelo" = "Caramelo et al.",
-                  "bello_chavolla", "Bello-Chavolla et al.")
+                  "bc_base" = "Bello-Chavolla et al.",
+                  "bc0" = "Bello-Chavolla et al.",
+                  "bc1" = "Bello-Chavolla et al.",
+                  "caramelo_base" = "Caramelo et al.",
+                  "caramelo1" = "Caramelo et al.",
+                  "caramelo2" = "Caramelo et al.",
+                  "age" = "Age only"
+)
 
 outcome_note_caption = c("incl" = " (inclusion crit.)")
 
-my_auc <- function(f, d, outcome_note = "", subgroup = "") {
+my_auc_summary <- function(f, d, outcome_note = "", subgroup = "") {
   roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
   ci = pROC::ci.auc(roc)
 
@@ -22,8 +28,7 @@ my_auc <- function(f, d, outcome_note = "", subgroup = "") {
   score = as.character(formula.tools::rhs(f))
 
 
-  list(roc = roc,
-       summary = tibble::tibble(
+  tibble::tibble(
          auc = ci[2],
          auc_low = ci[1],
          auc_high = ci[3],
@@ -31,79 +36,90 @@ my_auc <- function(f, d, outcome_note = "", subgroup = "") {
          score = score,
          outcome_note = outcome_note,
          subgroup = subgroup
-       )
+   )
+}
+
+my_auc_summary_from_estimate <- function(auc_estimate, outcome, score, outcome_note = "", subgroup = "") {
+  tibble::tibble(
+    auc = auc_estimate,
+    auc_low = NA_real_,
+    auc_high = NA_real_,
+    outcome = outcome,
+    score = score,
+    outcome_note = outcome_note,
+    subgroup = subgroup
   )
 }
 
-plot_my_auc <- function(auc_obj) {
-  if(!(auc_obj$summary$score %in% names(score_caption))) {
+
[…]

### 2020-10-11 manuscript/risk_scores.Rmd
Tweaking risk scores for consultation

@@
 
-In all of the following we focus on the C-statistic, also often called area under curve (AUC). We compute AUC using the  `pROC` package, which also reports bootstrapped confidence intervals.
+In all of the following we focus on the area under curve (AUC), also often called  C-statistic.
+We compute AUC using the  `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67], which also reports bootstrapped confidence intervals.
+
+For all risk scores we look at 12-day mortality and 30-day mortality - simply because those are the time horizons reported in the studies we looked at.
+When needed or when it is uncertain how the outcome was coded in the original study, we can also look at "indefinite horizon" where we only include patients that were not censored and had a known outcome.
 
@@
 
-li_et_al_data
+li_et_al_data %>% group_by(ACP_grade, alive_12) %>% summarise(N = n(), .groups = "drop")
 
@@
 acp_auc_orig <- my_auc_summary(alive_12 ~ ACP_grade, li_et_al_data)
-acp_auc_orig %>% transmute(`C-static` = auc, `95% CI` = paste0("(",auc_low, ", ",auc_high, ")"))
+acp_auc_orig %>% transmute(AUC = format(auc, digits = 3), `95% CI` = paste0("(",format(auc_low, digits = 3), ", ", format(auc_high, digits = 3), ")")) %>% knitr::kable(booktabs = TRUE)
 
@@
 ```{r, fig.cap="The AUC and it's 95% credible intervals for the ACP score.
-The thick blue vertical line marks AUC of 0.5, i.e.
+The thick blue line marks AUC of 0.5, i.e.
 the point where reverting the score would yield better predictions.
@@
 )
-compare_auc_to_orig(acp_auc_orig, acp_auc)
+compare_auc_to_orig(acp_auc_orig, acp_auc) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
 
-```{r}
-#acp_auc
-```
+We see that the score noticeably underperforms in the present dataset.
 
@@
 
+Chen and Liu [@http://zotero.org/users/5567156/items/4DDUVEKS] propose a score based on a logistic regression model as follows.
+
 $$
-Score = −4.211+0.013 \times CRP+0.059 \times Age +0.112 \times DD −1.984 \times LYM
+Score = -4.211+0.013 \times CRP+0.059 \times Age +0.112 \times DD -1.984 \times LYM
 $$
 
-The paper does not have units, but the web app
+Where CRP is C-reactive protein (mg/l) DD is D-dimer (the paper does not have units, but the associated web app is clear that it expects D-dimer as ug/ml FEU) and LYM is the Lympohocyte count (10^9 / L).
+The paper states:
 
@@
 
-Slightly unclear how they computed the outcome.
+However, reverse engineering the asociated web app (https://phenomics.fudan.edu.cn/risk_scores/), no transform appears to have been done to the variables that were finally selected.
+It is also slightly unclear how they computed the outcome (i.e.
+how they treated censoring).
 
-CRP: mg/l
-D-dimer: ug / mL FEU (we use ng/ml DDU, FEU value is double the DDU)
-Lymphocytes: 10^9 / L
+The units in our dataset are:
 
@@
 
-Results for the validation dataset:
-AUROC = 0.881
-Cutoff (0.5)
-Sensitivity = 0.839
-Specificity = 0.794
-
-Precise coefficient values and the CI range determined by reverse engineering the webapp at https://phenomics.fudan.edu.cn/risk_scores/
+Which matches the model except for D-dimer (we use ng/ml DDU, FEU value is double the DDU).
 
 ```{r}
+# More precise coefficient values and the CI range determined by reverse engineering the webapp
+
 chen_liu_data <- data$marker_data_wide %>%  
@@
     .groups = "drop") %>% 
-  inner_join(data$patient_data, by = c("patient_id")) %>%
+  inner_join(at_admission, by = c("patient_id")) %>%
   mutate(    
@@
     risk_upper = 1 / (1 + exp(-chen_liu - 1.96*0.1721)),
-    alive_12 = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
+    alive_12_data = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
                          outcome == "Death" ~ FALSE,
                          TRUE ~ NA),
-    alive_30 = case_when(last_record >= first_day + 30 | outcome == "Discharged" ~ TRUE,
+    alive_30_data = case_when(last_record >= first_day + 30 | outcome == "Discharged" ~ TRUE,
                          outcome == "Death" ~ FALSE,
@@
                          outcome == "Death" ~ FALSE,
-                         TRUE ~ NA),
-    alive_imputed = outcome != "Death"
+                         TRUE ~ NA)
 
@@
 
+A plot of the computed values for our dataset.
+
 ```{r}
@@
 ```
+
+We look at survival since first day when all of the values required for the score are available, or we ignore when the values were computed and look at survival since admission.
+Finally we also tried emulating the inclusion criteria from Chen & Lie
+
+Chen & Liu report AUC = 0.881 for the validation dataset and don't provide enough information to let us recompute it with credible intervals.
 
@@
   my_auc_summary(alive_30 ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_12_data ~ chen_liu, chen_liu_data),
+  my_auc_summary(alive_30_data ~ chen_liu, chen_liu_data),
   my_auc_summary(alive_all ~ chen_liu, chen_liu_data),
-  my_auc_summary(alive_imputed ~ chen_liu, chen_liu_data),
   my_auc_summary(alive_12 ~ chen_liu, chen_liu_data_inclusion, "incl"),
   my_auc_summary(alive_30 ~ chen_liu, chen_liu_data_inclusion, "incl"),
-  my_auc_summary(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl"),
-  my_auc_summary(alive_imputed ~ chen_liu, chen_liu_data_inclusion, "incl")
+  my_auc_summary(alive_12_data ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc_summary(alive_30_data ~ chen_liu, chen_liu_data_inclusion, "incl"),
+  my_auc_summary(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl")
 )
@@
 
-```{r}
+```{r, fig.cap = "The AUC and it's 95% credible intervals for the Chen & Liu score.
+The thick blue line marks AUC of 0.5, i.e.
+the point where reverting the score would yield better predictions.
+The thin gray dashed line marks the AUC from the original study for easier comparison.
+The outcomes ending with `_data` mean mortality was counted since the first day all the lab markers were measured, otherwise it is since hospitalization.
+`alive_all` mens that only patients that were not censored were included.
+"}
 chen_liu_auc_orig <- my_auc_summary_from_estimate(0.881, "alive", "chen_liu", "incl")
@@
 
-1 point for Age >= 50, Male, Hypertension
-
-Not sure what "severe" means, will test both "not ventilated", "not needing Oxygen"
+Shi et al.
+[@http://zotero.org/users/5567156/items/X4GWJIP5] report a simple score: 1 point for Age >= 50, Male and Hypertension each.
+They try to predict "severe" illness, but it is not clear what "severe" means.
+Here, we test both "not ventilated" (or worse, including death) and "not needing supplementary oxygen" as proxy for "severe".
+They also create a risk score for both "severe" on admission and "severe" during hospitalization.
+As for other scores we also look at 12- and 30- day mortality.
 
@@
   my_auc_summary(not_oxygen_all ~ shi, shi_et_al_data),
-  my_auc_summary(not_ventilated_admission ~ shi, shi_et_al_data, subgroup = " (at admission)"),
-  my_auc_summary(not_oxygen_admission ~ shi, shi_et_al_data, subgroup = " (at admission)")
+  my_auc_summary(not_ventilated_admission ~ shi, shi_et_al_data, subgroup = " (admission)"),
+  my_auc_summary(not_oxygen_admission ~ shi, shi_et_al_data, subgroup = " (admission)")
 )
@@
 
-From Fig1 B,C:
+We extract the original data from Fig1 B,C using WebPlotDigitizer.
+
 ```{r}
 shi_et_al_reported <- data.frame(
-  score = c(0,1,2,3), shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429),
+  score = c(0,1,2,3), 
+  shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429),
   mild_cases_admission = c(118, 199, 98, 21), all_cases_admission = c(118, 211, 121, 35),
@@
 )
+
+shi_et_al_reported %>% select(-starts_with("shi_")) %>% knitr::kable(booktabs = TRUE)
+
+```
+
+```{r}
 
@@
 
-shi_et_al_auc_reported_admission <- my_auc_summary(severe_admission ~ shi, shie_et_al_auc_data_admission, subgroup = " (at admission)")
+shi_et_al_auc_reported_admission <- my_auc_summary(severe_admission ~ shi, shie_et_al_auc_data_admission, subgroup = " (admission)")
 #shi_et_al_auc_reported_admission$summary
@@
 
-compare_auc_to_orig(rbind(shi_et_al_auc_reported_admission, shi_et_al_auc_reported_all), shi_et_al_auc, ncol = 1)
 ```
+
+```{r, fig.cap="The AUC and it's 95% credible intervals for the Shi et al.
+score.
+The thick blue line marks AUC of 0.5, i.e.
+the point where reverting the score would yield better predictions.
+The thin gray dashed line marks the AUC from the original study for easier comparison."}
+
+compare_auc_to_orig(rbind(shi_et_al_auc_reported_admission, shi_et_al_auc_reported_all), shi_et_al_auc, nrow = 1)
+```
+We see that the score consistently underperforms, although some of the CIs are quite wide.
+
+We can also compare the ratios of "severe" cases reported by Shi et al. to ratios in our dataset.
 
@@
 
-# Caramelo
+# Caramelo et al.
 
-Uses simulations to get patient-level characterstics from aggregate data.
+Caramelo et al.
+[@http://zotero.org/users/5567156/items/3A2RN4ZS] uses simulations to get patient-level characterstics from aggregate data and then fit a logistic regression model.
+We use the coefficients they report for the model (the model can't be used directly, because the intercept is not reported).
+The covariates in the model are age (by decades), sex, presence of hypertension, presence of diabetes, presence of cardiac disease, presence of chronic respiratory disease and presence of cancer.
+It is unclear whether NYHA > 1 would be considered "cardiac disease" and similarly whether asthma is chronic respiratory disease, so we compute two versions of the score.
+Our dataset also does not contain cancer, so we ignore it (cancer has one of the lowest reported OR in the model).
 
-```{r}
+Caramelo et al. have not computed AUC in their paper and do not give us any mean to reconstruct it, so we only present our results for the risk score.
+
+```{r, fig.cap="The AUC and it's 95% credible intervals for the Caramelo et al. score. The thick blue line marks AUC of 0.5, i.e. the point where reverting the score would yield better predictions. "}
 caramelo_data <- at_admission %>%
@@
 ```
+The results look somewhat encouraging for the score - the AUC values float around 0.8.
 
@@
 ```{r}
-bello_chavolla_data %>% ggplot(aes(x = bc0, y = as.integer(alive_30))) +
-  geom_smooth() + geom_jitter(width = 0.3, height = 0.3) 
-bello_chavolla_data %>% ggplot(aes(x = bc_base, y = as.integer(alive_30))) +
-  geom_smooth() + geom_jitter(width = 0.3, height = 0.3)
+# bello_chavolla_data %>% ggplot(aes(x = bc0, y = as.integer(alive_30))) +
+#   geom_smooth() + geom_jitter(width = 0.3, height = 0.3) 
+# bello_chavolla_data %>% ggplot(aes(x = bc_base, y = as.integer(alive_30))) +
+#   geom_smooth() + geom_jitter(width = 0.3, height = 0.3)
 
@@
 
-```{r}
+```{r, fig.cap = "The AUC and it's 95% credible intervals for the Bello-Chavolla et al.
+score.
+The thick blue line marks AUC of 0.5, i.e.
+the point where reverting the score would yield better predictions.
+The thin gray dashed line marks the AUC from the original study for easier comparison.
+We use 3 variants of the score - `base` means not inferring pneumonia at all, `day0` means that pneumonia is inferred from antibiotics at hospitalization, `day1` - inferring pneumonia from antibiotics within 1 day of hospitalization "}
 bc_auc <- rbind(
@@
   my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, outcome_note = "base"),
-  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data),
-  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data),
+  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data, outcome_note = "day0"),
+  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data, outcome_note = "day0"),
   my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, outcome_note = "day1"),
@@
 ```
+The main surprise here is that not inferring pneumonia (shown as "base" in the figure) works best. This indicates that our data may not be well suited to the use of the score.
+
 # Age only
 
-As a simple baseline we use age in years as the sole predictor.
+As a simple baseline we use age in years as the sole predictor and compute AUC for this.
 
@@
 )
+age_auc
 
@@
 
-```{r}
+```{r, fig.cap = "The AUC and it's 95% credible intervals for all of the scores considered so far.
+The thick blue line marks AUC of 0.5, i.e.
+the point where reverting the score would yield better predictions.
+The thin gray dashed line marks the AUC from the original study for easier comparison."}
+
+all_auc <- rbind(acp_auc, chen_liu_auc, shi_et_al_auc, caramelo_auc, bc_auc, age_auc)
 compare_auc_to_orig(rbind(acp_auc_orig, chen_liu_auc_orig, shi_et_al_auc_reported_admission,
                          shi_et_al_auc_reported_all, bc_auc_orig),
-                    rbind(acp_auc, chen_liu_auc, shi_et_al_auc, caramelo_auc, bc_auc), show_outcome = FALSE)
+                    all_auc, show_outcome = FALSE)
 
 ```
+We see that the "age only" baseline actually outperforms all of the other scores - although there is some uncertainty prohibiting us to be very sure about it.
+And the second best performing score (Caramelo et al.) is actually heavily based on age.
+Table of the results below as the small differences are not well visible in the plot.
+
+```{r}
+all_auc %>% mutate(score = score_caption[score]) %>%
+  group_by(score, subgroup) %>% 
+  summarise(best_auc = max(auc), .groups = "drop") %>%
+  arrange(desc(best_auc))
+```
+
+# References

### 2020-10-12 manuscript/risk_scores.Rmd
More tweaks for risk scores

@@
 
-```{r, fig.cap="The AUC and it's 95% credible intervals for the ACP score.
+```{r, fig.cap="The AUC and it's 95\\% credible intervals for the ACP score.
 The thick blue line marks AUC of 0.5, i.e.
@@
 )
-compare_auc_to_orig(acp_auc_orig, acp_auc) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+compare_auc_to_orig(acp_auc_orig, acp_auc)
 
@@
 
+Here is the comparison to original data.
+
 ```{r, fig.cap="The AUC and it's 95% credible intervals for the Shi et al.
@@
 ```
+
 We see that the score consistently underperforms, although some of the CIs are quite wide.
@@
   my_auc_summary(alive_30 ~ caramelo1, caramelo_data),
-  my_auc_summary(alive_12 ~ caramelo2, caramelo_data, outcome_note = "v2"),
-  my_auc_summary(alive_30 ~ caramelo2, caramelo_data, outcome_note = "v2"),
-  my_auc_summary(alive_12 ~ caramelo_base, caramelo_data, outcome_note = "base"),
-  my_auc_summary(alive_30 ~ caramelo_base, caramelo_data, outcome_note = "base")
+  my_auc_summary(alive_12 ~ caramelo2, caramelo_data, note = "v2"),
+  my_auc_summary(alive_30 ~ caramelo2, caramelo_data, note = "v2"),
+  my_auc_summary(alive_12 ~ caramelo_base, caramelo_data, note = "base"),
+  my_auc_summary(alive_30 ~ caramelo_base, caramelo_data, note = "base")
 )
@@
 ```
+
 The results look somewhat encouraging for the score - the AUC values float around 0.8.
 
-# Bello-Chavolla
+# Bello-Chavolla et al.
 
@@
 bc_auc <- rbind(
-  my_auc_summary(alive_30 ~ bc_base, bello_chavolla_data, outcome_note = "base"),
-  my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, outcome_note = "base"),
-  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data, outcome_note = "day0"),
-  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data, outcome_note = "day0"),
-  my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, outcome_note = "day1"),
-  my_auc_summary(alive_12 ~ bc1, bello_chavolla_data, outcome_note = "day1")
+  my_auc_summary(alive_30 ~ bc_base, bello_chavolla_data, note = "base"),
+  my_auc_summary(alive_12 ~ bc_base, bello_chavolla_data, note = "base"),
+  my_auc_summary(alive_30 ~ bc0, bello_chavolla_data, note = "day0"),
+  my_auc_summary(alive_12 ~ bc0, bello_chavolla_data, note = "day0"),
+  my_auc_summary(alive_30 ~ bc1, bello_chavolla_data, note = "day1"),
+  my_auc_summary(alive_12 ~ bc1, bello_chavolla_data, note = "day1")
 )
@@
 ```
+
 The main surprise here is that not inferring pneumonia (shown as "base" in the figure) works best. This indicates that our data may not be well suited to the use of the score.
@@
 
-As a simple baseline we use age in years as the sole predictor and compute AUC for this.
+As a simple baseline we use age in years and the decade of age as the sole predictor and compute AUC for this.
 
 ```{r}
+age_only_data <- at_admission %>%
+  mutate(age_decade = floor(age / 10))
 age_auc <-rbind(
-  my_auc_summary(alive_30 ~ age, at_admission),
-  my_auc_summary(alive_12 ~ age, at_admission)
+  my_auc_summary(alive_30 ~ age, age_only_data),
+  my_auc_summary(alive_12 ~ age, age_only_data),
+  my_auc_summary(alive_30 ~ age_decade, age_only_data, note = "decade"),
+  my_auc_summary(alive_12 ~ age_decade, age_only_data, note = "decade")
 )
@@
 
+# plot_my_auc(alive_12 ~ age, at_admission)
+
+```
+
+```{r, fig.cap = "The ROC curve for age in years as the sole predictor of 30-day mortality."}
 plot_my_auc(alive_30 ~ age, at_admission)
-plot_my_auc(alive_12 ~ age, at_admission)
-
 ```
@@
 ```
+
 We see that the "age only" baseline actually outperforms all of the other scores - although there is some uncertainty prohibiting us to be very sure about it.
-And the second best performing score (Caramelo et al.) is actually heavily based on age.
-Table of the results below as the small differences are not well visible in the plot.
+This holds even if we only take the decade of age.
+And the second best performing score (Caramelo et al.) is actually heavily based on the decade of age.
+Table of the results is shown below as the small differences are not well visible in the plot.
 
@@
 all_auc %>% mutate(score = score_caption[score]) %>%
-  group_by(score, subgroup) %>% 
+  group_by(score, subgroup, note) %>% 
   summarise(best_auc = max(auc), .groups = "drop") %>%
-  arrange(desc(best_auc))
+  arrange(desc(best_auc)) %>%
+  knitr::kable(booktabs = TRUE)
 ```

### 2020-10-12 R/auc_tools.R

@@ -15,12 +15,13 @@ score_caption = c("ACP_grade" = "Li et al. (ACP)",
                   "caramelo_base" = "Caramelo et al.",
                   "caramelo1" = "Caramelo et al.",
                   "caramelo2" = "Caramelo et al.",
-                  "age" = "Age only"
+                  "age" = "Age only",
+                  "age_decade" = "Age only"
 )
 
-outcome_note_caption = c("incl" = " (inclusion crit.)")
+note_caption = c("incl" = " (inclusion crit.)")
 
-my_auc_summary <- function(f, d, outcome_note = "", subgroup = "") {
+my_auc_summary <- function(f, d, note = "", subgroup = "") {
   roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
   ci = pROC::ci.auc(roc)
 
@@ -34,25 +35,25 @@ my_auc_summary <- function(f, d, outcome_note = "", subgroup = "") {
          auc_high = ci[3],
          outcome = outcome,
          score = score,
-         outcome_note = outcome_note,
+         note = note,
          subgroup = subgroup
    )
 }
 
-my_auc_summary_from_estimate <- function(auc_estimate, outcome, score, outcome_note = "", subgroup = "") {
+my_auc_summary_from_estimate <- function(auc_estimate, outcome, score, note = "", subgroup = "") {
   tibble::tibble(
     auc = auc_estimate,
     auc_low = NA_real_,
     auc_high = NA_real_,
     outcome = outcome,
     score = score,
-    outcome_note = outcome_note,
+    note = note,
     subgroup = subgroup
   )
 }
 
 
-plot_my_auc <- function(f, d, outcome_note = "") {
+plot_my_auc <- function(f, d, note = "") {
 
   outcome = as.character(formula.tools::lhs(f))
   score = as.character(formula.tools::rhs(f))
@@ -64,18 +65,18 @@ plot_my_auc <- function(f, d, outcome_note = "") {
     stop("Outcome does not have caption")
   }
 
-  if(outcome_note != "") {
-    if(!(outcome_note %in% names(outcome_note_caption))) {
+  if(note != "") {
+    if(!(note %in% names(note_caption))) {
       stop("Outcome note does not have caption")
     }
-    outcome_note_caption_value <- outcome_note_caption[outcome_note]
+    note_caption_value <- note_caption[note]
   } else {
-    outcome_note_caption_value <- ""
+    note_caption_value <- ""
[…]

### 2020-10-15 R/data_mapping.R
Special lab data processing - first working version

@@ -14,10 +14,12 @@ marker_map <- c(
         "CRP" = "CRP",
         "IL-6" = "IL_6",
         "PCT" = "procalcitonin",
+        "Prokalcitonin" = "procalcitonin",
         "Lymphocyte count" = "lymphocyte_count",
         "Leukocyte count" = "leukocyte_count",
         "Leukocytes" = "leukocyte_count",
         "Leukocyte" = "leukocyte_count",
+        "creatinine" = "creatinine",
         "urea" = "urea",
         "IgG" = "IgG",
         "ARO" = "ICU", # We know the patient was on ICU, but no details on breathing
@@ -29,17 +31,20 @@ marker_map <- c(
         "Covalescent plasma" = "convalescent_plasma",
         "Dexametason" = "dexamethasone",
         "Dexamethasone" = "dexamethasone",
+        "Dexamethason" = "dexamethasone",
         "Remdesivir" = "remdesivir",
       ## Antibiotics
         "Amoxicillin/Clavulanic" = "amoxiclav",
         "Amoxicillin/Clavulanate" = "amoxiclav",
         "amoxiclin/klavulanat" = "amoxiclav",
+        "Amoxicilin/klavulan\U00e1t" = "amoxiclav",
         "Augmentin" = "amoxiclav",
         "Piperacilin/tazobactam" = "piperacilin_tazobactam",
         "Piperacillin/Tazobactam" = "piperacilin_tazobactam",
         "Piperacilin Tazobactam" = "piperacilin_tazobactam",
         "Clarithromycin" = "clarithromycin", # Also a macrolide
         "Clarithomcin" = "clarithromycin",
+        "Klarithromycin" = "clarithromycin",
         "sulfamethoxazol" = "sulfamethoxazol", # Combines with trimethoprim to "Cotrimoxazol"
         "trimethoprim" = "trimethoprim",
         "ofloxacin" = "ofloxacin",
@@ -47,6 +52,7 @@ marker_map <- c(
         "Ceftriaxon" = "ceftriaxone",
         "cefuroxime" = "cefuroxime",
         "cefuroxim" = "cefuroxime",
+        "Cefuroxim axetim" = "cefuroxime",
         "cefotaxim" = "cefotaxime",
         "cefotaxime" = "cefotaxime",
         "cefepim" = "cefepime",
@@ -131,6 +137,11 @@ unit_overwrites <- list(
[…]

### 2020-10-15 R/raw_data_preprocessing.R

@@ -416,7 +416,8 @@ read_progression_data <- function(file, hospital_id, lang, file_version, patient
     mutate(censored = case_when(startsWith(value, "<") ~ "right",
                                 startsWith(value, ">") ~ "left",
                                 TRUE ~ "none"),
-           value_parsed = as.numeric(gsub("^[<>]? *", "", value)))
+           value_parsed =  as.numeric(if_else(tolower(value) == "no", "0",
+                                              gsub("^[<>]? *", "", value))))
 
 
   bad_parse_value <- marker_data %>% filter(is.na(value_parsed) & !is.na(value))
@@ -621,3 +622,4 @@ anonymize_for_analysis <- function(complete_data) {
 
   anonymized_data
 }
+

### 2020-10-15 R/special_lab_data.R

@@ -1,0 +1,289 @@
+special_lab_data_ignored_patterns <- c(
+  "^.adatel:",
+  "^Pl.tce ?:",
+  "^Protokol",
+  "^-*$",
+  "^Odd.leni klinick. mikrobiologie$",
+  "^DIF - STANOVEN. ANALYZ.TOREM$",
+  "^Nen. zn.mo, zda pacient ",
+  "^Hodn\\.:",
+  "^[0-9]{5} -"
+)
+
+special_lab_data_sections <- c(
+  "Plazma",
+  "KREVN\U00CD OBRAZ",
+  "DIF ANALYZ\U00C1TOR",
+  "KOAGULACE",
+  "Biochemie",
+  "Mo\U010D chemicky a sedim",
+  "Mo\U00E8 chemicky a sedim",
+  "Mo\U010D",
+  "Astrup",
+  "DIF MIKROSKOPICKY",
+  "Imunologie",
+  "RETIKULOCYTY",
+  "Transfuzn\U00ED",
+  "Punkt\U00E1t",
+  "L\U00E9\U010Diva",
+  "Markery hepatitid",
+  "Toxikologie"
+
+)
+
+special_lab_data_markers_to_collect <-
+  c("PT (QT) INR" = "PT",
+    "Ddim" = "d_dimer",
+    "Ly#" = "lymphocyte_count",
+    "LYa" = "lymphocyte_count",
+    "Kre" = "creatinine",
+    "CRP" = "CRP",
+    "PCT" = "procalcitonin",
+    "Alb" = "albumin",
+    "Ferr" = "ferritin",
+    "IL-6" = "IL_6",
+    "IL6" = "IL_6"
+  )
+
+special_lab_data_markers_to_collect_units <-
+  c("PT (QT) INR" = "INR",
+    "Ddim" = "ng/ml DDU",
+    "Ly#" = "10^9/l",
+    "LYa" = "10^9/l",
+    "Kre" = "\U03BCmol/l",
+    "CRP" = "mg/l",
+    "PCT" = "\U03BCg/l",
+    "Alb" = "g/l",
+    "Ferr" = "\U03BCg/l",
+    "IL-6" = "ng/ml",
+    "IL6" = "ng/ml"
+  )
+
+if(!identical(names(special_lab_data_markers_to_collect), names(special_lab_data_markers_to_collect_units))){
+  stop("Bad names")
+}
+
+read_special_lab_data <- function(input_file, patient_id, hospital_id, first_day) {
+  lines <- readLines(input_file)
[…]

### 2020-10-16 dev/raw_data_preprocessing.Rmd
Special lab data processing - first working version
Succesful read of special format data

@@
 
-```{r}
+```{r setup}
 devtools::load_all()
@@
 
-TODO: code adverse events, code discontinued_medication_reason
-TODO: check last day, add outcome to breathing data, possibly
+TODO: check last day
 
@@
 
-  hospital_data[[i]] <- read_raw_data(file, hospital_id, lang, file_version)
+  hospital_data[[hospital_id]] <- read_raw_data(file, hospital_id, lang, file_version)
 }
@@
 differing_units <- complete_data$marker_data %>% group_by(marker) %>% filter(length(unique(unit)) > 1) %>% 
-  group_by(marker, unit) %>% summarise(count = n(), .groups = "drop")
+  group_by(marker, unit) %>% summarise(count = n(), hospitals = paste0(unique(hospital_id), collapse = ","), .groups = "drop")
 
@@
 ```
-```{r}
-complete_data$marker_data %>% filter(marker == "amoxiclav", value == 0)
-```
-
-```{r}
-complete_data$marker_data %>% filter(marker == "CRP", unit == "mg/l") %>% pull(hospital_id) %>%   unique()
-
-```
 
@@
 ```{r}
-patients_to_merge <- readxl::read_excel(mapping_file, sheet = "PatientsToMerge")
+patients_to_merge <- readxl::read_excel(mapping_file, sheet = "PatientsToMerge", 
+                                        col_types = c("text", "text", "text", "text", "text", "numeric", "text"))
 hospitals_to_merge <- readxl::read_excel(mapping_file, sheet = "HospitalsToMerge")
@@
 for(i in 1:(length(hospital_data) - 1)) {
+  if(names(hospital_data)[i] %in% c("cTdij", "iVMlA","bGzOO", "tMdnA", "nDINR", "vQNDS")) {
+    next # Checked manually during date entry
+  }
   for(j in (i + 1):length(hospital_data)) {
     #cat(i,",", j,"\n")
+    if(names(hospital_data)[j] %in% c("cTdij", "iVMlA","bGzOO", "tMdnA", "nDINR","vQNDS")) {
+      next # Checked manually during date entry
+    }
 
@@
 ```{r}
-complete_data <- merge_patients(complete_data, patients_to_merge)
+typeof(patients_to_merge$patient_id1)
+complete_data_merged <- merge_patients(complete_data, patients_to_merge)
 
 complete_data$patient_data %>% filter(hospital_id == "Oriik", patient_id == 3)
+complete_data_merged$breathing_data %>% filter(hospital_id == "cTdij")
+complete_data_merged$patient_data %>% filter(hospital_id == "cTdij")
+complete_data$patient_data %>% filter(hospital_id == "cTdij")
+
 ```

### 2020-10-16 R/data_mapping.R

@@ -9,6 +9,7 @@ marker_map <- c(
         "SpO2 native" = "SpO2_native",
         "Horowitz index" = "Horowitz_index",
         "PEEP" = "PEEP",
+        "FiO2" = "FiO2",
         "Ferritin" = "ferritin",
         "D-dimer" = "d_dimer",
         "CRP" = "CRP",
@@ -22,22 +23,27 @@ marker_map <- c(
         "creatinine" = "creatinine",
         "urea" = "urea",
         "IgG" = "IgG",
-        "ARO" = "ICU", # We know the patient was on ICU, but no details on breathing
+        "ARO" = "ARO", # We know the patient was on anesthesia-resuscitation ward, but no details on breathing
       ## Covid-specific drugs
         "Hydroxychloroquine" = "hcq",
         "Kaletra" = "kaletra",
         "Azithromycin" = "az",
+        "Sumamed" = "az",
         "Tocilizumab" = "tocilizumab",
         "Covalescent plasma" = "convalescent_plasma",
+        "Convalescent plasma" = "convalescent_plasma",
         "Dexametason" = "dexamethasone",
         "Dexamethasone" = "dexamethasone",
         "Dexamethason" = "dexamethasone",
         "Remdesivir" = "remdesivir",
+        "Favipiravir" = "favipiravir",
       ## Antibiotics
         "Amoxicillin/Clavulanic" = "amoxiclav",
         "Amoxicillin/Clavulanate" = "amoxiclav",
         "amoxiclin/klavulanat" = "amoxiclav",
         "Amoxicilin/klavulan\U00e1t" = "amoxiclav",
+        "Amoksiklav" = "amoxiclav",
+        "Amoxiclav" = "amoxiclav",
         "Augmentin" = "amoxiclav",
         "Piperacilin/tazobactam" = "piperacilin_tazobactam",
         "Piperacillin/Tazobactam" = "piperacilin_tazobactam",
@@ -45,14 +51,23 @@ marker_map <- c(
         "Clarithromycin" = "clarithromycin", # Also a macrolide
         "Clarithomcin" = "clarithromycin",
         "Klarithromycin" = "clarithromycin",
+        "Clacid" = "clarithromycin",
+        "Klacid" = "clarithromycin",
+        "Biseptol" = "cotrimoxazole",
[…]

### 2020-10-16 R/raw_data_preprocessing.R

@@ -22,7 +22,7 @@ update_and_check_version <- function(file, lang, file_version) {
     file_version <- readxl::read_excel(file, sheet = version_sheet(lang), range = "C7:C7", col_types = "text", col_names = "version")$version
   }
 
-  if(!grepl("^(Old|1\\.[2-4]\\.[0-9])$", file_version)) {
+  if(!grepl("^(Old|Special|1\\.[2-4]\\.[0-9])$", file_version)) {
     print(file_version)
     stop("Invalid version")
   }
@@ -33,7 +33,7 @@ update_and_check_version <- function(file, lang, file_version) {
 read_patient_data <- function(file, hospital_id, lang, file_version, remove_examples) {
   patient_data_raw <- read_patient_data_raw(file, hospital_id, lang, file_version)
 
-  patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(Age), !is.na(Sex))
+  patient_data_raw_usable <- patient_data_raw %>% filter(!is.na(`Patient ID`), !is.na(Age), !is.na(Sex))
 
   if(remove_examples) {
     patient_data_raw_usable <- patient_data_raw_usable %>%
@@ -50,22 +50,6 @@ read_patient_data <- function(file, hospital_id, lang, file_version, remove_exam
 
   outcome_labels <- c("Discharged","Hospitalized","Transferred","Death")
 
-  # Special handling of badly filled outcomes
-  if(hospital_id == "QKuFp") {
-    mismatched_outcomes <- grepl("^439[0-9]{2}$", patient_data_raw_usable$`Outcome (Discharged/Hospitalized/Transferred/Death)`)
-    patient_data_raw_usable <- patient_data_raw_usable %>%
-      mutate(
-        `Date of last record` =
-          if_else(mismatched_outcomes,
-                  janitor::excel_numeric_to_date(
-                    as.integer(if_else(mismatched_outcomes, `Outcome (Discharged/Hospitalized/Transferred/Death)`, "0"))),
-                  lubridate::as_date(`Date of last record`)),
-        `Outcome (Discharged/Hospitalized/Transferred/Death)` =
-               if_else(mismatched_outcomes,
-                       "Hospitalized",
-                       `Outcome (Discharged/Hospitalized/Transferred/Death)`)
-        )
-  }
   if(hospital_id == "ZDYlY") {
     patient_data_raw_usable <- patient_data_raw_usable %>%
[…]

### 2020-10-16 R/special_lab_data.R

@@ -32,7 +32,7 @@ special_lab_data_sections <- c(
 )
 
 special_lab_data_markers_to_collect <-
-  c("PT (QT) INR" = "PT",
+  c("INR" = "pt_inr",
     "Ddim" = "d_dimer",
     "Ly#" = "lymphocyte_count",
     "LYa" = "lymphocyte_count",
@@ -46,7 +46,7 @@ special_lab_data_markers_to_collect <-
   )
 
 special_lab_data_markers_to_collect_units <-
-  c("PT (QT) INR" = "INR",
+  c("INR" = "INR",
     "Ddim" = "ng/ml DDU",
     "Ly#" = "10^9/l",
     "LYa" = "10^9/l",
@@ -55,8 +55,8 @@ special_lab_data_markers_to_collect_units <-
     "PCT" = "\U03BCg/l",
     "Alb" = "g/l",
     "Ferr" = "\U03BCg/l",
-    "IL-6" = "ng/ml",
-    "IL6" = "ng/ml"
+    "IL-6" = "ng/l",
+    "IL6" = "ng/l"
   )
 
 if(!identical(names(special_lab_data_markers_to_collect), names(special_lab_data_markers_to_collect_units))){
@@ -221,7 +221,7 @@ read_special_lab_data <- function(input_file, patient_id, hospital_id, first_day
         if(grepl("pozitivn.$", s)) {
           value = 1
         } else if(grepl("negativn.$", s)) {
-          value = "0"
+          value = 0
         } else {
           stop("Invalid Covid test res")
         }
@@ -264,11 +264,53 @@ read_special_lab_data <- function(input_file, patient_id, hospital_id, first_day
       }
     }
   }
-  do.call(rbind, res)
+  res_all <- do.call(rbind, res)
+
+  if(any(is.na(res_all$value))) {
+    print(res_all %>% filter(is.na(value)))
+    stop("NAs in values")
+  }
+
+  res_all
 }
 
 
+read_special_lab_data_all <- function(dir, hospital_id, patient_ids, first_days) {
+  res <- list()
+  for(i in 1:length(patient_ids)) {
+    filename <- paste0(dir, "/P", patient_ids[i], ".txt")
+    if(!file.exists(filename)) {
+      warning(paste0(filename, " does not exist\n"))
+    } else {
+      print(filename)
+      res[[i]] <- read_special_lab_data(filename, patient_ids[i], hospital_id, first_days[i])
+    }
+  }
 
[…]

### 2020-10-16 dev/raw_data_preprocessing.Rmd
Preprocessing final data

@@
 
+## Some more checks
+
+unknown_breathing
+
+```{r}
+max_days <- complete_data$breathing_data %>% group_by(hospital_id, patient_id) %>%
+  summarise(max_day = max(day))
+
+max_max_day = max(max_days$max_day)
+
+crossing(max_days, day = 0:max_max_day) %>% filter(day <= max_day, day >= 0) %>%
+  anti_join(complete_data$breathing_data, by = c("hospital_id", "patient_id", "day"))
+
+complete_data$breathing_data %>% filter(is.na(breathing)) 
+```
+
+Markers outside breathing
+
 ## Merging patients/wards
@@
 typeof(patients_to_merge$patient_id1)
-complete_data_merged <- merge_patients(complete_data, patients_to_merge)
+complete_data_merged_1 <- merge_patients(complete_data, patients_to_merge)
+```
 
-complete_data$patient_data %>% filter(hospital_id == "Oriik", patient_id == 3)
-complete_data_merged$breathing_data %>% filter(hospital_id == "cTdij")
-complete_data_merged$patient_data %>% filter(hospital_id == "cTdij")
-complete_data$patient_data %>% filter(hospital_id == "cTdij")
+```{r}
+complete_data_merged <- merge_hospitals(complete_data_merged_1, hospitals_to_merge)
+
+complete_data_merged$patient_data %>% filter(hospital_id == "Oriik")
 
@@
 ```{r}
-complete_data <- merge_hospitals(complete_data, hospitals_to_merge)
+complete_data <- complete_data_merged
+```
 
-complete_data$patient_data %>% filter(hospital_id == "Oriik")
+```{r}
 ```
@@
 
+Patient counts
+
+```{r}
+complete_data$patient_data %>% group_by(hospital_id) %>%
+  summarise(n())
+```
+
 ```{r}

### 2020-10-16 R/data_mapping.R

@@ -23,7 +23,6 @@ marker_map <- c(
         "creatinine" = "creatinine",
         "urea" = "urea",
         "IgG" = "IgG",
-        "ARO" = "ARO", # We know the patient was on anesthesia-resuscitation ward, but no details on breathing
       ## Covid-specific drugs
         "Hydroxychloroquine" = "hcq",
         "Kaletra" = "kaletra",

### 2020-10-16 R/data_preprocessing.R

@@ -19,6 +19,7 @@ read_data_for_analysis <- function() {
                              liver_disease = col_logical(),
                              smoking = col_logical(),
                              BMI = col_double(),
+                             obesity = col_character(),
                              NYHA = col_factor(levels = c(1,2,3,4), ordered = TRUE),
                              creatinin = col_double(),
                              pt_inr = col_double(),
@@ -31,15 +32,29 @@ read_data_for_analysis <- function() {
     mutate(age_norm = (age - 65) / 20,
            high_creatinin = if_else(sex == "M", creatinin > 115, creatinin > 97),
            high_pt_inr = pt_inr > 1.2,
-           low_albumin = albumin < 36)
+           low_albumin = albumin < 36,
+           obesity = factor(case_when(is.na(BMI) ~ obesity,
+                               BMI <= 25 ~ "No",
+                               BMI <= 30 ~ "Overweight",
+                               TRUE ~ "Obese")
+                            ,levels = c("No", "Overweight", "Obese"), ordered = TRUE)
+           )
 
+  collapse_breathing <- function(x) {
+    fct_collapse(x, Ventilated = c("NIPPV", "MV", "ECMO"))
+  }
   breathing_data <- read_csv(here::here("private_data", "breathing_data.csv"),
                            col_types = cols(
                             patient_id = col_character(),
                             hospital_id = col_character(),
                             day = col_integer(),
-                            breathing = col_factor(levels = disease_levels, ordered = TRUE)
-                          ))
+                            breathing = col_factor(levels = disease_levels, ordered = TRUE),
+                            breathing_low = col_factor(levels = disease_levels, ordered = TRUE),
+                            breathing_high = col_factor(levels = disease_levels, ordered = TRUE)
+                           )) %>%
+    mutate(breathing_s = collapse_breathing(breathing),
+           breathing_s_low = collapse_breathing(breathing_low),
+           breathing_s_high = collapse_breathing(breathing_low))
 
   marker_data <- read_csv(here::here("private_data", "marker_data.csv"),
                              col_types = cols(
[…]

### 2020-10-16 R/raw_data_preprocessing.R

@@ -6,6 +6,58 @@ read_raw_data <- function(file, hospital_id, lang, file_version, remove_examples
 
   progression_data <- read_progression_data(file, hospital_id, lang, file_version, patient_data)
 
+  if(file_version == "Special" && any(progression_data$marker_data$marker == "weight")) {
+    #Use Height/Weight from lab markers to calculate BMI
+    BMI_lab_data <-  progression_data$marker_data %>%
+      select(patient_id, marker, value, day) %>%
+      filter(marker %in% c("height", "weight")) %>%
+      pivot_wider(names_from = "marker", values_from = "value") %>%
+      group_by(patient_id) %>%
+      filter(day == min(day)) %>%
+      ungroup() %>%
+      mutate(BMI_lab = weight / (( height /100) ^ 2)) %>%
+      select(patient_id, BMI_lab)
+
+    nrow_before <- nrow(patient_data)
+
+    patient_data <- patient_data %>%
+      left_join(BMI_lab_data, by = "patient_id") %>%
+      mutate(BMI = if_else(!is.na(BMI_lab), BMI_lab, BMI)) %>%
+      select(-BMI_lab)
+
+    if(nrow(patient_data) != nrow_before) {
+      stop("Bad join")
+    }
+
+    progression_data$marker_data <-
+      progression_data$marker_data %>% filter(!(marker %in% c("height", "weight")))
+
+    # Use markers from first two days to impute markers at admission
+    markers_lab_data <- progression_data$marker_data %>%
+      select(patient_id, marker, value, day) %>%
+      filter(day >= 0, day <= 1, marker %in% c("pt_inr", "albumin", "creatinin")) %>%
+      pivot_wider(names_from = "marker", values_from = "value") %>%
+      group_by(patient_id) %>%
+      filter(day == min(day)) %>%
+      select(-day)
+
+    # Make sure all columns exist
+    for(m in c("pt_inr", "albumin", "creatinin")) {
+      if(is.null(markers_lab_data[[m]])) {
+        markers_lab_data[[m]] <- array(NA_real_, nrow(markers_lab_data))
+      }
+    }
+
+    patient_data <- patient_data %>%
+      select(-pt_inr, -albumin, -creatinin) %>% # I now those were not filled
+      left_join(markers_lab_data, by = "patient_id")
+
+    if(nrow(patient_data) != nrow_before) {
+      stop("Bad join")
+    }
+  }
+
+
   # TODO code adverse events
   loo::nlist(patient_data,
[…]

### 2020-10-16 manuscript/manuscript.Rmd
Further work on risk scores
Improved progression figure

@@
      toc: false
+  # bookdown::word_document2: 
+  #     toc: false
   # bookdown::pdf_book: 
@@
 
-sum_percent <- function(x) {
-  paste0(sum(x, na.rm = TRUE), " (", scales::percent(mean(x, na.rm = TRUE), accuracy = 1), ")")
-}
-
-mean_iqr <- function(x) {
-  paste0(round(mean(x, na.rm = TRUE)), " (", round(quantile(x, 0.25, na.rm = TRUE)), " - ", round(quantile(x, 0.75, na.rm = TRUE)), ")")  
-}
-
 data_table_1 %>% group_by(group) %>% 
@@
     N = n(),
-   # `Distinct sites` = length(unique(hospital_id)),
+    `Distinct sites` = length(unique(hospital_id)),
     Male = sum_percent(sex == "M"),
@@
     `Took Azithromycin` = sum_percent(ever_az),
+    `Took Dexamethasone` = sum_percent(ever_dexamethasone),
+    `Took Favipiravir` = sum_percent(ever_favipiravir),
+    `Took Remdesivir` = sum_percent(ever_remdesivir),
+    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
     `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
@@
 
-```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
-AA = Ambient air, Oxygen = Nasal oxygen, NIPPV = Non-invasive positive-pressure ventilation, ECMO = Extra-corporeal membrane oxygenation."}
+```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required."}
 my_paste <- function(x) {
@@
 }
-wide  %>% 
-  filter(!is.na(breathing)) %>%
+
+data_for_fig_progression <- wide  %>% 
+  filter(!is.na(breathing_s), day >= 0) %>%
   arrange(hospital_id, patient_id, day) %>%
-  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing,  .fun = my_paste, .desc = FALSE)) %>%
-  ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
+  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing_s,  .fun = my_paste, .desc = FALSE)) %>%
+  mutate(first_half = as.integer(patient_id) < length(levels(patient_id)) / 2 + 1, 
+         is_outcome =  breathing_s %in% c("Discharged", "Death"),
+         breathing_s = fct_recode(breathing_s, `Ambient air` = "AA")
+         )
+
+data_for_fig_progression %>% 
+  filter(!is_outcome) %>%
+  ggplot(aes(x = day, y = patient_id, fill = breathing_s)) + 
   geom_tile(height = 0.7) + 
+  geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing_s, ordered = FALSE)),
+             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome)) +
   scale_fill_viridis_d("", direction = -1) +
   scale_x_continuous("Days since hospitalization") +
-  theme(axis.text.y = element_blank()) +
-  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank())
+  scale_shape_manual("", values = c(1, 15) ) +
+  facet_wrap(~first_half, scales = "free_y") +
+  theme(axis.text.y = element_blank(), strip.background = element_blank(), strip.text = element_blank(),
+        legend.direction = "horizontal", legend.position = "bottom") +
+  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())
 
@@
 
+## Survey of antimicrobial treatment
+
 ## Evaluating prediction models

### 2020-10-17 manuscript/competing_risks.Rmd
Working with the new dataset

@@
 f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + hospital_id)
-fit_competing_hcq_az_hospital <- coxph(f_competing_hcq_az_hospital, data = data_surv, id = patient_id)
+data_surv_hospital <- data_surv %>% 
+  mutate(hospital_id = if_else(hospital_id %in% c("DJkhZ", "lRBHa", "nDINR"), "small", hospital_id))
+fit_competing_hcq_az_hospital <- coxph(f_competing_hcq_az_hospital, data = data_surv_hospital, id = patient_id)
 fit_competing_hcq_az_hospital
@@
 
-```
-```{r}
-hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = here::here("manuscript", "competing_risks_res.csv"))
 ```
@@
 
-```{r}
-hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
-```
-
 # Peak IL-6, peak D-dimer
@@
 # Lymphocyte count predicts?
+
+```{r}
+hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
+write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+```

### 2020-10-17 manuscript/manuscript.Rmd
Working with the new dataset

@@
   geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing_s, ordered = FALSE)),
-             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome)) +
+             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome), size = 0.9) +
   scale_fill_viridis_d("", direction = -1) +
   scale_x_continuous("Days since hospitalization") +
-  scale_shape_manual("", values = c(1, 15) ) +
+  scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
   facet_wrap(~first_half, scales = "free_y") +

### 2020-10-17 manuscript/risk_scores.Rmd
Working with the new dataset

@@
            (-6) * (age < 40) +
-           obesity +
+           is_obese +
            3 * renal_disease +
@@
 
-We see that the "age only" baseline actually outperforms all of the other scores - although there is some uncertainty prohibiting us to be very sure about it.
+We see that the "age only" baseline performs similarly to the best performing score (Caramelo et al.) all of the other scores - there is some uncertainty prohibiting us to be very sure about which works better.
 This holds even if we only take the decade of age.
-And the second best performing score (Caramelo et al.) is actually heavily based on the decade of age.
+Note that the score by (Caramelo et al.) is actually heavily based on the decade of age.
 Table of the results is shown below as the small differences are not well visible in the plot.

### 2020-10-17 R/data_mapping.R

@@ -197,6 +197,10 @@ unit_map <- c(
 breathing_levels <- c("AA","Oxygen", "NIPPV","MV","ECMO")
 disease_levels <- c("Discharged", breathing_levels, "Death")
 
+breathing_s_levels <- c("AA","Oxygen", "Ventilated")
+disease_s_levels <- c("Discharged", breathing_s_levels, "Death")
+
+
 unit_conversions <- list(
   list(markers = c(
     "cefotaxime","cefuroxime", "ampicillin_sulbactam", "amoxiclav",

### 2020-10-17 R/raw_data_preprocessing.R

@@ -125,6 +125,7 @@ read_patient_data <- function(file, hospital_id, lang, file_version, remove_exam
 
     patient_data_raw_usable$Obesity[special_match] <- "Overweight"
     patient_data_raw_usable$BMI <- BMI_new
+    patient_data_raw_usable$Obesity[is.na(patient_data_raw_usable$Obesity)] <- "No"
   }
 
   patient_data_typed <- patient_data_raw_usable %>%

### 2020-10-17 manuscript/analysis_brms.Rmd
Some suggestions for the brms analysis

@@
 # all patients for which we have data for at least 7 days
-data$surv_data_7 <- data$surv_data_last_day %>%
-  filter(!(day < 7 & outcome3 == "Hospitalized"))
+# data$surv_data_7 <- data$surv_data %>%
+#   group_by(patient_id) %>%
+#   filter(!(max(day) < 7 & outcome3 == "Hospitalized")) %>%
+#   ungroup()
+#  
+# 
+# # all patients for which we have data for at least 7 days
+# data$surv_data_28 <- data$surv_data_last_day %>%
+#   filter(!(day < 28 & outcome3 == "Hospitalized"))
 
-# all patients for which we have data for at least 7 days
-data$surv_data_28 <- data$surv_data_last_day %>%
-  filter(!(day < 28 & outcome3 == "Hospitalized"))
+# Note: the code above can IMHO introduce biases if there is some systematic skew that makes patients not have a final outcome after day 7 if they are more/less likely to have good outcomes.
+Instead, I _think_ we need to have "state at day 7", i.e.
+patients that died/were discharged after day 7 are still treated as "Hospitalized" for the analysis.
+Additionally, other studies (e.g.
+RECOVERY) work this way.
+This is how I would code this:
+
+surv_data_at_day <- function(patient_data, day) {
+  res <- patient_data %>%
+    filter(!(last_record < day & outcome3 == "Hospitalized"))
+
+  #Non tidy to avoid hassle with factor levels  
+  res$outcome3[res$last_record > day] <- "Hospitalized"
+  res
+}
+
+data$surv_data_7 <- surv_data_at_day(data$surv_data, 7)
+data$surv_data_28 <- surv_data_at_day(data$surv_data, 28)
 ```
@@
   prior = prior_survival_base,
-  file = "models/fit_survival_base"
+  file = paste0(cache_dir, "/fit_survival_base")
 )
@@
   prior = prior_survival_base,
-  file = "models/fit_survival_base_7"
+  file = paste0(cache_dir, "/fit_survival_base_7")
 )
@@
   prior = prior_survival_base,
-  file = "models/fit_survival_base_28"
+  file = paste0(cache_dir, "/fit_survival_base_28")
 )
@@
   inits = 0,
-  file = "models/fit_survival_time"
+  file = paste0(cache_dir, "/fit_survival_time")
 )
@@
   prior = prior_discharged_base,
-  file = "models/fit_discharged_base"
+  file = paste0(cache_dir, "/fit_discharged_base")
 )
@@
   prior = prior_discharged_base,
-  file = "models/fit_discharged_base_7"
+  file = paste0(cache_dir, "/fit_discharged_base_7")
 )
@@
   prior = prior_discharged_base,
-  file = "models/fit_discharged_base_28"
+  file = paste0(cache_dir, "/fit_discharged_base_28")
 )
@@
   control = list(adapt_delta = 0.95),
-  file = "models/fit_discharged_time"
+  file = paste0(cache_dir, "/fit_discharged_time")
 )
@@
   prior = prior_outcome_base,
-  file = "models/fit_outcome_base"
+  file = paste0(cache_dir, "/fit_outcome_base")
 )
@@
   prior = prior_outcome_base,
-  file = "models/fit_outcome_base_7"
+  file = paste0(cache_dir, "/fit_outcome_base_7")
 )
@@
   prior = prior_outcome_base,
-  file = "models/fit_outcome_base_28"
+  file = paste0(cache_dir, "/fit_outcome_base_28")
 )
@@
   prior = prior_worst_breathing_base,
-  file = "models/fit_worst_breathing_base"
+  file = paste0(cache_dir, "/fit_worst_breathing_base")
 )

### 2020-10-17 dev/hmm_model_devel.Rmd
Optimize the HMM model by avoiding computation for impossible states

@@
 ```{r}
+
 adapt_delta <- 0.9
 
-data <- hmm_simulator(5, 10, 2, use_noisy_states = TRUE)
+data <- hmm_simulator(N_series = 10, N_time = 10, N_mid_states = 2, use_noisy_states = FALSE)
+standata <- make_standata_hmm(data$observed)
+
+#Temporary hack
+registerS3method("stan_log_lik", class = "rate_hmm", method = stan_log_lik.rate_hmm, envir = asNamespace("brms"))
 
@@
 
-model_single <- stan_model(model_code = single_code)
-
-standata <- make_standata_hmm(data$observed)
+model_file <- here::here("local_temp_data", "devel_single_model.stan")
+write_lines(single_code, path = model_file)
+model_single <- cmdstanr::cmdstan_model(model_file)
 
@@
 } else {
+  standata$optimize_possible = 0
   cmdstan_fit <- model_single$sample(data = standata, adapt_delta = adapt_delta)

### 2020-10-17 R/make_standata_hmm.R

@@ -131,7 +131,8 @@ make_data_hmm <- function(brmshmmdata) {
     serie_max_time,
     obs_states_rect,
     predictor_sets_rect,
-    rate_predictors
+    rate_predictors,
+    optimize_possible = d$optimize_possible
   )
 
   loo::nlist(standata, brmsdata)

### 2020-10-18 dev/hmm_model_devel.Rmd
One more level of optimization

--- 2020-05-21 dev/hmm_model_devel.Rmd
@@
 ```{r}
+
 adapt_delta <- 0.9
 
-data <- hmm_simulator(5, 10, 2, use_noisy_states = TRUE)
+data <- hmm_simulator(N_series = 10, N_time = 12, N_mid_states = 4, use_noisy_states = FALSE)
+standata <- make_standata_hmm(data$observed)
+
+#Temporary hack
+registerS3method("stan_log_lik", class = "rate_hmm", method = stan_log_lik.rate_hmm, envir = asNamespace("brms"))
 

### 2020-10-18 R/brmshmmdata.R

@@ -1,8 +1,12 @@
+#' @param  optimize_possible Let's the algorithm optimize for models where some
+#' hidden states are impossible, given an observation. The optimization
+#' some overhead and so can be turned of. Takes values 0 - no such optimalization),
+#'  1 - avoid redundant rows in matrix to multiply and 2 - also avoid some multiplications.
 brmshmmdata <- function(formula,
                         serie_data, rate_data,
                         hidden_state_data, initial_states,
                         observed_state_data = NULL, prior = NULL,
-                        sensitivity_low_bound = 0.5, optimize_possible = TRUE) {
+                        sensitivity_low_bound = 0.5, optimize_possible = 2) {
 
   validate_brmshmmdata(
     structure(loo::nlist(

### 2020-10-18 R/make_stancode_hmm.R

@@ -125,7 +125,7 @@ rate_hmm_stanvars_data  <- function(standata) {
 
   brms::stanvar(x = standata$predictor_sets_rect, name = "predictor_sets_rect", scode = "  int<lower=0, upper=N_predictor_sets> predictor_sets_rect[N_series, N_time];", block = "data") +
   brms::stanvar(x = standata$rate_predictors, name = "rate_predictors", scode = "  int<lower=1, upper=N> rate_predictors[N_predictor_sets, N_rates];", block = "data") +
-  brms::stanvar(x = standata$optimize_possible, name = "optimize_possible", scode = "  int<lower=0, upper=1> optimize_possible;", block = "data")
+  brms::stanvar(x = standata$optimize_possible, name = "optimize_possible", scode = "  int<lower=0, upper=2> optimize_possible;", block = "data")
 }
 
 rate_hmm_tdata_code <- '
@@ -183,8 +183,6 @@ rate_hmm_parameters_code <- "
   simplex[N_other_observations] other_observations_probs[N_noisy_states];
 "
 
-##' @export stan_log_lik
-##' @method stan_log_lik rate_hmm
 stan_log_lik.rate_hmm <- function( ... ) {
 "
   matrix[N_states_hidden, N_states_observed] observation_probs = compute_observation_matrix(
@@ -211,6 +209,7 @@ stan_log_lik.rate_hmm <- function( ... ) {
     vector[serie_max_time[s]] alpha_log_norms;
     alpha[initial_states[s], 1] = observation_probs[initial_states[s], obs_states_rect[s, 1]];
 
+
     {
       real norm = max(alpha[, 1]);
       alpha[, 1] /= norm;
@@ -221,16 +220,31 @@ stan_log_lik.rate_hmm <- function( ... ) {
       int ps = predictor_sets_rect[s, t - 1];
       int obs = obs_states_rect[s,t];
       if(obs != 0) {
-        //Oberved something. Only update states that are possible
+        //Oberved something.
         int N_possible = N_possible_states[obs];
         if(!optimize_possible || N_possible == N_states_hidden) {
           vector[N_states_hidden] transition_probs =
             transition_matrices_t[ps] * alpha[, t - 1];
           alpha[, t] = observation_probs[, obs] .* transition_probs;
         } else {
-          for(p_state_id in 1:N_possible) {
-             int p_state = possible_states[obs, p_state_id];
[…]

### 2020-10-19 manuscript/analysis_brms.Rmd
update brms analysis

@@
 ```{r}
-# all patients for which we have data for at least 7 days
-# data$surv_data_7 <- data$surv_data %>%
-#   group_by(patient_id) %>%
-#   filter(!(max(day) < 7 & outcome3 == "Hospitalized")) %>%
-#   ungroup()
-#  
-# 
-# # all patients for which we have data for at least 7 days
-# data$surv_data_28 <- data$surv_data_last_day %>%
-#   filter(!(day < 28 & outcome3 == "Hospitalized"))
-
-# Note: the code above can IMHO introduce biases if there is some systematic skew that makes patients not have a final outcome after day 7 if they are more/less likely to have good outcomes.
-Instead, I _think_ we need to have "state at day 7", i.e.
-patients that died/were discharged after day 7 are still treated as "Hospitalized" for the analysis.
-Additionally, other studies (e.g.
-RECOVERY) work this way.
-This is how I would code this:
-
+# we need to have "state at day 7", i.e. patients that died/were discharged 
+# after day 7 are still treated as "Hospitalized" for the analysis. Other 
+# studies (e.g. RECOVERY) work this way. Code provided by Martin.
 surv_data_at_day <- function(patient_data, day) {
@@
 
-  #Non tidy to avoid hassle with factor levels  
+  # Non tidy to avoid hassle with factor levels  
   res$outcome3[res$last_record > day] <- "Hospitalized"
@@
 
-- first and last day invasive: Martin will check
 - days_from_symptom_onset. Days from symptom onset until hospitalization. 
@@
   survival ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az +
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
     hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, sex, age_norm, smoking, BMI"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
   ),
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary_28",
+    estimand = "logit(survival)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
 ```
-
-### TODOs
-
-- model varying slopes?
-- model interactions?
-- use horseshoe prior
-- use projpred
-- add ever_keltara as soon as there is a lest one patient who took it.
 
@@
   day | cens(censored) ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az + 
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
 ```
-
-### TODOs:
-
-- How does that relate to days of hospitalization? (Answer: could be analysed;
-could be an interesting end point when people are discharged)
 
@@
   discharged ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az +
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
     hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, sex, age_norm, smoking, BMI"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
   ),
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_binary_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
   day | cens(censored_discharged) ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az + 
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiraviTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravi_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
   outcome3 ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az +
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
 hypothesis_res_list[["hcq_az_outcome_base"]] <- rbind(
+  # Death 
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
   ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_favipiravirTRUE,
+    model = "brms_categorical",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+  ),
+  # Discharged
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    model = "brms_categorical",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
 hypothesis_res_list[["hcq_az_outcome_base_7"]] <- rbind(
+  # Death 
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
   ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_favipiravirTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+  ),
+  # Discharged
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
 hypothesis_res_list[["hcq_az_outcome_base_28"]] <- rbind(
+  # Death
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
   ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_favipiravirTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$favipiravir_reduces_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+  ),
+  # Discharged
   bayesian_hypothesis_res_from_draws(
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, sex, age_norm, smoking, BMI"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_increases_discharged,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
   )
@@
   worst_breathing ~ sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az +
+    smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir +
     (1 | hospital_id),
@@
 ```{r, results="hide"}
-form_worst_breathing_cs <- bf(
-  worst_breathing ~ cs(sex + age_norm + 
-    smoking + BMI + ever_hcq + ever_az) +
-    (1 | hospital_id),
-  family = sratio("logit")
-)
-
-prior_worst_breathing_cs <- 
-  prior(normal(0, 5), class = "b") +
-  prior(normal(0, 5), class = "sd")
-
-fit_worst_breathing_cs <- brm(
-  formula = form_worst_breathing_cs,
-  data = data$patient_data,
-  prior = prior_worst_breathing_cs,
-  inits = 0,
-  file = "models/fit_worst_breathing_cs"
-)
-fit_worst_breathing_cs <- add_criterion(fit_worst_breathing_cs, "loo")
+# form_worst_breathing_cs <- bf(
+#   worst_breathing ~ cs(sex + age_norm + 
+#     smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir) +
+#     (1 | hospital_id),
+#   family = sratio("logit")
+# )
+# 
+# prior_worst_breathing_cs <- 
+#   prior(normal(0, 5), class = "b") +
+#   prior(normal(0, 5), class = "sd")
+# 
+# fit_worst_breathing_cs <- brm(
+#   formula = form_worst_breathing_cs,
+#   data = data$patient_data,
+#   prior = prior_worst_breathing_cs,
+#   inits = 0,
+#   file = paste0(cache_dir, "/fit_worst_breathing_cs")
+# )
+# fit_worst_breathing_cs <- add_criterion(fit_worst_breathing_cs, "loo")
 ```
@@
 ```{r}
-summary(fit_worst_breathing_cs)
+# summary(fit_worst_breathing_cs)
 # ce <- conditional_effects(fit_worst_breathing_cs, categorical = TRUE)
@@
 ```{r}
-loo(fit_worst_breathing_base, fit_worst_breathing_cs)
+# loo(fit_worst_breathing_base, fit_worst_breathing_cs)
 ```
 
-### TODOs
-
-- How to handle missing values in `worst_breathing`?
-- decide which predictors to include 
-- model interactions?
-- model varying slopes?
-- use horseshoe or similar sparsity priors?
+## Store Hypotheses
 

### 2020-10-22 manuscript/hmm.Rmd
Some work on HMM models and associated tools

--- 2020-09-13 manuscript/hmm.Rmd
@@
 rstan::rstan_options(auto_write = TRUE)
+options(brms.backend = "cmdstanr")
 
@@
 data <- read_data_for_analysis()
+
+if(length(unique(data$patient_data$patient_id)) != length(data$patient_data$patient_id)) {
+  stop("FAil")
+}
+# warning("Subsetting data")
+# set.seed(4568224)
+# patients_to_use <- sample(data$patient_data$patient_id, size = 70)
+# data <- filter_complete_data(data, patient_id %in% patients_to_use)
+
 wide <- data$marker_data_wide
@@
 
+do_pp_checks <- FALSE
+hmm_hypothesis_res_list <- list()
+
 #Temporary hack
-registerS3method("stan_llh", class = "rate_hmm", method = stan_llh.rate_hmm, envir = asNamespace("brms"))
+registerS3method("stan_log_lik", class = "rate_hmm", method = stan_log_lik.rate_hmm, envir = asNamespace("brms"))
 ```
@@
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
-  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
   mutate(.time = .time +1) %>% 
@@
 
+serie_data_28 <- wide %>%
+  inner_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day")) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
+  mutate(.time = .time +1) %>% 
+  filter(.time >= 1) %>%
+  group_by(.serie) %>%
+  filter(sum(!is.na(.observed)) > 1) %>%
+  ungroup()
+
+# Extend terminal states for serie_data_28
+serie_data_28 <- serie_data_28 %>%
+  group_by(.serie) %>%
+  mutate(last_day = max(day[!is.na(.observed)]),
+         last_state = .observed[day == last_day],
+         .observed = if_else(last_state %in% c("Discharged", "Death") & is.na(.observed) & day > last_day,
+                             last_state,
+                             .observed)) %>%
+  ungroup()
+
+base_observed_state_data <- tibble(id = factor(disease_s_levels, levels = disease_s_levels), is_noisy = FALSE)
+
+simple_hidden_state_data <- tibble(id = factor(disease_s_levels, levels = disease_s_levels)) %>% mutate( corresponding_obs = id)
+
+simple_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 ```
@@
 ```{r}
-data$patient_data %>% filter(!is.na(best_supportive_care_from)) %>% select(outcome)
+#data$patient_data %>% filter(!is.na(best_supportive_care_from)) %>% select(outcome)
 ```
@@
 ```{r}
-hcq_m1_hidden_state_data <- tibble(id = factor(disease_levels, levels = disease_levels)) %>% mutate( corresponding_obs = id)
-
-hcq_m1_observed_state_data <- tibble(id = factor(disease_levels, levels = disease_levels), is_noisy = FALSE)
-
 recode_group <- function(rate_data, group_name) {
@@
 
-hcq_m1_rate_data <- rbind(
-  tibble(.from = breathing_levels, .to = "Death", rate_group = "death"),
+```
+
+```{r trt_m1}
+trt_m1_hidden_state_data <- simple_hidden_state_data
+
+trt_m1_observed_state_data <- base_observed_state_data
+
+trt_m1_rate_data <- rbind(
+  tibble(.from = breathing_s_levels, .to = "Death", rate_group = "death"),
   tibble(.from = "AA", .to = "Discharged", rate_group = "improve_one"),
-  tibble(.from = breathing_levels[2 : length(breathing_levels)], .to = breathing_levels[1 : (length(breathing_levels) - 1)], rate_group = "improve_one"),
-  tibble(.from = breathing_levels[1 : (length(breathing_levels) - 1)], .to = breathing_levels[2 : length(breathing_levels)], rate_group = "worsen_one")
+  tibble(.from = breathing_s_levels[2 : length(breathing_s_levels)], .to = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], rate_group = "improve_one"),
+  tibble(.from = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], .to = breathing_s_levels[2 : length(breathing_s_levels)], rate_group = "worsen_one")
 ) %>%
   mutate(.rate_id = factor(1:n()),
-         .from = factor(.from, levels = disease_levels),
-         .to = factor(.to, levels = disease_levels)
+         .from = factor(.from, levels = disease_s_levels),
+         .to = factor(.to, levels = disease_s_levels)
   ) %>%
@@
   recode_group("improve_one") %>%
-  recode_group("worsen_one")
+  recode_group("worsen_one") %>%
+  mutate(rate_group = fct_relevel(rate_group, "improve_one"))
 
-hcq_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+trt_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
+trt_m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
   brms::set_prior("normal(0, 5)", "b") +
@@
 
-hcq_m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group
-hcq_m1 <- brmshmmdata(hcq_m1_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+trt_m1_formula <- ~ is_worsen_one +  mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + (took_hcq + took_az+ took_favipiravir + took_convalescent_plasma | rate_group)
 
-#make_stancode_hmm(hcq_m1)
-#xx <- brms::make_standata(hcq_m1$formula, data = make_data_hmm(hcq_m1)$brmsdata)
-hcq_m1_fit <- brmhmm(hcq_m1, cache_file = paste0(cache_dir,"/hcq_m1.rds"))
+# trt_m1_formula <- ~ rate_group +  mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+#     (0 + is_improve_one | improve_one_id) + (0 + best_supportive_care + took_hcq + took_az+ took_favipiravir + took_convalescent_plasma | .rate_id) + ((best_supportive_care + took_hcq + took_az+ took_favipiravir + took_convalescent_plasma)* rate_group)
 
-summary(hcq_m1_fit$brmsfit)
+trt_m1 <- brmshmmdata(trt_m1_formula, serie_data, 
+                  trt_m1_rate_data, simple_hidden_state_data, simple_initial_states, prior = trt_m1_prior,
+                  observed_state_data = base_observed_state_data)
 
+#make_stancode_hmm(trt_m1)
+#xx <- brms::make_standata(trt_m1$formula, data = make_data_hmm(trt_m1)$brmsdata)
+trt_m1_fit <- brmhmm(trt_m1, cache_file = paste0(cache_dir,"/trt_m1.rds"))
+
+# Test code for running with reduce_sum. Achieves around 1500s instead of 2000s while using the full CPU...
+# Will possibly get worse as more effort goes to the matrix exp than to the HMM part.
+# test_code <- make_stancode_hmm(trt_m1)
+# 
+# test_model_file <- here::here("local_temp_data", "test_trt_m1.stan")
+# write_lines(test_code, path = test_model_file)
+# model_test <- cmdstanr::cmdstan_model(test_model_file, cpp_options = list(stan_threads = TRUE))
+# 
+# 
+# model_test$sample(data = make_standata_hmm(trt_m1), parallel_chains = 4, threads_per_chain = 2)
+
+summary(trt_m1_fit$brmsfit)
+bayesplot::mcmc_pairs(trt_m1_fit$brmsfit, regex_pars = "b_rate")  
 ```
+
 ```{r}
 
-counterfactual_max_days <- 90
-counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
-  inner_join(data$patient_data, by = c("patient_id")) %>% 
-  left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
-  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
-  mutate(.time = .time +1) %>% 
-  filter(.time >= 1) %>%
-  group_by(.serie) %>%
-  mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
-  ungroup()
-
-counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
-counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
-
-hcq_m1_counterfactual_took_hcq <- hcq_m1
-hcq_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
-
-hcq_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_took_hcq)
-
-hcq_m1_counterfactual_no_hcq <- hcq_m1
-hcq_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
-
-hcq_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_no_hcq)
+# counterfactual_max_days <- 90
+# counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
+#   inner_join(data$patient_data, by = c("patient_id")) %>% 
+#   left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
+#   mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+#   mutate(.time = .time +1) %>% 
+#   filter(.time >= 1) %>%
+#   group_by(.serie) %>%
+#   mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
+#   ungroup()
+# 
+# 
+# counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
+# counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
+# 
+# 
+# 
+# trt_m1_counterfactual_took_hcq <- trt_m1
+# trt_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
+# 
+# trt_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(trt_m1_fit, newdata = trt_m1_counterfactual_took_hcq)
+# 
+# trt_m1_counterfactual_no_hcq <- trt_m1
+# trt_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
+# 
+# trt_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(trt_m1_fit, newdata = trt_m1_counterfactual_no_hcq)
 
@@
 ```{r}
-hcq_m1_epred_rect <- posterior_epred_rect(hcq_m1_fit)
-hcq_m1_predicted_rect <- posterior_epred_to_predicted(hcq_m1_fit, hcq_m1_epred_rect)
-hcq_m1_predicted <- posterior_rect_to_long(hcq_m1_fit, hcq_m1_predicted_rect)
+draws <- tidybayes::tidy_draws(trt_m1_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m1_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "supportive"
+  )
+
+hmm_hypothesis_res_list[["trt_m1_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "supportive"
+  )
 ```
@@
 ```{r}
-hcq_m1_predicted_df <- posterior_long_to_df(hcq_m1_fit$data, hcq_m1_predicted)
+system.time(posterior_epred_rect(trt_m1_fit))
+system.time(posterior_epred_rect(trt_m1_fit, nsamples = 2))
+```
 
-all_ids <- hcq_m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+```{r}
+if(do_pp_checks) {
+  trt_m1_epred_rect <- posterior_epred_rect(trt_m1_fit)
+  trt_m1_predicted_rect <- posterior_epred_to_predicted(trt_m1_fit, trt_m1_epred_rect)
+  trt_m1_predicted <- posterior_rect_to_long(trt_m1_fit, trt_m1_predicted_rect)
+}
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m1_predicted_df <- posterior_long_to_df(trt_m1_fit$data, trt_m1_predicted)
+
+  all_ids <- trt_m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m1_predicted_df, all_ids[series_id], hcq_m1_fit$data) %>% print()
+    posterior_state_plot(trt_m1_predicted_df, all_ids[series_id], trt_m1_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m1_fit, hcq_m1_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect)
-pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m1_fit, trt_m1_predicted_rect) %>% print() #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m1_fit, trt_m1_predicted_rect) %>% print()
+  pp_check_transitions_direction(trt_m1_fit, trt_m1_predicted_rect, states_from = 2:6)  %>% print()
+}
 ```
@@
 ```{r}
-hcq_m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
-hcq_m2 <- brmshmmdata(hcq_m2_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+simple_rate_data <- rbind(
+  tibble(.from = breathing_s_levels, .to = "Death", rate_group = "death"),
+  tibble(.from = "AA", .to = "Discharged", rate_group = "not_death"),
+  tibble(.from = breathing_s_levels[2 : length(breathing_s_levels)], .to = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], rate_group = "not_death"),
+  tibble(.from = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], .to = breathing_s_levels[2 : length(breathing_s_levels)], rate_group = "not_death")
+) %>%
+  mutate(.rate_id = factor(1:n()),
+         .from = factor(.from, levels = disease_s_levels),
+         .to = factor(.to, levels = disease_s_levels)
+  ) %>%
+  recode_group("death") %>%
+  recode_group("not_death")
 
-hcq_m2_fit <- brmhmm(hcq_m2, cache_file = paste0(cache_dir,"/hcq_m2.rds"))
+base_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
+  brms::set_prior("normal(0, 5)", "b") +
+  brms::set_prior("normal(0, 2)", "sd")
 
-summary(hcq_m2_fit$brmsfit)
+simple_m1_formula <- ~ mo(death_id) + (0 + is_not_death | not_death_id) + 
+     + best_supportive_care * .rate_id + took_hcq * .rate_id
+simple_m1 <- brmshmmdata(simple_m1_formula, serie_data, 
+                  simple_rate_data, simple_hidden_state_data, simple_initial_states, prior = base_prior,
+                  observed_state_data = base_observed_state_data)
+
+#make_stancode_hmm(trt_m1)
+#xx <- brms::make_standata(trt_m1$formula, data = make_data_hmm(trt_m1)$brmsdata)
+simple_m1_fit <- brmhmm(simple_m1, cache_file = paste0(cache_dir,"/simple_m1.rds"))
+
+summary(simple_m1_fit$brmsfit)
+
 ```
@@
 ```{r}
-hcq_m2_epred_rect <- posterior_epred_rect(hcq_m2_fit)
-hcq_m2_predicted_rect <- posterior_epred_to_predicted(hcq_m2_fit, hcq_m2_epred_rect)
-hcq_m2_predicted <- posterior_rect_to_long(hcq_m2_fit, hcq_m2_predicted_rect)
+if(do_pp_checks) {
+  simple_m1_epred_rect <- posterior_epred_rect(simple_m1_fit)
+  simple_m1_predicted_rect <- posterior_epred_to_predicted(simple_m1_fit, simple_m1_epred_rect)
+  simple_m1_predicted <- posterior_rect_to_long(simple_m1_fit, simple_m1_predicted_rect)
+}
 ```
@@
 ```{r}
-hcq_m2_predicted_df <- posterior_long_to_df(hcq_m2_fit$data, hcq_m2_predicted)
+if(do_pp_checks) {
+  simple_m1_predicted_df <- posterior_long_to_df(simple_m1_fit$data, simple_m1_predicted)
 
-all_ids <- hcq_m2_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+  all_ids <- simple_m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m2_predicted_df, all_ids[series_id], hcq_m2_fit$data) %>% print()
+    posterior_state_plot(simple_m1_predicted_df, all_ids[series_id], simple_m1_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m2_fit, hcq_m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect)
-pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(simple_m1_fit, simple_m1_predicted_rect) %>% print()#TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+
+  pp_check_transitions_direction(simple_m1_fit, simple_m1_predicted_rect) %>% print()
+  pp_check_transitions_direction(simple_m1_fit, simple_m1_predicted_rect, states_from = 2:4) %>% print()
+}
 ```
@@
 ```{r}
-hcq_m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m3 <- brmshmmdata(hcq_m3_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+if(do_pp_checks) {
 
-hcq_m3_fit <- brmhmm(hcq_m3, cache_file = paste0(cache_dir,"/hcq_m3.rds"))
+  simple_m1_28 <- simple_m1
+  simple_m1_28$serie_data = serie_data_28
+  simple_m1_28 <- validate_brmshmmdata(simple_m1_28)
+  simple_m1_epred_rect_28 <- posterior_epred_rect(simple_m1_fit, newdata = simple_m1_28)
+  simple_m1_predicted_rect_28 <- posterior_epred_to_predicted(simple_m1_fit, simple_m1_epred_rect_28)
 
-summary(hcq_m3_fit$brmsfit)
-```
-```{r}
-hcq_m3_epred_rect <- posterior_epred_rect(hcq_m3_fit)
-hcq_m3_predicted_rect <- posterior_epred_to_predicted(hcq_m3_fit, hcq_m3_epred_rect)
-hcq_m3_predicted <- posterior_rect_to_long(hcq_m3_fit, hcq_m3_predicted_rect)
+  pp_check_state_at(simple_m1_fit, simple_m1_predicted_rect_28, 7, newdata = simple_m1_28) %>% print()
+  pp_check_state_at(simple_m1_fit, simple_m1_predicted_rect_28, 28, newdata = simple_m1_28)  %>% print()
+}
 ```
@@
 ```{r}
-hcq_m3_predicted_df <- posterior_long_to_df(hcq_m3_fit$data, hcq_m3_predicted)
+trt_m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
+trt_m2 <- brmshmmdata(trt_m2_formula, serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m1_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
 
-all_ids <- hcq_m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+trt_m2_fit <- brmhmm(trt_m2, cache_file = paste0(cache_dir,"/trt_m2.rds"))
+
+summary(trt_m2_fit$brmsfit)
+```
+
+```{r}
+draws <- tidybayes::tidy_draws(trt_m2_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m2_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive"
+  )
+
+hmm_hypothesis_res_list[["trt_m2_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive"
+  )
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m2_epred_rect <- posterior_epred_rect(trt_m2_fit)
+  trt_m2_predicted_rect <- posterior_epred_to_predicted(trt_m2_fit, trt_m2_epred_rect)
+  trt_m2_predicted <- posterior_rect_to_long(trt_m2_fit, trt_m2_predicted_rect)
+}
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m2_predicted_df <- posterior_long_to_df(trt_m2_fit$data, trt_m2_predicted)
+
+  all_ids <- trt_m2_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m3_predicted_df, all_ids[series_id], hcq_m3_fit$data) %>% print()
+    posterior_state_plot(trt_m2_predicted_df, all_ids[series_id], trt_m2_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m3_fit, hcq_m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect)
-pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m2_fit, trt_m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m2_fit, trt_m2_predicted_rect)
+  pp_check_transitions_direction(trt_m2_fit, trt_m2_predicted_rect, states_from = 2:4)
+}
 ```
@@
 ```{r}
-hcq_m4_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m4 <- brmshmmdata(hcq_m4_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+trt_m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
+trt_m3 <- brmshmmdata(trt_m3_formula, serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m1_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
 
-hcq_m4_fit <- brmhmm(hcq_m4, cache_file = paste0(cache_dir,"/hcq_m4.rds"))
+trt_m3_fit <- brmhmm(trt_m3, cache_file = paste0(cache_dir,"/trt_m3.rds"))
 
-summary(hcq_m4_fit$brmsfit)
+summary(trt_m3_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m4_epred_rect <- posterior_epred_rect(hcq_m4_fit)
-hcq_m4_predicted_rect <- posterior_epred_to_predicted(hcq_m4_fit, hcq_m4_epred_rect)
-hcq_m4_predicted <- posterior_rect_to_long(hcq_m4_fit, hcq_m4_predicted_rect)
+draws <- tidybayes::tidy_draws(trt_m3_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m3_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital"
+  )
+
+hmm_hypothesis_res_list[["trt_m3_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital"
+  )
 ```
@@
 ```{r}
-hcq_m4_predicted_df <- posterior_long_to_df(hcq_m4_fit$data, hcq_m4_predicted)
+if(do_pp_checks) {
+  trt_m3_epred_rect <- posterior_epred_rect(trt_m3_fit)
+  trt_m3_predicted_rect <- posterior_epred_to_predicted(trt_m3_fit, trt_m3_epred_rect)
+  trt_m3_predicted <- posterior_rect_to_long(trt_m3_fit, trt_m3_predicted_rect)
+}
+```
 
-all_ids <- hcq_m4_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+```{r}
+if(do_pp_checks) {
+  trt_m3_predicted_df <- posterior_long_to_df(trt_m3_fit$data, trt_m3_predicted)
+
+  all_ids <- trt_m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m4_predicted_df, all_ids[series_id], hcq_m4_fit$data) %>% print()
+    posterior_state_plot(trt_m3_predicted_df, all_ids[series_id], trt_m3_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m4_fit, hcq_m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect)
-pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect, states_from = 2:6)
-```
-
-## hcq_m5 -
-
-```{r}
-hcq_m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + comorbidities_sum * rate_group #mo(comorbidities_sum) * rate_group
-hcq_m5 <- brmshmmdata(hcq_m5_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
-
-hcq_m5_fit <- brmhmm(hcq_m5, cache_file = paste0(cache_dir,"/hcq_m5.rds"))
-
-summary(hcq_m5_fit$brmsfit)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m3_fit, trt_m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m3_fit, trt_m3_predicted_rect)
+  pp_check_transitions_direction(trt_m3_fit, trt_m3_predicted_rect, states_from = 2:4)
+}
 ```
@@
 ```{r}
-hcq_m5_epred_rect <- posterior_epred_rect(hcq_m5_fit)
-hcq_m5_predicted_rect <- posterior_epred_to_predicted(hcq_m5_fit, hcq_m5_epred_rect)
-hcq_m5_predicted <- posterior_rect_to_long(hcq_m5_fit, hcq_m5_predicted_rect)
+trt_m4_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + took_favipiravir * rate_group + took_convalescent_plasma * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
+trt_m4 <- brmshmmdata(trt_m4_formula, serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m1_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
+
+trt_m4_fit <- brmhmm(trt_m4, cache_file = paste0(cache_dir,"/trt_m4.rds"))
+
+summary(trt_m4_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m5_predicted_df <- posterior_long_to_df(hcq_m5_fit$data, hcq_m5_predicted)
+draws <- tidybayes::tidy_draws(trt_m4_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m4_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital, az"
+  )
 
-all_ids <- hcq_m5_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+hmm_hypothesis_res_list[["trt_m4_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, az"
+  )
+
+hmm_hypothesis_res_list[["trt_m4_death_az"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_azTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_reduces_death,
+    adjusted = "age, sex, supportive, hospital, hcq"
+  )
+
+hmm_hypothesis_res_list[["trt_m4_discharged_az"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_azTRUE + draws$`b_rate_groupimprove_one:took_azTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, hcq"
+  )
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m4_epred_rect <- posterior_epred_rect(trt_m4_fit)
+  trt_m4_predicted_rect <- posterior_epred_to_predicted(trt_m4_fit, trt_m4_epred_rect)
+  trt_m4_predicted <- posterior_rect_to_long(trt_m4_fit, trt_m4_predicted_rect)
+}
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m4_predicted_df <- posterior_long_to_df(trt_m4_fit$data, trt_m4_predicted)
+
+  all_ids <- trt_m4_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m5_predicted_df, all_ids[series_id], hcq_m5_fit$data) %>% print()
+    posterior_state_plot(trt_m4_predicted_df, all_ids[series_id], trt_m4_fit$data) %>% print()
 
+  }
 }  
@@
 ```{r}
-pp_check_last_state(hcq_m5_fit, hcq_m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect)
-pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m4_fit, trt_m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m4_fit, trt_m4_predicted_rect)
+  pp_check_transitions_direction(trt_m4_fit, trt_m4_predicted_rect, states_from = 2:4)
+}
 ```
 
-## hcq_m6 - D-dimer
+## trt_m5 -
 
 ```{r}
-hcq_m6_serie_data <- serie_data %>% 
+trt_m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + comorbidities_sum * rate_group #mo(comorbidities_sum) * rate_group
+trt_m5 <- brmshmmdata(trt_m5_formula, serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m1_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
+
+trt_m5_fit <- brmhmm(trt_m5, cache_file = paste0(cache_dir,"/trt_m5.rds"))
+
+summary(trt_m5_fit$brmsfit)
+```
+
+```{r}
+draws <- tidybayes::tidy_draws(trt_m5_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m5_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_reduces_death,
+    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+  )
+
+hmm_hypothesis_res_list[["trt_m5_discharged"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_increases_discharged,
+    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+  )
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m5_epred_rect <- posterior_epred_rect(trt_m5_fit)
+  trt_m5_predicted_rect <- posterior_epred_to_predicted(trt_m5_fit, trt_m5_epred_rect)
+  trt_m5_predicted <- posterior_rect_to_long(trt_m5_fit, trt_m5_predicted_rect)
+}
+```
+
+```{r}
+if(do_pp_checks) {
+  trt_m5_predicted_df <- posterior_long_to_df(trt_m5_fit$data, trt_m5_predicted)
+
+  all_ids <- trt_m5_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+    distinct() %>% arrange(hospital_id) %>% pull(.serie)
+  step_size <- 6
+  for(step in 1:ceiling(length(all_ids) / step_size)) {
+    series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+
+    posterior_state_plot(trt_m5_predicted_df, all_ids[series_id], trt_m5_fit$data) %>% print()
+
+  }
+}
+```
+
+```{r}
+if(do_pp_checks) {
+  pp_check_last_state(trt_m5_fit, trt_m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m5_fit, trt_m5_predicted_rect)
+  pp_check_transitions_direction(trt_m5_fit, trt_m5_predicted_rect, states_from = 2:4)
+}
+```
+
+## trt_m6 - D-dimer
+
+```{r}
+trt_m6_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m6_initial_states <- hcq_m6_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+trt_m6_initial_states <- trt_m6_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m6_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+trt_m6_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_d_dimer * rate_group
 
-hcq_m6 <- brmshmmdata(hcq_m6_formula, hcq_m6_serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m6_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+trt_m6 <- brmshmmdata(trt_m6_formula, trt_m6_serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m6_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
 
-hcq_m6_fit <- brmhmm(hcq_m6, cache_file = paste0(cache_dir,"/hcq_m6.rds"))
+trt_m6_fit <- brmhmm(trt_m6, cache_file = paste0(cache_dir,"/trt_m6.rds"))
 
-summary(hcq_m6_fit$brmsfit)
+summary(trt_m6_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m6_epred_rect <- posterior_epred_rect(hcq_m6_fit)
-hcq_m6_predicted_rect <- posterior_epred_to_predicted(hcq_m6_fit, hcq_m6_epred_rect)
-hcq_m6_predicted <- posterior_rect_to_long(hcq_m6_fit, hcq_m6_predicted_rect)
+draws <- tidybayes::tidy_draws(trt_m6_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m6_d_dimer_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_d_dimer,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$d_dimer_increases_death,
+    adjusted = "age, sex, hospital"
+  )
+
 ```
@@
 ```{r}
-hcq_m6_predicted_df <- posterior_long_to_df(hcq_m6_fit$data, hcq_m6_predicted)
+if(do_pp_checks) {
+  trt_m6_epred_rect <- posterior_epred_rect(trt_m6_fit)
+  trt_m6_predicted_rect <- posterior_epred_to_predicted(trt_m6_fit, trt_m6_epred_rect)
+  trt_m6_predicted <- posterior_rect_to_long(trt_m6_fit, trt_m6_predicted_rect)
+}
+```
 
-all_ids <- hcq_m6_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+```{r}
+if(do_pp_checks) {
+  trt_m6_predicted_df <- posterior_long_to_df(trt_m6_fit$data, trt_m6_predicted)
+
+  all_ids <- trt_m6_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m6_predicted_df, all_ids[series_id], hcq_m6_fit$data) %>% print()
+    posterior_state_plot(trt_m6_predicted_df, all_ids[series_id], trt_m6_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m6_fit, hcq_m6_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect)
-pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m6_fit, trt_m6_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m6_fit, trt_m6_predicted_rect)
+  pp_check_transitions_direction(trt_m6_fit, trt_m6_predicted_rect, states_from = 2:4)
+}
 ```
 
-## hcq_m7 - IL-6
+## trt_m7 - IL-6
 
 ```{r}
-hcq_m7_serie_data <- serie_data %>% 
+trt_m7_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m7_initial_states <- hcq_m7_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+trt_m7_initial_states <- trt_m7_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m7_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
+trt_m7_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
     (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_IL_6 * rate_group
 
-hcq_m7 <- brmshmmdata(hcq_m7_formula, hcq_m7_serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m7_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+trt_m7 <- brmshmmdata(trt_m7_formula, trt_m7_serie_data, 
+                  trt_m1_rate_data, trt_m1_hidden_state_data, trt_m7_initial_states, prior = trt_m1_prior,
+                  observed_state_data = trt_m1_observed_state_data)
 
-hcq_m7_fit <- brmhmm(hcq_m7, cache_file = paste0(cache_dir,"/hcq_m7.rds"))
+trt_m7_fit <- brmhmm(trt_m7, cache_file = paste0(cache_dir,"/trt_m7.rds"))
 
-summary(hcq_m7_fit$brmsfit)
+summary(trt_m7_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m7_epred_rect <- posterior_epred_rect(hcq_m7_fit)
-hcq_m7_predicted_rect <- posterior_epred_to_predicted(hcq_m7_fit, hcq_m7_epred_rect)
-hcq_m7_predicted <- posterior_rect_to_long(hcq_m7_fit, hcq_m7_predicted_rect)
+draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
+hmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <- 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_IL_6,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$IL_6_increases_death,
+    adjusted = "age, sex, hospital"
+  )
+
 ```
@@
 ```{r}
-hcq_m7_predicted_df <- posterior_long_to_df(hcq_m7_fit$data, hcq_m7_predicted)
+if(do_pp_checks) {
+  trt_m7_epred_rect <- posterior_epred_rect(trt_m7_fit)
+  trt_m7_predicted_rect <- posterior_epred_to_predicted(trt_m7_fit, trt_m7_epred_rect)
+  trt_m7_predicted <- posterior_rect_to_long(trt_m7_fit, trt_m7_predicted_rect)
+}
+```
 
-all_ids <- hcq_m7_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
+```{r}
+if(do_pp_checks) {
+  trt_m7_predicted_df <- posterior_long_to_df(trt_m7_fit$data, trt_m7_predicted)
+
+  all_ids <- trt_m7_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
     distinct() %>% arrange(hospital_id) %>% pull(.serie)
@@
 
-  posterior_state_plot(hcq_m7_predicted_df, all_ids[series_id], hcq_m7_fit$data) %>% print()
+    posterior_state_plot(trt_m7_predicted_df, all_ids[series_id], trt_m7_fit$data) %>% print()
 
+  }
 }
@@
 ```{r}
-pp_check_last_state(hcq_m7_fit, hcq_m7_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect)
-pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect, states_from = 2:6)
+if(do_pp_checks) {
+  pp_check_last_state(trt_m7_fit, trt_m7_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
+  pp_check_transitions_direction(trt_m7_fit, trt_m7_predicted_rect)
+  pp_check_transitions_direction(trt_m7_fit, trt_m7_predicted_rect, states_from = 2:4)
+}
 ```
+
+# Environment
+
+```{r}
+hmm_hypothesis_res_all <- do.call(rbind, hmm_hypothesis_res_list)
+write_csv(hmm_hypothesis_res_all, path = here::here("manuscript", "hmm_res.csv"))
+```
+
+```{r}
+sessionInfo()
+```

### 2020-10-22 R/posterior_plots_hmm.R

@@ -66,6 +66,32 @@ pp_check_last_state <- function(fit, predicted_rect) {
   bayesplot::ppc_bars(last_states_observed, t(last_states))
 }
 
+
+pp_check_state_at <- function(fit, predicted_rect, day, newdata = NULL) {
+  last_state_time <- function(x) {
+    max(which(!is.na(x) & x > 0))
+  }
+
+  if(is.null(newdata)) {
+    pred_rawdata <- fit$data
+  } else {
+    pred_rawdata <- newdata
+  }
+
+  data_hmm <- make_data_hmm(pred_rawdata)
+
+  last_state_times <- data_hmm$standata$obs_states_rect %>% apply(MARGIN = 1, last_state_time)
+
+  use_rows <- last_state_times >= day
+
+  states_at_day_observed <- data_hmm$standata$obs_states_rect[use_rows, day]
+
+  states_at <- predicted_rect[day, use_rows, ]
+
+  bayesplot::ppc_bars(states_at_day_observed, t(states_at))
+}
+
+
 pp_check_transitions <- function(fit, predicted_rect, states_from = NULL, states_to = NULL,
                                  binwidth = NULL, scale = "prob", ...) {
   N_states_observed <- fit$data_processed$standata$N_states_observed

### 2020-10-22 R/predict_hmm.R

@@ -1,3 +1,5 @@
+#'
+#' @return Array with dimensions time, serie, sample
 posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL) {
   validate_brmshmmfit(fit)
 

### 2020-10-22 manuscript/hmm.Rmd
Optimized and cleaned HMM models

--- 2020-09-16 manuscript/hmm.Rmd
@@
 data <- read_data_for_analysis()
+
+if(length(unique(data$patient_data$patient_id)) != length(data$patient_data$patient_id)) {
+  stop("FAil")
+}
+# warning("Subsetting data")
+# set.seed(4568224)
+# patients_to_use <- sample(data$patient_data$patient_id, size = 70)
+# data <- filter_complete_data(data, patient_id %in% patients_to_use)
+
 wide <- data$marker_data_wide
 
-cache_dir <- here::here("local_temp_data")
+cache_dir <- here::here("local_temp_data", "hmm")
 if(!dir.exists(cache_dir)) {
-  dir.create(cache_dir)
+  dir.create(cache_dir, recursive = TRUE)
 }
@@
 #Temporary hack
-registerS3method("stan_llh", class = "rate_hmm", method = stan_llh.rate_hmm, envir = asNamespace("brms"))
+registerS3method("stan_log_lik", class = "rate_hmm", method = stan_log_lik.rate_hmm, envir = asNamespace("brms"))
 ```
@@
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
-  mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
+  mutate(.time = .time +1) %>% 
+  filter(.time >= 1) %>%
+  group_by(.serie) %>%
+  filter(sum(!is.na(.observed)) > 1) %>%
+  ungroup() %>%
+  #Force explicit dummy coding for the models
+  mutate(across(tidyselect::starts_with("took_"), as.integer),
+         best_supportive_care = as.integer(best_supportive_care))
+  #mutate(best_supportive_care = as.integer(best_supportive_care))
+
+serie_data_28 <- wide %>%
+  inner_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day")) %>%
+  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
   mutate(.time = .time +1) %>% 
@@
 
+# Extend terminal states for serie_data_28
+serie_data_28 <- serie_data_28 %>%
+  group_by(.serie) %>%
+  mutate(last_day = max(day[!is.na(.observed)]),
+         last_state = .observed[day == last_day],
+         .observed = if_else(last_state %in% c("Discharged", "Death") & is.na(.observed) & day > last_day,
+                             last_state,
+                             .observed)) %>%
+  ungroup()
+
+base_observed_state_data <- tibble(id = factor(disease_s_levels, levels = disease_s_levels), is_noisy = FALSE)
+
+simple_hidden_state_data <- tibble(id = factor(disease_s_levels, levels = disease_s_levels)) %>% mutate( corresponding_obs = id)
+
+simple_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+
+default_control <- list(adapt_delta = 0.9)
 ```
@@
 
-hcq_m1_rate_data <- rbind(
-  tibble(.from = breathing_levels, .to = "Death", rate_group = "death"),
+```
+
+# Treatments - effects on rate groups
+
+## Only treatments
+
+```{r trt_group_m1}
+trt_group_m1_hidden_state_data <- simple_hidden_state_data
+
+trt_group_m1_observed_state_data <- base_observed_state_data
+
+trt_group_m1_rate_data <- rbind(
+  tibble(.from = c("Oxygen", "Ventilated"), .to = "Death", rate_group = "death"),
   tibble(.from = "AA", .to = "Discharged", rate_group = "improve_one"),
-  tibble(.from = breathing_levels[2 : length(breathing_levels)], .to = breathing_levels[1 : (length(breathing_levels) - 1)], rate_group = "improve_one"),
-  tibble(.from = breathing_levels[1 : (length(breathing_levels) - 1)], .to = breathing_levels[2 : length(breathing_levels)], rate_group = "worsen_one")
+  tibble(.from = breathing_s_levels[2 : length(breathing_s_levels)], .to = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], rate_group = "improve_one"),
+  tibble(.from = breathing_s_levels[1 : (length(breathing_s_levels) - 1)], .to = breathing_s_levels[2 : length(breathing_s_levels)], rate_group = "worsen_one")
 ) %>%
-  mutate(.rate_id = factor(1:n()),
-         .from = factor(.from, levels = disease_levels),
-         .to = factor(.to, levels = disease_levels)
+  mutate(.rate_id = factor(paste0(.from, "_", .to), levels = paste0(.from, "_", .to)),
+         .from = factor(.from, levels = disease_s_levels),
+         .to = factor(.to, levels = disease_s_levels)
   ) %>%
@@
   recode_group("improve_one") %>%
-  recode_group("worsen_one")
+  recode_group("worsen_one") %>%
+  mutate(rate_group = fct_relevel(rate_group, "improve_one"))
 
-hcq_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+trt_group_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m1_prior = brms::set_prior("normal(-2, 5)", "Intercept") +
-  brms::set_prior("normal(0, 5)", "b") +
+trt_group_m1_prior = #brms::set_prior("normal(-2, 5)", "Intercept") +
+  brms::set_prior("normal(-2, 5)", "b") +
   brms::set_prior("normal(0, 2)", "sd")
 
-hcq_m1_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group
-hcq_m1 <- brmshmmdata(hcq_m1_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+trt_group_m1_formula <- ~ 0 + .rate_id +  (0 + best_supportive_care + took_hcq + took_az + took_favipiravir + took_convalescent_plasma || rate_group)
 
-#make_stancode_hmm(hcq_m1)
-#xx <- brms::make_standata(hcq_m1$formula, data = make_data_hmm(hcq_m1)$brmsdata)
-hcq_m1_fit <- brmhmm(hcq_m1, cache_file = paste0(cache_dir,"/hcq_m1.rds"))
+trt_group_m1 <- brmshmmdata(trt_group_m1_formula, serie_data, 
+                  trt_group_m1_rate_data, simple_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
 
-summary(hcq_m1_fit$brmsfit)
+trt_group_m1_fit <- brmhmm(trt_group_m1, cache_file = paste0(cache_dir,"/trt_group_m1.rds"), control = default_control, init = 0.1)
 
+summary(trt_group_m1_fit$brmsfit)
 ```
+
 ```{r}
@@
 # 
-# hcq_m1_counterfactual_took_hcq <- hcq_m1
-# hcq_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
+# trt_group_m1_counterfactual_took_hcq <- trt_group_m1
+# trt_group_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
 # 
-# hcq_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_took_hcq)
+# trt_group_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(trt_group_m1_fit, newdata = trt_group_m1_counterfactual_took_hcq)
 # 
-# hcq_m1_counterfactual_no_hcq <- hcq_m1
-# hcq_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
+# trt_group_m1_counterfactual_no_hcq <- trt_group_m1
+# trt_group_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
 # 
-# hcq_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(hcq_m1_fit, newdata = hcq_m1_counterfactual_no_hcq)
+# trt_group_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(trt_group_m1_fit, newdata = trt_group_m1_counterfactual_no_hcq)
 
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m1_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m1_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "supportive"
-  )
-
-hmm_hypothesis_res_list[["hcq_m1_discharged"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "supportive"
-  )
+# draws <- tidybayes::tidy_draws(trt_group_m1_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_group_m1_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "supportive"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_group_m1_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "supportive"
+#   )
 ```
@@
 if(do_pp_checks) {
-  hcq_m1_epred_rect <- posterior_epred_rect(hcq_m1_fit)
-  hcq_m1_predicted_rect <- posterior_epred_to_predicted(hcq_m1_fit, hcq_m1_epred_rect)
-  hcq_m1_predicted <- posterior_rect_to_long(hcq_m1_fit, hcq_m1_predicted_rect)
+  do_common_pp_checks(trt_group_m1_fit)
 }
+```
+
+## Treatments + age + sex
+
+```{r trt_group_m2}
+trt_group_m2_formula <- update.formula(trt_group_m1_formula,  ~ . +  (age_norm + sex || rate_group))
+trt_group_m2 <- brmshmmdata(trt_group_m2_formula, serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
+
+trt_group_m2_fit <- brmhmm(trt_group_m2, cache_file = paste0(cache_dir,"/trt_group_m2.rds"), init = 0.1)
+
+summary(trt_group_m2_fit$brmsfit)
+```
+
+```{r}
+# draws <- tidybayes::tidy_draws(trt_group_m2_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_group_m2_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_group_m2_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive"
+#   )
 ```
@@
 if(do_pp_checks) {
-  hcq_m1_predicted_df <- posterior_long_to_df(hcq_m1_fit$data, hcq_m1_predicted)
+  do_common_pp_checks(trt_group_m2_fit)
+}
+```
 
-  all_ids <- hcq_m1_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-    distinct() %>% arrange(hospital_id) %>% pull(.serie)
-  step_size <- 6
-  for(step in 1:ceiling(length(all_ids) / step_size)) {
-    series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+## Treatments + age + sex + hospital
 
-    posterior_state_plot(hcq_m1_predicted_df, all_ids[series_id], hcq_m1_fit$data) %>% print()
+```{r trt_group_m3}
+trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
+trt_group_m3 <- brmshmmdata(trt_group_m3_formula, serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
 
-  }
-}
+trt_group_m3_fit <- brmhmm(trt_group_m3, cache_file = paste0(cache_dir,"/trt_group_m3.rds"), init = 0.1)
+
+summary(trt_group_m3_fit$brmsfit)
+```
+
+```{r}
+# draws <- tidybayes::tidy_draws(trt_group_m3_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_group_m3_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive, hospital"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_group_m3_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive, hospital"
+#   )
 ```
@@
 if(do_pp_checks) {
-  pp_check_last_state(hcq_m1_fit, hcq_m1_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-  pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect)
-  pp_check_transitions_direction(hcq_m1_fit, hcq_m1_predicted_rect, states_from = 2:6)
+  do_common_pp_checks(trt_group_m3_fit)
 }
@@
 
-```{r}
-hcq_m2_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group
-hcq_m2 <- brmshmmdata(hcq_m2_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+## Treatment + age + sex + hospital + comorbidities
 
-hcq_m2_fit <- brmhmm(hcq_m2, cache_file = paste0(cache_dir,"/hcq_m2.rds"))
+```{r trt_group_m4}
+trt_group_m4_formula <-  update.formula(trt_group_m3_formula ~ . + (comorbidities_sum || rate_group)) trt_group_m4 <- brmshmmdata(trt_group_m4_formula, serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
 
-summary(hcq_m2_fit$brmsfit)
+trt_group_m4_fit <- brmhmm(trt_group_m4, cache_file = paste0(cache_dir,"/trt_group_m4.rds"), init = 0.1)
+
+summary(trt_group_m4_fit$brmsfit)
 ```
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m2_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m2_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "age, sex, supportive"
-  )
-
-hmm_hypothesis_res_list[["hcq_m2_discharged"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "age, sex, supportive"
-  )
+# draws <- tidybayes::tidy_draws(trt_group_m4_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_group_m4_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_group_m4_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
 ```
@@
 ```{r}
-hcq_m2_epred_rect <- posterior_epred_rect(hcq_m2_fit)
-hcq_m2_predicted_rect <- posterior_epred_to_predicted(hcq_m2_fit, hcq_m2_epred_rect)
-hcq_m2_predicted <- posterior_rect_to_long(hcq_m2_fit, hcq_m2_predicted_rect)
+if(do_pp_checks) {
+  do_common_pp_checks(trt_group_m4_fit)
+}
+```
+
+# Treatments - effect on individual rates
+
+## Only treatments
+
+```{r trt_rate_m1}
+trt_rate_m1_hidden_state_data <- simple_hidden_state_data
+
+trt_rate_m1_observed_state_data <- base_observed_state_data
+
+trt_rate_m1_rate_data <- trt_group_m1_rate_data
+
+trt_rate_m1_initial_states <- trt_group_m1_initial_states
+
+trt_rate_m1_prior = trt_group_m1_prior
+
+trt_rate_m1_formula <- ~ 0 + .rate_id +  (0 + best_supportive_care + took_hcq + took_az + took_favipiravir + took_convalescent_plasma || .rate_id)
+
+trt_rate_m1 <- brmshmmdata(trt_rate_m1_formula, serie_data, 
+                  trt_rate_m1_rate_data, simple_hidden_state_data, simple_initial_states, prior = trt_rate_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+trt_rate_m1_fit <- brmhmm(trt_rate_m1, cache_file = paste0(cache_dir,"/trt_rate_m1.rds"), control = default_control, init = 0.1)
+
+summary(trt_rate_m1_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m2_predicted_df <- posterior_long_to_df(hcq_m2_fit$data, hcq_m2_predicted)
 
-all_ids <- hcq_m2_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
+# counterfactual_max_days <- 90
+# counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
+#   inner_join(data$patient_data, by = c("patient_id")) %>% 
+#   left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
+#   mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
+#   mutate(.time = .time +1) %>% 
+#   filter(.time >= 1) %>%
+#   group_by(.serie) %>%
+#   mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
+#   ungroup()
+# 
+# 
+# counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
+# counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
+# 
+# 
+# 
+# trt_rate_m1_counterfactual_took_hcq <- trt_rate_m1
+# trt_rate_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
+# 
+# trt_rate_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(trt_rate_m1_fit, newdata = trt_rate_m1_counterfactual_took_hcq)
+# 
+# trt_rate_m1_counterfactual_no_hcq <- trt_rate_m1
+# trt_rate_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
+# 
+# trt_rate_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(trt_rate_m1_fit, newdata = trt_rate_m1_counterfactual_no_hcq)
 
-  posterior_state_plot(hcq_m2_predicted_df, all_ids[series_id], hcq_m2_fit$data) %>% print()
+```
 
+```{r}
+# draws <- tidybayes::tidy_draws(trt_rate_m1_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_rate_m1_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "supportive"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_rate_m1_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "supportive"
+#   )
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(trt_rate_m1_fit)
 }
@@
 
-```{r}
-pp_check_last_state(hcq_m2_fit, hcq_m2_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect)
-pp_check_transitions_direction(hcq_m2_fit, hcq_m2_predicted_rect, states_from = 2:6)
+## Treatments + age + sex
+
+```{r trt_rate_m2}
+trt_rate_m2_formula <- update.formula(trt_rate_m1_formula,  ~ . +  (age_norm + sex || .rate_id))
+trt_rate_m2 <- brmshmmdata(trt_rate_m2_formula, serie_data, 
+                  trt_rate_m1_rate_data, trt_rate_m1_hidden_state_data, trt_rate_m1_initial_states, prior = trt_rate_m1_prior,
+                  observed_state_data = trt_rate_m1_observed_state_data)
+
+trt_rate_m2_fit <- brmhmm(trt_rate_m2, cache_file = paste0(cache_dir,"/trt_rate_m2.rds"), init = 0.1)
+
+summary(trt_rate_m2_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m3_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m3 <- brmshmmdata(hcq_m3_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
-
-hcq_m3_fit <- brmhmm(hcq_m3, cache_file = paste0(cache_dir,"/hcq_m3.rds"))
-
-summary(hcq_m3_fit$brmsfit)
+# draws <- tidybayes::tidy_draws(trt_rate_m2_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_rate_m2_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_rate_m2_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive"
+#   )
 ```
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m3_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m3_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "age, sex, supportive, hospital"
-  )
+if(do_pp_checks) {
+  do_common_pp_checks(trt_rate_m2_fit)
+}
+```
 
-hmm_hypothesis_res_list[["hcq_m3_discharged"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "age, sex, supportive, hospital"
-  )
+## Treatments + age + sex + hospital
+
+```{r trt_rate_m3}
+trt_rate_m3_formula <- update.formula(trt_rate_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
+trt_rate_m3 <- brmshmmdata(trt_rate_m3_formula, serie_data, 
+                  trt_rate_m1_rate_data, trt_rate_m1_hidden_state_data, trt_rate_m1_initial_states, prior = trt_rate_m1_prior,
+                  observed_state_data = trt_rate_m1_observed_state_data)
+
+trt_rate_m3_fit <- brmhmm(trt_rate_m3, cache_file = paste0(cache_dir,"/trt_rate_m3.rds"), init = 0.1)
+
+summary(trt_rate_m3_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m3_epred_rect <- posterior_epred_rect(hcq_m3_fit)
-hcq_m3_predicted_rect <- posterior_epred_to_predicted(hcq_m3_fit, hcq_m3_epred_rect)
-hcq_m3_predicted <- posterior_rect_to_long(hcq_m3_fit, hcq_m3_predicted_rect)
+# draws <- tidybayes::tidy_draws(trt_rate_m3_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_rate_m3_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive, hospital"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_rate_m3_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive, hospital"
+#   )
 ```
@@
 ```{r}
-hcq_m3_predicted_df <- posterior_long_to_df(hcq_m3_fit$data, hcq_m3_predicted)
-
-all_ids <- hcq_m3_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
-
-  posterior_state_plot(hcq_m3_predicted_df, all_ids[series_id], hcq_m3_fit$data) %>% print()
-
+if(do_pp_checks) {
+  do_common_pp_checks(trt_rate_m3_fit)
 }
@@
 
-```{r}
-pp_check_last_state(hcq_m3_fit, hcq_m3_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect)
-pp_check_transitions_direction(hcq_m3_fit, hcq_m3_predicted_rect, states_from = 2:6)
+## Treatment + age + sex + hospital + comorbidities
+
+```{r trt_rate_m4}
+trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula ~ . + (comorbidities_sum || .rate_id)) 
+trt_rate_m4 <- brmshmmdata(trt_rate_m4_formula, serie_data, 
+                  trt_rate_m1_rate_data, trt_rate_m1_hidden_state_data, trt_rate_m1_initial_states, prior = trt_rate_m1_prior,
+                  observed_state_data = trt_rate_m1_observed_state_data)
+
+trt_rate_m4_fit <- brmhmm(trt_rate_m4, cache_file = paste0(cache_dir,"/trt_rate_m4.rds"), init = 0.1)
+
+summary(trt_rate_m4_fit$brmsfit)
 ```
@@
 ```{r}
-hcq_m4_formula <- ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + took_az * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id)
-hcq_m4 <- brmshmmdata(hcq_m4_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
-
-hcq_m4_fit <- brmhmm(hcq_m4, cache_file = paste0(cache_dir,"/hcq_m4.rds"))
-
-summary(hcq_m4_fit$brmsfit)
+# draws <- tidybayes::tidy_draws(trt_rate_m4_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_rate_m4_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
+# 
+# hmm_hypothesis_res_list[["trt_rate_m4_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
 ```
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m4_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m4_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "age, sex, supportive, hospital, az"
-  )
-
-hmm_hypothesis_res_list[["hcq_m4_discharged"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "age, sex, supportive, hospital, az"
-  )
-
-hmm_hypothesis_res_list[["hcq_m4_death_az"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_azTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "age, sex, supportive, hospital, hcq"
-  )
-
-hmm_hypothesis_res_list[["hcq_m4_discharged_az"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_azTRUE + draws$`b_rate_groupimprove_one:took_azTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "age, sex, supportive, hospital, hcq"
-  )
-```
-
-```{r}
-hcq_m4_epred_rect <- posterior_epred_rect(hcq_m4_fit)
-hcq_m4_predicted_rect <- posterior_epred_to_predicted(hcq_m4_fit, hcq_m4_epred_rect)
-hcq_m4_predicted <- posterior_rect_to_long(hcq_m4_fit, hcq_m4_predicted_rect)
-```
-
-```{r}
-hcq_m4_predicted_df <- posterior_long_to_df(hcq_m4_fit$data, hcq_m4_predicted)
-
-all_ids <- hcq_m4_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
-
-  posterior_state_plot(hcq_m4_predicted_df, all_ids[series_id], hcq_m4_fit$data) %>% print()
-
+if(do_pp_checks) {
+  do_common_pp_checks(trt_rate_m4_fit)
 }
@@
 
-```{r}
-pp_check_last_state(hcq_m4_fit, hcq_m4_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect)
-pp_check_transitions_direction(hcq_m4_fit, hcq_m4_predicted_rect, states_from = 2:6)
-```
+# Markers
 
-## hcq_m5 -
+## D-dimer + hospital
 
-```{r}
-hcq_m5_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + took_hcq * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + comorbidities_sum * rate_group #mo(comorbidities_sum) * rate_group
-hcq_m5 <- brmshmmdata(hcq_m5_formula, serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m1_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
-
-hcq_m5_fit <- brmhmm(hcq_m5, cache_file = paste0(cache_dir,"/hcq_m5.rds"))
-
-summary(hcq_m5_fit$brmsfit)
-```
-
-```{r}
-draws <- tidybayes::tidy_draws(hcq_m5_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m5_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-  )
-
-hmm_hypothesis_res_list[["hcq_m5_discharged"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-  )
-```
-
-```{r}
-hcq_m5_epred_rect <- posterior_epred_rect(hcq_m5_fit)
-hcq_m5_predicted_rect <- posterior_epred_to_predicted(hcq_m5_fit, hcq_m5_epred_rect)
-hcq_m5_predicted <- posterior_rect_to_long(hcq_m5_fit, hcq_m5_predicted_rect)
-```
-
-```{r}
-hcq_m5_predicted_df <- posterior_long_to_df(hcq_m5_fit$data, hcq_m5_predicted)
-
-all_ids <- hcq_m5_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
-
-  posterior_state_plot(hcq_m5_predicted_df, all_ids[series_id], hcq_m5_fit$data) %>% print()
-
-}
-```
-
-```{r}
-pp_check_last_state(hcq_m5_fit, hcq_m5_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect)
-pp_check_transitions_direction(hcq_m5_fit, hcq_m5_predicted_rect, states_from = 2:6)
-```
-
-## hcq_m6 - D-dimer
-
-```{r}
-hcq_m6_serie_data <- serie_data %>% 
+```{r d_dimer_m1}
+d_dimer_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m6_initial_states <- hcq_m6_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+d_dimer_initial_states <- d_dimer_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m6_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_d_dimer * rate_group
+d_dimer_m1_formula <-  ~ 0 + .rate_id +  (0 + best_supportive_care || rate_group)  + (0 + .rate_id | hospital_id) + (log_peak_d_dimer || rate_group)
 
-hcq_m6 <- brmshmmdata(hcq_m6_formula, hcq_m6_serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m6_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+d_dimer_m1 <- brmshmmdata(d_dimer_m1_formula, d_dimer_serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
 
-hcq_m6_fit <- brmhmm(hcq_m6, cache_file = paste0(cache_dir,"/hcq_m6.rds"))
+d_dimer_m1_fit <- brmhmm(trt_m6, cache_file = paste0(cache_dir,"/d_dimer_m1.rds"), init = 0.1)
 
-summary(hcq_m6_fit$brmsfit)
+summary(d_dimer_m1_fit$brmsfit)
 ```
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m6_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m6_d_dimer_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_log_peak_d_dimer,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$d_dimer_increases_death,
-    adjusted = "age, sex, hospital"
-  )
+# draws <- tidybayes::tidy_draws(trt_m6_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_m6_d_dimer_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_log_peak_d_dimer,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$d_dimer_increases_death,
+#     adjusted = "age, sex, hospital"
+#   )
+#
 
@@
 ```{r}
-hcq_m6_epred_rect <- posterior_epred_rect(hcq_m6_fit)
-hcq_m6_predicted_rect <- posterior_epred_to_predicted(hcq_m6_fit, hcq_m6_epred_rect)
-hcq_m6_predicted <- posterior_rect_to_long(hcq_m6_fit, hcq_m6_predicted_rect)
-```
-
-```{r}
-hcq_m6_predicted_df <- posterior_long_to_df(hcq_m6_fit$data, hcq_m6_predicted)
-
-all_ids <- hcq_m6_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
-
-  posterior_state_plot(hcq_m6_predicted_df, all_ids[series_id], hcq_m6_fit$data) %>% print()
-
+if(do_pp_checks) {
+  do_common_pp_checks(d_dimer_m1_fit)
 }
@@
 
+## D-dimer + age + sex + comorbidities
+
 ```{r}
-pp_check_last_state(hcq_m6_fit, hcq_m6_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect)
-pp_check_transitions_direction(hcq_m6_fit, hcq_m6_predicted_rect, states_from = 2:6)
+d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (age_norm + sex + comorbidities_sum || rate_group))
+
+d_dimer_m2 <- brmshmmdata(d_dimer_m2_formula, d_dimer_serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
+
+d_dimer_m2_fit <- brmhmm(trt_m6, cache_file = paste0(cache_dir,"/d_dimer_m2.rds"), init = 0.1)
+
+summary(d_dimer_m2_fit$brmsfit)
+
 ```
 
-## hcq_m7 - IL-6
+## IL-6 + hospital
 
 ```{r}
-hcq_m7_serie_data <- serie_data %>% 
+IL_6_serie_data <- serie_data %>% 
   group_by(.serie) %>%
@@
 
-hcq_m7_initial_states <- hcq_m7_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+IL_6_initial_states <- IL_6_serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
-hcq_m7_formula <-  ~ is_improve_one + is_death + mo(death_id) + (0 + is_worsen_one | worsen_one_id) + 
-    (0 + is_improve_one | improve_one_id) + best_supportive_care * rate_group + age_norm * rate_group + sex * rate_group + (rate_group | hospital_id) + log_peak_IL_6 * rate_group
+IL_6_m1_formula <- ~ 0 + .rate_id +  (0 + best_supportive_care || rate_group)  + (0 + .rate_id | hospital_id) + (log_peak_IL_6 || rate_group)
 
-hcq_m7 <- brmshmmdata(hcq_m7_formula, hcq_m7_serie_data, 
-                  hcq_m1_rate_data, hcq_m1_hidden_state_data, hcq_m7_initial_states, prior = hcq_m1_prior,
-                  observed_state_data = hcq_m1_observed_state_data)
+IL_6_m1 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
 
-hcq_m7_fit <- brmhmm(hcq_m7, cache_file = paste0(cache_dir,"/hcq_m7.rds"))
+IL_6_m1_fit <- brmhmm(IL_6_m1, cache_file = paste0(cache_dir,"/IL_6_m1.rds"), init = 0.1)
 
-summary(hcq_m7_fit$brmsfit)
+summary(IL_6_m1_fit$brmsfit)
 ```
@@
 ```{r}
-draws <- tidybayes::tidy_draws(hcq_m7_fit$brmsfit)
-hmm_hypothesis_res_list[["hcq_m7_IL_6_death"]] <- 
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_log_peak_IL_6,
-    model = "HMM",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$IL_6_increases_death,
-    adjusted = "age, sex, hospital"
-  )
+# draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
+# hmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_log_peak_IL_6,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$IL_6_increases_death,
+#     adjusted = "age, sex, hospital"
+#   )
 
@@
 ```{r}
-hcq_m7_epred_rect <- posterior_epred_rect(hcq_m7_fit)
-hcq_m7_predicted_rect <- posterior_epred_to_predicted(hcq_m7_fit, hcq_m7_epred_rect)
-hcq_m7_predicted <- posterior_rect_to_long(hcq_m7_fit, hcq_m7_predicted_rect)
-```
-
-```{r}
-hcq_m7_predicted_df <- posterior_long_to_df(hcq_m7_fit$data, hcq_m7_predicted)
-
-all_ids <- hcq_m7_fit$data$serie_data %>% select(hospital_id, .serie) %>% 
-  distinct() %>% arrange(hospital_id) %>% pull(.serie)
-step_size <- 6
-for(step in 1:ceiling(length(all_ids) / step_size)) {
-  series_id <- ((step - 1) * step_size + 1) : (step* step_size)
-
-  posterior_state_plot(hcq_m7_predicted_df, all_ids[series_id], hcq_m7_fit$data) %>% print()
-
+if(do_pp_checks) {
+  do_common_pp_checks(IL_6_m1_fit)
 }
@@
 
+## IL_6 + age + sex + comorbidities
+
 ```{r}
-pp_check_last_state(hcq_m7_fit, hcq_m7_predicted_rect) #TODO separately terminal and non-terminal states. Or "state at time t" - this should be coherent
-pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect)
-pp_check_transitions_direction(hcq_m7_fit, hcq_m7_predicted_rect, states_from = 2:6)
+IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (age_norm + sex + comorbidities_sum || rate_group))
+
+IL_6_m2 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
+                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
+
+IL_6_m2_fit <- brmhmm(IL_6_m2, cache_file = paste0(cache_dir,"/IL_6_m2.rds"), init = 0.1)
+
+summary(IL_6_m2_fit$brmsfit)
 ```

### 2020-10-22 manuscript/competing_risks.Rmd
Work on manuscript

@@
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+write_csv(hypothesis_res_all, path = here("manuscript", "/competing_risks_res.csv"))
 ```

### 2020-10-22 manuscript/manuscript.Rmd
Work on manuscript

@@
 ---
-title: "Detailed disease progression of 97 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
+title: "Detailed disease progression of 191 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
 output: 
@@
   #pdf_document
-  bookdown::pdf_document2:
+  # bookdown::pdf_document2:
+  #    toc: false
+  bookdown::word_document2: 
      toc: false
-  # bookdown::word_document2: 
-  #     toc: false
   # bookdown::pdf_book: 
@@
 wide <- data$marker_data_wide
+
+n_patients <- nrow(data$patient_data)
+n_hospitals <- length(unique(data$patient_data$hospital_id))
 ```
@@
 Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
-The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths.
+The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighbouring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
+The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths.
 This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e.
-that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
-We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening.
-The states and available transitions are shown in Figure&nbsp;\@ref(fig:modelstates)b.
+that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
+We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction.
 
-```{r modelstates, fig.cap="States used in the competing risk model (a) and in the hidden Markov model (b).
-AA = Ambient air, Oxygen = Nasal oxygen, NIPPV = Non-invasive positive-pressure ventilation, ECMO = Extra-corporeal membrane oxygenation.
-In both models the 'Death' and 'Discharged' states are terminal.
-In the hidden Markov model, the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
-knitr::include_graphics(here::here("manuscript","static_figs","model_states.pdf"))
+```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c).
+AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation.
+In all models the 'Death' and 'Discharged' states are terminal.
+In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
+knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
 ```
@@
 The estimated value of the marker is then used as one of the predictors in the time-to-event component.
+
+Finally we used a set of Bayesian regression models implemented with the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
+Those included overall survival, state at day 7 or 28 as either binary or ordinal outcome, Bayesian version of the Cox proportional-hazards model and time-to-event models.
 
@@
 This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5].
+We used
 
 # Results
-
-```{r}
-n_patients <- nrow(data$patient_data)
-n_hospitals <- length(unique(data$patient_data$hospital_id))
-```
 
@@
   rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
+        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
         read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
@@
 
+```
+
+```{r}
+plot_multiverse(all_hypo %>% filter(group == "favipiravir"))
 ```

### 2020-10-22 manuscript/risk_scores.Rmd
Work on manuscript

@@
            COPD,
-         bc0 = bc_base + 7 * atb0,
-         bc1 = bc_base + 7 * atb1
+         bc0 = bc_base + 7 * (atb0 & admitted_for_covid),
+         bc1 = bc_base + 7 * (atb1 & admitted_for_covid)
   )
@@
 The thin gray dashed line marks the AUC from the original study for easier comparison.
-We use 3 variants of the score - `base` means not inferring pneumonia at all, `day0` means that pneumonia is inferred from antibiotics at hospitalization, `day1` - inferring pneumonia from antibiotics within 1 day of hospitalization "}
+We use 3 variants of the score - `base` means not inferring pneumonia at all, `day0` means that pneumonia is inferred from being admitted for covid and antibiotics at hospitalization, `day1` - inferring pneumonia from antibiotics within 1 day of hospitalization "}
 bc_auc <- rbind(

### 2020-10-22 manuscript/analysis_brms.Rmd
Minor tweaks to brms analysis

@@
 
-cache_dir <- here::here("local_temp_data")
+cache_dir <- here::here("local_temp_data", "analysis_brms")
 if (!dir.exists(cache_dir)) {
-  dir.create(cache_dir)
+  dir.create(cache_dir, recursive = TRUE)
 }
@@
 }
+
+default_control <- list(adapt_delta = 0.95)
 ```
@@
   prior = prior_survival_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base")
@@
 ```{r}
+#Marking the results as suspicious because there was a divergence...
 draws <- tidy_draws(fit_survival_base)
@@
   prior = prior_survival_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base_7")
@@
   prior = prior_survival_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base_28")
@@
   inits = 0,
+  control = default_control,
   file = paste0(cache_dir, "/fit_survival_time")
@@
   prior = prior_discharged_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base")
@@
   prior = prior_discharged_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base_7")
@@
   prior = prior_discharged_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base_28")
@@
   inits = 0,
-  control = list(adapt_delta = 0.95),
+  control = list(adapt_delta = 0.98),
   file = paste0(cache_dir, "/fit_discharged_time")
@@
   prior = prior_outcome_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base")
@@
   prior = prior_outcome_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base_7")
@@
   prior = prior_outcome_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base_28")
@@
   prior = prior_worst_breathing_base,
+  control = default_control,
   file = paste0(cache_dir, "/fit_worst_breathing_base")
@@
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/analysis_brms.csv"))
+write_csv(hypothesis_res_all, path = here::here("manuscript/analysis_brms.csv"))
 ```

### 2020-10-22 manuscript/hmm.Rmd
Fixingg issues with HMM models

@@
   mutate(across(tidyselect::starts_with("took_"), as.integer),
-         best_supportive_care = as.integer(best_supportive_care))
+         best_supportive_care = as.integer(best_supportive_care),
+         is_male = as.integer(sex == "M")
+         )
   #mutate(best_supportive_care = as.integer(best_supportive_care))
@@
 
-trt_group_m1_rate_data <- rbind(
+simple_rate_data <- rbind(
   tibble(.from = c("Oxygen", "Ventilated"), .to = "Death", rate_group = "death"),
@@
   recode_group("worsen_one") %>%
-  mutate(rate_group = fct_relevel(rate_group, "improve_one"))
+  mutate(rate_group = fct_relevel(rate_group, "improve_one"),
+         )
 
-trt_group_m1_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+simple_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
 
@@
 trt_group_m1 <- brmshmmdata(trt_group_m1_formula, serie_data, 
-                  trt_group_m1_rate_data, simple_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, simple_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = base_observed_state_data)
 
-trt_group_m1_fit <- brmhmm(trt_group_m1, cache_file = paste0(cache_dir,"/trt_group_m1.rds"), control = default_control, init = 0.1)
+trt_group_m1_fit <- brmhmm(trt_group_m1, cache_file = paste0(cache_dir,"/trt_group_m1.rds"), control = default_control, init = 0.1, iter = 1500)
 
 summary(trt_group_m1_fit$brmsfit)
+```
+
+```{r, fig.width=12, fig.height=12}
+#bayesplot::mcmc_pairs(xx$fit, regex_pars = c("b_.rate_id"))
+# bayesplot::mcmc_pairs(trt_group_m1_fit$brmsfit, regex_pars = "worsen_one", np = nuts_params(trt_group_m2_fit$brmsfit))
+
+```
+
+```{r, fig.width=12, fig.height=12}
+# bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, pars = c("b_.rate_idAA_Discharged", "b_.rate_idOxygen_AA"), regex_pars = "improve.*(age|male)")
+# 
+#bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, regex_pars = c("b_.rate_id"))
 ```
@@
 ```{r trt_group_m2}
-trt_group_m2_formula <- update.formula(trt_group_m1_formula,  ~ . +  (age_norm + sex || rate_group))
+trt_group_m2_formula <- update.formula(trt_group_m1_formula,  ~ . +  (0 + age_norm + is_male || rate_group))
 trt_group_m2 <- brmshmmdata(trt_group_m2_formula, serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-trt_group_m2_fit <- brmhmm(trt_group_m2, cache_file = paste0(cache_dir,"/trt_group_m2.rds"), init = 0.1)
+trt_group_m2_fit <- brmhmm(trt_group_m2, cache_file = paste0(cache_dir,"/trt_group_m2.rds"), init = 0.1, control = default_control, iter = 1200)
 
+#parnames(trt_group_m2_fit$brmsfit)
 summary(trt_group_m2_fit$brmsfit)
+```
+```{r, fig.width=12, fig.height=12}
+# 
+#  
+# bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, regex_pars = "improve_one", np = nuts_params(trt_group_m2_fit$brmsfit))
+# 
+# 
+# bayesplot::mcmc_parcoord(trt_group_m2_fit$brmsfit, regex_pars = "death", np = nuts_params(trt_group_m2_fit$brmsfit))
+# bayesplot::mcmc_parcoord(trt_group_m2_fit$brmsfit, regex_pars = "worsen_one", np = nuts_params(trt_group_m2_fit$brmsfit))
 ```
@@
 trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
+
 trt_group_m3 <- brmshmmdata(trt_group_m3_formula, serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-trt_group_m3_fit <- brmhmm(trt_group_m3, cache_file = paste0(cache_dir,"/trt_group_m3.rds"), init = 0.1)
+trt_group_m3_fit <- brmhmm(trt_group_m3, cache_file = paste0(cache_dir,"/trt_group_m3.rds"), init = 0.1, control = default_control, iter = 1200)
 
@@
 ```{r trt_group_m4}
-trt_group_m4_formula <-  update.formula(trt_group_m3_formula ~ . + (comorbidities_sum || rate_group)) trt_group_m4 <- brmshmmdata(trt_group_m4_formula, serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, trt_group_m1_initial_states, prior = trt_group_m1_prior,
+trt_group_m4_formula <-  update.formula(trt_group_m3_formula,  ~ . + (comorbidities_sum || rate_group))
+
+trt_group_m4 <- brmshmmdata(trt_group_m4_formula, serie_data, 
+                  simple_rate_data, trt_group_m1_hidden_state_data, simple_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-trt_group_m4_fit <- brmhmm(trt_group_m4, cache_file = paste0(cache_dir,"/trt_group_m4.rds"), init = 0.1)
+trt_group_m4_fit <- brmhmm(trt_group_m4, cache_file = paste0(cache_dir,"/trt_group_m4.rds"), init = 0.1, control = default_control, iter = 1200)
 
@@
 
-trt_rate_m1_rate_data <- trt_group_m1_rate_data
+trt_rate_m1_rate_data <- simple_rate_data
 
-trt_rate_m1_initial_states <- trt_group_m1_initial_states
+trt_rate_m1_initial_states <- simple_initial_states
 
@@
 ```{r trt_rate_m2}
-trt_rate_m2_formula <- update.formula(trt_rate_m1_formula,  ~ . +  (age_norm + sex || .rate_id))
+trt_rate_m2_formula <- update.formula(trt_rate_m1_formula,  ~ . +  (0 + age_norm + is_male || rate_group))
 trt_rate_m2 <- brmshmmdata(trt_rate_m2_formula, serie_data, 
@@
 
-trt_rate_m2_fit <- brmhmm(trt_rate_m2, cache_file = paste0(cache_dir,"/trt_rate_m2.rds"), init = 0.1)
+trt_rate_m2_fit <- brmhmm(trt_rate_m2, cache_file = paste0(cache_dir,"/trt_rate_m2.rds"), init = 0.1, control = default_control)
 
@@
 
-trt_rate_m3_fit <- brmhmm(trt_rate_m3, cache_file = paste0(cache_dir,"/trt_rate_m3.rds"), init = 0.1)
+trt_rate_m3_fit <- brmhmm(trt_rate_m3, cache_file = paste0(cache_dir,"/trt_rate_m3.rds"), init = 0.1, control = default_control, iter = 1500)
 
@@
 ```{r trt_rate_m4}
-trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula ~ . + (comorbidities_sum || .rate_id)) 
+trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula, ~ . + (comorbidities_sum || .rate_id))
+
 trt_rate_m4 <- brmshmmdata(trt_rate_m4_formula, serie_data, 
@@
 
-trt_rate_m4_fit <- brmhmm(trt_rate_m4, cache_file = paste0(cache_dir,"/trt_rate_m4.rds"), init = 0.1)
+trt_rate_m4_fit <- brmhmm(trt_rate_m4, cache_file = paste0(cache_dir,"/trt_rate_m4.rds"), init = 0.1, control = default_control, iter = 1500)
 
@@
 d_dimer_m1 <- brmshmmdata(d_dimer_m1_formula, d_dimer_serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-d_dimer_m1_fit <- brmhmm(trt_m6, cache_file = paste0(cache_dir,"/d_dimer_m1.rds"), init = 0.1)
+d_dimer_m1_fit <- brmhmm(d_dimer_m1, cache_file = paste0(cache_dir,"/d_dimer_m1.rds"), init = 0.1, control = default_control, iter = 1500)
 
@@
 ```{r}
-d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (age_norm + sex + comorbidities_sum || rate_group))
+d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
 
 d_dimer_m2 <- brmshmmdata(d_dimer_m2_formula, d_dimer_serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, d_dimer_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-d_dimer_m2_fit <- brmhmm(trt_m6, cache_file = paste0(cache_dir,"/d_dimer_m2.rds"), init = 0.1)
+d_dimer_m2_fit <- brmhmm(d_dimer_m2, cache_file = paste0(cache_dir,"/d_dimer_m2.rds"), init = 0.1, control = default_control, iter = 1500)
 
@@
 IL_6_m1 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-IL_6_m1_fit <- brmhmm(IL_6_m1, cache_file = paste0(cache_dir,"/IL_6_m1.rds"), init = 0.1)
+IL_6_m1_fit <- brmhmm(IL_6_m1, cache_file = paste0(cache_dir,"/IL_6_m1.rds"), init = 0.1, control = default_control, iter = 1500)
 
@@
 ```{r}
-IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (age_norm + sex + comorbidities_sum || rate_group))
+IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
 
 IL_6_m2 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
-                  trt_group_m1_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
+                  simple_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
                   observed_state_data = trt_group_m1_observed_state_data)
 
-IL_6_m2_fit <- brmhmm(IL_6_m2, cache_file = paste0(cache_dir,"/IL_6_m2.rds"), init = 0.1)
+IL_6_m2_fit <- brmhmm(IL_6_m2, cache_file = paste0(cache_dir,"/IL_6_m2.rds"), init = 0.1, control = default_control, iter = 1500)
 
 summary(IL_6_m2_fit$brmsfit)
+```
+
+# Complex: Treatments - effects on rate groups
+
+## Only treatments
+
+```{r c_trt_m1}
+complex_rate_data <- rbind(
+  tibble(.from = paste0(c("Oxygen", "Ventilated"), "_worsening"), .to = "Death", rate_group = "death"),
+  tibble(.from = "AA_improving", 
+         .to = "Discharged", 
+         rate_group = "improve_one"),
+  tibble(.from = paste0(breathing_s_levels[2 : length(breathing_s_levels)], "_improving"), 
+         .to = paste0(breathing_s_levels[1 : (length(breathing_s_levels) - 1)] , "_improving"), 
+         rate_group = "improve_one"),
+  tibble(.from = paste0(breathing_s_levels[1 : (length(breathing_s_levels) - 1)], "_worsening"),
+         .to = paste0(breathing_s_levels[2 : length(breathing_s_levels)], "_worsening"),
+         rate_group = "worsen_one"),
+  tibble(.from = paste0(breathing_s_levels, "_worsening"),
+         .to = paste0(breathing_s_levels, "_improving"),
+         rate_group = "to_improving"),
+  tibble(.from = paste0(breathing_s_levels, "_improving"),
+         .to = paste0(breathing_s_levels, "_worsening"),
+         rate_group = "to_worsening")
+) %>%
+  mutate(.rate_id = factor(paste0(.from, "_", .to), levels = paste0(.from, "_", .to)),
+  ) %>%
+  mutate(rate_group = fct_relevel(rate_group, "improve_one"))
+
+complex_hidden_state_data <- rbind(
+  tibble(base = breathing_s_levels) %>% mutate(corresponding_obs = base) %>% 
+    crossing(tibble(state_group = c("_improving", "_worsening"))) %>% 
+    mutate(id = factor(paste0(base, state_group))) %>%
+    select(-base, -state_group),
+  tibble(id = c("Death","Discharged")) %>% mutate(corresponding_obs = id)
+)
+
+complex_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")
+
+c_trt_m1 <- brmshmmdata(trt_group_m1_formula, serie_data, 
+                  complex_rate_data, complex_hidden_state_data, complex_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+c_trt_m1_fit <- brmhmm(c_trt_m1, cache_file = paste0(cache_dir,"/c_trt_m1.rds"), control = default_control, init = 0.1, iter = 1500)
+
+summary(c_trt_m1_fit$brmsfit)
+
+```
+```{r}
+do_common_pp_checks(c_trt_m1_fit)
 ```

### 2020-10-23 manuscript/analysis_brms.Rmd
Cleanups, tests, more complex HMM models

@@
 encoding: UTF-8
-editor_options: 
-  chunk_output_type: console
+#editor_options: 
+#  chunk_output_type: console
 ---
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiraviTRUE,
+    draws = draws$b_ever_favipiravirTRUE,
     model = "brms_cox",
     estimand = "log(HR)",
-    hypothesis = hypotheses$favipiravi_increases_discharged,
+    hypothesis = hypotheses$favipiravir_increases_discharged,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity"

### 2020-10-23 manuscript/hmm.Rmd
Cleanups, tests, more complex HMM models

@@
 ```{r}
-serie_data <- wide %>%
-  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
-  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
+transform_serie_data <- function(raw_serie_data) {
+  raw_serie_data  %>%
   mutate(.time = .time +1) %>% 
@@
          )
-  #mutate(best_supportive_care = as.integer(best_supportive_care))
+}
+
+serie_data <- wide %>%
+  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
+  mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
+  transform_serie_data()
 
 serie_data_28 <- wide %>%
-  inner_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day")) %>%
+  right_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day")) %>%
   mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
-  mutate(.time = .time +1) %>% 
-  filter(.time >= 1) %>%
-  group_by(.serie) %>%
-  filter(sum(!is.na(.observed)) > 1) %>%
-  ungroup()
+  transform_serie_data()
 
-# Extend terminal states for serie_data_28
+# Extend terminal states and took_XX data for serie_data_28
 serie_data_28 <- serie_data_28 %>%
@@
                              .observed)) %>%
+  mutate(across(tidyselect::starts_with("took_"), ~ if_else(is.na(.x), as.integer(any(.x != 0, na.rm = TRUE)), .x))) %>%
   ungroup()
@@
 ```{r}
+# fit <- trt_group_m1_fit
+# data_28 <- trt_group_m1
+# data_28$serie_data <- serie_data_28
+# epred_rect <- posterior_epred_rect(fit)#, newdata = data_28)
+# predicted_rect <- posterior_epred_to_predicted(fit, epred_rect)
+# # predicted <- posterior_rect_to_long(fit, predicted_rect)
+# 
+# pp_check_state_at(fit, predicted_rect, 28, newdata = data_28)
+```
 
-# counterfactual_max_days <- 90
-# counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
-#   inner_join(data$patient_data, by = c("patient_id")) %>% 
-#   left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
-#   mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
-#   mutate(.time = .time +1) %>% 
-#   filter(.time >= 1) %>%
-#   group_by(.serie) %>%
-#   mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
-#   ungroup()
-# 
-# 
-# counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
+```{r}
+# counterfactual_took_hcq <- serie_data_28 %>% mutate(took_hcq = TRUE)
 # counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
@@
 ```{r}
+# fit <- trt_group_m1_fit
+# epred_rect <- posterior_epred_rect(fit)
+# predicted_rect <- posterior_epred_to_predicted(fit, epred_rect)
+# predicted <- posterior_rect_to_long(fit, predicted_rect)
+
 if(do_pp_checks) {
@@
 ```{r trt_rate_m2}
-trt_rate_m2_formula <- update.formula(trt_rate_m1_formula,  ~ . +  (0 + age_norm + is_male || rate_group))
+trt_rate_m2_formula <- update.formula(trt_rate_m1_formula,  ~ . +  (0 + age_norm + is_male || .rate_id))
 trt_rate_m2 <- brmshmmdata(trt_rate_m2_formula, serie_data, 
@@
 ```{r}
+if(do_pp_checks) {
   do_common_pp_checks(c_trt_m1_fit)
+}
+```
+
+## Treatments + age + sex
+
+```{r c_trt_m2}
+c_trt_m2 <- brmshmmdata(trt_group_m2_formula, serie_data, 
+                  complex_rate_data, complex_hidden_state_data, complex_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+c_trt_m2_fit <- brmhmm(c_trt_m2, cache_file = paste0(cache_dir,"/c_trt_m2.rds"), init = 0.1, control = default_control, iter = 1200)
+
+#parnames(c_trt_m2_fit$brmsfit)
+summary(c_trt_m2_fit$brmsfit)
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(c_trt_m2_fit)
+}
+```
+
+## Treatments + age + sex + hospital
+
+```{r c_trt_m3}
+c_trt_m3 <- brmshmmdata(trt_group_m3_formula, serie_data, 
+                  complex_rate_data, complex_hidden_state_data, complex_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+c_trt_m3_fit <- brmhmm(c_trt_m3, cache_file = paste0(cache_dir,"/c_trt_m3.rds"), init = 0.1, control = default_control, iter = 1200)
+
+summary(c_trt_m3_fit$brmsfit)
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(c_trt_m3_fit)
+}
+```
+
+## Treatment + age + sex + hospital + comorbidities
+
+```{r c_trt_m4}
+c_trt_m4 <- brmshmmdata(trt_group_m4_formula, serie_data, 
+                  complex_rate_data, complex_hidden_state_data, complex_initial_states, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+c_trt_m4_fit <- brmhmm(c_trt_m4, cache_file = paste0(cache_dir,"/c_trt_m4.rds"), init = 0.1, control = default_control, iter = 1200)
+
+summary(c_trt_m4_fit$brmsfit)
+```
+
+```{r}
+# draws <- tidybayes::tidy_draws(c_trt_m4_fit$brmsfit)
+# hmm_hypothesis_res_list[["c_trt_m4_death"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_reduces_death,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
+# 
+# hmm_hypothesis_res_list[["c_trt_m4_discharged"]] <- 
+#   bayesian_hypothesis_res_from_draws(
+#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
+#     model = "HMM",
+#     estimand = "log(HR)",
+#     hypothesis = hypotheses$hcq_increases_discharged,
+#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
+#   )
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(c_trt_m4_fit)
+}
 ```

### 2020-10-26 manuscript/manuscript.Rmd
Work on manuscript and presentation

@@
 ---
-title: "Detailed disease progression of 191 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
+title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
 output: 
@@
     toc: false
-  # bookdown::pdf_book: 
-  #    base_format: rticles::arxiv_article
+    keep_md: true
+    reference_docx: "output_template.docx"
 abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
+Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination.
 To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
@@
 date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
+author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
 ---
@@
 ```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, fig.width = 8)
 
@@
 The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths.
-Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR; TODO].
-Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
-While initially promising, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
+Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
+Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting (TODO cite).
+Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
 Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
 Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
-No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US] For additional summary see [@http://zotero.org/users/5567156/items/YP62P6XJ].
+No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
 
@@
 
+Favipiravir is one of the less explored experimental treatmets, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
+Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/IHATMFTC; @http://zotero.org/users/5567156/items/894H9N9B].
+None of the results allow strong conclusions about benefits or harms of Favipiravir.
+For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].
+
 High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/ID7F3XIH].
-Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed TODO.
+Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the diesease.
 
@@
 ## Data Collection
-A convenience sample of patients...
+
+```{r}
+ethics_data <- readxl::read_excel(here::here("manuscript", "ethics.xlsx"), sheet = "Sheet1", na = "NA")
+ethics_string <- ethics_data %>% pull(Site) %>% paste0(collapse = ", ")
+```
+
+A convenience sample of patients from `r n_hospitals` sites was collected.
+For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period.
+The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.
 Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.
+We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
 Full protocol for data collection is attached in the supplementary material.
-The study was approved by the Ethical committees of TODO.
+The study was approved by the Ethical committees of `r ethics_string`.
 
@@
 For each model class we worked with several possible sets of adjustments and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC].
-All analysis wes performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualisations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
-Complete code for all analyses can be found at TODO.
+All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualisations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
 
@@
 where we have gathered all the input features).
-This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5].
-We used
+We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67].
+When there were multiple reasonable ways to evaluate either outcome or a predictor in our dataset, we computed and reported all of those options.
+As a baseline to compare the scores against we used age and the decade of age as the sole predictor.
+
+Complete code for all analyses is available at https://github.com/cas-bioinf/covid19retrospective/.
 
@@
 
-In Figure&nbsp;\@ref(fig:progression) we show the disease progression for all patients.
+In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure &nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
+The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization.
+We also see that the marker levels were not substantially stratified by study site.
+Those should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.
 
-```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required."}
+```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
+Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
 my_paste <- function(x) {
@@
   theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())
+
+```
+
+```{r selectedmarkers, fig.cap = "Values of selected markers over the course of the disease.
+Each line represents a patient, stratified by the worst breathing support required over the hospitalization.
+Color indicates study sites, notably  The vertical scale is logarithmic.
+Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
+#data$marker_data %>%
+# TODO: use SpO2 in AA as SpO2 native (and vice versa)
+plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {
+
+  if(length(markers) > 1) {
+    facet = facet_grid(marker ~ worst_breathing_s, scales = "free_y")
+    scale_y_name = "Value"
+  } else {
+    facet = facet_wrap( ~ worst_breathing_s, nrow = 1)
+    unit <- data$marker_data %>% filter(marker == names(markers)) %>% pull(unit) %>% unique()
+    scale_y_name <- markers
+  }
+
+  data_for_plot <- data$marker_data %>% 
+    filter(marker %in% names(markers), day >= 0) %>% 
+    group_by(marker, patient_id) %>%
+    filter(n() > 1) %>%
+    ungroup() %>%
+    mutate(marker = markers[marker])
+
+  if(!is.null(max_patients)) {
+    patient_markers_to_use <- data_for_plot %>% 
+      select(marker, patient_id) %>%
+      distinct() %>%
+      group_by(marker) %>%
+      slice_sample(n = max_patients)
+
+    data_for_plot <- data_for_plot %>%
+      semi_join(patient_markers_to_use, by = c("patient_id", "marker"))
+  }
+
+  if(log_transform) {
+    scale_y <- scale_y_log10(scale_y_name)
+    data_for_plot <- data_for_plot %>% 
+      group_by(marker) %>%
+      mutate(value = if_else(value == 0, min(value[value > 0]) / 2, value)) %>%
+      ungroup()
+  } else {
+    scale_y <- scale_y_continuous(scale_y_name)
+  }
+
+  if(smooth) {
+    main_geom <- geom_smooth(aes(group = outcome))
+  } else {
+    main_geom <- geom_line(alpha = 0.5)
+  }
+
+  if(!is.null(max_day)) {
+    expand <- expand_limits(x = max_day)
+  } else {
+    expand <- NULL
+  }
+
+  data_for_plot %>%
+    inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>%
+    mutate(outcome = fct_relevel(fct_collapse(outcome, Hospitalized = c("Hospitalized", "Transferred")),
+            "Discharged", "Hospitalized", "Death"),
+           worst_breathing_s =fct_recode(worst_breathing_s, `Ambient Air` = "AA")) %>%
+    ggplot(aes(x = day, y = value,  group = patient_id, color = hospital_id)) + 
+    main_geom +
+    facet + 
+    expand +
+    #scale_color_brewer(type = "qual", palette = 2, guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + 
+    scale_y +
+    scale_x_continuous("Days since hospitalization") +
+    scale_color_discrete(guide = FALSE)
+    #scale_color_viridis_d(guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + scale_y
+}
+
+max_day <- data$marker_data_wide %>% filter(!is.na(CRP)) %>% pull(day) %>% max()
+hide_axis_theme <- theme(axis.text.x = element_blank(),
+                         axis.line.x = element_blank(),
+                         axis.title.x = element_blank(),
+                         axis.ticks.x = element_blank())
+hide_strip_theme <- theme(strip.background = element_blank(), strip.text = element_blank())
+title_theme <- theme(axis.title.y = element_text(size = 10))
+
+((plot_marker_data(c("CRP" = "CRP"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme) /
+(plot_marker_data(c("d_dimer" = "D-dimer"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme + hide_strip_theme) /
+(plot_marker_data(c("lymphocyte_count" = "Ly"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme + hide_strip_theme) /
+(plot_marker_data(c("IL_6" = "IL-6"), log_transform = TRUE, max_day = max_day) + title_theme + hide_strip_theme))
+
+  #plot_marker_data(c("d_dimer" = "D-dimer", "IL_6" = "IL 6", "CRP" = "CRP"), log_transform = TRUE)
+#plot_marker_data(c("d_dimer", "IL_6", "CRP"), log_transform = FALSE)
+
+# plot_marker_data(c("CRP"), log_transform = TRUE)
+# plot_marker_data(c("CRP"), log_transform = TRUE, max_patients = 30)
+# plot_marker_data(c("CRP"), log_transform = TRUE, max_patients = 30)
 
@@
 
-## Association between treatment, markers and outcomes
+## Association between treatments and outcomes
+
+```{r}
+#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
+#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
+#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
+```
+
+Here, we focus on Hydroxychloroquine and Azithromycin as it is the only treatment with larger sample size.
+We also investigate Favipiravir and convalescent plasma as those therapies are less well reported in the literature.
+Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg.
+Majority of patients complemented Hydroxychloroquine with Azithromycin while Azithromycin was rarely used alone (see Table \@ref(tab:summarytable)).
+Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.
+Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose.
+Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
+
+The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
 
@@
 For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval."}
+
+#In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
+Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged).
+After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome.
+In all cases, the results seem to not be strongly affected by model choice.
+
 plot_multiverse <- function(hypo_res) {
@@
 
+## Association between markers and outcomes
+
+We additionally investigated the association of several markers with disease progression.
+Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice.
+All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)
+
 ```{r multiverseresultsmarkers, fig.height=2.5, fig.width=6, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
@@
 
-The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
-In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
-Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged).
-After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome.
-In all cases, the results seem to not be strongly affected by model choice.
+## Prognostic models are not better using age only
 
-We additionally investigated the association of several markers with disease progression.
-Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice.
-All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)
+Following [@http://zotero.org/users/5567156/items/TS57HJ3T] we found five prediction models we were able to recompute: Li et al.
+report the ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] combining CRP and age to form 3 grades, Chen & Liu [@http://zotero.org/users/5567156/items/4DDUVEKS] report a continuous score using age, CRP, D-dimer and lymphocyte count,
+Shi et al.
+[@http://zotero.org/users/5567156/items/X4GWJIP5] use age, sex and hypertension to form 4 grades, Caramelo et al.
+use age, sex, hypertension, diabetes, cardiac disease, chronic respiratory disease and cancer to form a continuous score [@http://zotero.org/users/5567156/items/3A2RN4ZS], Bello-Chavolla et al.
+[@http://zotero.org/users/5567156/items/INZVABQI] use age, diabetes, obesity, pneumonia, chronic kidney disease, COPD and immunosuppression to build a score ranging from -6 to 22.
+For the latter two scores we had to impute some of the predictors as they had no immediate equivalent in our dataset.
+The outcomes present in the studies were: 12-day mortality, 30-day mortality and mortality without any further details, here we report results for both 12-day and 30-day mortality.
+Full details on the scores and how we used our dataset to compute them is given in the supplementary material.
 
-## Survey of antimicrobial treatment
+Overall all prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
+Additionally, most publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed.
+We thus concur with [@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.
 
-## Evaluating prediction models
+```{r riskscores, fig.cap="Performance of tested prediction scores as measured by AUC.
+AUC = 1 means perfect prediction while AUC <= 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the predictions (marked by thick blue line).
+The line ranges represent the bootstrapped 95% confidence intervals.
+Red dots show results computed in present study - individual dots vary in the outcome measured (12-day or 30-day mortality) and potentially on how ambiguities in score computation were resolved, although this rarely makes a big difference.
+Cyan triangles show AUC as reported by the original authors or recomputed based on their published data.
+When the confidence interval or the AUC of the original study is not reported, it means that the value was not reported by the authors and not enough information to recompute it was given."}
+col_types_auc <- cols(
+  auc = col_double(),
+  auc_low = col_double(),
+  auc_high = col_double(),
+  outcome = col_character(),
+  score = col_character(),
+  note = col_character(),
+  subgroup = col_character()
+)
+all_auc <- read_csv(here::here("manuscript", "all_auc.csv"), col_types = col_types_auc, na = "NA") %>% filter(subgroup != "(admission)")
+all_auc_orig <- read_csv(here::here("manuscript", "all_auc_orig.csv"), col_types = col_types_auc, na = "NA") %>% filter(subgroup != "(admission)")
 
-The ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] is a simple score that categorizes the patients into 3 categories based on age and C-reactive protein (CRP): grade 1 (age $<$ 60 years and CRP $<$ 34 mg/L), grade 2 (age $\geq$ 60
-years and CRP $<$ 34 mg/L or age $<$ 60 years and CRP $\geq$ 34 mg/L), grade 3 (age $\geq$ 60 years and CRP $\geq$ 34 mg/L) and was evaluated primarily on 12-day mortality.
-Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in rough agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).
+compare_auc_to_orig(all_auc_orig,
+                    all_auc, show_outcome = FALSE)
 
-```{r acpresults, echo=FALSE}
-acp_data <- wide %>%  
-  group_by(patient_id) %>%
-  filter(any(!is.na(CRP))) %>%
-  summarise(
-    first_CRP_day = min(day[!is.na(CRP)]),
-    first_CRP = CRP[day == first_CRP_day],
-    worst_breathing_12 = max(breathing[day >= first_CRP_day & day <= first_CRP_day + 12]),
-    .groups = "drop") %>% 
-  inner_join(data$patient_data, by = c("patient_id")) %>%
-  filter(outcome %in% c("Death","Discharged") | last_record - first_CRP_day >= 12) %>%
-  mutate(    
-    alive_12 = last_record > first_CRP_day + 12 | outcome == "Discharged",
-    ACP_grade = case_when(
-       age < 60 & first_CRP < 34 ~ 1,
-       age >= 60 & first_CRP >= 34 ~ 3,
-       TRUE ~ 2
-    ) 
-  )
-
-acp_data %>%
-  group_by(ACP_grade) %>%
-  summarise(`patients` = n(), `12-day mortality` = scales::percent(mean(!alive_12), accuracy = 0.1), .groups = "drop") %>%
-  arrange(ACP_grade) %>%
-  mutate(`Li et al. 95% CI` = c("0%", "0 - 11.3%", "19.8 - 44.3%")) %>%
-  rename(`ACP grade` = ACP_grade) %>%
-  knitr::kable(booktabs = TRUE, caption = "Comparing the 12-day mortality based on ACP score (where available) with the 95\\% confidence intervals (CI) for 12-day mortality reported by Lu et al.")
-
-# acp_data %>%
-#   group_by(ACP_grade, worst_breathing_12) %>%
-#   summarise(mean(alive_12))
 ```
-
-(TODO: the other prediction models)
 

### 2020-10-26 R/data_preprocessing.R

@@ -69,7 +69,14 @@ read_data_for_analysis <- function() {
                               day = col_integer(),
                               value = col_double(),
                               censored = col_character()
-                            ))
+                            )) %>%
+    # One more bug quick fix (fixed in raw data for further imports)
+    mutate(marker = if_else(patient_id == "809N695Ho6" & marker == "SpO2_native" & day >= 4, "Horowitz_index", marker)) %>%
+    # Filtering harmless NAs (due to error in data)
+    filter(!(patient_id == "d2TgUar00l" & is.na(value)))
+
+
+
 
   multiple_units <- marker_data %>%
     group_by(marker) %>%
@@ -114,7 +121,8 @@ compute_derived_quantities_patients <- function(data) {
               ever_antibiotics = any(took_antibiotics),
               any_d_dimer = any(!is.na(d_dimer)),
               any_IL_6 = any(!is.na(IL_6)),
-              worst_breathing = max(breathing),
+              worst_breathing = max(breathing, na.rm = TRUE),
+              worst_breathing_s = max(breathing_s, na.rm = TRUE),
               had_invasive = any(breathing >= "MV" & breathing < "Death"),
               first_day_invasive = if_else(had_invasive, min(c(day[breathing >= "MV"], 1e6)), NA_real_),
               last_day_invasive = if_else(had_invasive, max(c(day[breathing >= "MV"], 0)), NA_real_),

### 2020-10-26 manuscript/manuscript.Rmd
More work on manuscript

@@
 
-```{r}
-#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
-#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
-#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
-```
-
 Here, we focus on Hydroxychloroquine and Azithromycin as it is the only treatment with larger sample size.
@@
 Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.
-Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose.
+Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose all but one of the patients receiving Favipiravir also received Hydroxychloroquine.
 Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
+Except for convalescent plasma, treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).
 
-The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
+```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites.
+Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
+#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
+#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
+#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
 
-```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between treatments and main outcomes.
+marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
+
+data$marker_data %>% 
+  filter(marker %in% c("az", "hcq", "favipiravir", "convalescent_plasma"), !is.na(value), value > 0) %>%
+  mutate(marker = marker_map[marker]) %>%
+  group_by(marker, patient_id) %>% 
+  summarise(first_day = min(day), .groups = "drop") %>% 
+  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
+```
+
+The results need to be interpreted very carefully as the treatme  The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
+
+```{r multiverseresultsdrugs, fig.height=9, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between treatments and main outcomes.
 Each row represents a model - HMM = Bayesian hidden-markov model, coxph = Frequentist competing risks model.
@@
 
-(multiverse_hcq / multiverse_az) + plot_layout(heights = c(2.2,1))
+(multiverse_hcq / multiverse_az) + plot_layout(heights = c(2.5,1.5))
 
@@
 
-```{r}
+```{r, fig.height=4}
 plot_multiverse(all_hypo %>% filter(group == "favipiravir"))
@@
 
+An important lesson from our analysis is that the assesment of treatment efficacy from observational data is heavily sensitive to modelling assumptions while it is usually almost impossible to determine which of the models is more likely to reflect reality (if any).
+We believe that using multiverse analysis is an appropriate way to explore data in such contexts as it lets us be transparent about this sensitivity.
+
 Where our resutls contradict randomized trials, the most likely explanation is systematic bias in our dataset.

### 2020-10-26 manuscript/authors.R

@@ -1,0 +1,20 @@
+get_authors_string <- function() {
+  authors_csv <- read.csv(here::here("private_data", "authors.csv"), stringsAsFactors = FALSE, encoding = "UTF-8")
+
+  authors_csv$Affiliation = factor(authors_csv$Affiliation, levels = unique(authors_csv$Affiliation))
+
+  max_affiliation <- 0
+
+  authors_strings <- character( nrow(authors_csv))
+  for(i in 1:nrow(authors_csv)) {
+    if(as.integer(authors_csv$Affiliation[i]) > max_affiliation) {
+      affiliation <- paste0("^[", authors_csv$Affiliation[i], "]")
+      max_affiliation <- as.integer(authors_csv$Affiliation[i])
+    } else {
+      affiliation <- paste0("^", as.integer(authors_csv$Affiliation[i]), "^")
+    }
+    authors_strings[i] <- paste0(gsub(" ", "&nbsp;", authors_csv$Name[i]), affiliation)
+  }
+
+  paste0(authors_strings, collapse = ", ")
+}

### 2020-10-26 manuscript/hmm.Rmd
Forgottent commit of HMM model fixes

@@
 
-IL_6_m1 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
+IL_6_m1 <- brmshmmdata(IL_6_m1_formula, IL_6_serie_data, 
                   simple_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,
@@
 
-IL_6_m2 <- brmshmmdata(IL_6_initial_states, IL_6_serie_data, 
+IL_6_m2 <- brmshmmdata(IL_6_m2_formula, IL_6_serie_data, 
                   simple_rate_data, trt_group_m1_hidden_state_data, IL_6_initial_states, prior = trt_group_m1_prior,

### 2020-10-27 manuscript/manuscript.Rmd
TS's stuff added: treatment combinations overview, baseline associations with treatment, convenience sample warning, coxph modeling

@@
 Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination.
-To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
+To ~~assess~~ *__explore__* association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
 Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
@@
 author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
----
 
@@
 library(patchwork)
+library(tsiMisc)
 theme_set(cowplot::theme_cowplot())
@@
 ## Statistical analysis
+
+It shall be noted that the character of the convenience sample did not allow to properly assess 
+the association between treatments, markers and patient outcomes,
+because the treatments had not been assigned to patients at random (as in a randomized clinical trial), 
+but were only observed retrospectively.
+Therefore, we could not exclude the possibility of accidental association of treatments to patients' baseline characteristics, 
+which could, in turn, demonstrate in spurious association between treatments and outcomes, 
+because both of these can be driven by the patients' characteristics. 
+Eventhough we attempted to account for possible associations of treatments to patients' baseline characteristics, 
+we still could not rule them out, and the results of our analyses should be considered as exploratory.
 
@@
 In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.
+Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
 
@@
   knitr::kable(caption = "Patient characteristics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
+```
+
+```{r treatment_comb_table, echo=FALSE}
+pd<-data$patient_data
+x<-sort(table(paste0(
+  as.numeric(pd$ever_hcq),
+  as.numeric(pd$ever_az),
+  as.numeric(pd$ever_convalescent_plasma),
+  as.numeric(pd$ever_favipiravir),
+  as.numeric(pd$ever_remdesivir),
+  as.numeric(pd$ever_dexamethasone),
+  as.numeric(pd$ever_antibiotics))),dec=TRUE)
+x2<-c()
+nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
+for (i in 1:length(x)) {
+  x3<-c()
+  for (i2 in 1:nchar(names(x)[i])) {
+    x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
+  }
+  x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
+}
+rownames(x2)<-c()
+colnames(x2)<-c('N',nms)
+x2<-x2[,c(1,2,3,8,4:7)]
+knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
 ```
@@
 res_frequentist <- 
-  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols))
+  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols),
+        read_csv(here::here("manuscript", "single_outcome_coxph_res.csv"), col_types = frequentist_cols))
 
@@
 ```
+
+## Association between patients' characteristics and treatments
+
+As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
+To illustrate this phenomenon, we give an example of the association of treatment with Hydroxychloroquine on patients' baseline characteristics.
+
+```{r hcq_assoc, message=FALSE, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be asociated with the status of patients upon hospitalization.
+Comorbidities were asociated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red).
+HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
+
+tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
+  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
+  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
+tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
+tmp$outcome<-dropLevels(tmp$outcome)
+tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
+tmp2$outcome<-dropLevels(tmp2$outcome)
+tmp2$outcome<-tmp2$outcome=='Death'
+
+m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
+m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
+m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
+m.copd<-glm(outcome~COPD,tmp2,family='binomial')
+m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
+m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
+m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
+m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
+m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
+m.s<-glm(outcome~smoking,tmp2,family='binomial')
+m.ob<-glm(outcome~obesity,tmp2,family='binomial')
+m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
+m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
+m.la<-glm(outcome~low_albumin,tmp2,family='binomial')
+
+m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
+m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
+m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
+m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
+m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
+m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
+m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
+m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
+m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
+m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
+m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
+m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
+m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
+m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')
+
+models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
+modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
+ys<-c(0)
+for (m in models) {
+  ys<-c(min(min(ys),confint(m)[2,1]),max(max(ys),confint(m)[2,2]))
+}
+models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
+modelNames2<-modelNames
+ys2<-c(0)
+for (m in models2) {
+  ys2<-c(min(min(ys2),confint(m)[2,1]),max(max(ys2),confint(m)[2,2]))
+}
+
+plot(ys,c(1,length(modelNames)),ty='n',yaxt='n',xlab='Log odds ratio',ylab='')
+abline(v=0,col='blue')
+col.hcq<-bfc()["Vermillion"]
+for (i in seq(along=models)) {
+  ii<-length(models)-i+1
+  m<-models[[i]]
+  m2<-models2[[i]]
+  lines(confint(m)[2,],rep(ii,2))
+  points(coef(m)[2],ii,pch=19)
+  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
+  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
+}
+axis(2,length(modelNames):1,modelNames,las=2)
+legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
+```
+
+Figure&nbsp;\@ref(fig:hcq_assoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
+were less likely to receive Hydroxychloroquine treatment. 
+This clearly creates a bias towards a positive effect of Hydroxychloroquine on the outcome, 
+which, however, could be false.
+
+```{r}
+# condition on the sum of comorbidities
+pdd<-within(pd,cnd<-ischemic_heart_disease+has_hypertension_drugs+heart_failure+COPD+    other_lung_disease+    renal_disease+  high_creatinin)
+m.hcq_treatment2<-glm(ever_hcq~cnd,pdd,family='binomial')
+m.hcq_treatment0<-glm(ever_hcq~1,pdd,subset=!is.na(cnd),family='binomial')
+```
+
+Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
+the score representing the cummulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
+other lung diseases, renal disease, or high creatinin 
+was asscociated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
+(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
+(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
+Chi-square test in the logistic regression model, 
+$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
+P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).
 

### 2020-10-27 manuscript/single_outcome_coxph.Rmd
TS's stuff added: treatment combinations overview, baseline associations with treatment, convenience sample warning, coxph modeling

---
title: "Survival - frequentist, single outcome"
output: html_notebook
---

```{r}
devtools::load_all()
library(tidyverse)
library(survival)

theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()

# add hcq_first
pd<-data$patient_data
pd$hcq_first<-NA
for (i in unique(pd$patient_id)) {
  tmp<-data$marker_data$day[data$marker_data$patient_id==i&data$marker_data$marker=='hcq']
  if (sum(!is.na(tmp))>0) {
    pd$hcq_first[pd$patient_id==i]<-min(tmp)
  }
}

# prepare surv data
d<-c()
# I know, the code is busy, but intuitive
for (i in unique(pd$patient_id)) {
  cnd<-pd$patient_id==i
  if (pd$last_record[cnd]==0) next
  hcq_first<-pd$hcq_first[cnd]
  if (!is.na(hcq_first)) {
    if (hcq_first>0) {
      if (pd$last_record[cnd]>hcq_first) {
        if (hcq_first>1) {
          d<-rbind(d,data.frame(t0=0,t1=hcq_first-1,outcome=0,patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
        }
        d<-rbind(d,data.frame(t0=hcq_first-1,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=1))
      } else {
        d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
      }
    } else { # hcq_first=0
      d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=1))
    }
  } else {
    d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
  }
}
d<-within(d,comorb_sum<-ischemic_heart_disease+has_hypertension_drugs+heart_failure+COPD+    other_lung_disease+    renal_disease)

hypothesis_res_list <- list()

cache_dir <- here::here("local_temp_data")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

# Risk of death, increased discharge

```{r}

m1<-with(d,coxph(S<-Surv(t0,t1,outcome=='Death')~hcq+age+sex+strata(hospital_id)+comorb_sum))
m0<-with(d,coxph(S<-Surv(t0,t1,outcome=='Death')~    age+sex+strata(hospital_id)+comorb_sum))

mt1<-with(d,coxph(S<-Surv(t0,t1,outcome=='Discharged')~hcq+age+sex+strata(hospital_id)+comorb_sum))
mt0<-with(d,coxph(S<-Surv(t0,t1,outcome=='Discharged')~    age+sex+strata(hospital_id)+comorb_sum))

hypothesis_res_list[["competing_hcq_comorbidities_sum"]] <-
  rbind(
  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_reduces_death, 
                                         point_estimate = coef(m1)[1],
                                         adjusted = "age, sex, comorbidities (sum)", 
                                         test_stat = anova(m1,m0,test='Chisq')[2,2],
                                         df = anova(m1,m0,test='Chisq')[2,3],
                                         p = anova(m1,m0,test='Chisq')[2,4],
                                         ci_low = confint(m1)[1,1],
                                         ci_high = confint(m1)[1,2]),
  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_increases_discharged, 
                                         point_estimate = coef(mt1)[1],
                                         adjusted = "age, sex, comorbidities (sum)", 
                                         test_stat = anova(mt1,mt0,test='Chisq')[2,2],
                                         df = anova(mt1,mt0,test='Chisq')[2,3],
                                         p = anova(mt1,mt0,test='Chisq')[2,4],
                                         ci_low = confint(mt1)[1,1],
                                         ci_high = confint(mt1)[1,2])
  )
hypothesis_res_list[["competing_hcq_comorbidities_sum"]]

```

```{r}
hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
write_csv(hypothesis_res_all, path = here::here("manuscript", "/single_outcome_coxph_res.csv"))
```

### 2020-10-27 R/hypotheses.R

@@ -37,6 +37,23 @@ frequentist_hypothesis_res_from_coxph <- function(
   )
 }
 
+# single outcome
+frequentist_hypothesis_res_from_coxph1 <- function(
+  hypothesis, adjusted, point_estimate, test_stat, df, p_value, ci_low, ci_high, model_check = "OK") {
+
+  data.frame(hypothesis = hypothesis$name,
+             model = "coxph",
+             adjusted = adjusted,
+             estimand = "log(HR)",
+             point_estimate = point_estimate,
+             p_value = p_value,
+             ci_low = ci_low,
+             ci_high = ci_high,
+             model_check = model_check,
+             data_version = get_data_version()
+  )
+}
+
 
 bayesian_hypothesis_res_from_jm <- function(
   hypothesis, jm_fit, coefficient_name, adjusted, model_check = "OK") {

### 2020-10-31 manuscript/hmm.Rmd
Somewhat working new hypothesis code

@@
 do_pp_checks <- FALSE
-hmm_hypothesis_res_list <- list()
+hhmm_hypothesis_res_list <- list()
 
@@
 serie_data <- wide %>%
-  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>% 
+  inner_join(data$patient_data, by = c("hospital_id", "patient_id"), suffix = c("_admission", "")) %>% 
   mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
@@
 serie_data_28 <- wide %>%
-  right_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day")) %>%
+  right_join(crossing(data$patient_data, day = 0:28), by = c("hospital_id", "patient_id", "day"),  suffix = c("_admission", "")) %>%
+  # group_by(patient_id) %>%
+  # mutate(
+  #   across(
+  #     c(starts_with("took_"), all_of("best_supportive_care")),
+  #     ~ if_else(is.na(.x) & day > max(day[!is.na(.x)]), .x[which.max(day[!is.na(.x)])], .x)
+  #     ),
+  # ) %>%
+  # ungroup() %>%
   mutate(.time = day, .serie = patient_id, .observed = breathing_s) %>%
@@
                              .observed)) %>%
-  mutate(across(tidyselect::starts_with("took_"), ~ if_else(is.na(.x), as.integer(any(.x != 0, na.rm = TRUE)), .x))) %>%
+  mutate(across(c(tidyselect::starts_with("took_"), all_of("best_supportive_care")), ~ if_else(is.na(.x), as.integer(any(.x != 0, na.rm = TRUE)), .x))) %>%
   ungroup()
+
+# Check succesful extension
+mismatches <- serie_data_28 %>% inner_join(serie_data, by = c("patient_id", "day")) %>%
+  select(-starts_with("."), -last_day,  -last_state) %>%
+  mutate(across(everything(), as.character)) %>%
+  pivot_longer(!any_of(c("patient_id", "day")), names_to = c("col", "source"), names_sep = "\\.") %>%
+  group_by(patient_id, day, col) %>%
+  summarise(n_vals = length(unique(value)), .groups = "drop") %>%
+  filter(n_vals > 1)
+
+if(nrow(mismatches) > 0) {
+  print(mismatches)
+  stop("Mismatches in serie_data_28")
+}
 
@@
 ```{r}
-# counterfactual_took_hcq <- serie_data_28 %>% mutate(took_hcq = TRUE)
-# counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
-# 
-# 
-# 
-# trt_group_m1_counterfactual_took_hcq <- trt_group_m1
-# trt_group_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
-# 
-# trt_group_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(trt_group_m1_fit, newdata = trt_group_m1_counterfactual_took_hcq)
-# 
-# trt_group_m1_counterfactual_no_hcq <- trt_group_m1
-# trt_group_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
-# 
-# trt_group_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(trt_group_m1_fit, newdata = trt_group_m1_counterfactual_no_hcq)
+compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_data, denominator_serie_data, day = 28) {
+  if(nrow(numerator_serie_data) != nrow(denominator_serie_data)) {
+    stop("Groups to compare must have equal number of rows")
+  }
+
+  numerator_data <- hmm_fit$data
+  numerator_data$serie_data = numerator_serie_data
+
+  numerator_state_probs <- posterior_epred_rect(hmm_fit, newdata = numerator_data, method = posterior_epred_state_prob)
+
+  denominator_data <- hmm_fit$data
+  denominator_data$serie_data = denominator_serie_data
+
+  denominator_state_probs <- posterior_epred_rect(hmm_fit, newdata = denominator_data, method = posterior_epred_state_prob)
+
+  mortality_state <- which(hmm_fit$data$hidden_state_data$id == "Death")
+  samples_probs_numerator <- 
+    rowMeans(numerator_state_probs[mortality_state, day + 1, ,])
+  samples_probs_denominator <- 
+    rowMeans(denominator_state_probs[mortality_state, day + 1, ,])
+
+  log_odds_numerator <- log(samples_probs_numerator) - log1p(-samples_probs_numerator)
+  log_odds_denominator <- log(samples_probs_denominator) - log1p(-samples_probs_denominator)
+
+  log_odds_numerator - log_odds_denominator
+}
+
+evaluate_treatment_hypothesis <- function(fit, hypothesis, treatment_column, model_subgroup, adjusted, model_check) {
+  counterfactual_took <- serie_data_28 %>% mutate({{ treatment_column }} := 1)
+  counterfactual_no <- serie_data_28 %>% mutate({{treatment_column }} := 0)
+
+  log_OR_samples <- compute_counterfactual_mortality_log_OR(fit, counterfactual_took, counterfactual_no, day = 28)
+
+  bayesian_hypothesis_res_from_draws(
+    draws = log_OR_samples,
+    model = paste0("HMM", model_subgroup),
+    estimand = "log(OR)",
+    hypothesis = hypothesis,
+    adjusted = adjusted,
+    model_check = model_check
+  )
+}
+
+evaluate_all_treatment_hypotheses <- function(fit, model_subgroup, adjusted, model_check) {
+  rbind(
+    evaluate_treatment_hypothesis(fit, hypotheses$hcq_reduces_death, took_hcq, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check),
+    evaluate_treatment_hypothesis(fit, hypotheses$az_reduces_death, took_az, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check),
+    evaluate_treatment_hypothesis(fit, hypotheses$favipiravir_reduces_death, took_favipiravir, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check)
+  )
+}
 
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_group_m1_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_group_m1_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "supportive"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_group_m1_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "supportive"
-#   )
+hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
 ```
@@
 ```{r}
-# fit <- trt_group_m1_fit
-# epred_rect <- posterior_epred_rect(fit)
-# predicted_rect <- posterior_epred_to_predicted(fit, epred_rect)
-# predicted <- posterior_rect_to_long(fit, predicted_rect)
-
 if(do_pp_checks) {
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_group_m2_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_group_m2_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_group_m2_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive"
-#   )
+hhmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_group_m3_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_group_m3_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive, hospital"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_group_m3_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive, hospital"
-#   )
+hhmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_group_m4_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_group_m4_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_group_m4_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
+hhmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
 ```
@@
 ```{r}
-
-# counterfactual_max_days <- 90
-# counterfactual_base <-  crossing(day = 0:counterfactual_max_days, patient_id = unique(serie_data$patient_id)) %>%
-#   inner_join(data$patient_data, by = c("patient_id")) %>% 
-#   left_join(wide, by = c("hospital_id", "patient_id", "day")) %>%
-#   mutate(.time = day, .serie = patient_id, .observed = breathing) %>%
-#   mutate(.time = .time +1) %>% 
-#   filter(.time >= 1) %>%
-#   group_by(.serie) %>%
-#   mutate(best_supportive_care = !is.na(best_supportive_care_from) & day >= best_supportive_care_from) %>%
-#   ungroup()
-# 
-# 
-# counterfactual_took_hcq <- counterfactual_base %>% mutate(took_hcq = TRUE)
-# counterfactual_no_hcq <- counterfactual_base %>% mutate(took_hcq = FALSE)
-# 
-# 
-# 
-# trt_rate_m1_counterfactual_took_hcq <- trt_rate_m1
-# trt_rate_m1_counterfactual_took_hcq$serie_data = counterfactual_took_hcq
-# 
-# trt_rate_m1_counterfactual_took_hcq_epred_rect <- posterior_epred_rect(trt_rate_m1_fit, newdata = trt_rate_m1_counterfactual_took_hcq)
-# 
-# trt_rate_m1_counterfactual_no_hcq <- trt_rate_m1
-# trt_rate_m1_counterfactual_no_hcq$serie_data = counterfactual_no_hcq
-# 
-# trt_rate_m1_counterfactual_no_hcq_epred_rect <- posterior_epred_rect(trt_rate_m1_fit, newdata = trt_rate_m1_counterfactual_no_hcq)
-
-```
-
-```{r}
-# draws <- tidybayes::tidy_draws(trt_rate_m1_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_rate_m1_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "supportive"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_rate_m1_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "supportive"
-#   )
+hhmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_rate_m2_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_rate_m2_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_rate_m2_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive"
-#   )
+hhmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_rate_m3_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_rate_m3_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive, hospital"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_rate_m3_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive, hospital"
-#   )
+hhmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_rate_m4_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_rate_m4_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
-# 
-# hmm_hypothesis_res_list[["trt_rate_m4_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
+hhmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_m6_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_m6_d_dimer_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_log_peak_d_dimer,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$d_dimer_increases_death,
-#     adjusted = "age, sex, hospital"
-#   )
-#
+draws <- tidybayes::tidy_draws(d_dimer_m1_fit$brmsfit)
+hhmm_hypothesis_res_list[["d_dimer_m1_death"]] <-
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_d_dimer,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$d_dimer_increases_death,
+    adjusted = "supportive, age, sex, hospital"
+  )
 
@@
 
+```
+
+```{r}
+stop("Calculate hypo")
 ```
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
-# hmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_log_peak_IL_6,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$IL_6_increases_death,
-#     adjusted = "age, sex, hospital"
-#   )
+stop("Calculate hypo")
+
+draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
+hhmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <-
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_log_peak_IL_6,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$IL_6_increases_death,
+    adjusted = "age, sex, hospital"
+  )
 
@@
 summary(IL_6_m2_fit$brmsfit)
+```
+
+```{r}
+stop("Calculate hypo")
+
 ```
@@
 ```
+
+```{r}
+hhmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
+```
+
 ```{r}
@@
 summary(c_trt_m2_fit$brmsfit)
+```
+
+```{r}
+hhmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+
 ```
@@
 ```{r}
+hhmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious")
+
+```
+
+```{r}
 if(do_pp_checks) {
@@
 ```{r}
-# draws <- tidybayes::tidy_draws(c_trt_m4_fit$brmsfit)
-# hmm_hypothesis_res_list[["c_trt_m4_death"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_reduces_death,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
-# 
-# hmm_hypothesis_res_list[["c_trt_m4_discharged"]] <- 
-#   bayesian_hypothesis_res_from_draws(
-#     draws = draws$b_took_hcqTRUE + draws$`b_rate_groupimprove_one:took_hcqTRUE`,
-#     model = "HMM",
-#     estimand = "log(HR)",
-#     hypothesis = hypotheses$hcq_increases_discharged,
-#     adjusted = "age, sex, supportive, hospital, comorbidities_sum"
-#   )
+hhmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious")
+
 ```
@@
 ```{r}
-hmm_hypothesis_res_all <- do.call(rbind, hmm_hypothesis_res_list)
+hmm_hypothesis_res_all <- do.call(rbind, hhmm_hypothesis_res_list)
 write_csv(hmm_hypothesis_res_all, path = here::here("manuscript", "hmm_res.csv"))

### 2020-10-31 R/predict_hmm.R

@@ -1,6 +1,6 @@
 #'
 #' @return Array with dimensions time, serie, sample
-posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL) {
+posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL, method = posterior_epred_rect_simulate) {
   validate_brmshmmfit(fit)
 
   brms:::contains_samples(fit$brmsfit)
@@ -14,20 +14,21 @@ posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL) {
   data_hmm <- make_data_hmm(pred_rawdata)
 
   epred_mu <- brms::posterior_epred(fit$brmsfit, newdata = data_hmm$brmsdata, nsamples = nsamples)
+  if(any(is.na(epred_mu))) {
+    stop("NAs in epred_mu")
+  }
   transition_matrices <- compute_all_transition_matrices(data_hmm, epred_mu)
 
   if(is.null(nsamples)) {
     nsamples <- dim(epred_mu)[1]
   }
 
-  result_rect <- array(NA_integer_, c(max(pred_rawdata$serie_data$.time), max(as.integer(pred_rawdata$serie_data$.serie)), nsamples))
-
   max_times <- pred_rawdata$serie_data %>% group_by(.serie) %>% summarise(max_time = max(.time)) %>%
     arrange(.serie) %>% pull(max_time)
 
-  posterior_epred_rect_simulate(
-    nsamples = nsampes,
-    initial_states = pred_rawdata$initial_states[s],
+  method(
+    nsamples = nsamples,
+    initial_states = pred_rawdata$initial_states,
     max_times = max_times,
     transition_matrices = transition_matrices,
     predictor_sets_rect = data_hmm$standata$predictor_sets_rect)
@@ -68,6 +69,7 @@ posterior_epred_rect_simulate <- function(nsamples, max_times, initial_states, t
 
 }
 
+#' @return array with dimensions n_states, n_time, n_samples, n_series
 posterior_epred_state_prob <- function(nsamples, max_times, initial_states, transition_matrices, predictor_sets_rect) {
   if(length(max_times) != length(initial_states)) {
     stop("Incompatible lengths")

### 2020-10-31 manuscript/hmm.Rmd
Optimizing HMM hypotheses WIP

@@
 do_pp_checks <- FALSE
-hhmm_hypothesis_res_list <- list()
+hmm_hypothesis_res_list <- list()
 
@@
 ```{r}
-compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_data, denominator_serie_data, day = 28) {
-  if(nrow(numerator_serie_data) != nrow(denominator_serie_data)) {
-    stop("Groups to compare must have equal number of rows")
-  }
-
-  numerator_data <- hmm_fit$data
-  numerator_data$serie_data = numerator_serie_data
-
-  numerator_state_probs <- posterior_epred_rect(hmm_fit, newdata = numerator_data, method = posterior_epred_state_prob)
-
-  denominator_data <- hmm_fit$data
-  denominator_data$serie_data = denominator_serie_data
-
-  denominator_state_probs <- posterior_epred_rect(hmm_fit, newdata = denominator_data, method = posterior_epred_state_prob)
-
-  mortality_state <- which(hmm_fit$data$hidden_state_data$id == "Death")
-  samples_probs_numerator <- 
-    rowMeans(numerator_state_probs[mortality_state, day + 1, ,])
-  samples_probs_denominator <- 
-    rowMeans(denominator_state_probs[mortality_state, day + 1, ,])
-
-  log_odds_numerator <- log(samples_probs_numerator) - log1p(-samples_probs_numerator)
-  log_odds_denominator <- log(samples_probs_denominator) - log1p(-samples_probs_denominator)
-
-  log_odds_numerator - log_odds_denominator
-}
-
-evaluate_treatment_hypothesis <- function(fit, hypothesis, treatment_column, model_subgroup, adjusted, model_check) {
-  counterfactual_took <- serie_data_28 %>% mutate({{ treatment_column }} := 1)
-  counterfactual_no <- serie_data_28 %>% mutate({{treatment_column }} := 0)
-
-  log_OR_samples <- compute_counterfactual_mortality_log_OR(fit, counterfactual_took, counterfactual_no, day = 28)
-
-  bayesian_hypothesis_res_from_draws(
-    draws = log_OR_samples,
-    model = paste0("HMM", model_subgroup),
-    estimand = "log(OR)",
-    hypothesis = hypothesis,
-    adjusted = adjusted,
-    model_check = model_check
-  )
-}
-
-evaluate_all_treatment_hypotheses <- function(fit, model_subgroup, adjusted, model_check) {
-  rbind(
-    evaluate_treatment_hypothesis(fit, hypotheses$hcq_reduces_death, took_hcq, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check),
-    evaluate_treatment_hypothesis(fit, hypotheses$az_reduces_death, took_az, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check),
-    evaluate_treatment_hypothesis(fit, hypotheses$favipiravir_reduces_death, took_favipiravir, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check)
-  )
-}
 
@@
 hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
+```
+
+```{r}
+read_csv("hmm_res.csv")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
+hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
+hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
+hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
+hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
 ```
@@
 draws <- tidybayes::tidy_draws(d_dimer_m1_fit$brmsfit)
-hhmm_hypothesis_res_list[["d_dimer_m1_death"]] <-
+hmm_hypothesis_res_list[["d_dimer_m1_death"]] <-
   bayesian_hypothesis_res_from_draws(
@@
 draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
-hhmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <-
+hmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <-
   bayesian_hypothesis_res_from_draws(
@@
 ```{r}
-hhmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
 ```
@@
 ```{r}
-hhmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
 
@@
 ```{r}
-hhmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious")
 
@@
 ```{r}
-hhmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious")
 
@@
 ```{r}
-hmm_hypothesis_res_all <- do.call(rbind, hhmm_hypothesis_res_list)
+hmm_hypothesis_res_all <- do.call(rbind, hmm_hypothesis_res_list)
 write_csv(hmm_hypothesis_res_all, path = here::here("manuscript", "hmm_res.csv"))

### 2020-10-31 R/predict_hmm.R

@@ -219,6 +219,7 @@ posterior_long_to_df <- function(data, prediction) {
 #'
 #' @return An array of dimensions `N_states_hidden` x `N_states_hidden` x `N_predictor_sets` x `N_samples`
 compute_all_transition_matrices <- function(data_hmm, epred_mu) {
+  start_time = proc.time()
   nsamples <- dim(epred_mu)[1]
   epred_mu_exp <- exp(epred_mu)
   standata <- data_hmm$standata
@@ -237,5 +238,7 @@ compute_all_transition_matrices <- function(data_hmm, epred_mu) {
       res[,, ps, i] <-  expm::expm(rate_matrix)
     }
   }
+  time_taken <- proc.time() - start_time
+  message(paste0("Transition matrices computed in ", time_taken["elapsed"], "s."))
   res
 }

### 2020-10-31 manuscript/hmm.Rmd
Updated HMM results

@@
 hmm_hypothesis_res_list <- list()
+
+cl <-  parallel::makePSOCKcluster(parallel::detectCores())
 
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
 ```
@@
 ```{r}
-read_csv("hmm_res.csv")
+read_csv("D:/Google Drive/MBÚ/Covid/hmm_res.csv")
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+
+hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
+hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
+hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK")
+hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK")
+hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl)
 ```
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_log_peak_d_dimer,
+    draws = draws$`r_rate_group[death,log_peak_d_dimer]`,
     model = "HMM",
@@
     hypothesis = hypotheses$d_dimer_increases_death,
-    adjusted = "supportive, age, sex, hospital"
+    adjusted = "supportive, hospital",
+    model_check = "Suspicious"
   )
@@
 ```{r}
-stop("Calculate hypo")
+draws <- tidybayes::tidy_draws(d_dimer_m2_fit$brmsfit)
+hmm_hypothesis_res_list[["d_dimer_m2_death"]] <-
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$`r_rate_group[death,log_peak_d_dimer]`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$d_dimer_increases_death,
+    adjusted = "supportive, age, sex, hospital",
+    model_check = "Suspicious"
+  )
+
 ```
@@
 ```{r}
-stop("Calculate hypo")
-
-draws <- tidybayes::tidy_draws(trt_m7_fit$brmsfit)
-hmm_hypothesis_res_list[["trt_m7_IL_6_death"]] <-
+draws <- tidybayes::tidy_draws(IL_6_m1_fit$brmsfit)
+hmm_hypothesis_res_list[["IL_6_m1_death"]] <-
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_log_peak_IL_6,
+    draws = draws$`r_rate_group[death,log_peak_IL_6]`,
     model = "HMM",
@@
     hypothesis = hypotheses$IL_6_increases_death,
-    adjusted = "age, sex, hospital"
+    adjusted = "supportive, hospital"
   )
@@
 ```{r}
-stop("Calculate hypo")
+draws <- tidybayes::tidy_draws(IL_6_m2_fit$brmsfit)
+hmm_hypothesis_res_list[["IL_6_m2_death"]] <-
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$`r_rate_group[death,log_peak_IL_6]`,
+    model = "HMM",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$IL_6_increases_death,
+    adjusted = "supportive, age, sex, hospital"
+  )
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious", cl = cl)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious")
+hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl)
 

### 2020-10-31 manuscript/manuscript.Rmd
Updated HMM results

@@
 Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination.
-To ~~assess~~ *__explore__* association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
+To explore association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
 Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
@@
 author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
+---
 
@@
 library(patchwork)
-library(tsiMisc)
+library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")
 theme_set(cowplot::theme_cowplot())

### 2020-10-31 R/predict_hmm.R

@@ -1,6 +1,7 @@
 #'
 #' @return Array with dimensions time, serie, sample
-posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL, method = posterior_epred_rect_simulate) {
+posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL, method = posterior_epred_rect_simulate,
+                                 cl = NULL, cores = parallel::detectCores()) {
   validate_brmshmmfit(fit)
 
   brms:::contains_samples(fit$brmsfit)
@@ -17,7 +18,7 @@ posterior_epred_rect <- function(fit, nsamples = NULL, newdata = NULL, method = 
   if(any(is.na(epred_mu))) {
     stop("NAs in epred_mu")
   }
-  transition_matrices <- compute_all_transition_matrices(data_hmm, epred_mu)
+  transition_matrices <- compute_all_transition_matrices(data_hmm, epred_mu, cl = cl, cores = cores)
 
   if(is.null(nsamples)) {
     nsamples <- dim(epred_mu)[1]
@@ -218,14 +219,15 @@ posterior_long_to_df <- function(data, prediction) {
 
 #'
 #' @return An array of dimensions `N_states_hidden` x `N_states_hidden` x `N_predictor_sets` x `N_samples`
-compute_all_transition_matrices <- function(data_hmm, epred_mu) {
+compute_all_transition_matrices <- function(data_hmm, epred_mu, cl = NULL, cores = parallel::detectCores()) {
   start_time = proc.time()
   nsamples <- dim(epred_mu)[1]
   epred_mu_exp <- exp(epred_mu)
   standata <- data_hmm$standata
-  res <- array(NA_real_, c(standata$N_states_hidden, standata$N_states_hidden, standata$N_predictor_sets, nsamples))
-  for(i in 1:nsamples) {
-    for(ps in 1:standata$N_predictor_sets) {
+
+  compute_transition_matrices_ps <- function(ps, epred_mu_exp, standata) {
+    res <- array(NA_real_, c(standata$N_states_hidden, standata$N_states_hidden, nsamples))
+    for(i in 1:nsamples) {
       rate_matrix <- matrix(0, standata$N_states_hidden, standata$N_states_hidden)
 
       rate_values <- epred_mu_exp[i, standata$rate_predictors[ps,]]
@@ -235,9 +237,36 @@ compute_all_transition_matrices <- function(data_hmm, epred_mu) {
[…]

### 2020-11-03 manuscript/analysis_brms.Rmd
Inching towards a complete draft

@@
     hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   )
@@
     hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   )
@@
     hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),

### 2020-11-03 manuscript/competing_risks.Rmd
Inching towards a complete draft

--- 2020-10-17 manuscript/competing_risks.Rmd
@@
 ```{r}
+f_competing_hcq_az_fpv <- update.formula(f_competing_hcq_az, . ~ . + took_favipiravir)
+fit_competing_hcq_az_fpv <- coxph(f_competing_hcq_az_fpv, data = data_surv, id = patient_id)
+
+fit_competing_hcq_az_fpv
+
+hypothesis_res_list[["competing_hcq_az_fpv"]] <-
+  rbind(
+  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_reduces_death, fit_competing_hcq_az_fpv,
+                                        coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az"),
+  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_increases_discharged, fit_competing_hcq_az_fpv,
+                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az")
+  )
+hypothesis_res_list[["competing_hcq_az_fpv"]]
+```
+
+```{r}
 f_competing_hcq_az_admitted <- update.formula(f_competing_hcq_az, . ~ . + admitted_for_covid)
@@
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = paste0(cache_dir, "/competing_risks_res.csv"))
+write_csv(hypothesis_res_all, path = here::here("manuscript", "/competing_risks_res.csv"))
 ```

### 2020-11-03 manuscript/manuscript.Rmd
Inching towards a complete draft

@@
 Finally we used a set of Bayesian regression models implemented with the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
-Those included overall survival, state at day 7 or 28 as either binary or ordinal outcome, Bayesian version of the Cox proportional-hazards model and time-to-event models.
+Those included overall survival, state at day 7 or 28 as either binary or categorical outcome and a Bayesian version of the Cox proportional-hazards model.
 
@@
 
-In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis.
-Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
+In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis, including treatments taken.
+The dataset includes 19 patients already reported in a study of inflammatory signatures of Covid-19 [@http://zotero.org/users/5567156/items/IA4FQCF3].
 
@@
 ```{r treatment_comb_table, echo=FALSE}
-pd<-data$patient_data
-x<-sort(table(paste0(
-  as.numeric(pd$ever_hcq),
-  as.numeric(pd$ever_az),
-  as.numeric(pd$ever_convalescent_plasma),
-  as.numeric(pd$ever_favipiravir),
-  as.numeric(pd$ever_remdesivir),
-  as.numeric(pd$ever_dexamethasone),
-  as.numeric(pd$ever_antibiotics))),dec=TRUE)
-x2<-c()
-nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
-for (i in 1:length(x)) {
-  x3<-c()
-  for (i2 in 1:nchar(names(x)[i])) {
-    x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
-  }
-  x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
-}
-rownames(x2)<-c()
-colnames(x2)<-c('N',nms)
-x2<-x2[,c(1,2,3,8,4:7)]
-knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
+# Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
+# Currently not showing - almost all of this information is visible in the table above. Will put in supplement
+#pd<-data$patient_data
+# x<-sort(table(paste0(
+#   as.numeric(pd$ever_hcq),
+#   as.numeric(pd$ever_az),
+#   as.numeric(pd$ever_convalescent_plasma),
+#   as.numeric(pd$ever_favipiravir),
+#   as.numeric(pd$ever_remdesivir),
+#   as.numeric(pd$ever_dexamethasone),
+#   as.numeric(pd$ever_antibiotics))),dec=TRUE)
+# x2<-c()
+# nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
+# for (i in 1:length(x)) {
+#   x3<-c()
+#   for (i2 in 1:nchar(names(x)[i])) {
+#     x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
+#   }
+#   x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
+# }
+# rownames(x2)<-c()
+# colnames(x2)<-c('N',nms)
+# x2<-x2[,c(1,2,3,8,4:7)]
+# knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
 ```
@@
 
-data_for_fig_progression %>% 
+fig_progression <- data_for_fig_progression %>% 
   filter(!is_outcome) %>%
@@
 
+fig_progression
 ```
@@
 
-((plot_marker_data(c("CRP" = "CRP"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme) /
+fig_markers <- ((plot_marker_data(c("CRP" = "CRP"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme) /
 (plot_marker_data(c("d_dimer" = "D-dimer"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme + hide_strip_theme) /
@@
 
+fig_markers
   #plot_marker_data(c("d_dimer" = "D-dimer", "IL_6" = "IL 6", "CRP" = "CRP"), log_transform = TRUE)
@@
 # plot_marker_data(c("CRP"), log_transform = TRUE, max_patients = 30)
-
-```
-
-```{r}
-bayesian_cols <- cols(
-  hypothesis = col_character(),
-  model = col_character(),
-  adjusted = col_character(),
-  estimand = col_character(),
-  widest_CI_excl_reference = col_double(),
-  point_estimate = col_double(),
-  ci_low = col_double(),
-  ci_high = col_double()
-)
-res_bayesian <- 
-  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
-        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
-        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
-        )
-
-frequentist_cols <- cols(
-  hypothesis = col_character(),
-  model = col_character(),
-  adjusted = col_character(),
-  p_value = col_double(),
-  ci_low = col_double(),
-  ci_high = col_double()
-)
-
-res_frequentist <- 
-  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols),
-        read_csv(here::here("manuscript", "single_outcome_coxph_res.csv"), col_types = frequentist_cols))
-
-all_hypo <- rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
-      res_frequentist %>% select(-p_value)) %>%
-    inner_join(hypotheses_df, by = c("hypothesis" = "name"))
 
@@
 # condition on the sum of comorbidities
+pd<-data$patient_data
 pdd<-within(pd,cnd<-ischemic_heart_disease+has_hypertension_drugs+heart_failure+COPD+    other_lung_disease+    renal_disease+  high_creatinin)
@@
 
+```{r}
+bayesian_cols <- cols(
+  hypothesis = col_character(),
+  model = col_character(),
+  adjusted = col_character(),
+  estimand = col_character(),
+  widest_CI_excl_reference = col_double(),
+  point_estimate = col_double(),
+  ci_low = col_double(),
+  ci_high = col_double()
+)
+res_bayesian <- 
+  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
+        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
+        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
+        )
+
+frequentist_cols <- cols(
+  hypothesis = col_character(),
+  model = col_character(),
+  adjusted = col_character(),
+  p_value = col_double(),
+  ci_low = col_double(),
+  ci_high = col_double()
+)
+
+res_frequentist <- 
+  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols),
+        read_csv(here::here("manuscript", "single_outcome_coxph_res.csv"), col_types = frequentist_cols))
+
+all_hypo <- rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
+      res_frequentist %>% select(-p_value)) %>%
+    inner_join(hypotheses_df, by = c("hypothesis" = "name"))
+
+all_hypo_drugs <- all_hypo  %>%
+  filter(model_check == "OK")
+
+```
+
 Here, we focus on Hydroxychloroquine and Azithromycin as it is the only treatment with larger sample size.
-We also investigate Favipiravir and convalescent plasma as those therapies are less well reported in the literature.
+We also investigate Favipiravir as it is less well reported in the literature.
 Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg.
@@
 Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose all but one of the patients receiving Favipiravir also received Hydroxychloroquine.
-Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
-Except for convalescent plasma, treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).
+Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).
 
@@
 
+#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
+
 marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
@@
 data$marker_data %>% 
-  filter(marker %in% c("az", "hcq", "favipiravir", "convalescent_plasma"), !is.na(value), value > 0) %>%
+  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
   mutate(marker = marker_map[marker]) %>%
@@
 
-The results need to be interpreted very carefully as the treatme  The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs).
+The results of the multiverse analysis for association between Hydrxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - we only report models that we did not find to have immediate problems representing the data well.
+Results for all models we tested are reported in the supplementary material.
+Notably, most models report that using Hydroxychloroquine is associated with lower risk of death, we must however bear in mind the potential bias noted in the previous section.
+Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero, indicating that quite likely better adjustments could drive the effect towards zero.
+The case of Hydroxychloroquine serves as a "control group" for our other results - since we know from randomized trials that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the observed effects are just a measure of bias in the data.
+Also, the models we used cannot discern the sign of association between Hydroxychloroquine and in hospital.
+Additionally, our models either cannot determine the sign of association between Azithromycin and the outcome or even show an increase in risk of death or length of hospitalization.
 
-```{r multiverseresultsdrugs, fig.height=9, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between treatments and main outcomes.
-Each row represents a model - HMM = Bayesian hidden-markov model, coxph = Frequentist competing risks model.
-Additionally the factors the model adjusted for are listed - supportive = best supportive care initiated, comorbidities_sum = total number of comorbidities, admitted = admitted for Covid-19, az = took Azithromycin, hcq = took Hydroxychloroquine.
-For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval."}
+```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients for association between treatments and main outcomes.
+Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)c.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
+Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
+= total number of comorbidities, AZ = took Azithromycin, HCQ = took Hydroxychloroquine, FPV = took Favipiravir, C.
+plasma = received convalescent plasma."}
 
@@
 
-plot_multiverse <- function(hypo_res) {
-  hypo_res %>% mutate(
-    id = rev(1:n()),
-    model_adjusted = paste0(model, " - ", adjusted),
-    label = factor(model_adjusted, levels = model_adjusted, labels = str_wrap(model_adjusted, 30)),
-    model_adjusted_fct = fct_reorder(factor(model_adjusted),  id)) %>%
-  ggplot(aes(x = point_estimate, xmin = ci_low, xmax = ci_high, y = model_adjusted_fct)) +
-  geom_vline(xintercept = 0, color = "blue") +
-  geom_point() +
-  geom_errorbarh() + 
-  facet_wrap(~caption, ncol = 1, scales = "free_y") +
-  scale_x_continuous("Model coefficient estimate") +
-  scale_y_discrete("")
-  #theme(axis.text.y = element_text(size = 6))
-}
+my_lims <- c(-5.5,3.5)
+multiverse_hcq1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "hcq_reduces_death"), x_limits = my_lims)
+multiverse_hcq2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "hcq_increases_discharged"), x_limits = my_lims)
+multiverse_az1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "az_reduces_death"), x_limits = my_lims)
+multiverse_az2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "az_increases_discharged"), x_limits = my_lims)
 
-multiverse_hcq <- plot_multiverse(all_hypo %>% filter(group == "hcq"))
-multiverse_az <- plot_multiverse(all_hypo %>% filter(group == "az"))
-
-(multiverse_hcq / multiverse_az) + plot_layout(heights = c(2.5,1.5))
+((multiverse_hcq1 & hide_axis_theme) / (multiverse_hcq2 & hide_axis_theme) / (multiverse_az1 & hide_axis_theme) / multiverse_az2) + plot_layout(heights = c(1, 0.65,0.8,0.45)) & theme(axis.title.y = element_blank())
 
@@
 
-```{r, fig.height=4}
-plot_multiverse(all_hypo %>% filter(group == "favipiravir"))
+Figure \@ref{multiverseresultsfpv} shows a similar analysis for Favipiravir.
+Most models exclude very strong association between negative outcomes and using Favipiravir, but our data are necessarily quite limited, which is reflected in the very wide uncertainties around estimates.
+The only exception is the Categorical model for hospitalization at 28 days and the Cox competing hazards model which allow for even strongly negative associations.
+
+```{r multiverseresultsfpv, fig.height=5, fig.width=8.5, fig.cap="Estimates of model coefficients for association between Favipiravir and main outcomes.
+Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)c.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
+Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
+= total number of comorbidities, AZ = took Azithromycin, HCQ = took Hydroxychloroquine, FPV = took Favipiravir, C.
+plasma = received convalescent plasma."}
+
+multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(group == "favipiravir"), adjustments_to_hide = "favipiravir") & theme(axis.title.y = element_blank())
+multiverse_fpv1
 ```
@@
 
-```{r multiverseresultsmarkers, fig.height=2.5, fig.width=6, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
-Each row represents a model - HMM = Bayesian hidden-markov model, jm = Joint longitudinal and time-to-event model.
-Additionally the factors the model adjusted for are listed - hcq = took Hydroxychloroquine.
+```{r multiverseresultsmarkers, fig.height=3, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
+Each row represents a model - HMM A = Bayesian hidden-markov model, JM = Joint longitudinal and time-to-event model.
+Additionally the factors the model adjusted for are listed - Site = the study site, Supportive = best supportive care initiated, HCQ = took Hydroxychloroquine.
 We show posterior mean and 95% credible interval."}
-plot_multiverse(all_hypo %>% filter(group == "markers"))
+plot_multiverse(all_hypo %>% filter(group == "markers", model_check != "Problematic"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"))
 ```
 
-## Prognostic models are not better using age only
+## Prognostic models are not better than just using age
 
@@
 
-Will be written once we have the full dataset.
+Our data support substantial between-patient variability in progression of the disease in terms of both length of hospital stay, duration of various types of breathing support and basic markers.
 
-An important lesson from our analysis is that the assesment of treatment efficacy from observational data is heavily sensitive to modelling assumptions while it is usually almost impossible to determine which of the models is more likely to reflect reality (if any).
+We provide very weak observational evidence against a substantial beneficiary effect of using Azithromycin (both with or without Hydroxychloroquine) and against substantial negative effect of using Favipiravir in hospitalized Covid-19 patients.
+
+Where our results contradict randomized trials, the most likely explanation is systematic bias in our dataset.
+
+A lesson from our analysis is that the assessment of treatment efficacy from observational data is very sensitive to modelling assumptions while it is usually almost impossible to determine which of the models is more likely to reflect reality (if any).
 We believe that using multiverse analysis is an appropriate way to explore data in such contexts as it lets us be transparent about this sensitivity.
 
-Where our resutls contradict randomized trials, the most likely explanation is systematic bias in our dataset.
+Additionally our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial.
+We suggest that comparing the prognostic models against simple baselines (e.g.
+decade of age as the single predictor) should be a first step in validation.
+Further, many of the published scores lack enough information to let others exactly implement the score.
 

### 2020-11-03 R/multiverse_tools.R

@@ -1,0 +1,131 @@
+plot_multiverse <- function(hypo_res, adjustments_to_hide = c(), x_limits = NULL) {
+  all_adjustments_dict <- c(
+                       "hospital" = "Site",
+                       "admitted" = "Admitted",
+                       "supportive" = "Supportive",
+                       "age" = "Age",
+                       "age_norm" = "Age",
+                       "sex" = "Sex",
+                       "smoking" = "Smoking",
+                       "obesity" = "Obesity",
+                       "comorbidities \\(sum\\)" = "Comorb.",
+                       "comorbidities" = "Comorb.",
+                       "hcq" = "HCQ",
+                       "az" = "AZ",
+                       "favipiravir" = "FPV",
+                       "all treatments" = "All treatments",
+                       "convalescent_plasma" = "C. plasma")
+  all_adjustments = names(all_adjustments_dict)
+
+  model_labels <- c("brms_categorical" = "Categorical (all)",
+                    "brms_categorical_7" = "Categorical 7",
+                    "brms_categorical_28" = "Categorical 28",
+                    "brms_cox" = "Bayes Cox",
+                    "coxph1" = "Cox (single)",
+                    "coxph" = "Cox (competing)",
+                    "HMMsimple_group" = "HMM A",
+                    "HMMsimple_rate" = "HMM B",
+                    "HMMcomplex_group" = "HMM C",
+                    "jm" = "JM"
+  )
+
+  single_adjustment_regex <- paste(all_adjustments, collapse = "|")
+  adjustment_regex <- paste0("^((",single_adjustment_regex,")(, |$))+")
+  recognized_adjustments <- grepl(adjustment_regex, hypo_res$adjusted)
+  if(!all(recognized_adjustments)) {
+    print(hypo_res$adjusted[!recognized_adjustments])
+    stop("Unexpected adjustment")
+  }
+
+  spacers <- hypo_res %>% group_by(model, hypothesis, estimand, caption, group, type) %>%
+    summarise(is_spacer = TRUE, point_estimate = NA, ci_low = NA, ci_high = NA, adjusted = NA,
+              model_check = NA, data_version = NA)
+
+  for(adj in all_adjustments) {
[…]

### 2020-11-05 manuscript/analysis_brms.Rmd
add some model checks

--- 2020-10-22 manuscript/analysis_brms.Rmd
@@
 
-cache_dir <- here::here("local_temp_data", "analysis_brms")
+cache_dir <- here::here("local_temp_data")
 if (!dir.exists(cache_dir)) {
-  dir.create(cache_dir, recursive = TRUE)
+  dir.create(cache_dir)
 }
@@
 }
-
-default_control <- list(adapt_delta = 0.95)
 ```
@@
   prior = prior_survival_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base")
@@
 summary(fit_survival_base)
+pp_check(fit_survival_base, "error_binned")
+loo(fit_survival_base)
 # ce <- conditional_effects(fit_survival_base)
@@
 ```{r}
-#Marking the results as suspicious because there was a divergence...
 draws <- tidy_draws(fit_survival_base)
@@
   prior = prior_survival_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base_7")
@@
 summary(fit_survival_base_7)
+pp_check(fit_survival_base_7, "error_binned")
+loo(fit_survival_base_7)
 # ce <- conditional_effects(fit_survival_base_7)
@@
     hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   ),
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   ),
@@
     hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   )
@@
   prior = prior_survival_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_survival_base_28")
@@
 summary(fit_survival_base_28)
+pp_check(fit_survival_base_28, "error_binned")
+loo(fit_survival_base_28)
 # ce <- conditional_effects(fit_survival_base_28)
@@
   inits = 0,
-  control = default_control,
   file = paste0(cache_dir, "/fit_survival_time")
@@
 summary(fit_survival_time)
+loo(fit_survival_time)
 ```
@@
     hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   ),
@@
     hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   ),
@@
     hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   )
@@
   prior = prior_discharged_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base")
@@
 summary(fit_discharged_base)
+pp_check(fit_discharged_base, "error_binned")
+loo(fit_discharged_base)
 # ce <- conditional_effects(fit_discharged_base)
@@
   prior = prior_discharged_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base_7")
@@
 summary(fit_discharged_base_7)
+pp_check(fit_discharged_base_7, "error_binned")
+loo(fit_discharged_base_7)
 # ce <- conditional_effects(fit_discharged_base_7)
@@
     hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   ),
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   ),
@@
     hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   )
@@
   prior = prior_discharged_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_discharged_base_28")
@@
 summary(fit_discharged_base_28)
+pp_check(fit_discharged_base_28, "error_binned")
+loo(fit_discharged_base_28)
 # ce <- conditional_effects(fit_discharged_base_28)
@@
   inits = 0,
-  control = list(adapt_delta = 0.98),
+  control = list(adapt_delta = 0.95),
   file = paste0(cache_dir, "/fit_discharged_time")
@@
 summary(fit_discharged_time)
+loo(fit_discharged_time)
 ```
@@
     hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   ),
@@
     hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   ),
@@
     hypothesis = hypotheses$favipiravi_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   )
@@
   prior = prior_outcome_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base")
@@
 summary(fit_outcome_base)
+loo(fit_outcome_base)
 # ce <- conditional_effects(fit_outcome_base)
@@
   prior = prior_outcome_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base_7")
@@
 summary(fit_outcome_base_7)
+pp_check(fit_outcome_base_7)
+loo(fit_outcome_base_7)
 # ce <- conditional_effects(fit_outcome_base_7)
@@
     hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
   ),
@@
   prior = prior_outcome_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_outcome_base_28")
@@
 summary(fit_outcome_base_28)
+pp_check(fit_outcome_base_28)
+loo(fit_outcome_base_28)
 # ce <- conditional_effects(fit_outcome_base_28)
@@
   prior = prior_worst_breathing_base,
-  control = default_control,
   file = paste0(cache_dir, "/fit_worst_breathing_base")
@@
 summary(fit_worst_breathing_base)
+loo(fit_worst_breathing_base)
 # ce <- conditional_effects(fit_worst_breathing_base, categorical = TRUE)
@@
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)
-write_csv(hypothesis_res_all, path = here::here("manuscript/analysis_brms.csv"))
+write_csv(hypothesis_res_all, path = paste0(cache_dir, "/analysis_brms.csv"))
 ```

### 2020-11-05 manuscript/analysis_brms.Rmd

--- 2020-11-03 manuscript/analysis_brms.Rmd
@@
     hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity"
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "some influencial observations"
   )

### 2020-11-05 manuscript/competing_risks.Rmd

@@
   frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_reduces_death, fit_competing_hcq_az_fpv,
-                                        coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az"),
+                                        coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az", model_check = "Problematic"),
   frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_increases_discharged, fit_competing_hcq_az_fpv,
-                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az")
+                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az", model_check = "Problematic")
   )
@@
 
-# PCR positivity predicts
+```{r}
+f_competing_marker_base <- Surv(daym1, day,status) ~ age + sex
+f_competing_il_6 <- update.formula(f_competing_marker_base, . ~ . + log_peak_IL_6)
 
-# Lymphocyte count predicts?
+data_IL_6 <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(IL_6))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "IL_6", new_column_name = "peak_IL_6", initial_value = 1) %>%
+  mutate(log_peak_IL_6 = log(peak_IL_6))
+
+fit_competing_il_6 <- coxph(f_competing_il_6, data = data_IL_6, id = patient_id, control = coxph.control(iter.max = 200))
+fit_competing_il_6
+
+hypothesis_res_list[["competing_il_6"]] <-
+  frequentist_hypothesis_res_from_coxph(hypotheses$IL_6_increases_death, fit_competing_il_6,
+                                        coefficient_name = "log_peak_IL_6", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
+hypothesis_res_list[["competing_il_6"]]
+```
+```{r}
+f_competing_d_dimer <- update.formula(f_competing_marker_base, . ~ . + log_peak_d_dimer)
+
+data_d_dimer <- data_surv %>%
+  group_by(patient_id) %>%
+  filter(any(!is.na(d_dimer))) %>%
+  ungroup() %>%
+  compute_marker_peak(column_name = "d_dimer", new_column_name = "peak_d_dimer", initial_value = 0) %>%
+  mutate(log_peak_d_dimer = log(peak_d_dimer + 1) - 4)
+
+fit_competing_d_dimer <- coxph(f_competing_d_dimer, data = data_d_dimer, id = patient_id, control = coxph.control(iter.max = 2000))
+fit_competing_d_dimer
+
+hypothesis_res_list[["competing_d_dimer"]] <-
+  frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_increases_death, fit_competing_d_dimer,
+                                        coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Problematic")
+hypothesis_res_list[["competing_d_dimer"]]
+```
 

### 2020-11-05 manuscript/hmm.Rmd

@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
 ```
@@
 
-hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl)
 ```
@@
     draws = draws$`r_rate_group[death,log_peak_d_dimer]`,
-    model = "HMM",
+    model = "HMMsimple_group",
     estimand = "log(HR)",
@@
     draws = draws$`r_rate_group[death,log_peak_IL_6]`,
-    model = "HMM",
+    model = "HMMsimple_group",
     estimand = "log(HR)",
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
 

### 2020-11-05 manuscript/manuscript.Rmd

@@
 Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination.
-To explore association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
-Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
+To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
+Weak evidence to constrain the potential efficacy of Azithromycin and Favipiravir can be extracted from the data.
+Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset."
 bibliography: Covid19.json
@@
 
-The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of September 2020 led to over 28 million cases and over 900 000 deaths.
-Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteriods, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
-Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteriod Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting (TODO cite).
-Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure profylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
+The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020 led to over 46 million cases and over 1.1 million deaths.
+Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteroids, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
+Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/NSTX4EZM].
+Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/NSTX4EZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
 Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
 Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
-No benefit was observed for convalscent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
+No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
 
-In observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e.
-g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9]
+Interestingly, in observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e.
+g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].
 
-Favipiravir is one of the less explored experimental treatmets, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
-Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/IHATMFTC; @http://zotero.org/users/5567156/items/894H9N9B].
+Favipiravir is one of the less explored experimental treatments, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
+Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B].
 None of the results allow strong conclusions about benefits or harms of Favipiravir.
@@
 
-High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/ID7F3XIH].
-Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the diesease.
+High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
+Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the disease.
 
-An ongoing challenge in evalutating Covid-19 treatments is that the analysis and interpretation of the data is often inappropriate or misleading, most notably interpreting lack of evidence due to small sample size as evidence of no effect [see e.g., @http://zotero.org/users/5567156/items/5PZEEPJV; @http://zotero.org/users/5567156/items/AZDZHLJV].
+An ongoing challenge in evaluating Covid-19 treatments is that the analysis and interpretation of the data is often inappropriate or misleading, most notably interpreting lack of evidence due to small sample size as evidence of no effect [see e.g., @http://zotero.org/users/5567156/items/5PZEEPJV; @http://zotero.org/users/5567156/items/AZDZHLJV].
 
 Additionally, many methods for predicting disease severity of Covid-19 were published, but the methods are at high risk of bias and lack external validation [@http://zotero.org/users/5567156/items/TS57HJ3T].
+
+The present study aims to describe the outcomes and disease course of hospitalized patients with mild to severe clinical presentation in a multicentric Czech cohort, explore the association between the outcomes and pharmacological interventions and to provide external validation to previously published prognostic models for Covid-19 severity.
 
@@
 ethics_string <- ethics_data %>% pull(Site) %>% paste0(collapse = ", ")
+
+n_ecmo <- data$marker_data_wide %>% filter(breathing == "ECMO") %>% pull(patient_id) %>% unique() %>% length()
+n_nippv <- data$marker_data_wide %>% filter(breathing == "NIPPV") %>% pull(patient_id) %>% unique() %>% length()
 ```
@@
 Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.
-We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
+
+We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
 Full protocol for data collection is attached in the supplementary material.
+Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we collapsed those categories with mechanical ventilation.
+
+Not all patients developed pneumonia or other symptoms of Covid-19.
+All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections.
+Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of attending physician and underlying patient condition.
+We note that the participating sites were not homogeneous in either patient population or treatment protocols.
+The choice of treatment was based on the decision of the attending clinician and availability.
+
 The study was approved by the Ethical committees of `r ethics_string`.
@@
 
-It shall be noted that the character of the convenience sample did not allow to properly assess 
-the association between treatments, markers and patient outcomes,
-because the treatments had not been assigned to patients at random (as in a randomized clinical trial), 
+The character of the convenience sample does not allow for a proper assessment of
+the association between treatments and patient outcomes,
+because the treatments had not been assigned to patients at random, 
 but were only observed retrospectively.
-Therefore, we could not exclude the possibility of accidental association of treatments to patients' baseline characteristics, 
-which could, in turn, demonstrate in spurious association between treatments and outcomes, 
-because both of these can be driven by the patients' characteristics. 
-Eventhough we attempted to account for possible associations of treatments to patients' baseline characteristics, 
-we still could not rule them out, and the results of our analyses should be considered as exploratory.
+This can be partially remedied by adjusting for patient characteristic in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.
 
 Since many details of analysis may influence the conclusions made, we performed multiverse analysis [@http://zotero.org/users/5567156/items/DXWM259V] and report results for all the hypothesis tested across multiple different models using both frequentist and Bayesian paradigms.
-For each model class we worked with several possible sets of adjustments and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC].
-All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualisations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
+For each model class we worked with several possible sets of adjustments.
+All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualizations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
+
+```{r}
+ # and - where applicable - with multiple estimands of the outcome as different tasks require different estimands [@http://zotero.org/users/5567156/items/MH797GNC]
+```
 
@@
 Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
-The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighbouring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
+The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
 The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths.
@@
 We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction.
+All of the hidden Markov models take into account whether best supportive care was initiated and a patient was thus not indicated to progress to more intensive treatment modalities.
 
 ```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c).
-AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation.
+AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation.
 In all models the 'Death' and 'Discharged' states are terminal.
@@
 
-When markers are used as predictors in both the hidden Markov models and the competing risks model, we impute the missing values using either the maximum of the marker so far or the last-known value of the marker.
+<!-- When markers are used as predictors in both the hidden Markov models and the competing risks model, we impute the missing values using either the maximum of the marker so far or the last-known value of the marker.
+-->
 
-Additionally we used joint models for longitudinal and time-to-event data from `stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5].
+<!-- Additionally we used joint models for longitudinal and time-to-event data via the `stan_jm` function in the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5].
 Those models include a longitudinal component to model the evolution over time of marker for each patient and a time-to-event component which assumes proportional-hazards for an event.
 The estimated value of the marker is then used as one of the predictors in the time-to-event component.
+-->
 
@@
 where we have gathered all the input features).
-We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67].
-When there were multiple reasonable ways to evaluate either outcome or a predictor in our dataset, we computed and reported all of those options.
-As a baseline to compare the scores against we used age and the decade of age as the sole predictor.
+We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and its bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67].
+When there were multiple reasonable ways to evaluate the outcome or a predictor in our dataset, we computed and reported all of those options.
+We used two simple scores with age or the decade of age as the sole predictor to have a simple baseline to compare the scores against.
 
@@
   data$patient_data %>% mutate(group = if_else(ever_hcq, "HCQ", "No HCQ")),
-  data$patient_data %>% filter(any_d_dimer) %>% mutate(group = "Measured D-dimer"),
-  data$patient_data %>% filter(any_IL_6) %>% mutate(group = "Measured IL-6")
+  data$patient_data %>% filter(ever_favipiravir) %>% mutate(group = "Favipiravir")
+  # data$patient_data %>% filter(any_d_dimer) %>% mutate(group = "Measured D-dimer"),
+  # data$patient_data %>% filter(any_IL_6) %>% mutate(group = "Measured IL-6")
   )
@@
     `BMI (mean, IQR)` = mean_iqr(BMI),
+    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
     Deceased = sum_percent(outcome == "Death"),
@@
   ) %>%
-  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Measured D-dimer","Measured IL-6"), ordered = TRUE)) %>%
+  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
   arrange(group) %>%
@@
   t() %>% 
-  knitr::kable(caption = "Patient characteristics. Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.", booktabs = TRUE)
+  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups.
+Note that the Favipiravir subgroup is not exclusive with either the HCQ or No HCQ group.
+IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)
+
+# Note that the 'Measured D-dimer' and 'Measured IL-6' are neither exclusive nor exhaustive.
 ```
@@
 
-In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure &nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
+In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
 The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization.
+The limited nature of the data however does not allow for any statistically robust conclusions.
 We also see that the marker levels were not substantially stratified by study site.
-Those should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.
+Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.
+However we believe this kind of patient-level view is useful to appreciate both the extent of between- and within-patient variability.
 
 ```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
-Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
+Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
 my_paste <- function(x) {
@@
 Each line represents a patient, stratified by the worst breathing support required over the hospitalization.
-Color indicates study sites, notably  The vertical scale is logarithmic.
-Ventilated = any form of vantilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
+Color indicates study sites.
+The vertical scale is logarithmic.
+Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
 #data$marker_data %>%
@@
 
-```{r hcq_assoc, message=FALSE, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be asociated with the status of patients upon hospitalization.
-Comorbidities were asociated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red).
+```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be associated with the status of patients upon hospitalization.
+Comorbidities were associated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red).
 HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
@@
 
-Figure&nbsp;\@ref(fig:hcq_assoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
+Figure&nbsp;\@ref(fig:hcqassoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
 were less likely to receive Hydroxychloroquine treatment. 
@@
 Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
-the score representing the cummulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
+the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
 other lung diseases, renal disease, or high creatinin 
-was asscociated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
+was associated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
 (the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
@@
 
-Here, we focus on Hydroxychloroquine and Azithromycin as it is the only treatment with larger sample size.
+Here, we focus on Hydroxychloroquine and Azithromycin as those are the only treatment with larger sample size.
 We also investigate Favipiravir as it is less well reported in the literature.
@@
 Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.
-Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose all but one of the patients receiving Favipiravir also received Hydroxychloroquine.
+Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose.
+All but one of the patients receiving Favipiravir also received Hydroxychloroquine.
 Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).
@@
 
-The results of the multiverse analysis for association between Hydrxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - we only report models that we did not find to have immediate problems representing the data well.
+The results of the multiverse analysis for association between Hydroxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that we did not find to have immediate problems representing the data well or computational issues.
 Results for all models we tested are reported in the supplementary material.
 Notably, most models report that using Hydroxychloroquine is associated with lower risk of death, we must however bear in mind the potential bias noted in the previous section.
-Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero, indicating that quite likely better adjustments could drive the effect towards zero.
-The case of Hydroxychloroquine serves as a "control group" for our other results - since we know from randomized trials that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the observed effects are just a measure of bias in the data.
-Also, the models we used cannot discern the sign of association between Hydroxychloroquine and in hospital.
+Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero.
+This is a weak indication that further adjustments could drive the effect towards zero.
+We did not attempt to model additional adjustments as the models became computationally intractable.
+The case of Hydroxychloroquine serves as a "control group" for our other results - since randomized trials give us high confidence that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the associations we observe for Hydroxychloroquine are just a measure of bias in the data.
+The models we used cannot discern the sign of association between Hydroxychloroquine and length of hospital stay.
 Additionally, our models either cannot determine the sign of association between Azithromycin and the outcome or even show an increase in risk of death or length of hospitalization.
+This serves as a weak evidence that a substantial improvement from Azithromycin is unlikely.
 
@@
 The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
+Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
 Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
 = total number of comorbidities, AZ = took Azithromycin, HCQ = took Hydroxychloroquine, FPV = took Favipiravir, C.
-plasma = received convalescent plasma."}
+plasma = received convalescent plasma.
+"}
 
@@
 
-((multiverse_hcq1 & hide_axis_theme) / (multiverse_hcq2 & hide_axis_theme) / (multiverse_az1 & hide_axis_theme) / multiverse_az2) + plot_layout(heights = c(1, 0.65,0.8,0.45)) & theme(axis.title.y = element_blank())
+# multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_reduces_death"))
+# multiverse_fpv2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_increases_discharged"))
+
+# ((multiverse_hcq1 & hide_axis_theme) / (multiverse_hcq2 & hide_axis_theme) / (multiverse_az1 & hide_axis_theme) / (multiverse_az2 & hide_axis_theme) / (multiverse_fpv1 & hide_axis_theme) / (multiverse_fpv2)) + plot_layout(heights = c(1, 0.65,0.8,0.4, 0.8,0.3)) & theme(axis.title.y = element_blank())
+
+((multiverse_hcq1 & hide_axis_theme) / (multiverse_hcq2 & hide_axis_theme) / (multiverse_az1 & hide_axis_theme) / (multiverse_az2))  + plot_layout(heights = c(1, 0.65,0.8,0.4)) & theme(axis.title.y = element_blank())
 
@@
 
-Figure \@ref{multiverseresultsfpv} shows a similar analysis for Favipiravir.
-Most models exclude very strong association between negative outcomes and using Favipiravir, but our data are necessarily quite limited, which is reflected in the very wide uncertainties around estimates.
-The only exception is the Categorical model for hospitalization at 28 days and the Cox competing hazards model which allow for even strongly negative associations.
+Figure \@ref(fig:multiverseresultsfpv) shows a similar analysis for Favipiravir.
+Most models exclude very strong association between death/hospitalization and using Favipiravir, but our data are necessarily quite limited, which is reflected in the very wide uncertainties around estimates.
+The only exception is the Categorical model  at 28 days which allows for substantially higher risk of being hospitalized in the Favipiravir group.
 
-```{r multiverseresultsfpv, fig.height=5, fig.width=8.5, fig.cap="Estimates of model coefficients for association between Favipiravir and main outcomes.
-Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)c.
-For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
-The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
-Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
-= total number of comorbidities, AZ = took Azithromycin, HCQ = took Hydroxychloroquine, FPV = took Favipiravir, C.
-plasma = received convalescent plasma."}
+```{r multiverseresultsfpv, fig.height=3.3, fig.width=8.5, fig.cap="Estimates of model coefficients for association between Favipiravir and main outcomes.
+Abbreviations and other elements as in Figure&nbsp;\\@ref(fig:multiverseresultsdrugs)."}
 
-multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(group == "favipiravir"), adjustments_to_hide = "favipiravir") & theme(axis.title.y = element_blank())
-multiverse_fpv1
+fpv_lims <- c(-12,2)
+
+multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_reduces_death"), x_limits = fpv_lims, adjustments_to_hide = "FPV")
+multiverse_fpv2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_increases_discharged"), x_limits = fpv_lims, adjustments_to_hide = "FPV")
+
+((multiverse_fpv1 & hide_axis_theme) / (multiverse_fpv2)) + plot_layout(heights = c(0.8,0.3)) & theme(axis.title.y = element_blank())
+
 ```
 
-## Association between markers and outcomes
-
-We additionally investigated the association of several markers with disease progression.
-Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice.
+```{r}
+# ## Association between markers and outcomes
+# 
+# We additionally investigated the association of several markers with disease progression.
+Due to low number of patients in the current dataset, the results are inconclusive and dependent on model choice, many models also proved to be problematic to fit to this reduced dataset.
 All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)
-
-```{r multiverseresultsmarkers, fig.height=3, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
-Each row represents a model - HMM A = Bayesian hidden-markov model, JM = Joint longitudinal and time-to-event model.
+# 
+# 
+# ``{r multiverseresultsmarkers, fig.height=3.5, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
+Each row represents a model - Cox (competing) =  frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, JM = Bayesian joint longitudinal and time-to-event model.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
 Additionally the factors the model adjusted for are listed - Site = the study site, Supportive = best supportive care initiated, HCQ = took Hydroxychloroquine.
 We show posterior mean and 95% credible interval."}
-plot_multiverse(all_hypo %>% filter(group == "markers", model_check != "Problematic"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"))
+# plot_multiverse(all_hypo %>% filter(group == "markers", model_check != "Problematic"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"))
+# ``
+
 ```
-
 ## Prognostic models are not better than just using age
@@
 Overall all prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
-Additionally, most publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed.
+Additionally, some publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed.
 We thus concur with [@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.
@@
 ```{r riskscores, fig.cap="Performance of tested prediction scores as measured by AUC.
-AUC = 1 means perfect prediction while AUC <= 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the predictions (marked by thick blue line).
+AUC = 1 means perfect prediction while AUC <= 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the score (marked by thick blue line).
 The line ranges represent the bootstrapped 95% confidence intervals.
@@
 Cyan triangles show AUC as reported by the original authors or recomputed based on their published data.
-When the confidence interval or the AUC of the original study is not reported, it means that the value was not reported by the authors and not enough information to recompute it was given."}
+When the confidence interval or the AUC of the original study is not shown, it means that the value was not reported by the authors and not enough information to recompute it was given."}
 col_types_auc <- cols(
@@
 
-Our data support substantial between-patient variability in progression of the disease in terms of both length of hospital stay, duration of various types of breathing support and basic markers.
+Our data show the extent of between-patient variability in progression of the disease in terms of both length of hospital stay, duration of various types of breathing support and basic markers.
 
 We provide very weak observational evidence against a substantial beneficiary effect of using Azithromycin (both with or without Hydroxychloroquine) and against substantial negative effect of using Favipiravir in hospitalized Covid-19 patients.
-
 Where our results contradict randomized trials, the most likely explanation is systematic bias in our dataset.
@@
 We believe that using multiverse analysis is an appropriate way to explore data in such contexts as it lets us be transparent about this sensitivity.
+We further believe that using Hidden Markov models is a promising complement to the standard Cox proportional hazards analysis when detailed information on disease progression is available, particularly because it lets us impose additional structure on the model and thus make inferences with more disease states than would be possible to handle in the Cox framework, making better use of the available data.
 
@@
 decade of age as the single predictor) should be a first step in validation.
-Further, many of the published scores lack enough information to let others exactly implement the score.
+Further, some of the published scores lack enough information to let others exactly implement the score.
+
+# Acknowledgements
+
+This work was supported by ELIXIR CZ research infrastructure project (MEYS Grant No: LM2018131) including access to computing and storage facilities.
 

### 2020-11-05 manuscript/risk_scores.Rmd

@@
 all_auc <- rbind(acp_auc, chen_liu_auc, shi_et_al_auc, caramelo_auc, bc_auc, age_auc)
-compare_auc_to_orig(rbind(acp_auc_orig, chen_liu_auc_orig, shi_et_al_auc_reported_admission,
-                         shi_et_al_auc_reported_all, bc_auc_orig),
+all_auc_orig <- rbind(acp_auc_orig, chen_liu_auc_orig, shi_et_al_auc_reported_admission,
+                         shi_et_al_auc_reported_all, bc_auc_orig)
+
+write_csv(all_auc, here::here("manuscript", "all_auc.csv"), na = "")
+write_csv(all_auc_orig, here::here("manuscript", "all_auc_orig.csv"), na = "")
+
+compare_auc_to_orig(all_auc_orig,
                     all_auc, show_outcome = FALSE)

### 2020-11-14 manuscript/hmm.Rmd
Added first wave checks to HMM

@@
 
+## Treatments + age + sex + hospital, first wave only
+
+```{r trt_group_m3_fw}
+serie_data_fw <- serie_data %>% filter(first_wave)
+simple_initial_states_fw <- serie_data_fw %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+
+trt_group_m3_fw <- brmshmmdata(trt_group_m3_formula, serie_data_fw, 
+                  simple_rate_data, trt_group_m1_hidden_state_data, simple_initial_states_fw, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
+
+trt_group_m3_fw_fit <- brmhmm(trt_group_m3_fw, cache_file = paste0(cache_dir,"/trt_group_m3_fw.rds"), init = 0.1, control = default_control, iter = 1200)
+
+summary(trt_group_m3_fw_fit$brmsfit)
+```
+
+```{r}
+hmm_hypothesis_res_list[["trt_group_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(trt_group_m3_fw_fit)
+}
+```
+
 ## Treatment + age + sex + hospital + comorbidities
@@
 hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl)
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(trt_group_m4_fit)
+}
+```
+
+## Treatment + age + sex + hospital + comorbidities, first wave
+
+```{r trt_group_m4_fw}
+trt_group_m4_fw <- brmshmmdata(trt_group_m4_formula, serie_data_fw, 
+                  simple_rate_data, trt_group_m1_hidden_state_data, simple_initial_states_fw, prior = trt_group_m1_prior,
+                  observed_state_data = trt_group_m1_observed_state_data)
+
+trt_group_m4_fw_fit <- brmhmm(trt_group_m4_fw, cache_file = paste0(cache_dir,"/trt_group_m4_fw.rds"), init = 0.1, control = default_control, iter = 1200)
+
+summary(trt_group_m4_fw_fit$brmsfit)
+```
+
+```{r}
+hmm_hypothesis_res_list[["trt_group_m4_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities, first_wave", model_check = "OK", cl = cl)
 ```
@@
   do_common_pp_checks(trt_rate_m3_fit)
+}
+```
+
+## Treatments + age + sex + hospital, first_Wave
+
+```{r trt_rate_m3_fw}
+trt_rate_m3_fw <- brmshmmdata(trt_rate_m3_formula, serie_data_fw, 
+                  trt_rate_m1_rate_data, trt_rate_m1_hidden_state_data, simple_initial_states_fw, prior = trt_rate_m1_prior,
+                  observed_state_data = trt_rate_m1_observed_state_data)
+
+trt_rate_m3_fw_fit <- brmhmm(trt_rate_m3_fw, cache_file = paste0(cache_dir,"/trt_rate_m3_fw.rds"), init = 0.1, control = default_control, iter = 1500)
+
+summary(trt_rate_m3_fw_fit$brmsfit)
+```
+
+```{r}
+hmm_hypothesis_res_list[["trt_rate_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fw_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(trt_rate_m3_fw_fit)
 }
@@
 
+## Treatments + age + sex + hospital, first wave
+
+```{r c_trt_m3_fw}
+complex_initial_states_fw <- serie_data_fw %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed) %>% paste0(., "_worsening")
+
+c_trt_m3_fw <- brmshmmdata(trt_group_m3_formula, serie_data_fw, 
+                  complex_rate_data, complex_hidden_state_data, complex_initial_states_fw, prior = trt_group_m1_prior,
+                  observed_state_data = base_observed_state_data)
+
+c_trt_m3_fw_fit <- brmhmm(c_trt_m3_fw, cache_file = paste0(cache_dir,"/c_trt_m3_fw.rds"), init = 0.1, control = default_control, iter = 1200)
+
+summary(c_trt_m3_fw_fit$brmsfit)
+```
+
+```{r}
+hmm_hypothesis_res_list[["c_trt_rate_m3_fw"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fw_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+
+```
+
+```{r}
+if(do_pp_checks) {
+  do_common_pp_checks(c_trt_m3_fit)
+}
+```
+
 ## Treatment + age + sex + hospital + comorbidities

### 2020-11-19 manuscript/hmm.Rmd
HMM for supplementary, WIP

@@
 ```
+
+# The overall approach for hidden Markov models
+
+To complement the more traditionally used proportional hazards models we also use several variants of hidden Markov models (HMMs).
+A discrete random process $X_1, X_2, X_3, ...$  satisfies the Markov property whenever for all states $x$ and times $t$ we have $P(X_{t + 1} = x | X_1 = x_1, X_2 = x_2, ..., X_t = x_t) = P(X_{t+1} = x | X_t = x_t)$, i.e.
+when the process is "memoryless".
+If this process is unobservable, but we can observe another process $Y$ such that $P(Y_{t} = y | X_1 = x_1, X_2 = x_2, ..., X_n = x_n) = P(Y_{t} = y | X_t = x_t)$ we call the pair $(X, Y)$ a hidden Markov process.
+
+We treat the breathing support required as a Markov process, either with states directly corresponding to the breathing support and directly observable (Fig.
+\@ref(fig:hmmstates)a), or with the states carrying an unobserved binary improving worsening component which is unobservable (Fig.
+\@ref(fig:hmmstates)b) while the breathing support dimension still being completely observable.
+The former completely observable model is called _simple_ in the following text while the latter is called _complex_.
+This is unlike many uses of HMMs where the observation process is usually described by a general observation matrix.
+The reason is that the breathing support actually used is very likely to correspond to the breathing support needed by the patient.
+In initial versions of our models we also tested models that treated the breathing support observed as noisy realisation of the actual state, but the fitted probabilities of such imprecise observations were always very low, so we ended up not using them.
+
+Directly modelling the full transition matrix of the Markov process would not make best use of the data as we know there is additional structure expressed in the ordering of states, e.g.
+that the probability of being discharged from the "AA" state is higher than from the "Oxygen" state, or that transitioning to "Ventilated" is more likely from the "Oxygen" state than from the "AA" state.
+To incorporate this structure, we follow the approach outlined in [@http://zotero.org/users/5567156/items/F9K2L5Q8].
+
+Here we define a _rate matrix_ $R$ so that for any two states $i \neq j$ $R_{i,j}$ is the rate of transition from $i$ to $j$.
+$R$ is intended to be sparse, i.e.
+$R_{i,j} \neq 0$ only for transitions explicitly intended to be modelled directly.
+Additionally the diagonal elements are set as $R_{i,i} = -\sum_{j \neq i} R_{i,j}$.
+
+```{r hmmstates, fig.cap="The Markov models used a) simple, b) complex.
+AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation.
+In all models the 'Death' and 'Discharged' states are terminal.
+In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
+knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
+```
+
+The evolution of the vector $p(t)$ of the state probabilities in continuous time $t$ is then given by the differential equation
+
+$$
+\frac{dp(t)}{dt} = Rp
+$$
+
+Given $p(0)$, the solution to this equation is:
+
+$$
+p(t) = \exp(tR)p(0)
+$$
+
+where $\exp$ is the matrix exponential.
+We can thus compute a discrete-time transition matrix $S = \exp(R)$ so that $p(t + 1) = Sp(t)$.
+The appeal of this approach is that we can enforce a lot of structure on the transition matrix while allowing positive probability for almost all transitions.
+In the case of the complex model, the full transition matrix would have 42 free parameters (no transitions from the "Death" and "Discharged states) while the rate matrix has only 13 free parameters.
+We could obviously have a sparse transition matrix, but that would make the model overly rigid as transitions that are not modeled directly but actually occur in the data (e.g.
+from "Oxygen" to "Discharged") would have zero probability.
+
+Finally we put a mixed-effects linear predictor on each of the rates we model so that $R_{i,j} = \exp(\mu_{i,j})$.
+We build the model in the Stan language,  using the `brms` package to express the linear predictors.
+
+As an additional structure, we consider _rate groups_ which are: "Improving" (from a breathing level to a better one, including "Discharged"), "Worsening" (from a breathing level to a worse one, not including "Death"), and "Death" (any transition to the "Death" state).
+In addition, the complex model has "To improving" and "To worsening" which correspond to the switches between the improving and worsening variants of each breathing level.
+
+Below we use three types of models: simple with predictors acting on rate groups (i.e.
+assuming that any predictor has the same multiplicative effect on all rates in a group), simple with predictors acting on rates directly (i.e.
+a predictor can have different effect on each rate) and complex with predictors acting on rate groups.
 
@@
 
-```{r}
-#data$patient_data %>% filter(!is.na(best_supportive_care_from)) %>% select(outcome)
-```
-
-```{r}
-recode_group <- function(rate_data, group_name) {
-  is_column <- paste0("is_", group_name) 
-  id_column <- paste0(group_name, "_id")
-
-  rate_data %>% mutate(
-    !!is_column := as.double(rate_group == group_name),
-    !!id_column := factor(
-      if_else(rate_group == group_name, as.integer(.rate_id) - min(as.integer(.rate_id[rate_group == group_name])), as.integer(0)), 
-      levels = 0:(sum(rate_group == group_name) - 1), 
-      ordered = TRUE)
-  )
-}
-
-```
-
-# Treatments - effects on rate groups
+# Treatments - simple, effects on rate groups
 
@@
 
-```{r trt_group_m1}
+```{r}
 trt_group_m1_hidden_state_data <- simple_hidden_state_data
@@
          .to = factor(.to, levels = disease_s_levels)
-  ) %>%
-  recode_group("death") %>%
-  recode_group("improve_one") %>%
-  recode_group("worsen_one") %>%
-  mutate(rate_group = fct_relevel(rate_group, "improve_one"),
   )
@@
 
+```
+
+For the simple model, the rates we model are:
+
+```{r}
+simple_rate_data  
+```
+
+The $\log(R_{i,j})$ elements are then modeled via the `brms` formula
+
+```{r}
+trt_group_m1_formula
+```
+I.e. there is a separate coefficient for each rate and then the other predictors act uniformly on each rate group.
+
+This is the summary of the fitted coefficients.
+
+```{r trt_group_m1}
 trt_group_m1 <- brmshmmdata(trt_group_m1_formula, serie_data, 
@@
 summary(trt_group_m1_fit$brmsfit)
-```
-
-```{r, fig.width=12, fig.height=12}
-#bayesplot::mcmc_pairs(xx$fit, regex_pars = c("b_.rate_id"))
-# bayesplot::mcmc_pairs(trt_group_m1_fit$brmsfit, regex_pars = "worsen_one", np = nuts_params(trt_group_m2_fit$brmsfit))
-
-```
-
-```{r, fig.width=12, fig.height=12}
-# bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, pars = c("b_.rate_idAA_Discharged", "b_.rate_idOxygen_AA"), regex_pars = "improve.*(age|male)")
-# 
-#bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, regex_pars = c("b_.rate_id"))
-```
-
-```{r}
-# fit <- trt_group_m1_fit
-# data_28 <- trt_group_m1
-# data_28$serie_data <- serie_data_28
-# epred_rect <- posterior_epred_rect(fit)#, newdata = data_28)
-# predicted_rect <- posterior_epred_to_predicted(fit, epred_rect)
-# # predicted <- posterior_rect_to_long(fit, predicted_rect)
-# 
-# pp_check_state_at(fit, predicted_rect, 28, newdata = data_28)
-```
-
-```{r}
 
@@
 hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
-```
-
-```{r}
-read_csv("D:/Google Drive/MBÚ/Covid/hmm_res.csv")
 ```
@@
 
+```{r}
+trt_group_m2_formula <- update.formula(trt_group_m1_formula,  ~ . +  (0 + age_norm + is_male || rate_group))
+
+```
+
 ```{r trt_group_m2}
-trt_group_m2_formula <- update.formula(trt_group_m1_formula,  ~ . +  (0 + age_norm + is_male || rate_group))
 trt_group_m2 <- brmshmmdata(trt_group_m2_formula, serie_data, 

### 2020-11-19 manuscript/hmm.Rmd
Fixes for FW HMM models

--- 2020-11-14 manuscript/hmm.Rmd
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 
-hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+serie_data_28_fw = serie_data_28 %>% filter(first_wave)
+
+hmm_hypothesis_res_list[["trt_group_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m4_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities, first_wave", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_group_m4_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m1"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m1_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fw_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fw_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 ```
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m3"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m3_fw"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fw_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m3_fw"]] <- evaluate_all_treatment_hypotheses(c_trt_m3_fw_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
 
@@
 ```{r}
-hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl)
+hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl, serie_data_28 = serie_data_28)
 

### 2020-11-19 manuscript/manuscript.Rmd
Fixes for FW HMM models

@@
     toc: false
-    keep_md: true
     reference_docx: "output_template.docx"
@@
 For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period.
-The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.
+The data collection started at the onset of the Covid-19 pandemic (except for one site where some older records were inaccessible), but the end data for collection differs between sites.
 Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.

### 2020-11-19 R/hmm_tools.R

@@ -34,8 +34,9 @@ compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_dat
   log_odds_numerator - log_odds_denominator
 }
 
-evaluate_treatment_hypothesis <- function(fit, hypothesis, treatment_column, model_subgroup,
-                                          adjusted, model_check, cl = NULL, cores = parallel::detectCores()) {
+evaluate_treatment_hypothesis <- function(fit, serie_data_28, hypothesis, treatment_column, model_subgroup,
+                                          adjusted, model_check, cl = NULL, cores = parallel::detectCores()
+                                          ) {
   counterfactual_took <- serie_data_28 %>% mutate({{ treatment_column }} := 1)
   counterfactual_no <- serie_data_28 %>% mutate({{treatment_column }} := 0)
 
@@ -51,10 +52,10 @@ evaluate_treatment_hypothesis <- function(fit, hypothesis, treatment_column, mod
   )
 }
 
-evaluate_all_treatment_hypotheses <- function(fit, model_subgroup, adjusted, model_check, cl = NULL, cores = parallel::detectCores()) {
+evaluate_all_treatment_hypotheses <- function(fit, model_subgroup, adjusted, model_check, serie_data_28 = serie_data_28, cl = NULL, cores = parallel::detectCores()) {
   rbind(
-    evaluate_treatment_hypothesis(fit, hypotheses$hcq_reduces_death, took_hcq, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check, cl = cl, cores = cores),
-    evaluate_treatment_hypothesis(fit, hypotheses$az_reduces_death, took_az, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check, cl = cl, cores = cores),
-    evaluate_treatment_hypothesis(fit, hypotheses$favipiravir_reduces_death, took_favipiravir, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check, cl = cl, cores = cores)
+    evaluate_treatment_hypothesis(fit, serie_data_28, hypotheses$hcq_reduces_death, took_hcq, model_subgroup = model_subgroup, adjusted = adjusted, model_check = model_check, cl = cl, cores = cores),
[…]

### 2020-11-21 manuscript/hmm.Rmd
Infrastructure for supplement

@@
 ---
-title: "HMM Model"
-output: html_document
+title: "Hidden Markov Models"
+output: 
+  bookdown::pdf_document2:
+    toc: false
+bibliography: "Covid19.json"
 ---
 
-```{r setup}
+# Hidden Markov Models
+
+```{r setup_hmm, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+knitr::opts_chunk$set(echo = FALSE)
+
 devtools::load_all()
@@
 
-# The overall approach for hidden Markov models
+## The overall approach for hidden Markov models
 
@@
 We treat the breathing support required as a Markov process, either with states directly corresponding to the breathing support and directly observable (Fig.
-\@ref(fig:hmmstates)a), or with the states carrying an unobserved binary improving worsening component which is unobservable (Fig.
+\@ref(fig:hmmstates)a), or with the states carrying an unobserved binary improving/worsening component which is unobservable (Fig.
 \@ref(fig:hmmstates)b) while the breathing support dimension still being completely observable.
 The former completely observable model is called _simple_ in the following text while the latter is called _complex_.
-This is unlike many uses of HMMs where the observation process is usually described by a general observation matrix.
+This is unlike many uses of HMMs where the observation process is described by a general observation matrix.
 The reason is that the breathing support actually used is very likely to correspond to the breathing support needed by the patient.
-In initial versions of our models we also tested models that treated the breathing support observed as noisy realisation of the actual state, but the fitted probabilities of such imprecise observations were always very low, so we ended up not using them.
+In initial versions of our models we also tested models that treated the breathing support observed as a noisy realisation of the actual state, but the fitted probabilities of such imprecise observations were always very low, so we ended up not using them.
 
-Directly modelling the full transition matrix of the Markov process would not make best use of the data as we know there is additional structure expressed in the ordering of states, e.g.
+Directly modeling the full transition matrix of the Markov process would not make best use of the data as we know there is additional structure expressed in the ordering of states, e.g.
 that the probability of being discharged from the "AA" state is higher than from the "Oxygen" state, or that transitioning to "Ventilated" is more likely from the "Oxygen" state than from the "AA" state.
@@
 
-Here we define a _rate matrix_ $R$ so that for any two states $i \neq j$ $R_{i,j}$ is the rate of transition from $i$ to $j$.
+We setup a _rate matrix_ $R$ so that for any two states $i \neq j$ $R_{i,j}$ is the rate of transition from $i$ to $j$.
 $R$ is intended to be sparse, i.e.
-$R_{i,j} \neq 0$ only for transitions explicitly intended to be modelled directly.
+$R_{i,j} \neq 0$ only for transitions explicitly intended to be modeled directly.
 Additionally the diagonal elements are set as $R_{i,i} = -\sum_{j \neq i} R_{i,j}$.
@@
 In all models the 'Death' and 'Discharged' states are terminal.
-In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
-knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
+In the complex model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed." }
+knitr::include_graphics(here::here("manuscript","static_figs","model_states_hmm.png"), auto_pdf = TRUE)
 ```
@@
 
-Given $p(0)$, the solution to this equation is:
+Given the initial state probabilities $p(0)$, the solution to this equation is:
 
@@
 We can thus compute a discrete-time transition matrix $S = \exp(R)$ so that $p(t + 1) = Sp(t)$.
-The appeal of this approach is that we can enforce a lot of structure on the transition matrix while allowing positive probability for almost all transitions.
-In the case of the complex model, the full transition matrix would have 42 free parameters (no transitions from the "Death" and "Discharged states) while the rate matrix has only 13 free parameters.
+The appeal of this approach is that we can enforce a lot of structure on the transition matrix $S$ while allowing positive probability for almost all transitions.
+In the case of the complex model, the full transition matrix between the 8 states would have 42 free parameters (no transitions from the "Death" and "Discharged states) while the rate matrix has only 13 free parameters.
 We could obviously have a sparse transition matrix, but that would make the model overly rigid as transitions that are not modeled directly but actually occur in the data (e.g.
@@
 
-Finally we put a mixed-effects linear predictor on each of the rates we model so that $R_{i,j} = \exp(\mu_{i,j})$.
-We build the model in the Stan language,  using the `brms` package to express the linear predictors.
+We also find the rate formulation to be theoretically appealing - the disease progression takes place in continuous time, the discretization into individual days is an artifact of the way we collected data, not the reality.
+
+Finally we put a mixed-effects linear predictor on each of the rates we model so that they are allowed to differ between patients (indexed by $k$) and over (discrete) time (indexed by $t$) $R_{k,t;i,j} = \exp(\mu_{k,t;i,j})$ and $S_{k,t} = \exp(R_{k,t})$.
+We build the model in the Stan language,  using the `brms` package to express the linear predictors $\mu_{k,t;i,j}$ - we use baseline patient characteristics as time-constant predictors and the treatments administered, initiation of best supportive care and markers as time-varying predictors.
 
@@
 a predictor can have different effect on each rate) and complex with predictors acting on rate groups.
+
+Models will be presented in a unified structure that we explain with the first model and won't be repeated
 
@@
 
-# Treatments - simple, effects on rate groups
-
-## Only treatments
+## Treatments - simple, effects on rate groups
 
 ```{r}
-trt_group_m1_hidden_state_data <- simple_hidden_state_data
-
-trt_group_m1_observed_state_data <- base_observed_state_data
-
 simple_rate_data <- rbind(
@@
          .from = factor(.from, levels = disease_s_levels),
-         .to = factor(.to, levels = disease_s_levels)
+         .to = factor(.to, levels = disease_s_levels),
+          rate_group = fct_relevel(rate_group, "improve_one")
        )
@@
 simple_initial_states <- serie_data %>% filter(.time == 1) %>% arrange(.serie) %>% pull(.observed)
+```
+
+The rates included in the simple model are shown in Table&nbsp;\@ref(tab:statessimple).
+
+```{r statessimple}
+simple_rate_data  %>% knitr::kable(booktabs = TRUE, caption="Rates used in the simple model.")
+```
+
+### Only treatments
+
+```{r}
+trt_group_m1_hidden_state_data <- simple_hidden_state_data
+
+trt_group_m1_observed_state_data <- base_observed_state_data
 
@@
 
-For the simple model, the rates we model are:
-
-```{r}
-simple_rate_data  
-```
-
-The $\log(R_{i,j})$ elements are then modeled via the `brms` formula
+For this model, the $\log(R_{i,j})$ elements are then modeled via the `brms` formula
 
@@
 ```
-I.e. there is a separate coefficient for each rate and then the other predictors act uniformly on each rate group.
+Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:statessimple)) and those taken from patient's data.
+There is a separate coefficient for each rate and then the other predictors act uniformly on each rate group.
+The `took_XXX` variables are boolean predictors that set to true for all days since the first dose of the treatment was given - i.e.
+we assume the treatment alters the disease overall disease progression irreversibly.
 
@@
 
-summary(trt_group_m1_fit$brmsfit)
+print_my_fit_summary(trt_group_m1_fit)
 
 ```
+Note that the coefficients are  hard to interpret directly.
+E.g.
+initiated best supportive care (`best_supportive_care`) increases the rate of "Death" but it also might decrease the rate of "worsening" transitions.
+How should those be combined?
+To make the results easy to interpret we use the model to build counterfactual predictions for each patient and treatment: what is (according to the model) the probability of the patient being alive at 28 days if they didn't receive the treatment at all and what if the received the treatment upon admission.
+Those two probabilities can then be used to compute log odds-ratio, which is the value we report.
+Similarly, we compute the log odds-ratio for being discharged.
+For this model this results in the following estimates of log ORs.
 
@@
 hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+hmm_hypothesis_res_list[["trt_group_m1"]] %>% select(hypothesis, point_estimate, ci_low, ci_high) %>%
+  rename(`95% CI Low` = ci_low, `95% CI High` = ci_high)
 ```
@@
 }
+rm(trt_group_m1_fit)
 ```
 
-## Treatments + age + sex
+### Treatments + age + sex
+
+In this model the formula is expanded to include age and sex:
 
@@
 
+trt_group_m2_formula
 ```
+
+This is the summary of the model parameters
 
@@
 
-trt_group_m2_fit <- brmhmm(trt_group_m2, cache_file = paste0(cache_dir,"/trt_group_m2.rds"), init = 0.1, control = default_control, iter = 1200)
+trt_group_m2_fit <- brmhmm(trt_group_m2, cache_file = paste0(cache_dir,"/trt_group_m2.rds"), init = 0.1, control = list(adapt_delta = 0.95), iter = 1200)
 
 #parnames(trt_group_m2_fit$brmsfit)
-summary(trt_group_m2_fit$brmsfit)
-```
-```{r, fig.width=12, fig.height=12}
-# 
-#  
-# bayesplot::mcmc_pairs(trt_group_m2_fit$brmsfit, regex_pars = "improve_one", np = nuts_params(trt_group_m2_fit$brmsfit))
-# 
-# 
-# bayesplot::mcmc_parcoord(trt_group_m2_fit$brmsfit, regex_pars = "death", np = nuts_params(trt_group_m2_fit$brmsfit))
-# bayesplot::mcmc_parcoord(trt_group_m2_fit$brmsfit, regex_pars = "worsen_one", np = nuts_params(trt_group_m2_fit$brmsfit))
+print_my_fit_summary(trt_group_m2_fit)
 ```
@@
 }
+rm(trt_group_m2_fit)
 ```
 
-## Treatments + age + sex + hospital
+### Treatments + age + sex + hospital
 
@@
 
-trt_group_m3_fit <- brmhmm(trt_group_m3, cache_file = paste0(cache_dir,"/trt_group_m3.rds"), init = 0.1, control = default_control, iter = 1200)
+trt_group_m3_fit <- brmhmm(trt_group_m3, cache_file = paste0(cache_dir,"/trt_group_m3.rds"), init = 0.1, control = list(adapt_delta = 0.95), iter = 1200)
 
@@
 }
+rm(trt_group_m3_fit)
 ```
 
-## Treatments + age + sex + hospital, first wave only
+### Treatments + age + sex + hospital, first wave only
 
@@
 }
+rm(trt_group_m3_fw_fit)
 ```
 
-## Treatment + age + sex + hospital + comorbidities
+### Treatment + age + sex + hospital + comorbidities
 
@@
 }
+rm(trt_group_m4_fit)
 ```
 
-## Treatment + age + sex + hospital + comorbidities, first wave
+### Treatment + age + sex + hospital + comorbidities, first wave
 
@@
 }
+rm(trt_group_m4_fit)
 ```
 
-# Treatments - effect on individual rates
+## Treatments - effect on individual rates
 
-## Only treatments
+### Only treatments
 
@@
 }
+rm(trt_rate_m1_fit)
 ```
 
-## Treatments + age + sex
+### Treatments + age + sex
 
@@
 }
+rm(trt_rate_m2_fit)
 ```
 
-## Treatments + age + sex + hospital
+### Treatments + age + sex + hospital
 
@@
 
-trt_rate_m3_fit <- brmhmm(trt_rate_m3, cache_file = paste0(cache_dir,"/trt_rate_m3.rds"), init = 0.1, control = default_control, iter = 1500)
+trt_rate_m3_fit <- brmhmm(trt_rate_m3, cache_file = paste0(cache_dir,"/trt_rate_m3.rds"), init = 0.1, control = list(adapt_delta = 0.95), iter = 1500)
 
@@
 }
+rm(trt_rate_m3_fit)
 ```
 
-## Treatments + age + sex + hospital, first_Wave
+### Treatments + age + sex + hospital, first_Wave
 
@@
 
-trt_rate_m3_fw_fit <- brmhmm(trt_rate_m3_fw, cache_file = paste0(cache_dir,"/trt_rate_m3_fw.rds"), init = 0.1, control = default_control, iter = 1500)
+trt_rate_m3_fw_fit <- brmhmm(trt_rate_m3_fw, cache_file = paste0(cache_dir,"/trt_rate_m3_fw.rds"), init = 0.1, control = list(adapt_delta = 0.95), iter = 1500)
 
@@
 }
+rm(trt_rate_m3_fw_fit)
 ```
 
-## Treatment + age + sex + hospital + comorbidities
+### Treatment + age + sex + hospital + comorbidities
 
@@
 }
+rm(trt_group_m4_fit)
 ```
 
-# Markers
+## Markers
 
-## D-dimer + hospital
+### D-dimer + hospital
 
@@
 }
+rm(d_dimer_m1_fit)
 ```
 
-## D-dimer + age + sex + comorbidities
+### D-dimer + age + sex + comorbidities
 
@@
 
-## IL-6 + hospital
+### IL-6 + hospital
 
@@
 }
+rm(IL_6_m1_fit)
 ```
 
-## IL_6 + age + sex + comorbidities
+### IL_6 + age + sex + comorbidities
 
@@
 
-# Complex: Treatments - effects on rate groups
+## Complex: Treatments - effects on rate groups
 
-## Only treatments
+### Only treatments
 
@@
 }
+rm(c_trt_m1_fit)
 ```
 
-## Treatments + age + sex
+### Treatments + age + sex
 
@@
 }
+rm(c_trt_m2_fit)
 ```
 
-## Treatments + age + sex + hospital
+### Treatments + age + sex + hospital
 
@@
 }
+rm(c_trtr_m3_fit)
 ```
 
-## Treatments + age + sex + hospital, first wave
+### Treatments + age + sex + hospital, first wave
 
@@
 if(do_pp_checks) {
-  do_common_pp_checks(c_trt_m3_fit)
+  do_common_pp_checks(c_trt_m3_fw_fit)
 }
+rm(c_trt_m3_fw_fit)
 ```
 
-## Treatment + age + sex + hospital + comorbidities
+### Treatment + age + sex + hospital + comorbidities
 
@@
 }
+rm(c_trt_m4_fit)
 ```
-
-# Environment
 
@@
 ```
-
-```{r}
-sessionInfo()
-```

### 2020-11-21 manuscript/supplement.Rmd
Infrastructure for supplement

---
site: "bookdown::bookdown_site"
title: "Supplementary material for Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis"
output:
  bookdown::pdf_document2
bibliography: "Covid19.json"
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
#author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
---

```{r setup, echo=FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=FALSE, cache = TRUE, cache.lazy = FALSE)
```

\newpage

```{r child_hmm, child = 'hmm.Rmd'}
```

### 2020-11-21 R/hmm_tools.R

@@ -1,4 +1,4 @@
-compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_data, denominator_serie_data, day = 28,
+compute_counterfactual_state_log_OR <- function(hmm_fit, numerator_serie_data, denominator_serie_data, target_state, day = 28,
                                                     cl = NULL, cores = parallel::detectCores()) {
   if(nrow(numerator_serie_data) != nrow(denominator_serie_data)) {
     stop("Groups to compare must have equal number of rows")
@@ -13,7 +13,7 @@ compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_dat
 
   combined_data$initial_states = rep(hmm_fit$data$initial_states, 2)
 
-  mortality_state <- which(hmm_fit$data$hidden_state_data$id == "Death")
+  target_state_id <- which(hmm_fit$data$hidden_state_data$id == target_state)
   numerator_indices <- unique(as.integer(combined_data$serie_data$.serie)[combined_data$serie_data$is_numerator])
   denominator_indices <- unique(as.integer(combined_data$serie_data$.serie)[!combined_data$serie_data$is_numerator])
   if(!identical(sort(c(numerator_indices, denominator_indices)), 1:length(unique(combined_data$serie_data$.serie)))) {
@@ -24,9 +24,9 @@ compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_dat
   state_probs <- posterior_epred_rect(hmm_fit, newdata = combined_data, method = posterior_epred_state_prob, cl = cl, cores = cores)
 
   samples_probs_numerator <-
-    rowMeans(state_probs[mortality_state, day + 1, , numerator_indices])
+    rowMeans(state_probs[target_state_id, day + 1, , numerator_indices])
   samples_probs_denominator <-
-    rowMeans(state_probs[mortality_state, day + 1, , denominator_indices])
+    rowMeans(state_probs[target_state_id, day + 1, , denominator_indices])
 
   log_odds_numerator <- log(samples_probs_numerator) - log1p(-samples_probs_numerator)
   log_odds_denominator <- log(samples_probs_denominator) - log1p(-samples_probs_denominator)
@@ -34,28 +34,70 @@ compute_counterfactual_mortality_log_OR <- function(hmm_fit, numerator_serie_dat
   log_odds_numerator - log_odds_denominator
 }
 
[…]

### 2020-11-21 R/predict_hmm.R

@@ -84,16 +84,18 @@ posterior_epred_state_prob <- function(nsamples, max_times, initial_states, tran
   }
 
   res_internal <- posterior_epred_state_prob_internal(n_samples = nsamples,
-                                      max_times = max_times,
-                                      initial_states = initial_states,
-                                      n_states = n_states,
-                                      n_time = n_time,
-                                      n_predictor_sets = n_predictor_sets,
-                                      transition_matrices = transition_matrices,
-                                      predictor_sets_rect = predictor_sets_rect)
+                                    max_times = max_times,
+                                    initial_states = initial_states,
+                                    n_states = n_states,
+                                    n_time = n_time,
+                                    n_predictor_sets = n_predictor_sets,
+                                    transition_matrices = transition_matrices,
+                                    predictor_sets_rect = predictor_sets_rect)
 
   dim(res_internal) <- c(n_states, n_time, nsamples, n_series);
+
   res_internal
+
 }
 
 posterior_rect_to_long <- function(fit, posterior_rect, newdata = NULL) {
@@ -219,55 +221,77 @@ posterior_long_to_df <- function(data, prediction) {
 
 #'
 #' @return An array of dimensions `N_states_hidden` x `N_states_hidden` x `N_predictor_sets` x `N_samples`
-compute_all_transition_matrices <- function(data_hmm, epred_mu, cl = NULL, cores = parallel::detectCores()) {
-  start_time = proc.time()
-  nsamples <- dim(epred_mu)[1]
-  epred_mu_exp <- exp(epred_mu)
-  standata <- data_hmm$standata
+compute_all_transition_matrices <- function(data_hmm, epred_mu, cl = NULL, cores = parallel::detectCores(), cache = "auto") {
+  res <- NULL
+  if(cache == "auto") {
+    all_params <- list(data_hmm, epred_mu)
 
-  compute_transition_matrices_ps <- function(ps, epred_mu_exp, standata) {
-    res <- array(NA_real_, c(standata$N_states_hidden, standata$N_states_hidden, nsamples))
-    for(i in 1:nsamples) {
[…]

### 2020-11-22 manuscript/hmm.Rmd
More work on supplement

@@
 ```
-Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:statessimple)) and those taken from patient's data.
+
+Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:statessimple)) and those taken from patient's data (both the fixed baseline characteristics and time-varying values).
 There is a separate coefficient for each rate and then the other predictors act uniformly on each rate group.
-The `took_XXX` variables are boolean predictors that set to true for all days since the first dose of the treatment was given - i.e.
-we assume the treatment alters the disease overall disease progression irreversibly.
+The `took_XXX` variables are boolean predictors that are set to true for all days since the first dose of the treatment was given - i.e.
+we assume the treatment alters the overall disease progression irreversibly.
 
-This is the summary of the fitted coefficients.
+This is the summary of the fitted coefficients. `Q2.5` and `Q97.5` are the boundaries of the central 95% credible interval.
 
@@
 
-print_my_fit_summary(trt_group_m1_fit)
+print_hmm_fit_summary(trt_group_m1_fit)
 
 ```
+
 Note that the coefficients are  hard to interpret directly.
@@
 How should those be combined?
-To make the results easy to interpret we use the model to build counterfactual predictions for each patient and treatment: what is (according to the model) the probability of the patient being alive at 28 days if they didn't receive the treatment at all and what if the received the treatment upon admission.
+To make the results easy to interpret we use the model to build counterfactual predictions for each patient and treatment: what is (according to the model) the probability of the patient being alive at 28 days if they did not receive the treatment at all and what if they received the treatment immediately upon admission.
 Those two probabilities can then be used to compute log odds-ratio, which is the value we report.
@@
 
-hmm_hypothesis_res_list[["trt_group_m1"]] %>% select(hypothesis, point_estimate, ci_low, ci_high) %>%
-  rename(`95% CI Low` = ci_low, `95% CI High` = ci_high)
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m1"]])
 ```
@@
 #parnames(trt_group_m2_fit$brmsfit)
-print_my_fit_summary(trt_group_m2_fit)
+print_hmm_fit_summary(trt_group_m2_fit)
 ```
+
+And here are the estimates of the log ORs:
 
@@
 hmm_hypothesis_res_list[["trt_group_m2"]] <- evaluate_all_treatment_hypotheses(trt_group_m2_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m2"]])
 ```
@@
 
+Here the model is further expanded to allow all the rates to vary between hospitals.
+
+```{r}
+trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id || hospital_id))
+trt_group_m3_formula
+```
+
+Summary of the coefficients:
+
 ```{r trt_group_m3}
-trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
 
@@
 
-summary(trt_group_m3_fit$brmsfit)
+print_hmm_fit_summary(trt_group_m3_fit)
 ```
+We see that the data are consistent with quite high between-site differences (measured in standard deviations of the varying intercepts).
+
+And the resulting log ORs are:
 
@@
 hmm_hypothesis_res_list[["trt_group_m3"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m3"]])
 ```
@@
 
+This is the same model as above, but using only the patients from the first wave. Here is the summary of the coefficients.
+
 ```{r trt_group_m3_fw}
@@
 
-summary(trt_group_m3_fw_fit$brmsfit)
+print_hmm_fit_summary(trt_group_m3_fw_fit)
 ```
+
+And the resulting log ORs:
 
@@
 hmm_hypothesis_res_list[["trt_group_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m3_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m3_fw"]])
 ```
@@
 
+For this model, the sum of comorbidities is added as a continuous predictor, resulting in the formula:
+
+```{r}
+trt_group_m4_formula <-  update.formula(trt_group_m3_formula,  ~ . + (comorbidities_sum || rate_group)) 
+trt_group_m4_formula
+```
+
+Below is the summary of the coefficients:
+
 ```{r trt_group_m4}
-trt_group_m4_formula <-  update.formula(trt_group_m3_formula,  ~ . + (comorbidities_sum || rate_group))
 
@@
 
-trt_group_m4_fit <- brmhmm(trt_group_m4, cache_file = paste0(cache_dir,"/trt_group_m4.rds"), init = 0.1, control = default_control, iter = 1200)
+trt_group_m4_fit <- brmhmm(trt_group_m4, cache_file = paste0(cache_dir,"/trt_group_m4.rds"), init = 0.1, control = list(adapt_delta = 0.95), iter = 1200)
 
-summary(trt_group_m4_fit$brmsfit)
+print_hmm_fit_summary(trt_group_m4_fit)
 ```
+The sampler warns us of divergent transitions
+
+And the resulting log ORs:
 
@@
 hmm_hypothesis_res_list[["trt_group_m4"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m4"]])
 ```
@@
 
-summary(trt_group_m4_fw_fit$brmsfit)
+print_hmm_fit_summary(trt_group_m4_fw_fit)
 ```
@@
 rm(trt_group_m4_fit)
+invisible(parallel::clusterEvalQ(cl, gc()))
 ```
@@
 
-summary(trt_rate_m1_fit$brmsfit)
+print_hmm_fit_summary(trt_rate_m1_fit)
 ```
@@
 
-summary(trt_rate_m2_fit$brmsfit)
+print_hmm_fit_summary(trt_rate_m2_fit)
 ```
@@
 
-summary(trt_rate_m3_fit$brmsfit)
+print_hmm_fit_summary(trt_rate_m3_fit)
 ```
@@
 
-summary(trt_rate_m3_fw_fit$brmsfit)
+print_hmm_fit_summary(trt_rate_m3_fw_fit)
 ```
@@
 
-summary(trt_rate_m4_fit$brmsfit)
+print_hmm_fit_summary(trt_rate_m4_fit)
 ```
@@
 }
-rm(trt_group_m4_fit)
+rm(trt_rate_m4_fit)
+invisible(parallel::clusterEvalQ(cl, gc()))
+
 ```
@@
 
-summary(d_dimer_m1_fit$brmsfit)
+print_hmm_fit_summary(d_dimer_m1_fit)
 ```
@@
 
-summary(d_dimer_m2_fit$brmsfit)
+print_hmm_fit_summary(d_dimer_m2_fit)
 
@@
   )
+rm(d_dimer_m2_fit)
 
@@
 
-summary(IL_6_m1_fit$brmsfit)
+print_hmm_fit_summary(IL_6_m1_fit)
 ```
@@
 
-summary(IL_6_m2_fit$brmsfit)
+print_hmm_fit_summary(IL_6_m2_fit)
 ```
@@
   )
+
+```
+
+```{r}
+rm(IL_6_m2_fit)
+invisible(parallel::clusterEvalQ(cl, gc()))
 
@@
 
-summary(c_trt_m1_fit$brmsfit)
+print_hmm_fit_summary(c_trt_m1_fit)
 
@@
 #parnames(c_trt_m2_fit$brmsfit)
-summary(c_trt_m2_fit$brmsfit)
+print_hmm_fit_summary(c_trt_m2_fit)
 ```
@@
 
-summary(c_trt_m3_fit$brmsfit)
+print_hmm_fit_summary(c_trt_m3_fit)
 ```
@@
 
-summary(c_trt_m3_fw_fit$brmsfit)
+print_hmm_fit_summary(c_trt_m3_fw_fit)
 ```
@@
 
-summary(c_trt_m4_fit$brmsfit)
+print_hmm_fit_summary(c_trt_m4_fit)
 ```
@@
 rm(c_trt_m4_fit)
+invisible(parallel::clusterEvalQ(cl, gc()))
 ```

### 2020-11-22 manuscript/hmm.Rmd
Jm and risks for supplement

--- 2020-11-21 manuscript/hmm.Rmd
@@
 }
-# warning("Subsetting data")
-# set.seed(4568224)
-# patients_to_use <- sample(data$patient_data$patient_id, size = 70)
-# data <- filter_complete_data(data, patient_id %in% patients_to_use)
 

### 2020-11-22 manuscript/jm.Rmd
Jm and risks for supplement

@@
 title: "Joint modelling"
-output: html_notebook
+output: html_document
 ---
+
+# Joint longitudinal and time-to-event models
 
@@
 ```{r}
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[1:11])
-plot(p1, plot_observed = TRUE)                     
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[12:14])
-plot(p1, plot_observed = TRUE)                     
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[15:18])
-plot(p1, plot_observed = TRUE)
-
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[19:20])
-plot(p1, plot_observed = TRUE)                     
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[21:22])
-plot(p1, plot_observed = TRUE)                     
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[22:26])
-plot(p1, plot_observed = TRUE)
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[27:29])
-plot(p1, plot_observed = TRUE)                     
-p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[30:32])
-plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[1:11])
+# plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[12:14])
+# plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[15:18])
+# plot(p1, plot_observed = TRUE)                     
+# 
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[19:20])
+# plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[21:22])
+# plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[22:26])
+# plot(p1, plot_observed = TRUE)
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[27:29])
+# plot(p1, plot_observed = TRUE)                     
+# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[30:32])
+# plot(p1, plot_observed = TRUE)                     
 ```

### 2020-11-26 dev/raw_data_preprocessing.Rmd
Working with the new dataset
Cleaning up, finishing touches on manuscript, work on supplement

@@
 ```
-
-TODO: check last day
 
@@
 ```
+Data collection periods
+
+```{r}
+data_collection_period <- complete_data$patient_data %>% group_by(hospital_id) %>%
+  summarise(first_admission = min(admission), last_admission = max(admission), .groups = "drop") %>%
+  arrange(first_admission) %>%
+  mutate(hospital_id_anon = 1:n()) %>% 
+  select(-hospital_id)
+
+write_csv(data_collection_period, here::here("manuscript", "data_collection_period.csv"))
+```
+
+```{r}
+complete_data$patient_data <- complete_data$patient_data  %>% mutate(first_wave = admission < lubridate::ymd("2020-09-01"))
+```
 

### 2020-11-26 manuscript/analysis_brms.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

@@
 
-# {.tabset}
-
-## Setup
+# Bayesian models with brms
 
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
+    draws = -draws$b_ever_hcqTRUE,
     model = "brms_binary",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
     model = "brms_binary",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
     model = "brms_binary",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$favipiravir_reduces_death,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_death,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
+    draws = -draws$b_ever_hcqTRUE,
     model = "brms_binary_7",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
     model = "brms_binary_7",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
     model = "brms_binary_7",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$favipiravir_reduces_death,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_death,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_hcqTRUE,
+    model = "brms_binary_28",
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
+    model = "brms_binary_28",
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
+    model = "brms_binary_28",
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  )
+)
+```
+
+## Analysis: Survival time
+
+```{r, results="hide"}
+
+prior_survival_time <- 
+  prior(normal(0, 5), class = "b") +
+  prior(normal(0, 5), class = "sd")
+
+```
+
+### Fewer adjustments
+
+```{r}
+form_survival_time_small <- bf(
+  day | cens(censored) ~ sex + age_norm  + ever_hcq + ever_az + ever_favipiravir +
+    (1 | hospital_id),
+  family = brmsfamily("cox", "log")
+)
+
+fit_survival_time_small <- brm(
+  formula = form_survival_time_small,
+  data = data$surv_data_last_day,
+  prior = prior_survival_time,
+  inits = 0,
+  control = default_control,
+  file = paste0(cache_dir, "/fit_survival_time_small")
+)
+fit_survival_time_small <- add_criterion(fit_survival_time_small, "loo")
+
+```
+
+```{r}
+summary(fit_survival_time_small)
+loo(fit_survival_time_small)
+```
+
+```{r}
+draws <- tidy_draws(fit_survival_time_small)
+hypothesis_res_list[["hcq_az_survival_time_small"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
     draws = draws$b_ever_hcqTRUE,
-    model = "brms_binary_28",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm",
+    model_check = "some influential observations"
   ),
@@
     draws = draws$b_ever_azTRUE,
-    model = "brms_binary_28",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm",
+    model_check = "some influential observations"
   ),
@@
     draws = draws$b_ever_favipiravirTRUE,
-    model = "brms_binary_28",
-    estimand = "logit(survival)",
-    hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  )
-)
-```
-
-## Analysis: Survival time
-
-```{r, results="hide"}
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm",
+    model_check = "some influential observations"
+  )
+)
+```
+
+### Fewer adjustments - First wave only
+
+```{r}
+
+fit_survival_time_small_fw <- brm(
+  formula = form_survival_time_small,
+  data = data$surv_data_last_day %>% filter(first_wave),
+  prior = prior_survival_time,
+  inits = 0,
+  control = default_control,
+  file = paste0(cache_dir, "/fit_survival_time_small_fw")
+)
+fit_survival_time_small_fw <- add_criterion(fit_survival_time_small_fw, "loo")
+
+```
+
+```{r}
+summary(fit_survival_time_small_fw)
+loo(fit_survival_time_small_fw)
+```
+
+```{r}
+draws <- tidy_draws(fit_survival_time_small_fw)
+hypothesis_res_list[["hcq_az_survival_time_small_fw"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, first_wave",
+    model_check = "some influential observations"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, first_wave",
+    model_check = "some influential observations"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, first_wave",
+    model_check = "some influential observations"
+  )
+)
+```
+
+### More adjustments
+
+```{r}
 form_survival_time <- bf(
@@
 
-prior_survival_time <- 
-  prior(normal(0, 5), class = "b") +
-  prior(normal(0, 5), class = "sd")
-
 fit_survival_time <- brm(
@@
 fit_survival_time <- add_criterion(fit_survival_time, "loo")
+
 ```
@@
 ```
+There were some very influential observations.
 
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_reduces_death,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$az_reduces_death,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
   ),
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$favipiravir_reduces_death,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  )
+)
+```
+
+### More adjustments - First wave only
+
+```{r, results="hide"}
+fit_survival_time_fw <- brm(
+  formula = form_survival_time,
+  data = data$surv_data_last_day %>% filter(first_wave),
+  prior = prior_survival_time,
+  inits = 0,
+  control = default_control,
+  file = paste0(cache_dir, "/fit_survival_time_fw")
+)
+fit_survival_time_fw <- add_criterion(fit_survival_time_fw, "loo")
+```
+
+```{r}
+summary(fit_survival_time_fw)
+loo(fit_survival_time_fw)
+```
+
+There were some very influential observations.
+
+```{r}
+draws <- tidy_draws(fit_survival_time_fw)
+hypothesis_res_list[["hcq_az_survival_time_fw"]] <- rbind(
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_hcqTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_azTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_ever_favipiravirTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "Suspicious"
   )
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
+    draws = -draws$b_ever_hcqTRUE,
     model = "brms_binary",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
     model = "brms_binary",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
     model = "brms_binary",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_hospital,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
+    draws = -draws$b_ever_hcqTRUE,
     model = "brms_binary_7",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
     model = "brms_binary_7",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
     model = "brms_binary_7",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_hospital,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
+    draws = -draws$b_ever_hcqTRUE,
     model = "brms_binary_28",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
     model = "brms_binary_28",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
     model = "brms_binary_28",
-    estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
+    estimand = "log(OR)",
+    hypothesis = hypotheses$favipiravir_hospital,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
 
+There were some very influential observations
+
 ```{r}
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_hcqTRUE,
-    model = "brms_cox",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_azTRUE,
-    model = "brms_cox",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_ever_favipiravirTRUE,
-    model = "brms_cox",
-    estimand = "log(HR)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
-    model_check = "some influencial observations"
+    draws = -draws$b_ever_hcqTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influential observations"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_azTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "some influential observations"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_ever_favipiravirTRUE,
+    model = "brms_cox",
+    estimand = "log(HR)",
+    hypothesis = hypotheses$favipiravir_hospital,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "some influential observations"
   )
@@
     estimand = "logit(death)",
-    hypothesis = hypotheses$hcq_reduces_death,
+    hypothesis = hypotheses$hcq_death,
     adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
@@
     estimand = "logit(death)",
-    hypothesis = hypotheses$az_reduces_death,
+    hypothesis = hypotheses$az_death,
     adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
@@
     estimand = "logit(death)",
-    hypothesis = hypotheses$favipiravir_reduces_death,
+    hypothesis = hypotheses$favipiravir_death,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_hcqTRUE,
+    draws = -draws$b_muDischarged_ever_hcqTRUE,
     model = "brms_categorical",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_azTRUE,
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_azTRUE,
     model = "brms_categorical",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "Suspicious"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "Suspicious"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_favipiravirTRUE,
     model = "brms_categorical",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
+    hypothesis = hypotheses$favipiravir_hospital,
     adjusted = "hcq, az, sex, age_norm, smoking, obesity",
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_hcqTRUE,
+    draws = -draws$b_muDischarged_ever_hcqTRUE,
     model = "brms_categorical_7",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "probably overconfident"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_azTRUE,
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_azTRUE,
     model = "brms_categorical_7",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "probably overconfident"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_favipiravirTRUE,
     model = "brms_categorical_7",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    hypothesis = hypotheses$favipiravir_hospital,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  )
+)
+```
+
+### Basic Categorical model (7 Days), first wave only
+
+```{r, results="hide"}
+fit_outcome_base_7_fw <- brm(
+  formula = form_outcome_base,
+  data = data$surv_data_7 %>% filter(first_wave),
+  prior = prior_outcome_base,
+  control = default_control,
+  file = paste0(cache_dir, "/fit_outcome_base_7_fw")
+)
+fit_outcome_base_7_fw <- add_criterion(fit_outcome_base_7_fw, "loo")
+```
+
+```{r}
+summary(fit_outcome_base_7_fw)
+pp_check(fit_outcome_base_7_fw)
+loo(fit_outcome_base_7_fw)
+# ce <- conditional_effects(fit_outcome_base_7_fw)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_outcome_base_7_fw)
+hypothesis_res_list[["hcq_az_outcome_base_7_fw"]] <- rbind(
+  # Death 
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_hcqTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_azTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_favipiravirTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  # Discharged
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_hcqTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_azTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_favipiravirTRUE,
+    model = "brms_categorical_7",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_hospital,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity, first_wave",
     model_check = "probably overconfident"
@@
   bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_hcqTRUE,
+    draws = -draws$b_muDischarged_ever_hcqTRUE,
     model = "brms_categorical_28",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$hcq_increases_discharged,
-    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "probably overconfident"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_azTRUE,
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_azTRUE,
     model = "brms_categorical_28",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$az_increases_discharged,
-    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
-    model_check = "probably overconfident"
-  ),
-  bayesian_hypothesis_res_from_draws(
-    draws = draws$b_muDischarged_ever_favipiravirTRUE,
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_favipiravirTRUE,
     model = "brms_categorical_28",
     estimand = "logit(discharged)",
-    hypothesis = hypotheses$favipiravir_increases_discharged,
-    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    hypothesis = hypotheses$favipiravir_hospital,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity",
+    model_check = "probably overconfident"
+  )
+)
+```
+
+### Basic Categorical model (28 Days), first wave only
+
+```{r, results="hide"}
+fit_outcome_base_28_fw <- brm(
+  formula = form_outcome_base,
+  data = data$surv_data_28,
+  prior = prior_outcome_base,
+  control = default_control,
+  file = paste0(cache_dir, "/fit_outcome_base_28_fw")
+)
+fit_outcome_bas_28 <- add_criterion(fit_outcome_base_28_fw, "loo")
+```
+
+```{r}
+summary(fit_outcome_base_28_fw)
+pp_check(fit_outcome_base_28_fw)
+loo(fit_outcome_base_28_fw)
+# ce <- conditional_effects(fit_outcome_base_28_fw)
+# plot(ce, ask = FALSE)
+```
+
+```{r}
+draws <- tidy_draws(fit_outcome_base_28_fw)
+hypothesis_res_list[["hcq_az_outcome_base_28_fw"]] <- rbind(
+  # Death
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_hcqTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$hcq_death,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_azTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$az_death,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = draws$b_muDeath_ever_favipiravirTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(death)",
+    hypothesis = hypotheses$favipiravir_death,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  # Discharged
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_hcqTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$hcq_hospital,
+    adjusted = "az, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_azTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$az_hospital,
+    adjusted = "hcq, favipiravir, sex, age_norm, smoking, obesity, first_wave",
+    model_check = "probably overconfident"
+  ),
+  bayesian_hypothesis_res_from_draws(
+    draws = -draws$b_muDischarged_ever_favipiravirTRUE,
+    model = "brms_categorical_28",
+    estimand = "logit(discharged)",
+    hypothesis = hypotheses$favipiravir_hospital,
+    adjusted = "hcq, az, sex, age_norm, smoking, obesity, first_wave",
     model_check = "probably overconfident"
@@
 
-### Model Comparison
-
-```{r}
-# loo(fit_worst_breathing_base, fit_worst_breathing_cs)
-```
-
-## Store Hypotheses
-
-```{r}
+```{r store_hypothesis_brms}
 hypothesis_res_all <- do.call(rbind, hypothesis_res_list)

### 2020-11-26 manuscript/competing_risks.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq,
                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex")
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex", multiplier = -1)
   )
@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az,
                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az", multiplier = -1),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az,
                                         coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az,
-                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq")
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq", multiplier = -1)
   )
@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_reduces_death, fit_competing_hcq_az_fpv,
+  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_death, fit_competing_hcq_az_fpv,
                                         coefficient_name = "took_favipiravirTRUE", transition = "1:3", adjusted = "age, sex, hcq, az", model_check = "Problematic"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_increases_discharged, fit_competing_hcq_az_fpv,
-                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az", model_check = "Problematic")
+  frequentist_hypothesis_res_from_coxph(hypotheses$favipiravir_hospital, fit_competing_hcq_az_fpv,
+                                        coefficient_name = "took_favipiravirTRUE", transition = "1:2", adjusted = "age, sex, hcq, az", model_check = "Problematic", multiplier = -1)
   )
@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_admitted,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_admitted,
                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, admitted, az"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_admitted,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, admitted, az"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az_admitted,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, admitted, az", multiplier = -1),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_admitted,
                                         coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, admitted, hcq"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az_admitted,
-                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, admitted, hcq")
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_admitted,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, admitted, hcq", multiplier = -1)
   )
@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_hospital,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_hospital,
                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, hospital"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_hospital,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_reduces_death, fit_competing_hcq_az_hospital,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_hospital,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, hospital", multiplier = -1),
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_death, fit_competing_hcq_az_hospital,
                                         coefficient_name = "took_azTRUE", transition = "1:3", adjusted = "age, sex, hcq, hospital"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$az_increases_discharged, fit_competing_hcq_az_hospital,
-                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq, hospital")
+  frequentist_hypothesis_res_from_coxph(hypotheses$az_hospital, fit_competing_hcq_az_hospital,
+                                        coefficient_name = "took_azTRUE", transition = "1:2", adjusted = "age, sex, hcq, hospital", multiplier = -1)
   )
@@
   rbind(
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_reduces_death, fit_competing_hcq_az_comorbidities_sum,
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_death, fit_competing_hcq_az_comorbidities_sum,
                                         coefficient_name = "took_hcqTRUE", transition = "1:3", adjusted = "age, sex, az, comorbidities (sum)"),
-  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_increases_discharged, fit_competing_hcq_az_comorbidities_sum,
-                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)")
+  frequentist_hypothesis_res_from_coxph(hypotheses$hcq_hospital, fit_competing_hcq_az_comorbidities_sum,
+                                        coefficient_name = "took_hcqTRUE", transition = "1:2", adjusted = "age, sex, az, comorbidities (sum)", multiplier = -1)
   )
@@
 hypothesis_res_list[["competing_il_6"]] <-
-  frequentist_hypothesis_res_from_coxph(hypotheses$IL_6_increases_death, fit_competing_il_6,
+  frequentist_hypothesis_res_from_coxph(hypotheses$IL_6_death, fit_competing_il_6,
                                         coefficient_name = "log_peak_IL_6", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
@@
 hypothesis_res_list[["competing_d_dimer"]] <-
-  frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_increases_death, fit_competing_d_dimer,
+  frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_death, fit_competing_d_dimer,
                                         coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Problematic")

### 2020-11-26 manuscript/hmm.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

@@
 
-The rates included in the simple model are shown in Table&nbsp;\@ref(tab:statessimple).
+The rates included in the simple model are shown in Table&nbsp;\@ref(tab:ratessimple).
 
-```{r statessimple}
+```{r ratessimple}
 simple_rate_data  %>% knitr::kable(booktabs = TRUE, caption="Rates used in the simple model.")
@@
 Those two probabilities can then be used to compute log odds-ratio, which is the value we report.
-Similarly, we compute the log odds-ratio for being discharged.
-For this model this results in the following estimates of log ORs.
+Similarly, we compute the log odds-ratio for being still hospitalized against being discharged.
+In both cases log odds ratio > 0 is worse for the patient (higher chance of death, higher chance of being hospitalized) in the treatment group.
+
+Additionally we also look on the (counterfactual) time of hospitalization for patients discharged by day 28 in both groups and - to put it on the same scale, compare the mean of the log of the per-patient ratios of those times, i.e.
+estimate > 0 implies worse outcome (longer hospitalization) in the treatment group.
+This value is not reported in the main manuscript, only in the supplement.
+
+For this model this results in the following estimates:
 
 ```{r}
+
+#
+
 hmm_hypothesis_res_list[["trt_group_m1"]] <- evaluate_all_treatment_hypotheses(trt_group_m1_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
@@
 ```{r}
-trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id || hospital_id))
+trt_group_m3_formula <- update.formula(trt_group_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
 trt_group_m3_formula
@@
 ```
-The sampler warns us of divergent transitions
+The sampler warns us of divergent transitions, indicating the posterior might not have been fully explored.
+This is most likely because we do not have enough data to estimate all the parameters.
+Here and in the following, we mark models that had such problems as "Suspicious" and do not report their results in the main text of the manuscript.
 
-And the resulting log ORs:
+The resulting log ORs (which cannot be fully trusted):
 
@@
 
+Once again, we run the same model but for the first wave data only. Similarly to the previous case this results in divergent transitions and the model should not be fully trusted.
+
 ```{r trt_group_m4_fw}
@@
 ```
+And here are the resulting log ORs
 
 ```{r}
-hmm_hypothesis_res_list[["trt_group_m4_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
+hmm_hypothesis_res_list[["trt_group_m4_fw"]] <-  evaluate_all_treatment_hypotheses(trt_group_m4_fw_fit, model_subgroup = "simple_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities, first_wave", model_check = "Suspicious", cl = cl, serie_data_28 = serie_data_28_fw)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_group_m4_fw"]])
 ```
@@
 if(do_pp_checks) {
-  do_common_pp_checks(trt_group_m4_fit)
+  do_common_pp_checks(trt_group_m4_fw_fit)
 }
-rm(trt_group_m4_fit)
+rm(trt_group_m4_fw_fit)
 invisible(parallel::clusterEvalQ(cl, gc()))
@@
 
+In this set of models the effects (especially the treatments) act directly on the individual rates.
+
 ### Only treatments
 
-```{r trt_rate_m1}
+For treatments only, this boils down to the following `brms` formula:
+
+```{r}
 trt_rate_m1_hidden_state_data <- simple_hidden_state_data
@@
 
+trt_rate_m1_formula
+```
+
+Resulting in the following fitted coefficients:
+
+```{r trt_rate_m1}
 trt_rate_m1 <- brmshmmdata(trt_rate_m1_formula, serie_data, 
@@
 ```
+And those log ORs for the outcomes:
 
@@
 hmm_hypothesis_res_list[["trt_rate_m2"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m2_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_rate_m2"]])
 ```
@@
 
+We allow the rates to differ by hospital site with this formula:
+
+```{r}
+trt_rate_m3_formula <- update.formula(trt_rate_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
+trt_rate_m3_formula
+```
+
+Those are the resulting fitted coefficients:
+
 ```{r trt_rate_m3}
-trt_rate_m3_formula <- update.formula(trt_rate_m2_formula,  ~  . + (0 + .rate_id | hospital_id))
 trt_rate_m3 <- brmshmmdata(trt_rate_m3_formula, serie_data, 
@@
 
+And the resulting estimates:
+
 ```{r}
 hmm_hypothesis_res_list[["trt_rate_m3"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_rate_m3"]])
 ```
@@
 
+This is same as the previous model, but limiting only to first wave patients, the fitted coefficients:
+
 ```{r trt_rate_m3_fw}
@@
 
+And the log(OR) estimates:
+
 ```{r}
 hmm_hypothesis_res_list[["trt_rate_m3_fw"]] <-  evaluate_all_treatment_hypotheses(trt_rate_m3_fw_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, first_wave", model_check = "OK", cl = cl, serie_data_28 = serie_data_28_fw)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_rate_m3_fw"]])
 ```
@@
 
+Adding comorbidities, the complete formula:
+
+```{r}
+trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula, ~ . + (comorbidities_sum || .rate_id))
+trt_rate_m4_formula
+```
+
+Coefficient estimates - note the divergent transitions, indicating the results are not completely trustworthy:
+
 ```{r trt_rate_m4}
-trt_rate_m4_formula <-  update.formula(trt_rate_m3_formula, ~ . + (comorbidities_sum || .rate_id))
 
@@
 
+Log odds-ratio estimates:
+
 ```{r}
 hmm_hypothesis_res_list[["trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(trt_rate_m4_fit, model_subgroup = "simple_rate", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["trt_rate_m4"]])
 ```
@@
 
+In the marker models, we use the simple model and investigate effects on rate groups.
+We filter only for patients that at had the marker measured at least once.
+We use the peak of the marker so far as a predictor in the model and investigate its estimated coefficient.
+Unfortunately, none of the results are trustworthy as the sampler encountered divergent transitions.
+
 ### D-dimer + hospital
 
-```{r d_dimer_m1}
+The formula is:
+
+```{r}
 d_dimer_serie_data <- serie_data %>% 
@@
 d_dimer_m1_formula <-  ~ 0 + .rate_id +  (0 + best_supportive_care || rate_group)  + (0 + .rate_id | hospital_id) + (log_peak_d_dimer || rate_group)
+
+d_dimer_m1_formula
+```
+
+And here are the coefficient estimates  - note the divergent transitions, indicating the results are not completely trustworthy:
+
+```{r d_dimer_m1}
 
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$d_dimer_increases_death,
+    hypothesis = hypotheses$d_dimer_death,
     adjusted = "supportive, hospital",
@@
 
+Including comorbidites in the model gives the formula:
+
 ```{r}
 d_dimer_m2_formula <-  update.formula(d_dimer_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
+d_dimer_m2_formula
+```
+
+And here are the coefficient estimates - note the divergent transitions, indicating the results are not completely trustworthy:
+
+```{r}
 
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$d_dimer_increases_death,
+    hypothesis = hypotheses$d_dimer_death,
     adjusted = "supportive, age, sex, hospital",
@@
 ### IL-6 + hospital
+
+We further look into Interleukin 6 using the formula:
 
@@
 IL_6_m1_formula <- ~ 0 + .rate_id +  (0 + best_supportive_care || rate_group)  + (0 + .rate_id | hospital_id) + (log_peak_IL_6 || rate_group) 
+IL_6_m1_formula
+```
+
+Giving us the following coefficients - note the divergent transitions, indicating the results are not completely trustworthy:
+
+```{r}
 
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$IL_6_increases_death,
+    hypothesis = hypotheses$IL_6_death,
     adjusted = "supportive, hospital"
@@
 
+And finally adding comorbidities with the formula:
+
 ```{r}
 IL_6_m2_formula <-  update.formula(IL_6_m1_formula, ~ . + (0 + age_norm + is_male || rate_group))
+IL_6_m2_formula
+```
+
+Gives us those estimates - note the divergent transitions, indicating the results are not completely trustworthy:
+
+```{r}
 
@@
     estimand = "log(HR)",
-    hypothesis = hypotheses$IL_6_increases_death,
+    hypothesis = hypotheses$IL_6_death,
     adjusted = "supportive, age, sex, hospital"
@@
 
-### Only treatments
-
-```{r c_trt_m1}
+```{r}
 complex_rate_data <- rbind(
@@
 
+```
+
+```{r ratescomplex}
+complex_rate_data %>% knitr::kable(booktabs = TRUE, caption="Rates used in the complex model.") 
+```
+
+Here we use the complex model and let treatments have effect on the rate groups.
+The models have the same formulae as the simple models, only the rate (and state) definitions change.
+The rates used are shown in Table&nbsp;\@ref(tab:ratescomplex).
+
+### Only treatments
+
+Using treatments only with the formula:
+
+```{r}
+trt_group_m1_formula
+```
+
+Fitted coefficients:
+
+```{r c_trt_m1}
+
 c_trt_m1 <- brmshmmdata(trt_group_m1_formula, serie_data, 
@@
 
+Estimates:
+
 ```{r}
 hmm_hypothesis_res_list[["c_trt_rate_m1"]] <- evaluate_all_treatment_hypotheses(c_trt_m1_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["c_trt_rate_m1"]])
 ```
@@
 ### Treatments + age + sex
+
+Adding age and sex with formula:
+
+```{r}
+trt_group_m2_formula
+```
+
+Fitted coefficients:
 
@@
 
+Estimates:
+
 ```{r}
 hmm_hypothesis_res_list[["c_trt_rate_m2"]] <- evaluate_all_treatment_hypotheses(c_trt_m2_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex", model_check = "OK", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["c_trt_rate_m2"]])
 
@@
 
+Including hospital as well with the formula:
+
+```{r}
+trt_group_m3_formula
+```
+
+Fitted coefficients:
+
 ```{r c_trt_m3}
@@
 
+Estimates:
+
 ```{r}
@@
 
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["c_trt_rate_m3"]])
 ```
@@
 }
-rm(c_trtr_m3_fit)
+rm(c_trt_m3_fit)
 ```
@@
 ### Treatments + age + sex + hospital, first wave
+
+The same model as above, but using only first wave patients. Fitted coefficients:
 
@@
 
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["c_trt_rate_m3_fw"]])
 ```
@@
 
+Adding comorbidities with the formula:
+
+```{r}
+trt_group_m4_formula
+```
+
+Fitted coefficients - note the divergent transitions, indicating the results are not completely trustworthy:
+
 ```{r c_trt_m4}
@@
 
+Estimates (not completely trustworthy due to the divergent transitions):
+
 ```{r}
 hmm_hypothesis_res_list[["c_trt_rate_m4"]] <- evaluate_all_treatment_hypotheses(c_trt_m4_fit, model_subgroup = "complex_group", adjusted ="all treatments, supportive, age, sex, hospital, comorbidities", model_check = "Suspicious", cl = cl, serie_data_28 = serie_data_28)
+
+print_hmm_hypothesis_res(hmm_hypothesis_res_list[["c_trt_rate_m4"]])
 

### 2020-11-26 manuscript/jm.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

@@
   bayesian_hypothesis_res_from_jm(
-    hypotheses$d_dimer_increases_death,
+    hypotheses$d_dimer_death,
     fit_d_dimer_age_sex,
@@
   bayesian_hypothesis_res_from_jm(
-    hypotheses$d_dimer_increases_death,
+    hypotheses$d_dimer_death,
     fit_d_dimer_age_sex_hcq,
@@
   bayesian_hypothesis_res_from_jm(
-    hypotheses$IL_6_increases_death,
+    hypotheses$IL_6_death,
     fit_IL_6_age_sex,
@@
   bayesian_hypothesis_res_from_jm(
-    hypotheses$IL_6_increases_death,
+    hypotheses$IL_6_death,
     fit_IL_6_age_sex_hcq,

### 2020-11-26 manuscript/manuscript.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

--- 2020-11-05 manuscript/manuscript.Rmd
@@
     toc: false
-    keep_md: true
+    number_sections: false
     reference_docx: "output_template.docx"
-abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2.
+abstract: |  
+  **Abstract:** We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
-Patients in the dataset were treated with Hydroxychloroquine (N = 108), Azithromycin (N = 72), Favipiravir (N = 9), convalescent plasma (N = 7), Dexamethasone (N = 4) and Remdesivir (N = 3), often in combination.
+Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination.
+Most patients were admitted during the first wave of the epidemic.
 To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
-Weak evidence to constrain the potential efficacy of Azithromycin and Favipiravir can be extracted from the data.
-Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset."
+Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data.
+Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
 bibliography: Covid19.json
+csl: journal-of-medical-virology.csl
 date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
@@
 
-The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020 led to over 46 million cases and over 1.1 million deaths.
-Proposed treatments include antivirals approved for other indications (Chloroquine, Hydroxychloroquine, Lopinavir/Ritonavir, Remdesivir, Favipiravir, Umifenovir), Azithromycin, corticosteroids, immunoglobulins, Tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
-Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid Dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], while results on Remdesivir have been conflicting [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/NSTX4EZM].
-Despite initial promise, Hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/NSTX4EZM], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
-Further, no improvement could be discerned when adding Azithromycin to Hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
+The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths.
+Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
+Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
+Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA.
+However, during spring and summer it was available only in limited amounts via compassionate use programme.
+Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3].
+Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
+Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
 Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
@@
 
-Interestingly, in observational studies, Hydroxychloroquine was often found to be associated with better outcomes [e.
+Interestingly, in observational studies, hydroxychloroquine was often found to be associated with better outcomes [e.
 g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].
 
-Favipiravir is one of the less explored experimental treatments, with only one larger study known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
+Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
 Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B].
-None of the results allow strong conclusions about benefits or harms of Favipiravir.
+None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
+
  For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].
@@
 
-# Methods
+# Patients and Methods
 
@@
 n_nippv <- data$marker_data_wide %>% filter(breathing == "NIPPV") %>% pull(patient_id) %>% unique() %>% length()
+
+n_second_wave <- data$patient_data %>% filter(!first_wave) %>% nrow()
 ```
@@
 For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period.
-The data collection started at the onset of the Covid-19 pandemic, but the end data for collection differs between sites.
-Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical evaluation of any Covid-19 pharmacotherapy.
+The data collection started at the onset of the Covid-19 pandemic (except for one site where some older records were inaccessible), but the end date for collection differs between sites due to time constraints of the participating physicians.
+Three sites included total of `r n_second_wave` patients that could be considered part of "second wave", i.e.
+admitted after September 1st, last patient included in the dataset was admitted on October 12th.
+See Supplementary Figure 1 for per-site data collection periods.
+Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical trial of any Covid-19 pharmacotherapy.
 
@@
 Full protocol for data collection is attached in the supplementary material.
-Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we collapsed those categories with mechanical ventilation.
+Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we merged those categories with mechanical ventilation.
 
@@
 All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections.
-Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of attending physician and underlying patient condition.
+Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition.
 We note that the participating sites were not homogeneous in either patient population or treatment protocols.
@@
 but were only observed retrospectively.
-This can be partially remedied by adjusting for patient characteristic in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.
+This can be partially remedied by adjusting for patient characteristics in the analysis, but such adjustments will always be imperfect and the analysis  needs to be treated as exploratory and interpreted cautiously.
 
@@
 
-First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
+First class of models are frequentist survival and multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
 We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).
 
-Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
-The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
+Second class of models are Bayesian hidden Markov models (HMM) of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
+The parametrization of the HMM is inspired by Williams et al.[@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
 The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths.
@@
 
-Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g. the effect of taking a drug is only included for the days after the drug was taken.
+Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g., the effect of taking a drug is only included for the days after the drug was taken.
+More details on the exact model formulations can be found in the supplementary statistical analysis.
 
@@
 
-We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e.
+We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for those that could be applied to our dataset (i.e.
 where we have gathered all the input features).
@@
 When there were multiple reasonable ways to evaluate the outcome or a predictor in our dataset, we computed and reported all of those options.
-We used two simple scores with age or the decade of age as the sole predictor to have a simple baseline to compare the scores against.
+We used two simple scores with age or the decade of age as the sole predictor to have a baseline to compare the scores against.
 
@@
     `Admitted for Covid` = sum_percent(admitted_for_covid),
-    `Took Azithromycin` = sum_percent(ever_az),
-    `Took Dexamethasone` = sum_percent(ever_dexamethasone),
-    `Took Favipiravir` = sum_percent(ever_favipiravir),
-    `Took Remdesivir` = sum_percent(ever_remdesivir),
+    `Took azithromycin` = sum_percent(ever_az),
+    `Took dexamethasone` = sum_percent(ever_dexamethasone),
+    `Took favipiravir` = sum_percent(ever_favipiravir),
+    `Took remdesivir` = sum_percent(ever_remdesivir),
     `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
@@
   knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups.
-Note that the Favipiravir subgroup is not exclusive with either the HCQ or No HCQ group.
+Note that the favipiravir subgroup is not exclusive with either the HCQ or No HCQ group.
 IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)
@@
 In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
-The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients were hospitalized with high CRP and for many the CRP levels stayed elevated over the whole hospitalization.
+The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization.
 The limited nature of the data however does not allow for any statistically robust conclusions.
@@
 Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.
-However we believe this kind of patient-level view is useful to appreciate both the extent of between- and within-patient variability.
+However we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.
 
@@
 As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
-To illustrate this phenomenon, we give an example of the association of treatment with Hydroxychloroquine on patients' baseline characteristics.
-
-```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with Hydroxychloroquine seemed to be associated with the status of patients upon hospitalization.
-Comorbidities were associated with both worse outcome (black) and lower chance of treatment with Hydroxychloroquine (red).
-HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
-
-tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
-  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
-  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
-tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
-tmp$outcome<-dropLevels(tmp$outcome)
-tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
-tmp2$outcome<-dropLevels(tmp2$outcome)
-tmp2$outcome<-tmp2$outcome=='Death'
-
-m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
-m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
-m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
-m.copd<-glm(outcome~COPD,tmp2,family='binomial')
-m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
-m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
-m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
-m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
-m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
-m.s<-glm(outcome~smoking,tmp2,family='binomial')
-m.ob<-glm(outcome~obesity,tmp2,family='binomial')
-m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
-m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
-m.la<-glm(outcome~low_albumin,tmp2,family='binomial')
-
-m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
-m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
-m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
-m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
-m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
-m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
-m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
-m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
-m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
-m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
-m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
-m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
-m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
-m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')
-
-models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
-modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
-ys<-c(0)
-for (m in models) {
-  ys<-c(min(min(ys),confint(m)[2,1]),max(max(ys),confint(m)[2,2]))
-}
-models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
-modelNames2<-modelNames
-ys2<-c(0)
-for (m in models2) {
-  ys2<-c(min(min(ys2),confint(m)[2,1]),max(max(ys2),confint(m)[2,2]))
-}
-
-plot(ys,c(1,length(modelNames)),ty='n',yaxt='n',xlab='Log odds ratio',ylab='')
-abline(v=0,col='blue')
-col.hcq<-bfc()["Vermillion"]
-for (i in seq(along=models)) {
-  ii<-length(models)-i+1
-  m<-models[[i]]
-  m2<-models2[[i]]
-  lines(confint(m)[2,],rep(ii,2))
-  points(coef(m)[2],ii,pch=19)
-  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
-  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
-}
-axis(2,length(modelNames):1,modelNames,las=2)
-legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
-```
-
-Figure&nbsp;\@ref(fig:hcqassoc) shows that patients with worse baseline characteristics, which lead to worse outcome, 
-were less likely to receive Hydroxychloroquine treatment. 
-This clearly creates a bias towards a positive effect of Hydroxychloroquine on the outcome, 
-which, however, could be false.
+In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 3).
+This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.
 
@@
 
-Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future Hydroxychloroquine treatment:
+Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future hydroxychloroquine treatment:
 the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
-other lung diseases, renal disease, or high creatinin 
-was associated with a lower chance of taking Hydroxychloroquine over the course of the hospitalization 
+other lung diseases, renal disease, or high creatinine 
+was associated with a lower chance of taking hydroxychloroquine over the course of the hospitalization 
 (the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
@@
 ```{r}
-bayesian_cols <- cols(
-  hypothesis = col_character(),
-  model = col_character(),
-  adjusted = col_character(),
-  estimand = col_character(),
-  widest_CI_excl_reference = col_double(),
-  point_estimate = col_double(),
-  ci_low = col_double(),
-  ci_high = col_double()
-)
-res_bayesian <- 
-  rbind(read_csv(here::here("manuscript", "hmm_res.csv"), col_types = bayesian_cols),
-        read_csv(here::here("manuscript", "analysis_brms.csv"), col_types = bayesian_cols),
-        read_csv(here::here("manuscript", "jm_res.csv"), col_types = bayesian_cols)
-        )
 
-frequentist_cols <- cols(
-  hypothesis = col_character(),
-  model = col_character(),
-  adjusted = col_character(),
-  p_value = col_double(),
-  ci_low = col_double(),
-  ci_high = col_double()
-)
+all_hypo_drugs <- read_all_multiverse_res()  %>%
+  filter(model_check == "OK", !adj.first_wave, estimand_group == "")
 
-res_frequentist <- 
-  rbind(read_csv(here::here("manuscript", "competing_risks_res.csv"), col_types = frequentist_cols),
-        read_csv(here::here("manuscript", "single_outcome_coxph_res.csv"), col_types = frequentist_cols))
-
-all_hypo <- rbind(res_bayesian %>% select(-widest_CI_excl_reference), 
-      res_frequentist %>% select(-p_value)) %>%
-    inner_join(hypotheses_df, by = c("hypothesis" = "name"))
-
-all_hypo_drugs <- all_hypo  %>%
-  filter(model_check == "OK")
+first_fig_multiverse_drugs <- 4
+last_fig_multiverse_drugs <- 6
+supplementary_figure_markers <- last_fig_multiverse_drugs + 1 
+supplementary_figures_multiverse <-  paste0(first_fig_multiverse_drugs, "-", last_fig_multiverse_drugs)
 
@@
 
-Here, we focus on Hydroxychloroquine and Azithromycin as those are the only treatment with larger sample size.
-We also investigate Favipiravir as it is less well reported in the literature.
+Here, we focus on hydroxychloroquine and azithromycin as those are the only treatments with larger sample size.
+We also investigate favipiravir as it is less well reported in the literature.
 Hydroxychloroquine was dosed almost exclusively in a 5-day regime starting with a loading dose of 800mg on the first day and followed by 400mg.
-Majority of patients complemented Hydroxychloroquine with Azithromycin while Azithromycin was rarely used alone (see Table \@ref(tab:summarytable)).
+Majority of patients complemented hydroxychloroquine with azithromycin while azithromycin was rarely used alone (see Table \@ref(tab:summarytable)).
 Azithromycin was most frequently dosed 250 or 500mg/day, but doses ranging from 100mg/day to 1500 mg/day were observed.
 Favipiravir was used only at one site with a loading dose of 3600mg on the first day, followed by at most 9 days with a 1600mg dose.
-All but one of the patients receiving Favipiravir also received Hydroxychloroquine.
-Treatment was initiated mostly within two days of admission (see Figure&nbsp;\@ref(fig:initiation)).
+All but one of the patients receiving favipiravir also received hydroxychloroquine.
+Treatment was initiated mostly within two days of admission (see Supplementary Figure 2).
 
-```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites.
-Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
-#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
-#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
-#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
+The results of the multiverse analysis for association between hydroxochloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues.
+Results for all models we tested are reported in  supplementary Figures `r supplementary_figures_multiverse`, with additional details in supplementary statistical analysis.
+The results do not change noticeably when only patients from the first wave are included (supplementary Figures `r supplementary_figures_multiverse`).
 
-#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
-
-marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
-
-data$marker_data %>% 
-  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
-  mutate(marker = marker_map[marker]) %>%
-  group_by(marker, patient_id) %>% 
-  summarise(first_day = min(day), .groups = "drop") %>% 
-  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
-```
-
-The results of the multiverse analysis for association between Hydroxochloroquine and Azithromycin usage and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that we did not find to have immediate problems representing the data well or computational issues.
-Results for all models we tested are reported in the supplementary material.
-Notably, most models report that using Hydroxychloroquine is associated with lower risk of death, we must however bear in mind the potential bias noted in the previous section.
+Most models report that using hydroxychloroquine is associated with lower risk of death.
+We must however bear in mind the potential bias noted in the previous section.
 Also, we see that for the HMM models, as we add adjustments the credible intervals do not widen but instead shift towards zero.
 This is a weak indication that further adjustments could drive the effect towards zero.
-We did not attempt to model additional adjustments as the models became computationally intractable.
-The case of Hydroxychloroquine serves as a "control group" for our other results - since randomized trials give us high confidence that Hydroxychloroquine does not substantially reduce mortality, we can be quite certain the associations we observe for Hydroxychloroquine are just a measure of bias in the data.
-The models we used cannot discern the sign of association between Hydroxychloroquine and length of hospital stay.
-Additionally, our models either cannot determine the sign of association between Azithromycin and the outcome or even show an increase in risk of death or length of hospitalization.
-This serves as a weak evidence that a substantial improvement from Azithromycin is unlikely.
+We did not attempt to model additional adjustments as the models became computationally unstable.
+The case of hydroxychloroquine serves as a "control group" for our other results - since randomized trials give us high confidence that hydroxychloroquine does not substantially reduce mortality, we can be quite certain the associations we observe for hydroxychloroquine are just a measure of bias in the data.
+Additionally, our models either cannot determine the sign of association between azithromycin and  risk of death or even show an increase in risk of death.
+This serves as a weak evidence that a substantial improvement in mortality from azithromycin is unlikely.
+
+Most models exclude very strong association between increased risk of death and using favipiravir, but our data are necessarily quite limited, which is reflected in the very wide uncertainties around estimates.
+We also cannot put strict bounds on the association between favipiravir and length of hospitalization.
+
+We also examined the association between treatments and length of hospital stay for patients that survived.
+Almost all models cannot discern the sign of the association for all treatments examined (Supplementary Figures `r supplementary_figures_multiverse`).
+Similarly, we studied the association between D-dimer and Interleukin 6 and outcomes, with unconclusive results as well (Supplementary Figure `r supplementary_figure_markers` )
 
 ```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients for association between treatments and main outcomes.
-Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)c.
+Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a), HMM A = Bayesian hidden-Markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Figure \\@ref(fig:modelstates)c.
 For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
 The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
+In all cases coefficient < 0 means better patient outcome in the treatment group.
 Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
 Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
-= total number of comorbidities, AZ = took Azithromycin, HCQ = took Hydroxychloroquine, FPV = took Favipiravir, C.
+= total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C.
 plasma = received convalescent plasma.
@@
 
-#In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
-Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged).
-After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome.
+#In the current dataset, we see a possibly strong negative association between using hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength.
+Across all models, the data offer no definitive conclusions on the association between taking hydroxychloroquine and length of hospital stay (represented as risk for being discharged).
+After adjusting for hydroxychloroquine usage, we also cannot make any strong claims on the association of using azithromycin with either outcome.
 In all cases, the results seem to not be strongly affected by model choice.
@@
 my_lims <- c(-5.5,3.5)
-multiverse_hcq1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "hcq_reduces_death"), x_limits = my_lims)
-multiverse_hcq2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "hcq_increases_discharged"), x_limits = my_lims)
-multiverse_az1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "az_reduces_death"), x_limits = my_lims)
-multiverse_az2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "az_increases_discharged"), x_limits = my_lims)
+multiverse_hcq1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "hcq_death"), x_limits = my_lims, adjustments_to_hide = "first_wave")
+multiverse_az1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "az_death"), x_limits = my_lims, adjustments_to_hide = "first_wave")
 
-# multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_reduces_death"))
-# multiverse_fpv2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_increases_discharged"))
+# multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_death"))
+# multiverse_fpv2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_hospital"))
 
@@
 
-((multiverse_hcq1 & hide_axis_theme) / (multiverse_hcq2 & hide_axis_theme) / (multiverse_az1 & hide_axis_theme) / (multiverse_az2))  + plot_layout(heights = c(1, 0.65,0.8,0.4)) & theme(axis.title.y = element_blank())
+fpv_lims <- c(-12,2)
+
+multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_death"), x_limits = fpv_lims, adjustments_to_hide = "FPV")
+
+((multiverse_hcq1 & hide_axis_theme) / (multiverse_az1 & theme(axis.title.x = element_blank())) / multiverse_fpv1)  + plot_layout(heights = c(0.8, 0.7,0.70)) & theme(axis.title.y = element_blank())
 
@@
 
-Figure \@ref(fig:multiverseresultsfpv) shows a similar analysis for Favipiravir.
-Most models exclude very strong association between death/hospitalization and using Favipiravir, but our data are necessarily quite limited, which is reflected in the very wide uncertainties around estimates.
-The only exception is the Categorical model  at 28 days which allows for substantially higher risk of being hospitalized in the Favipiravir group.
-
-```{r multiverseresultsfpv, fig.height=3.3, fig.width=8.5, fig.cap="Estimates of model coefficients for association between Favipiravir and main outcomes.
-Abbreviations and other elements as in Figure&nbsp;\\@ref(fig:multiverseresultsdrugs)."}
-
-fpv_lims <- c(-12,2)
-
-multiverse_fpv1 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_reduces_death"), x_limits = fpv_lims, adjustments_to_hide = "FPV")
-multiverse_fpv2 <- plot_multiverse(all_hypo_drugs %>% filter(hypothesis == "favipiravir_increases_discharged"), x_limits = fpv_lims, adjustments_to_hide = "FPV")
-
-((multiverse_fpv1 & hide_axis_theme) / (multiverse_fpv2)) + plot_layout(heights = c(0.8,0.3)) & theme(axis.title.y = element_blank())
-
-```
-
-```{r}
-# ## Association between markers and outcomes
-# 
-# We additionally investigated the association of several markers with disease progression.
-Due to low number of patients in the current dataset, the results are inconclusive and dependent on model choice, many models also proved to be problematic to fit to this reduced dataset.
-All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)
-# 
-# 
-# ``{r multiverseresultsmarkers, fig.height=3.5, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
-Each row represents a model - Cox (competing) =  frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a, HMM A = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, JM = Bayesian joint longitudinal and time-to-event model.
-For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
-Additionally the factors the model adjusted for are listed - Site = the study site, Supportive = best supportive care initiated, HCQ = took Hydroxychloroquine.
-We show posterior mean and 95% credible interval."}
-# plot_multiverse(all_hypo %>% filter(group == "markers", model_check != "Problematic"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"))
-# ``
-
-```
 ## Prognostic models are not better than just using age
 
-Following [@http://zotero.org/users/5567156/items/TS57HJ3T] we found five prediction models we were able to recompute: Li et al.
+Following Wynants et al.[@http://zotero.org/users/5567156/items/TS57HJ3T] we found five prediction models we were able to recompute: Li et al.
 report the ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] combining CRP and age to form 3 grades, Chen & Liu [@http://zotero.org/users/5567156/items/4DDUVEKS] report a continuous score using age, CRP, D-dimer and lymphocyte count,
@@
 The outcomes present in the studies were: 12-day mortality, 30-day mortality and mortality without any further details, here we report results for both 12-day and 30-day mortality.
-Full details on the scores and how we used our dataset to compute them is given in the supplementary material.
+Full details on the scores and how we used our dataset to compute them is given in the supplementary statistical analysis.
 
-Overall all prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
+All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
 Additionally, some publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed.
-We thus concur with [@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.
+We thus concur with Wynants et al.
+[@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.
 
 ```{r riskscores, fig.cap="Performance of tested prediction scores as measured by AUC.
-AUC = 1 means perfect prediction while AUC <= 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the score (marked by thick blue line).
+AUC = 1 means perfect prediction while AUC $\\leq$ 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the score (marked by thick blue line).
 The line ranges represent the bootstrapped 95% confidence intervals.
@@
 
-We provide very weak observational evidence against a substantial beneficiary effect of using Azithromycin (both with or without Hydroxychloroquine) and against substantial negative effect of using Favipiravir in hospitalized Covid-19 patients.
+We provide very weak observational evidence against a substantial beneficiary effect of using azithromycin (both with or without hydroxychloroquine) and against substantial negative effect of using favipiravir in hospitalized Covid-19 patients.
+We also observed better outcomes associated with taking hydroxychloroquine, which is likely linked to substantial confounding by indication.
 Where our results contradict randomized trials, the most likely explanation is systematic bias in our dataset.
@@
 We believe that using multiverse analysis is an appropriate way to explore data in such contexts as it lets us be transparent about this sensitivity.
-We further believe that using Hidden Markov models is a promising complement to the standard Cox proportional hazards analysis when detailed information on disease progression is available, particularly because it lets us impose additional structure on the model and thus make inferences with more disease states than would be possible to handle in the Cox framework, making better use of the available data.
+We further believe that using hidden Markov models is a promising complement to the standard Cox proportional hazards analysis when detailed information on disease progression is available, particularly because it lets us impose additional structure on the model and thus make inferences with more disease states than would be possible to handle in the Cox framework, making better use of the available data.
 
@@
 
+# Conflict of Interest
+
+The authors declare that there is no conflict of interest.
+
 # Acknowledgements

### 2020-11-26 manuscript/single_outcome_coxph.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

@@
   rbind(
-  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_reduces_death, 
+  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_death, 
                                          point_estimate = coef(m1)[1],
-                                         adjusted = "age, sex, comorbidities (sum)", 
+                                         adjusted = "age, sex, hospital, comorbidities (sum)", 
                                          test_stat = anova(m1,m0,test='Chisq')[2,2],
@@
                                          ci_high = confint(m1)[1,2]),
-  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_increases_discharged, 
-                                         point_estimate = coef(mt1)[1],
-                                         adjusted = "age, sex, comorbidities (sum)", 
+  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_hospital, 
+                                         point_estimate = -coef(mt1)[1],
+                                         adjusted = "age, sex, hospital, comorbidities (sum)", 
                                          test_stat = anova(mt1,mt0,test='Chisq')[2,2],
@@
                                          p = anova(mt1,mt0,test='Chisq')[2,4],
-                                         ci_low = confint(mt1)[1,1],
-                                         ci_high = confint(mt1)[1,2])
+                                         ci_low = -confint(mt1)[1,2],
+                                         ci_high = -confint(mt1)[1,1])
   )

### 2020-11-26 manuscript/supplementary_figures.Rmd
Cleaning up, finishing touches on manuscript, work on supplement

---
title: "Supplementary figures for Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  # bookdown::pdf_document2:
  #    toc: false
  bookdown::word_document2: 
    toc: false
    reference_docx: "output_template.docx"
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8, cache = TRUE)

devtools::load_all()
library(tidyverse)
library(patchwork)
library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")

theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide

```

```{r datacollectionperiod, fig.height=2.5, fig.cap= "Data collection periods at individual sites, showing the range of admission dates of patients included in the study. Note that we cannot provide additional information to link the sites here with data shown elsewhere as that would  increase the risk of deanonymization of the patients."}
data_collection_period <- read_csv(here::here("manuscript", "data_collection_period.csv"), col_types = cols(
  first_admission = col_date(format = "%Y-%m-%d"),
  last_admission = col_date(format = "%Y-%m-%d"),
  hospital_id_anon = col_integer()
))

data_collection_period %>% ggplot(aes(xmin = first_admission, xmax = last_admission, y = hospital_id_anon)) + geom_linerange(size = 3) + scale_y_continuous("Study site")+
  scale_x_date("Data collection period", date_breaks = "2 weeks")  + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  )
```

```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites. Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())

#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.

marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")

data$marker_data %>% 
  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
  mutate(marker = marker_map[marker]) %>%
  group_by(marker, patient_id) %>% 
  summarise(first_day = min(day), .groups = "drop") %>% 
  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
```

```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with hydroxychloroquine seemed to be associated with the status of patients upon hospitalization. Comorbidities were associated with both worse outcome (black) and lower chance of treatment with hydroxychloroquine (red). Dots and lines represent the estimates and the 95% confidence intervals of the log odds ratio of the respective outcome. HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}

tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
tmp$outcome<-dropLevels(tmp$outcome)
tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
tmp2$outcome<-dropLevels(tmp2$outcome)
tmp2$outcome<-tmp2$outcome=='Death'

m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
m.copd<-glm(outcome~COPD,tmp2,family='binomial')
m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
m.s<-glm(outcome~smoking,tmp2,family='binomial')
m.ob<-glm(outcome~obesity,tmp2,family='binomial')
m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
m.la<-glm(outcome~low_albumin,tmp2,family='binomial')

m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')

models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
ys<-c(0)
for (m in models) {
  ys<-c(min(min(ys),confint(m)[2,1]),max(max(ys),confint(m)[2,2]))
}
models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
modelNames2<-modelNames
ys2<-c(0)
for (m in models2) {
  ys2<-c(min(min(ys2),confint(m)[2,1]),max(max(ys2),confint(m)[2,2]))
}

plot(ys,c(1,length(modelNames)),ty='n',yaxt='n',xlab='Log odds ratio',ylab='')
abline(v=0,col='blue')
col.hcq<-bfc()["Vermillion"]
for (i in seq(along=models)) {
  ii<-length(models)-i+1
  m<-models[[i]]
  m2<-models2[[i]]
  lines(confint(m)[2,],rep(ii,2))
  points(coef(m)[2],ii,pch=19)
  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
}
axis(2,length(modelNames):1,modelNames,las=2)
legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
```

```{r allmulthcq, fig.height=11, fig.cap = "Estimates of model coefficients for association between hydroxychloroquine and main outcomes. The \"Suspicious\" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.  Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure 1a), HMM A = Bayesian hidden-Markov model as in Figure 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Figure 1c. For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval. The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration). In all cases coefficient < 0 means better patient outcome in the treatment group. Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green). Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb. = total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C. plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included. "}
plot_multiverse(read_all_multiverse_res() %>% filter(grepl("hcq", hypothesis)), show_check = TRUE)
```

```{r allmultaz, fig.height=11, fig.cap = "Estimates of model coefficients for association between azithromycin and main outcomes. The \"Suspicious\" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.  Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure 1a), HMM A = Bayesian hidden-Markov model as in Figure 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Figure 1c. For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval. The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration). In all cases coefficient < 0 means better patient outcome in the treatment group. Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green). Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb. = total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C. plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included. "}
plot_multiverse(read_all_multiverse_res() %>% filter(grepl("az", hypothesis)), show_check = TRUE)
```
```{r allmultfpv, fig.height=11, fig.cap = "Estimates of model coefficients for association between favipiravir and main outcomes. The \"Suspicious\" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.  Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure 1a), HMM A = Bayesian hidden-Markov model as in Figure 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Figure 1c. For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval. The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration). In all cases coefficient < 0 means better patient outcome in the treatment group. Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green). Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb. = total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C. plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included. "}
plot_multiverse(read_all_multiverse_res() %>% filter(grepl("favipiravir", hypothesis), model_check != "Problematic"), show_check = TRUE)
```

```{r multiverseresultsmarkers, fig.height=3.5, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death. The \"Suspicious\" section shows models that were found to not fit the data well or have computational issues, \"Problematic\" section shows models that were completely broken - see supplementary statistical analysis for details. Each row represents a model - Cox (competing) =  frequentist Cox model using competing risks (as in Figure 1a), HMM A = Bayesian hidden-markov model as in Figure 1b with predictors for rate groups, JM = Bayesian joint longitudinal and time-to-event model. For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval. Additionally the factors the model adjusted for are listed - Site = the study site, Supportive = best supportive care initiated, HCQ = took Hydroxychloroquine. We show posterior mean and 95% credible interval."}
plot_multiverse(read_all_multiverse_res() %>% filter(group == "markers"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"), show_check = TRUE)

```

### 2020-11-26 R/hypotheses.R

@@ -1,15 +1,15 @@
 hypotheses <-
   list(
-       hcq_reduces_death = list(caption = "HCQ associated with risk of death", group = "hcq", type = "death"),
-       hcq_increases_discharged = list(caption = "HCQ associated with time in hospital", group = "hcq", type = "hospital"),
-       az_reduces_death = list(caption = "Azithromycin associated with risk of death", group = "az", type = "death"),
-       az_increases_discharged = list(caption = "Azithromycin associated with time in hospital", group = "az", type = "hospital"),
-       favipiravir_reduces_death = list(caption = "Favipiravir associated with risk of death", group = "favipiravir", type = "death"),
-       favipiravir_increases_discharged = list(caption = "Favipiravir associated with time in hospital", group = "favipiravir", type = "hospital"),
-       convalescent_plasma_reduces_death = list(caption = "Conv. plasma associated with risk of death", group = "convalescent_plasma", type = "death"),
-       convalescent_plasma_increases_discharged = list(caption = "Conv. plasma associated with time in hospital", group = "convalescent_plasma", type = "hospital"),
-       d_dimer_increases_death = list(caption = "High D-dimer associated with risk of death", group = "markers", type = "death"),
-       IL_6_increases_death = list(caption = "High IL-6 associated with risk of death", group = "markers", type = "death")
+       hcq_death = list(caption = "HCQ associated with risk of death", group = "hcq", type = "death"),
+       hcq_hospital = list(caption = "HCQ associated with time in hospital", group = "hcq", type = "hospital"),
+       az_death = list(caption = "Azithromycin associated with risk of death", group = "az", type = "death"),
+       az_hospital = list(caption = "Azithromycin associated with time in hospital", group = "az", type = "hospital"),
+       favipiravir_death = list(caption = "Favipiravir associated with risk of death", group = "favipiravir", type = "death"),
+       favipiravir_hospital = list(caption = "Favipiravir associated with time in hospital", group = "favipiravir", type = "hospital"),
+       convalescent_plasma_death = list(caption = "Conv. plasma associated with risk of death", group = "convalescent_plasma", type = "death"),
+       convalescent_plasma_hospital = list(caption = "Conv. plasma associated with time in hospital", group = "convalescent_plasma", type = "hospital"),
[…]

### 2020-11-26 R/multiverse_tools.R

@@ -1,63 +1,152 @@
-plot_multiverse <- function(hypo_res, adjustments_to_hide = c(), x_limits = NULL) {
-  all_adjustments_dict <- c(
-                       "hospital" = "Site",
-                       "admitted" = "Admitted",
-                       "supportive" = "Supportive",
-                       "age" = "Age",
-                       "age_norm" = "Age",
-                       "sex" = "Sex",
-                       "smoking" = "Smoking",
-                       "obesity" = "Obesity",
-                       "comorbidities \\(sum\\)" = "Comorb.",
-                       "comorbidities" = "Comorb.",
-                       "hcq" = "HCQ",
-                       "az" = "AZ",
-                       "favipiravir" = "FPV",
-                       "all treatments" = "All treatments",
-                       "convalescent_plasma" = "C. plasma")
-  all_adjustments = names(all_adjustments_dict)
+#Metadat about models and hypothese
 
-  model_labels <- c("brms_categorical" = "Categorical (all)",
-                    "brms_categorical_7" = "Categorical 7",
-                    "brms_categorical_28" = "Categorical 28",
-                    "brms_cox" = "Bayes Cox",
-                    "coxph1" = "Cox (single)",
-                    "coxph" = "Cox (competing)",
-                    "HMMsimple_group" = "HMM A",
-                    "HMMsimple_rate" = "HMM B",
-                    "HMMcomplex_group" = "HMM C",
-                    "jm" = "JM"
+all_adjustments_dict <- c(
+  "hospital" = "Site",
+  "admitted" = "Admitted",
+  "supportive" = "Supportive",
+  "age" = "Age",
+  "age_norm" = "Age",
+  "sex" = "Sex",
+  "smoking" = "Smoking",
+  "obesity" = "Obesity",
+  "comorbidities \\(sum\\)" = "Comorb.",
+  "comorbidities" = "Comorb.",
+  "hcq" = "HCQ",
+  "az" = "AZ",
+  "favipiravir" = "FPV",
+  "all treatments" = "All treatments",
+  "convalescent_plasma" = "C. plasma",
+  "first_wave" = "First wave")
+all_adjustments = names(all_adjustments_dict)
+
+model_labels <- c("brms_categorical" = "Categorical (all)",
+                  "brms_categorical_7" = "Categorical 7",
[…]

### 2020-11-26 manuscript/authors.R

@@ -1,5 +1,5 @@
 get_authors_string <- function() {
-  authors_file <- here::here("private_data", "authors.csv")
+  authors_file <- here::here("manuscript", "authors.csv")
   if(!file.exists(authors_file)) {
     return("Authors file not found")
   }

### 2020-11-26 manuscript/analysis_brms.Rmd
Version as sent for approval to co-authors

@@
 author: "Paul Bürkner"
-output:
-  html_document:
-    theme: default
+output: bookdown::pdf_document2
+
 encoding: UTF-8
@@
 
-```{r setup, include=FALSE, cache=FALSE}
-knitr::opts_chunk$set(cache = TRUE, fig.width = 10)
-```
-
-```{r general_options, cache=FALSE}
+```{r setup_brms, include=FALSE, cache=FALSE}
+knitr::opts_chunk$set(cache = TRUE, echo=FALSE, fig.width = 10)
+```
+
+```{r general_options, cache=FALSE, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
 devtools::load_all()
@@
 options(mc.cores = 4)
-theme_set(bayesplot::theme_default())
+#theme_set(bayesplot::theme_default())
 
@@
 
-## Data preparation
+Here we use a set of Bayesian linear models using the `brms` package. There are several classes of model:
+
+- Binary survival models - logistic regression with survival as the outcome.
+- Survival time models - Bayesian version of the single-outcome Cox proportional hazards model for the "Death" event.
+- Binary discharged models - logistic regression with discharged as the outcome (differs from binary survival in the handling of censored patients)
+- Discharged time models - Bayesian version of the single-outcome Cox proportional hazards model for the "Discharged" event
+- Categorical models - categorical (multinomial) regression of the "Discharged"/"Hospitalized"/"Death" outcome.
+
+For the final outcome models (binary and categorical) we use three variants of dealing with censoring:
+
+- All: Remove censored patients and keep only those that have data until final outcome
+- 7 days: Remove patients with less than 7 days of data and treat the state at day 7 post admission as the outcome
+- 28 days: Same as 7 days, but with a longer time frame
+
+The 7 and 28 days models should be somewhat less biased as censored patients are more likely those that spent longer time in hospital and thus have likely more severe disease.
+We marked the "All" models as "Suspicious" for the purpose of the multiverse analysis.
+
+For each model we show the `brms` formula used, summary of fitted coefficients.
+See the `brms` manual for detailed explanation of each model.
+Additionally we show some _posterior predictive checks_ [@http://zotero.org/users/5567156/items/42DSAQX9; @http://zotero.org/users/5567156/items/ZKWU8INU] to verify that models represent the data well.
+Consult the `bayesplot` package manual for details on the checks used.
+As another way to check for potential overfitting we use the `loo` package [@http://zotero.org/users/5567156/items/88ANHBVF] and its Pareto-k diagnostic.
+
+All hypotheses are evaluated via model coefficients of predictors representing treatments, except for sign corrections (e.g.
+in  binary survival models positive effects represent increased survival while for the main analysis we take positive effect as increased mortality, so we flip the sign).
+
+Some derived predictors are used `age_norm` (age, normalized to have roughly mean 0 and standard deviation of 1), `ever_az`, `ever_hcq`, `ever_favipiravir` (whether the patient took the respective medications at any time) and `comorbidities_sum` - the sum of comorbidities as described in Section \@ref(comorbiditiessum).
 
@@
 
-### Notes:
-
-- days_from_symptom_onset. Days from symptom onset until hospitalization. 
-Uncertain marker but contains some information. 
-Required for anti viral (< 7 days) hypothesis.
-- first_day_*: time from day 0 (arrived at hospital) to getting antivirals
-- some sites only support patients with mild disease and some only with
-strong disease symptoms
-
-## Analysis: Survival
-
-### Basic binary model
-
-```{r, results="hide"}
+## Binary survival models
+
+All variants of the model use the same `brms` formula, with family `bernoulli("logit")`.
+
+```{r}
 form_survival_base <- bf(
@@
 
+print_formula(form_survival_base)
+```
+
+### Basic binary model
+
+```{r, results="hide"}
+
 prior_survival_base <- 
@@
 
+The fitted coefficients:
+
 ```{r}
 summary(fit_survival_base)
-pp_check(fit_survival_base, "error_binned")
-loo(fit_survival_base)
-# ce <- conditional_effects(fit_survival_base)
-# plot(ce, ask = FALSE)
-```
-
-```{r}
-#Marking the results as suspicious because there was a divergence...
+invisible(loo(fit_survival_base)) #loo used primarily for checks not for the actual values
+```
+
+The `error_binned` posterior predictive check - showing the distribution of errors per predicted probability bin.
+
+```{r}
+pp_check(fit_survival_base, "error_binned", nsamples = 20)
+
+```
+
+```{r}
+#Marking the results as suspicious because of the potential bias
 draws <- tidy_draws(fit_survival_base)
@@
 )
+
 ```
@@
 
+The fitted coefficients:
+
 ```{r}
 summary(fit_survival_base_7)
-pp_check(fit_survival_base_7, "error_binned")
-loo(fit_survival_base_7)
+
 # ce <- conditional_effects(fit_survival_base_7)
 # plot(ce, ask = FALSE)
+```
+
+The `error_binned` check:
+
+```{r}
+pp_check(fit_survival_base_7, "error_binned", nsamples = 20)
+invisible(loo(fit_survival_base_7))
 ```
@@
 
+The fitted coefficients:
+
 ```{r}
 summary(fit_survival_base_28)
-pp_check(fit_survival_base_28, "error_binned")
-loo(fit_survival_base_28)
+
 # ce <- conditional_effects(fit_survival_base_28)
@@
 ```
+Note that there is a divergent transition indicating potential computational problems, we will mark the model as "Suspicious".
+
+The `error_binned` check:
+
+```{r}
+pp_check(fit_survival_base_28, "error_binned", nsamples = 20)
+invisible(loo(fit_survival_base_28))
+```
 
@@
 
-## Analysis: Survival time
+## Survival time models
 
@@
 ### Fewer adjustments
+
+Those models are described by the `brms` formula with family `brmsfamily("cox", "log")`:
 
@@
 )
+print_formula(form_survival_time_small)
+
+```
+
+```{r}
 
@@
 
+Here are the fitted coefficients:
+
 ```{r}
 summary(fit_survival_time_small)
-loo(fit_survival_time_small)
+invisible(loo(fit_survival_time_small))
 ```
@@
 ### Fewer adjustments - First wave only
+
+The same model as above, but using data from the first wave only. Here is the summary of the fit:
 
@@
 
+For those models the formula changes to:
+
 ```{r}
@@
 )
+print_formula(form_survival_time)
+```
+
+```{r}
 
@@
 
+Here are the fitted coefficients - there is a divergent transition so we are possibly fitting an overparemtrized model and we will mark it as "Suspicious":
+
 ```{r}
 summary(fit_survival_time)
+```
+
+In addition, the LOO diagnostics warn us that there were some very influential observations, indicating possible overfitting.
+
+```{r}
 loo(fit_survival_time)
-```
-There were some very influential observations.
+
+```
 
@@
 ### More adjustments - First wave only
+
+The same model as above, but fitting only to patients from first wave. Here are the fitted model coefficients:
 
@@
 summary(fit_survival_time_fw)
+```
+
+ We do not see divergent transitions, but some overly influential observations (according to `loo`) remain, the model will be marked as "Suspicious".
+
+```{r}
 loo(fit_survival_time_fw)
 ```
-
-There were some very influential observations.
 
@@
 
-## Analysis: Discharged
-
-### Basic binary model
-
-```{r, results="hide"}
+## Binary discharged models
+
+All the models in this section use the `bernoulli("logit")` family (i.e. logistic regression) and follow the `brms` formula
+
+```{r}
 form_discharged_base <- bf(
@@
 
+print_formula(form_discharged_base)
+```
+
+### Basic binary model
+
+```{r, results="hide"}
+
 prior_discharged_base <- 
@@
 
+Here is the summary of the fitted coefficients
+
 ```{r}
 summary(fit_discharged_base)
-pp_check(fit_discharged_base, "error_binned")
-loo(fit_discharged_base)
+invisible(loo(fit_discharged_base))
 # ce <- conditional_effects(fit_discharged_base)
 # plot(ce, ask = FALSE)
+```
+And the `error_binned` check:
+
+```{r}
+pp_check(fit_discharged_base, "error_binned", nsamples = 20)
+
 ```
@@
 
+Here is the summary of the fitted coefficients:
+
 ```{r}
 summary(fit_discharged_base_7)
-pp_check(fit_discharged_base_7, "error_binned")
-loo(fit_discharged_base_7)
+invisible(loo(fit_discharged_base_7))
 # ce <- conditional_effects(fit_discharged_base_7)
 # plot(ce, ask = FALSE)
+```
+
+And the `error_binned` check.
+
+```{r}
+pp_check(fit_discharged_base_7, "error_binned", nsamples = 20)
+
 ```
@@
 
+Here are the fitted coefficients:
+
 ```{r}
 summary(fit_discharged_base_28)
-pp_check(fit_discharged_base_28, "error_binned")
-loo(fit_discharged_base_28)
+invisible(loo(fit_discharged_base_28))
 # ce <- conditional_effects(fit_discharged_base_28)
@@
 
+And the `error_binned` check:
+
+```{r}
+pp_check(fit_discharged_base_28, "error_binned")
+
+```
+
 ```{r}
@@
 
-## Analysis: discharged time
-
-```{r, results="hide"}
+## Discharged time model
+
+The model uses the family `brmsfamily("cox", "log")` and the formula:
+
+```{r}
 form_discharged_time <- bf(
@@
 
+print_formula(form_discharged_time)
+```
+
+```{r, results="hide"}
 prior_discharged_time <- 
@@
 
+The fitted coefficients are:
+
 ```{r}
 summary(fit_discharged_time)
+
+```
+
+The `loo` diagnostics warn us that there were some very influential observations.
+
+```{r}
 loo(fit_discharged_time)
 ```
-
-There were some very influential observations
 
@@
 
-## Analysis: Hospitalized vs. Discharged vs. Death
-
-### Basic Categorical model
-
-```{r, results="hide"}
+## Categorical models
+
+All the categorical models use the `categorical("logit")` family and the following formula:
+
+```{r}
 form_outcome_base <- bf(
@@
 )
+print_formula(form_outcome_base)
+```
+
+### Basic Categorical model
+
+```{r, results="hide"}
 
@@
 
+Fitted coefficients below:
+
 ```{r}
 summary(fit_outcome_base)
-loo(fit_outcome_base)
+invisible(loo(fit_outcome_base))
 # ce <- conditional_effects(fit_outcome_base)
@@
   prior = prior_outcome_base,
-  control = default_control,
+  control = list(adapt_delta = 0.98),
   file = paste0(cache_dir, "/fit_outcome_base_7")
@@
 
+Summary of model fit:
+
 ```{r}
 summary(fit_outcome_base_7)
-pp_check(fit_outcome_base_7)
-loo(fit_outcome_base_7)
+invisible(loo(fit_outcome_base_7))
 # ce <- conditional_effects(fit_outcome_base_7)
 # plot(ce, ask = FALSE)
+```
+
+```{r, fig.width=4, fig.height=2.5}
+#pp_check(fit_outcome_base_7, type = "bars", nsamples = 200)
+
 ```
@@
 ### Basic Categorical model (7 Days), first wave only
+
+The same model as above, using only data from the first wave. Summary below:
 
@@
 summary(fit_outcome_base_7_fw)
-pp_check(fit_outcome_base_7_fw)
-loo(fit_outcome_base_7_fw)
+invisible(loo(fit_outcome_base_7_fw))
 # ce <- conditional_effects(fit_outcome_base_7_fw)
 # plot(ce, ask = FALSE)
+```
+
+```{r, fig.width=4, fig.height=2.5}
+#pp_check(fit_outcome_base_7_fw, type = "bars", nsamples = 200)
+
 ```
@@
 
+Summary of the model is:
+
 ```{r}
 summary(fit_outcome_base_28)
-pp_check(fit_outcome_base_28)
-loo(fit_outcome_base_28)
-# ce <- conditional_effects(fit_outcome_base_28)
-# plot(ce, ask = FALSE)
+#pp_check(fit_outcome_base_28)
+invisible(loo(fit_outcome_base_28))
 ```
@@
 ### Basic Categorical model (28 Days), first wave only
+
+The same model, but only for patients from the first wave, summary of the fit:
 
@@
 summary(fit_outcome_base_28_fw)
-pp_check(fit_outcome_base_28_fw)
-loo(fit_outcome_base_28_fw)
-# ce <- conditional_effects(fit_outcome_base_28_fw)
-# plot(ce, ask = FALSE)
+#pp_check(fit_outcome_base_28_fw)
+invisible(loo(fit_outcome_base_28_fw))
 ```
@@
 
-## Analysis: Worst Breathing {.tabset}
-
-### Basic ordinal model
+## Worst Breathing {.tabset}
+
+Not reported in the main manuscript but we also experimented with ordinal (stopping ratio) models for the worst breathing level experienced by a patient.
 
@@
 summary(fit_worst_breathing_base)
-loo(fit_worst_breathing_base)
+invisible(loo(fit_worst_breathing_base))
 # ce <- conditional_effects(fit_worst_breathing_base, categorical = TRUE)
-# plot(ce, ask = FALSE)
-```
-
-### Category-specific ordinal model
-
-```{r, results="hide"}
-# form_worst_breathing_cs <- bf(
-#   worst_breathing ~ cs(sex + age_norm + 
-#     smoking + mo(obesity) + ever_hcq + ever_az + ever_favipiravir) +
-#     (1 | hospital_id),
-#   family = sratio("logit")
-# )
-# 
-# prior_worst_breathing_cs <- 
-#   prior(normal(0, 5), class = "b") +
-#   prior(normal(0, 5), class = "sd")
-# 
-# fit_worst_breathing_cs <- brm(
-#   formula = form_worst_breathing_cs,
-#   data = data$patient_data,
-#   prior = prior_worst_breathing_cs,
-#   inits = 0,
-#   file = paste0(cache_dir, "/fit_worst_breathing_cs")
-# )
-# fit_worst_breathing_cs <- add_criterion(fit_worst_breathing_cs, "loo")
-```
-
-```{r}
-# summary(fit_worst_breathing_cs)
-# ce <- conditional_effects(fit_worst_breathing_cs, categorical = TRUE)
 # plot(ce, ask = FALSE)

### 2020-11-26 manuscript/competing_risks.Rmd
Version as sent for approval to co-authors

@@
 title: "Survival - frequentist"
-output: html_notebook
+output: bookdown::pdf_document2
 ---
 
-```{r}
+```{r setup_competing, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+knitr::opts_chunk$set(echo = FALSE)
 devtools::load_all()
@@
 
-# Competing risks death vs. discharged
+# Competing risks proportional hazards models
+
+In all of those models we use the `coxph` function from the `Survival` package and treat death and discharged as competing risks.
+The `status` column in the data corresponds to the three-valued state "Hospitalized" (1) / "Discharged" (2) / "Death" (3).
+I.e.
+in all model outputs below, there are two transitions "1:2" is the transition to "Discharged" and "1:3" is the transition to "Death".
+We evaluate the hypotheses about treatment effects or marker associations directly via the estimated log hazard ratio of the treatment for the relevant transition.
+Notably, the sign of the effect on the transition to "Discharged" is flipped for the multiverse analysis because it assumes that positive effect means increased hospital stay.
+
+We do not work with models with more than about 5 predictors, as those become impossible to estimate due to insufficient amount of events.
 
@@
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) 
+```
 
-f_competing_hcq <- Surv(daym1, day,status) ~ age + sex + took_hcq
+## Treatment effects
 
-#survcheck(f_competing_hcq, data = data_surv, id = patient_id)
+### Age + sex + hydroxychloroquine
+
+The model uses the formula:
+
+```{r}
+f_competing_hcq <- Surv(daym1, day, status) ~ age + sex + took_hcq
+f_competing_hcq
+```
+Fitted coefficients below:
+
+```{r}
+
 fit_competing_hcq <- coxph(f_competing_hcq, data = data_surv, id = patient_id)
@@
 
-hypothesis_res_list[["competing_hcq"]]
+#hypothesis_res_list[["competing_hcq"]]
 ```
+
+### Age + sex + hydroxychloroquine + azithromycin
+
+The formula becomes:
 
@@
 f_competing_hcq_az <- update.formula(f_competing_hcq, . ~ . + took_az)
+f_competing_hcq_az
+```
+
+The fitted coefficients are:
+
+```{r}
+
 fit_competing_hcq_az <- coxph(f_competing_hcq_az, data = data_surv, id = patient_id)
@@
   )
-hypothesis_res_list[["competing_hcq_az"]]
+#hypothesis_res_list[["competing_hcq_az"]]
 ```
+
+### Age + sex + hydroxychloroquine + azithromycin + favipiravir
+
+The formula is:
 
@@
 f_competing_hcq_az_fpv <- update.formula(f_competing_hcq_az, . ~ . + took_favipiravir)
+f_competing_hcq_az_fpv
+```
+
+And the fitted model is:
+
+```{r}
+
 fit_competing_hcq_az_fpv <- coxph(f_competing_hcq_az_fpv, data = data_surv, id = patient_id)
@@
   )
-hypothesis_res_list[["competing_hcq_az_fpv"]]
+#hypothesis_res_list[["competing_hcq_az_fpv"]]
 ```
+
+### Age + sex + hydroxychloroquine + azithromycin + admitted
+
+The formula is:
 
@@
 f_competing_hcq_az_admitted <- update.formula(f_competing_hcq_az, . ~ . + admitted_for_covid)
+f_competing_hcq_az_admitted
+```
+
+The fitted coefficients are:
+
+```{r}
 fit_competing_hcq_az_admitted <- coxph(f_competing_hcq_az_admitted, data = data_surv, id = patient_id)
@@
   )
-hypothesis_res_list[["competing_hcq_az_admitted"]]
+#hypothesis_res_list[["competing_hcq_az_admitted"]]
 ```
 
+### Age + sex + hydroxychloroquine + azithromycin + hospital
+
+The new formula is:
+
 ```{r}
-f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + hospital_id)
+f_competing_hcq_az_hospital <- update.formula(f_competing_hcq_az, . ~ . + strata(hospital_id))
+f_competing_hcq_az_hospital
+```
+
+The resulting coefficients are:
+
+```{r}
+
 data_surv_hospital <- data_surv %>% 
@@
   )
-hypothesis_res_list[["competing_hcq_az_hospital"]]
+#hypothesis_res_list[["competing_hcq_az_hospital"]]
 
@@
 
+### Age + sex + hydroxychloroquine + azithromycin + comorbidities
+
+The formula becomes:
+
 ```{r}
 f_competing_hcq_az_comorbidities_sum <- update.formula(f_competing_hcq_az, . ~ . + comorbidities_sum)
+f_competing_hcq_az_comorbidities_sum
+```
+
+Model fit below:
+
+```{r}
 fit_competing_hcq_az_comorbidities_sum <- coxph(f_competing_hcq_az_comorbidities_sum, data = data_surv, id = patient_id)
@@
   )
-hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
+#hypothesis_res_list[["competing_hcq_az_comorbidities_sum"]]
 ```
 
-# Peak IL-6, peak D-dimer
+## Marker effects
+
+### Age + sex + IL-6
+
+The formula is:
 
@@
 f_competing_il_6 <- update.formula(f_competing_marker_base, . ~ . + log_peak_IL_6)
+f_competing_il_6
+```
+
+where `log_peak_IL_6` is the logarithm of the peak IL-6 value seen up to the given time. The fitted model is:
+
+```{r}
 
@@
                                         coefficient_name = "log_peak_IL_6", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
-hypothesis_res_list[["competing_il_6"]]
+#hypothesis_res_list[["competing_il_6"]]
 ```
+Note the warning on convergence and the huge `se` for the `log_peak_IL_6` coefficient for the 1:2 transition. We mark this model as "Suspicious".
+
+### Age + sex + D-dimer
+
+The formula is:
+
 ```{r}
 f_competing_d_dimer <- update.formula(f_competing_marker_base, . ~ . + log_peak_d_dimer)
+f_competing_d_dimer
+```
+
+where `log_peak_d_dimer` is the logarithm of the peak D-dimer value seen up to the given time. The fitted model is:
+
+```{r}
 
@@
   frequentist_hypothesis_res_from_coxph(hypotheses$d_dimer_death, fit_competing_d_dimer,
-                                        coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Problematic")
-hypothesis_res_list[["competing_d_dimer"]]
+                                        coefficient_name = "log_peak_d_dimer", transition = "1:3", adjusted = "age, sex", model_check = "Suspicious")
+#hypothesis_res_list[["competing_d_dimer"]]
 ```
+We see similar, but even more extreme convergence issues as before, we mark the model "Suspicious".
 

### 2020-11-26 manuscript/hmm.Rmd
Version as sent for approval to co-authors

@@
 
-```{r setup_hmm, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+```{r setup_hmm, echo=FALSE, message=FALSE, warning = FALSE, results = "hide",cache=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
@@
 We treat the breathing support required as a Markov process, either with states directly corresponding to the breathing support and directly observable (Fig.
-\@ref(fig:hmmstates)a), or with the states carrying an unobserved binary improving/worsening component which is unobservable (Fig.
+\@ref(fig:hmmstates)a), or with the states carrying a binary improving/worsening component which is unobservable (Fig.
 \@ref(fig:hmmstates)b) while the breathing support dimension still being completely observable.
 The former completely observable model is called _simple_ in the following text while the latter is called _complex_.
-This is unlike many uses of HMMs where the observation process is described by a general observation matrix.
+This results in a very restricted observation matrix.
 The reason is that the breathing support actually used is very likely to correspond to the breathing support needed by the patient.
@@
 
-Models will be presented in a unified structure that we explain with the first model and won't be repeated
+Models will be presented in a unified structure that we explain with the first model and will not be repeated
 
@@
 
-For this model, the $\log(R_{i,j})$ elements are then modeled via the `brms` formula
+For this model, the $\log(R_{i,j})$ elements are modeled via the `brms` formula
 
@@
 
-Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:statessimple)) and those taken from patient's data (both the fixed baseline characteristics and time-varying values).
+Our code is setup so that the formula can combine predictors drawn from the description of rates (see Table \@ref(tab:ratessimple)) and those taken from patient's data (both the fixed baseline characteristics and time-varying values).
 There is a separate coefficient for each rate and then the other predictors act uniformly on each rate group.
@@
 
-And the resulting log ORs:
+And the resulting log ORs/log duration ratios:
 
@@
 ```
+Where `comorbidities_sum` is a comorbidity score as described in Section \@ref(comorbiditiessum).
 

### 2020-11-26 manuscript/jm.Rmd
Version as sent for approval to co-authors

@@
 title: "Joint modelling"
-output: html_document
+output: bookdown::pdf_document2
+bibliography: Covid19.json
 ---
@@
 
-```{r setup}
+Joint models have two components:
+
+- _longitudinal model_ describes the evolution of some quantity/quantities over time (in our case a biological marker)
+- _time-to-event model_ describes the risk of event over time assuming proportional hazards. This risk can be affected by the value of the marker.
+
+Modelling those two jointly means that the marker values are imputed between measurements according to the longitudinal model, but information also flows in the other direction.
+We use the implemention from the `rstanarm` package [@http://zotero.org/users/5567156/items/KHJGLQS5] - consult the reference for an introduction to this class of models.
+
+We use those models only for association between markers (IL-6 and D-dimer) and disease progression. Those analyses ended up not being reported in the main paper as the results are inconclusive.
+
+We evaluate the associations directly via the coefficient describing the association between the longitudinal and time-to-event component. We focus solely on the "Death" event.
+
+```{r setup_jm, cache=FALSE, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+knitr::opts_chunk$set(cache = TRUE, echo=FALSE)
+
 devtools::load_all()
@@
 
+## D-dimer
+
+### Age + sex
+
+The longitudinal formula is `log_d_dimer ~ day + (1  | patient_id)` and the time-to-event formula is `Surv(daym1, day, event_death) ~ age + sex`. The fitted coefficients are:
+
 ```{r}
@@
 
-#data_d_dimer %>% filter(d_dimer == 0)
-# data_d_dimer %>% group_by(sex, ever_hcq) %>% summarise(pts = length(unique(patient_id)), n_dimer = sum(!is.na(d_dimer)))
 fit_d_dimer_age_sex <- stan_jm_with_cache(formulaLong = log_d_dimer ~ day + (1  | patient_id),
@@
 fit_d_dimer_age_sex
-
-#length(unique(data_d_dimer$patient_id))
-# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
-#   filter(n_d_dimer > 1) %>% pull(patient_id)
-#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])
 
@@
 
-jm_hypothesis_res_list[["d_dimer_age_sex"]]
+#jm_hypothesis_res_list[["d_dimer_age_sex"]]
 ```
+
+### Age + sex + hydroxychloroquine
+
+The time-to-event formula is expanded to `Surv(daym1, day, event_death) ~ age + sex + took_hcq`. The fitted coefficients are:
 
@@
 fit_d_dimer_age_sex_hcq
-
-#length(unique(data_d_dimer$patient_id))
-# to_pred <- data_d_dimer %>% group_by(patient_id) %>% summarise(n_d_dimer = sum(!is.na(d_dimer))) %>%
-#   filter(n_d_dimer > 1) %>% pull(patient_id)
-#data_d_dimer %>% filter(patient_id == unique(data_d_dimer$patient_id)[31])
-
 ```
@@
 
-jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]]
+#jm_hypothesis_res_list[["d_dimer_age_sex_hcq"]]
 ```
 
-```{r}
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[1:11])
-# plot(p1, plot_observed = TRUE)                     
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[12:14])
-# plot(p1, plot_observed = TRUE)                     
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[15:18])
-# plot(p1, plot_observed = TRUE)                     
-# 
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[19:20])
-# plot(p1, plot_observed = TRUE)                     
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[21:22])
-# plot(p1, plot_observed = TRUE)                     
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[22:26])
-# plot(p1, plot_observed = TRUE)
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[27:29])
-# plot(p1, plot_observed = TRUE)                     
-# p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[30:32])
-# plot(p1, plot_observed = TRUE)                     
+The problems in the model are visible when visualising the fitted longitudinal trajectories for some patients: there are very few measurements and thus huge uncertainties:
+
+```{r, warning = FALSE, message = FALSE}
+ p1 <- posterior_traj(fit_d_dimer_age_sex_hcq, ids = unique(data_d_dimer$patient_id)[23:28])
+ plot(p1, plot_observed = TRUE)                     
 ```
+
+## IL-6
+
+### Age + sex
+
+The longitudinal formula is `log_IL_6 ~ day + (1 + day | patient_id)` and the time-to-event formula is `Surv(daym1, day, event_death) ~ age + sex`. The fitted coefficients are:
 
@@
 
-jm_hypothesis_res_list[["IL_6_age_sex"]]
+#jm_hypothesis_res_list[["IL_6_age_sex"]]
 
 ```
+
+### Age + sex + hydroxychloroquine
+
+The time-to-event formula becomes `Surv(daym1, day, event_death) ~ age + sex + took_hcq`. The fitted model is:
 
@@
 
-jm_hypothesis_res_list[["IL_6_age_sex_hcq"]]
+#jm_hypothesis_res_list[["IL_6_age_sex_hcq"]]
 
@@
 
-```{r}
-for(id in unique(data_IL_6$patient_id)) {
-  p1 <- posterior_traj(fit_IL_6_age_sex_hcq, ids = id)
-  print(plot(p1, plot_observed = TRUE)                     )
-}
+Similarly, we can explore the posterior trajectories, showing similar issues as before: too wide uncertainty.
 
+```{r, message = FALSE}
+ p1 <- posterior_traj(fit_IL_6_age_sex_hcq, ids = unique(data_IL_6$patient_id)[c(8,14,23,13,6,1)])
+ plot(p1, plot_observed = TRUE)                     
 ```

### 2020-11-26 manuscript/manuscript.Rmd
Version as sent for approval to co-authors

@@
 Weak evidence to constrain the potential efficacy of azithromycin and favipiravir can be extracted from the data.
-Additionally we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
+Additionally, we performed external validation of several proposed prognostic models for Covid-19 severity showing that they mostly perform unsatisfactorily on our dataset.
 bibliography: Covid19.json

### 2020-11-26 manuscript/single_outcome_coxph.Rmd
Version as sent for approval to co-authors

--- 2020-10-27 manuscript/single_outcome_coxph.Rmd
@@
 title: "Survival - frequentist, single outcome"
-output: html_notebook
+output:
+  bookdown::pdf_document2
 ---
 
-```{r}
-devtools::load_all()
+# Single outcome proprtional hazards model
+
+Here we use a single outcome survival model via the `coxph` function of the `Survival` package. Hydroxychloroquine is treated as a time-varying covariate, other covariates are fixed.
+
+```{r setup_coxph_single, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+knitr::opts_chunk$set(echo = FALSE)
+devtools::load_all(here::here())
 library(tidyverse)
@@
         if (hcq_first>1) {
-          d<-rbind(d,data.frame(t0=0,t1=hcq_first-1,outcome=0,patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
+          d<-rbind(d,data.frame(t0=0,t1=hcq_first-1,outcome=0,patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],first_wave=pd$first_wave[cnd],hcq=0))
         }
-        d<-rbind(d,data.frame(t0=hcq_first-1,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=1))
+        d<-rbind(d,data.frame(t0=hcq_first-1,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],first_wave=pd$first_wave[cnd],hcq=1))
       } else {
-        d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
+        d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],first_wave=pd$first_wave[cnd],hcq=0))
       }
     } else { # hcq_first=0
-      d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=1))
+      d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],first_wave=pd$first_wave[cnd],hcq=1))
     }
   } else {
-    d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],hcq=0))
+    d<-rbind(d,data.frame(t0=0,t1=pd$last_record[cnd],outcome=pd$outcome[cnd],patientId=i,hcq_first=pd$hcq_first[cnd],last_record=pd$last_record[cnd],age=pd$age[cnd],sex=pd$sex[cnd],hospital_id=pd$hospital_id[cnd],admitted_for_covid=pd$admitted_for_covid[cnd],ischemic_heart_disease=pd$ischemic_heart_disease[cnd],has_hypertension_drugs=pd$has_hypertension_drugs[cnd],heart_failure=pd$heart_failure[cnd],renal_disease=pd$renal_disease[cnd],high_creatinin=pd$high_creatinin[cnd],COPD=pd$COPD[cnd],other_lung_disease=pd$other_lung_disease[cnd],renal_disease=pd$renal_disease[cnd],obesity=pd$obesity[cnd],first_wave=pd$first_wave[cnd],hcq=0))
   }
@@
 
-cache_dir <- here::here("local_temp_data")
-if(!dir.exists(cache_dir)) {
-  dir.create(cache_dir)
-}
+```
+
+## Risk of death, increased discharge
+
+Here, we compare the model with hydroxychloroquine as a covariate (`m1`) with model not including hydroxychloroquine (`m0`), the formulas used are:
 
 ```
+m1: Surv(t0,t1,outcome=='Death')~hcq+age+sex+strata(hospital_id)+comorb_sum)
+m0: Surv(t0,t1,outcome=='Death')~    age+sex+strata(hospital_id)+comorb_sum)
+```
 
-# Risk of death, increased discharge
+where `comorb_sum` is a score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, other lung diseases, renal disease, or high creatinine.
+A separate model is fitted for the "Discharged" outcome.
+
+The point estimate of the effect and its confidence intervals are taken directly from the `m1` model, a p-value is computed via a call `anova` comparing the two models.
+
+Here are the fitted models for the "Death" outcome:
 
@@
 
+m1
+m0
+
+```
+
+And the fitted models for the "Discharged" outcome:
+
+```{r}
 mt1<-with(d,coxph(S<-Surv(t0,t1,outcome=='Discharged')~hcq+age+sex+strata(hospital_id)+comorb_sum))
@@
 
+mt1
+mt0
+```
+
+And here are the resulting evaluation of individual hypotheses:
+
+```{r}
 hypothesis_res_list[["competing_hcq_comorbidities_sum"]] <-
   rbind(
-  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_reduces_death, 
+  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_death, 
                                          point_estimate = coef(m1)[1],
-                                         adjusted = "age, sex, comorbidities (sum)", 
+                                         adjusted = "age, sex, hospital, comorbidities (sum)", 
                                          test_stat = anova(m1,m0,test='Chisq')[2,2],
@@
                                          p = anova(mt1,mt0,test='Chisq')[2,4],
-                                         ci_low = confint(mt1)[1,1],
-                                         ci_high = confint(mt1)[1,2])
+                                         ci_low = -confint(mt1)[1,2],
+                                         ci_high = -confint(mt1)[1,1])
   )
-hypothesis_res_list[["competing_hcq_comorbidities_sum"]]
+hypothesis_res_list[["competing_hcq_comorbidities_sum"]] %>% select(hypothesis, point_estimate, p_value, ci_low, ci_high)
+
+```
+
+## Risk of death, increased discharge - first wave only
+
+Here, we perform the same analysis as above, but including only patients from the first wave, here are the fitted models for the "Death" outcome:
+
+```{r}
+d_first <- d %>% filter(first_wave)
+m1_first<-with(d_first,coxph(S<-Surv(t0,t1,outcome=='Death')~hcq+age+sex+strata(hospital_id)+comorb_sum))
+m0_first<-with(d_first,coxph(S<-Surv(t0,t1,outcome=='Death')~    age+sex+strata(hospital_id)+comorb_sum))
+
+m1_first
+m0_first
+
+```
+
+And here are the fits for the "Discharged" outcome:
+
+```{r}
+mt1_first<-with(d_first,coxph(S<-Surv(t0,t1,outcome=='Discharged')~hcq+age+sex+strata(hospital_id)+comorb_sum))
+mt0_first<-with(d_first,coxph(S<-Surv(t0,t1,outcome=='Discharged')~    age+sex+strata(hospital_id)+comorb_sum))
+
+mt1_first
+mt0_first
+```
+
+Which give us the following evaluation of the hypotheses:
+
+```{r}
+hypothesis_res_list[["competing_hcq_comorbidities_sum_fw"]] <-
+  rbind(
+  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_death, 
+                                         point_estimate = coef(m1_first)[1],
+                                         adjusted = "age, sex, hospital, comorbidities (sum), first_wave", 
+                                         test_stat = anova(m1_first,m0_first,test='Chisq')[2,2],
+                                         df = anova(m1_first,m0_first,test='Chisq')[2,3],
+                                         p = anova(m1_first,m0_first,test='Chisq')[2,4],
+                                         ci_low = confint(m1_first)[1,1],
+                                         ci_high = confint(m1_first)[1,2]),
+  frequentist_hypothesis_res_from_coxph1(hypotheses$hcq_hospital, 
+                                         point_estimate = -coef(mt1_first)[1],
+                                         adjusted = "age, sex, hospital, comorbidities (sum), first_wave", 
+                                         test_stat = anova(mt1_first,mt0_first,test='Chisq')[2,2],
+                                         df = anova(mt1_first,mt0_first,test='Chisq')[2,3],
+                                         p = anova(mt1_first,mt0_first,test='Chisq')[2,4],
+                                         ci_low = -confint(mt1_first)[1,2],
+                                         ci_high = -confint(mt1_first)[1,1])
+  )
+
+hypothesis_res_list[["competing_hcq_comorbidities_sum_fw"]] %>% select(hypothesis, point_estimate, p_value, ci_low, ci_high)
 

### 2020-11-26 manuscript/supplementary_statistical_analysis.Rmd
Version as sent for approval to co-authors

---
site: "bookdown::bookdown_site"
title: "Supplementary statistical analysis for \"Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis\""
output:
  bookdown::pdf_document2:
    toc_depth: 2
bibliography: "Covid19.json"
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
abstract: This documents describes additional details about the models and statistical analyses used in the paper, including fitted coefficient values. Additionally, details on the risk scores and how we evaluated them are provided. Chapters of the document correspond to individual model families we used, final chapter is devoted to risk scores evaluation. Overall description of the models and the outputs presented is given at the beginning of each chapter and is not repeated for each model. To avoid any ambiguities, complete code to compile this document and for all analyses is available at https://github.com/cas-bioinf/covid19retrospective/.
---

```{r setup, echo=FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=FALSE, cache = TRUE, cache.lazy = FALSE)
devtools::load_all()
```

# General remarks on models

For all treatments under investigation, we try to quantify their association with both mortality and length of hospitalization. The exact ways those are represented differs between models.

We sort models into 3 groups:

- "Problematic" - model is almost certainly not useful, especially because of computational issues.
- "Suspicious" - model has some problems representing the data or minor computational issues. It is likely the results are not completely trustworthy, but they are also unlikely to be completely useless. 
- "OK" - we found no problems with the model.

In the main text, we report only "OK" models, but supplementary figures show the multiverse analysis performed for all models we tried.

To make sure our inferences are not skewed by including patients from both first and second waves of the pandemic, we run most analyses twice: once with those patients included and once without them.

## Comorbidity score {#comorbiditiessum}

Multiple models share the same comorbidity score that is computed as the count of the following 9 conditions that are met (missing data are treated as 0):

- Heart problems (NYHA > 1 OR ischemic heart disease OR heart failur)
- Hypertension (uses at least 1 hypertension drug)
- Lung problems (COPD OR asthma OR other lung disease)
- Diabetes
- Renal disease
- Liver disease
- Smoking
- Markers out of range (creatinine > 115 for males, creatinine > 97 for females OR PT INR > 1.2 OR albumin < 36)
- Obesity (as ascertained by a clinician OR BMI > 30)

All of those quantities were evaluated upon admission. See the protocol of the study for more details on the definitions.

\newpage

```{r child_coxph_single, child = 'single_outcome_coxph.Rmd'}
```

\newpage

```{r child_competing, child = 'competing_risks.Rmd'}
```

\newpage

```{r child_brms, child = 'analysis_brms.Rmd'}
```

\newpage

```{r child_hmm, child = 'hmm.Rmd'}
```

\newpage

```{r child_jm, child = 'jm.Rmd'}
```

\newpage

```{r child_risk_scores, child = 'risk_scores.Rmd'}
```

# Original computing environment

```{r}
sessionInfo()
```

# References

### 2020-11-27 manuscript/supplementary_figures.Rmd
switched to ggplot2

@@
 models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
+models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
 modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
-ys<-c(0)
-for (m in models) {
-  ys<-c(min(min(ys),confint(m)[2,1]),max(max(ys),confint(m)[2,2]))
-}
-models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
-modelNames2<-modelNames
-ys2<-c(0)
-for (m in models2) {
-  ys2<-c(min(min(ys2),confint(m)[2,1]),max(max(ys2),confint(m)[2,2]))
-}
-
-plot(ys,c(1,length(modelNames)),ty='n',yaxt='n',xlab='Log odds ratio',ylab='')
-abline(v=0,col='blue')
-col.hcq<-bfc()["Vermillion"]
-for (i in seq(along=models)) {
+gtmp<-c()
+for (i in rev(seq(along=models))) {
   ii<-length(models)-i+1
@@
   m2<-models2[[i]]
-  lines(confint(m)[2,],rep(ii,2))
-  points(coef(m)[2],ii,pch=19)
-  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
-  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
+  ci<-confint(m)[2,]
+  ci2<-confint(m2)[2,]
+  gtmp<-rbind(gtmp,data.frame(est=coef(m)[2],low=ci[1],high=ci[2],name=modelNames[i],type='outcome'))
+  gtmp<-rbind(gtmp,data.frame(est=-coef(m2)[2],low=-ci2[1],high=-ci2[2],name=modelNames[i],type='hcq'))
 }
-axis(2,length(modelNames):1,modelNames,las=2)
-legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
+ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
+  geom_vline(xintercept=0,color='blue')+
+  geom_point(position=position_dodge(width=0.5))+
+  geom_errorbarh(position=position_dodge(width=0.5),height= 0.5)+
+  scale_x_continuous('Log odds ratio')+
+  scale_y_discrete('')+
+  scale_color_manual('Model of:',values=c('black','red'),labels=c('Death','No ACQ'))+
+  theme(legend.background = element_rect(fill="gray95"),legend.position=c(.83,.7))
 ```

### 2020-11-30 manuscript/manuscript.Rmd
Minor tweaks based on final feedback

@@
 Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
-Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
+As of this writing, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
 Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA.
@@
 Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3].
-Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
+Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
 Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
-Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
+Similarly, Lopinavir/Ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
 No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
@@
 
-High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
+High IL-6, D-dimer values were observed to be associated with worse outcome and increased disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
 Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the disease.
@@
 Not all patients developed pneumonia or other symptoms of Covid-19.
-All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections.
+All patients received the standard of care including supplemental oxygen and ventilation and antibiotics for bacterial superinfections as determined by the attending physician.
 Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition.
 We note that the participating sites were not homogeneous in either patient population or treatment protocols.
-The choice of treatment was based on the decision of the attending clinician and availability.
+The choice of pharmacological treatment was based on the decision of the attending clinician and its availability.
 
@@
 ```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c).
-AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation.
+AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation).
 In all models the 'Death' and 'Discharged' states are terminal.
@@
 Note that the favipiravir subgroup is not exclusive with either the HCQ or No HCQ group.
-IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)
+IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated to undergo more intensive treatment modality.", booktabs = TRUE)
 
@@
 The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization.
-The limited nature of the data however does not allow for any statistically robust conclusions.
+However, the limited nature of the data does not allow for any statistically robust conclusions.
 We also see that the marker levels were not substantially stratified by study site.
@@
 ```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
-Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
+Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation)."}
 my_paste <- function(x) {
@@
 The vertical scale is logarithmic.
-Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
+Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation), CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
 #data$marker_data %>%
@@
 
-Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future hydroxychloroquine treatment:
+Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to the future hydroxychloroquine treatment:
 the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
@@
 
-We also examined the association between treatments and length of hospital stay for patients that survived.
+We also examined the association between treatments and length of hospital stay for all the patients that survived.
 Almost all models cannot discern the sign of the association for all treatments examined (Supplementary Figures `r supplementary_figures_multiverse`).
@@
 All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
-Additionally, some publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed.
+Additionally, some publications did not provide enough detail to unambiguously reconstruct how the score and/or outcome was assessed.
 We thus concur with Wynants et al.
@@
 
-Additionally our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial.
+Additionally, our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial.
 We suggest that comparing the prognostic models against simple baselines (e.g.
 decade of age as the single predictor) should be a first step in validation.
-Further, some of the published scores lack enough information to let others exactly implement the score.
+Furthermore, some of the published scores lack enough information to let others implement the score in the same way.
 

### 2020-12-01 manuscript/supplementary_figures.Rmd
supplementary tables added: treatment combinations and outcomes by hospital

@@
   ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
+```
+
+# Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
+```{r treatment_comb_table, echo=FALSE}
+pd<-data$patient_data
+x<-sort(table(paste0(
+ as.numeric(pd$ever_hcq),
+ as.numeric(pd$ever_az),
+ as.numeric(pd$ever_convalescent_plasma),
+ as.numeric(pd$ever_favipiravir),
+ as.numeric(pd$ever_remdesivir),
+ as.numeric(pd$ever_dexamethasone),
+ as.numeric(pd$ever_antibiotics))),dec=TRUE)
+x2<-c()
+nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
+for (i in 1:length(x)) {
+ x3<-c()
+ for (i2 in 1:nchar(names(x)[i])) {
+   x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
+ }
+ x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
+}
+rownames(x2)<-c()
+colnames(x2)<-c('N',nms)
+x2<-x2[,c(1,2,3,8,4:7)]
+knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
+```
+
+```{r outcome_by_hospital_prepare, echo=FALSE}
+tmp<-data$patient_data[,c('patient_id','hospital_id','outcome')]
+tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
+tmp$outcome<-dropLevels(tmp$outcome)
+tmp$hospital_id<-factor(tmp$hospital_id)
+levels(tmp$hospital_id)<-paste('H',1:length(levels(tmp$hospital_id)),sep='')
+```
+
+There were large differences between hospitals in the ratio of fatal outcomes 
+(P=`r round(fisher.test(with(tmp,table(outcome,hospital_id)),simulate.p.value=T)$p.value,5)`).
+```{r outcome_by_hospital, echo=FALSE}
+knitr::kable(with(tmp,table(outcome,hospital_id)),caption = "Overview of outcomes in different hospitals.
+Note that at H6, 9 out of 19 patients died, while at H9, no patient died out of 9 patients treated.", booktabs = TRUE)
 ```

### 2020-12-01 manuscript/single_outcome_coxph.Rmd

@@
 
-And here are the resulting evaluation of individual hypotheses:
+And here are the resulting evaluation of individual hypotheses - note that we change sign of the coefficient for the "length of hospitalization" hypothesis (modeled here as the risk of discharge) to conform to the overall rule that negative coefficient implies better outcomes for treated patients:
 

### 2020-12-01 other_outputs/fpv_only.Rmd

---
title: "Favipiravir-only images"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8)

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide

```

```{r}
fpv_hospital_id <- wide %>% filter(took_favipiravir) %>% pull(hospital_id) %>% unique()

data_fpv <- filter_complete_data(data, hospital_id == fpv_hospital_id)
wide_fpv <- data_fpv$marker_data_wide
```

```{r, fig.width = 8, fig.height=4}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}

data_for_fig_progression <- wide_fpv  %>% 
  filter(!is.na(breathing), day >= 0) %>%
  arrange(hospital_id, patient_id, day) %>%
  mutate(patient_id = factor(patient_id), patient_id = fct_reorder(patient_id, breathing,  .fun = my_paste, .desc = FALSE)) %>%
  mutate(
         is_outcome =  breathing %in% c("Discharged", "Death"),
         breathing = fct_recode(breathing, `Ambient air` = "AA")
         ) %>% 
  group_by(patient_id) %>%
  mutate(group = if_else(any(took_favipiravir), "Favipiravir", "Other"))

fpv_start <- wide_fpv %>% filter(took_favipiravir) %>%
  group_by(patient_id) %>% 
  summarise(min_day = min(day), group = "Favipiravir")

fig_progression <- data_for_fig_progression %>% 
  filter(!is_outcome) %>%
  ggplot(aes(x = day, y = patient_id, fill = breathing)) + 
  geom_tile(height = 0.7) + 
  geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing, ordered = FALSE)),
             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome), size = 2) +
  geom_text(label = "I", data = fpv_start, size = 10,  hjust = 0.5, inherit.aes = FALSE, aes(x = min_day, y = patient_id) ) +
  scale_fill_viridis_d("", direction = -1) +
  scale_x_continuous("Days since hospitalization") +
  scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
  facet_wrap(~group, scales = "free_y") +
  theme(axis.text.y = element_blank(),
        legend.direction = "horizontal", legend.position = "bottom") +
  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())

fig_progression
```

```{r}
ggsave(here::here("local_temp_data", "progression_fpv.png"), fig_progression, width = 8, height = 4, dpi = 300)
```

```{r, fig.width=8, fig.height=4}
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

  if(length(markers) > 1) {
    facet = facet_grid(marker ~ group, scales = "free_y")
    scale_y_name = "Value"
  } else {
    facet = facet_wrap( ~ group, nrow = 1)
    unit <- data$marker_data %>% filter(marker == names(markers)) %>% pull(unit) %>% unique()
    scale_y_name <- markers
  }

  data_for_plot <- data_fpv$marker_data %>% 
    group_by(patient_id) %>%
    mutate(group = if_else(any(marker == "favipiravir" & value > 0), "Favipiravir", "Other")) %>%
    ungroup() %>%
    filter(marker %in% names(markers), day >= 0) %>% 
    group_by(marker, patient_id) %>%
    filter(n() > 1) %>%
    ungroup() %>%
    mutate(marker = markers[marker])

  if(!is.null(max_patients)) {
    patient_markers_to_use <- data_for_plot %>% 
      select(marker, patient_id) %>%
      distinct() %>%
      group_by(marker) %>%
      slice_sample(n = max_patients)

    data_for_plot <- data_for_plot %>%
      semi_join(patient_markers_to_use, by = c("patient_id", "marker"))
  }

  if(log_transform) {
    scale_y <- scale_y_log10(scale_y_name)
    data_for_plot <- data_for_plot %>% 
      group_by(marker) %>%
      mutate(value = if_else(value == 0, min(value[value > 0]) / 2, value)) %>%
      ungroup()
  } else {
    scale_y <- scale_y_continuous(scale_y_name)
  }

  if(smooth) {
    main_geom <- geom_smooth(aes(group = outcome))
  } else {
    main_geom <- geom_line(alpha = 0.5)
  }

  if(!is.null(max_day)) {
    expand <- expand_limits(x = max_day)
  } else {
    expand <- NULL
  }

  data_for_plot %>%
    inner_join(data_fpv$patient_data, by = c("hospital_id", "patient_id")) %>%
    mutate(outcome = fct_relevel(fct_collapse(outcome, Hospitalized = c("Hospitalized", "Transferred")),
            "Discharged", "Hospitalized", "Death"),
           worst_breathing_s =fct_recode(worst_breathing_s, `Ambient Air` = "AA")) %>%
    ggplot(aes(x = day, y = value,  group = patient_id)) + 
    main_geom +
    facet + 
    expand +
    #scale_color_brewer(type = "qual", palette = 2, guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + 
    scale_y +
    scale_x_continuous("Days since hospitalization") 
    #scale_color_viridis_d(guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + scale_y
}

max_day <- data$marker_data_wide %>% filter(!is.na(CRP)) %>% pull(day) %>% max()
hide_axis_theme <- theme(axis.text.x = element_blank(),
                         axis.line.x = element_blank(),
                         axis.title.x = element_blank(),
                         axis.ticks.x = element_blank())
hide_strip_theme <- theme(strip.background = element_blank(), strip.text = element_blank())
title_theme <- theme(axis.title.y = element_text(size = 10))

fig_markers <- ((plot_marker_data(c("CRP" = "CRP"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme) /
(plot_marker_data(c("d_dimer" = "D-dimer"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme + hide_strip_theme) /
(plot_marker_data(c("lymphocyte_count" = "Ly"), log_transform = TRUE, max_day = max_day) + hide_axis_theme + title_theme + hide_strip_theme) /
(plot_marker_data(c("IL_6" = "IL-6"), log_transform = TRUE, max_day = max_day) + title_theme + hide_strip_theme))

fig_markers
```

```{r}
ggsave(here::here("local_temp_data", "markers_fpv.png"), fig_markers, width = 8, height = 4, dpi = 300)
```

### 2020-12-02 manuscript/manuscript.Rmd
supplementary tables added: treatment combinations and outcomes by hospital
More tweaks from feedback

@@
 abstract: |  
-  **Abstract:** We collected a multi-centric retrospective dataset of patients who were admitted to 10 hospitals in Czech Republic and tested positive for SARS-CoV-2.
+  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
@@
 
-The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths.
+The Covid-19 pandemic caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths.
 Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
@@
 Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3].
-Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
+Despite initial promise, hydroxychloroquine has failed to show benefit for hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
 Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
@@
 Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
-Even considering studies or case series with only several treated patients, very few are available [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B].
+Other studies are scarce - even when considering very small studies and case series [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B].
 None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
 
- For additional summary of treatments under investigation see [@http://zotero.org/users/5567156/items/YP62P6XJ].
+ For additional summary of treatments under investigation see Siemieniuk et al. [@http://zotero.org/users/5567156/items/YP62P6XJ]
 
@@
 ```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
+Each horizontal strip is a single patient, the ordering on the vertical axis is by disease severity, the split into two panels is for visual purposes only.
 Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation)."}
@@
 As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
-In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 3).
+In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 2).
 This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.
@@
 All but one of the patients receiving favipiravir also received hydroxychloroquine.
-Treatment was initiated mostly within two days of admission (see Supplementary Figure 2).
+Treatment was initiated mostly within two days of admission (see Supplementary Figure 3).
 
-The results of the multiverse analysis for association between hydroxochloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues.
+The results of the multiverse analysis for association between hydroxochloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues (see Supplementary statistical analysis for details).
 Results for all models we tested are reported in  supplementary Figures `r supplementary_figures_multiverse`, with additional details in supplementary statistical analysis.
@@
 
-## Prognostic models are not better than just using age
+## Published prognostic models are not better than using age as the sole predictor of outcome
 
@@
 The line ranges represent the bootstrapped 95% confidence intervals.
-Red dots show results computed in present study - individual dots vary in the outcome measured (12-day or 30-day mortality) and potentially on how ambiguities in score computation were resolved, although this rarely makes a big difference.
+Red dots show results computed in present study - model variants (horizontal axis) vary in the outcome measured (12-day or 30-day mortality, severe disease) and potentially on how ambiguities in score computation were resolved, although this rarely makes a big difference - see supplementary statistical analysis for details.
 Cyan triangles show AUC as reported by the original authors or recomputed based on their published data.

### 2020-12-02 manuscript/risk_scores.Rmd
Jm and risks for supplement
Version as sent for approval to co-authors
More tweaks from feedback

@@
 
-```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
+# Validating published prognostic models
+
+```{r setup_risk_scores, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
 knitr::opts_chunk$set(echo = FALSE)
@@
 
-# ACP
+## ACP
 
@@
 
-# Chen and Liu
+## Chen and Liu
 
@@
 
-# Shi et al.
+## Shi et al.
 
@@
 
-# Caramelo et al.
+## Caramelo et al.
 
@@
 
-# Bello-Chavolla et al.
+## Bello-Chavolla et al.
 
@@
 
-# Age only
+## Age only
 
@@
 
-# All together now
+## Summary
 
@@
 the point where reverting the score would yield better predictions.
-The thin gray dashed line marks the AUC from the original study for easier comparison."}
+The thin gray dashed line marks the AUC from the original study for easier comparison.", fig.width = 8, fig.height=6}
 
@@
 ```
-
-# References

### 2020-12-02 manuscript/supplementary_figures.Rmd
More tweaks from feedback

--- 2020-11-26 manuscript/supplementary_figures.Rmd
@@
   scale_x_date("Data collection period", date_breaks = "2 weeks")  + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  )
-```
-
-```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites.
-Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
-#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
-#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
-#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
-
-#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
-
-marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
-
-data$marker_data %>% 
-  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
-  mutate(marker = marker_map[marker]) %>%
-  group_by(marker, patient_id) %>% 
-  summarise(first_day = min(day), .groups = "drop") %>% 
-  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
 ```
@@
   m2<-models2[[i]]
-  lines(confint(m)[2,],rep(ii,2))
-  points(coef(m)[2],ii,pch=19)
-  lines(-confint(m2)[2,],rep(ii,2)-.3,col=col.hcq)
-  points(-coef(m2)[2],ii-.3,col=col.hcq,pch=19)
+  ci<-confint(m)[2,]
+  ci2<-confint(m2)[2,]
+  gtmp<-rbind(gtmp,data.frame(est=coef(m)[2],low=ci[1],high=ci[2],name=modelNames[i],type='outcome'))
+  gtmp<-rbind(gtmp,data.frame(est=-coef(m2)[2],low=-ci2[1],high=-ci2[2],name=modelNames[i],type='hcq'))
 }
-axis(2,length(modelNames):1,modelNames,las=2)
-legend('right',c('Death','No HCQ'),lty=1,col=c('black',col.hcq))
+ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
+  geom_vline(xintercept=0,color='blue')+
+  geom_point(position=position_dodge(width=0.5))+
+  geom_errorbarh(position=position_dodge(width=0.5),height= 0.5)+
+  scale_x_continuous('Log odds ratio')+
+  scale_y_discrete('')+
+  scale_color_manual('Model of:',values=c('black','red'),labels=c('Death','No HCQ'))+
+  theme(legend.background = element_rect(fill="gray95"),legend.position=c(.83,.7))
+```
+
+```{r initiation, fig.cap="Histogram of timing of first treatment relative to admission into one of the study sites.
+Two patients initiated treatment before admission, which is shown as the negative numbers.", fig.height=2}
+#data$marker_data %>% filter(marker == "favipiravir") %>% group_by(patient_id) %>% summarise(days = n(), .groups = "drop") %>% summarise(min_days = min(days), max_days = max(days))
+#data$marker_data %>% filter(marker == "az") %>% group_by(patient_id) %>% summarise(days = sum(!is.na(value) & value > 0), .groups = "drop") %>% filter(days > 0) %>% ggplot(aes(x = days)) + geom_histogram(binwidth = 1)
+#data$marker_data%>% filter(marker == "az") %>% group_by(value) %>% summarise(n())
+
+#Patients treated with convalescent plasma received 1-3 units of plasma over 1-2 days.
+
+marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
+
+data$marker_data %>% 
+  filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
+  mutate(marker = marker_map[marker]) %>%
+  group_by(marker, patient_id) %>% 
+  summarise(first_day = min(day), .groups = "drop") %>% 
+  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
 ```
@@
 ```
+
 ```{r allmultfpv, fig.height=11, fig.cap = "Estimates of model coefficients for association between favipiravir and main outcomes.
@@
 
-```{r multiverseresultsmarkers, fig.height=3.5, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
+```{r multiverseresultsmarkers, fig.height=5, fig.width=8.5, fig.cap="Estimates of model coefficients (log hazard ratios) for association between markers and death.
 The \"Suspicious\" section shows models that were found to not fit the data well or have computational issues, \"Problematic\" section shows models that were completely broken - see supplementary statistical analysis for details.

### 2020-12-02 R/auc_tools.R

@@ -103,10 +103,17 @@ compare_auc_to_orig <- function(orig_auc, our_auc, show_outcome = TRUE, ...) {
     data_for_plot_base <- all_auc
     all_orig <- all_auc %>% head(0) # Just to enforce correct columns
   }
-  data_for_plot <- data_for_plot_base %>%
+  data_for_plot <- data_for_plot_base  %>%
     mutate(
-      id = interaction(note, outcome, source),
            size = if_else(source == "orig", 2, 1))
+
+  if(show_outcome) {
+    data_for_plot <- data_for_plot %>% mutate(id = interaction(note, outcome, source))
+  } else {
+    data_for_plot <- data_for_plot %>% group_by(score, source) %>%
+      mutate(id = if_else(source == source_factor("orig"), "O", LETTERS[1:n()])) %>%
+      ungroup()
+  }
 
   res <- data_for_plot %>% ggplot(aes(y = auc, ymin = auc_low, ymax = auc_high, x = id, color = source, shape = source)) +
     geom_hline(yintercept = 0.5, color = "darkblue", size = 2) +
@@ -114,13 +121,14 @@ compare_auc_to_orig <- function(orig_auc, our_auc, show_outcome = TRUE, ...) {
     geom_errorbar(width = 0, na.rm = TRUE) + geom_point(aes(size = size)) +
     scale_size(range = c(2,4), guide = FALSE) +
     scale_y_continuous("AUC") +
+    scale_x_discrete("Score/outcome variant") +
     scale_shape_discrete("Source") +
     scale_color_discrete("Source") +
-    facet_wrap(~score, scales = "free_x", ...) +
-    theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.2))
+    facet_wrap(~score, scales = "free_x", ...)
 
-  if(!show_outcome) {
-    res <- res + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
+  if(show_outcome) {
+    res <- res +     theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.2))
+
   }
 
   res

### 2020-12-02 manuscript/supplementary_figures.Rmd

@@
   scale_x_date("Data collection period", date_breaks = "2 weeks")  + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  )
+```
+
+# Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
+```{r treatment_comb_table, echo=FALSE}
+pd<-data$patient_data
+x<-sort(table(paste0(
+ as.numeric(pd$ever_hcq),
+ as.numeric(pd$ever_az),
+ as.numeric(pd$ever_convalescent_plasma),
+ as.numeric(pd$ever_favipiravir),
+ as.numeric(pd$ever_remdesivir),
+ as.numeric(pd$ever_dexamethasone),
+ as.numeric(pd$ever_antibiotics))),dec=TRUE)
+x2<-c()
+nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
+for (i in 1:length(x)) {
+ x3<-c()
+ for (i2 in 1:nchar(names(x)[i])) {
+   x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
+ }
+ x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
+}
+rownames(x2)<-c()
+colnames(x2)<-c('N',nms)
+x2<-x2[,c(1,2,3,8,4:7)]
+knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
+```
+
+```{r outcome_by_hospital_prepare, echo=FALSE}
+tmp<-data$patient_data[,c('patient_id','hospital_id','outcome')]
+tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
+tmp$outcome<-dropLevels(tmp$outcome)
+tmp$hospital_id<-factor(tmp$hospital_id)
+levels(tmp$hospital_id)<-paste('H',1:length(levels(tmp$hospital_id)),sep='')
+```
+
+There were large differences between hospitals in the ratio of fatal outcomes 
+(P=`r round(fisher.test(with(tmp,table(outcome,hospital_id)),simulate.p.value=T)$p.value,5)`).
+```{r outcome_by_hospital, echo=FALSE}
+knitr::kable(with(tmp,table(outcome,hospital_id)),caption = "Overview of outcomes in different hospitals.
+Note that at H6, 9 out of 19 patients died, while at H9, no patient died out of 9 patients treated.", booktabs = TRUE)
 ```

### 2020-12-03 manuscript/manuscript.Rmd
Tweaks from KF, TS suggestions

@@
 csl: journal-of-medical-virology.csl
-date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
+#date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
 author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
@@
 Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
-Similarly, Lopinavir/Ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
+Similarly, lopinavir/ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
 No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
@@
 None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
-
 For additional summary of treatments under investigation see Siemieniuk et al. [@http://zotero.org/users/5567156/items/YP62P6XJ]
@@
 In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis, including treatments taken.
+Counts of all treatment combinations are shown in Supplementary Figure 2 and Supplementary Figure 3 shows outcomes by study site, demonstrating quite large hospital-specific differences.
 The dataset includes 19 patients already reported in a study of inflammatory signatures of Covid-19 [@http://zotero.org/users/5567156/items/IA4FQCF3].
@@
   scale_x_continuous("Days since hospitalization") +
+  # scale_color_manual("", values = c("black", "darkred"))+
   scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
@@
 As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
-In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 2).
+In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 4).
 This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.
@@
 
-first_fig_multiverse_drugs <- 4
-last_fig_multiverse_drugs <- 6
+first_fig_multiverse_drugs <- 6
+last_fig_multiverse_drugs <- 8
 supplementary_figure_markers <- last_fig_multiverse_drugs + 1 
@@
 All but one of the patients receiving favipiravir also received hydroxychloroquine.
-Treatment was initiated mostly within two days of admission (see Supplementary Figure 3).
+Treatment was initiated mostly within two days of admission (see Supplementary Figure 5).
 

### 2020-12-03 manuscript/supplementary_figures.Rmd
Tweaks from KF, TS suggestions

--- 2020-12-02 manuscript/supplementary_figures.Rmd
@@
 library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")
+library(UpSetR)
 
@@
   scale_x_date("Data collection period", date_breaks = "2 weeks")  + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  )
+```
+
+```{r upsettreatment, fig.cap = "Upset plot of treatment combinations - each vertical bar displays the number of patients that received the combination indicated by filled dots in the matrix.
+Horizontal bars show the total number of patients receiveing the given treatment.", fig.width=8, fig.height=6 }
+upset_data <- data$patient_data %>% select(starts_with("ever")) %>% mutate(across(starts_with("ever"), as.integer)) %>% as.data.frame()
+
+names(upset_data) <- gsub("ever_", "", names(upset_data)) %>% gsub("_", " ", .) %>% gsub("hcq", "hydroxychloroquine", .) %>% gsub("az", "azithromycin", .)
+
+names(upset_data) <- paste0(toupper(substring(names(upset_data),1,1)), substring(names(upset_data),2))
+
+UpSetR::upset(upset_data, sets = names(upset_data), text.scale = 2)
+```
+
+```{r outcome_by_hospital_prepare, echo=FALSE}
+tmp<-data$patient_data[,c('patient_id','hospital_id','outcome')]
+tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
+tmp$outcome<-dropLevels(tmp$outcome)
+tmp$hospital_id<-factor(tmp$hospital_id)
+levels(tmp$hospital_id)<-paste('H',1:length(levels(tmp$hospital_id)),sep='')
+```
+
+```{r outcomebyhospital, echo=FALSE}
+# knitr::kable(with(tmp,table(outcome,hospital_id)),caption = paste0( "Overview of outcomes in different hospitals.
+Note that at H6, 9 out of 19 patients died, while at H9, no patient died out of 9 patients treated.
+There were large differences between hospitals in the ratio of fatal outcomes (P=", round(fisher.test(with(tmp,table(outcome,hospital_id)),simulate.p.value=T)$p.value,5), ")"), booktabs = TRUE)
+```
+
+```{r hospitaloutcomes,fig.cap="Number of patients and outcomes at the individual sites.
+The numbers above bars are the exact counts.
+Hospitalized = still hospitalized at the end of data collection at the site.
+Sites are anonymized to preserve patient privacy."}
+data$patient_data %>% mutate(
+  hospital_label = factor(paste0("H", as.integer(factor(hospital_id))), levels = paste0("H", 1:10)),
+  outcome = fct_collapse(outcome, Hospitalized = c("Hospitalized", "Transferred")) %>% fct_relevel("Death", "Hospitalized", "Discharged")
+) %>%
+  ggplot(aes(x = hospital_label,  fill = outcome)) + geom_bar(position = position_dodge()) + 
+  geom_text(stat='count', aes(label=..count..), vjust=-1, position = position_dodge(width = 1)) +
+  #scale_fill_viridis_d("Outcome") +
+  scale_fill_manual("Outcome", values = c(Death = as.character(bfc()["Black"]), Hospitalized = as.character(bfc()["Bluish green"]), Discharged = as.character(bfc()["Yellow"]))) +
+  scale_x_discrete("Study site") +
+  scale_y_continuous("Count") +
+  expand_limits(y = 36)
 ```
@@
 
-marker_map <- c("az" = "Azithromycin", "hcq" = "Hydroxychloroquine", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
+marker_map <- c("hcq" = "Hydroxychloroquine", "az" = "Azithromycin", "favipiravir" = "Favipiravir", "convalescent_plasma" = "Conv. plasma")
 
@@
   filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
-  mutate(marker = marker_map[marker]) %>%
+  mutate(marker = factor(marker_map[marker], levels = marker_map)) %>%
   group_by(marker, patient_id) %>% 
   summarise(first_day = min(day), .groups = "drop") %>% 
-  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
+  ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free_y", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
 ```

### 2020-12-03 manuscript/manuscript.Rmd
Tweaks VK, formatting fixes for the supplements

@@
 
-We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
+We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow rate, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
 Full protocol for data collection is attached in the supplementary material.
@@
 Not all patients developed pneumonia or other symptoms of Covid-19.
-All patients received the standard of care including supplemental oxygen and ventilation and antibiotics for bacterial superinfections as determined by the attending physician.
+All patients received the standard of care which could include supplemental oxygen and ventilation and antibiotics for bacterial superinfections, as determined by the attending physician.
 Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition.

### 2020-12-03 manuscript/risk_scores.Rmd
Tweaks VK, formatting fixes for the supplements

@@
   transmute(score = shi, `Shi et al.
-(admission)` = shi_severe_admission, `Ventilated (admission)` = 1 - our_not_ventilated_admission, `Oxygen (admission)` = 1 - our_not_oxygen_admission, N = n_our_admission)
+(admission)` = shi_severe_admission, `Ventilated (admission)` = 1 - our_not_ventilated_admission, `Oxygen (admission)` = 1 - our_not_oxygen_admission, N = n_our_admission) %>% knitr::kable(booktabs = TRUE)
 
@@
   transmute(score = shi, `Shi et al.
-(hospitalization)` = shi_severe_all, `Ventilated (hospitalization)` = 1 - our_not_ventilated_all, `Oxygen (hospitalization)` = 1 - our_not_oxygen_all, N = n_our_all)
+(hospitalization)` = shi_severe_all, `Ventilated (hospitalization)` = 1 - our_not_ventilated_all, `Oxygen (hospitalization)` = 1 - our_not_oxygen_all, N = n_our_all) %>% knitr::kable(booktabs = TRUE)
 
@@
 )
-age_auc
+age_auc %>% knitr::kable(booktabs = TRUE)
 
@@
 
-We see that the "age only" baseline performs similarly to the best performing score (Caramelo et al.) all of the other scores - there is some uncertainty prohibiting us to be very sure about which works better.
+We see that the "age only" baseline performs similarly to the best performing score (Caramelo et al.) and likely better than all of the other scores - there is however some uncertainty prohibiting us to be very sure about the ordering.
 This holds even if we only take the decade of age.

### 2020-12-03 manuscript/manuscript.Rmd
Data availability in manuscript for medrxiv

@@
 
+# Data availability
+
+The data that support the findings of this study are available on request from the corresponding author, if legal obligations under GDPR can be met.
+The data are not publicly available due to privacy restrictions.
+We will be happy to arrange to run any analytical code locally and share the results, provided the code and the results do not leak personal information.
+
 # Acknowledgements

### 2020-12-18 manuscript/manuscript.Rmd
Version sent to co-authors to reapprove for PLoS

@@
 Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
-As of this writing, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58].
+As of this writing, the only treatment that exhibited positive effects in published randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], Janus kinase inhibitor baricitinib for hospitalized patients[@http://zotero.org/users/5567156/items/PDEPFSAU] and the monoclonal antibody bamlanivimab for non-hospitalized patients [@http://zotero.org/users/5567156/items/TZFUM6JH].
+Other monoclonal antibodies have also announced, but not yet published positive results of randomized trials, including tocilizumab[@http://zotero.org/users/5567156/items/4SLQXUJM] and combination of casirivimab and imdevimab[@http://zotero.org/users/5567156/items/VTR54LI9].
 Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA.
@@
 
-```{r treatment_comb_table, echo=FALSE}
-# Table \@ref(tab:treatment_comb_table} gives overview of combinations of treatments used in individual patients.
-# Currently not showing - almost all of this information is visible in the table above. Will put in supplement
-#pd<-data$patient_data
-# x<-sort(table(paste0(
-#   as.numeric(pd$ever_hcq),
-#   as.numeric(pd$ever_az),
-#   as.numeric(pd$ever_convalescent_plasma),
-#   as.numeric(pd$ever_favipiravir),
-#   as.numeric(pd$ever_remdesivir),
-#   as.numeric(pd$ever_dexamethasone),
-#   as.numeric(pd$ever_antibiotics))),dec=TRUE)
-# x2<-c()
-# nms<-c('HCQ','AZ','Plasma','Favipiravir','Remdesivir','Dexamethasone','ATB')
-# for (i in 1:length(x)) {
-#   x3<-c()
-#   for (i2 in 1:nchar(names(x)[i])) {
-#     x3<-c(x3,ifelse(substring(names(x)[i],i2,i2)=='1',nms[i2],''))
-#   }
-#   x2<-rbind(x2,cbind(n=x[i],rbind(x3)))
-# }
-# rownames(x2)<-c()
-# colnames(x2)<-c('N',nms)
-# x2<-x2[,c(1,2,3,8,4:7)]
-# knitr::kable(x2,caption = "Treatment combinations in individual patients.", booktabs = TRUE)
+```{r}
+median_ratios <-  data$marker_data %>% 
+  filter(marker %in% c("lymphocyte_count", "d_dimer", "CRP", "IL_6")) %>%
+  group_by(patient_id, marker) %>%
+  filter(n() > 2) %>%
+  summarise(ratio = max(value) / min(value), .groups = "drop") %>%
+  group_by(marker) %>%
+  summarise(med_ratio = median(ratio), .groups = "drop") %>%
+  column_to_rownames("marker") %>% as.matrix() %>% round(1)
 ```
@@
 
+The between-patient variability is notable even across outcomes - when ordering the patients by the highest CRP levels experienced throughout the hospital stay, the top 20% of patients that breathed ambient air for the whole hospitalization experienced higher levels than the bottom 20% of patients that required ventilation or died.
+This overlap is even larger when comparing only against the patients that died and D-dimer, Interleukin-6 and lymphocyte count also show a notably larger overlap than CRP (Supplementary Figure 4).
+
 ```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required.
-Each horizontal strip is a single patient, the ordering on the vertical axis is by disease severity, the split into two panels is for visual purposes only.
+Each vertical strip is a single patient, the ordering on the horizontal axis is by disease severity.
 Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation)."}
@@
 fig_progression <- data_for_fig_progression %>% 
+  mutate(patient_id = fct_reorder(patient_id, breathing_s,  .fun = my_paste, .desc = TRUE)) %>%
   filter(!is_outcome) %>%
-  ggplot(aes(x = day, y = patient_id, fill = breathing_s)) + 
-  geom_tile(height = 0.7) + 
+  ggplot(aes(x = patient_id, y = day, fill = breathing_s)) + 
+  geom_tile(width = 0.7) + 
   geom_point(data = data_for_fig_progression %>% filter(is_outcome) %>% mutate(outcome = factor(breathing_s, ordered = FALSE)),
-             inherit.aes = FALSE, aes(x = day, y = patient_id, shape = outcome), size = 0.9) +
+             inherit.aes = FALSE, aes(x = patient_id, y = day, shape = outcome), size = 0.9) +
   scale_fill_viridis_d("", direction = -1) +
-  scale_x_continuous("Days since hospitalization") +
+  scale_y_continuous("Days since hospitalization") +
   # scale_color_manual("", values = c("black", "darkred"))+
   scale_shape_manual("", values = c(1, 15), guide = guide_legend(override.aes = list(size = 4))) +
-  facet_wrap(~first_half, scales = "free_y") +
-  theme(axis.text.y = element_blank(), strip.background = element_blank(), strip.text = element_blank(),
+  theme(axis.text.x = element_blank(), strip.background = element_blank(), strip.text = element_blank(),
         legend.direction = "horizontal", legend.position = "bottom") +
-  theme(axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())
+  theme(axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.line.x = element_blank())
 
@@
 As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
-In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Supplementary Figure 4).
+In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Figure&nbsp;\@ref(fig:hcqassoc)).
 This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.
+
+```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with hydroxychloroquine seemed to be associated with the status of patients upon hospitalization.
+Comorbidities were associated with both worse outcome (black) and lower chance of treatment with hydroxychloroquine (red).
+Dots and lines represent the estimates and the 95% confidence intervals of the log odds ratio of the respective outcome.
+HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
+
+tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
+  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
+  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
+tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
+tmp$outcome<-dropLevels(tmp$outcome)
+tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
+tmp2$outcome<-dropLevels(tmp2$outcome)
+tmp2$outcome<-tmp2$outcome=='Death'
+
+m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
+m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
+m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
+m.copd<-glm(outcome~COPD,tmp2,family='binomial')
+m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
+m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
+m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
+m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
+m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
+m.s<-glm(outcome~smoking,tmp2,family='binomial')
+m.ob<-glm(outcome~obesity,tmp2,family='binomial')
+m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
+m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
+m.la<-glm(outcome~low_albumin,tmp2,family='binomial')
+
+m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
+m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
+m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
+m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
+m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
+m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
+m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
+m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
+m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
+m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
+m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
+m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
+m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
+m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')
+
+models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
+models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
+modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
+gtmp<-c()
+for (i in rev(seq(along=models))) {
+  ii<-length(models)-i+1
+  m<-models[[i]]
+  m2<-models2[[i]]
+  ci<-confint(m)[2,]
+  ci2<-confint(m2)[2,]
+  gtmp<-rbind(gtmp,data.frame(est=coef(m)[2],low=ci[1],high=ci[2],name=modelNames[i],type='outcome'))
+  gtmp<-rbind(gtmp,data.frame(est=-coef(m2)[2],low=-ci2[1],high=-ci2[2],name=modelNames[i],type='hcq'))
+}
+ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
+  geom_vline(xintercept=0,color='blue')+
+  geom_point(position=position_dodge(width=0.5))+
+  geom_errorbarh(position=position_dodge(width=0.5),height= 0.5)+
+  scale_x_continuous('Log odds ratio')+
+  scale_y_discrete('')+
+  scale_color_manual('Model of:',values=c('black','red'),labels=c('Death','No HCQ'))+
+  theme(legend.background = element_rect(fill="gray95"),legend.position=c(.83,.7))
+```
 

### 2020-12-18 manuscript/supplementary_figures.Rmd
Tweaks VK, formatting fixes for the supplements
Version sent to co-authors to reapprove for PLoS

@@
 ```{r upsettreatment, fig.cap = "Upset plot of treatment combinations - each vertical bar displays the number of patients that received the combination indicated by filled dots in the matrix.
-Horizontal bars show the total number of patients receiveing the given treatment.", fig.width=8, fig.height=6 }
+Horizontal bars show the total number of patients receiveing the given treatment.", fig.width=8, fig.height=5.8 }
 upset_data <- data$patient_data %>% select(starts_with("ever")) %>% mutate(across(starts_with("ever"), as.integer)) %>% as.data.frame()
@@
 The numbers above bars are the exact counts.
-Hospitalized = still hospitalized at the end of data collection at the site.
-Sites are anonymized to preserve patient privacy."}
+Hospitalized = still hospitalized at the end of data collection at the site or transferred to other site and lost to followup.
+Sites are anonymized to preserve patient privacy.", fig.height=3.5}
 data$patient_data %>% mutate(
@@
 
-```{r hcqassoc, message=FALSE, fig.width=7, fig.height=4.5, fig.cap="The choice of treatment with hydroxychloroquine seemed to be associated with the status of patients upon hospitalization.
-Comorbidities were associated with both worse outcome (black) and lower chance of treatment with hydroxychloroquine (red).
-Dots and lines represent the estimates and the 95% confidence intervals of the log odds ratio of the respective outcome.
-HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
+```{r markeroverlaps, fig.cap = "Density plots of worst marker values per patient, stratified by worst condition experienced by the patient.", fig.height=6}
+frac <- 0.2
+data$marker_data %>% filter(marker %in% c("lymphocyte_count", "d_dimer", "CRP", "IL_6")) %>%
+  inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>%
+  #filter(worst_breathing %in% c("AA", "Death")) %>%
+  group_by(marker, patient_id, worst_breathing_s) %>%
+  summarise(worst_value = case_when(marker == "lymphocyte_count" ~ min(log10(value)),
+                                    marker == "IL_6" ~ max(log10(value)),
+                                    TRUE ~ max(value)),
+            .groups = "drop") %>% 
+  filter(!is.infinite(worst_value)) %>%
+  mutate(marker = factor(marker, levels = c("lymphocyte_count", "d_dimer", "CRP", "IL_6"),
+         labels = c("Lowest Log10(lymphocyte count)", "Highest D-dimer", "Highest CRP", "Highest Log10(IL_6)"))) %>%
+  ggplot(aes(x = worst_value, color = worst_breathing_s, fill = worst_breathing_s, group = worst_breathing_s)) +
+  geom_density(adjust = 2, alpha = 0.3) +
+  scale_color_viridis_d("Worst condition") +
+  scale_fill_viridis_d("Worst condition") +
+  scale_x_continuous("Marker value") +
+  facet_wrap(~marker, scales = "free", ncol = 1)
 
-tmp<-data$patient_data[,c('age','sex','days_from_symptom_onset','admitted_for_covid','ischemic_heart_disease','has_hypertension_drugs','heart_failure',
-  'COPD','asthma','other_lung_disease','diabetes','renal_disease','liver_disease','smoking','BMI','obesity','last_record','high_creatinin','high_pt_inr','low_albumin',
-  'heart_problems','outcome','ever_hcq','ever_az','worst_breathing','had_invasive'),]
-tmp$outcome[tmp$outcome=='Transferred']<-'Hospitalized'
-tmp$outcome<-dropLevels(tmp$outcome)
-tmp2<-tmp[tmp$outcome%in%c('Death','Discharged'),]
-tmp2$outcome<-dropLevels(tmp2$outcome)
-tmp2$outcome<-tmp2$outcome=='Death'
+#Validating the claim about overlapping proportions
+# frac <- 0.2
+# data$marker_data %>% filter(marker %in% c("lymphocyte_count", "d_dimer", "CRP")) %>%
+#   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>%
+#   filter(worst_breathing_s != "Oxygen") %>%
+#   mutate(worst_breathing_s = fct_collapse(worst_breathing_s, "Ventilated or Death" = c("Ventilated", "Death"))) %>%
+#   group_by(marker, patient_id, worst_breathing_s) %>%
+#   summarise(min_value = min(value), max_value = max(value), .groups = "drop") %>%
+#   group_by(marker, worst_breathing_s) %>%
+#   summarise(qlowmin = quantile(min_value, frac), qhigh_min = quantile(min_value, 1 - frac), qlow_max = quantile(max_value, frac), qhigh_max = quantile(max_value, 1 - frac), .groups = "drop")
 
-m.ihd<-glm(outcome~ischemic_heart_disease,tmp2,family='binomial')
-m.hd<-glm(outcome~has_hypertension_drugs,tmp2,family='binomial')
-m.hf<-glm(outcome~heart_failure,tmp2,family='binomial')
-m.copd<-glm(outcome~COPD,tmp2,family='binomial')
-m.asthma<-glm(outcome~asthma,tmp2,family='binomial')
-m.old<-glm(outcome~other_lung_disease,tmp2,family='binomial')
-m.dia<-glm(outcome~diabetes,tmp2,family='binomial')
-m.rd<-glm(outcome~renal_disease,tmp2,family='binomial')
-m.ld<-glm(outcome~liver_disease,tmp2,family='binomial')
-m.s<-glm(outcome~smoking,tmp2,family='binomial')
-m.ob<-glm(outcome~obesity,tmp2,family='binomial')
-m.hc<-glm(outcome~high_creatinin,tmp2,family='binomial')
-m.hi<-glm(outcome~high_pt_inr,tmp2,family='binomial')
-m.la<-glm(outcome~low_albumin,tmp2,family='binomial')
-
-m2.ihd<-glm(ever_hcq~ischemic_heart_disease,tmp2,family='binomial')
-m2.hd<-glm(ever_hcq~has_hypertension_drugs,tmp2,family='binomial')
-m2.hf<-glm(ever_hcq~heart_failure,tmp2,family='binomial')
-m2.copd<-glm(ever_hcq~COPD,tmp2,family='binomial')
-m2.asthma<-glm(ever_hcq~asthma,tmp2,family='binomial')
-m2.old<-glm(ever_hcq~other_lung_disease,tmp2,family='binomial')
-m2.dia<-glm(ever_hcq~diabetes,tmp2,family='binomial')
-m2.rd<-glm(ever_hcq~renal_disease,tmp2,family='binomial')
-m2.ld<-glm(ever_hcq~liver_disease,tmp2,family='binomial')
-m2.s<-glm(ever_hcq~smoking,tmp2,family='binomial')
-m2.ob<-glm(outcome~obesity,tmp2,family='binomial')
-m2.hc<-glm(ever_hcq~high_creatinin,tmp2,family='binomial')
-m2.hi<-glm(ever_hcq~high_pt_inr,tmp2,family='binomial')
-m2.la<-glm(ever_hcq~low_albumin,tmp2,family='binomial')
-
-models<-list(m.ihd,m.hd,m.hf,m.copd,m.asthma,m.old,m.dia,m.rd,m.ld,m.s,m.ob,m.hc,m.hi,m.la)
-models2<-list(m2.ihd,m2.hd,m2.hf,m2.copd,m2.asthma,m2.old,m2.dia,m2.rd,m2.ld,m2.s,m2.ob,m2.hc,m2.hi,m2.la)
-modelNames<-c('IHD','HD','HF','COPD','Asthma','LungD','Dia','RenalD','LiverD','Smoking','Obesity','HighCr','HighInr','LowAlb')
-gtmp<-c()
-for (i in rev(seq(along=models))) {
-  ii<-length(models)-i+1
-  m<-models[[i]]
-  m2<-models2[[i]]
-  ci<-confint(m)[2,]
-  ci2<-confint(m2)[2,]
-  gtmp<-rbind(gtmp,data.frame(est=coef(m)[2],low=ci[1],high=ci[2],name=modelNames[i],type='outcome'))
-  gtmp<-rbind(gtmp,data.frame(est=-coef(m2)[2],low=-ci2[1],high=-ci2[2],name=modelNames[i],type='hcq'))
-}
-ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
-  geom_vline(xintercept=0,color='blue')+
-  geom_point(position=position_dodge(width=0.5))+
-  geom_errorbarh(position=position_dodge(width=0.5),height= 0.5)+
-  scale_x_continuous('Log odds ratio')+
-  scale_y_discrete('')+
-  scale_color_manual('Model of:',values=c('black','red'),labels=c('Death','No HCQ'))+
-  theme(legend.background = element_rect(fill="gray95"),legend.position=c(.83,.7))
+```
+```{r}
+# data$marker_data %>% 
+#   filter(marker %in% c("lymphocyte_count", "CRP", "d_dimer","IL_6"), value > 0, day >=0) %>%
+#   inner_join(data$breathing_data, by = c("hospital_id", "patient_id", "day")) %>%
+#   filter(breathing_s != "Death") %>%
+#   ggplot(aes(x = day, y = value, color = breathing_s, group = patient_id)) +
+#     geom_line(alpha = 0.8, position = position_jitter(width = 0.2, height = 0)) +
+#    # geom_point(, alpha = 0.5, position = position_jitter(width = 0.2, height = 0)) +
+#   scale_y_log10() + 
+#   scale_color_manual("Breathing", values = c(Ventilated = as.character(bfc()["Black"]), Oxygen = as.character(bfc()["Bluish green"]), AA = as.character(bfc()["Orange"]))) + #scale_color_brewer(type = "qual") +
+#   # scale_color_viridis_d(guide = guide_legend(override.aes = list(alpha = 1)), direction = -1, end = 0.9) +
+#   facet_wrap(~marker, scales = "free_y")
+#   
 ```

### 2020-12-22 manuscript/manuscript.Rmd
Applied minor tweks from feedback

@@
 None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
-For additional summary of treatments under investigation see Siemieniuk et al. [@http://zotero.org/users/5567156/items/YP62P6XJ]
+For additional summary of treatments under investigation see Siemieniuk et al.
+[@http://zotero.org/users/5567156/items/YP62P6XJ].
+There is also an emerging consensus on the importance of prophylactic anticoagulation therapy in managing hospitalized patients with Covid-19, although clinical trials to validate this approach are still ongoing[@http://zotero.org/users/5567156/items/5ZW5JM8R].
 

### 2020-12-22 manuscript/supplementary_figures.Rmd
Applied minor tweks from feedback

@@
     reference_docx: "output_template.docx"
-date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
+#date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
 ---
@@
 
-```{r markeroverlaps, fig.cap = "Density plots of worst marker values per patient, stratified by worst condition experienced by the patient.", fig.height=6}
+```{r markeroverlaps, fig.cap = "Density plots of worst marker values per patient, stratified by worst condition experienced by the patient.
+For each patient that had a given marker measured, the worst value was taken.
+Additionally the patients are classified by the worst condition (regardless of the timing relative to the worst marker levels).
+For each set of patients and marker an empirical density plot of the worst marker values is shown.", fig.height=6}
 frac <- 0.2

### 2020-12-22 R/data_preprocessing.R

@@ -38,10 +38,7 @@ read_data_for_analysis <- function() {
                                BMI <= 25 ~ "No",
                                BMI <= 30 ~ "Overweight",
                                TRUE ~ "Obese")
-                            ,levels = c("No", "Overweight", "Obese"), ordered = TRUE),
-           #TODO 8PGr47dGTx - quick fix for bug in preprocessing, incorrectly merged after transfer
-           last_record = if_else(patient_id == "8PGr47dGTx", last_record + 12L, last_record),
-           outcome = if_else(patient_id == "8PGr47dGTx", factor("Discharged", levels = levels(outcome)), outcome)
+                            ,levels = c("No", "Overweight", "Obese"), ordered = TRUE)
            )
 
   collapse_breathing <- function(x) {

### 2021-01-06 dev/risk_scores_consortium_tests.Rmd
Work on risk scores

---
title: "Collaboration on risk scores"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE)

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
```

Get marker counts

```{r}
relevant_markers <- c("lactate_dehydrogenase", "d_dimer", "platelet_count", "neutrophile_count", "lymphocyte_count", "AST")

subset_data <- data$marker_data %>% 
  filter(marker %in% relevant_markers & !is.na(value)) %>%
  group_by(marker, patient_id) %>% summarise(present = TRUE, .groups = "drop") %>%
  pivot_wider(names_from = "marker", values_from = "present") %>%
  mutate(across(all_of(relevant_markers), ~ if_else(is.na(.x), 0L, 1L)))

UpSetR::upset(subset_data %>% as.data.frame(), sets = relevant_markers)
```

No GFR (no Wang lab)

### 2021-01-06 other_outputs/risk_scores_consortium.Rmd
Work on risk scores

---
title: "Validating risk scores - Czech data"
output: html_notebook
---

# Prepare data

```{r setup}

devtools::load_all()
library(tidyverse)

library(mice)
library(pROC)
library(metamisc)

data <- read_data_for_analysis()
```

```{r}
first_non_NA <- function(x) {
  if(length(x) > 2) {
    stop("Wrong length")
  } else if(length(x) == 1) {
    x
  } else {
    if(is.na(x[1])) {
      x[2]
    } else {
      x[1]
    }
  }

}

marker_data_admission_all <- data$marker_data_wide %>% 
  filter(day == 0 | day == 1) %>%
  arrange(hospital_id, patient_id,  day) %>%
  group_by(patient_id, hospital_id) %>%
  summarise(across(everything(), first_non_NA), .groups = "drop")

marker_data_admission <- marker_data_admission_all %>%
  select(-ends_with("censored"),-breathing, -breathing_low, -breathing_high, -breathing_s_high, -breathing_s_low, -pcr_value, -hcq, -favipiravir, -convalescent_plasma, -remdesivir,-dexamethasone, -day) %>%
  select( -all_of(c(all_antibiotics, other_drugs)), -kaletra, -tocilizumab)
```

```{r}
data_risk <- data$patient_data %>%
  mutate(mortality  = case_when(
    last_record >= 30 ~ "alive",
    outcome == "Discharged" ~ "alive",
    outcome == "Death" ~ "dead",
    TRUE ~ "alive")) %>%
  inner_join(marker_data_admission, by = c("patient_id", "hospital_id")) %>%
  mutate(creatinin = if_else(is.na(creatinin.x), creatinin.y, creatinin.x),
         pt_inr = if_else(is.na(pt_inr.x), pt_inr.y, pt_inr.x),
         albumin = if_else(is.na(albumin.x), albumin.y, albumin.x),
         ) %>%
  select(-creatinin.x, -creatinin.y, -pt_inr.x, -pt_inr.y, -albumin.x, -albumin.y) %>%
  #avoid including some derived quantities
  select(-age_norm, -high_creatinin, -high_pt_inr, -low_albumin, -heart_problems, -lung_problems, -markers_out_of_range, -antibiotics, -macrolides) %>%
  #avoid including indicators that are not available at hospitalization 
  select(-discontinued_medication, -last_record, -outcome,  -best_supportive_care_from,
         -starts_with("ever_"), -starts_with("any_"), -starts_with("worst"), -had_invasive,
         -first_day_invasive, -last_day_invasive) %>%
  #avoid columns that are not useful
  select(-hospital_id, -patient_id, -pcr_positive, -FiO2, -urea,
         -took_convalescent_plasma, -obesity, -took_favipiravir, -took_az,
         -took_dexamethasone, -took_remdesivir, -IgG, -Horowitz_index,
         -has_hypertension_drugs
         ) %>%
  #convert to factors to use with mice
  mutate(sex = factor(sex), 
         breathing_s = factor(breathing_s, ordered = FALSE), 
         requires_oxygen = breathing_s != "AA",
         across(all_of(names(.)[sapply(., typeof) == "logical"]), factor))
```

```{r}
where_impute <- is.na(data_risk)
columns_to_impute <- c("lactate_dehydrogenase", 
                       "lymphocyte_count",
                       "SpO2",
                       "SpO2_native",
                       "CRP",
                       "d_dimer",
                       "COPD",
                       "other_lung_disease",
                       "has_hypertension_drugs",
                       "ischemic_heart_disease",
                       "heart_failure")

where_impute[, setdiff(names(data_risk), columns_to_impute)] <- FALSE

# predictor_matrix <- matrix(1, ncol = ncol(data_risk), 
#                            nrow = ncol(data_risk),  dimnames = list(paste0("b_", names(data_risk), names(data_risk))))

block_vars <- setdiff(names(data_risk), "mortality")
blocks <- make.blocks(block_vars)
predictor_matrix <- make.predictorMatrix(data_risk, blocks)

predictor_matrix[, "requires_oxygen"] <- 0
low_obs_count <- c("lactate_dehydrogenase", "AST", "ferritin", "IL_6", "leukocyte_count", "PEEP")
predictor_matrix[low_obs_count, ] <- 0
predictor_matrix[low_obs_count, c("age","sex")] <- 1

mid_obs_count <- c("neutrophile_count", "platelet_count", "procalcitonin")
predictor_matrix[mid_obs_count, ] <- 0
predictor_matrix[mid_obs_count, c("age","sex", "requires_oxygen", "admitted_for_covid", "comorbidities_sum", "BMI")] <- 1
predictor_matrix["procalcitonin", "admitted_for_covid"] <- 0

```

```{r}
setup_imp <- mice(data_risk, maxit=0, blocks = blocks, predictorMatrix = predictor_matrix)
setup_imp$loggedEvents
```

```{r}
setup_imp$method
```
```{r}
set.seed(22546824)
data_risk_imp <- mice(data_risk, method = setup_imp$method, m = 50, maxit = 10, blocks = blocks, predictorMatrix = predictor_matrix, printFlag = F )
```

```{r}
if(!is.null(data_risk_imp$loggedEvents)) {
  data_risk_imp$loggedEvents %>% select(-it, -im) %>% distinct() %>% arrange(dep)
}
```

# Define prediction models

## Hu

The units in our dataset are:

```{r}
data$marker_data %>% filter(marker %in% c("CRP", "d_dimer", "lymphocyte_count")) %>%
  select(marker, unit) %>% distinct()
```

So only need to conver D-dimer from ng/ml DDU to ug/ml FEU

```{r}
hu <- function(data) {
  lp <- data %>%
    mutate(d_dimer_converted = d_dimer * 2 / 1000,
           lp = -4.21167 + 0.058699 * age + 0.013391 * CRP + 0.112055 * d_dimer_converted - 1.983733 * lymphocyte_count) %>%
    pull(lp)

  lp
}
```

## Xie

```{r}
xie <- function(data) {
  lp <- data %>%
    mutate(lp = 4.559 + 0.047 * age + 0.003 * lactate_dehydrogenase -1.094 * lymphocyte_count - 0.098 * SpO2) %>%
    pull(lp)

  lp
}
```

## Wang clinicial

```{r}
wang_clinical <- function(data) {
  data %>%
    mutate(hypertension = n_hypertension_drugs > 0,
           lp =
  -2.6551 + 0.10 * age + 0.60 * hypertension + 1.11 * (ischemic_heart_disease == "TRUE")) %>%
    pull(lp)
}
```

## Age baseline

```{r}
age_baseline <- function(data) {
  data %>% mutate(age_norm = (age - 65) / 20 - 2) %>%
    pull(age_norm)
}
```

# Model validation

```{r}
validation_res <- list()
for(model in c("hu", "xie", "wang_clinical", "age_baseline")) {
  cat("\n===== Validating ", model, " ==============\n")
  single_res <- list()
  cat("- AUC\n")
  single_res$est_auc <- validate.mids(data = data_risk_imp,
                       model = model,
                       measure = auc,
                       quiet = F)

  cat("- Calibration slope\n")
  single_res$est_slope <- validate.mids(data = data_risk_imp,
                       model = model,
                       measure = calibration_slope)

  cat("- Calibration intercept\n")
  single_res$est_int <- validate.mids(data = data_risk_imp,
                       model = model,
                       measure = calibration_intercept)

  cat("- OE ratio\n")
  single_res$est_log_oe <- validate.mids(data = data_risk_imp,
                              model = model,
                              measure = oe_ratio,
                              log = TRUE)

  cat("- Net benefit\n")
  single_res$est_nb <- validate.mids(data = data_risk_imp,
                       model = model,
                       measure = net_benefit_bootstrap)

  validation_res[[model]] <- single_res
}
saveRDS(validation_res, file = here::here("local_temp_data", "validation_res.rds"))
```

```{r}
aucs <- list()
slopes <- list()
oes <- list()
nbs <- list()

for(model in names(validation_res)) {
  auc_pooled <- pool.auc.mids(validation_res[[model]]$est_auc)$pooled$auc
  print(auc_pooled)
  aucs[[model]] <- auc_pooled

  slope_pooled <- pool.validate.mids(validation_res[[model]]$est_slope)$pooled  
  print(slope_pooled)
  slopes[[model]] <- slope_pooled

  print(pool.validate.mids(validation_res[[model]]$est_int)$pooled)

  oe_pooled <- pool.oe.mids(validation_res[[model]]$est_log_oe)$oe$pooled 
  print(oe_pooled)
  oes[[model]] <- oe_pooled

  nb_pooled <- pool.validate.mids(validation_res[[model]]$est_nb)$pooled
  print(nb_pooled)
  nbs[[model]] <- nb_pooled
}
```

```{r}
plot_results <- function(results_list, title, estimate_column) {
  purrr::imap_dfr(results_list, ~ .x %>% mutate(model = .y)) %>% 
    ggplot(aes(x = model, y = {{estimate_column}}, ymin = ci.lb, ymax = ci.ub)) + geom_pointrange()  + scale_y_continuous(title)
}

plot_results(aucs, "AUC", est)
plot_results(slopes, "Calibration slope", estimate)
plot_results(oes, "OE", estimate)
plot_results(nbs, "Net benefit", estimate)

```

```{r}
```

Xie - not sure if using SpO2 or SpO2 native
DDU vs FEU D-dimer

### 2021-01-06 R/data_preprocessing.R

@@ -1,5 +1,5 @@
 read_data_for_analysis <- function() {
-  patient_data <- read_csv(here::here("private_data", "patient_data.csv"),
+  patient_data <- readr::read_csv(here::here("private_data", "patient_data.csv"),
                            col_types = cols(
                              hospital_id = col_character(),
                              patient_id = col_character(),
@@ -44,7 +44,7 @@ read_data_for_analysis <- function() {
   collapse_breathing <- function(x) {
     fct_collapse(x, Ventilated = c("NIPPV", "MV", "ECMO"))
   }
-  breathing_data <- read_csv(here::here("private_data", "breathing_data.csv"),
+  breathing_data <- readr::read_csv(here::here("private_data", "breathing_data.csv"),
                            col_types = cols(
                             patient_id = col_character(),
                             hospital_id = col_character(),
@@ -57,7 +57,7 @@ read_data_for_analysis <- function() {
            breathing_s_low = collapse_breathing(breathing_low),
            breathing_s_high = collapse_breathing(breathing_low))
 
-  marker_data <- read_csv(here::here("private_data", "marker_data.csv"),
+  marker_data <- readr::read_csv(here::here("private_data", "marker_data.csv"),
                              col_types = cols(
                               patient_id = col_character(),
                               marker = col_character(),

### 2021-01-06 R/risk_scores_tools.R

@@ -1,0 +1,180 @@
+inv_logit <- function(x) {  1/(1+exp(-x)) }
+
+calibration_intercept <- function(lp, y, family = binomial, ...) {
+  #lp <- log(p / (1 - p))
+  resp = y == "dead"
+  fit <- glm(resp ~ 1, family = family, offset = lp, ...)
+  variance <- diag(vcov(fit))
+  c(estimate = coef(fit), se = sqrt(variance), var = variance)
+}
+
+calibration_slope <- function(lp, y, family = binomial, ...) {
+  #lp <- log(p / (1 - p))
+  resp = y == "dead"
+  fit <- glm(resp ~ lp + 1, family = family, ...)
+  variance <- diag(vcov(fit))[2]
+  c(estimate = coef(fit)[2], se = sqrt(variance), var = variance)
+}
+
+auc <- function(lp, y, method = "delong", direction = "<", ...) {
+  fit <- pROC::auc(response = y, predictor = lp, direction = direction, method = method,  ...)
+  variance <- var(fit)
+  ci <- confint(fit)
+  c(estimate = fit[[1]][[1]], se = sqrt(variance), var = variance, ci.lb = ci$ci.lb, ci.ub = ci$ci.ub)
+}
+
+net_benefit <- function(lp, y, threshold = .50, test_harm = 0, case_label = "dead", ...) {
+  p = inv_logit(lp)
+  tp <- sum((p >= threshold) & (y == case_label))
+  fp <- sum((p >= threshold) & (y != case_label))
+  n <- length(p)
+
+  if (threshold == 1) { # by definition
+    return(0)
+  } else{return(tp/n - fp/n * threshold / (1-threshold) - test_harm)}
+}
+
+net_benefit_bootstrap <- function(lp, y, threshold = .50, test_harm = 0, case_label = "dead", I = 100, ...) {
+  p = inv_logit(lp)
+  nb <- rep(NA, I)
+  for (i in seq_len(I)) {
+    s <- sample.int(length(p), length(p), replace = TRUE)
+    nb[i] <- net_benefit(p[s], y[s], threshold, test_harm, case_label, ...)
+  }
+  se <- sd(nb)
+  est <- net_benefit(p, y, threshold, test_harm, case_label, ...)
+
+  c(estimate = est, se = se, var = se^2)
+}
+
+oe_ratio <- function(lp, y, case_label = "dead", log = TRUE, ...) {
[…]

### 2021-01-06 R/special_lab_data.R

@@ -36,6 +36,9 @@ special_lab_data_markers_to_collect <-
     "Ddim" = "d_dimer",
     "Ly#" = "lymphocyte_count",
     "LYa" = "lymphocyte_count",
+    "Ne#" = "neutrophile_count",
+    "NEa" = "neutrophile_count",
+    "PLT" = "platelet_count",
     "Kre" = "creatinin",
     "CRP" = "CRP",
     "PCT" = "procalcitonin",
@@ -43,6 +46,8 @@ special_lab_data_markers_to_collect <-
     "Ferr" = "ferritin",
     "IL-6" = "IL_6",
     "IL6" = "IL_6",
+    "AST" = "AST",
+    "LD" = "lactate_dehydrogenase",
     "Vyska" = "height",
     "Vaha" = "weight"
   )
@@ -52,6 +57,9 @@ special_lab_data_markers_to_collect_units <-
     "Ddim" = "ng/ml DDU",
     "Ly#" = "10^9/l",
     "LYa" = "10^9/l",
+    "Ne#" = "10^9/l",
+    "NEa" = "10^9/l",
+    "PLT" = "10^9/l",
     "Kre" = "\U03BCmol/l",
     "CRP" = "mg/l",
     "PCT" = "\U03BCg/l",
@@ -59,6 +67,8 @@ special_lab_data_markers_to_collect_units <-
     "Ferr" = "\U03BCg/l",
     "IL-6" = "ng/l",
     "IL6" = "ng/l",
+    "AST" = "\U03BCkat/l",
+    "LD" = "\U03BCkat/l",
     "Vyska" = "cm",
     "Vaha" = "kg"
   )
@@ -155,11 +165,11 @@ read_special_lab_data <- function(input_file, patient_id, hospital_id, first_day
             vals <- strsplit(marker_value_part[2], "; ", fixed = TRUE)[[1]]
 
             #Ignored values
-            vals <- vals[!grepl("^(Txt\\+Hisviz koment..|krev sra.en.)$", vals)]
+            vals <- vals[!grepl("^(Txt\\+Hisviz koment..|krev sra.en.|hemol.za|chyl.zn.|nelze)$", vals)]
             if(length(vals) == 0) {
               next
             }
-            value_pattern <- "^(<?-?[0-9]+,?[0-9]*|(Txt\\+His)?v.ce ne. [0-9]+ .*)$"
[…]

### 2021-01-11 other_outputs/risk_scores_consortium.Rmd
Updates for risk scores based on feedback by V.M.T.deJong

@@
   #avoid including some derived quantities
-  select(-age_norm, -high_creatinin, -high_pt_inr, -low_albumin, -heart_problems, -lung_problems, -markers_out_of_range, -antibiotics, -macrolides) %>%
-  #avoid including indicators that are not available at hospitalization 
-  select(-discontinued_medication, -last_record, -outcome,  -best_supportive_care_from,
-         -starts_with("ever_"), -starts_with("any_"), -starts_with("worst"), -had_invasive,
-         -first_day_invasive, -last_day_invasive) %>%
+  select(-age_norm, -high_creatinin, -high_pt_inr, -low_albumin, -heart_problems, -lung_problems, -markers_out_of_range, -antibiotics, -macrolides, -comorbidities_sum_na) %>%
   #avoid columns that are not useful
   select(-hospital_id, -patient_id, -pcr_positive, -FiO2, -urea,
+          -discontinued_medication, -last_record, -outcome,  
+         #-best_supportive_care_from,
+         -starts_with("ever_"), ever_hcq,
+         -starts_with("any_"), 
+         -starts_with("worst"), worst_breathing_s,
+         -had_invasive,
+         -first_day_invasive, -last_day_invasive,
+      #   -startsWith("took"),
+         -took_hcq,
          -took_convalescent_plasma, -obesity, -took_favipiravir, -took_az,
@@
          breathing_s = factor(breathing_s, ordered = FALSE), 
+         worst_breathing_s = factor(worst_breathing_s, ordered = FALSE), 
          requires_oxygen = breathing_s != "AA",
          across(all_of(names(.)[sapply(., typeof) == "logical"]), factor))
+```
+
+These are the final columns used:
+```{r}
+names(data_risk)
 ```
@@
 
+# I don't use mortality as it is already included in worst_breathing_s
 block_vars <- setdiff(names(data_risk), "mortality")
@@
 predictor_matrix[, "requires_oxygen"] <- 0
+predictor_matrix[, "mortality"] <- 0
+predictor_matrix["oxygen_flow", "worst_breathing_s"] <- 0
 low_obs_count <- c("lactate_dehydrogenase", "AST", "ferritin", "IL_6", "leukocyte_count", "PEEP")
@@
                        model = model,
-                       measure = auc,
+                       measure = covid19retrospective::auc,
                        quiet = F)
@@
                        model = model,
-                       measure = calibration_intercept)
+                       measure = calibration_intercept,
+                       use_logistf = (model == "hu"))
 

### 2021-01-25 other_outputs/risk_scores_consortium.Rmd
Another update to risk scores

@@
 data <- read_data_for_analysis()
+
+columns_to_impute <- c("lactate_dehydrogenase", 
+                       "lymphocyte_count",
+                       "SpO2",
+                       "SpO2_native",
+                       "CRP",
+                       "d_dimer",
+                       "COPD",
+                       "other_lung_disease",
+                       "has_hypertension_drugs",
+                       "ischemic_heart_disease",
+                       "heart_failure")
 ```
@@
 first_non_NA <- function(x) {
-  if(length(x) > 2) {
-    stop("Wrong length")
-  } else if(length(x) == 1) {
-    x
+  if(all(is.na(x))) {
+    NA
   } else {
-    if(is.na(x[1])) {
-      x[2]
-    } else {
-      x[1]
+    ind <- which(!is.na(x))[1]
+    x[ind]
   }
-  }
-
 }
@@
   summarise(across(everything(), first_non_NA), .groups = "drop")
+
+marker_data_2_5_to_impute <- data$marker_data_wide %>%
+  select(hospital_id, patient_id,  day, any_of(columns_to_impute)) %>%
+  filter(day >= 2 & day <= 5) %>%
+  arrange(hospital_id, patient_id,  day) %>%
+  group_by(patient_id, hospital_id) %>%
+  summarise(across(everything(), first_non_NA), .groups = "drop") %>%
+  select(-day)
+
+columns_2_5 <- setdiff(names(marker_data_2_5_to_impute), c("day", "hospital_id", "patient_id"))
 
@@
          ) %>%
-  select(-creatinin.x, -creatinin.y, -pt_inr.x, -pt_inr.y, -albumin.x, -albumin.y) %>%
+  select(-creatinin.x, -creatinin.y, -pt_inr.x, -pt_inr.y, -albumin.x, -albumin.y)
+
+# Add data 2 to 5
+data_risk <- data_risk %>%
+  inner_join(marker_data_2_5_to_impute, by = c("patient_id", "hospital_id"),
+           suffix = c("", "_2_5"))
+
   #avoid including some derived quantities
+data_risk <- data_risk %>%
   select(-age_norm, -high_creatinin, -high_pt_inr, -low_albumin, -heart_problems, -lung_problems, -markers_out_of_range, -antibiotics, -macrolides, -comorbidities_sum_na) %>%
@@
   select(-hospital_id, -patient_id, -pcr_positive, -FiO2, -urea,
-          -discontinued_medication, -last_record, -outcome,  
+          -discontinued_medication, -outcome,  
          #-best_supportive_care_from,
@@
          -first_day_invasive, -last_day_invasive,
+         -first_wave,
       #   -startsWith("took"),
@@
          across(all_of(names(.)[sapply(., typeof) == "logical"]), factor))
+
+# Nelson-Aalen for outcome
+data_risk <- data_risk %>% 
+  mutate(mortality_l = mortality == "dead") %>%
+  mutate(
+         nelsonaalen = mice::nelsonaalen(., timevar = "last_record", statusvar = "mortality_l")) %>%
+  select(-last_record, -mortality_l)
 ```
@@
 where_impute <- is.na(data_risk)
-columns_to_impute <- c("lactate_dehydrogenase", 
-                       "lymphocyte_count",
-                       "SpO2",
-                       "SpO2_native",
-                       "CRP",
-                       "d_dimer",
-                       "COPD",
-                       "other_lung_disease",
-                       "has_hypertension_drugs",
-                       "ischemic_heart_disease",
-                       "heart_failure")
 
@@
 predictor_matrix[low_obs_count, c("age","sex")] <- 1
+predictor_matrix["lactate_dehydrogenase", "sex"] <- 0
 
@@
 predictor_matrix["procalcitonin", "admitted_for_covid"] <- 0
+
+no_nelsonaalen <- c("best_supportive_care_from")
+predictor_matrix[no_nelsonaalen, "nelsonaalen"] <- 0
+
+# Use future values only for the corresponding predictors
+for(c in columns_2_5) {
+  c_2_5 <- paste0(c, "_2_5")
+  # The same predictor as for the original column
+  predictor_matrix[c_2_5, ] <- predictor_matrix[c, ] 
+  predictor_matrix[, c_2_5] <- 0
+
+  predictor_matrix[c, paste0(c, "_2_5")] <- 1 
+}
 
@@
 age_baseline <- function(data) {
-  data %>% mutate(age_norm = (age - 65) / 20 - 2) %>%
-    pull(age_norm)
+  # Comparison of the characteristics, morbidity, and mortality of COVID-19 and seasonal influenza: a nationwide, population-based retrospective cohort study
+  # DOI: 10.1016/S2213-2600(20)30527-0
+  # https://www.thelancet.com/journals/lanres/article/PIIS2213-2600(2030527-0/fulltext#sec1
+  # Supplementary Table 3
+  data %>% mutate(ifr_perc = case_when(
+    age < 18 ~ 0.7,
+    age <= 30 ~ 0.5,
+    age <= 40 ~ 1.4,
+    age <= 50 ~ 2.7,
+    age <= 60 ~ 6.1,
+    age <= 70 ~ 13.7,
+    age <= 80 ~ 23.1,
+    age <= 90 ~ 32.4,
+    TRUE ~ 36.3),
+    ifr = ifr_perc / 100,
+    lp = log(ifr) - log1p(-ifr)) %>%
+    pull(lp) 
 }

### 2021-04-06 other_outputs/risk_scores_consortium.Rmd
Updating Wang model, summary

@@
            lp =
-  -2.6551 + 0.10 * age + 0.60 * hypertension + 1.11 * (ischemic_heart_disease == "TRUE")) %>%
+  -8.6 + 0.10 * age + 0.60 * hypertension + 1.11 * (ischemic_heart_disease == "TRUE")) %>%
     pull(lp)
@@
 
+# Summaries
+
 ```{r}
+data_risk_sum <- data_risk %>% mutate(sex_male = sex == "M",
+                                      hypertension = n_hypertension_drugs > 0,
+                                      deceased = mortality == "dead")
+
+binary_columns <- c(
+ "sex_male",
+ "hypertension",
+ "heart_failure",
+ "COPD",
+ "asthma",
+ "other_lung_disease",
+ "diabetes",
+ "renal_disease",
+ "liver_disease",
+ "smoking",                  
+ "admitted_for_covid",
+ "ischemic_heart_disease",
+ "deceased"
+)
+
+continuous_columns <- c(
+ "age" ,
+ "days_from_symptom_onset",
+ "BMI",                
+ "AST",
+ "CRP",
+ "lymphocyte_count",
+ "neutrophile_count",
+ "platelet_count",
+ "d_dimer",
+ "ferritin",
+ "IL_6",
+ "lactate_dehydrogenase",
+ "procalcitonin",
+ "leukocyte_count",
+ "creatinin",
+ "pt_inr", 
+ "albumin"
+)
+
+summary_binary_list <- list()
+
+for(col in binary_columns) {
+  vals <- data_risk_sum[[col]]
+  if(is.factor(vals)) {
+    if(identical(levels(vals), c("FALSE", "TRUE"))) {
+      vals <- vals == "TRUE"
+    } else {
+      stop("Factor with invalid levels")
+    }
+  }
+  if(!is.logical(vals)) {
+    stop(col, " is not logical")
+  }
+  summary_binary_list[[col]] <- data.frame(
+    name = col,
+    type = "binary",
+    n_missing = sum(is.na(vals)),
+    mean = mean(vals, na.rm = TRUE)
+  )
+}
+
+summary_binary <- do.call(rbind, summary_binary_list)
+
+summary_continuous_list <- list()
+for(col in continuous_columns) {
+  vals <- data_risk_sum[[col]]
+  if(!is.numeric(vals)) {
+    stop(col, " is not numeric")
+  }
+  summary_continuous_list[[col]] <- data.frame(
+    name = col,
+    type = "continuous",
+    n_missing = sum(is.na(vals)),
+    mean = mean(vals, na.rm = TRUE)
+  )
+}
+
+summary_continuous <- do.call(rbind, summary_continuous_list)
+
+summary_all <- rbind(summary_binary, summary_continuous)
+summary_all
+
+write_csv(summary_all, path = here::here("local_temp_data", "table_1_summary.csv"))
 ```
-
-Xie - not sure if using SpO2 or SpO2 native
-DDU vs FEU D-dimer

### 2021-06-03 other_outputs/fpv_metaanalysis.Rmd
Computing data for FPV metaanalysis

---
title: "Favipiravir - data for metaanalysis"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8)

devtools::load_all()
library(tidyverse)
library(patchwork)
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide
```

```{r}
fpv_hospital_id <- wide %>% filter(took_favipiravir) %>% pull(hospital_id) %>% unique()

non_censored_patients <- data$patient_data %>% filter(outcome %in% c("Death", "Discharged"))
data_fpv <- filter_complete_data(data, 
                                 hospital_id == fpv_hospital_id & 
                                 patient_id %in% non_censored_patients$patient_id)
wide_fpv <- data_fpv$marker_data_wide
```

```{r}
data_fpv$marker_data %>% group_by(marker, patient_id) %>%
  summarise(n_days = n(), units = unique(unit)) %>%
  summarise(n_patients = n(), median_days = median(n_days), units = unique(units)) %>%
  arrange(desc(n_patients))
```

```{r}
markers_for_adverse_effects <- tribble(
  ~marker,            ~low_bound,      ~high_bound,
  "creatinin",        49,              90,
 # "CRP",
  "platelet_count",   150,             450,
  #"AST"
  "eosinophile_count",0,               0.5,
  "lymphocyte_count", 0.8,             4,
  "neutrophile_count",2,               7,
 # "albumin",          36,              45, # Removed, too few events/at risk
  #"pt_inr",                                # Removed - hard to put good bounds 
 # "procalcitonin",    0,               0.5 # Removed, too few events/at risk
)
```

```{r}
fpv_patients <- wide_fpv %>% 
  group_by(patient_id) %>%
  filter(any(took_favipiravir)) %>%
  summarise(min_day_fpv = min(day[took_favipiravir]), max_day_fpv = max(day[!is.na(favipiravir) & favipiravir > 0])) %>% 
  inner_join(data_fpv$patient_data, by = "patient_id") %>%
  select(patient_id, age, sex, min_day_fpv, max_day_fpv) %>%
  mutate(match_group = 1)

fpv_patients
```
```{r}
patients_matching <- data_fpv$patient_data %>%
  filter(!ever_favipiravir) %>%
  inner_join(data_fpv$marker_data, by = "patient_id") %>%
  group_by(patient_id, age, sex) %>%
  summarise(days_total.control = max(day), .groups = "drop") %>%
  mutate(match_group = 1)

potential_matches <- patients_matching %>%
  inner_join(fpv_patients, by = c("match_group"), suffix = c(".control", ".fpv")) %>%
  filter(days_total.control > min_day_fpv, abs(age.control - age.fpv) < 15)
```

```{r}
set.seed(48632485)
matches <- potential_matches %>%
  group_by(patient_id.control) %>%
#  filter(patient_id.fpv == sample(patient_id.fpv, size = 1, prob = exp(-abs(age.control - age.fpv))))
  filter(patient_id.fpv == sample(patient_id.fpv, size = 1, prob = 1/(1 + abs(age.control - age.fpv))))

matches_by_source <- matches %>% group_by(patient_id.fpv) %>% 
  summarise(count = n(), .groups = "drop")

if(nrow(matches_by_source) != nrow(fpv_patients)) {
  stop("Some FPV patients figure in no matching")
}
```

```{r}
groups_and_days_fpv <- 
  fpv_patients %>% select(patient_id, min_day_fpv, age, sex) %>%
  rename(exposure_day = min_day_fpv) %>%
  mutate(group = "FPV")

groups_and_days_control <- 
  matches %>% select(patient_id.control, min_day_fpv, age.control, sex.control) %>%
  rename(patient_id = patient_id.control, 
         age = age.control,
         sex = sex.control,
         exposure_day = min_day_fpv) %>%
  mutate(group = "Control")

groups_and_days <- rbind(groups_and_days_fpv, groups_and_days_control)

data_for_model <- data_fpv$marker_data %>%
  inner_join(groups_and_days, by = "patient_id") %>%
  inner_join(markers_for_adverse_effects, by = "marker") %>%
  ungroup() %>%
  mutate(outside_bounds = value < low_bound | value > high_bound) %>%
  group_by(patient_id, group, marker, age, sex) %>%
  summarise(at_risk = any(!outside_bounds[day <= exposure_day]) & any(!is.na(value[day > exposure_day & day < exposure_day + 10])),
            event = at_risk & any(outside_bounds[day >= exposure_day]), .groups = "drop")

data_for_model %>% 
  group_by(marker) %>%
  summarise(n_at_risk = sum(at_risk), n_events = sum(event), .groups = "drop")

data_for_model_all <- data_for_model %>% group_by(patient_id, age, sex, group) %>%
  summarise(at_risk = any(at_risk), event = any(event), .groups = "drop")

data_for_model_all %>% 
  summarise(n_at_risk = sum(at_risk), n_events = sum(event), .groups = "drop")

```

```{r}
fit_adverse_markers <- glm( event ~ age + sex + group, family = "binomial", data = data_for_model_all %>% filter(at_risk))

fpv_ci <-  confint(fit_adverse_markers)["groupFPV", ]
fpv_coeff <- summary(fit_adverse_markers)$coefficients["groupFPV", ]

other_adverse <- tibble::tribble(
  ~event, ~estimate, ~se, ~`lower95`, ~`upper95`,
  "Lab marker out of range", fpv_coeff["Estimate"], fpv_coeff["Std. Error"], fpv_ci[1], fpv_ci[2]
)

other_adverse
write_csv(other_adverse, here::here("local_temp_data", "fpv_meta", "lab_events.csv"))

```

```{r}
all_event_markers <- data_fpv$marker_data %>%
  inner_join(markers_for_adverse_effects, by = "marker") %>%
  mutate(outside_bounds = value < low_bound | value > high_bound) %>%
  group_by()

took_fpv <-  wide_fpv %>% select(patient_id, day,  took_favipiravir) %>%
  crossing(markers_for_adverse_effects)

all_event_markers_fpv <- took_fpv %>% 
  left_join(all_event_markers, by = c("patient_id", "day", "low_bound", "high_bound", "marker")) %>%
  mutate(group = if_else(took_favipiravir, "FPV", "Control"))

if(nrow(all_event_markers_fpv) != nrow(took_fpv)) {
  stop("bad join")
}
```

```{r}
all_res <- list()
for(exposure_day in 1:6) {

  all_events_risk <- all_event_markers_fpv %>%
    group_by(patient_id, marker) %>%
    summarise(at_risk = any(!outside_bounds[day <= exposure_day]) & any(!is.na(value[day > exposure_day & day < exposure_day + 10])),
              event = at_risk & any(outside_bounds[day >= exposure_day]), 
              group = group[day == exposure_day],
              .groups = "drop")

  by_marker <- all_events_risk %>%
    group_by(group, marker) %>%
    summarise(N = n(), at_risk = sum(at_risk, na.rm = TRUE), event = sum(event, na.rm = TRUE), .groups = "drop")

  any_marker <- all_events_risk %>%
    group_by(group) %>%
    summarise(N = length(unique(patient_id)), marker = "Any", at_risk = length(unique(patient_id[at_risk])), event = length(unique(patient_id[event])), .groups = "drop")

  all_res[[exposure_day]] <- rbind(any_marker, by_marker) %>% mutate(day = exposure_day)
}

do.call(rbind, all_res)

```

```{r}
all_event_markers_fpv %>% filter(is.na(group))

all_event_markers_fpv %>% ungroup() %>%
  group_by(day) %>%
  summarise(length(unique(patient_id)), sum(took_favipiravir), sum(is.na(took_favipiravir)), .groups = "drop")

dd <- all_event_markers_fpv %>%
  group_by(patient_id, marker) %>% summarise(at_risk = any(!outside_bounds[day <= exposure_day]) & any(!is.na(value[day > exposure_day & day < exposure_day + 10])), gr = if_else(took_favipiravir[day == exposure_day], "FPV", "Control"))

length(unique(dd$patient_id))
```

## Simple summaries

```{r}
table_fpv <- data_fpv$patient_data %>% 
  mutate(group = if_else(ever_favipiravir, "FPV", "Control")) %>% 
  group_by(group) %>% 
  summarise(
    N = n(),
    Male = sum(sex == "M"),
    `Age mean` = mean(age),
    `Age sd` = sd(age),
    `Age min` = min(age),
    `Age max` = max(age),
    `Admitted for Covid` = sum(admitted_for_covid),
    `Took hydroxychloroquine` = sum(ever_hcq),
    `Took azithromycin` = sum(ever_az),
    `Took dexamethasone` = sum(ever_dexamethasone),
    `Took remdesivir` = sum(ever_remdesivir),
    `Convalescent plasma` = sum(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum(ischemic_heart_disease),
    `Takes antihypertensives` = sum(has_hypertension_drugs),	
    `Heart Failure` = sum(heart_failure),
    COPD = sum(COPD),
    Asthma = sum(asthma),
    `Other lung disease` = sum(other_lung_disease),
    Diabetes = sum(diabetes),
    `Renal Disease` = sum(renal_disease),
    `Liver Disease` = sum(liver_disease),
    `Smoking` = sum(smoking),
    `BMI mean` = mean(BMI, na.rm = TRUE),
    `BMI sd` = sd(BMI, na.rm = TRUE),
    `Best supportive care` = sum(!is.na(best_supportive_care_from)),
    Deceased = sum(outcome == "Death"),
    `Required ventilation` = sum(worst_breathing_s >= "Ventilated"),
    `Required oxygen` = sum(worst_breathing_s >= "Oxygen"),
    Discharged = sum(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  as.data.frame()%>%
  column_to_rownames("group") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("characteristic")

table_fpv

write_csv(table_fpv, here::here("local_temp_data", "fpv_meta", "characteristics.csv"))

```

### 2021-06-04 other_outputs/eosinophiles.Rmd
Exploration of eosinophiles

---
title: "Eosinophile"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = FALSE, fig.width = 8)

devtools::load_all()
library(tidyverse)
library(patchwork)
library(tsiMisc) # devtools::install_github("tsieger/tsiMisc")
theme_set(cowplot::theme_cowplot())
data <- read_data_for_analysis()
wide <- data$marker_data_wide

```

```{r}
wide %>% filter(!is.na(eosinophile_count)) %>% pull(patient_id) %>% unique() %>% length()
wide %>% filter(!is.na(eosinophile_count)) %>% group_by(patient_id) %>% summarise(count = n()) %>% group_by(count) %>% summarise(n())
```

```{r}
wide %>% filter(!is.na(eosinophile_count)) %>% group_by(eosinophile_count == 0) %>% summarise(n())
```

```{r}
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

  if(length(markers) > 1) {
    facet = facet_grid(marker ~ worst_breathing_s, scales = "free_y")
    scale_y_name = "Value"
  } else {
    facet = facet_wrap( ~ worst_breathing_s, ncol = 1)
    unit <- data$marker_data %>% filter(marker == names(markers)) %>% pull(unit) %>% unique()
    scale_y_name <- markers
  }

  data_for_plot <- data$marker_data %>% 
    filter(marker %in% names(markers), day >= 0) %>% 
    group_by(marker, patient_id) %>%
    filter(n() > 1) %>%
    ungroup() %>%
    mutate(marker = markers[marker])

  if(!is.null(max_patients)) {
    patient_markers_to_use <- data_for_plot %>% 
      select(marker, patient_id) %>%
      distinct() %>%
      group_by(marker) %>%
      slice_sample(n = max_patients)

    data_for_plot <- data_for_plot %>%
      semi_join(patient_markers_to_use, by = c("patient_id", "marker"))
  }

  if(log_transform) {
    scale_y <- scale_y_log10(scale_y_name)
    data_for_plot <- data_for_plot %>% 
      group_by(marker) %>%
      mutate(value = if_else(value == 0, min(value[value > 0]) / 2, value)) %>%
      ungroup()
  } else {
    scale_y <- scale_y_continuous(scale_y_name)
  }

  if(smooth) {
    main_geom <- geom_smooth(aes(group = outcome))
  } else {
    main_geom <- geom_line(alpha = 0.5)
  }

  if(!is.null(max_day)) {
    expand <- expand_limits(x = max_day)
  } else {
    expand <- NULL
  }

  data_for_plot %>%
    inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>%
    mutate(outcome = fct_relevel(fct_collapse(outcome, Hospitalized = c("Hospitalized", "Transferred")),
            "Discharged", "Hospitalized", "Death"),
           worst_breathing_s =fct_recode(worst_breathing_s, `Ambient Air` = "AA")) %>%
    ggplot(aes(x = day, y = value,  group = patient_id)) + 
    main_geom +
    facet + 
    expand +
    #scale_color_brewer(type = "qual", palette = 2, guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + 
    scale_y +
    scale_x_continuous("Days since hospitalization") +
    scale_color_discrete(guide = FALSE)
    #scale_color_viridis_d(guide = guide_legend(override.aes = list(alpha = 1, size = 2))) + scale_y
}

max_day <- data$marker_data_wide %>% filter(!is.na(CRP)) %>% pull(day) %>% max()
hide_axis_theme <- theme(axis.text.x = element_blank(),
                         axis.line.x = element_blank(),
                         axis.title.x = element_blank(),
                         axis.ticks.x = element_blank())
hide_strip_theme <- theme(strip.background = element_blank(), strip.text = element_blank())
title_theme <- theme(axis.title.y = element_text(size = 10))

#  (plot_marker_data(c("eosinophile_count" = "Eo"), log_transform = TRUE, max_day = max_day))

fig_eo <-   (plot_marker_data(c("eosinophile_count" = "Eo"), log_transform = FALSE, max_day = max_day))

fig_eo

ggsave(here::here("local_temp_data/fig_eo.png"), plot = fig_eo, width = 8, heigh = 6)

```

```{r}
names(data$marker_data_wide)
```

```{r}
fig_neu <-   (plot_marker_data(c("neutrophile_count" = "Neu"), log_transform = FALSE, max_day = max_day))

fig_neu

ggsave(here::here("local_temp_data/fig_neu.png"), plot = fig_neu, width = 8, heigh = 6)

```

```{r}
eos_age <- wide %>% filter(!is.na(eosinophile_count)) %>%
  group_by(patient_id) %>% summarise(first_eos = eosinophile_count[day == min(day)], max_eos = max(eosinophile_count)) %>% inner_join(data$patient_data, by = "patient_id")

fig_eos_age <- ggplot(eos_age, aes(x = age, y = first_eos, color = worst_breathing_s, shape = worst_breathing_s, group = 1)) + geom_point()  + scale_color_brewer("Worst condition", type = "qual", palette = 2) + scale_shape("Worst condition") + scale_y_continuous("First measured Eos")

fig_eos_age

ggsave(here::here("local_temp_data/fig_eos_age.png"), plot = fig_eos_age, width = 8, heigh = 6)

ggplot(eos_age, aes(x = age, y = max_eos, color = worst_breathing_s, shape = worst_breathing_s)) + geom_point()  + scale_color_brewer(type = "qual", palette = 2)

```

### 2021-06-04 manuscript/hmm.Rmd

@@
 
+```{r}
+# Estimates for FPV for metaanalysis
+hmm_fit <- c_trt_m3_fw_fit
+numerator_serie_data <- serie_data_28_fw %>% mutate(took_favipiravir = 1)
+denominator_serie_data <- serie_data_28_fw %>% mutate(took_favipiravir = 0)
+
+  combined_data <- hmm_fit$data
+  combined_data$serie_data = rbind(
+    numerator_serie_data %>% mutate(.serie = paste0("__num_", .serie), is_numerator = TRUE),
+    denominator_serie_data %>% mutate(.serie = paste0("__denom_", .serie), is_numerator = FALSE)
+  ) %>%
+    mutate(.serie = factor(.serie))
+
+  combined_data$initial_states = rep(hmm_fit$data$initial_states, 2)
+
+  oxygen_states_ids <- which(hmm_fit$data$hidden_state_data$corresponding_obs == "Oxygen")
+  ventilation_states_ids <- which(hmm_fit$data$hidden_state_data$corresponding_obs == "Ventilated")
+
+  numerator_indices <- 1:length(hmm_fit$data$initial_states) + length(hmm_fit$data$initial_states)
+  denominator_indices <- 1:length(hmm_fit$data$initial_states)
+
+  numerator_indices_oxygen <- numerator_indices[hmm_fit$data$initial_states == "AA_worsening"]
+  denominator_indices_oxygen <- denominator_indices[hmm_fit$data$initial_states == "AA_worsening"]
+
+  numerator_indices_ventilated <- numerator_indices[hmm_fit$data$initial_states != "Ventilated_worsening"]
+  denominator_indices_ventilated <- denominator_indices[hmm_fit$data$initial_states != "Ventilated_worsening"]
+
+states_rep <- 4
+states_list <- list()
+for(i in 1:states_rep) {
+  states_list[[i]] <- posterior_epred_rect(hmm_fit, newdata = combined_data, method = posterior_epred_rect_simulate, cores = 6)
+}
+
+states <- do.call(abind::abind, args = c(states_list, list(along = 3)))
+
+  needed_ventilation <- apply(states, MARGIN = c(2,3), FUN = function(x) { any(x %in% ventilation_states_ids) })
+  needed_oxygen <- apply(states, MARGIN = c(2,3), FUN = function(x) { any(x %in% oxygen_states_ids) })
+
+  samples_oxygen_probs_numerator <- 
+    colMeans(needed_oxygen[numerator_indices_oxygen, ])
+  samples_oxygen_probs_denominator <- 
+    colMeans(needed_oxygen[denominator_indices_oxygen, ])
+
+  log_oxygen_odds_numerator <- log(samples_oxygen_probs_numerator) - log1p(-samples_oxygen_probs_numerator)
+  log_oxygen_odds_denominator <- log(samples_oxygen_probs_denominator) - log1p(-samples_oxygen_probs_denominator)
+
+  log_OR_oxygen <- log_oxygen_odds_numerator - log_oxygen_odds_denominator
+
+    samples_ventilation_probs_numerator <- 
+    colMeans(needed_ventilation[numerator_indices_ventilated, ])
+  samples_ventilation_probs_denominator <- 
+    colMeans(needed_ventilation[denominator_indices_ventilated, ])
+
+  log_ventilation_odds_numerator <- log(samples_ventilation_probs_numerator) - log1p(-samples_ventilation_probs_numerator)
+  log_ventilation_odds_denominator <- log(samples_ventilation_probs_denominator) - log1p(-samples_ventilation_probs_denominator)
+
+  log_OR_ventilation <- log_ventilation_odds_numerator - log_ventilation_odds_denominator
+
+  # There is a very tiny fraction of NaNs where no patient is predicted to need ventilation. remove those.
+  log_OR_ventilation <- log_OR_ventilation[is.finite(log_OR_ventilation)]
+
+  death_res <- hmm_hypothesis_res_list[["c_trt_rate_m3_fw"]] %>% filter(hypothesis == "favipiravir_death")
+
+res_fpv <- tibble::tribble(
+  ~event, ~estimate, ~se, ~`lower95`, ~`upper95`,
+  "Death",  death_res$point_estimate, NA, death_res$ci_low, death_res$ci_high,
+  "Oxygen", mean(log_OR_oxygen), sd(log_OR_oxygen), quantile(log_OR_oxygen, 0.025), quantile(log_OR_oxygen, 0.975),
+  "Ventilated", mean(log_OR_ventilation), sd(log_OR_ventilation), quantile(log_OR_ventilation, 0.025, na.rm=TRUE), quantile(log_OR_ventilation, 0.975, na.rm=TRUE)
+  )
+
+write_csv(res_fpv, file = here::here("local_temp_data", "fpv_meta","main_events.csv"))
+```
+
 ### Treatment + age + sex + hospital + comorbidities

### 2021-06-04 manuscript/manuscript.Rmd

@@
 ---
-title: "Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
+title: "Disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic in early 2020: An exploratory analysis" 
 output: 
@@
 abstract: |  
-  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2.
+  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2 during the early phases of the pandemic in March - October 2020.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
@@
 For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period.
-The data collection started at the onset of the Covid-19 pandemic (except for one site where some older records were inaccessible), but the end date for collection differs between sites due to time constraints of the participating physicians.
+The data collection started at the onset of the Covid-19 pandemic in March 2020 (except for one site where some older records were inaccessible), but the end date for collection differs between sites due to time constraints of the participating physicians.
 Three sites included total of `r n_second_wave` patients that could be considered part of "second wave", i.e.

### 2021-06-06 manuscript/manuscript.Rmd
Integrating reviewer comments

@@
 ---
-title: "Disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic in early 2020: An exploratory analysis" 
+title: "Disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic in March - October 2020: An exploratory analysis" 
 output: 
@@
 Patients in the dataset were treated with hydroxychloroquine (N = 108), azithromycin (N = 72), favipiravir (N = 9), convalescent plasma (N = 7), dexamethasone (N = 4) and remdesivir (N = 3), often in combination.
-Most patients were admitted during the first wave of the epidemic.
 To explore association between treatments and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom.
@@
 n_hospitals <- length(unique(data$patient_data$hospital_id))
+
+epidemiology_cz <- read_csv(here::here("public_data", "nakazeni-vyleceni-umrti-testy.csv"), col_types = cols(
+  datum = col_date(format = ""),
+  kumulativni_pocet_nakazenych = col_double(),
+  kumulativni_pocet_vylecenych = col_double(),
+  kumulativni_pocet_umrti = col_double(),
+  kumulativni_pocet_testu = col_double(),
+  kumulativni_pocet_ag_testu = col_double()
+))
+
+epidemiology_cz_october <- epidemiology_cz %>% filter(datum == "2020-10-01")
+
+# hospitalized_cz <- read_csv(here::here("public_data", "hospitalizace.csv"), col_types = cols(
+#   datum = col_date(format = ""),
+#   pacient_prvni_zaznam = col_double(),
+#   kum_pacient_prvni_zaznam = col_double(),
+#   pocet_hosp = col_double(),
+#   stav_bez_priznaku = col_double(),
+#   stav_lehky = col_double(),
+#   stav_stredni = col_double(),
+#   stav_tezky = col_double(),
+#   jip = col_double(),
+#   kyslik = col_double(),
+#   hfno = col_double(),
+#   upv = col_double(),
+#   ecmo = col_double(),
+#   tezky_upv_ecmo = col_double(),
+#   umrti = col_double(),
+#   kum_umrti = col_double()
+# ))
+# 
+# hospitalized_cz_october <- hospitalized_cz %>% filter(datum == "2020-10-01")
+# hospitalized_cz_october$kum_pacient_prvni_zaznam
+
+# epidemiology_cz_october$kumulativni_pocet_nakazenych
+# epidemiology_cz_october$kumulativni_pocet_umrti
 ```
@@
 
-The Covid-19 pandemic caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths.
-Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
-As of this writing, the only treatment that exhibited positive effects in published randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58], Janus kinase inhibitor baricitinib for hospitalized patients[@http://zotero.org/users/5567156/items/PDEPFSAU] and the monoclonal antibody bamlanivimab for non-hospitalized patients [@http://zotero.org/users/5567156/items/TZFUM6JH].
-Other monoclonal antibodies have also announced, but not yet published positive results of randomized trials, including tocilizumab[@http://zotero.org/users/5567156/items/4SLQXUJM] and combination of casirivimab and imdevimab[@http://zotero.org/users/5567156/items/VTR54LI9].
-Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA.
-However, during spring and summer it was available only in limited amounts via compassionate use programme.
-Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3].
-Despite initial promise, hydroxychloroquine has failed to show benefit for hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR].
-Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ].
-Similarly, lopinavir/ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU].
-No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].
+The Covid-19 pandemic caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of June 2021, led to over 172 million cases and over 3.7 million deaths.
+The present study was designed and conducted during March - October 2020, when Czech Republic experienced a relatively mild first wave of the pandemic due to early and strict lockdowns.
+Low numbers of cases continued throughout the summer but during September and October, after most of the data collection for this study concluded, the number of cases was raising again.
+On October 1st 2020, Czech Republic had accumulated `r as.character(epidemiology_cz_october$kumulativni_pocet_nakazenych)` total confirmed Covid-19 cases and `r epidemiology_cz_october$kumulativni_pocet_umrti` confirmed Covid-19 related deaths [@http://zotero.org/users/5567156/items/SWNTWFLU].
 
+At the time the study was conducted, the proposed treatments included antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR].
+Notably, the anti-malarial and anti-rheumatic drug hydroxychloroquine and the macrolide antibiotic  azithromycin showed promise in early data and were broadly available and thus were frequently used in the early stages of the pandemic.
+Remdesivir, previously designed and approved for Ebola, SARS and MERS, also reported good initial results.
+However, during spring and summer 2020 remdesivir was available in Czech Republic only in limited amounts via compassionate use programme.
+The RECOVERY trial reported positive results of croticosteriod dexamethasone for severe cases in June 2020 [@http://zotero.org/users/5567156/items/W8KK3TX2], but at this point the number of Covid-19 patients hospitalized in Czech Republic was low and dexamethasone thus did not see wider use until later in the pandemic.
+
+Our understanding of the efficacy of Covid-19 treatments has improved substantially since the present study was conducted.
+As of April 2021, the pharmacological treatments that were deemed to be beneficial for at least one outcome in a systematic review of randomized trials  were the corticosteroid dexamethasone (mortality, mechanical ventilation), colchicine (mortality, length of hospital stay),  the antiviral remdesivir (mechanical ventilation), Janus kinase inhibitors (mechanical ventilation, duration of ventilation), IL-6 inhibitors (mechanical ventilation, length of hospital stay), the antiviral favipiravir (length of hospital stay, resolution of symptoms) and the anti-androgen proxalutimide (admission to hospital).
+Hydroxychloroquine, interferon beta, lopinavir-ritonavir, azithromycin, vitamin C, vitamin D, anticoagulants and ACE inhibitors  were considered to not be better than standard of care and lopinvair-ritonavir showed evidence of harm, although most of the conclusions were considered to be of low certainty [@http://zotero.org/users/5567156/items/YP62P6XJ].
 Interestingly, in observational studies, hydroxychloroquine was often found to be associated with better outcomes [e.
 g., @http://zotero.org/users/5567156/items/WZPNBRYN; @http://zotero.org/users/5567156/items/4HR7MUJ6; @http://zotero.org/users/5567156/items/NXC9CDI9].
-
-Favipiravir is one of the less explored experimental treatments, previously used for influenza including avian types, with only one larger RCT in Covid- 19 known to us [@http://zotero.org/users/5567156/items/ISLSL7UC], the study however has many potentially severe problems [@http://zotero.org/users/5567156/items/7HDXLWEA].
-Other studies are scarce - even when considering very small studies and case series [@http://zotero.org/users/5567156/items/F3VNTQ92; @http://zotero.org/users/5567156/items/FSUQMUP8; @http://zotero.org/users/5567156/items/894H9N9B].
-None of the results allow strong conclusions about benefits or harms of favipiravir, although a  metaanalysis of all available small studies showed an effect of favipiravir on clinical improvement in mild and moderate disease [@http://zotero.org/users/5567156/items/7595KBWR].
-For additional summary of treatments under investigation see Siemieniuk et al.
-[@http://zotero.org/users/5567156/items/YP62P6XJ].
-There is also an emerging consensus on the importance of prophylactic anticoagulation therapy in managing hospitalized patients with Covid-19, although clinical trials to validate this approach are still ongoing[@http://zotero.org/users/5567156/items/5ZW5JM8R].
+No benefit was also observed in a meta-analysis of randomized trials of convalescent plasma treatment [@http://zotero.org/users/5567156/items/JCV6R3RC].
 
@@
 
-The present study aims to describe the outcomes and disease course of hospitalized patients with mild to severe clinical presentation in a multicentric Czech cohort, explore the association between the outcomes and pharmacological interventions and to provide external validation to previously published prognostic models for Covid-19 severity.
+The present study aims to describe the outcomes and disease course of hospitalized patients with mild to severe clinical presentation in a multicentric Czech cohort during the early stages of the pandemic, explore the association between the outcomes and pharmacological interventions and to provide external validation to previously published prognostic models for Covid-19 severity.
 
-# Patients and Methods
+# Methods
 
-## Data Collection
+## Patients and study design
 
@@
 n_second_wave <- data$patient_data %>% filter(!first_wave) %>% nrow()
+
 ```
@@
 A convenience sample of patients from `r n_hospitals` sites was collected.
+The study sites span the whole spectrum of sizes from large university hospitals in major cities with multiple dedicated Covid-19 wards (Thomayer Hospital in Prague, Motol University Hospital in Prague,  Kralovské Vinohrady University Hospital in Prague, General University Hospital in Prague, University Hospital in Pilsen) through major regional/specialized hospitals  (Na Homolce Hospital in Prague, Military Hospital Olomouc) as well as smaller hospitals caring for just several Covid-19 patients at a time (AGEL Hospital Nový Jičín, Hořovice Hospital, Třebíč Hospital).
+The sites were chosen based on availability and willingness of the personnel to participate in data collection.
+None of the study sites was exclusively dedicated to treating Covid-19 patients.
 For each site, the dataset contains all patients hospitalized in the participating wards over the data collection period.
 The data collection started at the onset of the Covid-19 pandemic in March 2020 (except for one site where some older records were inaccessible), but the end date for collection differs between sites due to time constraints of the participating physicians.
-Three sites included total of `r n_second_wave` patients that could be considered part of "second wave", i.e.
-admitted after September 1st, last patient included in the dataset was admitted on October 12th.
+Three sites included total of `r n_second_wave` patients that could be considered part of "second wave" (admitted after September 1st).
+Last patient included in the dataset was admitted on October 12th.
 See Supplementary Figure 1 for per-site data collection periods.
 Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical trial of any Covid-19 pharmacotherapy.
-
-We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow rate, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
-Full protocol for data collection is attached in the supplementary material.
-Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we merged those categories with mechanical ventilation.
 
@@
 
-The study was approved by the Ethical committees of `r ethics_string`.
+The study was approved by the Ethical committees of `r ethics_string`, all data were collected in fully anonymized form.
+Data was collected between June and October 2020 for patients that were treated between March and October 2020.
+
+## Data collection
+
+We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow rate, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
+Full protocol for data collection is attached in the supplementary material.
+Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we merged those categories with mechanical ventilation.
 
@@
 the association between treatments and patient outcomes,
-because the treatments had not been assigned to patients at random, 
+because the treatments had not been assigned to patients at random 
 but were only observed retrospectively.
@@
 For each model class we worked with several possible sets of adjustments.
-All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualizations and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
+All analyses were performed in the R language [@http://zotero.org/users/5567156/items/67XKBC2W], visualization and data cleaning was run with the `tidyverse` package [@http://zotero.org/users/5567156/items/8EWY8TKM].
 
@@
 The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths.
-This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e.
-that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
+This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e., that the (potentially unobserved) state and the covariates at a given day contain all the information to determine probabilities of the states on the next day.
 We use two versions of such models, one working solely with the observed breathing support and one assuming a hidden improving/worsening distinction.
@@
 
-Finally we used a set of Bayesian regression models implemented with the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
+Finally, we used a set of Bayesian regression models implemented with the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
 Those included overall survival, state at day 7 or 28 as either binary or categorical outcome and a Bayesian version of the Cox proportional-hazards model.
@@
 
-We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for those that could be applied to our dataset (i.e.
-where we have gathered all the input features).
+We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for those that could be applied to our dataset (i.e., where we have gathered all the input features).
 We primarily focused on the Area Under Receiver Operating Characteristic Curve (AUC), and its bootstrapped 95% confidence intervals which we computed using the `pROC` package [@http://zotero.org/users/5567156/items/UN6BIC67].
@@
 # Results
+
+## Baseline characteristics
 
@@
     `Admitted for Covid` = sum_percent(admitted_for_covid),
+    `Took hydroxychloroquine` = sum_percent(ever_hcq),
     `Took azithromycin` = sum_percent(ever_az),
@@
     `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
-    Deceased = sum_percent(outcome == "Death"),
+    Death = sum_percent(outcome == "Death"),
     Discharged = sum_percent(outcome == "Discharged"),
@@
 
+## Disease progression
+
 In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
@@
 Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured.
-However we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.
+However, we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.
 
@@
 
-The results of the multiverse analysis for association between hydroxochloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues (see Supplementary statistical analysis for details).
+The results of the multiverse analysis for association between hydroxychloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues (see Supplementary statistical analysis for details).
 Results for all models we tested are reported in  supplementary Figures `r supplementary_figures_multiverse`, with additional details in supplementary statistical analysis.
@@
 
-# Conclusions
+# Discussion
 
 Our data show the extent of between-patient variability in progression of the disease in terms of both length of hospital stay, duration of various types of breathing support and basic markers.
+A direct comparison with other studies is hard to perform as almost always only summaries of measurements are reported.
+
+For multiple candidate Covid-19 treatments, observational data repeatedly contradicted results of randomized controlled trials (contrast e.g.
+Catteau et al.
+[@http://zotero.org/users/5567156/items/WZPNBRYN] to the RECOVERY trial [@http://zotero.org/users/5567156/items/9X3UH8ZM] for hydroxychloroquine and Liu et al.
+[@http://zotero.org/users/5567156/items/ES7EXX6D] to Agarwal et al.
+[@http://zotero.org/users/5567156/items/UEXLDJCB] for convalescent plasma).
+Our results for hydroxychloroquine also fit into this pattern.
+This should make us wary about over-interpreting the results of this study for azithromycin and favipiravir, although some higher-quality evidence that suggests clinical benefit of favipiravir has been reported  [@http://zotero.org/users/5567156/items/YP62P6XJ].
+
+The current (April 2021) Covid-19 treatment guidelines in Czech Republic recommend monoclonal antibodies and in some cases convalescent plasma or favipiravir as early treatment for high-risk patients with mild or no symptoms.
+For more severe cases dexamethason and anticoagulants are recommended while remdesivir is recommended only for patients that have severe disease but do not require mechanical ventilation [@http://zotero.org/users/5567156/items/YA3UAMVY].
+This is similar to recommendations from the National Institute of Health in the USA who additionally recommend tocilizumab in some cases while not recommending convalescent plasma and favipiravir [@http://zotero.org/users/5567156/items/PCJBPPW7].
+We do not believe our results should directly inform clinical decision, though we see some potential for inclusion of our results in future meta-analyses.
+
+Regarding methodology, there are multiple approaches that are - at least in principle - capable of deriving causal conclusions from observational data, most notably the DAG framework [@http://zotero.org/users/5567156/items/IERDIX65; @http://zotero.org/users/5567156/items/HUSZBEA6] and target trial framework [@http://zotero.org/users/5567156/items/XRTF2VCR; @http://zotero.org/users/5567156/items/WK8XQTZQ].
+In all approaches - and also in some randomized designs - it is common that substantial uncertainty about the best statistical model for the task at hand remains and cannot be eliminated.
+Nevertheless, most published papers present results only from a single statistical model.
+We believe that this uncertainty about model choice should not be ignored, rather we should embrace the uncertainty and employ a multiverse analysis or other forms of robustness checks to explore how our conclusions would differ had different assumptions been adopted.
+In this work we tried to show how such an analysis could be performed and reported in practice.
+We note that modelling choices that are often made semi-arbitrarily, e.g., logistic regression for survival at 28 days vs.
+a Cox proportional hazards model for time to event, did in our case lead to substantially different results.
+
+The hidden-Markov models (HMMs) used in this study are of some interest because they fitted the data well and allowed for inclusion of a larger number of predictors than the Cox proportional hazards model without making the posterior uncertainty unreasonably large.
+We believe this is because such models make better use of the available detailed data.
+Additionally, HMMs can be used even when the outcomes are observed only indirectly or noisily --- as in the original study that motivated our models which concerns the progression of Alzheimer's disease [@http://zotero.org/users/5567156/items/F9K2L5Q8].
+Noisily observed outcomes can pose problems for the proportional hazards model and require some special care [@http://zotero.org/users/5567156/items/YM2LGA3Y; @http://zotero.org/users/5567156/items/RTERDC3X].
+We should however note that the Markov property assumed in HMMs is likely to be a reasonable approximation in fewer settings than the proportional hazards assumption of the Cox model.
+
+Common problems with prognostic models in medicine are small sample sizes used to develop the models, weak or problematic statistical methods and lack of external validation on independent datasets [@http://zotero.org/users/5567156/items/E76N3Y5U].
+Those problems are prevalent also in prediction models for Covid-19 [@http://zotero.org/users/5567156/items/TS57HJ3T].
+We used our dataset to validate several models and observed very poor performance for four out of the five models tested.
+In fact, the simple assumption that older patients tend to have worse outcomes provides better or similar results to all of the models we were able to implement.
+This is despite all of the scores including age as a predictor.
+There seem to be two causes - three of the models dichotomize the age into just two groups which is known to lose information [@http://zotero.org/users/5567156/items/59L7MJPE; @http://zotero.org/users/5567156/items/N4552Y8Q].
+Of the other two models Chen & Liu [@http://zotero.org/users/5567156/items/4DDUVEKS] use stepwise variable selection which is known to be a problematic approch [@http://zotero.org/users/5567156/items/J5VWRSEJ].
+The resulting model puts largest relative weight on laboratory markers and deemphasizes age.
+Caramelo et al.
+[@http://zotero.org/users/5567156/items/3A2RN4ZS] take the decade of age as a very strong predictor and  perform the best on our data.
+Still their results are not distinguishable from just using age.
+Our findings are almost the same as in a similar but larger validation study using 22 models and 411 patients from the United Kingdom where no tested model provided better prediction for mortality than age alone [@http://zotero.org/users/5567156/items/9XLSLVWJ].
+
+# Conclusions
 
@@
 
-A lesson from our analysis is that the assessment of treatment efficacy from observational data is very sensitive to modelling assumptions while it is usually almost impossible to determine which of the models is more likely to reflect reality (if any).
+A lesson from our analysis is that the assessment of treatment efficacy from observational data is sensitive to modelling assumptions while it is usually almost impossible to determine which of the models is more likely to reflect reality (if any).
 We believe that using multiverse analysis is an appropriate way to explore data in such contexts as it lets us be transparent about this sensitivity.
@@
 Additionally, our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial.
-We suggest that comparing the prognostic models against simple baselines (e.g.
-decade of age as the single predictor) should be a first step in validation.
+We suggest that comparing the prognostic models against simple baselines (e.g., decade of age as the single predictor) should be a first step in validation.
 Furthermore, some of the published scores lack enough information to let others implement the score in the same way.

### 2021-06-14 manuscript/manuscript.Rmd
Tiny tweaks before reformatting

@@
 However, during spring and summer 2020 remdesivir was available in Czech Republic only in limited amounts via compassionate use programme.
-The RECOVERY trial reported positive results of croticosteriod dexamethasone for severe cases in June 2020 [@http://zotero.org/users/5567156/items/W8KK3TX2], but at this point the number of Covid-19 patients hospitalized in Czech Republic was low and dexamethasone thus did not see wider use until later in the pandemic.
+The RECOVERY trial reported positive results of coroticosteriod dexamethasone for severe cases in June 2020 [@http://zotero.org/users/5567156/items/W8KK3TX2], but at this point the number of Covid-19 patients hospitalized in Czech Republic was low and dexamethasone thus did not see wider use until later in the pandemic.
 
@@
 A convenience sample of patients from `r n_hospitals` sites was collected.
-The study sites span the whole spectrum of sizes from large university hospitals in major cities with multiple dedicated Covid-19 wards (Thomayer Hospital in Prague, Motol University Hospital in Prague,  Kralovské Vinohrady University Hospital in Prague, General University Hospital in Prague, University Hospital in Pilsen) through major regional/specialized hospitals  (Na Homolce Hospital in Prague, Military Hospital Olomouc) as well as smaller hospitals caring for just several Covid-19 patients at a time (AGEL Hospital Nový Jičín, Hořovice Hospital, Třebíč Hospital).
+The study sites span the whole spectrum of sizes from large university hospitals in major cities with multiple dedicated Covid-19 wards (Thomayer University Hospital in Prague, Motol University Hospital in Prague,  Kralovské Vinohrady University Hospital in Prague, General University Hospital in Prague, University Hospital in Pilsen) through major regional/specialized hospitals  (Na Homolce Hospital in Prague, Military Hospital Olomouc) as well as smaller hospitals caring for just several Covid-19 patients at a time (AGEL Hospital Nový Jičín, Hořovice Hospital, Třebíč Hospital).
 The sites were chosen based on availability and willingness of the personnel to participate in data collection.

### 2021-06-14 manuscript/manuscript.Rmd
Changed formatting to rticles::plos

@@
 title: "Disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic in March - October 2020: An exploratory analysis" 
-output: 
-  #rticles::arxiv_article
-  #pdf_document
-  # bookdown::pdf_document2:
-  #    toc: false
-  bookdown::word_document2: 
-    toc: false
-    number_sections: false
-    reference_docx: "output_template.docx"
 abstract: |  
-  **Abstract:** We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2 during the early phases of the pandemic in March - October 2020.
+  We collected a multi-centric retrospective dataset of patients (N = 213) who were admitted to ten hospitals in Czech Republic and tested positive for SARS-CoV-2 during the early phases of the pandemic in March - October 2020.
 The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution.
@@
 bibliography: Covid19.json
-csl: journal-of-medical-virology.csl
+output: 
+  bookdown::pdf_book:
+    base_format: rticles::plos_article
+    includes:
+      in_header: preamble.tex
+
+  # rticles::plos_article:
+  #   includes:
+  #     in_header: preamble.tex
+csl: plos.csl
 #date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
@@
 ```{r setup, echo=FALSE, message=FALSE, warning = FALSE, results = "hide"}
-knitr::opts_chunk$set(echo = FALSE, fig.width = 8)
+knitr::opts_chunk$set(echo = FALSE, fig.width=8,out.width = "100%", cache=TRUE)
 
@@
 Last patient included in the dataset was admitted on October 12th.
-See Supplementary Figure 1 for per-site data collection periods.
+See S1 Fig  for per-site data collection periods.
 Patients over the age of 18 were included if they had PCR-confirmed infection of SARS-CoV-2 and were not participating in a clinical trial of any Covid-19 pharmacotherapy.
@@
 We collected data on comorbidities and information about disease progression on daily resolution including breathing support required, oxygen flow rate, experimental anti-Covid-19 and antimicrobial drugs taken and several laboratory markers (PCR positivity for SARS-CoV-2, C-reactive protein, D-dimer, Interleukin 6, Ferritin, lymphocyte count).
-Full protocol for data collection is attached in the supplementary material.
+Full protocol for data collection is attached in S1 file and the data collection table in S2 file.
 Due to very low number of patients using extra-corporeal membrane oxygenation (N = `r n_ecmo`) or non-invasive positive pressure ventilation (N = `r n_nippv`) in our sample, we merged those categories with mechanical ventilation.
@@
 First class of models are frequentist survival and multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF].
-We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).
+We primarily use a model with competing risks for death and discharge from hospital (see Fig&nbsp;\@ref(fig:modelstates)a).
 
 Second class of models are Bayesian hidden Markov models (HMM) of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER].
-The parametrization of the HMM is inspired by Williams et al.[@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Figure&nbsp;\@ref(fig:modelstates)b, c).
+The parametrization of the HMM is inspired by Williams et al.[@http://zotero.org/users/5567156/items/F9K2L5Q8]: the actual process of disease is assumed to be continuous and allow only for transitions between neighboring states (as shown in Fig&nbsp;\@ref(fig:modelstates)b, c).
 The total probability of transition between any two states over the period of a day is then computed as the total probability of transition across all possible paths.
@@
 Except for age, sex and comorbidities, all covariates are treated as time-varying, e.g., the effect of taking a drug is only included for the days after the drug was taken.
-More details on the exact model formulations can be found in the supplementary statistical analysis.
+More details on the exact model formulations can be found in the supplementary statistical analysis in S3 file.
 
@@
 In total, we were able to gather data for `r n_patients` patients, see Table \@ref(tab:summarytable) for the overall characteristics of the patient sample and several subgroups we used in the analysis, including treatments taken.
-Counts of all treatment combinations are shown in Supplementary Figure 2 and Supplementary Figure 3 shows outcomes by study site, demonstrating quite large hospital-specific differences.
+Counts of all treatment combinations are shown in S2 Fig and S3 Fig shows outcomes by study site, demonstrating quite large hospital-specific differences.
 The dataset includes 19 patients already reported in a study of inflammatory signatures of Covid-19 [@http://zotero.org/users/5567156/items/IA4FQCF3].
@@
 
-In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
+In Fig&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Fig&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured.
 The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization.
@@
 The between-patient variability is notable even across outcomes - when ordering the patients by the highest CRP levels experienced throughout the hospital stay, the top 20% of patients that breathed ambient air for the whole hospitalization experienced higher levels than the bottom 20% of patients that required ventilation or died.
-This overlap is even larger when comparing only against the patients that died and D-dimer, Interleukin-6 and lymphocyte count also show a notably larger overlap than CRP (Supplementary Figure 4).
+This overlap is even larger when comparing only against the patients that died and D-dimer, Interleukin-6 and lymphocyte count also show a notably larger overlap than CRP (S4 Fig).
 
@@
 The vertical scale is logarithmic.
-Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation), CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
+Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation), CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [$10^9$/l], IL-6 = Interleukin 6 [ng/l]."}
 #data$marker_data %>%
@@
 As noted above, the nature of the convenience sample did not enforce random assignment of treatments to patients.
-In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Figure&nbsp;\@ref(fig:hcqassoc)).
+In fact, patients with worse baseline characteristics, which lead to worse outcomes, were less likely to receive hydroxychloroquine (see Fig&nbsp;\@ref(fig:hcqassoc)).
 This clearly creates a bias towards a positive effect of hydroxychloroquine on the outcome (and potentially for other treatments as well - most were used in combination with hydroxychloroquine), which, however, could be false.
@@
 Comorbidities were associated with both worse outcome (black) and lower chance of treatment with hydroxychloroquine (red).
-Dots and lines represent the estimates and the 95% confidence intervals of the log odds ratio of the respective outcome.
+Dots and lines represent the estimates and the 95\\% confidence intervals of the log odds ratio of the respective outcome.
 HCQ: hydroxychloroquine, IHD: ischemic heart disease, HD: hypertension drugs, HF: heart failure history, COPD: chronic obstructive pulmonary disease, LungD: other lung disease, Dia: diabetes, RenalD: renal disease, LiverD: liver disease, HighCr: creatinin above 115 for males or above 97 for females, HighInr: Prothrombin time (Quick test) as International Normalized Ratio above 1.2, LowAlb: albumin in serum/plasma below 36 g/l."}
@@
 All but one of the patients receiving favipiravir also received hydroxychloroquine.
-Treatment was initiated mostly within two days of admission (see Supplementary Figure 5).
+Treatment was initiated mostly within two days of admission (see S5 Fig).
 
-The results of the multiverse analysis for association between hydroxychloroquine, azithromycin and favipiravir usage and death is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues (see Supplementary statistical analysis for details).
-Results for all models we tested are reported in  supplementary Figures `r supplementary_figures_multiverse`, with additional details in supplementary statistical analysis.
-The results do not change noticeably when only patients from the first wave are included (supplementary Figures `r supplementary_figures_multiverse`).
+The results of the multiverse analysis for association between hydroxychloroquine, azithromycin and favipiravir usage and death is shown in Fig&nbsp;\@ref(fig:multiverseresultsdrugs) - here, we only show models that were not found to have immediate problems representing the data well or computational issues (see S3 file for details).
+Results for all models we tested are reported in  S`r supplementary_figures_multiverse` Figs, with additional details in S3 file.
+The results do not change noticeably when only patients from the first wave are included (S`r supplementary_figures_multiverse` Figs).
 
@@
 We also examined the association between treatments and length of hospital stay for all the patients that survived.
-Almost all models cannot discern the sign of the association for all treatments examined (Supplementary Figures `r supplementary_figures_multiverse`).
-Similarly, we studied the association between D-dimer and Interleukin 6 and outcomes, with unconclusive results as well (Supplementary Figure `r supplementary_figure_markers` )
+Almost all models cannot discern the sign of the association for all treatments examined (S`r supplementary_figures_multiverse` Figs).
+Similarly, we studied the association between D-dimer and Interleukin 6 and outcomes, with unconclusive results as well (S`r supplementary_figure_markers` Fig)
 
-```{r multiverseresultsdrugs, fig.height=7, fig.width=8.5, fig.cap="Estimates of model coefficients for association between treatments and main outcomes.
-Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Figure \\@ref(fig:modelstates)a), HMM A = Bayesian hidden-Markov model as in Figure \\@ref(fig:modelstates)b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Figure \\@ref(fig:modelstates)b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Figure \\@ref(fig:modelstates)c.
-For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+```{r multiverseresultsdrugs, fig.height=7, fig.width=8, fig.cap="Estimates of model coefficients for association between treatments and main outcomes.
+Each row represents a model - Categorical 7/28 = Bayesian categorical regression for state at day 7/28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Fig 1a), HMM A = Bayesian hidden-Markov model as in Fig 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Fig 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Fig 1c.
+For frequentist models, we show maximum likelihood estimate and 95\\% confidence interval, for Bayesian models we show posterior mean and 95\\% credible interval.
 The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
@@
 The outcomes present in the studies were: 12-day mortality, 30-day mortality and mortality without any further details, here we report results for both 12-day and 30-day mortality.
-Full details on the scores and how we used our dataset to compute them is given in the supplementary statistical analysis.
+Full details on the scores and how we used our dataset to compute them is given in the S3 file.
 
-All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)).
+All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Fig&nbsp;\@ref(fig:riskscores)).
 Additionally, some publications did not provide enough detail to unambiguously reconstruct how the score and/or outcome was assessed.
@@
 AUC = 1 means perfect prediction while AUC $\\leq$ 0.5 means that the score is worse than random guess and a better prediction would be obtained by reversing the score (marked by thick blue line).
-The line ranges represent the bootstrapped 95% confidence intervals.
-Red dots show results computed in present study - model variants (horizontal axis) vary in the outcome measured (12-day or 30-day mortality, severe disease) and potentially on how ambiguities in score computation were resolved, although this rarely makes a big difference - see supplementary statistical analysis for details.
+The line ranges represent the bootstrapped 95\\% confidence intervals.
+Red dots show results computed in present study - model variants (horizontal axis) vary in the outcome measured (12-day or 30-day mortality, severe disease) and potentially on how ambiguities in score computation were resolved, although this rarely makes a big difference - see S3 file for details.
 Cyan triangles show AUC as reported by the original authors or recomputed based on their published data.
@@
 
+# Supporting information
+
+* **S1 Fig.
+Data collection timeline.** Data collection periods at individual sites, showing the range of admission dates of patients included in the study.
+Note that we cannot provide additional information to link the sites here with data shown elsewhere as that would  increase the risk of deanonymization of the patients.
+
+* **S2 Fig.
+Treatment combinations.
+** Upset plot of treatment combinations - each vertical bar displays the number of patients that received the combination indicated by filled dots in the matrix.
+Horizontal bars show the total number of patients receiveing the given treatment.
+
+* **S3 Fig.
+Outcomes per site.** Number of patients and outcomes at the individual sites.
+The numbers above bars are the exact counts.
+Hospitalized = still hospitalized at the end of data collection at the site or transferred to other site and lost to followup.
+Sites are anonymized to preserve patient privacy.
+
+* **S4 Fig.
+Markers and outcomes.** Density plots of worst marker values per patient, stratified by worst condition experienced by the patient.
+For each patient that had a given marker measured, the worst value was taken.
+Additionally the patients are classified by the worst condition (regardless of the timing relative to the worst marker levels).
+For each set of patients and marker an empirical density plot of the worst marker values is shown.
+
+* **S5 Fig.
+Treatment onset.** Histogram of timing of first treatment relative to admission into one of the study sites.
+Two patients initiated treatment before admission, which is shown as the negative numbers.
+
+* **S6 Fig.
+Association of HCQ with outcomes.** Estimates of model coefficients for association between hydroxychloroquine and main outcomes.
+The "Suspicious" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.
+Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Fig 1a), HMM A = Bayesian hidden-Markov model as in Fig 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Fig 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Fig 1c.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration).
+In all cases coefficient < 0 means better patient outcome in the treatment group.
+Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
+Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
+= total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C.
+plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included.
+
+* **S7 Fig.
+Association of azithromycin with outcomes.** Estimates of model coefficients for association between azithromycin and main outcomes.
+The "Suspicious" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.
+Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Fig 1a), HMM A = Bayesian hidden-Markov model as in Fig 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Fig 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Fig 1c.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration).
+In all cases coefficient < 0 means better patient outcome in the treatment group.
+Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
+Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
+= total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C.
+plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included.
+
+* **S8 Fig.
+Association of favipiravir with outcomes.** Estimates of model coefficients for association between favipiravir and main outcomes.
+The "Suspicious" section shows models that were found to not fit the data well or have computational issues - see supplementary statistical analysis for details.
+Each row represents a model - Categorical All/7/28 = Bayesian categorical regression for state at last observed day/day 7/day 28, Binary All/7/28 = Bayesian logistic regression for state at last observed day/day 7/day 28, Bayes Cox = Bayesian version of the Cox proportional hazards model with a binary outcome, Cox (single) = frequentist Cox model with a binary outcome, Cox (competing) = frequentist Cox model using competing risks (as in Fig 1a), HMM A = Bayesian hidden-Markov model as in Fig 1b with predictors for rate groups, HMM B = Bayesian hidden-markov model as in Fig 1b with predictors for individual rates, HMM C = Bayesian hidden-Markov model as in Fig 1c.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration).
+In all cases coefficient < 0 means better patient outcome in the treatment group.
+Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
+Additionally the factors the model adjusted for are listed - Site = the study site, admitted = Admitted for Covid-19, Supportive = best supportive care initiated, Comorb.
+= total number of comorbidities, AZ = took azithromycin, HCQ = took hydroxychloroquine, FPV = took favipiravir, C.
+plasma = received convalescent plasma, first wave = only patients admitted before September 1st were included.
+
+* **S9 Fig.
+Association of markers with outcomes.** Estimates of model coefficients (log hazard ratios) for association between markers and death.
+The "Suspicious" section shows models that were found to not fit the data well or have computational issues, "Problematic" section shows models that were completely broken - see supplementary statistical analysis for details.
+Each row represents a model - Cox (competing) =  frequentist Cox model using competing risks (as in Fig 1a), HMM A = Bayesian hidden-markov model as in Fig 1b with predictors for rate groups, JM = Bayesian joint longitudinal and time-to-event model.
+For frequentist models, we show maximum likelihood estimate and 95% confidence interval, for Bayesian models we show posterior mean and 95% credible interval.
+Additionally the factors the model adjusted for are listed - Site = the study site, Supportive = best supportive care initiated, HCQ = took Hydroxychloroquine.
+We show posterior mean and 95% credible interval.
+
+* **S1 File. Data collection protocol.**
+
+* **S2 File. MS Excel table used for data collection.**
+
+* **S3 File. Supplementary statistical analysis. Contains details on all statistical models and procedures used.**
+
 # References

### 2021-07-02 manuscript/hmm.Rmd
Forgotten commit for FPV metaanalysis data

@@
   ~event, ~estimate, ~se, ~`lower95`, ~`upper95`,
-  "Death",  death_res$point_estimate, NA, death_res$ci_low, death_res$ci_high,
+  "Death",  death_res$point_estimate, death_res$sd, death_res$ci_low, death_res$ci_high,
   "Oxygen", mean(log_OR_oxygen), sd(log_OR_oxygen), quantile(log_OR_oxygen, 0.025), quantile(log_OR_oxygen, 0.975),

### 2021-07-02 other_outputs/fpv_metaanalysis.Rmd
Forgotten commit for FPV metaanalysis data

@@
 all_res <- list()
-for(exposure_day in 1:6) {
+for(exposure_day in 0:6) {
 
@@
 
-  all_res[[exposure_day]] <- rbind(any_marker, by_marker) %>% mutate(day = exposure_day)
+  all_res[[exposure_day + 1]] <- rbind(any_marker, by_marker) %>% mutate(day = exposure_day)
 }
 
-do.call(rbind, all_res)
+res_to_write <- do.call(rbind, all_res) %>%
+  select(marker, day, group, N, at_risk, event) %>%
+  arrange(marker, day)
+
+write_csv(res_to_write, here::here("local_temp_data", "fpv_meta", "lab_events_detailed.csv"))
 

### 2021-07-02 manuscript/manuscript.Rmd
Updated data availability statement

@@
 
-The data that support the findings of this study are available on request from the corresponding author, if legal obligations under GDPR can be met.
-The data are not publicly available due to privacy restrictions.
+The data that support the findings of this study are available on request from the corresponding author or the secretariat of the Institute of Microbiology of the Czech Academy of Sciences (contact via mbu@biomed.cas.cz) for researchers who meet the criteria for access to confidential data.
+The data are not publicly available due to privacy restrictions imposed by the Ethical committee of General University Hospital in Prague and the GDPR regulation of the European Union.
 We will be happy to arrange to run any analytical code locally and share the results, provided the code and the results do not leak personal information.

### 2021-07-02 manuscript/manuscript.Rmd
Fixed authors for PLoS

@@
 #date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
-author: "`r { source(here::here('manuscript', 'authors.R')); get_authors_string() }`"
+author: 
+  - name: Martin Modrák
+    email: martin.modrak@biomed.cas.cz
+    corresponding: martin.modrak@biomed.cas.cz
+    affiliation: Institute of Microbiology of the Czech Academy of Sciences
+  - name: Paul-Christian Bürkner
+    affiliation: University of Stuttgart
+  - name: Tomáš Sieger
+    affiliation: cvutCyber
+  - name: Tomáš Slisz
+    affiliation: 
+      - resp1st
+      - thomayer
+  - name: Martina Vašáková
+    affiliation: 
+      - resp1st
+      - thomayer
+  - name: Grigorij Mesežnikov
+    affiliation: Motol
+  - name: Luis Fernando Casas-Mendez
+    affiliation: Motol
+  - name: Jaromír Vajter
+    affiliation: Motol
+  - name: Jan Táborský
+    affiliation: AGEL
+  - name: Viktor Kubricht
+    affiliation: FNKV
+  - name: Daniel Suk
+    affiliation: VFN
+  - name: Jan Horejsek
+    affiliation: VFN
+  - name: Martin Jedlička
+    affiliation: Olomouc
+  - name: Adriana Mifková
+    affiliation: Olomouc
+  - name: Adam Jaroš
+    affiliation: Homolka
+  - name: Miroslav Kubiska
+    affiliation: Pilsen
+  - name: Jana Váchalová
+    affiliation: Pilsen
+  - name: Robin Šín
+    affiliation: Pilsen
+  - name: Markéta Veverková
+    affiliation: Horovice
+  - name: Zbyšek Pospíšil
+    affiliation: Trebic
+  - name: Julie Vohryzková
+    affiliation: charles2nd 
+  - name: Rebeka Pokrievková
+    affiliation: charles3rd
+  - name: Kristián Hrušák
+    affiliation: charles2nd
+  - name: Kristína Christozova
+    affiliation: charles2nd
+  - name: Vianey Leos-Barajas
+    affiliation: toronto
+  - name: Karel Fišer
+    affiliation: bioinfo
+  - name: Tomáš Hyánek
+    affiliation: Homolka
+address:
+  - code: Institute of Microbiology of the Czech Academy of Sciences
+    address: Institute of Microbiology of the Czech Academy of Sciences, Bioinformatics Core Facility, Vídeňská 1083, Prague, Czech Republic, 142 20
+  - code: University of Stuttgart
+    address: University of Stuttgart, Cluster of Excellence SimTech, 	Universitätsstraße 32, Stuttgart, Deutschland, 70569
+  - code: cvutCyber
+    address: Dept. of Cybernetics, Faculty of Electrical Engineering, Czech Technical University in Prague, Technická 1902/2, Prague, Czech Republic, 166 27
+  - code: resp1st
+    address: Department of Respiratory Medicine, 1st Faculty of Medicine, Charles University in Prague, Vídeňská 800, Prague, Czech Republic, 140 59
+  - code: thomayer  
+    address: Thomayer University Hospital, Vídeňská 800, Prague, Czech Republic, 140 59
+  - code: Motol
+    address: Motol University Hospital, V Úvalu 84/1, Prague, Czech Republic, 150 06
+  - code: AGEL
+    address: AGEL Hospital Nový Jičín, Purkyňova 2138/16, Nový Jičín, Czech Republic, 741 01  
+  - code: FNKV
+    address: Kralovské Vinohrady University Hospital, Šrobárova 1150/50, Prague, Czech Republic, 100 34
+  - code: VFN
+    address: General University Hospital in Prague, U Nemocnice 499/2, Prague, Czech Republic, 128 08
+  - code: Olomouc
+    address: Military Hospital Olomouc, Sušilovo nám. 5, Olomouc, Czech Republic, 771 11
+  - code: Homolka
+    address: Na Homolce Hospital, Roentgenova 37, Prague, Czech Republic, 150 00
+  - code: Pilsen
+    address: Department of Infectious Diseases and Travel Medicine, Faculty of Medicine in Pilsen, Charles University, University Hospital in Pilsen, Edvarda Beneše 13, Pilsen, Czech Republic, 305 99
+  - code: Horovice
+    address: Hořovice Hospital, K Nemocnici 1106/14, Hořovice, Czech Republic, 268 01
+  - code: Trebic
+    address: Třebíč Hospital,  Purkyňovo nám. 133, Třebíč, Czech Republic, 674 01
+  - code: charles2nd 
+    address: 2nd Faculty of Medicine, Charles University in Prague, V Úvalu 84, Prague, Czech Republic, 150 06
+  - code: charles3rd
+    address: 3rd Faculty of Medicine, Charles University in Prague, Ruská 2411, Prague, Czech Republic, 100 00
+  - code: toronto
+    address: Department of Statistical Sciences, University of Toronto,  100 St George St, Toronto, Canada, ON M5S 3G3
+  - code: bioinfo
+    address: Department of Bioinformatics, 2nd Faculty of Medicine, Charles University in Prague, V Úvalu 84, Prague, Czech Republic, 150 06
 ---

### 2021-09-03 manuscript/manuscript.Rmd
Fixed bungled math
Forgotten commit
Exporting figures as tiff

@@
 fig_progression
+ggsave(fig_progression, filename = "supp_figs_production/fig2.tiff", width = 7.5)
 ```
@@
 fig_markers
+
+ggsave(fig_markers, filename = "supp_figs_production/fig3.tiff", width = 7.5)
   #plot_marker_data(c("d_dimer" = "D-dimer", "IL_6" = "IL 6", "CRP" = "CRP"), log_transform = TRUE)
@@
 }
-ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
+fig_assoc <- ggplot(gtmp,aes(y=name,x=est,xmin=low,xmax=high,color=type))+
   geom_vline(xintercept=0,color='blue')+
@@
   theme(legend.background = element_rect(fill="gray95"),legend.position=c(.83,.7))
+
+fig_assoc 
+ggsave(fig_assoc, filename = "supp_figs_production/fig4.tiff", width = 7, height = 4.5)
 ```
@@
 The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants).
-In all cases coefficient < 0 means better patient outcome in the treatment group.
+In all cases coefficient $< 0$ means better patient outcome in the treatment group.
 Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).
@@
 
-((multiverse_hcq1 & hide_axis_theme) / (multiverse_az1 & theme(axis.title.x = element_blank())) / multiverse_fpv1)  + plot_layout(heights = c(0.8, 0.7,0.70)) & theme(axis.title.y = element_blank())
+fig_drugs <- ((multiverse_hcq1 & hide_axis_theme) / (multiverse_az1 & theme(axis.title.x = element_blank())) / multiverse_fpv1)  + plot_layout(heights = c(0.8, 0.7,0.70)) & theme(axis.title.y = element_blank())
 
+fig_drugs
+
+ggsave(fig_drugs, filename = "supp_figs_production/fig5.tiff", width = 7.5, height = 7 * (7.5/8))
 ```
@@
 
-compare_auc_to_orig(all_auc_orig,
+fig_risk <- compare_auc_to_orig(all_auc_orig,
                     all_auc, show_outcome = FALSE)
 
+fig_risk
+ggsave(fig_risk, filename = "supp_figs_production/fig6.tiff", width = 7.5)
 ```
@@
 
-# Data availability
-
-The data that support the findings of this study are available on request from the corresponding author or the secretariat of the Institute of Microbiology of the Czech Academy of Sciences (contact via mbu@biomed.cas.cz) for researchers who meet the criteria for access to confidential data.
-The data are not publicly available due to privacy restrictions imposed by the Ethical committee of General University Hospital in Prague and the GDPR regulation of the European Union.
-We will be happy to arrange to run any analytical code locally and share the results, provided the code and the results do not leak personal information.
-
-# Acknowledgements
-
-This work was supported by ELIXIR CZ research infrastructure project (MEYS Grant No: LM2018131) including access to computing and storage facilities.
-
 # Supporting information
@@
 The estimands are either log odds-ratio (Categorical, HMM) or log hazard ratio (Cox variants) or log ratio of mean duration of hospitalization (HMM duration).
-In all cases coefficient < 0 means better patient outcome in the treatment group.
+In all cases coefficient $< 0$ means better patient outcome in the treatment group.
 Vertical lines indicate zero (blue) and substantial increase or decrease with odds or hazard ratio of 3:2 or 2:3 (green).

### 2021-09-03 manuscript/supplementary_figures.Rmd
Exporting figures as tiff

@@
 
-data_collection_period %>% ggplot(aes(xmin = first_admission, xmax = last_admission, y = hospital_id_anon)) + geom_linerange(size = 3) + scale_y_continuous("Study site")+
+s1_fig <- data_collection_period %>% ggplot(aes(xmin = first_admission, xmax = last_admission, y = hospital_id_anon)) + geom_linerange(size = 3) + scale_y_continuous("Study site")+
   scale_x_date("Data collection period", date_breaks = "2 weeks")  + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  )
+
+s1_fig
+ggsave(s1_fig, filename = "supp_figs_production/S1_fig.tiff", width = 7.5, height = 2.5 * (7.5/8))
 ```
@@
 
-UpSetR::upset(upset_data, sets = names(upset_data), text.scale = 2)
+s2_fig <- UpSetR::upset(upset_data, sets = names(upset_data), text.scale = 2)
+s2_fig
+tiff(filename = "supp_figs_production/S2_fig.tiff", width = 7.5, height = 5.8 * (7.5/8), units = "in", res = 300)
+s2_fig
+dev.off()
 ```
@@
 Sites are anonymized to preserve patient privacy.", fig.height=3.5}
-data$patient_data %>% mutate(
+s3_fig <- data$patient_data %>% mutate(
   hospital_label = factor(paste0("H", as.integer(factor(hospital_id))), levels = paste0("H", 1:10)),
@@
   expand_limits(y = 36)
+
+s3_fig
+ggsave(s3_fig, filename = "supp_figs_production/S3_fig.tiff", width = 7.5, height = 3.5 * (7.5/8))
 ```
@@
 frac <- 0.2
-data$marker_data %>% filter(marker %in% c("lymphocyte_count", "d_dimer", "CRP", "IL_6")) %>%
+s4_fig <- data$marker_data %>% filter(marker %in% c("lymphocyte_count", "d_dimer", "CRP", "IL_6")) %>%
   inner_join(data$patient_data, by = c("hospital_id", "patient_id")) %>%
@@
 
+s4_fig
+ggsave(s4_fig, filename = "supp_figs_production/S4_fig.tiff", width = 7.5, height = 6 * (7.5/8))
 #Validating the claim about overlapping proportions
@@
 
-data$marker_data %>% 
+s5_fig <- data$marker_data %>% 
   filter(marker %in% c("az", "hcq", "favipiravir"), !is.na(value), value > 0) %>%
@@
   ggplot(aes(x = first_day)) + geom_histogram(binwidth = 1) + facet_wrap(~marker, scales = "free_y", nrow = 1) + scale_x_continuous("First day of treatment relative to admission")
+
+s5_fig
+ggsave(s5_fig, filename = "supp_figs_production/S5_fig.tiff", width = 7.5, height = 2 * (7.5/8))
 ```
@@
 "}
-plot_multiverse(read_all_multiverse_res() %>% filter(grepl("hcq", hypothesis)), show_check = TRUE)
+s6_fig <- plot_multiverse(read_all_multiverse_res() %>% filter(grepl("hcq", hypothesis)), show_check = TRUE)
+s6_fig
+ggsave(s6_fig, filename = "supp_figs_production/S6_fig.tiff", width = 7.5, height = 11 * (7.5/8))
 ```
@@
 "}
-plot_multiverse(read_all_multiverse_res() %>% filter(grepl("az", hypothesis)), show_check = TRUE)
+s7_fig <- plot_multiverse(read_all_multiverse_res() %>% filter(grepl("az", hypothesis)), show_check = TRUE)
+s7_fig
+ggsave(s7_fig, filename = "supp_figs_production/S7_fig.tiff", width = 7.5, height = 11 * (7.5/8))
 ```
@@
 "}
-plot_multiverse(read_all_multiverse_res() %>% filter(grepl("favipiravir", hypothesis), model_check != "Problematic"), show_check = TRUE)
+s8_fig <- plot_multiverse(read_all_multiverse_res() %>% filter(grepl("favipiravir", hypothesis), model_check != "Problematic"), show_check = TRUE)
+s8_fig
+ggsave(s8_fig, filename = "supp_figs_production/S8_fig.tiff", width = 7.5, height = 11 * (7.5/8))
 ```

### 2021-09-07 manuscript/supplementary_figures.Rmd
More tweaks for production

@@
 We show posterior mean and 95% credible interval."}
-plot_multiverse(read_all_multiverse_res() %>% filter(group == "markers"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"), show_check = TRUE)
+s9_fig <- plot_multiverse(read_all_multiverse_res() %>% filter(group == "markers"), adjustments_to_hide = c("admitted", "smoking", "obesity", "comorbidities \\(sum\\)", "az", "favipiravir", "convalescent_plasma"), show_check = TRUE)
 
+s9_fig
+ggsave(s9_fig, filename = "supp_figs_production/S9_fig.tiff", width = 8.5, height = 5)
 ```
